### 2022-02-21 manuscript/manuscript.tex
First draft of an Abstract

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% LIVECOMS ARTICLE TEMPLATE FOR BEST PRACTICES GUIDE
%%% ADAPTED FROM ELIFE ARTICLE TEMPLATE (8/10/2017)
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% PREAMBLE
\documentclass[9pt,tutorial]{livecoms}
% Use the 'onehalfspacing' option for 1.5 line spacing
% Use the 'doublespacing' option for 2.0 line spacing
% Use the 'lineno' option for adding line numbers.
% Use the "ASAPversion' option following article acceptance to add the DOI and relevant dates to the document footer.
% Use the 'pubversion' option for adding the citation and publication information to the document footer, when the LiveCoMS issue is finalized.
% The 'bestpractices' option for indicates that this is a best practices guide.
% Omit the bestpractices option to remove the marking as a LiveCoMS paper.
% Please note that these options may affect formatting.

\usepackage{lipsum} % Required to insert dummy text
\usepackage[version=4]{mhchem}
\usepackage{siunitx}
\DeclareSIUnit\Molar{M}
\usepackage[italic]{mathastext}
\graphicspath{{figures/}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% IMPORTANT USER CONFIGURATION
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\newcommand{\versionnumber}{1.3}  % you should update the minor version number in preprints and major version number of submissions.
\newcommand{\githubrepository}{\url{https://github.com/myaccount/homegithubrepository}}  %this should be the main github repository for this article

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% ARTICLE SETUP
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{This is the title [Article v\versionnumber]}

\author[1*]{Firstname Middlename Surname}
\author[1,2\authfn{1}\authfn{3}]{Firstname Middlename Familyname}
\author[2\authfn{1}\authfn{4}]{Firstname Initials Surname}
\author[2*]{Firstname Surname}
\affil[1]{Institution 1}
\affil[2]{Institution 2}

\corr{email1@example.com}{FMS}  % Correspondence emails.  FMS and FS are the appropriate authors initials.
\corr{email2@example.com}{FS}

\orcid{Author 1 name}{AAAA-BBBB-CCCC-DDDD}
\orcid{Author 2 name}{EEEE-FFFF-GGGG-HHHH}

\contrib[\authfn{1}]{These authors contributed equally to this work}
\contrib[\authfn{2}]{These authors also contributed equally to this work}

\presentadd[\authfn{3}]{Department, Institute, Country}
\presentadd[\authfn{4}]{Department, Institute, Country}

\blurb{This LiveCoMS document is maintained online on GitHub at \githubrepository; to provide feedback, suggestions, or help improve it, please visit the GitHub repository and participate via the issue tracker.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% PUBLICATION INFORMATION
%%% Fill out these parameters when available
%%% These are used when the "pubversion" option is invoked
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\pubDOI{10.XXXX/YYYYYYY}
\pubvolume{<volume>}
\pubissue{<issue>}
\pubyear{<year>}
\articlenum{<number>}
\datereceived{Day Month Year}
\dateaccepted{Day Month Year}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% ARTICLE START
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

\begin{frontmatter}
\maketitle

\begin{abstract}
Grid inhomogeneous solvation theory (GIST) is a method to compute the free energy of hydration of a compound on a 3-dimensional grid. The high spacial resolution of the GIST output, as well as the decomposition into energy and entropy contributions, allow for highly detailed analyses on both proteins and small molecules. However, this versatility also comes with a higher entry barrier for new users.

In this tutorial, we aim to guide the reader through the most common steps involved in a GIST analysis at the example of the streptavidin-biotin complex. Furthermore, we discuss the theory of GIST with a focus on practical aspects, and show several pitfalls and technical difficulties that may occur during a GIST study. We assume familiarity with molecular dynamics (MD) simulations as well as the AmberTools package.

%This particular document provides a skeleton illustrating key sections for a Tutorial document.
%Please see the sample \texttt{sample-document.tex} in \url{github.com/livecomsjournal/article_templates/templates} for additional information on and examples of using the LiveCoMS LaTeX class.
%Here we also assume familiarity with LaTeX and knowledge of how to include figures, tables, etc.; if you want examples, see the sample just referenced.
%
%In your work, in this particular slot, please provide an abstract of no more than 250 words.
%Your abstract should explain the main contributions of your article, and should not contain any material that is not included in the main text.
%Please note that your abstract, plus the authorship material following it, must not extend beyond the title page or modifications to the LaTeX class will likely be needed.
\end{abstract}

\end{frontmatter}

\section{Introduction}

Here you would explain what problem you are tackling and briefly motivate your work.

In this particular template, we have removed most of the usage examples which occur in \texttt{sample-document.tex} to provide a minimal template you can modify; however, we retain a couple of examples illustrating more unusual features of our templates/article class, such as the checklists, and information on algorithms and pseudocode.

Keep in mind, as you prepare your manuscript, that you should plan for a representative image  which will be used to highlight your article on the journal website and publications. Usually, this would be one of your figures, but it must also be uploaded separately upon article submission. We give specific guidelines for this image on the journal website in the section on article submission (see \url{https://livecomsjournal.github.io/authors/policies/index.html#article-submission}).

Additionally, for well-formatted manuscripts, we recommend that you let LaTeX handle figure/table placement for you as much as possible, so please avoid specifying strenuous float instructions like `[h!]` and `[H]` as much as possible.

\subsection{Scope}

Tutorials should endeavor to cover the specific task at hand, and also highlight how the steps might need to be modified (or additional care might need to be taken at particular points) to handle more general cases.

The scope of the tutorial, as well as the expected proficiencies / outcomes for researchers who complete the tutorial, should be clearly defined.
This will often happen in a specific section or subsection in the article itself.

\section{Prerequisites}

Here you would identify prerequisites/background knowledge that are assumed by your work, as well as any software/license requirements.

\subsection{Background knowledge}
Tutorials should clearly define what concepts or abilities researchers will need to complete the tutorial (e.g., some proficiency in Python; experience with Jupyter notebooks; knowledge of classical MD; etc).

\subsection{Software/system requirements}
Tutorials should clearly define what system and/or software requirements the researcher will need to complete the tutorial (e.g., VMD version 1.9 or newer, AMBER, etc.). Tutorials requiring specific software packages must provide instructions and files for the referenced version of the software.

\section{Content and links}

A tutorial will normally draw on additional files and materials; clearly indicate where and how these are available, with links, and how they are being archived for the long-term and maintained so they stay current.
You will likely want to reference your GitHub repository as a central point to access all of this information, and then the GitHub repository may link out to other content as needed.

\section{Checklists}
Tutorials do not necessarily require the use of a checklist as in Best Practices documents; however, they can include these if desired.
Several useful checklist formats are available, with examples presented in \texttt{sample-document.tex} in \url{github.com/livecomsjournal/article_templates/templates}.
One example is shown here.

% Here is a single-column checklist that consists of multiple sub-checklists
\begin{Checklists}

\begin{checklist}{A list}
\textbf{Single-column checklists are also straightforward by removing the asterisk}
\begin{itemize}
\item First thing let's do an item which breaks across lines to see how that looks
\item Also remember
\item And finally
\end{itemize}
\end{checklist}

\begin{checklist}{Another list}
\textbf{This is some further description.}
\begin{itemize}
\item First thing
\item Also remember
\item And finally
\end{itemize}
\end{checklist}

\end{Checklists}

\section*{Author Contributions}
%%%%%%%%%%%%%%%%
% This section mustt describe the actual contributions of
% author. Since this is an electronic-only journal, there is
% no length limit when you describe the authors' contributions,
% so we recommend describing what they actually did rather than
% simply categorizing them in a small number of
% predefined roles as might be done in other journals.
%
% See the policies ``Policies on Authorship'' section of https://livecoms.github.io
% for more information on deciding on authorship and author order.
%%%%%%%%%%%%%%%%

(Explain the contributions of the different authors here)

% We suggest you preserve this comment:
For a more detailed description of author contributions,
see the GitHub issue tracking and changelog at \githubrepository.

\section*{Other Contributions}
%%%%%%%%%%%%%%%
% You should include all people who have filed issues that were
% accepted into the paper, or that upon discussion altered what was in the paper.
% Multiple significant contributions might mean that the contributor
% should be moved to authorship at the discretion of the a
%
% See the policies ``Policies on Authorship'' section of https://livecoms.github.io for
% more information on deciding on authorship and author order.
%%%%%%%%%%%%%%%

(Explain the contributions of any non-author contributors here)
% We suggest you preserve this comment:
For a more detailed description of contributions from the community and others, see the GitHub issue tracking and changelog at \githubrepository.

\section*{Potentially Conflicting Interests}
%%%%%%%
%Declare any potentially competing interests, financial or otherwise
%%%%%%%

Declare any potentially conflicting interests here, whether or not they pose an actual conflict in your view.

\section*{Funding Information}
%%%%%%%
% Authors should acknowledge funding sources here. Reference specific grants.
%%%%%%%
FMS acknowledges the support of NSF grant CHE-1111111.

\section*{Author Information}
\makeorcid

\bibliography{livecoms-sample}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% APPENDICES
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%\appendix

\end{document}

### 2022-02-21 manuscript/manuscript.tex
Fix a typo

@@
 Grid inhomogeneous solvation theory (GIST) is a method to compute the free energy of hydration of a compound on a 3-dimensional grid.
-The high spacial resolution of the GIST output, as well as the decomposition into energy and entropy contributions, allow for highly detailed analyses on both proteins and small molecules.
+The high spatial resolution of the GIST output, as well as the decomposition into energy and entropy contributions, allow for highly detailed analyses on both proteins and small molecules.
 However, this versatility also comes with a higher entry barrier for new users.

### 2022-02-21 manuscript/manuscript.tex
Add newlines in Abstract
I plan to format the TeX source code with newlines after every sentence.
This will make Git merges easier compared to one line per paragraph, or
breaking after a fixed line width.
Comment out most of the dummy text, first Theory paragraph, bibliography

@@
 
+%%% Shortcuts and macros
+\newcommand{\dgsolv}{\Delta G_\textup{solv}}
+\newcommand{\software}{\emph}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-Here you would explain what problem you are tackling and briefly motivate your work.
-
-In this particular template, we have removed most of the usage examples which occur in \texttt{sample-document.tex} to provide a minimal template you can modify; however, we retain a couple of examples illustrating more unusual features of our templates/article class, such as the checklists, and information on algorithms and pseudocode.
-
-Keep in mind, as you prepare your manuscript, that you should plan for a representative image  which will be used to highlight your article on the journal website and publications.
+%Here you would explain what problem you are tackling and briefly motivate your work.
+%
+%In this particular template, we have removed most of the usage examples which occur in \texttt{sample-document.tex} to provide a minimal template you can modify; however, we retain a couple of examples illustrating more unusual features of our templates/article class, such as the checklists, and information on algorithms and pseudocode.
+%
+%Keep in mind, as you prepare your manuscript, that you should plan for a representative image  which will be used to highlight your article on the journal website and publications.
 Usually, this would be one of your figures, but it must also be uploaded separately upon article submission.
 We give specific guidelines for this image on the journal website in the section on article submission (see \url{https://livecomsjournal.github.io/authors/policies/index.html#article-submission}).
-
-Additionally, for well-formatted manuscripts, we recommend that you let LaTeX handle figure/table placement for you as much as possible, so please avoid specifying strenuous float instructions like `[h!]` and `[H]` as much as possible.
+%
+%Additionally, for well-formatted manuscripts, we recommend that you let LaTeX handle figure/table placement for you as much as possible, so please avoid specifying strenuous float instructions like `[h!]` and `[H]` as much as possible.
 
@@
 
-Tutorials should endeavor to cover the specific task at hand, and also highlight how the steps might need to be modified (or additional care might need to be taken at particular points) to handle more general cases.
-
-The scope of the tutorial, as well as the expected proficiencies / outcomes for researchers who complete the tutorial, should be clearly defined.
-This will often happen in a specific section or subsection in the article itself.
+%Tutorials should endeavor to cover the specific task at hand, and also highlight how the steps might need to be modified (or additional care might need to be taken at particular points) to handle more general cases.
+%
+%The scope of the tutorial, as well as the expected proficiencies / outcomes for researchers who complete the tutorial, should be clearly defined.
+%This will often happen in a specific section or subsection in the article itself.
 
@@
 
-Here you would identify prerequisites/background knowledge that are assumed by your work, as well as any software/license requirements.
+%Here you would identify prerequisites/background knowledge that are assumed by your work, as well as any software/license requirements.
 
 \subsection{Background knowledge}
-Tutorials should clearly define what concepts or abilities researchers will need to complete the tutorial (e.g., some proficiency in Python; experience with Jupyter notebooks; knowledge of classical MD; etc).
+%Tutorials should clearly define what concepts or abilities researchers will need to complete the tutorial (e.g., some proficiency in Python; experience with Jupyter notebooks; knowledge of classical MD; etc).
 
 \subsection{Software/system requirements}
-Tutorials should clearly define what system and/or software requirements the researcher will need to complete the tutorial (e.g., VMD version 1.9 or newer, AMBER, etc.).
+%Tutorials should clearly define what system and/or software requirements the researcher will need to complete the tutorial (e.g., VMD version 1.9 or newer, AMBER, etc.).
 Tutorials requiring specific software packages must provide instructions and files for the referenced version of the software.
 
-\section{Content and links}
+\section{Theory}
+GIST is an implementation of Inhomogeneous Solvation Theory (IST) \cite{Lazaridis1998} that calculates the free energy of solvation $\dgsolv$ on a 3-dimensional grid.
+It was first devised by Nguyen et al. \cite{Nguyen2012} and its implementation in \software{cpptraj} was thoroughly described in \cite{Ramsey2016}.
+Here, we only provide a short overview of the theory behind GIST.
+For more detailed information, we recommend one of the more recent publications on developments in GIST. \cite{Kraml2020}\cite{Chen2021}
+\section{GIST study of streptavidin/biotin}
 
-A tutorial will normally draw on additional files and materials; clearly indicate where and how these are available, with links, and how they are being archived for the long-term and maintained so they stay current.
-You will likely want to reference your GitHub repository as a central point to access all of this information, and then the GitHub repository may link out to other content as needed.
+%\section{Content and links}
+%
+%A tutorial will normally draw on additional files and materials; clearly indicate where and how these are available, with links, and how they are being archived for the long-term and maintained so they stay current.
+%You will likely want to reference your GitHub repository as a central point to access all of this information, and then the GitHub repository may link out to other content as needed.
 
 \section{Checklists}
-Tutorials do not necessarily require the use of a checklist as in Best Practices documents; however, they can include these if desired.
-Several useful checklist formats are available, with examples presented in \texttt{sample-document.tex} in \url{github.com/livecomsjournal/article_templates/templates}.
-One example is shown here.
+%Tutorials do not necessarily require the use of a checklist as in Best Practices documents; however, they can include these if desired.
+%Several useful checklist formats are available, with examples presented in \texttt{sample-document.tex} in \url{github.com/livecomsjournal/article_templates/templates}.
+%One example is shown here.
 
@@
 \begin{Checklists}
+
+\begin{checklist}{Simulation settings}
+\end{checklist}
+
+\begin{checklist}{Obtaining absolute $\Delta G_\textup{solv}$ }
+\end{checklist}
 
@@
 \end{checklist}
-
-\begin{checklist}{Another list}
-\textbf{This is some further description.}
-\begin{itemize}
-\item First thing
-\item Also remember
-\item And finally
-\end{itemize}
-\end{checklist}
+%
+%\begin{checklist}{Another list}
+%\textbf{This is some further description.}
+%\begin{itemize}
+%\item First thing
+%\item Also remember
+%\item And finally
+%\end{itemize}
+%\end{checklist}
 
@@
 
-\bibliography{livecoms-sample}
+\bibliography{bibliography}
 

### 2022-03-09 manuscript/manuscript.tex

@@
 \newcommand{\software}{\emph}
+\newcommand{\todo}{\textcolor{red}}
+
+\newenvironment{code}{\par \noindent \ttfamily}{}
 
@@
 \subsection{Scope}
+The aim of this tutorial is to provide an example of a GIST workflow that can be easily adapted towards different systems.
+We present a GIST study of the streptavidin-biotin complex, which allows us to show many common aspects of GIST studies.
+We present solvation analysis of a small molecule (biotin) as well as a biomolecule (streptavidin).
+For both systems, we show how to interpret the three-dimensional contributions of solvation free energy in a binding pocket.
+Furthermore, we compute the contribution of solvation free energy to binding, which requires accurate post-processing of the GIST output to avoid unfavorable summation of bias in the GIST calculation.
+We therefore believe that this example study covers a wide range of applications of the GIST method.
 
+Furthermore, we discuss several technical aspects of GIST analyses, such as the normalization of voxel values and how to deal with double counting of energy contributions. 
 %Tutorials should endeavor to cover the specific task at hand, and also highlight how the steps might need to be modified (or additional care might need to be taken at particular points) to handle more general cases.
@@
 \section{GIST study of streptavidin/biotin}
+
+Here, we aim to guide the reader through a GIST analysis of the streptavidin-biotin complex. The goal is to
+\begin{enumerate}
+	\item compute the binding free energy of the complex from the individual energy contributions combined with the contribution of hydration computed using GIST.
+	\item visualize local contributions to $\dgsolv$ in the streptavidin binding pocket as well as around biotin.
+\end{enumerate}
+
+\subsection{Streptavidin/Biotin}
+\subsection{Tutorial data}
+We will use the 1STP crystal structure to start our calculations.
+For the sake of simplicity, you can download prepared and solvated Amber topologies and structures based on 1STP by running
+\begin{code}
+git clone git@github.com:liedllab/gist-tutorial.git
+\end{code}
+This will download the tutorial files (including both the manuscript and the examples in the \software{code} folder).
+\subsection{System Preparation and Equilibration}
+When calculating solvation free energy differences between a complex and the corresponding monomers (the dissociated state), there are two options for dealing with configurational diversity in the structures.
+\todo{Maybe Valentin knows a citation for that?}
+One can choose to sample the complex and dissociated states individually, or to sample only one of them and assume that there is sufficient overlap in the sampled conformations.
+%One can choose to sample only the complex, assuming that all relevant conformations of the dissociated state are also found in the complex.
+%The probability distribution of the dissociated state can be obtained by reweighting the probabilities in the complex using the solvation free energies.
+%Alternatively, one can sample the complex and dissociated states individually.
+%This approach does not require an overlap between the conformational probabilities of the complex and dissociated states.
+%However, it is prone to statistical errors due to uncertainties in the ensembles, and requires the free energy contributions from the internal degrees of freedom of the monomers (in both the complex and the monomer states) to be known.
+\todo{This paragraph does not fit here well.
+Maybe put it somewhere else.}
+
+Here, we choose not to include any sampling of the conformational states.
+It is therefore important that the conformations of the monomers in the complex and dissociated states match exactly.
+We provide a script called \software{cpptraj\_remove\_mol.sh} that removes a single molecule from a topology and structure using \software{cpptraj} without altering the rest of the system.
+Use this to create monomer topologies of biotin and streptavidin.
+
+\subsection{Running GIST}
+\subsection{Visualizing $\dgsolv$}
+\subsection{Contribution of Hydration to Binding}
 

### 2022-03-09 manuscript/manuscript.tex
More text for system preparation/equil in the Tutorial

@@
 
-\newenvironment{code}{\par \noindent \ttfamily}{}
+\newenvironment{code}{\par \noindent \ttfamily \ignorespaces}{\par}
+
+\DeclareSIUnit{\calorie}{cal}
+\DeclareSIUnit{\kcalPerMolASqr}{\kilo\calorie\per\mole\per\angstrom\squared}
 
@@
 \subsection{System Preparation and Equilibration}
+If you want to skip this section, run 
+\begin{code}
+	make equilibration-targets
+\end{code}
+in the code folder. This assumes that you have Python and Amber set up properly and that \software{pmemd.cuda} is in the PATH.
+
 When calculating solvation free energy differences between a complex and the corresponding monomers (the dissociated state), there are two options for dealing with configurational diversity in the structures.
@@
 We provide a script called \software{cpptraj\_remove\_mol.sh} that removes a single molecule from a topology and structure using \software{cpptraj} without altering the rest of the system.
-Use this to create monomer topologies of biotin and streptavidin.
+\todo{Explain cpptraj rather than our script.
+The script is for the automated workflow only.}
+
+Use the \software{equilibration.py} script (or your own) to perform short NVT and NpT equilibration runs on the starting structure.
+Note that the barostat might interfere with restrained equilibration of the complex, since different molecules are moved relative to each other when the volume changes.
+But since our equilibration script also contains unrestrained pressure equilibration, we should not run into any trouble.
+
+Then split the equilibrated structure into biotin and streptavidin parts using \software{cpptraj\_remove\_mol.sh}.
+After that, you can use \software{equilibration.py -R} to run short restrained equilibrations (where only the water is allowed to move) on the individual systems.
 
 \subsection{Running GIST}
+
+Now, you can run restrained MD simulations based on the equilibrated structures for subsequent GIST analyses.
+We apply a harmonic restraint of \SI{100}{\kilo\calorie\per\mole\per\angstrom\squared} to all heavy atoms to keep the system in place.
+We run \SI{100}{\nano\second} of simulation and store the conformations every \SI{100}{\pico\second}.
+A simulation length of at least \SIrange{10}{20}{\nano\second} is recommended, to obtain statistically independent snapshots of the water.
+
 \subsection{Visualizing $\dgsolv$}

### 2022-03-16 manuscript/manuscript.tex
Add more tutorial text, switch to lstlistings
* dont use the code environment any more, use lstlistings instead
* split subsection: Running MD for GIST and Running the GIST analysis

@@
 \graphicspath{{figures/}}
+
+\usepackage{listings}
 
@@
 
-\newenvironment{code}{\par \noindent \ttfamily \ignorespaces}{\par}
+\lstdefinestyle{code}{
+	basicstyle=\ttfamily
+}
+\lstset{style=code}
 
@@
 For the sake of simplicity, you can download prepared and solvated Amber topologies and structures based on 1STP by running
-\begin{code}
+\begin{lstlisting}
 git clone git@github.com:liedllab/gist-tutorial.git
-\end{code}
+\end{lstlisting}
 This will download the tutorial files (including both the manuscript and the examples in the \software{code} folder).
@@
 If you want to skip this section, run 
-\begin{code}
+\begin{lstlisting}
 make equilibration-targets
-\end{code}
+\end{lstlisting}
 in the code folder. This assumes that you have Python and Amber set up properly and that \software{pmemd.cuda} is in the PATH.
@@
 It is therefore important that the conformations of the monomers in the complex and dissociated states match exactly. 
-We provide a script called \software{cpptraj\_remove\_mol.sh} that removes a single molecule from a topology and structure using \software{cpptraj} without altering the rest of the system.
+Use \software{cpptraj} to produce biotin and streptavidin structures based on the complex by stripping the respective other molecule like this:
+\begin{lstlisting}
+parm complex/solvated.parm7
+trajin complex/equil/EQUIL-DONE.rst
+strip ^2 parmout streptavidin/solvated.parm7
+trajout streptavidin/pre-equil.ncrst
+\end{lstlisting}
+(Adapt for biotin as necessary.)
+
+%We provide a script called \software{cpptraj\_remove\_mol.sh} that removes a single molecule from a topology and structure using \software{cpptraj} without altering the rest of the system.
 \todo{Explain cpptraj rather than our script.
@@
 
-\subsection{Running GIST}
+\subsection{Running MD for GIST}
 
@@
 
+An example Amber input file for GIST might look like this:
+\begin{lstlisting}
+
+\end{lstlisting}
+\subsection{Running the GIST analysis}
+
+For easier post-processing, we will use grids that contain the whole molecule.
+This could in theory be avoided for streptavidin and the complex, since only the surrounding of the biotin molecule (i.e., the binding pocket) is required for the analysis.
+We also recommend that you center the molecule in space.
+For each system (biotin, streptavidin, complex), produce a centered pdb file using \software{cpptraj}:
+\begin{lstlisting}
+parm ../solvated.parm7
+trajin EQUIL-DONE.rst
+autoimage !(:WAT) origin
+strip :WAT
+trajout equil-centered.pdb
+\end{lstlisting}
+Then, find the extent (minimum and maximum) of the x-, y-, and z-coordinates for each system.
+This can be done e.g., using the \software{mdtraj} package in Python.
+\begin{lstlisting}[basicstyle=\ttfamily]
+import mdtraj as md
+pdb = md.load("equil-centered.pdb")
+min = pdb.xyz[0].min(0) * 10.
+max = pdb.xyz[0].max(0) * 10.
+print(*list(zip("xyz", min, max)), sep="\n")
+\end{lstlisting}
+Then, decide on grid dimensions for every system.
+For a system centered at the origin with a spacing of \SI{0.5}{\angstrom} and a buffer of, e.g., \SI{7}{\angstrom} between the solute and the grid wall, you need $4.0\times (d+7)$ bins in the respective direction, where $d$ is the larger of $-$min and $+$max in the respective direction.
 \subsection{Visualizing $\dgsolv$}

### 2022-03-17 manuscript/manuscript.tex
Make code smaller
Update tutorial ("gist" command) and the python styling

@@
 
-\lstdefinestyle{code}{
-	basicstyle=\ttfamily
+\lstset{
+	basicstyle=\ttfamily\footnotesize
+}
+\lstdefinestyle{code}{}
+\lstdefinestyle{python}{
+	language=Python,
+	stringstyle=\color[rgb]{1.0, 0, 0},
+	commentstyle=\color[rgb]{0, 0, 0.6}
 }
@@
 \section{Prerequisites}
+Before doing this tutorial, the reader should have a solid background knowledge of molecular dynamics (MD) simulations, as well as basic knowledge of Python, and should be familiar with working on the command line.
 
+A recent version of \software{cpptraj} should be used, such as the one in AmberTools22 or an installation from GitHub.
+Older versions of the Amber simulation engine can be used as well.
+If no Amber installation is available, the MD can be performed using Gromacs.
+In that case, we recommend preparing the structures and topologies using AmberTools and then converting to Gromacs using \software{acpype.py}.
+In that way, a full Amber topology is available for the GIST analysis.
+
+Furthermore, a recent Python version (>3.6) should be available with the following packages.
+
+To run the MD simulation, a PC with recent GPU is recommended.
+If the reader prefers to skip the MD and GIST calculation, we also provide the GIST output files with this tutorial, such that the post-processing can be done without any expensive calculations.
+
+\begin{itemize}
+	\item \software{gisttools} (and its prerequisites)
+	\item \software{mdtraj}
+\end{itemize}
 %Here you would identify prerequisites/background knowledge that are assumed by your work, as well as any software/license requirements.
@@
 \begin{lstlisting}
+restrained 100 ns NpT
+&cntrl
+  ntx=5, irest=1,
+  ioutfm=1,
+  ntb=2, iwrap=1,
+  ntr=1, restraint_wt=100.0,
+    restraintmask=\"!@H=&!:WAT\",
+  ntp=1, pres0=1.0, taup=1.0,
+  ntc=2, ntf=2,
+  ntt=3, tempi=300.0, temp0=300.0, gamma_ln=2,
+  nstlim=50000000, dt=0.002,
+  ntwr=50000, ntwx=5000,
+  ntpr=5000,
+/
+\end{lstlisting}
+Note that these are very standard MD settings except for the line that defines the restraints.
 
+Run the MD using \software{pmemd.cuda}:
+\begin{lstlisting}
+pmemd.cuda \
+	-O \
+	-i 100ns-npt-restraint.in \
+	-o md-01.out \
+	-p solvated.parm7 \
+	-c EQUIL-DONE.rst \
+	-x md-01.nc \
+	-r md-01.ncrst \
+	-ref EQUIL-DONE.rst
 \end{lstlisting}
@@
 \begin{lstlisting}
-parm ../solvated.parm7
-trajin EQUIL-DONE.rst
+parm solvated.parm7
+trajin md-01.nc 1 1 1
 autoimage !(:WAT) origin
 strip :WAT
-trajout equil-centered.pdb
+trajout solute-centered.pdb
 \end{lstlisting}
@@
 This can be done e.g., using the \software{mdtraj} package in Python.
-\begin{lstlisting}[basicstyle=\ttfamily]
+\begin{lstlisting}[style=python]
 import mdtraj as md
-pdb = md.load("equil-centered.pdb")
+pdb = md.load("solute-centered.pdb")
+# mdtraj converts to nm.
+# *10 to convert back to Angstrom.
 min = pdb.xyz[0].min(0) * 10.
@@
 For a system centered at the origin with a spacing of \SI{0.5}{\angstrom} and a buffer of, e.g., \SI{7}{\angstrom} between the solute and the grid wall, you need $4.0\times (d+7)$ bins in the respective direction, where $d$ is the larger of $-$min and $+$max in the respective direction.
+We note that this centered grid is usually larger than it needs to be in some directions.
+If you want to save some extra memory and calculation time, feel free to align the grid more precisely to the minimum and maximum coordinates, or use a smaller grid that only covers the binding pocket.
+You can also rotate the molecule by its principal axes to fit the cuboid grid more exactly.
+However, this needs to be done \emph{before} the MD run, since rotating the trajectory damages the periodic box information.
+
+Now, you can finally run the actual GIST analysis. Center the molecule exactly in the same way as above, to make sure that its position matches the grid parameters that you decided on.
+We recommend using either the PME or GPU implementation of GIST, depending on whether you have a fast GPU.
+A further advantage of the PME implementation is that it matches the energy calculation of the MD engine more closely.
+\todo{At the time of writing, it is important to use an up-to-date version of PME-GIST, or the Esw-Eww decomposition will not be available.
+Maybe remove this for the final version, when Amber22 is out.}
+\begin{lstlisting}
+parm solvated.parm7
+trajin md-01.nc
+autoimage !(:WAT) origin
+# remove the "pme" option if you want to use the GPU
+gist griddim 100 100 100 out gistoutlong.dat \
+  refdens 0.03287 pme
+# other settings you might want to change
+# (those are the defaults)
+# gridcntr 0 0 0 gridspacn 0.5 temp 300
+\end{lstlisting}
+\subsection{Reference Values and Radial Convergence}
 \subsection{Visualizing $\dgsolv$}

### 2022-03-17 manuscript/manuscript.tex

@@
 # remove the "pme" option if you want to use the GPU
-gist griddim 100 100 100 out gistoutlong.dat \
+gist griddim 100 100 100 out gist.dat \
   refdens 0.03287 pme
@@
 \subsection{Reference Values and Radial Convergence}
+All GIST quantities should be expressed relative to bulk.
+In practice, this is not problematic for the entropies, since they are computed relative to a distribution of randomly oriented molecules at the reference density.
+The solute-solvent energy Esw is also no problem, because it naturally tends to zero in bulk. 
+However, the solvent-solvent energy Eww needs to be referenced.
+
+The Amber manual provides reference energies for several solvents.
+However, the exact reference value is different when using PME-GIST, and also depends on the box size.
+For exact analyses such as computing the total $\dgsolv$ of a compound, it is important to compute an exact reference energy.
+
+The most accurate method to compute a reference energy is to run a separate GIST calculations with the same energy method (PME/GPU) and a similar box size, and then compute the average solvent-solvent energy per molecule.
+
+\begin{equation}
+\label{eq_ewwref}
+E_{ww}^\textit{ref} = \frac{\int{E_{ww}} \mathrm{d}x}{\left<N_\textit{solvent}\right>}
+  = \frac{\sum_\textit{voxels}E_{ww}^\textit{dens} V_\textit{vox}}{\sum_\textit{voxels}{N_\textit{solvent}} / N_\textit{frames}}
+\end{equation}
+
+Alternatively, one can note that the solvent-solvent energy in a sufficiently large GIST grid converges to the correct bulk value at large distance to the solute.
+This means that the reference energy can be obtained from a large GIST grid by binning the voxels by their distance to the solute, and evaluate Equation \ref{eq_ewwref} within each bin.
+If this value converges to a constant, this can be used as $E_{ww}^\textit{ref}$.
+
+\software{gisttools} contains functionality to perform this analysis easily.
+There is also a method ``detect\_reference\_value'' that automatically tries to find the converged value fo the rdf.
+Although it has a simple convergence check built-in, it is always recommended to check the convergence by hand as well.
+
+\begin{lstlisting}[style=python]
+import gisttools as gt
+import matplotlib.pyplot as plt
+gist = gt.gist.load_gist_file(
+    "gist.dat", struct="solute-centered.pdb")
+bins, eww = gist.rdf("Eww_unref", bins=100, rmax=20.)
+_, n = gist.rdf(
+    "population", bins=100, rmax=20., col_suffix="")
+plt.plot(
+    bins,
+    eww / (n/gist.n_frames/gist.grid.voxel_volume))
+eww_auto = gist.detect_reference_value().values[0]
+plt.axhline(eww_auto, color='k')
+plt.gca().set(
+    xlabel='distance [A]', ylabel='E_ww per mol')
+plt.show()
+\end{lstlisting}
+
+The expected output of this code is shown in Figure \ref{fig_ewwref}.
+
+\begin{figure}
+	\centering
+	\includegraphics[width=0.8\linewidth]{figures/Eww_convergence.eps}
+	\caption{Convergence of $E_{ww}$ with increasing distance to the solute. The horizontal line shows the automatically computed reference energy.}
+	\label{fig_ewwref}
+\end{figure}
+
 \subsection{Visualizing $\dgsolv$}

### 2022-03-17 manuscript/manuscript.tex
Update manuscript.tex
outline for theory and entropy equations to be discussed

--- 2022-03-16 manuscript/manuscript.tex
@@
 \graphicspath{{figures/}}
-
-\usepackage{listings}
 
@@
 
-\lstdefinestyle{code}{
-	basicstyle=\ttfamily
-}
-\lstset{style=code}
-
-\DeclareSIUnit{\calorie}{cal}
-\DeclareSIUnit{\kcalPerMolASqr}{\kilo\calorie\per\mole\per\angstrom\squared}
+\newenvironment{code}{\par \noindent \ttfamily}{}
 
@@
 
+\subsection{Inhomegeneous Solvation Theory}
+
+\subsection{Solvation Entropy}
+
+\begin{equation}
+    \Delta S_\textit{solv} = \Delta S_\textit{sw} + \Delta S_\textit{ww}
+\end{equation}
+
+\begin{equation}
+    \Delta S_\textit{solv} \approx \Delta S_\textit{sw} \equiv -k_\textit{B} \frac{\rho^\textit{0}}{8\pi^\textit{2}} \int g_\textit{sw} \left(\textbf{r}, \omega \right) d\textbf{r}d\omega
+\end{equation}
+
+\subsection{Entropy Calculations in Cpptraj}
+\subsubsection{Translational and Orientatonal Entropy}
+
+\begin{equation}
+    g_\textit{vox} \left( \textbf{r}, \mathbf{\omega} \right) \approx g_\textit{vox} \left( \textbf{r})g_\textit{vox}(\mathbf{\omega} \right)
+\end{equation}
+
+\begin{equation}
+        S_{k}^\textit{trans} \approx \frac{-R}{N_\textit{k}} \left( \sum _{i=1}^{N_k} ln g_{NN, \textit{i}} \left( \textbf{r} \right) + \gamma \right)
+\end{equation}
+
+\begin{equation}
+        S_{k}^\textit{orient} \approx \frac{-R}{N_\textit{k}} \left( \sum _{i=1}^{N_k} ln g_{NN, \textit{i}} \left( \mathbf {\omega} \right) + \gamma \right)
+\end{equation}
+
+\subsubsection{Sixth Order Entropy}
+\begin{equation}
+    S_\textit{k} \approx \frac{-R}{N_\textit{k}} \left( \sum _{i=1}^{N_k} ln g_{NN, \textit{i}} \left( \textbf{r}, \mathbf {\omega} \right) + \gamma \right)
+\end{equation}
+
+\subsection{Solvation Energy}
+\subsubsection{PME implementation}
+
 Here, we aim to guide the reader through a GIST analysis of the streptavidin-biotin complex. The goal is to
@@
 For the sake of simplicity, you can download prepared and solvated Amber topologies and structures based on 1STP by running
-\begin{lstlisting}
+\begin{code}
 git clone git@github.com:liedllab/gist-tutorial.git
-\end{lstlisting}
+\end{code}
 This will download the tutorial files (including both the manuscript and the examples in the \software{code} folder).
 \subsection{System Preparation and Equilibration}
-If you want to skip this section, run 
-\begin{lstlisting}
-make equilibration-targets
-\end{lstlisting}
-in the code folder. This assumes that you have Python and Amber set up properly and that \software{pmemd.cuda} is in the PATH.
-
 When calculating solvation free energy differences between a complex and the corresponding monomers (the dissociated state), there are two options for dealing with configurational diversity in the structures.
@@
 It is therefore important that the conformations of the monomers in the complex and dissociated states match exactly.
-Use \software{cpptraj} to produce biotin and streptavidin structures based on the complex by stripping the respective other molecule like this:
-\begin{lstlisting}
-parm complex/solvated.parm7
-trajin complex/equil/EQUIL-DONE.rst
-strip ^2 parmout streptavidin/solvated.parm7
-trajout streptavidin/pre-equil.ncrst
-\end{lstlisting}
-(Adapt for biotin as necessary.)
+We provide a script called \software{cpptraj\_remove\_mol.sh} that removes a single molecule from a topology and structure using \software{cpptraj} without altering the rest of the system.
+Use this to create monomer topologies of biotin and streptavidin.
 
-%We provide a script called \software{cpptraj\_remove\_mol.sh} that removes a single molecule from a topology and structure using \software{cpptraj} without altering the rest of the system.
-\todo{Explain cpptraj rather than our script.
-The script is for the automated workflow only.}
-
-Use the \software{equilibration.py} script (or your own) to perform short NVT and NpT equilibration runs on the starting structure.
-Note that the barostat might interfere with restrained equilibration of the complex, since different molecules are moved relative to each other when the volume changes.
-But since our equilibration script also contains unrestrained pressure equilibration, we should not run into any trouble.
-
-Then split the equilibrated structure into biotin and streptavidin parts using \software{cpptraj\_remove\_mol.sh}.
-After that, you can use \software{equilibration.py -R} to run short restrained equilibrations (where only the water is allowed to move) on the individual systems.
-
-\subsection{Running MD for GIST}
-
-Now, you can run restrained MD simulations based on the equilibrated structures for subsequent GIST analyses.
-We apply a harmonic restraint of \SI{100}{\kilo\calorie\per\mole\per\angstrom\squared} to all heavy atoms to keep the system in place.
-We run \SI{100}{\nano\second} of simulation and store the conformations every \SI{100}{\pico\second}.
-A simulation length of at least \SIrange{10}{20}{\nano\second} is recommended, to obtain statistically independent snapshots of the water.
-
-An example Amber input file for GIST might look like this:
-\begin{lstlisting}
-
-\end{lstlisting}
-\subsection{Running the GIST analysis}
-
-For easier post-processing, we will use grids that contain the whole molecule.
-This could in theory be avoided for streptavidin and the complex, since only the surrounding of the biotin molecule (i.e., the binding pocket) is required for the analysis.
-We also recommend that you center the molecule in space.
-For each system (biotin, streptavidin, complex), produce a centered pdb file using \software{cpptraj}:
-\begin{lstlisting}
-parm ../solvated.parm7
-trajin EQUIL-DONE.rst
-autoimage !(:WAT) origin
-strip :WAT
-trajout equil-centered.pdb
-\end{lstlisting}
-Then, find the extent (minimum and maximum) of the x-, y-, and z-coordinates for each system.
-This can be done e.g., using the \software{mdtraj} package in Python.
-\begin{lstlisting}[basicstyle=\ttfamily]
-import mdtraj as md
-pdb = md.load("equil-centered.pdb")
-min = pdb.xyz[0].min(0) * 10.
-max = pdb.xyz[0].max(0) * 10.
-print(*list(zip("xyz", min, max)), sep="\n")
-\end{lstlisting}
-Then, decide on grid dimensions for every system.
-For a system centered at the origin with a spacing of \SI{0.5}{\angstrom} and a buffer of, e.g., \SI{7}{\angstrom} between the solute and the grid wall, you need $4.0\times (d+7)$ bins in the respective direction, where $d$ is the larger of $-$min and $+$max in the respective direction.
+\subsection{Running GIST}
 \subsection{Visualizing $\dgsolv$}

### 2022-03-17 manuscript/manuscript.tex

--- 2022-03-17 manuscript/manuscript.tex
@@
 \section{Introduction}
+\todo{ (vah) Sources and better description of all these methods, more literature research, better flow }
+
+Solvation thermodynamics govers any process involving solutes in a solvent.
+Solutes are constantly interacting with their environment, necessitating a good description of these interactions for accurate computational predictions. 
+Especially the interaction with water is of utmost importance, as many biological processes occur in a aquaeous environment. 
+The free energy of solvation thereby influences real-world problems of interest such as hydrophobic effects, the binding of a ligand to a protein, protein-protein interactions, protein dynamics or even more generally, a protein's fold.
+The computational calculation of thermodynamical solvation properties is therefore of particular interest and has been tackled by a wide array of different methods.   
+Quantum mechanical methods provide the most rigorous description of the underlying potential energy surface (PES) but suffer from high computational cost, allowing only the treatment of small systems and few solvent molecules.
+More recently, machine learning approaches and QM/MM approaches are trying to bridge the gap to computationally less demanding methods while keeping the accuracy of ab initio QM methods.
+Molecular mechanics based approaches tackle this problem by describing the underlying PES through empirically fit force-fields, which reduce the computational demand of these calculations tremendously.
+This allows for the treatment of even large biological systems. 
+Implicit solvent methods such as MM/PBSA  or MM/GBSA can quickly provide results for a large amount of supplied structures, but yield inaccurate results when compared to more rigorous explicit solvent methods.
+Somewhat inbetween the implicit solvent methods and the fully explicit solvent methods are Reference Interaction Site Models such as 3D-RISM.
+Starting with an explicit water model, 3D-RISM foregoes the explicit simulation and uses 3D solvent distributions based on the Ornstein--Zernike integral equations to predict solvation thermodynamics instead.
+The 3D-RISM approach allows for a decomposition into entropic and enthalpic terms as well as a full localization of the underlying distributions to the solute structure.
+The most statisticaly rigorous of the explicit solvent MM methods are alchemical methods such as FEP or TI. 
+Assuming sufficient sampling, they can accurately predict the free energy of solvation in the boundaries of the assumptions underlying their forcefield. 
+However, the seperation of enthalpic and entropic contributtions to the free energy of solvation is difficult with these methods and localizing these contributions not easily possible.
+In this tutorial, the Grid Inhomogenous Solvation Theory (GIST) is used.
+It is based on Lazaridis' Inhomogenous Solvation Theory (IST), solving the arising integral equations on a grid.
+This allows for a full localization of entropic and enthalpic contributions on the grid around the solute.
+Recent improvements to the method have improved the calculation speed tremendously, allowing for the treatment of large macromolecules of biological interest such as proteases or antibodies.
+For small organic molecules, the method provides results compareable to TI in accuracy, provided some small correction factor is applied.
 

### 2022-03-31 manuscript/manuscript.tex
Update manuscript.tex
Discussion on the entropy portion

--- 2022-03-17 manuscript/manuscript.tex
@@
 \section{Theory}
-GIST is an implementation of Inhomogeneous Solvation Theory (IST) \cite{Lazaridis1998} that calculates the free energy of solvation $\dgsolv$ on a 3-dimensional grid.
-It was first devised by Nguyen et al. \cite{Nguyen2012} and its implementation in \software{cpptraj} was thoroughly described in \cite{Ramsey2016}.
+GIST is an implementation of Inhomogeneous Solvation Theory (IST) \cite{Lazaridis1998} that discretizes  the free energy of solvation $\dgsolv$ on a three-dimensional grid. 
+It was first devised by Nguyen et al.
+\cite{Nguyen2012} to overcome the limitation of IST which was commonly used for regions of high water occupancy and its implementation in \software{cpptraj} was thoroughly described in \cite{Ramsey2016}.
 Here, we only provide a short overview of the theory behind GIST.
 For more detailed information, we recommend one of the more recent publications on developments in GIST. \cite{Kraml2020}\cite{Chen2021}
-\section{GIST study of streptavidin/biotin}
-
-\subsection{Inhomegeneous Solvation Theory}
 
 \subsection{Solvation Entropy}
+The solvation entropy of a system is totality of the contribution of the  solute-water entropy and the water-water entropy.
 
@@
 \end{equation}
+Here the solvation entropy is approximated from the contribution of solute-water interaction. 
+\textbf{STILLNEEDS REVISON + ADD CITATIONS} The quantity $k_\textit{b}$represents the Boltzman constant, 
+$\rho^\textit{o}$ is the bulk number density, $g_\textit{sw}\left(\textbf{r},\omega \right)$ is the solute-water 
+pair correlation function in the solute's frame of reference, \textbf{r} is the coordinates of the oxygen atom of the water molecule,  
+$\omega$ is the Euler angle, and the factor $\frac{1}{8\pi^2}$ is the normalization constant. In bulk and in regions where the orientational 
+entropy is uniform, the quantity $g_\textit{sw}\left(\textbf{r},\omega \right)$ is unity which leads to the first-order solvation entropy of 
+bulk to be zero. Also, this quantity reaches unity as the distance from the solute molecule increases which leads for the calculation of 
+the solvation entropy to be an approximation of the integral around the solute atom.
 
@@
 
+The solvation entropy is further broken down in terms of the contribution from translational and orientational terms.
+\begin{equation}
+	\Delta S_\textit{solv} = \Delta S_\textit{trans} + \Delta S_\textit{orient}
+\end{equation}
+
 \subsection{Entropy Calculations in Cpptraj}
-\subsubsection{Translational and Orientatonal Entropy}
+In \software{cpptraj} the calculation of  solvation entropy is handled into two methods.
 
+The first method assumes that the position ($r$) is independent of the orientation ($\omega$). This approximation allows for the 
+splitting of the contributions of the translational and the orientational entropy. The nearest neighbor (NN) approach is used to 
+evaluate each expression where $N_\textit{k}$ is the number of water molecules found in voxel k, $\gamma$ is the Euler's constant 
+that accounts for the bias in naive entropy estimator, and $g_\textit{NN}$ is the nearest neighbor estimate. In the nearest neighbor
+estimate, a voxel is considered to be a neighboring voxel when it shares any vertices of the voxel of interest. 
 \begin{equation}
@@
 
-\subsubsection{Sixth Order Entropy}
+The second method directly calculates the solvation entropy by evaluating the six-dimensional integral (3 for position and 3 for orientation) 
+using the nearest-neighbor approach. Unlike the first method, the nearest neighbor estimate for the sixth-order entropy considers all the water 
+found within the same and neighboring voxels.  
 \begin{equation}
@@
 \subsection{Solvation Energy}
-\subsubsection{PME implementation}
+\begin{equation}
+	\Delta E_\textit{solv} = \Delta E_\textit{sw} + \Delta E_\textit{ww}
+\end{equation}
+
+\begin{equation}
+	\Delta E_\textit{sw} = \rho^\textit{o} \int g_\textit{sw}\left(\textbf{r}\right) \Delta E_\textit{sw} \left(\textbf{r}\right) d\textbf{r}
+\end{equation}
+
+\begin{equation}
+	\Delta E_\textit{ww} = \rho^\textit{o} \int g_\textit{sw}\left(\textbf{r}\right) \Delta E_\textit{ww} \left(\textbf{r}\right) d\textbf{r}
+\end{equation}
+
+\subsection{PME implementation}
 

### 2022-04-07 manuscript/manuscript.tex

--- 2022-03-17 manuscript/manuscript.tex
@@
 \newcommand{\todo}{\textcolor{red}}
+\newcommand\inlinecode{\texttt}
 
@@
 \section{GIST study of streptavidin/biotin}
+
+\subsection{Inhomegeneous Solvation Theory}
+
+\subsection{Solvation Entropy}
+
+\begin{equation}
+    \Delta S_\textit{solv} = \Delta S_\textit{sw} + \Delta S_\textit{ww}
+\end{equation}
+
+\begin{equation}
+    \Delta S_\textit{solv} \approx \Delta S_\textit{sw} \equiv -k_\textit{B} \frac{\rho^\textit{0}}{8\pi^\textit{2}} \int g_\textit{sw} \left(\textbf{r}, \omega \right) d\textbf{r}d\omega
+\end{equation}
+
+\subsection{Entropy Calculations in Cpptraj}
+\subsubsection{Translational and Orientatonal Entropy}
+
+\begin{equation}
+    g_\textit{vox} \left( \textbf{r}, \mathbf{\omega} \right) \approx g_\textit{vox} \left( \textbf{r})g_\textit{vox}(\mathbf{\omega} \right)
+\end{equation}
+
+\begin{equation}
+        S_{k}^\textit{trans} \approx \frac{-R}{N_\textit{k}} \left( \sum _{i=1}^{N_k} ln g_{NN, \textit{i}} \left( \textbf{r} \right) + \gamma \right)
+\end{equation}
+
+\begin{equation}
+        S_{k}^\textit{orient} \approx \frac{-R}{N_\textit{k}} \left( \sum _{i=1}^{N_k} ln g_{NN, \textit{i}} \left( \mathbf {\omega} \right) + \gamma \right)
+\end{equation}
+
+\subsubsection{Sixth Order Entropy}
+\begin{equation}
+    S_\textit{k} \approx \frac{-R}{N_\textit{k}} \left( \sum _{i=1}^{N_k} ln g_{NN, \textit{i}} \left( \textbf{r}, \mathbf {\omega} \right) + \gamma \right)
+\end{equation}
+
+\subsection{Solvation Energy}
+\subsubsection{PME implementation}
 
@@
 \software{gisttools} contains functionality to perform this analysis easily.
-There is also a method ``detect\_reference\_value'' that automatically tries to find the converged value fo the rdf.
+There is also a method \inlinecode{detect\_reference\_value} that automatically tries to find the converged value fo the rdf.
 Although it has a simple convergence check built-in, it is always recommended to check the convergence by hand as well.
@@
     "gist.dat", struct="solute-centered.pdb")
-bins, eww = gist.rdf("Eww_unref", bins=100, rmax=20.)
-_, n = gist.rdf(
-    "population", bins=100, rmax=20., col_suffix="")
-plt.plot(
-    bins,
-    eww / (n/gist.n_frames/gist.grid.voxel_volume))
+bins, eww = gist.rdf(
+    "Eww_unref", bins=100, rmax=20., normalize="norm")
+plt.plot(bins, eww)
 eww_auto = gist.detect_reference_value().values[0]
@@
 	\includegraphics[width=0.8\linewidth]{figures/Eww_convergence.eps}
-	\caption{Convergence of $E_{ww}$ with increasing distance to the solute. The horizontal line shows the automatically computed reference energy.}
+	\caption{Convergence of $E_{ww}$ with increasing distance to the solute. The horizontal line shows the automatically computed reference energy. Computed from the complex calculation.}
 	\label{fig_ewwref}
@@
 
+After choosing an appropriate reference value, we need to subtract this value from Eww.
+Make sure to subtract the reference from the normalized (\_norm) data.
+In \software{gisttools}, this is done simply by setting \inlinecode{gist.eww\_ref}.
+In \software{gistpp}, this can be done as follows: \todo{This might become invalid with future GIST versions, depending on whether we adopt the COM as molecular center.}
+
+\begin{lstlisting}
+rho0=0.03287465
+neg_eww_ref=9.533
+gistpp -op multconst -i gist-gO.dx -opt const $rho0 \
+    -o gist-gO-abs.dx 
+gistpp -op div -i gist-Eww-dens.dx -i2 gist-gO-abs.dx \
+    -o gist-Eww-norm.dx
+gistpp -op addconst -i gist-Eww-norm.dx \
+    -opt const $neg_eww_ref -o gist-Eww-ref-norm.dx
+gistpp -op mult -i gist-Eww-ref-norm.dx \
+    -i2 gist-gO-abs.dx -o gist-Eww-ref-dens.dx
+\end{lstlisting}
+
+Now, add all free energy contributions to obtain the spatially resolved $\dgsolv$.
+In \software{gisttools}, this is done automatically (simply use the \inlinecode{A\_dens} and \inlinecode{A\_norm} columns).
+In \software{gistpp}, use the \inlinecode{add} command, with the same syntax as the \inlinecode{div} command used above.
+After that, check whether your free energy contributions become negligible at high distances to the solute.
+In a plot like Figure \ref{fig_ewwref} based on the \inlinecode{A\_dens} column, the values should tend to zero.
+However, it is more informative to plot the cumulative free energy contribution against the distance to the solute, to check the convergence.
+It the curve flattens out, this is your final $\dgsolv$.
+If it does not, you might need to tweak the Eww reference, or introduce a reference value for the entropy (especially with older versions of \software{cpptraj}).
+
+\begin{lstlisting}[style=python]
+import gisttools as gt
+import matplotlib.pyplot as plt
+# adapt eww_ref!
+gist = gt.gist.load_gist_file("gist.dat",
+    struct="solute-centered.pdb", eww_ref=-9.533)
+bins, (dg, eww, esw, s) = gist.rdf(
+    ["A_dens", "Esw_dens", "Eww_dens", "dTSsix_dens"],
+    bins=100, rmax=20., normalize="none")
+plt.plot(bins, np.cumsum(dg), label="dG")
+plt.plot(bins, np.cumsum(eww), label="Eww")
+plt.plot(bins, np.cumsum(esw), label="Esw")
+plt.plot(bins, np.cumsum(s), label="dS")
+\end{lstlisting}
+
+The expected output of this code is shown in Figure \ref{fig_radial_convergence}.
+In this case, both the Eww energy and the entropy are not perfectly flat, indicating that the reference value is not optimal.
+The slow energy convergence indicates that a non-optimal reference value was used, but is also complicated by the fact that a charged solute is used without neutralizing counter ions.
+
+\begin{figure}
+	\centering
+	\includegraphics[width=0.8\linewidth]{figures/A_E_S_convergence.eps}
+	\caption{Convergence of $\dgsolv$ and its contributions with increasing distance to the solute.
+All quantities cumulative, i.e., summed up to the respective radius.
+Computed from the biotin calculation.}
+	\label{fig_radial_convergence}
+\end{figure}
+
 \subsection{Visualizing $\dgsolv$}
+Next, we want to visualize the 3-dimensional contributions to the free energy of hydration, as well as the entropy and enthalpy contributions.
+You can use PyMol\cite{pymol} or VMD\cite{vmd} to visualize the .dx files from GIST.
+With \software{gistpp}, you can use the file from the convergence check. Additional OpenDX files can be produced using \inlinecode{gistpp -op makedx}.
+With \software{gisttools}, you can create one using e.g., \inlinecode{gist.save\_dx("A\_dens", "A\_dens.dx")}.
+
+Visualize the free energy, the total energy, and the entropy, at several iso-levels.
+The expected result is shown in Figure ???.
+
 \subsection{Contribution of Hydration to Binding}

### 2022-04-26 manuscript/manuscript.tex
Finish first draft of introduction
- Reorganize Scope/Requirements section
- Add some Remarks here and there
- Fix some typos

--- 2022-04-07 manuscript/manuscript.tex
@@
 \newcommand{\todo}{\textcolor{red}}
-\newcommand\inlinecode{\texttt}
 
@@
 
-Solvation thermodynamics govers any process involving solutes in a solvent.
-Solutes are constantly interacting with their environment, necessitating a good description of these interactions for accurate computational predictions. 
+Solvation thermodynamics govern any process involving solutes in a solvent.
+Solutes are constantly interacting with solvents in their environment, necessitating a good description of these interactions for accurate computational predictions. 
 Especially the interaction with water is of utmost importance, as many biological processes occur in a aquaeous environment. 
-The free energy of solvation thereby influences real-world problems of interest such as hydrophobic effects, the binding of a ligand to a protein, protein-protein interactions, protein dynamics or even more generally, a protein's fold.
+The free energy of solvation thereby strongly influences real-world problems of interest such as hydrophobic effects, the binding of a ligand to a protein, protein-protein interactions, protein dynamics or even more generally, a protein's fold.
 The computational calculation of thermodynamical solvation properties is therefore of particular interest and has been tackled by a wide array of different methods.   
@@
 More recently, machine learning approaches and QM/MM approaches are trying to bridge the gap to computationally less demanding methods while keeping the accuracy of ab initio QM methods.
-Molecular mechanics based approaches tackle this problem by describing the underlying PES through empirically fit force-fields, which reduce the computational demand of these calculations tremendously.
-This allows for the treatment of even large biological systems. 
+Molecular mechanics based approaches describe the underlying PES through empirically fit force-fields, which reduce the computational demand of these calculations tremendously.
+This allows for the treatment of even large biological systems like fully solvated proteins. 
 Implicit solvent methods such as MM/PBSA  or MM/GBSA can quickly provide results for a large amount of supplied structures, but yield inaccurate results when compared to more rigorous explicit solvent methods.
-Somewhat inbetween the implicit solvent methods and the fully explicit solvent methods are Reference Interaction Site Models such as 3D-RISM.
-Starting with an explicit water model, 3D-RISM foregoes the explicit simulation and uses 3D solvent distributions based on the Ornstein--Zernike integral equations to predict solvation thermodynamics instead.
+Somewhat inbetween the implicit solvent methods and the fully explicit solvent methods are Reference Interaction Site Models.
+One of these methods, 3D-RISM, foregoes the explicit simulation and uses 3D solvent distributions based on the Ornstein--Zernike integral equations to predict solvation thermodynamics instead.
 The 3D-RISM approach allows for a decomposition into entropic and enthalpic terms as well as a full localization of the underlying distributions to the solute structure.
@@
 Assuming sufficient sampling, they can accurately predict the free energy of solvation in the boundaries of the assumptions underlying their forcefield. 
-However, the seperation of enthalpic and entropic contributtions to the free energy of solvation is difficult with these methods and localizing these contributions not easily possible.
-In this tutorial, the Grid Inhomogenous Solvation Theory (GIST) is used.
-It is based on Lazaridis' Inhomogenous Solvation Theory (IST), solving the arising integral equations on a grid.
-This allows for a full localization of entropic and enthalpic contributions on the grid around the solute.
-Recent improvements to the method have improved the calculation speed tremendously, allowing for the treatment of large macromolecules of biological interest such as proteases or antibodies.
-For small organic molecules, the method provides results compareable to TI in accuracy, provided some small correction factor is applied.
+However, the seperation of enthalpic and entropic contributions to the free energy of solvation is difficult with these methods and localizing these contributions not easily possible.
+In this tutorial, the Grid Inhomogenous Solvation Theory (GIST) is applied to calculate solution thermodynamic properties of interest. 
+GIST is based on Lazaridis' Inhomogenous Solvation Theory (IST), solving the arising integral equations on a grid. 
+This is highly advantegous compared to other IST-based methods such as e.g. WaterMap,\todo{Add more methdods!} as this is not restricted to high-occopancy solvation sites but provides a 
+By simulating the dynamics of a solvent around the solute, GIST calculates thermodynamic properties for a standard solvation process as described by Ben-Naim.
+The method allows for a full localization of entropic and enthalpic contributions on the grid around the solute.
+For small organic molecules in water, the method provided results compareable to TI in accuracy, assuming some small correction factor to the entropy is applied. 
+Recent improvements to the method have improved the calculation speed tremendously, allowing for the treatment of larger systems of biological interest such as proteases or antibodies. 
+Other improvements worked on decreasing differences in the treatment of long-range interaction energies between GIST and popular molecular dynamics engines. 
+Yet further changes allow the inclusion of other rigid solvents than water or allow for the treatment of solvent mixtures and ions in solution. 
+While previous tutorials gave insightful pointers as to how to setup and run GIST calculations, the post-processing of such calculations was often treated less rigorously.
+While building on these tutorials, an updated guide to calculating solvation properties using GIST is therefore presented here to address the recent improvements to the method.
 
@@
 \subsection{Scope}
-The aim of this tutorial is to provide an example of a GIST workflow that can be easily adapted towards different systems.
-We present a GIST study of the streptavidin-biotin complex, which allows us to show many common aspects of GIST studies.
+
+The tutorial provides an example of a GIST workflow that can be easily adapted towards different systems. 
+In addition to a short introduction to the method and the theory behind it an overview of the various different flavours of GIST methods is provided.
+Towards this, a GIST study of the streptavidin-biotin complex is presented, which reflects on many common aspects of GIST studies.
 We present solvation analysis of a small molecule (biotin) as well as a biomolecule (streptavidin).
 For both systems, we show how to interpret the three-dimensional contributions of solvation free energy in a binding pocket.
-Furthermore, we compute the contribution of solvation free energy to binding, which requires accurate post-processing of the GIST output to avoid unfavorable summation of bias in the GIST calculation.
-We therefore believe that this example study covers a wide range of applications of the GIST method.
+The contribution of solvation free energy to binding is computed, which requires accurate post-processing of the GIST output to avoid unfavorable summation of bias in the GIST calculation.
+We believe that this example study covers a wide range of applications of the GIST method.
+Furthermore, we discuss several technical aspects  and typical pitfalls of GIST analyses, such as the normalization of voxel values and how to deal with double counting of energy contributions. 
+Additionally, we provide a python library to unify the analysis of GIST data produced by various versions of GIST and make the post-processing of GIST results more accessible.
 
-Furthermore, we discuss several technical aspects of GIST analyses, such as the normalization of voxel values and how to deal with double counting of energy contributions. 
+After completing this tutorial, we expect the reader to be able to implement their own GIST study.
+
 %Tutorials should endeavor to cover the specific task at hand, and also highlight how the steps might need to be modified (or additional care might need to be taken at particular points) to handle more general cases.
@@
 \section{Prerequisites}
-Before doing this tutorial, the reader should have a solid background knowledge of molecular dynamics (MD) simulations, as well as basic knowledge of Python, and should be familiar with working on the command line.
+\todo{ (vah)  Improve flow and logical progression of prerquisites.}
 
+%Here you would identify prerequisites/background knowledge that are assumed by your work, as well as any software/license requirements.
+
+\subsection{Background knowledge}
+This tutorial is aimed at advanced users with a solid background knowledge of molecular dynamics (MD) simulations. The user should be able to run MD simulations and work with their output data.
+Previous knowledge of the method is not necessary, but we recommend the perusal of the primary sources on the method outlined in the Theory section.
+
+The presented analyses are run in Python.
+A working understanding of Python is therefore necessary to follow the tutorial. 
+A Jupyter notebook showing directly the herein presented analyses is provided as well. 
+To easily follow the analyses an understanding of working with Jupyter notebooks is therefore beneficial, though not strictly necessary.
+
+\subsection{Software/system requirements}
+%Tutorials should clearly define what system and/or software requirements the researcher will need to complete the tutorial (e.g., VMD version 1.9 or newer, AMBER, etc.).
+Tutorials requiring specific software packages must provide instructions and files for the referenced version of the software.
+GIST's main implementation is part of the MD analysis software \software{cpptraj} as part of the AmberTools software suite.
 A recent version of \software{cpptraj} should be used, such as the one in AmberTools22 or an installation from GitHub.
-Older versions of the Amber simulation engine can be used as well.
-If no Amber installation is available, the MD can be performed using Gromacs.
-In that case, we recommend preparing the structures and topologies using AmberTools and then converting to Gromacs using \software{acpype.py}.
+The closed-source Amber simulation engine is used in this tutorial to run the necessary MD simulations. 
+If no Amber installation is available, the MD can be performed other simulation engines compatible with \software{cpptraj}, we recommend GROMACS.
+If using GROMACS, we propose to prepare the structures and topologies using AmberTools and then converting to GROMACS using \software{acpype.py}.
 In that way, a full Amber topology is available for the GIST analysis.
@@
 Furthermore, a recent Python version (>3.6) should be available with the following packages.
-
-To run the MD simulation, a PC with recent GPU is recommended.
-If the reader prefers to skip the MD and GIST calculation, we also provide the GIST output files with this tutorial, such that the post-processing can be done without any expensive calculations.
-
 \begin{itemize}
@@
 \end{itemize}
-%Here you would identify prerequisites/background knowledge that are assumed by your work, as well as any software/license requirements.
-
-\subsection{Background knowledge}
-%Tutorials should clearly define what concepts or abilities researchers will need to complete the tutorial (e.g., some proficiency in Python; experience with Jupyter notebooks; knowledge of classical MD; etc).
-
-\subsection{Software/system requirements}
-%Tutorials should clearly define what system and/or software requirements the researcher will need to complete the tutorial (e.g., VMD version 1.9 or newer, AMBER, etc.).
-Tutorials requiring specific software packages must provide instructions and files for the referenced version of the software.
+If the reader prefers to skip the MD and GIST calculation, we also provide the GIST output files with this tutorial, such that the post-processing can be done without any expensive calculations.
+A Jupyter notebook showing the herein presented analyses is provided as well, necessitating a current installation of Jupyter to run.
 
@@
 \software{gisttools} contains functionality to perform this analysis easily.
-There is also a method \inlinecode{detect\_reference\_value} that automatically tries to find the converged value fo the rdf.
+There is also a method ``detect\_reference\_value'' that automatically tries to find the converged value fo the rdf.
 Although it has a simple convergence check built-in, it is always recommended to check the convergence by hand as well.
@@
     "gist.dat", struct="solute-centered.pdb")
-bins, eww = gist.rdf(
-    "Eww_unref", bins=100, rmax=20., normalize="norm")
-plt.plot(bins, eww)
+bins, eww = gist.rdf("Eww_unref", bins=100, rmax=20.)
+_, n = gist.rdf(
+    "population", bins=100, rmax=20., col_suffix="")
+plt.plot(
+    bins,
+    eww / (n/gist.n_frames/gist.grid.voxel_volume))
 eww_auto = gist.detect_reference_value().values[0]
@@
 	\includegraphics[width=0.8\linewidth]{figures/Eww_convergence.eps}
-	\caption{Convergence of $E_{ww}$ with increasing distance to the solute. The horizontal line shows the automatically computed reference energy. Computed from the complex calculation.}
+	\caption{Convergence of $E_{ww}$ with increasing distance to the solute. The horizontal line shows the automatically computed reference energy.}
 	\label{fig_ewwref}
@@
 
-After choosing an appropriate reference value, we need to subtract this value from Eww.
-Make sure to subtract the reference from the normalized (\_norm) data.
-In \software{gisttools}, this is done simply by setting \inlinecode{gist.eww\_ref}.
-In \software{gistpp}, this can be done as follows: \todo{This might become invalid with future GIST versions, depending on whether we adopt the COM as molecular center.}
-
-\begin{lstlisting}
-rho0=0.03287465
-neg_eww_ref=9.533
-gistpp -op multconst -i gist-gO.dx -opt const $rho0 \
-    -o gist-gO-abs.dx 
-gistpp -op div -i gist-Eww-dens.dx -i2 gist-gO-abs.dx \
-    -o gist-Eww-norm.dx
-gistpp -op addconst -i gist-Eww-norm.dx \
-    -opt const $neg_eww_ref -o gist-Eww-ref-norm.dx
-gistpp -op mult -i gist-Eww-ref-norm.dx \
-    -i2 gist-gO-abs.dx -o gist-Eww-ref-dens.dx
-\end{lstlisting}
-
-Now, add all free energy contributions to obtain the spatially resolved $\dgsolv$.
-In \software{gisttools}, this is done automatically (simply use the \inlinecode{A\_dens} and \inlinecode{A\_norm} columns).
-In \software{gistpp}, use the \inlinecode{add} command, with the same syntax as the \inlinecode{div} command used above.
-After that, check whether your free energy contributions become negligible at high distances to the solute.
-In a plot like Figure \ref{fig_ewwref} based on the \inlinecode{A\_dens} column, the values should tend to zero.
-However, it is more informative to plot the cumulative free energy contribution against the distance to the solute, to check the convergence.
-It the curve flattens out, this is your final $\dgsolv$.
-If it does not, you might need to tweak the Eww reference, or introduce a reference value for the entropy (especially with older versions of \software{cpptraj}).
-
-\begin{lstlisting}[style=python]
-import gisttools as gt
-import matplotlib.pyplot as plt
-# adapt eww_ref!
-gist = gt.gist.load_gist_file("gist.dat",
-    struct="solute-centered.pdb", eww_ref=-9.533)
-bins, (dg, eww, esw, s) = gist.rdf(
-    ["A_dens", "Esw_dens", "Eww_dens", "dTSsix_dens"],
-    bins=100, rmax=20., normalize="none")
-plt.plot(bins, np.cumsum(dg), label="dG")
-plt.plot(bins, np.cumsum(eww), label="Eww")
-plt.plot(bins, np.cumsum(esw), label="Esw")
-plt.plot(bins, np.cumsum(s), label="dS")
-\end{lstlisting}
-
-The expected output of this code is shown in Figure \ref{fig_radial_convergence}.
-In this case, both the Eww energy and the entropy are not perfectly flat, indicating that the reference value is not optimal.
-The slow energy convergence indicates that a non-optimal reference value was used, but is also complicated by the fact that a charged solute is used without neutralizing counter ions.
-
-\begin{figure}
-	\centering
-	\includegraphics[width=0.8\linewidth]{figures/A_E_S_convergence.eps}
-	\caption{Convergence of $\dgsolv$ and its contributions with increasing distance to the solute.
-All quantities cumulative, i.e., summed up to the respective radius.
-Computed from the biotin calculation.}
-	\label{fig_radial_convergence}
-\end{figure}
-
 \subsection{Visualizing $\dgsolv$}
-Next, we want to visualize the 3-dimensional contributions to the free energy of hydration, as well as the entropy and enthalpy contributions.
-You can use PyMol\cite{pymol} or VMD\cite{vmd} to visualize the .dx files from GIST.
-With \software{gistpp}, you can use the file from the convergence check. Additional OpenDX files can be produced using \inlinecode{gistpp -op makedx}.
-With \software{gisttools}, you can create one using e.g., \inlinecode{gist.save\_dx("A\_dens", "A\_dens.dx")}.
-
-Visualize the free energy, the total energy, and the entropy, at several iso-levels.
-The expected result is shown in Figure ???.
-
 \subsection{Contribution of Hydration to Binding}

### 2022-04-26 manuscript/manuscript.tex
Update manuscript.tex
solvation energy discussion

--- 2022-04-07 manuscript/manuscript.tex
@@
 \newcommand{\todo}{\textcolor{red}}
-\newcommand\inlinecode{\texttt}
 
-\lstset{
-	basicstyle=\ttfamily\footnotesize
-}
-\lstdefinestyle{code}{}
-\lstdefinestyle{python}{
-	language=Python,
-	stringstyle=\color[rgb]{1.0, 0, 0},
-	commentstyle=\color[rgb]{0, 0, 0.6}
-}
-\lstset{style=code}
-
-\DeclareSIUnit{\calorie}{cal}
-\DeclareSIUnit{\kcalPerMolASqr}{\kilo\calorie\per\mole\per\angstrom\squared}
+\newenvironment{code}{\par \noindent \ttfamily}{}
 
@@
 \section{Introduction}
-\todo{ (vah) Sources and better description of all these methods, more literature research, better flow }
-
-Solvation thermodynamics govers any process involving solutes in a solvent.
-Solutes are constantly interacting with their environment, necessitating a good description of these interactions for accurate computational predictions. 
-Especially the interaction with water is of utmost importance, as many biological processes occur in a aquaeous environment. 
-The free energy of solvation thereby influences real-world problems of interest such as hydrophobic effects, the binding of a ligand to a protein, protein-protein interactions, protein dynamics or even more generally, a protein's fold.
-The computational calculation of thermodynamical solvation properties is therefore of particular interest and has been tackled by a wide array of different methods.   
-Quantum mechanical methods provide the most rigorous description of the underlying potential energy surface (PES) but suffer from high computational cost, allowing only the treatment of small systems and few solvent molecules.
-More recently, machine learning approaches and QM/MM approaches are trying to bridge the gap to computationally less demanding methods while keeping the accuracy of ab initio QM methods.
-Molecular mechanics based approaches tackle this problem by describing the underlying PES through empirically fit force-fields, which reduce the computational demand of these calculations tremendously.
-This allows for the treatment of even large biological systems. 
-Implicit solvent methods such as MM/PBSA  or MM/GBSA can quickly provide results for a large amount of supplied structures, but yield inaccurate results when compared to more rigorous explicit solvent methods.
-Somewhat inbetween the implicit solvent methods and the fully explicit solvent methods are Reference Interaction Site Models such as 3D-RISM.
-Starting with an explicit water model, 3D-RISM foregoes the explicit simulation and uses 3D solvent distributions based on the Ornstein--Zernike integral equations to predict solvation thermodynamics instead.
-The 3D-RISM approach allows for a decomposition into entropic and enthalpic terms as well as a full localization of the underlying distributions to the solute structure.
-The most statisticaly rigorous of the explicit solvent MM methods are alchemical methods such as FEP or TI. 
-Assuming sufficient sampling, they can accurately predict the free energy of solvation in the boundaries of the assumptions underlying their forcefield. 
-However, the seperation of enthalpic and entropic contributtions to the free energy of solvation is difficult with these methods and localizing these contributions not easily possible.
-In this tutorial, the Grid Inhomogenous Solvation Theory (GIST) is used.
-It is based on Lazaridis' Inhomogenous Solvation Theory (IST), solving the arising integral equations on a grid.
-This allows for a full localization of entropic and enthalpic contributions on the grid around the solute.
-Recent improvements to the method have improved the calculation speed tremendously, allowing for the treatment of large macromolecules of biological interest such as proteases or antibodies.
-For small organic molecules, the method provides results compareable to TI in accuracy, provided some small correction factor is applied.
 
@@
 \section{Prerequisites}
-Before doing this tutorial, the reader should have a solid background knowledge of molecular dynamics (MD) simulations, as well as basic knowledge of Python, and should be familiar with working on the command line.
 
-A recent version of \software{cpptraj} should be used, such as the one in AmberTools22 or an installation from GitHub.
-Older versions of the Amber simulation engine can be used as well.
-If no Amber installation is available, the MD can be performed using Gromacs.
-In that case, we recommend preparing the structures and topologies using AmberTools and then converting to Gromacs using \software{acpype.py}.
-In that way, a full Amber topology is available for the GIST analysis.
-
-Furthermore, a recent Python version (>3.6) should be available with the following packages.
-
-To run the MD simulation, a PC with recent GPU is recommended.
-If the reader prefers to skip the MD and GIST calculation, we also provide the GIST output files with this tutorial, such that the post-processing can be done without any expensive calculations.
-
-\begin{itemize}
-	\item \software{gisttools} (and its prerequisites)
-	\item \software{mdtraj}
-\end{itemize}
 %Here you would identify prerequisites/background knowledge that are assumed by your work, as well as any software/license requirements.
@@
 \subsection{Solvation Energy}
+The solvation energy is readily calculated by summing up the contributions of the water-water interaction and the water-solute interaction.
 \begin{equation}
@@
 \begin{equation}
-	\Delta E_\textit{sw} = \rho^\textit{o} \int g_\textit{sw}\left(\textbf{r}\right) \Delta E_\textit{sw} \left(\textbf{r}\right) d\textbf{r}
+	\Delta E_\textit{sw} \equiv \frac{1}{8\pi^2} \int g_\textit{sw}\left(\omega|\textbf{r}\right) U_\textit{sw}\left(\textbf{r}, \omega\right) d\omega
 \end{equation}
 
+Similar to the entropy integrals, the solute-water solvation energy integral also decays with increasing distance from the solute.
+Hence, the solvation energy can be approximated by local spatial integrals.
+However, there is a complication arising from this approximation.
+Water molecules outside the region of interest will also be accounted for in the pairwise water term.
+For example, given two different regions $R_a$ and $R_b$, will not sum up to the total energy of the two regions due to double counting.
+
 \begin{equation}
-	\Delta E_\textit{ww} = \rho^\textit{o} \int g_\textit{sw}\left(\textbf{r}\right) \Delta E_\textit{ww} \left(\textbf{r}\right) d\textbf{r}
+\begin{aligned}
+	\Delta E_\textit{ww} \equiv & \left(\frac{1}{8\pi^2} \right)^2 \rho^o \int g_\textit{sw}\left(\omega|\textbf{r}\right) \\
+	& \times \left[g_\textit{sw}\left(\textbf{r}^\prime , \omega^\prime \right) - g^o_\textit{ww} \left(\textbf{r}, \omega, \textbf{r}^\prime, \omega^\prime \right)\right] \\
+	& \times U_\textit{ww}\left(\textbf{r}, \omega, \textbf{r}^\prime, \omega^\prime\right)d\omega d\textbf{r}^\prime d\omega^\prime
+\end{aligned}
+\end{equation}
+There are two ways in which this double counting was accounted for in the original paper.
+First, using the water-displacement method.
+This method is applicable when a determining the energetic cost of kicking out a water molecule.
+\begin{equation}
+	\Delta E^R_\textit{ww} = n^\textit{R} E^\textit{bulk}_\textit{ww} - E^\textit{R,corr}_\textit{ww}
+\end{equation}
+
+The second method uses normalized-water properties to account the difference in water properties of two different regions.
+The normalized water property is defined as the water property per water molecule.
+\begin{equation}
+	\Delta E^\textit{R,norm}_\textit{ww} = E^\textit{R,norm}_\textit{ww} - 2E^\textit{bulk}_\textit{ww}
 \end{equation}
@@
 It is therefore important that the conformations of the monomers in the complex and dissociated states match exactly.
-Use \software{cpptraj} to produce biotin and streptavidin structures based on the complex by stripping the respective other molecule like this:
-\begin{lstlisting}
-parm complex/solvated.parm7
-trajin complex/equil/EQUIL-DONE.rst
-strip ^2 parmout streptavidin/solvated.parm7
-trajout streptavidin/pre-equil.ncrst
-\end{lstlisting}
-(Adapt for biotin as necessary.)
+We provide a script called \software{cpptraj\_remove\_mol.sh} that removes a single molecule from a topology and structure using \software{cpptraj} without altering the rest of the system.
+Use this to create monomer topologies of biotin and streptavidin.
 
-%We provide a script called \software{cpptraj\_remove\_mol.sh} that removes a single molecule from a topology and structure using \software{cpptraj} without altering the rest of the system.
-\todo{Explain cpptraj rather than our script.
-The script is for the automated workflow only.}
-
-Use the \software{equilibration.py} script (or your own) to perform short NVT and NpT equilibration runs on the starting structure.
-Note that the barostat might interfere with restrained equilibration of the complex, since different molecules are moved relative to each other when the volume changes.
-But since our equilibration script also contains unrestrained pressure equilibration, we should not run into any trouble.
-
-Then split the equilibrated structure into biotin and streptavidin parts using \software{cpptraj\_remove\_mol.sh}.
-After that, you can use \software{equilibration.py -R} to run short restrained equilibrations (where only the water is allowed to move) on the individual systems.
-
-\subsection{Running MD for GIST}
-
-Now, you can run restrained MD simulations based on the equilibrated structures for subsequent GIST analyses.
-We apply a harmonic restraint of \SI{100}{\kilo\calorie\per\mole\per\angstrom\squared} to all heavy atoms to keep the system in place.
-We run \SI{100}{\nano\second} of simulation and store the conformations every \SI{100}{\pico\second}.
-A simulation length of at least \SIrange{10}{20}{\nano\second} is recommended, to obtain statistically independent snapshots of the water.
-
-An example Amber input file for GIST might look like this:
-\begin{lstlisting}
-restrained 100 ns NpT
-&cntrl
-  ntx=5, irest=1,
-  ioutfm=1,
-  ntb=2, iwrap=1,
-  ntr=1, restraint_wt=100.0,
-    restraintmask=\"!@H=&!:WAT\",
-  ntp=1, pres0=1.0, taup=1.0,
-  ntc=2, ntf=2,
-  ntt=3, tempi=300.0, temp0=300.0, gamma_ln=2,
-  nstlim=50000000, dt=0.002,
-  ntwr=50000, ntwx=5000,
-  ntpr=5000,
-/
-\end{lstlisting}
-Note that these are very standard MD settings except for the line that defines the restraints.
-
-Run the MD using \software{pmemd.cuda}:
-\begin{lstlisting}
-pmemd.cuda \
-	-O \
-	-i 100ns-npt-restraint.in \
-	-o md-01.out \
-	-p solvated.parm7 \
-	-c EQUIL-DONE.rst \
-	-x md-01.nc \
-	-r md-01.ncrst \
-	-ref EQUIL-DONE.rst
-\end{lstlisting}
-\subsection{Running the GIST analysis}
-
-For easier post-processing, we will use grids that contain the whole molecule.
-This could in theory be avoided for streptavidin and the complex, since only the surrounding of the biotin molecule (i.e., the binding pocket) is required for the analysis.
-We also recommend that you center the molecule in space.
-For each system (biotin, streptavidin, complex), produce a centered pdb file using \software{cpptraj}:
-\begin{lstlisting}
-parm solvated.parm7
-trajin md-01.nc 1 1 1
-autoimage !(:WAT) origin
-strip :WAT
-trajout solute-centered.pdb
-\end{lstlisting}
-Then, find the extent (minimum and maximum) of the x-, y-, and z-coordinates for each system.
-This can be done e.g., using the \software{mdtraj} package in Python.
-\begin{lstlisting}[style=python]
-import mdtraj as md
-pdb = md.load("solute-centered.pdb")
-# mdtraj converts to nm.
-# *10 to convert back to Angstrom.
-min = pdb.xyz[0].min(0) * 10.
-max = pdb.xyz[0].max(0) * 10.
-print(*list(zip("xyz", min, max)), sep="\n")
-\end{lstlisting}
-Then, decide on grid dimensions for every system.
-For a system centered at the origin with a spacing of \SI{0.5}{\angstrom} and a buffer of, e.g., \SI{7}{\angstrom} between the solute and the grid wall, you need $4.0\times (d+7)$ bins in the respective direction, where $d$ is the larger of $-$min and $+$max in the respective direction.
-We note that this centered grid is usually larger than it needs to be in some directions.
-If you want to save some extra memory and calculation time, feel free to align the grid more precisely to the minimum and maximum coordinates, or use a smaller grid that only covers the binding pocket.
-You can also rotate the molecule by its principal axes to fit the cuboid grid more exactly.
-However, this needs to be done \emph{before} the MD run, since rotating the trajectory damages the periodic box information.
-
-Now, you can finally run the actual GIST analysis. Center the molecule exactly in the same way as above, to make sure that its position matches the grid parameters that you decided on.
-We recommend using either the PME or GPU implementation of GIST, depending on whether you have a fast GPU.
-A further advantage of the PME implementation is that it matches the energy calculation of the MD engine more closely.
-\todo{At the time of writing, it is important to use an up-to-date version of PME-GIST, or the Esw-Eww decomposition will not be available.
-Maybe remove this for the final version, when Amber22 is out.}
-\begin{lstlisting}
-parm solvated.parm7
-trajin md-01.nc
-autoimage !(:WAT) origin
-# remove the "pme" option if you want to use the GPU
-gist griddim 100 100 100 out gist.dat \
-  refdens 0.03287 pme
-# other settings you might want to change
-# (those are the defaults)
-# gridcntr 0 0 0 gridspacn 0.5 temp 300
-\end{lstlisting}
-\subsection{Reference Values and Radial Convergence}
-All GIST quantities should be expressed relative to bulk.
-In practice, this is not problematic for the entropies, since they are computed relative to a distribution of randomly oriented molecules at the reference density.
-The solute-solvent energy Esw is also no problem, because it naturally tends to zero in bulk. 
-However, the solvent-solvent energy Eww needs to be referenced.
-
-The Amber manual provides reference energies for several solvents.
-However, the exact reference value is different when using PME-GIST, and also depends on the box size.
-For exact analyses such as computing the total $\dgsolv$ of a compound, it is important to compute an exact reference energy.
-
-The most accurate method to compute a reference energy is to run a separate GIST calculations with the same energy method (PME/GPU) and a similar box size, and then compute the average solvent-solvent energy per molecule.
-
-\begin{equation}
-\label{eq_ewwref}
-E_{ww}^\textit{ref} = \frac{\int{E_{ww}} \mathrm{d}x}{\left<N_\textit{solvent}\right>}
-  = \frac{\sum_\textit{voxels}E_{ww}^\textit{dens} V_\textit{vox}}{\sum_\textit{voxels}{N_\textit{solvent}} / N_\textit{frames}}
-\end{equation}
-
-Alternatively, one can note that the solvent-solvent energy in a sufficiently large GIST grid converges to the correct bulk value at large distance to the solute.
-This means that the reference energy can be obtained from a large GIST grid by binning the voxels by their distance to the solute, and evaluate Equation \ref{eq_ewwref} within each bin.
-If this value converges to a constant, this can be used as $E_{ww}^\textit{ref}$.
-
-\software{gisttools} contains functionality to perform this analysis easily.
-There is also a method \inlinecode{detect\_reference\_value} that automatically tries to find the converged value fo the rdf.
-Although it has a simple convergence check built-in, it is always recommended to check the convergence by hand as well.
-
-\begin{lstlisting}[style=python]
-import gisttools as gt
-import matplotlib.pyplot as plt
-gist = gt.gist.load_gist_file(
-    "gist.dat", struct="solute-centered.pdb")
-bins, eww = gist.rdf(
-    "Eww_unref", bins=100, rmax=20., normalize="norm")
-plt.plot(bins, eww)
-eww_auto = gist.detect_reference_value().values[0]
-plt.axhline(eww_auto, color='k')
-plt.gca().set(
-    xlabel='distance [A]', ylabel='E_ww per mol')
-plt.show()
-\end{lstlisting}
-
-The expected output of this code is shown in Figure \ref{fig_ewwref}.
-
-\begin{figure}
-	\centering
-	\includegraphics[width=0.8\linewidth]{figures/Eww_convergence.eps}
-	\caption{Convergence of $E_{ww}$ with increasing distance to the solute. The horizontal line shows the automatically computed reference energy. Computed from the complex calculation.}
-	\label{fig_ewwref}
-\end{figure}
-
-After choosing an appropriate reference value, we need to subtract this value from Eww.
-Make sure to subtract the reference from the normalized (\_norm) data.
-In \software{gisttools}, this is done simply by setting \inlinecode{gist.eww\_ref}.
-In \software{gistpp}, this can be done as follows: \todo{This might become invalid with future GIST versions, depending on whether we adopt the COM as molecular center.}
-
-\begin{lstlisting}
-rho0=0.03287465
-neg_eww_ref=9.533
-gistpp -op multconst -i gist-gO.dx -opt const $rho0 \
-    -o gist-gO-abs.dx 
-gistpp -op div -i gist-Eww-dens.dx -i2 gist-gO-abs.dx \
-    -o gist-Eww-norm.dx
-gistpp -op addconst -i gist-Eww-norm.dx \
-    -opt const $neg_eww_ref -o gist-Eww-ref-norm.dx
-gistpp -op mult -i gist-Eww-ref-norm.dx \
-    -i2 gist-gO-abs.dx -o gist-Eww-ref-dens.dx
-\end{lstlisting}
-
-Now, add all free energy contributions to obtain the spatially resolved $\dgsolv$.
-In \software{gisttools}, this is done automatically (simply use the \inlinecode{A\_dens} and \inlinecode{A\_norm} columns).
-In \software{gistpp}, use the \inlinecode{add} command, with the same syntax as the \inlinecode{div} command used above.
-After that, check whether your free energy contributions become negligible at high distances to the solute.
-In a plot like Figure \ref{fig_ewwref} based on the \inlinecode{A\_dens} column, the values should tend to zero.
-However, it is more informative to plot the cumulative free energy contribution against the distance to the solute, to check the convergence.
-It the curve flattens out, this is your final $\dgsolv$.
-If it does not, you might need to tweak the Eww reference, or introduce a reference value for the entropy (especially with older versions of \software{cpptraj}).
-
-\begin{lstlisting}[style=python]
-import gisttools as gt
-import matplotlib.pyplot as plt
-# adapt eww_ref!
-gist = gt.gist.load_gist_file("gist.dat",
-    struct="solute-centered.pdb", eww_ref=-9.533)
-bins, (dg, eww, esw, s) = gist.rdf(
-    ["A_dens", "Esw_dens", "Eww_dens", "dTSsix_dens"],
-    bins=100, rmax=20., normalize="none")
-plt.plot(bins, np.cumsum(dg), label="dG")
-plt.plot(bins, np.cumsum(eww), label="Eww")
-plt.plot(bins, np.cumsum(esw), label="Esw")
-plt.plot(bins, np.cumsum(s), label="dS")
-\end{lstlisting}
-
-The expected output of this code is shown in Figure \ref{fig_radial_convergence}.
-In this case, both the Eww energy and the entropy are not perfectly flat, indicating that the reference value is not optimal.
-The slow energy convergence indicates that a non-optimal reference value was used, but is also complicated by the fact that a charged solute is used without neutralizing counter ions.
-
-\begin{figure}
-	\centering
-	\includegraphics[width=0.8\linewidth]{figures/A_E_S_convergence.eps}
-	\caption{Convergence of $\dgsolv$ and its contributions with increasing distance to the solute.
-All quantities cumulative, i.e., summed up to the respective radius.
-Computed from the biotin calculation.}
-	\label{fig_radial_convergence}
-\end{figure}
-
+\subsection{Running GIST}
 \subsection{Visualizing $\dgsolv$}
-Next, we want to visualize the 3-dimensional contributions to the free energy of hydration, as well as the entropy and enthalpy contributions.
-You can use PyMol\cite{pymol} or VMD\cite{vmd} to visualize the .dx files from GIST.
-With \software{gistpp}, you can use the file from the convergence check. Additional OpenDX files can be produced using \inlinecode{gistpp -op makedx}.
-With \software{gisttools}, you can create one using e.g., \inlinecode{gist.save\_dx("A\_dens", "A\_dens.dx")}.
-
-Visualize the free energy, the total energy, and the entropy, at several iso-levels.
-The expected result is shown in Figure ???.
-
 \subsection{Contribution of Hydration to Binding}

### 2022-04-26 manuscript/manuscript.tex
More text for tutorial

--- 2022-04-07 manuscript/manuscript.tex
@@
 \section{Theory}
-GIST is an implementation of Inhomogeneous Solvation Theory (IST) \cite{Lazaridis1998} that discretizes  the free energy of solvation $\dgsolv$ on a three-dimensional grid. 
-It was first devised by Nguyen et al.
-\cite{Nguyen2012} to overcome the limitation of IST which was commonly used for regions of high water occupancy and its implementation in \software{cpptraj} was thoroughly described in \cite{Ramsey2016}.
+GIST is an implementation of Inhomogeneous Solvation Theory (IST) \cite{Lazaridis1998} that calculates the free energy of solvation $\dgsolv$ on a 3-dimensional grid.
+It was first devised by Nguyen et al. \cite{Nguyen2012} and its implementation in \software{cpptraj} was thoroughly described in \cite{Ramsey2016}.
 Here, we only provide a short overview of the theory behind GIST.
 For more detailed information, we recommend one of the more recent publications on developments in GIST. \cite{Kraml2020}\cite{Chen2021}
+\section{GIST study of streptavidin/biotin}
+
+\subsection{Inhomegeneous Solvation Theory}
 
 \subsection{Solvation Entropy}
-The solvation entropy of a system is totality of the contribution of the  solute-water entropy and the water-water entropy.
 
@@
 \end{equation}
-Here the solvation entropy is approximated from the contribution of solute-water interaction. 
-\textbf{STILLNEEDS REVISON + ADD CITATIONS} The quantity $k_\textit{b}$represents the Boltzman constant, 
-$\rho^\textit{o}$ is the bulk number density, $g_\textit{sw}\left(\textbf{r},\omega \right)$ is the solute-water 
-pair correlation function in the solute's frame of reference, \textbf{r} is the coordinates of the oxygen atom of the water molecule,  
-$\omega$ is the Euler angle, and the factor $\frac{1}{8\pi^2}$ is the normalization constant. In bulk and in regions where the orientational 
-entropy is uniform, the quantity $g_\textit{sw}\left(\textbf{r},\omega \right)$ is unity which leads to the first-order solvation entropy of 
-bulk to be zero. Also, this quantity reaches unity as the distance from the solute molecule increases which leads for the calculation of 
-the solvation entropy to be an approximation of the integral around the solute atom.
 
@@
 
-The solvation entropy is further broken down in terms of the contribution from translational and orientational terms.
-\begin{equation}
-	\Delta S_\textit{solv} = \Delta S_\textit{trans} + \Delta S_\textit{orient}
-\end{equation}
+\subsection{Entropy Calculations in Cpptraj}
+\subsubsection{Translational and Orientatonal Entropy}
 
-\subsection{Entropy Calculations in Cpptraj}
-In \software{cpptraj} the calculation of  solvation entropy is handled into two methods.
-
-The first method assumes that the position ($r$) is independent of the orientation ($\omega$). This approximation allows for the 
-splitting of the contributions of the translational and the orientational entropy. The nearest neighbor (NN) approach is used to 
-evaluate each expression where $N_\textit{k}$ is the number of water molecules found in voxel k, $\gamma$ is the Euler's constant 
-that accounts for the bias in naive entropy estimator, and $g_\textit{NN}$ is the nearest neighbor estimate. In the nearest neighbor
-estimate, a voxel is considered to be a neighboring voxel when it shares any vertices of the voxel of interest. 
 \begin{equation}
@@
 
-The second method directly calculates the solvation entropy by evaluating the six-dimensional integral (3 for position and 3 for orientation) 
-using the nearest-neighbor approach. Unlike the first method, the nearest neighbor estimate for the sixth-order entropy considers all the water 
-found within the same and neighboring voxels.  
+\subsubsection{Sixth Order Entropy}
 \begin{equation}
@@
 \subsection{Solvation Energy}
-\begin{equation}
-	\Delta E_\textit{solv} = \Delta E_\textit{sw} + \Delta E_\textit{ww}
-\end{equation}
-
-\begin{equation}
-	\Delta E_\textit{sw} = \rho^\textit{o} \int g_\textit{sw}\left(\textbf{r}\right) \Delta E_\textit{sw} \left(\textbf{r}\right) d\textbf{r}
-\end{equation}
-
-\begin{equation}
-	\Delta E_\textit{ww} = \rho^\textit{o} \int g_\textit{sw}\left(\textbf{r}\right) \Delta E_\textit{ww} \left(\textbf{r}\right) d\textbf{r}
-\end{equation}
-
-\subsection{PME implementation}
+\subsubsection{PME implementation}
 
@@
 In \software{gisttools}, this is done simply by setting \inlinecode{gist.eww\_ref}.
-In \software{gistpp}, this can be done as follows: \todo{This might become invalid with future GIST versions, depending on whether we adopt the COM as molecular center.}
+In \software{gistpp}, this can be done as follows: \todo{This might become invalid with future GIST versions, depending on whether we adopt the COM as molecular center.} \todo{Better use the population.}
 
@@
 
+It might also be interesting to visualize the average energy of a water molecule at each grid voxel.
+This quantity is given by 2*Eww + Esw.
+The energy referencing leads to non-zero normalized values where the population is zero.
+While this is irrelevant for further post-processing, it is advantageous for visualization to set those empty regions to zero.
+A .dx file can be produced using gisttools as follows:
+
+\begin{lstlisting}[style=python]
+import gisttools as gt
+gist = gt.gist.load_gist_file("gist.dat",
+    struct="solute-centered.pdb", eww_ref=-9.533)
+gist['E_norm'] = gist['Eww_norm'] * 2 + gist['Esw_norm']
+gist.loc[gist['population'] == 0, 'E_norm'] = 0
+gist.save_dx('E_norm', 'gist-E-per-mol-norm.dx')
+\end{lstlisting}
+
 Visualize the free energy, the total energy, and the entropy, at several iso-levels.
-The expected result is shown in Figure ???.
+An minimal PyMOL input might look like this:
+
+\begin{lstlisting}
+load solute-centered.pdb
+show sticks, resname BTN
+as surface, not resname BTN
+load gist-gO.dx, gO
+isosurface gO_high, gO, 2.
+\end{lstlisting}
+
+Adapt the above PyMOL script to visualize the average energy per solvent at a positive and a negative isolevel, as well as regions of negative entropy per solvent.
+The expected representation of the streptavidin binding pocket is shown in Figure \ref{fig_binding_pocket_pymol}.
+
+\begin{figure}
+	\centering
+	\includegraphics[width=0.8\linewidth]{figures/binding_pocket_S_-2_E_-2_E_3.png}
+	\caption{Free energy contributions in the streptavidin binding pocket.
+Red: high solvent energy (at +3 kcal/mol).
+Blue: low solvent energy (at -2 kcal/mol).
+Green: low solvent entropy $T\Delta S^{six}$ (at -2 kcal/mol).}
+	\label{fig_binding_pocket_pymol}
+\end{figure}
+
+You should find that there are several regions in the binding pocket that feature a high solvent energy and/or low solvent entropy.
+However, such regions also exist outside of the binding pocket, and there are also low-energy regions within the binding pocket.
+Therefore, the strong affinity to biotin can not be explained solely by the hydration contribution, and the direct interaction energy also needs to be taken into account.
 
 \subsection{Contribution of Hydration to Binding}
+In the next step, calculate the $\dgsolv$ contributions around the biotin ligand in each of the three systems (biotin, streptavidin, and complex).
+Note that our integration region does not fully reach into bulk, but also comprises regions that are close to the streptavidin protein. 
+Therefore, our results for each system will depend on the exact position of the integration boundary. 
+To avoid inconsistencies, it is important to choose exactly the same integration region for each system.
+Although we kept the systems rigid, the center of mass might have been shifted during pressure equilibration.
+Therefore, we align the complex structure to the streptavidin system and use the shifted biotin coordinates to define the integration region.
 
+We recommend using a \software{Jupyter Notebook} for the following analyses, although they can also be done in a script or an interactive Python prompt.
+The code is not explained in detail here, but should be clear after reading the previous sections.
+
+Load the files and double check the frame numbers and reference density.
+\begin{lstlisting}[style=python]
+import numpy as np
+from gisttools.gist import load_gist_file
+import matplotlib.pyplot as plt
+
+compl = load_gist_file('complex/gist/gist.dat', \
+    struct='complex/gist/solute-centered.pdb')
+print(compl.n_frames, compl.rho0)
+biotin = load_gist_file('biotin/gist/gist.dat', \
+    struct='biotin/gist/solute-centered.pdb')
+print(biotin.n_frames, biotin.rho0)
+strept = load_gist_file('streptavidin/gist/gist.dat', \
+    struct='streptavidin/gist/solute-centered.pdb')
+print(strept.n_frames, strept.rho0)
+\end{lstlisting}
+
+Assign reference energies and check them for plausibility.
+\begin{lstlisting}[style=python]
+biotin.eww_ref = biotin.detect_reference_value()
+print("Biotin:", biotin.eww_ref)
+strept.eww_ref = strept.detect_reference_value()
+print("Streptavidin:", strept.eww_ref)
+compl.eww_ref = compl.detect_reference_value()
+print("Complex:", compl.eww_ref)
+\end{lstlisting}
+
+Subtract a reference entropy from the dTSsix columns.
+\begin{lstlisting}[style=python]
+def reference_entropy(gf):
+    if 'dTSsix_unref_norm' not in gf.data.columns:
+        gf['dTSsix_unref_norm'] = gf['dTSsix_norm']
+        gf['dTSsix_unref_dens'] = gf['dTSsix_dens']
+    gf['dTSsix_norm'] = gf['dTSsix_unref_norm'] \
+        - gf.detect_reference_value('dTSsix_unref_dens')
+    gf['dTSsix_dens'] = gf.norm2dens(gf['dTSsix_norm'])
+reference_entropy(biotin)
+reference_entropy(strept)
+reference_entropy(compl)
+\end{lstlisting}
+
+Compute the atom positions that define the integration region.
+For the streptavidin integral, we align the complex structure onto the streptavidin structure and then use the biotin positions as centers.
+Note that we use \software{mdtraj} here, since \software{gisttools} stores the reference structure as a \inlinecode{mdtraj.Trajectory} object, but any other library for post-processing of trajectories works as well.
+\begin{lstlisting}[style=python]
+col = 'dTSsix_dens'
+def select(traj, sel):
+    "Slice a Trajectory by selection mask."
+    return traj.atom_slice(traj.top.select(sel))
+# we multiply by 10 to convert nm to Angstrom.
+compl_x = select(compl.struct, biotin_mask).xyz[0] * 10.
+biotin_x = select(biotin.struct, biotin_mask).xyz[0] * 10.
+aligned = compl.struct[:].superpose(strept.struct, \
+    atom_indices=strept.struct.top.select(strept_mask))
+aligned = select(aligned, biotin_mask)
+strept_x = aligned.xyz[0] * 10.
+
+bins, biotin_rdf = biotin.rdf( \
+    col, centers=biotin_x, bins=100, rmax=24)
+bins, strept_rdf = strept.rdf( \
+    col, centers=strept_x, bins=100, rmax=24)
+bins, compl_rdf = compl.rdf( \
+    col, centers=compl_x, bins=100, rmax=24)
+\end{lstlisting}
+
+Now, subtract the monomer rdfs from the complex, and compute the integrals.
+If you also visualize the individual rdfs, you will notice that the difference converges much better with increasing radius than the contributions.
+This is because the cutoff contains regions that are close to the atom, since the integration region does not comprise the whole protein.
+\begin{lstlisting}[style=python]
+difference = compl_rdf - biotin_rdf - strept_rdf
+cutoff = 12
+integral = difference[bins < cutoff].sum()
+print("Integral = {}".format(integral))
+plt.plot(bins, np.cumsum(difference))
+plt.axvline(cutoff)
+plt.xlabel('distance to biotin [A]')
+plt.ylabel('dG contribution [kcal/mol]')
+\end{lstlisting}
 %\section{Content and links}
@@
 \section*{Potentially Conflicting Interests}
-%%%%%%%
+%%%%%%%j
 %Declare any potentially competing interests, financial or otherwise

### 2022-04-26 manuscript/manuscript.tex

@@
 \subsection{Solvation Energy}
-\subsubsection{PME implementation}
+The solvation energy is readily calculated by summing up the contributions of the water-water interaction and the water-solute interaction.
+\begin{equation}
+	\Delta E_\textit{solv} = \Delta E_\textit{sw} + \Delta E_\textit{ww}
+\end{equation}
+
+\begin{equation}
+	\Delta E_\textit{sw} \equiv \frac{1}{8\pi^2} \int g_\textit{sw}\left(\omega|\textbf{r}\right) U_\textit{sw}\left(\textbf{r}, \omega\right) d\omega
+\end{equation}
+
+Similar to the entropy integrals, the solute-water solvation energy integral also decays with increasing distance from the solute.
+Hence, the solvation energy can be approximated by local spatial integrals.
+However, there is a complication arising from this approximation.
+Water molecules outside the region of interest will also be accounted for in the pairwise water term.
+For example, given two different regions $R_a$ and $R_b$, will not sum up to the total energy of the two regions due to double counting.
+
+\begin{equation}
+\begin{aligned}
+	\Delta E_\textit{ww} \equiv & \left(\frac{1}{8\pi^2} \right)^2 \rho^o \int g_\textit{sw}\left(\omega|\textbf{r}\right) \\
+	& \times \left[g_\textit{sw}\left(\textbf{r}^\prime , \omega^\prime \right) - g^o_\textit{ww} \left(\textbf{r}, \omega, \textbf{r}^\prime, \omega^\prime \right)\right] \\
+	& \times U_\textit{ww}\left(\textbf{r}, \omega, \textbf{r}^\prime, \omega^\prime\right)d\omega d\textbf{r}^\prime d\omega^\prime
+\end{aligned}
+\end{equation}
+There are two ways in which this double counting was accounted for in the original paper.
+First, using the water-displacement method.
+This method is applicable when a determining the energetic cost of kicking out a water molecule.
+\begin{equation}
+	\Delta E^R_\textit{ww} = n^\textit{R} E^\textit{bulk}_\textit{ww} - E^\textit{R,corr}_\textit{ww}
+\end{equation}
+
+The second method uses normalized-water properties to account the difference in water properties of two different regions.
+The normalized water property is defined as the water property per water molecule.
+\begin{equation}
+	\Delta E^\textit{R,norm}_\textit{ww} = E^\textit{R,norm}_\textit{ww} - 2E^\textit{bulk}_\textit{ww}
+\end{equation}
+
+\subsection{PME implementation}
 
@@
 In \software{gisttools}, this is done simply by setting \inlinecode{gist.eww\_ref}.
-In \software{gistpp}, this can be done as follows: \todo{This might become invalid with future GIST versions, depending on whether we adopt the COM as molecular center.} \todo{Better use the population.}
+In \software{gistpp}, this can be done as follows: \todo{This might become invalid with future GIST versions, depending on whether we adopt the COM as molecular center.}
 
@@
 
-It might also be interesting to visualize the average energy of a water molecule at each grid voxel.
-This quantity is given by 2*Eww + Esw.
-The energy referencing leads to non-zero normalized values where the population is zero.
-While this is irrelevant for further post-processing, it is advantageous for visualization to set those empty regions to zero.
-A .dx file can be produced using gisttools as follows:
-
-\begin{lstlisting}[style=python]
-import gisttools as gt
-gist = gt.gist.load_gist_file("gist.dat",
-    struct="solute-centered.pdb", eww_ref=-9.533)
-gist['E_norm'] = gist['Eww_norm'] * 2 + gist['Esw_norm']
-gist.loc[gist['population'] == 0, 'E_norm'] = 0
-gist.save_dx('E_norm', 'gist-E-per-mol-norm.dx')
-\end{lstlisting}
-
 Visualize the free energy, the total energy, and the entropy, at several iso-levels.
-An minimal PyMOL input might look like this:
-
-\begin{lstlisting}
-load solute-centered.pdb
-show sticks, resname BTN
-as surface, not resname BTN
-load gist-gO.dx, gO
-isosurface gO_high, gO, 2.
-\end{lstlisting}
-
-Adapt the above PyMOL script to visualize the average energy per solvent at a positive and a negative isolevel, as well as regions of negative entropy per solvent.
-The expected representation of the streptavidin binding pocket is shown in Figure \ref{fig_binding_pocket_pymol}.
-
-\begin{figure}
-	\centering
-	\includegraphics[width=0.8\linewidth]{figures/binding_pocket_S_-2_E_-2_E_3.png}
-	\caption{Free energy contributions in the streptavidin binding pocket.
-Red: high solvent energy (at +3 kcal/mol).
-Blue: low solvent energy (at -2 kcal/mol).
-Green: low solvent entropy $T\Delta S^{six}$ (at -2 kcal/mol).}
-	\label{fig_binding_pocket_pymol}
-\end{figure}
-
-You should find that there are several regions in the binding pocket that feature a high solvent energy and/or low solvent entropy.
-However, such regions also exist outside of the binding pocket, and there are also low-energy regions within the binding pocket.
-Therefore, the strong affinity to biotin can not be explained solely by the hydration contribution, and the direct interaction energy also needs to be taken into account.
+The expected result is shown in Figure ???.
 
 \subsection{Contribution of Hydration to Binding}
-In the next step, calculate the $\dgsolv$ contributions around the biotin ligand in each of the three systems (biotin, streptavidin, and complex).
-Note that our integration region does not fully reach into bulk, but also comprises regions that are close to the streptavidin protein. 
-Therefore, our results for each system will depend on the exact position of the integration boundary. 
-To avoid inconsistencies, it is important to choose exactly the same integration region for each system.
-Although we kept the systems rigid, the center of mass might have been shifted during pressure equilibration.
-Therefore, we align the complex structure to the streptavidin system and use the shifted biotin coordinates to define the integration region.
 
-We recommend using a \software{Jupyter Notebook} for the following analyses, although they can also be done in a script or an interactive Python prompt.
-The code is not explained in detail here, but should be clear after reading the previous sections.
-
-Load the files and double check the frame numbers and reference density.
-\begin{lstlisting}[style=python]
-import numpy as np
-from gisttools.gist import load_gist_file
-import matplotlib.pyplot as plt
-
-compl = load_gist_file('complex/gist/gist.dat', \
-    struct='complex/gist/solute-centered.pdb')
-print(compl.n_frames, compl.rho0)
-biotin = load_gist_file('biotin/gist/gist.dat', \
-    struct='biotin/gist/solute-centered.pdb')
-print(biotin.n_frames, biotin.rho0)
-strept = load_gist_file('streptavidin/gist/gist.dat', \
-    struct='streptavidin/gist/solute-centered.pdb')
-print(strept.n_frames, strept.rho0)
-\end{lstlisting}
-
-Assign reference energies and check them for plausibility.
-\begin{lstlisting}[style=python]
-biotin.eww_ref = biotin.detect_reference_value()
-print("Biotin:", biotin.eww_ref)
-strept.eww_ref = strept.detect_reference_value()
-print("Streptavidin:", strept.eww_ref)
-compl.eww_ref = compl.detect_reference_value()
-print("Complex:", compl.eww_ref)
-\end{lstlisting}
-
-Subtract a reference entropy from the dTSsix columns.
-\begin{lstlisting}[style=python]
-def reference_entropy(gf):
-    if 'dTSsix_unref_norm' not in gf.data.columns:
-        gf['dTSsix_unref_norm'] = gf['dTSsix_norm']
-        gf['dTSsix_unref_dens'] = gf['dTSsix_dens']
-    gf['dTSsix_norm'] = gf['dTSsix_unref_norm'] \
-        - gf.detect_reference_value('dTSsix_unref_dens')
-    gf['dTSsix_dens'] = gf.norm2dens(gf['dTSsix_norm'])
-reference_entropy(biotin)
-reference_entropy(strept)
-reference_entropy(compl)
-\end{lstlisting}
-
-Compute the atom positions that define the integration region.
-For the streptavidin integral, we align the complex structure onto the streptavidin structure and then use the biotin positions as centers.
-Note that we use \software{mdtraj} here, since \software{gisttools} stores the reference structure as a \inlinecode{mdtraj.Trajectory} object, but any other library for post-processing of trajectories works as well.
-\begin{lstlisting}[style=python]
-col = 'dTSsix_dens'
-def select(traj, sel):
-    "Slice a Trajectory by selection mask."
-    return traj.atom_slice(traj.top.select(sel))
-# we multiply by 10 to convert nm to Angstrom.
-compl_x = select(compl.struct, biotin_mask).xyz[0] * 10.
-biotin_x = select(biotin.struct, biotin_mask).xyz[0] * 10.
-aligned = compl.struct[:].superpose(strept.struct, \
-    atom_indices=strept.struct.top.select(strept_mask))
-aligned = select(aligned, biotin_mask)
-strept_x = aligned.xyz[0] * 10.
-
-bins, biotin_rdf = biotin.rdf( \
-    col, centers=biotin_x, bins=100, rmax=24)
-bins, strept_rdf = strept.rdf( \
-    col, centers=strept_x, bins=100, rmax=24)
-bins, compl_rdf = compl.rdf( \
-    col, centers=compl_x, bins=100, rmax=24)
-\end{lstlisting}
-
-Now, subtract the monomer rdfs from the complex, and compute the integrals.
-If you also visualize the individual rdfs, you will notice that the difference converges much better with increasing radius than the contributions.
-This is because the cutoff contains regions that are close to the atom, since the integration region does not comprise the whole protein.
-\begin{lstlisting}[style=python]
-difference = compl_rdf - biotin_rdf - strept_rdf
-cutoff = 12
-integral = difference[bins < cutoff].sum()
-print("Integral = {}".format(integral))
-plt.plot(bins, np.cumsum(difference))
-plt.axvline(cutoff)
-plt.xlabel('distance to biotin [A]')
-plt.ylabel('dG contribution [kcal/mol]')
-\end{lstlisting}
 %\section{Content and links}
@@
 \section*{Potentially Conflicting Interests}
-%%%%%%%j
+%%%%%%%
 %Declare any potentially competing interests, financial or otherwise

### 2022-05-11 manuscript/manuscript.tex
More tutorial text, update references
* remove note from the references
* remove "type = {Journal Article}" from the references. This messes
with the bibtex style somehow...
* extra text: expected results, with some interpretation.
* some formatting updates: Figure~\ref..., \@ in sentences that end with
GIST.

@@
   ntr=1, restraint_wt=100.0,
-    restraintmask=\"!@H=&!:WAT\",
+    restraintmask="!@H=&!:WAT",
   ntp=1, pres0=1.0, taup=1.0,
@@
 
-The expected output of this code is shown in Figure \ref{fig_ewwref}.
+The expected output of this code is shown in Figure~\ref{fig_ewwref}.
 
@@
 	\includegraphics[width=0.8\linewidth]{figures/Eww_convergence.eps}
-	\caption{Convergence of $E_{ww}$ with increasing distance to the solute. The horizontal line shows the automatically computed reference energy. Computed from the complex calculation.}
-	\label{fig_ewwref}
+	\caption{Convergence of $E_{ww}$ with increasing distance to the solute.
+The horizontal line shows the automatically computed reference energy.
+Computed from the complex calculation.}\label{fig_ewwref}
 \end{figure}
@@
 After that, check whether your free energy contributions become negligible at high distances to the solute.
-In a plot like Figure \ref{fig_ewwref} based on the \inlinecode{A\_dens} column, the values should tend to zero.
+In a plot like Figure~\ref{fig_ewwref} based on the \inlinecode{A\_dens} column, the values should tend to zero.
 However, it is more informative to plot the cumulative free energy contribution against the distance to the solute, to check the convergence.
@@
 
-The expected output of this code is shown in Figure \ref{fig_radial_convergence}.
+The expected output of this code is shown in Figure~\ref{fig_radial_convergence}.
 In this case, both the Eww energy and the entropy are not perfectly flat, indicating that the reference value is not optimal.
@@
 All quantities cumulative, i.e., summed up to the respective radius.
-Computed from the biotin calculation.}
-	\label{fig_radial_convergence}
+Computed from the biotin calculation.}\label{fig_radial_convergence}
 \end{figure}
@@
 Next, we want to visualize the 3-dimensional contributions to the free energy of hydration, as well as the entropy and enthalpy contributions.
-You can use PyMol\cite{pymol} or VMD\cite{vmd} to visualize the .dx files from GIST.
+You can use PyMol\cite{pymol} or VMD\cite{vmd} to visualize the .dx files from GIST\@.
 With \software{gistpp}, you can use the file from the convergence check. Additional OpenDX files can be produced using \inlinecode{gistpp -op makedx}.
@@
 While this is irrelevant for further post-processing, it is advantageous for visualization to set those empty regions to zero.
-A .dx file can be produced using gisttools as follows:
+An OpenDX file can be produced using \software{gisttools} as follows:
 
@@
 Adapt the above PyMOL script to visualize the average energy per solvent at a positive and a negative isolevel, as well as regions of negative entropy per solvent.
-The expected representation of the streptavidin binding pocket is shown in Figure \ref{fig_binding_pocket_pymol}.
+The expected representation of the streptavidin binding pocket is shown in Figure~\ref{fig_binding_pocket_pymol}.
 
@@
 Blue: low solvent energy (at -2 kcal/mol).
-Green: low solvent entropy $T\Delta S^{six}$ (at -2 kcal/mol).}
-	\label{fig_binding_pocket_pymol}
+Green: low solvent entropy $T\Delta S^{six}$ (at -2 kcal/mol).}\label{fig_binding_pocket_pymol}
 \end{figure}
@@
 \end{lstlisting}
+
+Compute the energy (Eall) and entropy (dTSsix) contributions separately.
+You will find that the energy is strongly unfavorable to binding, because we do not yet include the interaction energy between biotin and streptavidin.
+
+Using the \inlinecode{energy} command in \software{cpptraj}, you can compute this energy difference.
+We recommend using PME in combination with PME-GIST, but not with GPU-GIST\@.
+An example \software{cpptraj} input might look like:
+\begin{lstlisting}
+parm solvated.parm7
+trajin md-01.nc 1 last 100
+energy complex ^1,2 # etype pme
+energy strept ^1 # etype pme
+energy biotin ^2 # etype pme
+go
+diff = complex[total] - strept[total] - biotin[total]
+writedata energy.dat diff complex[total] \
+strept[total] biotin[total]
+avg(complex[total])
+avg(strept[total])
+avg(biotin[total])
+avg(diff)
+\end{lstlisting}
+
+The expected results are summarized in Table~\ref{tab_dg_monomers_dimer}.
+
+\begin{table}
+	\caption{Free energy contributions for the monomers and the dimer.}\label{tab_dg_monomers_dimer}
+	\begin{tabular}{lrrrr}
+		System       & E(internal) & $\Delta$E(GIST) & $\Delta$S(GIST) & total \\
+		complex      & -1303.4 & -361.2 & -233.1 & -226.7 \\
+		streptavidin & -1180.7 & -360.8 & -242.4 & -217.1 \\
+		biotin       & -24.0   &  -98.1 &  -35.1 & -161.1 \\
+		Diff         & -98.7   &   97.7 &   44.4 &  -45.4
+	\end{tabular}
+\end{table}
+
+Even though streptavidin-biotin is known to be a very stable complex, the solvation energy favors the dissociation.
+This is expected: if the molecules were not able to form any interactions, but interacted fully with the water, they would not form a complex.
+
+In this case, we find that the energy of solvation and the internal energy of the complex more or less cancel out, indicating that binding is largely entropy-driven in this case.
+This is somewhat surprising, since isothermal titration calorimetry (ITC) of biotin-streptavidin shows strong enthalpic binding contributions.\cite{mpye2020-biotin-itc}\cite{hyre2006-biotin-itc}
 %\section{Content and links}

### 2022-05-19 manuscript/manuscript.tex
Bugfixes in the tutorial part

@@
 % Please note that these options may affect formatting.
+
+% NOTES
+% when importing references from EndNote, remove the type = {Journal Article}. It messes with the livecoms bibtex style somehow...
+% Place a \@ between all-caps words and punctuation ("This is a sentence about GIST\@. This is ...")
+% Place a protected space (~) between "Figure" and the reference. The same with tables.
 
@@
 For a system centered at the origin with a spacing of \SI{0.5}{\angstrom} and a buffer of, e.g., \SI{7}{\angstrom} between the solute and the grid wall, you need $4.0\times (d+7)$ bins in the respective direction, where $d$ is the larger of $-$min and $+$max in the respective direction.
+\todo{this does not work with the default gisttools values for the automatic detection of the energy reference. Better use a larger value. }
 We note that this centered grid is usually larger than it needs to be in some directions.
@@
 Alternatively, one can note that the solvent-solvent energy in a sufficiently large GIST grid converges to the correct bulk value at large distance to the solute.
-This means that the reference energy can be obtained from a large GIST grid by binning the voxels by their distance to the solute, and evaluate Equation \ref{eq_ewwref} within each bin.
+This means that the reference energy can be obtained from a large GIST grid by binning the voxels by their distance to the solute, and evaluate Equation~\ref{eq_ewwref} within each bin.
 If this value converges to a constant, this can be used as $E_{ww}^\textit{ref}$.
@@
 bins, eww = gist.rdf(
-    "Eww_unref", bins=100, rmax=20., normalize="norm")
+    "Eww_unref_norm", bins=100, rmax=20., normalize="norm")
 plt.plot(bins, eww)
@@
 However, it is more informative to plot the cumulative free energy contribution against the distance to the solute, to check the convergence.
-It the curve flattens out, this is your final $\dgsolv$.
+It the curve flattens out, this is your final $\dgsolv$\@.
 If it does not, you might need to tweak the Eww reference, or introduce a reference value for the entropy (especially with older versions of \software{cpptraj}).
@@
     struct="solute-centered.pdb", eww_ref=-9.533)
-bins, (dg, eww, esw, s) = gist.rdf(
+bins, (dg, esw, eww, s) = gist.rdf(
     ["A_dens", "Esw_dens", "Eww_dens", "dTSsix_dens"],
@@
 
+% The current font makes the O look like a tofu / placeholder
 \begin{lstlisting}

### 2022-05-19 manuscript/manuscript.tex
Update manuscript.tex
PME implementation discussion

--- 2022-04-26 manuscript/manuscript.tex
@@
 \section{Introduction}
-\todo{ (vah) Sources and better description of all these methods, more literature research, better flow }
-
-Solvation thermodynamics govern any process involving solutes in a solvent.
-Solutes are constantly interacting with solvents in their environment, necessitating a good description of these interactions for accurate computational predictions. 
-Especially the interaction with water is of utmost importance, as many biological processes occur in a aquaeous environment. 
-The free energy of solvation thereby strongly influences real-world problems of interest such as hydrophobic effects, the binding of a ligand to a protein, protein-protein interactions, protein dynamics or even more generally, a protein's fold.
-The computational calculation of thermodynamical solvation properties is therefore of particular interest and has been tackled by a wide array of different methods.   
-Quantum mechanical methods provide the most rigorous description of the underlying potential energy surface (PES) but suffer from high computational cost, allowing only the treatment of small systems and few solvent molecules.
-More recently, machine learning approaches and QM/MM approaches are trying to bridge the gap to computationally less demanding methods while keeping the accuracy of ab initio QM methods.
-Molecular mechanics based approaches describe the underlying PES through empirically fit force-fields, which reduce the computational demand of these calculations tremendously.
-This allows for the treatment of even large biological systems like fully solvated proteins. 
-Implicit solvent methods such as MM/PBSA  or MM/GBSA can quickly provide results for a large amount of supplied structures, but yield inaccurate results when compared to more rigorous explicit solvent methods.
-Somewhat inbetween the implicit solvent methods and the fully explicit solvent methods are Reference Interaction Site Models.
-One of these methods, 3D-RISM, foregoes the explicit simulation and uses 3D solvent distributions based on the Ornstein--Zernike integral equations to predict solvation thermodynamics instead.
-The 3D-RISM approach allows for a decomposition into entropic and enthalpic terms as well as a full localization of the underlying distributions to the solute structure.
-The most statisticaly rigorous of the explicit solvent MM methods are alchemical methods such as FEP or TI. 
-Assuming sufficient sampling, they can accurately predict the free energy of solvation in the boundaries of the assumptions underlying their forcefield. 
-However, the seperation of enthalpic and entropic contributions to the free energy of solvation is difficult with these methods and localizing these contributions not easily possible.
-In this tutorial, the Grid Inhomogenous Solvation Theory (GIST) is applied to calculate solution thermodynamic properties of interest. 
-GIST is based on Lazaridis' Inhomogenous Solvation Theory (IST), solving the arising integral equations on a grid. 
-This is highly advantegous compared to other IST-based methods such as e.g. WaterMap,\todo{Add more methdods!} as this is not restricted to high-occopancy solvation sites but provides a 
-By simulating the dynamics of a solvent around the solute, GIST calculates thermodynamic properties for a standard solvation process as described by Ben-Naim.
-The method allows for a full localization of entropic and enthalpic contributions on the grid around the solute.
-For small organic molecules in water, the method provided results compareable to TI in accuracy, assuming some small correction factor to the entropy is applied. 
-Recent improvements to the method have improved the calculation speed tremendously, allowing for the treatment of larger systems of biological interest such as proteases or antibodies. 
-Other improvements worked on decreasing differences in the treatment of long-range interaction energies between GIST and popular molecular dynamics engines. 
-Yet further changes allow the inclusion of other rigid solvents than water or allow for the treatment of solvent mixtures and ions in solution. 
-While previous tutorials gave insightful pointers as to how to setup and run GIST calculations, the post-processing of such calculations was often treated less rigorously.
-While building on these tutorials, an updated guide to calculating solvation properties using GIST is therefore presented here to address the recent improvements to the method.
 
@@
 \subsection{Scope}
-
-The tutorial provides an example of a GIST workflow that can be easily adapted towards different systems. 
-In addition to a short introduction to the method and the theory behind it an overview of the various different flavours of GIST methods is provided.
-Towards this, a GIST study of the streptavidin-biotin complex is presented, which reflects on many common aspects of GIST studies.
+The aim of this tutorial is to provide an example of a GIST workflow that can be easily adapted towards different systems.
+We present a GIST study of the streptavidin-biotin complex, which allows us to show many common aspects of GIST studies.
 We present solvation analysis of a small molecule (biotin) as well as a biomolecule (streptavidin).
 For both systems, we show how to interpret the three-dimensional contributions of solvation free energy in a binding pocket.
-The contribution of solvation free energy to binding is computed, which requires accurate post-processing of the GIST output to avoid unfavorable summation of bias in the GIST calculation.
-We believe that this example study covers a wide range of applications of the GIST method.
-Furthermore, we discuss several technical aspects  and typical pitfalls of GIST analyses, such as the normalization of voxel values and how to deal with double counting of energy contributions. 
-Additionally, we provide a python library to unify the analysis of GIST data produced by various versions of GIST and make the post-processing of GIST results more accessible.
+Furthermore, we compute the contribution of solvation free energy to binding, which requires accurate post-processing of the GIST output to avoid unfavorable summation of bias in the GIST calculation.
+We therefore believe that this example study covers a wide range of applications of the GIST method.
 
-After completing this tutorial, we expect the reader to be able to implement their own GIST study.
-
+Furthermore, we discuss several technical aspects of GIST analyses, such as the normalization of voxel values and how to deal with double counting of energy contributions. 
 %Tutorials should endeavor to cover the specific task at hand, and also highlight how the steps might need to be modified (or additional care might need to be taken at particular points) to handle more general cases.
@@
 \section{Prerequisites}
-\todo{ (vah)  Improve flow and logical progression of prerquisites.}
 
@@
 \subsection{Background knowledge}
-This tutorial is aimed at advanced users with a solid background knowledge of molecular dynamics (MD) simulations. The user should be able to run MD simulations and work with their output data.
-Previous knowledge of the method is not necessary, but we recommend the perusal of the primary sources on the method outlined in the Theory section.
-
-The presented analyses are run in Python.
-A working understanding of Python is therefore necessary to follow the tutorial. 
-A Jupyter notebook showing directly the herein presented analyses is provided as well. 
-To easily follow the analyses an understanding of working with Jupyter notebooks is therefore beneficial, though not strictly necessary.
+%Tutorials should clearly define what concepts or abilities researchers will need to complete the tutorial (e.g., some proficiency in Python; experience with Jupyter notebooks; knowledge of classical MD; etc).
 
@@
 Tutorials requiring specific software packages must provide instructions and files for the referenced version of the software.
-GIST's main implementation is part of the MD analysis software \software{cpptraj} as part of the AmberTools software suite.
-A recent version of \software{cpptraj} should be used, such as the one in AmberTools22 or an installation from GitHub.
-The closed-source Amber simulation engine is used in this tutorial to run the necessary MD simulations. 
-If no Amber installation is available, the MD can be performed other simulation engines compatible with \software{cpptraj}, we recommend GROMACS.
-If using GROMACS, we propose to prepare the structures and topologies using AmberTools and then converting to GROMACS using \software{acpype.py}.
-In that way, a full Amber topology is available for the GIST analysis.
-
-Furthermore, a recent Python version (>3.6) should be available with the following packages.
-\begin{itemize}
-	\item \software{gisttools} (and its prerequisites)
-	\item \software{mdtraj}
-\end{itemize}
-If the reader prefers to skip the MD and GIST calculation, we also provide the GIST output files with this tutorial, such that the post-processing can be done without any expensive calculations.
-A Jupyter notebook showing the herein presented analyses is provided as well, necessitating a current installation of Jupyter to run.
 
@@
 
+In the latest implementation of GIST, the particle mesh ewald (PME) functionaliy was added that decomposed the contributions of the PME-based electrostatic energy and the long-range Lennard-Jones energy to the total solvation energy.
+This allows  for a faster and more accurate calculation of the solvation energy.
+
+\begin{equation}
+	E_\textit{total} = E_\textit{elec} + E_\textit{lj}
+\end{equation}
+
+In the original version of GIST, the energies are calculated based on the shortest distance or the minimum image convention.
+In doing so, long range interactions are not accounted for.
+In the PME functionality, the contribution of the long-range interactions are accounted for.
+
+In terms of the electrostatic energy, in the original version, the electrostatic energy longer than half of the cell's diagonal are not accounted for.
+In the PME functionality, the long-range interactions are accounted for by decomposing the electrostatic interaction into three parts: the direct term, the reciprocal term, and the correction term.
+
+\begin{equation}
+	E_\textit{elec} = E_\textit{dir} + E_\textit{rec} + E_\textit{corr}
+\end{equation}
+
+On the other hand, in PME-GIST, the long-range interactions for the Lennard-Jones contribution are treated by accounting for the short-range and the long-range correction contributions separately.
+The short-range contribution is the LJ energy within the set cutoff and the long-range correction term accounts for the contributions above this cutoff.
+
+\begin{equation}
+	E_\textit{lj} = E_\texit{lj\_short} +  E_\textit{corr}
+\end{equation}
+
 Here, we aim to guide the reader through a GIST analysis of the streptavidin-biotin complex. The goal is to
@@
 It is therefore important that the conformations of the monomers in the complex and dissociated states match exactly.
-Use \software{cpptraj} to produce biotin and streptavidin structures based on the complex by stripping the respective other molecule like this:
-\begin{lstlisting}
-parm complex/solvated.parm7
-trajin complex/equil/EQUIL-DONE.rst
-strip ^2 parmout streptavidin/solvated.parm7
-trajout streptavidin/pre-equil.ncrst
-\end{lstlisting}
-(Adapt for biotin as necessary.)
+We provide a script called \software{cpptraj\_remove\_mol.sh} that removes a single molecule from a topology and structure using \software{cpptraj} without altering the rest of the system.
+Use this to create monomer topologies of biotin and streptavidin.
 
-%We provide a script called \software{cpptraj\_remove\_mol.sh} that removes a single molecule from a topology and structure using \software{cpptraj} without altering the rest of the system.
-\todo{Explain cpptraj rather than our script.
-The script is for the automated workflow only.}
+\subsection{Running GIST}
+\subsection{Visualizing $\dgsolv$}
+\subsection{Contribution of Hydration to Binding}
 
-Use the \software{equilibration.py} script (or your own) to perform short NVT and NpT equilibration runs on the starting structure.
-Note that the barostat might interfere with restrained equilibration of the complex, since different molecules are moved relative to each other when the volume changes.
-But since our equilibration script also contains unrestrained pressure equilibration, we should not run into any trouble.
-
-Then split the equilibrated structure into biotin and streptavidin parts using \software{cpptraj\_remove\_mol.sh}.
-After that, you can use \software{equilibration.py -R} to run short restrained equilibrations (where only the water is allowed to move) on the individual systems.
-
-\subsection{Running MD for GIST}
-
-Now, you can run restrained MD simulations based on the equilibrated structures for subsequent GIST analyses.
-We apply a harmonic restraint of \SI{100}{\kilo\calorie\per\mole\per\angstrom\squared} to all heavy atoms to keep the system in place.
-We run \SI{100}{\nano\second} of simulation and store the conformations every \SI{100}{\pico\second}.
-A simulation length of at least \SIrange{10}{20}{\nano\second} is recommended, to obtain statistically independent snapshots of the water.
-
-An example Amber input file for GIST might look like this:
-\begin{lstlisting}
-restrained 100 ns NpT
-&cntrl
-  ntx=5, irest=1,
-  ioutfm=1,
-  ntb=2, iwrap=1,
-  ntr=1, restraint_wt=100.0,
-    restraintmask=\"!@H=&!:WAT\",
-  ntp=1, pres0=1.0, taup=1.0,
-  ntc=2, ntf=2,
-  ntt=3, tempi=300.0, temp0=300.0, gamma_ln=2,
-  nstlim=50000000, dt=0.002,
-  ntwr=50000, ntwx=5000,
-  ntpr=5000,
-/
-\end{lstlisting}
-Note that these are very standard MD settings except for the line that defines the restraints.
-
-Run the MD using \software{pmemd.cuda}:
-\begin{lstlisting}
-pmemd.cuda \
-	-O \
-	-i 100ns-npt-restraint.in \
-	-o md-01.out \
-	-p solvated.parm7 \
-	-c EQUIL-DONE.rst \
-	-x md-01.nc \
-	-r md-01.ncrst \
-	-ref EQUIL-DONE.rst
-\end{lstlisting}
-\subsection{Running the GIST analysis}
-
-For easier post-processing, we will use grids that contain the whole molecule.
-This could in theory be avoided for streptavidin and the complex, since only the surrounding of the biotin molecule (i.e., the binding pocket) is required for the analysis.
-We also recommend that you center the molecule in space.
-For each system (biotin, streptavidin, complex), produce a centered pdb file using \software{cpptraj}:
-\begin{lstlisting}
-parm solvated.parm7
-trajin md-01.nc 1 1 1
-autoimage !(:WAT) origin
-strip :WAT
-trajout solute-centered.pdb
-\end{lstlisting}
-Then, find the extent (minimum and maximum) of the x-, y-, and z-coordinates for each system.
-This can be done e.g., using the \software{mdtraj} package in Python.
-\begin{lstlisting}[style=python]
-import mdtraj as md
-pdb = md.load("solute-centered.pdb")
-# mdtraj converts to nm.
-# *10 to convert back to Angstrom.
-min = pdb.xyz[0].min(0) * 10.
-max = pdb.xyz[0].max(0) * 10.
-print(*list(zip("xyz", min, max)), sep="\n")
-\end{lstlisting}
-Then, decide on grid dimensions for every system.
-For a system centered at the origin with a spacing of \SI{0.5}{\angstrom} and a buffer of, e.g., \SI{7}{\angstrom} between the solute and the grid wall, you need $4.0\times (d+7)$ bins in the respective direction, where $d$ is the larger of $-$min and $+$max in the respective direction.
-We note that this centered grid is usually larger than it needs to be in some directions.
-If you want to save some extra memory and calculation time, feel free to align the grid more precisely to the minimum and maximum coordinates, or use a smaller grid that only covers the binding pocket.
-You can also rotate the molecule by its principal axes to fit the cuboid grid more exactly.
-However, this needs to be done \emph{before} the MD run, since rotating the trajectory damages the periodic box information.
-
-Now, you can finally run the actual GIST analysis. Center the molecule exactly in the same way as above, to make sure that its position matches the grid parameters that you decided on.
-We recommend using either the PME or GPU implementation of GIST, depending on whether you have a fast GPU.
-A further advantage of the PME implementation is that it matches the energy calculation of the MD engine more closely.
-\todo{At the time of writing, it is important to use an up-to-date version of PME-GIST, or the Esw-Eww decomposition will not be available.
-Maybe remove this for the final version, when Amber22 is out.}
-\begin{lstlisting}
-parm solvated.parm7
-trajin md-01.nc
-autoimage !(:WAT) origin
-# remove the "pme" option if you want to use the GPU
-gist griddim 100 100 100 out gist.dat \
-  refdens 0.03287 pme
-# other settings you might want to change
-# (those are the defaults)
-# gridcntr 0 0 0 gridspacn 0.5 temp 300
-\end{lstlisting}
-\subsection{Reference Values and Radial Convergence}
-All GIST quantities should be expressed relative to bulk.
-In practice, this is not problematic for the entropies, since they are computed relative to a distribution of randomly oriented molecules at the reference density.
-The solute-solvent energy Esw is also no problem, because it naturally tends to zero in bulk. 
-However, the solvent-solvent energy Eww needs to be referenced.
-
-The Amber manual provides reference energies for several solvents.
-However, the exact reference value is different when using PME-GIST, and also depends on the box size.
-For exact analyses such as computing the total $\dgsolv$ of a compound, it is important to compute an exact reference energy.
-
-The most accurate method to compute a reference energy is to run a separate GIST calculations with the same energy method (PME/GPU) and a similar box size, and then compute the average solvent-solvent energy per molecule.
-
-\begin{equation}
-\label{eq_ewwref}
-E_{ww}^\textit{ref} = \frac{\int{E_{ww}} \mathrm{d}x}{\left<N_\textit{solvent}\right>}
-  = \frac{\sum_\textit{voxels}E_{ww}^\textit{dens} V_\textit{vox}}{\sum_\textit{voxels}{N_\textit{solvent}} / N_\textit{frames}}
-\end{equation}
-
-Alternatively, one can note that the solvent-solvent energy in a sufficiently large GIST grid converges to the correct bulk value at large distance to the solute.
-This means that the reference energy can be obtained from a large GIST grid by binning the voxels by their distance to the solute, and evaluate Equation \ref{eq_ewwref} within each bin.
-If this value converges to a constant, this can be used as $E_{ww}^\textit{ref}$.
-
-\software{gisttools} contains functionality to perform this analysis easily.
-There is also a method \inlinecode{detect\_reference\_value} that automatically tries to find the converged value fo the rdf.
-Although it has a simple convergence check built-in, it is always recommended to check the convergence by hand as well.
-
-\begin{lstlisting}[style=python]
-import gisttools as gt
-import matplotlib.pyplot as plt
-gist = gt.gist.load_gist_file(
-    "gist.dat", struct="solute-centered.pdb")
-bins, eww = gist.rdf(
-    "Eww_unref", bins=100, rmax=20., normalize="norm")
-plt.plot(bins, eww)
-eww_auto = gist.detect_reference_value().values[0]
-plt.axhline(eww_auto, color='k')
-plt.gca().set(
-    xlabel='distance [A]', ylabel='E_ww per mol')
-plt.show()
-\end{lstlisting}
-
-The expected output of this code is shown in Figure \ref{fig_ewwref}.
-
-\begin{figure}
-	\centering
-	\includegraphics[width=0.8\linewidth]{figures/Eww_convergence.eps}
-	\caption{Convergence of $E_{ww}$ with increasing distance to the solute. The horizontal line shows the automatically computed reference energy. Computed from the complex calculation.}
-	\label{fig_ewwref}
-\end{figure}
-
-After choosing an appropriate reference value, we need to subtract this value from Eww.
-Make sure to subtract the reference from the normalized (\_norm) data.
-In \software{gisttools}, this is done simply by setting \inlinecode{gist.eww\_ref}.
-In \software{gistpp}, this can be done as follows: \todo{This might become invalid with future GIST versions, depending on whether we adopt the COM as molecular center.} \todo{Better use the population.}
-
-\begin{lstlisting}
-rho0=0.03287465
-neg_eww_ref=9.533
-gistpp -op multconst -i gist-gO.dx -opt const $rho0 \
-    -o gist-gO-abs.dx 
-gistpp -op div -i gist-Eww-dens.dx -i2 gist-gO-abs.dx \
-    -o gist-Eww-norm.dx
-gistpp -op addconst -i gist-Eww-norm.dx \
-    -opt const $neg_eww_ref -o gist-Eww-ref-norm.dx
-gistpp -op mult -i gist-Eww-ref-norm.dx \
-    -i2 gist-gO-abs.dx -o gist-Eww-ref-dens.dx
-\end{lstlisting}
-
-Now, add all free energy contributions to obtain the spatially resolved $\dgsolv$.
-In \software{gisttools}, this is done automatically (simply use the \inlinecode{A\_dens} and \inlinecode{A\_norm} columns).
-In \software{gistpp}, use the \inlinecode{add} command, with the same syntax as the \inlinecode{div} command used above.
-After that, check whether your free energy contributions become negligible at high distances to the solute.
-In a plot like Figure \ref{fig_ewwref} based on the \inlinecode{A\_dens} column, the values should tend to zero.
-However, it is more informative to plot the cumulative free energy contribution against the distance to the solute, to check the convergence.
-It the curve flattens out, this is your final $\dgsolv$.
-If it does not, you might need to tweak the Eww reference, or introduce a reference value for the entropy (especially with older versions of \software{cpptraj}).
-
-\begin{lstlisting}[style=python]
-import gisttools as gt
-import matplotlib.pyplot as plt
-# adapt eww_ref!
-gist = gt.gist.load_gist_file("gist.dat",
-    struct="solute-centered.pdb", eww_ref=-9.533)
-bins, (dg, eww, esw, s) = gist.rdf(
-    ["A_dens", "Esw_dens", "Eww_dens", "dTSsix_dens"],
-    bins=100, rmax=20., normalize="none")
-plt.plot(bins, np.cumsum(dg), label="dG")
-plt.plot(bins, np.cumsum(eww), label="Eww")
-plt.plot(bins, np.cumsum(esw), label="Esw")
-plt.plot(bins, np.cumsum(s), label="dS")
-\end{lstlisting}
-
-The expected output of this code is shown in Figure \ref{fig_radial_convergence}.
-In this case, both the Eww energy and the entropy are not perfectly flat, indicating that the reference value is not optimal.
-The slow energy convergence indicates that a non-optimal reference value was used, but is also complicated by the fact that a charged solute is used without neutralizing counter ions.
-
-\begin{figure}
-	\centering
-	\includegraphics[width=0.8\linewidth]{figures/A_E_S_convergence.eps}
-	\caption{Convergence of $\dgsolv$ and its contributions with increasing distance to the solute.
-All quantities cumulative, i.e., summed up to the respective radius.
-Computed from the biotin calculation.}
-	\label{fig_radial_convergence}
-\end{figure}
-
-\subsection{Visualizing $\dgsolv$}
-Next, we want to visualize the 3-dimensional contributions to the free energy of hydration, as well as the entropy and enthalpy contributions.
-You can use PyMol\cite{pymol} or VMD\cite{vmd} to visualize the .dx files from GIST.
-With \software{gistpp}, you can use the file from the convergence check. Additional OpenDX files can be produced using \inlinecode{gistpp -op makedx}.
-With \software{gisttools}, you can create one using e.g., \inlinecode{gist.save\_dx("A\_dens", "A\_dens.dx")}.
-
-It might also be interesting to visualize the average energy of a water molecule at each grid voxel.
-This quantity is given by 2*Eww + Esw.
-The energy referencing leads to non-zero normalized values where the population is zero.
-While this is irrelevant for further post-processing, it is advantageous for visualization to set those empty regions to zero.
-A .dx file can be produced using gisttools as follows:
-
-\begin{lstlisting}[style=python]
-import gisttools as gt
-gist = gt.gist.load_gist_file("gist.dat",
-    struct="solute-centered.pdb", eww_ref=-9.533)
-gist['E_norm'] = gist['Eww_norm'] * 2 + gist['Esw_norm']
-gist.loc[gist['population'] == 0, 'E_norm'] = 0
-gist.save_dx('E_norm', 'gist-E-per-mol-norm.dx')
-\end{lstlisting}
-
-Visualize the free energy, the total energy, and the entropy, at several iso-levels.
-An minimal PyMOL input might look like this:
-
-\begin{lstlisting}
-load solute-centered.pdb
-show sticks, resname BTN
-as surface, not resname BTN
-load gist-gO.dx, gO
-isosurface gO_high, gO, 2.
-\end{lstlisting}
-
-Adapt the above PyMOL script to visualize the average energy per solvent at a positive and a negative isolevel, as well as regions of negative entropy per solvent.
-The expected representation of the streptavidin binding pocket is shown in Figure \ref{fig_binding_pocket_pymol}.
-
-\begin{figure}
-	\centering
-	\includegraphics[width=0.8\linewidth]{figures/binding_pocket_S_-2_E_-2_E_3.png}
-	\caption{Free energy contributions in the streptavidin binding pocket.
-Red: high solvent energy (at +3 kcal/mol).
-Blue: low solvent energy (at -2 kcal/mol).
-Green: low solvent entropy $T\Delta S^{six}$ (at -2 kcal/mol).}
-	\label{fig_binding_pocket_pymol}
-\end{figure}
-
-You should find that there are several regions in the binding pocket that feature a high solvent energy and/or low solvent entropy.
-However, such regions also exist outside of the binding pocket, and there are also low-energy regions within the binding pocket.
-Therefore, the strong affinity to biotin can not be explained solely by the hydration contribution, and the direct interaction energy also needs to be taken into account.
-
-\subsection{Contribution of Hydration to Binding}
-In the next step, calculate the $\dgsolv$ contributions around the biotin ligand in each of the three systems (biotin, streptavidin, and complex).
-Note that our integration region does not fully reach into bulk, but also comprises regions that are close to the streptavidin protein. 
-Therefore, our results for each system will depend on the exact position of the integration boundary. 
-To avoid inconsistencies, it is important to choose exactly the same integration region for each system.
-Although we kept the systems rigid, the center of mass might have been shifted during pressure equilibration.
-Therefore, we align the complex structure to the streptavidin system and use the shifted biotin coordinates to define the integration region.
-
-We recommend using a \software{Jupyter Notebook} for the following analyses, although they can also be done in a script or an interactive Python prompt.
-The code is not explained in detail here, but should be clear after reading the previous sections.
-
-Load the files and double check the frame numbers and reference density.
-\begin{lstlisting}[style=python]
-import numpy as np
-from gisttools.gist import load_gist_file
-import matplotlib.pyplot as plt
-
-compl = load_gist_file('complex/gist/gist.dat', \
-    struct='complex/gist/solute-centered.pdb')
-print(compl.n_frames, compl.rho0)
-biotin = load_gist_file('biotin/gist/gist.dat', \
-    struct='biotin/gist/solute-centered.pdb')
-print(biotin.n_frames, biotin.rho0)
-strept = load_gist_file('streptavidin/gist/gist.dat', \
-    struct='streptavidin/gist/solute-centered.pdb')
-print(strept.n_frames, strept.rho0)
-\end{lstlisting}
-
-Assign reference energies and check them for plausibility.
-\begin{lstlisting}[style=python]
-biotin.eww_ref = biotin.detect_reference_value()
-print("Biotin:", biotin.eww_ref)
-strept.eww_ref = strept.detect_reference_value()
-print("Streptavidin:", strept.eww_ref)
-compl.eww_ref = compl.detect_reference_value()
-print("Complex:", compl.eww_ref)
-\end{lstlisting}
-
-Subtract a reference entropy from the dTSsix columns.
-\begin{lstlisting}[style=python]
-def reference_entropy(gf):
-    if 'dTSsix_unref_norm' not in gf.data.columns:
-        gf['dTSsix_unref_norm'] = gf['dTSsix_norm']
-        gf['dTSsix_unref_dens'] = gf['dTSsix_dens']
-    gf['dTSsix_norm'] = gf['dTSsix_unref_norm'] \
-        - gf.detect_reference_value('dTSsix_unref_dens')
-    gf['dTSsix_dens'] = gf.norm2dens(gf['dTSsix_norm'])
-reference_entropy(biotin)
-reference_entropy(strept)
-reference_entropy(compl)
-\end{lstlisting}
-
-Compute the atom positions that define the integration region.
-For the streptavidin integral, we align the complex structure onto the streptavidin structure and then use the biotin positions as centers.
-Note that we use \software{mdtraj} here, since \software{gisttools} stores the reference structure as a \inlinecode{mdtraj.Trajectory} object, but any other library for post-processing of trajectories works as well.
-\begin{lstlisting}[style=python]
-col = 'dTSsix_dens'
-def select(traj, sel):
-    "Slice a Trajectory by selection mask."
-    return traj.atom_slice(traj.top.select(sel))
-# we multiply by 10 to convert nm to Angstrom.
-compl_x = select(compl.struct, biotin_mask).xyz[0] * 10.
-biotin_x = select(biotin.struct, biotin_mask).xyz[0] * 10.
-aligned = compl.struct[:].superpose(strept.struct, \
-    atom_indices=strept.struct.top.select(strept_mask))
-aligned = select(aligned, biotin_mask)
-strept_x = aligned.xyz[0] * 10.
-
-bins, biotin_rdf = biotin.rdf( \
-    col, centers=biotin_x, bins=100, rmax=24)
-bins, strept_rdf = strept.rdf( \
-    col, centers=strept_x, bins=100, rmax=24)
-bins, compl_rdf = compl.rdf( \
-    col, centers=compl_x, bins=100, rmax=24)
-\end{lstlisting}
-
-Now, subtract the monomer rdfs from the complex, and compute the integrals.
-If you also visualize the individual rdfs, you will notice that the difference converges much better with increasing radius than the contributions.
-This is because the cutoff contains regions that are close to the atom, since the integration region does not comprise the whole protein.
-\begin{lstlisting}[style=python]
-difference = compl_rdf - biotin_rdf - strept_rdf
-cutoff = 12
-integral = difference[bins < cutoff].sum()
-print("Integral = {}".format(integral))
-plt.plot(bins, np.cumsum(difference))
-plt.axvline(cutoff)
-plt.xlabel('distance to biotin [A]')
-plt.ylabel('dG contribution [kcal/mol]')
-\end{lstlisting}
 %\section{Content and links}

### 2022-05-19 manuscript/manuscript.tex

--- 2022-05-19 manuscript/manuscript.tex
@@
 
+\section{Tutorial}
 Here, we aim to guide the reader through a GIST analysis of the streptavidin-biotin complex. The goal is to
@@
 Compute the atom positions that define the integration region.
-For the streptavidin integral, we align the complex structure onto the streptavidin structure and then use the biotin positions as centers.
+For the streptavidin integral, we align the complex structure onto streptavidin and then use the biotin positions as centers.
 Note that we use \software{mdtraj} here, since \software{gisttools} stores the reference structure as a \inlinecode{mdtraj.Trajectory} object, but any other library for post-processing of trajectories works as well.
@@
 	\caption{Free energy contributions for the monomers and the dimer.}\label{tab_dg_monomers_dimer}
-	\begin{tabular}{lrrrr}
-		System       & E(internal) & $\Delta$E(GIST) & $\Delta$S(GIST) & total \\
-		complex      & -1303.4 & -361.2 & -233.1 & -226.7 \\
-		streptavidin & -1180.7 & -360.8 & -242.4 & -217.1 \\
-		biotin       & -24.0   &  -98.1 &  -35.1 & -161.1 \\
-		Diff         & -98.7   &   97.7 &   44.4 &  -45.4
+	\small
+	\begin{tabular}{lrrrrr}
+		\toprule
+		System       & E(internal) & $\Delta E^\textit{GIST}$ & $\Delta S^\textit{GIST}$ & $\Delta S^\textit{scaled}$ & total \\
+		\midrule
+		complex      & -1303.4 & -361.2 & -233.1 & -139.9 & -192.0 \\
+		streptavidin & -1180.7 & -360.8 & -242.4 & -145.4 & -188.4 \\
+		biotin       & -24.0   &  -98.1 &  -35.1 &  -21.1 & -105.4 \\
+		Diff         & -98.7   &   97.7 &   44.4 &   26.6 &  -27.6
 	\end{tabular}
@@
 This is somewhat surprising, since isothermal titration calorimetry (ITC) of biotin-streptavidin shows strong enthalpic binding contributions.\cite{mpye2020-biotin-itc}\cite{hyre2006-biotin-itc}
+
+\subsection{Further steps}
 %\section{Content and links}

### 2022-05-19 manuscript/manuscript.tex

@@
 \subsection{PME implementation}
+
+In the latest implementation of GIST, the particle mesh ewald (PME) functionaliy was added that decomposed the contributions of the PME-based electrostatic energy and the long-range Lennard-Jones energy to the total solvation energy.
+This allows  for a faster and more accurate calculation of the solvation energy.
+
+\begin{equation}
+	E_\textit{total} = E_\textit{elec} + E_\textit{lj}
+\end{equation}
+
+In the original version of GIST, the energies are calculated based on the shortest distance or the minimum image convention.
+In doing so, long range interactions are not accounted for.
+In the PME functionality, the contribution of the long-range interactions are accounted for.
+
+In terms of the electrostatic energy, in the original version, the electrostatic energy longer than half of the cell's diagonal are not accounted for.
+In the PME functionality, the long-range interactions are accounted for by decomposing the electrostatic interaction into three parts: the direct term, the reciprocal term, and the correction term.
+
+\begin{equation}
+	E_\textit{elec} = E_\textit{dir} + E_\textit{rec} + E_\textit{corr}
+\end{equation}
+
+On the other hand, in PME-GIST, the long-range interactions for the Lennard-Jones contribution are treated by accounting for the short-range and the long-range correction contributions separately.
+The short-range contribution is the LJ energy within the set cutoff and the long-range correction term accounts for the contributions above this cutoff.
+
+\begin{equation}
+	E_\textit{lj} = E_\texit{lj\_short} +  E_\textit{corr}
+\end{equation}
 

### 2022-05-20 manuscript/manuscript.tex
Update some paragraphs in intro and theory.
* add a few citations
* add a makeshift \neq command because the default one is messed up.

@@
 \DeclareSIUnit\Molar{M}
-\usepackage[italic]{mathastext}
+\usepackage[italic,LGRgreek]{mathastext}
 \graphicspath{{figures/}}
@@
 
+% The \neq command looks messy by default. Replace it by our own makeshift version.
+\renewcommand{\neq}{\setbox0\hbox{=}\rlap{\hbox to \wd0{\hss /\hss}}\box0}
+
 \lstset{
@@
 Solvation thermodynamics govern any process involving solutes in a solvent.
-Solutes are constantly interacting with solvents in their environment, necessitating a good description of these interactions for accurate computational predictions. 
+%Solutes are constantly interacting with solvents in their environment, necessitating a good description of these interactions for accurate computational predictions. 
 Especially the interaction with water is of utmost importance, as many biological processes occur in a aquaeous environment. 
-The free energy of solvation thereby strongly influences real-world problems of interest such as hydrophobic effects, the binding of a ligand to a protein, protein-protein interactions, protein dynamics or even more generally, a protein's fold.
-The computational calculation of thermodynamical solvation properties is therefore of particular interest and has been tackled by a wide array of different methods.   
+The free energy of solvation thereby strongly influences real-world problems of interest such as hydrophobic effects, the binding of ligands to a biomolecule, as well as the dynamics and folding of proteins.
+
+The computational calculation of thermodynamical solvation properties is therefore of particular interest and has been tackled by a wide array of methods.
 Quantum mechanical methods provide the most rigorous description of the underlying potential energy surface (PES) but suffer from high computational cost, allowing only the treatment of small systems and few solvent molecules.
-More recently, machine learning approaches and QM/MM approaches are trying to bridge the gap to computationally less demanding methods while keeping the accuracy of ab initio QM methods.
+More recently, QM/MM approaches and potentials based on deep learning \cite{Smith2017-ani} are trying to bridge the gap to computationally less demanding methods while keeping the accuracy of DFT methods.
+
 Molecular mechanics based approaches describe the underlying PES through empirically fit force-fields, which reduce the computational demand of these calculations tremendously.
-This allows for the treatment of even large biological systems like fully solvated proteins. 
-Implicit solvent methods such as MM/PBSA  or MM/GBSA can quickly provide results for a large amount of supplied structures, but yield inaccurate results when compared to more rigorous explicit solvent methods.
-Somewhat inbetween the implicit solvent methods and the fully explicit solvent methods are Reference Interaction Site Models.
-One of these methods, 3D-RISM, foregoes the explicit simulation and uses 3D solvent distributions based on the Ornstein--Zernike integral equations to predict solvation thermodynamics instead.
-The 3D-RISM approach allows for a decomposition into entropic and enthalpic terms as well as a full localization of the underlying distributions to the solute structure.
-The most statisticaly rigorous of the explicit solvent MM methods are alchemical methods such as FEP or TI. 
-Assuming sufficient sampling, they can accurately predict the free energy of solvation in the boundaries of the assumptions underlying their forcefield. 
-However, the seperation of enthalpic and entropic contributions to the free energy of solvation is difficult with these methods and localizing these contributions not easily possible.
-In this tutorial, the Grid Inhomogenous Solvation Theory (GIST) is applied to calculate solution thermodynamic properties of interest. 
-GIST is based on Lazaridis' Inhomogenous Solvation Theory (IST), solving the arising integral equations on a grid. 
-This is highly advantegous compared to other IST-based methods such as e.g. WaterMap,\todo{Add more methdods!} as this is not restricted to high-occopancy solvation sites but provides a 
+This allows for the treatment of large biological systems such as proteins in aqueous solution.
+Implicit solvent methods such as MM/PBSA or MM/GBSA can quickly provide results for a large number of structures, but yield inaccurate results when compared to more rigorous explicit solvent methods. 
+On the other hand, the free energy of solvation can be calculated in a statistically rigorous way using explicit solvent simulations combined with alchemical methods such as free energy perturbation (FEP) or thermodynamic integration(TI).
+However, they are unable to provide a spatial interpretation of the results.
+Furthermore, splitting the resulting free energy into contributions for the enthalpy and entropy is challenging \cite{Peter2004-alchemical-entropy}.
+%Assuming sufficient sampling, they can accurately predict the free energy of solvation in the boundaries of the assumptions underlying their forcefield. 
+%However, the seperation of enthalpic and entropic contributions to the free energy of solvation is difficult with these methods and localizing these contributions not easily possible.
+
+Methods derived from statistical mechanics bridge this gap by providing three-dimensional resolution.
+Two major groups of methods that fall into this group are those based on inhomogeneous solvation theory (IST) \cite{Lazaridis1998} and those based on the Ornstein--Zernike (OZ) equation.
+The most important example of OZ-based methods is the reference interaction site model (RISM) as well as its extension to three dimensions (3D-RISM).
+Those methods compute the 3D solvent distributions by self-consistent solution of the OZ integral equation to predict solvation thermodynamics.
+%The 3D-RISM approach allows for a decomposition into entropic and enthalpic terms as well as a full localization of the underlying distributions to the solute structure.
+
+On the other hand, methods based on IST generally use a molecular solvent distribution obtained from an MD simulation to compute the free energy of hydration from its enthalpic and entropic contributions.
+Examples of those methods include grid inhomogeneous solvation theory (GIST) \cite{Nguyen2012,Ramsey2016}, WaterMap, SSTMap \todo{what else?}.
+The main limitation of those methods is that they rely on an infinite expansion of the entropy in terms of the density distribution, which is truncated after the first or, sometimes, the second term.
+Nevertheless, they have been applied on a wide variety of problems involving both small molecules and biological systems.
+
+In this tutorial, GIST is applied to calculate solution thermodynamics properties of interest.
+
+GIST is based on IST, solving the integral equations on a grid. 
+This approach is more general than other IST-based methods because it accounts for all solvent regions rather than focusing on high-occopancy solvation sites.
 By simulating the dynamics of a solvent around the solute, GIST calculates thermodynamic properties for a standard solvation process as described by Ben-Naim.
 The method allows for a full localization of entropic and enthalpic contributions on the grid around the solute.
-For small organic molecules in water, the method provided results compareable to TI in accuracy, assuming some small correction factor to the entropy is applied. 
-Recent improvements to the method have improved the calculation speed tremendously, allowing for the treatment of larger systems of biological interest such as proteases or antibodies. 
-Other improvements worked on decreasing differences in the treatment of long-range interaction energies between GIST and popular molecular dynamics engines. 
+It has been shown that GIST can produce comparable accuracy to TI by applying a linear correction factor (0.6 for water) to the first-order entropy as follows.
+\begin{equation}
+S^\textit{total} = S^\textit{1st-order} \times f
+\end{equation}
+
+Recent work has improved the calculation speed tremendously, allowing for the treatment of large biological systems such as proteases or antibodies.
+Furthermore, an implementation based on the particle mesh Ewald (PME) method was presented, which improves the consistency between GIST and popular molecular dynamics engines. 
 Yet further changes allow the inclusion of other rigid solvents than water or allow for the treatment of solvent mixtures and ions in solution.
+
 While previous tutorials gave insightful pointers as to how to setup and run GIST calculations, the post-processing of such calculations was often treated less rigorously.
@@
 The tutorial provides an example of a GIST workflow that can be easily adapted towards different systems. 
-In addition to a short introduction to the method and the theory behind it an overview of the various different flavours of GIST methods is provided.
+In addition to a short introduction to the method and the theory behind it, an overview of the various different flavors of GIST methods is provided.
 Towards this, a GIST study of the streptavidin-biotin complex is presented, which reflects on many common aspects of GIST studies.
 We present solvation analysis of a small molecule (biotin) as well as a biomolecule (streptavidin).
-For both systems, we show how to interpret the three-dimensional contributions of solvation free energy in a binding pocket.
+We show how to interpret the three-dimensional contributions of solvation free energy in a binding pocket and around the ligand. \todo{use active voice over passive.}
 The contribution of solvation free energy to binding is computed, which requires accurate post-processing of the GIST output to avoid unfavorable summation of bias in the GIST calculation.
@@
 
-After completing this tutorial, we expect the reader to be able to implement their own GIST study.
+After completing this tutorial, we expect the reader to be able to run their own GIST study.
 
@@
 \subsection{Background knowledge}
-This tutorial is aimed at advanced users with a solid background knowledge of molecular dynamics (MD) simulations. The user should be able to run MD simulations and work with their output data.
-Previous knowledge of the method is not necessary, but we recommend the perusal of the primary sources on the method outlined in the Theory section.
+This tutorial is aimed at users with a solid knowledge of molecular dynamics (MD) simulations.
+The user should be able to run MD simulations and work with their output data.
+Previous experience of GIST is not necessary, but the reader should be familiar with the primary sources on the method outlined in the Theory section.
+\todo{shouldn't a tutorial be the entry point to a new method?}
 
 The presented analyses are run in Python.
-A working understanding of Python is therefore necessary to follow the tutorial. 
+A basic understanding of Python is therefore necessary to follow the tutorial. 
 A Jupyter notebook showing directly the herein presented analyses is provided as well. 
@@
 Tutorials requiring specific software packages must provide instructions and files for the referenced version of the software.
-GIST's main implementation is part of the MD analysis software \software{cpptraj} as part of the AmberTools software suite.
+The main implementation of GIST is part of the MD analysis software \software{cpptraj} as part of AmberTools \cite{amber22}.
 A recent version of \software{cpptraj} should be used, such as the one in AmberTools22 or an installation from GitHub.
@@
 \begin{equation}
-	\Delta S_\textit{solv} \approx \Delta S_\textit{sw} \equiv -k_\textit{B} \frac{\rho^\textit{0}}{8\pi^\textit{2}} \int g_\textit{sw} \left(\textbf{r}, \omega \right) d\textbf{r}d\omega
+	\Delta S_\textit{solv}
+	\approx \Delta S_\textit{sw}
+	\newcommand{\gro}{g_{sw}(\textbf{r},\omega)}
+	\equiv -k_\textit{B} \frac{\rho^\textit{0}}{8\pi^\textit{2}} \int \gro \ln\gro d\textbf{r}d\omega
 \end{equation}
+\todo{We are inconsistent regarding R/kB}
 
@@
 \begin{equation}
-	g_\textit{vox} \left( \textbf{r}, \mathbf{\omega} \right) \approx g_\textit{vox} \left( \textbf{r})g_\textit{vox}(\mathbf{\omega} \right)
+	g_\textit{vox} \left( \textbf{r}, \omega \right) \approx g_\textit{vox} \left( \textbf{r})g_\textit{vox}\omega \right)
 \end{equation}
@@
 \begin{equation}
-	S_{k}^\textit{orient} \approx \frac{-R}{N_\textit{k}} \left( \sum _{i=1}^{N_k} ln g_{NN, \textit{i}} \left( \mathbf {\omega} \right) + \gamma \right)
+	S_{k}^\textit{orient} \approx \frac{-R}{N_k} \left( \sum _{i=1}^{N_k} \ln g_{NN, i} ( \omega ) + \gamma \right)
 \end{equation}
@@
 \begin{equation}
-	S_\textit{k} \approx \frac{-R}{N_\textit{k}} \left( \sum _{i=1}^{N_k} ln g_{NN, \textit{i}} \left( \textbf{r}, \mathbf {\omega} \right) + \gamma \right)
+	S_\textit{k} \approx \frac{-R}{N_\textit{k}} \left( \sum _{i=1}^{N_k} ln g_{NN, \textit{i}} \left( \textbf{r}, \omega \right) + \gamma \right)
 \end{equation}
@@
 
+The solute-solvent energy $E_{sw}$ can be expressed in terms of the solvent density and the potential $U_{sw}(\textbf{r},\omega)$ induced by the solute.
+In practice, it is evaluated as the expectation value $\langle\cdot\rangle$ of the pairwise force field energy $U_{ij}$ between all solvent molecules in voxel $k$ and all solute atoms.
+
 \begin{equation}
-	\Delta E_\textit{sw} \equiv \frac{1}{8\pi^2} \int g_\textit{sw}\left(\omega|\textbf{r}\right) U_\textit{sw}\left(\textbf{r}, \omega\right) d\omega
+\begin{aligned}
+\label{eq-esw}
+	\Delta E_{sw}(\textbf{r}) \equiv& \frac{1}{8\pi^2} \int g_\textit{sw}\left(\omega|\textbf{r}\right) U_\textit{sw}\left(\textbf{r}, \omega\right) d\omega \\
+	\Delta E_{sw,k}=& \left\langle \sum_i^{\textit{solvent},k} \; \sum_j^\textit{solute} U_{ij}\right\rangle
+\end{aligned}
 \end{equation}
 
+To localize $E_{sw}$ on the three-dimensional grid, every energy term is assigned to the voxel $k$ holding the solvent, and the expectation value in Equation~\ref{eq-esw} is computed per voxel.
 Similar to the entropy integrals, the solute-water solvation energy integral also decays with increasing distance from the solute.
 Hence, the solvation energy can be approximated by local spatial integrals.
-However, there is a complication arising from this approximation.
+
+The solvent-solvent energy $E_{ww}$ is computed in a similar manner.
+Again, it can be expressed in terms of density functions, but is practically computed as a sum over solvent-solvent interactions where one solvent is in voxel $k$.
+In contrast to $E_{sw}$, $E_{ww}$ does not tend to zero in bulk.
+Therefore, a reference energy corresponding to the average energy of a water molecule in bulk needs to be subtracted.
+
+%However, there is a complication arising from this approximation.
 Water molecules outside the region of interest will also be accounted for in the pairwise water term.
@@
 \begin{aligned}
-	\Delta E_\textit{ww} \equiv & \left(\frac{1}{8\pi^2} \right)^2 \rho^o \int g_\textit{sw}\left(\omega|\textbf{r}\right) \\
+	\Delta E_{ww}(\textbf{r}) \equiv & \left(\frac{1}{8\pi^2} \right)^2 \rho^o \int g_\textit{sw}\left(\omega|\textbf{r}\right) \\
 	& \times \left[g_\textit{sw}\left(\textbf{r}^\prime , \omega^\prime \right) - g^o_\textit{ww} \left(\textbf{r}, \omega, \textbf{r}^\prime, \omega^\prime \right)\right] \\
-	& \times U_\textit{ww}\left(\textbf{r}, \omega, \textbf{r}^\prime, \omega^\prime\right)d\omega d\textbf{r}^\prime d\omega^\prime
+	& \times U_\textit{ww}\left(\textbf{r}, \omega, \textbf{r}^\prime, \omega^\prime\right)d\omega d\textbf{r}^\prime d\omega^\prime \\
+	\Delta E_{ww,k} =& \frac{1}{N_k} \left\langle \sum_i^\textit{solvent,k} \; \sum_{j \neq i}^\textit{solvent} U_{ij}\right\rangle - \left\langle E_{ww}^\textit{bulk}\right\rangle
 \end{aligned}
 \end{equation}
-There are two ways in which this double counting was accounted for in the original paper.
+
+%AFAIK, this was never done, since the pair interaction matrix Eij would have to be written out.
+%There are two ways in which this double counting was accounted for in the original paper.
 First, using the water-displacement method.
 This method is applicable when a determining the energetic cost of kicking out a water molecule.
+%\begin{equation}
+%	\Delta E^R_\textit{ww} = n^\textit{R} E^\textit{bulk}_\textit{ww} - E^\textit{R,corr}_\textit{ww}
+%\end{equation}
+
+%The second method uses normalized-water properties to account the difference in water properties of two different regions.
+The normalized water property is defined as the water property per water molecule.
+%\begin{equation}
+%	\Delta E^\textit{R,norm}_\textit{ww} = E^\textit{R,norm}_\textit{ww} - 2E^\textit{bulk}_\textit{ww}
+%\end{equation}
+
+When computing the total $E_{ww}$ of the system, double-counting of interactions must be avoided.
+The total solvent-solvent energy of the system is as follows.
+
 \begin{equation}
-	\Delta E^R_\textit{ww} = n^\textit{R} E^\textit{bulk}_\textit{ww} - E^\textit{R,corr}_\textit{ww}
+\Delta E_{ww} = \int \frac{\Delta E_{ww}(\textbf{r})}{2} d\textbf{r}
 \end{equation}
 
-The second method uses normalized-water properties to account the difference in water properties of two different regions.
-The normalized water property is defined as the water property per water molecule.
+When water is replaced from a small region $R$ of interest, such as a single water molecule, almost all interactions are with water molecules outside of this region.
+Therefore, there is no double counting, and the full solvent-solvent energy should be used.
+
 \begin{equation}
-	\Delta E^\textit{R,norm}_\textit{ww} = E^\textit{R,norm}_\textit{ww} - 2E^\textit{bulk}_\textit{ww}
+\Delta E_{ww,R} = \int_R E_{ww}(\textbf{r}) d\textbf{r}
 \end{equation}
+
+When the region of interest $R$ comprises more than one molecule, interactions within this region are double-counted while interactions to the outside are not.
+To take this into account, the solvent-solvent energy between each pair of voxels $E_{ww,kl}$ is needed.
+While this is supported by the standard GIST implementation, it is rarely done due to the large size of the $E_{ww,kl}$ matrix.
+
+We note that replacement of a small or medium solvent region would lead to reorganization of the surrounding solvent.
+However, this depends on the local environment and on \emph{what} the molecule would be replaced by.
+Therefore, the effect of reorganization can not be directly included in the GIST method, but must be judged for each case individually.
+This is not a problem when computing the total free energy of solvation $\dgsolv$, since no additional boundary is created.
+\todo{Do we need this paragraph?}
 
@@
 \begin{equation}
-	E_\textit{lj} = E_\texit{lj\_short} +  E_\textit{corr}
+	E_\textit{lj} = E_\textit{lj\_short} +  E_\textit{corr}
 \end{equation}

### 2022-05-30 manuscript/manuscript.tex
Update Text on PME implementation

@@
 
-In the latest implementation of GIST, the particle mesh ewald (PME) functionaliy was added that decomposed the contributions of the PME-based electrostatic energy and the long-range Lennard-Jones energy to the total solvation energy.
-This allows  for a faster and more accurate calculation of the solvation energy.
+In the original version of GIST, the energies are calculated based on the the minimum image convention.
+In PME-GIST, the electrostatics are calculated using the particle mesh Ewald (PME) method, which yields energies that are highly consistent with the treatment in the Amber MD engine \cite{Chen2021}.
+The electrostatic and Lennard-Jones parts $E_\text{elec}$ and $E_\text{lj}$ are computed separately.
 
 \begin{equation}
-	E_\textit{total} = E_\textit{elec} + E_\textit{lj}
+	E_\text{total} = E_\text{elec} + E_\text{lj}
 \end{equation}
 
-In the original version of GIST, the energies are calculated based on the shortest distance or the minimum image convention.
-In doing so, long range interactions are not accounted for.
-In the PME functionality, the contribution of the long-range interactions are accounted for.
-
-In terms of the electrostatic energy, in the original version, the electrostatic energy longer than half of the cell's diagonal are not accounted for.
-In the PME functionality, the long-range interactions are accounted for by decomposing the electrostatic interaction into three parts: the direct term, the reciprocal term, and the correction term.
+In PME-GIST, the system is treated as periodic, and the electrostatic energy is split into a short-range term $E_\text{dir}$, which is calculated in direct space, and a long-range term $E_\text{rec}$, which is calculated in reciprocal space.
+Additionally, there is a correction term $E_\text{corr}$, which corrects for the self-interaction of each solvent molecule in the reciprocal term.
 
@@
 
-On the other hand, in PME-GIST, the long-range interactions for the Lennard-Jones contribution are treated by accounting for the short-range and the long-range correction contributions separately.
+The Lennard-Jones contribution is computed in the direct space, using a distance cutoff.
 The short-range contribution is the LJ energy within the set cutoff and the long-range correction term accounts for the contributions above this cutoff.
@@
 \software{gisttools} contains functionality to perform this analysis easily.
-There is also a method \inlinecode{detect\_reference\_value} that automatically tries to find the converged value fo the rdf.
+There is also a method \inlinecode{detect\_reference\_value} that automatically tries to find the converged value from the rdf.
 Although it has a simple convergence check built-in, it is always recommended to check the convergence by hand as well.
@@
 However, it is more informative to plot the cumulative free energy contribution against the distance to the solute, to check the convergence.
-It the curve flattens out, this is your final $\dgsolv$\@.
+It the curve flattens out, the converged value is your final $\dgsolv$\@.
 If it does not, you might need to tweak the Eww reference, or introduce a reference value for the entropy (especially with older versions of \software{cpptraj}).
@@
 
+It has been shown in literature \cite{Chen2021}\todo{cite Kraml-Waibl paper} that the total solvation entropy in water is best approximated by 0.6 times the first order entropy.
+Also compute the scaled entropy and check the effect on the binding affinity.
 The expected results are summarized in Table~\ref{tab_dg_monomers_dimer}.
@@
 In this case, we find that the energy of solvation and the internal energy of the complex more or less cancel out, indicating that binding is largely entropy-driven in this case.
-This is somewhat surprising, since isothermal titration calorimetry (ITC) of biotin-streptavidin shows strong enthalpic binding contributions.\cite{mpye2020-biotin-itc}\cite{hyre2006-biotin-itc}
+At first, this is surprising, since isothermal titration calorimetry (ITC) of biotin-streptavidin shows strong enthalpic binding contributions \cite{mpye2020-biotin-itc,hyre2006-biotin-itc}.
+However, we do not take the conformational transition of the streptavidin binding pocket into account.
+Prior computational studies suggest a $\Delta G$ of \SI{-26.6}{\kilo\calorie\per\mole} for the binding of biotin into the closed conformation of streptavidin \cite{Bansal2018-biotin}.
+This indicates that our result is in good agreement with literature.
+However, the conformational changes of the binding pocket would need to be described to improve the agreement with experiments.
 

### 2022-06-02 manuscript/manuscript.tex
Change nothing, add snarky comments by yours truly all over the place

@@
 The electrostatic and Lennard-Jones parts $E_\text{elec}$ and $E_\text{lj}$ are computed separately.
-
+\todo{E$_{LJ}$ seems more intuitive to me, need to check what the GIST literature is using -vah}
 \begin{equation}
@@
 \end{equation}
-
+\todo{How is E$_{dir}$ treated? Is there a cutoff as well (PME GIST paper seems to indicate so)? If so, we should probably note that. }
 The Lennard-Jones contribution is computed in the direct space, using a distance cutoff.
@@
 \end{lstlisting}
-
+\todo{While not massively important, I don't like that we call the gist object here "gist", as that's the same name as gisttools subpackage gist.
+"gt.gist.bla" and "gist.bla" therefore look quite similiar, but do very different things.
+-vah}
 The expected output of this code is shown in Figure~\ref{fig_ewwref}.
@@
 
-After choosing an appropriate reference value, we need to subtract this value from Eww.
+After choosing an appropriate reference value, we need to subtract this value from E$_{ww}$.
 Make sure to subtract the reference from the normalized (\_norm) data.
 In \software{gisttools}, this is done simply by setting \inlinecode{gist.eww\_ref}.
-In \software{gistpp}, this can be done as follows: \todo{This might become invalid with future GIST versions, depending on whether we adopt the COM as molecular center.} \todo{Better use the population.}
+In \software{gistpp}, this can be done as follows: 
+\todo{Is there some official documentation for gistpp? We should probably link that somewhere, I feel that the code presented here is not really helpful/legible without further documentation. -vah}
+\todo{This might become invalid with future GIST versions, depending on whether we adopt the COM as molecular center.} 
+\todo{Better use the population.}
 
@@
 If it does not, you might need to tweak the Eww reference, or introduce a reference value for the entropy (especially with older versions of \software{cpptraj}).
+\todo{Eww or E$_{ww}$? Doesn't seem consistent so far. -vah}
 
@@
 The slow energy convergence indicates that a non-optimal reference value was used, but is also complicated by the fact that a charged solute is used without neutralizing counter ions.
+\todo{The reader is left hanging here - what should be done about the convergence issue? Is that bad or can it be ignored? -vah}
 
@@
 \end{lstlisting}
-
+\todo{Maybe add more complete print calls? E.g. print('Biotin: Frames=\{biotin.n\_frames\}, Rho=\{biotin.rho0\} -vah)}
 Assign reference energies and check them for plausibility.
@@
 Subtract a reference entropy from the dTSsix columns.
+\todo{Is there actually a reason why this is not in gisttools itself implemented? -vah}
 \begin{lstlisting}[style=python]
@@
 \end{lstlisting}
-
+\todo{\$\textbackslash AA\$ should produce an angstrom sign in matplotlib (what a pain to have this printed out in \LaTeX\ tho) -vah}
 Compute the energy (Eall) and entropy (dTSsix) contributions separately.
@@
 The expected results are summarized in Table~\ref{tab_dg_monomers_dimer}.
-
+\todo{Table 1 is placed quite akwardly at the moment for me, maybe enforce its position? -vah}
 \begin{table}
@@
 However, the conformational changes of the binding pocket would need to be described to improve the agreement with experiments.
-
+\todo{Should we include a short introduction to other solvents with GIST? What about solvent/ion mixtures? Can we somehow distinguish python code from input files optically? -vah}
 \subsection{Further steps}

### 2022-06-14 manuscript/manuscript.tex

@@
 \newcommand{\dgsolv}{\Delta G_\textup{solv}}
-\newcommand{\software}{\emph}
+\newcommand{\software}{\texttt}
 \newcommand{\todo}{\textcolor{red}}
@@
 \lstset{
-	basicstyle=\ttfamily\footnotesize
+	basicstyle=\ttfamily\footnotesize,
+	inputencoding=utf8,
+	extendedchars=true,
+	literate={Å}{{\AA}}1
 }
@@
 
+\todo{IMO, our article looks too dry. We should have something nice (e.g., an illustration of the GIST method) on the 1st or 2nd page -fwa.}
+
 Solvation thermodynamics govern any process involving solutes in a solvent.
@@
 Quantum mechanical methods provide the most rigorous description of the underlying potential energy surface (PES) but suffer from high computational cost, allowing only the treatment of small systems and few solvent molecules.
-More recently, QM/MM approaches and potentials based on deep learning \cite{Smith2017-ani} are trying to bridge the gap to computationally less demanding methods while keeping the accuracy of DFT methods.
+%More recently, QM/MM approaches and potentials based on deep learning \cite{Smith2017-ani} are trying to bridge the gap to computationally less demanding methods while keeping the accuracy of DFT methods.
 
@@
 This allows for the treatment of large biological systems such as proteins in aqueous solution.
-Implicit solvent methods such as MM/PBSA or MM/GBSA can quickly provide results for a large number of structures, but yield inaccurate results when compared to more rigorous explicit solvent methods. 
-On the other hand, the free energy of solvation can be calculated in a statistically rigorous way using explicit solvent simulations combined with alchemical methods such as free energy perturbation (FEP) or thermodynamic integration(TI).
+Implicit solvent methods such as MM/PBSA or MM/GBSA \cite{Sitkoff1994-pbsa,Kollman2000-mmpbsa} can quickly provide results for a large number of structures, but yield inaccurate results when compared to more rigorous explicit solvent methods \cite{Genheden2015-mmpbsa-review}.
+On the other hand, the free energy of solvation can be calculated in a statistically rigorous way using explicit solvent simulations combined with alchemical methods such as free energy perturbation (FEP) \cite{Zwanzig1954-reweighting} or thermodynamic integration(TI) \cite{Kirkwood1935-ti}.
 However, they are unable to provide a spatial interpretation of the results.
@@
 Methods derived from statistical mechanics bridge this gap by providing three-dimensional resolution.
-Two major groups of methods that fall into this group are those based on inhomogeneous solvation theory (IST) \cite{Lazaridis1998} and those based on the Ornstein--Zernike (OZ) equation.
-The most important example of OZ-based methods is the reference interaction site model (RISM) as well as its extension to three dimensions (3D-RISM).
-Those methods compute the 3D solvent distributions by self-consistent solution of the OZ integral equation to predict solvation thermodynamics.
+Two major groups of methods that fall into this group are those based on inhomogeneous solvation theory (IST) \cite{Lazaridis1998} and those based on the Ornstein--Zernike (OZ) equation \cite{Hansen2013-simple-liquids}.
+The most important example of OZ-based methods is the reference interaction site model (RISM) \cite{Chandler1972-rism} as well as its extension to three dimensions (3D-RISM) \cite{Kovalenko1998-3drism}.
+Those methods compute the solvent distributions by self-consistent solution of the OZ integral equation to predict solvation thermodynamics.
 %The 3D-RISM approach allows for a decomposition into entropic and enthalpic terms as well as a full localization of the underlying distributions to the solute structure.
@@
 On the other hand, methods based on IST generally use a molecular solvent distribution obtained from an MD simulation to compute the free energy of hydration from its enthalpic and entropic contributions.
-Examples of those methods include grid inhomogeneous solvation theory (GIST) \cite{Nguyen2012,Ramsey2016}, WaterMap, SSTMap \todo{what else?}.
-The main limitation of those methods is that they rely on an infinite expansion of the entropy in terms of the density distribution, which is truncated after the first or, sometimes, the second term.
+Examples of those methods include grid inhomogeneous solvation theory (GIST) \cite{Nguyen2012,Ramsey2016}, WaterMap \cite{Young2007-watermap,Abel2008-watermap}, SSTMap \cite{Haider2018-sstmap}, and STOW \cite{Li2012-stow}.
+The main limitation of those methods is that they rely on an infinite expansion of the entropy in terms of the density distribution, which is truncated after the first or, sometimes, the second \cite{Nguyen2016-gist-second-order,Waibl2022-gist-solvents} term.
 Nevertheless, they have been applied on a wide variety of problems involving both small molecules and biological systems.
@@
 This approach is more general than other IST-based methods because it accounts for all solvent regions rather than focusing on high-occopancy solvation sites.
-By simulating the dynamics of a solvent around the solute, GIST calculates thermodynamic properties for a standard solvation process as described by Ben-Naim.
-The method allows for a full localization of entropic and enthalpic contributions on the grid around the solute.
-It has been shown that GIST can produce comparable accuracy to TI by applying a linear correction factor (0.6 for water) to the first-order entropy as follows.
+By simulating the dynamics of a solvent around the solute, GIST calculates thermodynamic properties for a standard solvation process as described by Ben-Naim \cite{ben-naim-book}.
+It allows for a full localization of entropic and enthalpic contributions on the grid around the solute.
+It has been shown that GIST can produce comparable accuracy to TI by applying a linear correction factor (0.6 for water) to the first-order entropy as follows \cite{Chen2021,Waibl2022-gist-solvents}:
 \begin{equation}
@@
 Recent work has improved the calculation speed tremendously, allowing for the treatment of large biological systems such as proteases or antibodies.
-Furthermore, an implementation based on the particle mesh Ewald (PME) method was presented, which improves the consistency between GIST and popular molecular dynamics engines. 
+Furthermore, an implementation based on the particle mesh Ewald (PME) \cite{Darden1993-pme} method was presented, which improves the consistency between GIST and popular molecular dynamics engines. 
 Yet further changes allow the inclusion of other rigid solvents than water or allow for the treatment of solvent mixtures and ions in solution.
@@
 
-The tutorial provides an example of a GIST workflow that can be easily adapted towards different systems. 
-In addition to a short introduction to the method and the theory behind it, an overview of the various different flavors of GIST methods is provided.
-Towards this, a GIST study of the streptavidin-biotin complex is presented, which reflects on many common aspects of GIST studies.
-We present solvation analysis of a small molecule (biotin) as well as a biomolecule (streptavidin).
-We show how to interpret the three-dimensional contributions of solvation free energy in a binding pocket and around the ligand. \todo{use active voice over passive.}
-The contribution of solvation free energy to binding is computed, which requires accurate post-processing of the GIST output to avoid unfavorable summation of bias in the GIST calculation.
+This tutorial provides an example of a GIST workflow that can be easily adapted towards different systems. 
+In addition to a short introduction of the method and the theory behind it, we present an overview of the various different flavors of GIST methods.
+
+Furthermore, a GIST study of the streptavidin-biotin complex is presented, which reflects on many common aspects of GIST studies.
+Using this test system, we present solvation analysis of a small molecule (biotin) as well as a biomolecule (streptavidin).
+We show how to interpret the three-dimensional contributions of solvation free energy in a binding pocket and around the ligand.
+
+Additionally, we compute the binding contribution of the free energy of solvation, which requires accurate post-processing of the GIST outputs to avoid unfavorable summation of bias in the GIST calculation.
 We believe that this example study covers a wide range of applications of the GIST method.
+
 Furthermore, we discuss several technical aspects  and typical pitfalls of GIST analyses, such as the normalization of voxel values and how to deal with double counting of energy contributions. 
@@
 The user should be able to run MD simulations and work with their output data.
-Previous experience of GIST is not necessary, but the reader should be familiar with the primary sources on the method outlined in the Theory section.
-\todo{shouldn't a tutorial be the entry point to a new method?}
+Previous experience of GIST is not necessary.
+However, we only provide a short explanation of the theory behind GIST and point the reader towards the primary sources for more detailed information.
 
 The presented analyses are run in Python.
-A basic understanding of Python is therefore necessary to follow the tutorial. 
-A Jupyter notebook showing directly the herein presented analyses is provided as well. 
-To easily follow the analyses an understanding of working with Jupyter notebooks is therefore beneficial, though not strictly necessary.
+While no programming experience should be necessary to follow the tutorial, a basic understanding of Python is advantageous to adjust the code towards different requirements and use-cases. 
+We provide a Jupyter notebook which contains the presented analyses. 
+Therefore, an understanding of Jupyter notebooks is beneficial to reproduce the analyses.
 
@@
 The closed-source Amber simulation engine is used in this tutorial to run the necessary MD simulations.
-If no Amber installation is available, the MD can be performed other simulation engines compatible with \software{cpptraj}, we recommend GROMACS.
-If using GROMACS, we propose to prepare the structures and topologies using AmberTools and then converting to GROMACS using \software{acpype.py}.
-In that way, a full Amber topology is available for the GIST analysis.
+
+If no Amber installation is available, the MD can be performed other simulation engines compatible with \software{cpptraj}.
+If using GROMACS, we propose to prepare the structures and topologies with AmberTools and converting them using \software{acpype.py}.
+In that way, a full Amber topology is available for GIST.
+Also note that compressed trajectory formats such as \inlinecode{xtc} might bias the entropy calculation.
 
@@
 \end{itemize}
+
+We also show examples using the \software{gistpp} program, which can be obtained from \url{https://github.com/KurtzmanLab/Gist-Post-Processing} (code and documentation).
+
 If the reader prefers to skip the MD and GIST calculation, we also provide the GIST output files with this tutorial, such that the post-processing can be done without any expensive calculations.
-A Jupyter notebook showing the herein presented analyses is provided as well, necessitating a current installation of Jupyter to run.
+We also provide a Jupyter notebook \cite{Kluyver2016-jupyter,Granger2021-jupyter} with the presented analyses.
 
@@
 Here, we only provide a short overview of the theory behind GIST.
-For more detailed information, we recommend one of the more recent publications on developments in GIST. \cite{Kraml2020}\cite{Chen2021}
+For more detailed information, we recommend one of the more recent publications on developments in GIST. \cite{Kraml2020,Chen2021}
 
@@
 pair correlation function in the solute's frame of reference, \textbf{r} is the coordinates of the oxygen atom of the water molecule,  
-$\omega$ is the Euler angle, and the factor $\frac{1}{8\pi^2}$ is the normalization constant. In bulk and in regions where the orientational 
+$\omega$ represents the Euler angles, and $\frac{1}{8\pi^2}$ is the volume of the orientational space, which acts as a normalization constant.
+In bulk and in regions where the orientational 
 entropy is uniform, the quantity $g_\textit{sw}\left(\textbf{r},\omega \right)$ is unity which leads to the first-order solvation entropy of 
@@
 	& \times U_\textit{ww}\left(\textbf{r}, \omega, \textbf{r}^\prime, \omega^\prime\right)d\omega d\textbf{r}^\prime d\omega^\prime \\
-	\Delta E_{ww,k} =& \frac{1}{N_k} \left\langle \sum_i^\textit{solvent,k} \; \sum_{j \neq i}^\textit{solvent} U_{ij}\right\rangle - \left\langle E_{ww}^\textit{bulk}\right\rangle
+	\Delta E_{ww,k} =& \frac{1}{N_k} \left\langle \sum_i^{\text{solvent},k} \; \sum_{j \neq i}^\text{solvent} U_{ij}\right\rangle - \left\langle E_{ww}^\text{bulk}\right\rangle
 \end{aligned}
@@
 The electrostatic and Lennard-Jones parts $E_\text{elec}$ and $E_\text{lj}$ are computed separately.
-\todo{E$_{LJ}$ seems more intuitive to me, need to check what the GIST literature is using -vah}
+\todo{E$_{LJ}$ seems more intuitive to me, need to check what the GIST literature is using -vah. They use $E_{lj}$, but LJ in the text -fwa.}
 \begin{equation}
@@
 
-In PME-GIST, the system is treated as periodic, and the electrostatic energy is split into a short-range term $E_\text{dir}$, which is calculated in direct space, and a long-range term $E_\text{rec}$, which is calculated in reciprocal space.
+In PME-GIST, the system is treated as periodic, and the electrostatic energy is split into a short-range term $E_\text{dir}$, which is calculated in direct space using a distance cutoff, and a long-range term $E_\text{rec}$, which is calculated in reciprocal space.
 Additionally, there is a correction term $E_\text{corr}$, which corrects for the self-interaction of each solvent molecule in the reciprocal term.
@@
 \end{equation}
-\todo{How is E$_{dir}$ treated? Is there a cutoff as well (PME GIST paper seems to indicate so)? If so, we should probably note that. }
+%\todo{How is E$_{dir}$ treated? Is there a cutoff as well (PME GIST paper seems to indicate so)? If so, we should probably note that. } - ok? -fwa
 The Lennard-Jones contribution is computed in the direct space, using a distance cutoff.
@@
 A further advantage of the PME implementation is that it matches the energy calculation of the MD engine more closely.
-\todo{At the time of writing, it is important to use an up-to-date version of PME-GIST, or the Esw-Eww decomposition will not be available.
-Maybe remove this for the final version, when Amber22 is out.}
+
 \begin{lstlisting}
@@
 In practice, this is not problematic for the entropies, since they are computed relative to a distribution of randomly oriented molecules at the reference density.
-The solute-solvent energy Esw is also no problem, because it naturally tends to zero in bulk. 
-However, the solvent-solvent energy Eww needs to be referenced.
+The solute-solvent energy $E_{sw}$ is also no problem, because it naturally tends to zero in bulk. 
+However, the solvent-solvent energy $E_{ww}$ needs to be referenced.
 
@@
 
-After choosing an appropriate reference value, we need to subtract this value from E$_{ww}$.
+After choosing an appropriate reference value, we need to subtract this value from $E_{ww}$.
 Make sure to subtract the reference from the normalized (\_norm) data.
@@
 It the curve flattens out, the converged value is your final $\dgsolv$\@.
-If it does not, you might need to tweak the Eww reference, or introduce a reference value for the entropy (especially with older versions of \software{cpptraj}).
-\todo{Eww or E$_{ww}$? Doesn't seem consistent so far. -vah}
+If it does not, you might need to tweak the $E_{ww}$ reference, or introduce a reference value for the entropy (especially with older versions of \software{cpptraj}).
+% \todo{Eww or E$_{ww}$? Doesn't seem consistent so far. -vah} - let's go with all math mode $E_{ww}$ -fwa
 
@@
 The expected output of this code is shown in Figure~\ref{fig_radial_convergence}.
-In this case, both the Eww energy and the entropy are not perfectly flat, indicating that the reference value is not optimal.
-The slow energy convergence indicates that a non-optimal reference value was used, but is also complicated by the fact that a charged solute is used without neutralizing counter ions.
-\todo{The reader is left hanging here - what should be done about the convergence issue? Is that bad or can it be ignored? -vah}
+Both the solvent-solvent energy and the entropy are not perfectly flat, indicating that the reference value is not optimal.
+The slow energy convergence indicates that a non-optimal reference value was used, but is also complicated by the fact that a charged solute is used.
+In this case, both the entropy and $E_{sw}$ can be considered converged around \SI{10}{\angstrom}.
+While this is not true for $E_{ww}$, we will see below that the energy difference upon binding converges more readily than its individual contributions.
+%\todo{The reader is left hanging here - what should be done about the convergence issue? Is that bad or can it be ignored? -vah} - ok? -fwa
 
@@
 Visualize the free energy, the total energy, and the entropy, at several iso-levels.
-An minimal PyMOL input might look like this:
+A minimal PyMOL input might look like this:
 
@@
 This is because the cutoff contains regions that are close to the atom, since the integration region does not comprise the whole protein.
+
 \begin{lstlisting}[style=python]
@@
 plt.axvline(cutoff)
-plt.xlabel('distance to biotin [A]')
+plt.xlabel('distance to biotin [Å]')
 plt.ylabel('dG contribution [kcal/mol]')
 \end{lstlisting}
-\todo{\$\textbackslash AA\$ should produce an angstrom sign in matplotlib (what a pain to have this printed out in \LaTeX\ tho) -vah}
-Compute the energy (Eall) and entropy (dTSsix) contributions separately.
+% \todo{\$\textbackslash AA\$ should produce an angstrom sign in matplotlib (what a pain to have this printed out in \LaTeX\ tho) -vah} -- Å produces an angstrom sign ;-)
+Compute the energy (\inlinecode{Eall}) and entropy (\inlinecode{dTSsix}) contributions separately.
 You will find that the energy is strongly unfavorable to binding, because we do not yet include the interaction energy between biotin and streptavidin.
@@
 
-It has been shown in literature \cite{Chen2021}\todo{cite Kraml-Waibl paper} that the total solvation entropy in water is best approximated by 0.6 times the first order entropy.
+It has been shown in literature \cite{Chen2021,Waibl2022-gist-solvents} that the total solvation entropy in water is best approximated by 0.6 times the first order entropy.
 Also compute the scaled entropy and check the effect on the binding affinity.
 The expected results are summarized in Table~\ref{tab_dg_monomers_dimer}.
-\todo{Table 1 is placed quite akwardly at the moment for me, maybe enforce its position? -vah}
-\begin{table}
+%\todo{Table 1 is placed quite akwardly at the moment for me, maybe enforce its position? -vah. Ok? -fwa}
+\begin{table}[h]
 	\caption{Free energy contributions for the monomers and the dimer.}\label{tab_dg_monomers_dimer}
@@
 		biotin       & -24.0   &  -98.1 &  -35.1 &  -21.1 & -105.4 \\
-		Diff         & -98.7   &   97.7 &   44.4 &   26.6 &  -27.6
+		Diff         & -98.7   &   97.7 &   44.4 &   26.6 &  -27.6 \\
+		\bottomrule
 	\end{tabular}
@@
 \subsection{Further steps}
+In this tutorial, we obtained a value for the binding of biotin into the closed conformation of the streptavidin binging pocket.
+To investigate the effect of different conformations on the binding affinity, you can perform molecular dynamics simulations of the complex and/or monomers and perform GIST on multiple cluster representatives.
+To include the effect of the lid closing transition in the binding process, free energy calculation methods such as umbrella sampling can be used.
 %\section{Content and links}
@@
 %You will likely want to reference your GitHub repository as a central point to access all of this information, and then the GitHub repository may link out to other content as needed.
+
+\section{Recent developments}
+\subsection{GIST with non-water solvents}
+\subsection{GIST with salt-water mixtures}
 
@@
 % Here is a single-column checklist that consists of multiple sub-checklists
-\begin{Checklists}
+\begin{Checklists}[h]
 
 \begin{checklist}{Simulation settings}
+\textbf{Each number denotes the minimum setting, with the optimum in brackets.}
+\begin{itemize}
+	\item Simulation time: \SIrange{10}{20}{\nano\second} (\SI{100}{\nano\second})
+	\item Number of analyzed frames: \num{10000} (\num{100000})
+	\item Restraints: typically \SI{100}{\kcalPerMolASqr} on solute heavy atoms.
+\end{itemize}
 \end{checklist}
 
-\begin{checklist}{Obtaining absolute $\Delta G_\textup{solv}$ }
+\begin{checklist}{Obtaining absolute $\Delta G_\textup{solv}$}
+\begin{itemize}
+	\item Check the radial convergence (see Fig.~\ref{fig_radial_convergence})
+	\item Choose a sufficient distance cutoff
+	\item Choose an optimal $E_{ww}$ reference value
+	\item Tweak simulation length and number of frames to obtain smooth free energy contributions and unbiased (i.e., zero) bulk entropy.
+	\item If necessary, subtract an entropy reference
+\end{itemize}
 \end{checklist}
 
-\begin{checklist}{A list}
-\textbf{Single-column checklists are also straightforward by removing the asterisk}
-\begin{itemize}
-\item First thing let's do an item which breaks across lines to see how that looks
-\item Also remember
-\item And finally
-\end{itemize}
-\end{checklist}
+%\begin{checklist}{A list}
+%\textbf{Single-column checklists are also straightforward by removing the asterisk}
+%\begin{itemize}
+%\item First thing let's do an item which breaks across lines to see how that looks
+%\item Also remember
+%\item And finally
+%\end{itemize}
+%\end{checklist}
 %

### 2022-06-28 manuscript/manuscript.tex
Checklist for energy methods, update intro + theory

@@
 %Solutes are constantly interacting with solvents in their environment, necessitating a good description of these interactions for accurate computational predictions. 
-Especially the interaction with water is of utmost importance, as many biological processes occur in a aquaeous environment. 
+Especially the interaction with water is of utmost importance, as many biological processes occur in a aqueous environment. 
 The free energy of solvation thereby strongly influences real-world problems of interest such as hydrophobic effects, the binding of ligands to a biomolecule, as well as the dynamics and folding of proteins.
@@
 This approach is more general than other IST-based methods because it accounts for all solvent regions rather than focusing on high-occopancy solvation sites.
-By simulating the dynamics of a solvent around the solute, GIST calculates thermodynamic properties for a standard solvation process as described by Ben-Naim \cite{ben-naim-book}.
-It allows for a full localization of entropic and enthalpic contributions on the grid around the solute.
-It has been shown that GIST can produce comparable accuracy to TI by applying a linear correction factor (0.6 for water) to the first-order entropy as follows \cite{Chen2021,Waibl2022-gist-solvents}:
-\begin{equation}
-S^\textit{total} = S^\textit{1st-order} \times f
-\end{equation}
+%By simulating the dynamics of a solvent around the solute, GIST calculates thermodynamic properties for a standard solvation process as described by Ben-Naim \cite{ben-naim-book}.
+It estimates the solvent density from a molecular dynamics (MD) trajectory and allows for a full localization of entropic and enthalpic contributions on the grid around the solute.
+It has been shown that GIST can produce comparable accuracy to TI by approximating the higher-order entropies as a fraction of the first-order entropy (using a factor of \SI{-0.4} for water) \cite{Chen2021,Waibl2022-gist-solvents}.
+%applying a linear correction factor (0.6 for water) to the first-order entropy as follows \cite{Chen2021,Waibl2022-gist-solvents}:
+%\begin{equation}
+%S^\textit{total} = S^\textit{1st-order} \times f
+%\end{equation}
 
@@
 A recent version of \software{cpptraj} should be used, such as the one in AmberTools22 or an installation from GitHub.
+
 The closed-source Amber simulation engine is used in this tutorial to run the necessary MD simulations.
-
 If no Amber installation is available, the MD can be performed other simulation engines compatible with \software{cpptraj}.
@@
 \section{Theory}
-GIST is an implementation of Inhomogeneous Solvation Theory (IST) \cite{Lazaridis1998} that discretizes  the free energy of solvation $\dgsolv$ on a three-dimensional grid. 
+GIST is an implementation of inhomogeneous fluid solvation theory (IST) \cite{Lazaridis1998} that discretizes the free energy of solvation $\dgsolv$ on a three-dimensional grid. 
 It was first devised by Nguyen et al.
-\cite{Nguyen2012} to overcome the limitation of IST which was commonly used for regions of high water occupancy and its implementation in \software{cpptraj} was thoroughly described in \cite{Ramsey2016}.
+\cite{Nguyen2012} and its implementation in \software{cpptraj} was thoroughly described by Ramsey et al.
+\cite{Ramsey2016}.
+% to overcome the limitation of IST which was commonly used for regions of high water occupancy and its implementation in \software{cpptraj} was thoroughly described in \cite{Ramsey2016}.
 Here, we only provide a short overview of the theory behind GIST.
@@
 \subsection{Solvation Entropy}
-The solvation entropy of a system is totality of the contribution of the  solute-water entropy and the water-water entropy.
+%The solvation entropy of a system is totality of the contribution of the  solute-water entropy and the water-water entropy. -- this is inconsistent with IST theory (fwa)
 
-\begin{equation}
-	\Delta S_\textit{solv} = \Delta S_\textit{sw} + \Delta S_\textit{ww}
-\end{equation}
-Here the solvation entropy is approximated from the contribution of solute-water interaction. 
-\textbf{STILLNEEDS REVISON + ADD CITATIONS} The quantity $k_\textit{b}$represents the Boltzman constant, 
-$\rho^\textit{o}$ is the bulk number density, $g_\textit{sw}\left(\textbf{r},\omega \right)$ is the solute-water 
-pair correlation function in the solute's frame of reference, \textbf{r} is the coordinates of the oxygen atom of the water molecule,  
-$\omega$ represents the Euler angles, and $\frac{1}{8\pi^2}$ is the volume of the orientational space, which acts as a normalization constant.
-In bulk and in regions where the orientational 
-entropy is uniform, the quantity $g_\textit{sw}\left(\textbf{r},\omega \right)$ is unity which leads to the first-order solvation entropy of 
-bulk to be zero. Also, this quantity reaches unity as the distance from the solute molecule increases which leads for the calculation of 
-the solvation entropy to be an approximation of the integral around the solute atom.
+%\begin{equation}
+%	\Delta S_\textit{solv} = \Delta S_\textit{sw} + \Delta S_\textit{ww}
+%\end{equation}
+%Here the solvation entropy is approximated from the contribution of solute-water interaction. 
+%\textbf{STILLNEEDS REVISON + ADD CITATIONS} The quantity $k_\textit{b}$represents the Boltzman constant, 
+IST expresses the free energy of solvation in terms of the solvent density in a coordinate system defined by the solute.
+The solute is kept fixed in space, following the standard solvation process established by Ben-Naim \cite{ben-naim-book}.
 
+For the entropy, this leads to an infinite expansion of correlations of increasing order.
+In a slightly simplified view, the first order is the solvent density $g_\textit{sw}\left(\textbf{r},\omega \right)$ at each position $\textbf{r}$ and orientation $\omega$.
+The first-order entropy omits all solvent-solvent (and higher) correlations, while the solute-solvent correlation is taken into account because the coordinates are relative to the solute.
+Thus, it is called the solute-water entropy $S_{sw}$.
+%The second order is the pair density, i.e., the probability of finding two solvent molecules in positions $\textbf{r}$ and $\textbf{r}^\prime$.
+
+The solvent density is expressed relative to the bulk density $\rho^\textit{o}$. %$g_\textit{sw}\left(\textbf{r},\omega \right)$ is the solute-water 
+%pair correlation function in the solute's frame of reference, \textbf{r} is the coordinates of the oxygen atom of the water molecule,  
+%$\omega$ represents the Euler angles, and 
+The entropy integral is also normalized by $8\pi^2$, which is the volume of the orientational space.
+In bulk or at a high distance to the solute, the quantity $g_\textit{sw}\left(\textbf{r},\omega \right)$ is unity which leads to the first-order entropy of 
+bulk to be zero. 
+Therefore, no reference entropy needs to be computed unless there is a numeric bias, e.g., when the MD frames are not statistically independent due to insufficient spacing.
+%Also, this quantity reaches unity as the distance from the solute molecule increases which leads for the calculation of 
+%the solvation entropy to be an approximation of the integral around the solute atom. 
 \begin{equation}
@@
 
-The first method assumes that the position ($r$) is independent of the orientation ($\omega$). This approximation allows for the 
-splitting of the contributions of the translational and the orientational entropy. The nearest neighbor (NN) approach is used to 
-evaluate each expression where $N_\textit{k}$ is the number of water molecules found in voxel k, $\gamma$ is the Euler's constant 
-that accounts for the bias in naive entropy estimator, and $g_\textit{NN}$ is the nearest neighbor estimate. In the nearest neighbor
-estimate, a voxel is considered to be a neighboring voxel when it shares any vertices of the voxel of interest. 
+The first method assumes that the distribution of the orientation $\omega$ is constant within a voxel $k$.
+This allows for the splitting of the contributions of the translational and the orientational entropy. 
+A first nearest neighbor (NN) approach is used to evaluate the density.
+%The translational density $g_{NN}(\textbf{r})$ is expressed in reference to a homogeneous bulk solvent at the reference density, while the orientational density $g_{NN}(\omega)$ is expressed to a random placement of the number of solvents found in this voxel.
+Both the translational ($g_{NN}(\textbf{r})$) and orientational ($g_{NN}(\omega)$) density are computed relative to a bulk solvent at the reference density $\rho^0$.
+
+The average entropy contributions per solvent molecule in voxel $k$ are calculated as
 \begin{equation}
-	g_\textit{vox} \left( \textbf{r}, \omega \right) \approx g_\textit{vox} \left( \textbf{r})g_\textit{vox}\omega \right)
+	S_{k}^\textit{trans} = -R \left( \gamma + \frac{1}{N_\textit{k}} \sum _{i=1}^{N_k} \ln g_{NN, \textit{i}}(\textbf{r}) \right)
+\end{equation}
+and
+\begin{equation}
+	S_{k}^\textit{orient} = -R \left( \gamma + \frac{1}{N_k} \sum _{i=1}^{N_k} \ln g_{NN, i}(\omega) \right)
+	,
+\end{equation}
+where $N_\textit{k}$ is the number of water molecules found in voxel k, $\gamma$ is Euler's constant that accounts for the bias in naive entropy estimator, and $g_\textit{NN}$ is the nearest neighbor estimate.
+
+The second method directly calculates the solvation entropy by evaluating the six-dimensional integral (3 for position and 3 for orientation) 
+using the nearest-neighbor approach. 
+% Unlike the first method, the nearest neighbor estimate for the sixth-order entropy considers all the water found within the same and neighboring voxels.
+-- misleading because this is also done for S_trans (fwa)
+
+\begin{equation}
+g_\textit{vox} \left( \textbf{r}, \omega \right) \approx g_\textit{vox}(\textbf{r}) g_\textit{vox}(\omega)
 \end{equation}
@@
 \begin{equation}
-	S_{k}^\textit{trans} \approx \frac{-R}{N_\textit{k}} \left( \sum _{i=1}^{N_k} ln g_{NN, \textit{i}} \left( \textbf{r} \right) + \gamma \right)
-\end{equation}
-
-\begin{equation}
-	S_{k}^\textit{orient} \approx \frac{-R}{N_k} \left( \sum _{i=1}^{N_k} \ln g_{NN, i} ( \omega ) + \gamma \right)
-\end{equation}
-
-The second method directly calculates the solvation entropy by evaluating the six-dimensional integral (3 for position and 3 for orientation) 
-using the nearest-neighbor approach. Unlike the first method, the nearest neighbor estimate for the sixth-order entropy considers all the water 
-found within the same and neighboring voxels.  
-\begin{equation}
-	S_\textit{k} \approx \frac{-R}{N_\textit{k}} \left( \sum _{i=1}^{N_k} ln g_{NN, \textit{i}} \left( \textbf{r}, \omega \right) + \gamma \right)
+	S_\textit{k} = -R \left( \gamma + \frac{1}{N_\textit{k}} \sum _{i=1}^{N_k} \ln g_{NN, \textit{i}}(\textbf{r},\omega) \right)
 \end{equation}
@@
 \begin{aligned}
-	\Delta E_{ww}(\textbf{r}) \equiv & \left(\frac{1}{8\pi^2} \right)^2 \rho^o \int g_\textit{sw}\left(\omega|\textbf{r}\right) \\
-	& \times \left[g_\textit{sw}\left(\textbf{r}^\prime , \omega^\prime \right) - g^o_\textit{ww} \left(\textbf{r}, \omega, \textbf{r}^\prime, \omega^\prime \right)\right] \\
-	& \times U_\textit{ww}\left(\textbf{r}, \omega, \textbf{r}^\prime, \omega^\prime\right)d\omega d\textbf{r}^\prime d\omega^\prime \\
+	\Delta E_{ww}(\textbf{r}) \equiv & \left(\frac{1}{8\pi^2} \right)^2 \rho^o \int g_\textit{sw}(\omega|\textbf{r}) \\
+	& \times \left[g_\textit{sw}(\textbf{r}^\prime , \omega^\prime) - g^o_\textit{ww} (\textbf{r}, \omega, \textbf{r}^\prime, \omega^\prime )\right] \\
+	& \times U_\textit{ww}(\textbf{r}, \omega, \textbf{r}^\prime, \omega^\prime)d\omega d\textbf{r}^\prime d\omega^\prime \\
 	\Delta E_{ww,k} =& \frac{1}{N_k} \left\langle \sum_i^{\text{solvent},k} \; \sum_{j \neq i}^\text{solvent} U_{ij}\right\rangle - \left\langle E_{ww}^\text{bulk}\right\rangle
@@
 
+\begin{checklist}{Choosing an energy method}
+\begin{itemize}
+	\item GPU, direct space: generally the fastest
+	\item CPU, PME: highly consistent with Amber MD, fast
+	\item CPU: slow, most general (can write out $E_{ij}$ matrices)
+\end{itemize}
+\end{checklist}
+
 \begin{checklist}{Obtaining absolute $\Delta G_\textup{solv}$}

### 2022-07-12 manuscript/manuscript.tex
Update manuscript.tex
added a general overview and tutorial to run gist

--- 2022-06-14 manuscript/manuscript.tex
@@
 If using GROMACS, we propose to prepare the structures and topologies with AmberTools and converting them using \software{acpype.py}.
+\todo {If the simulation was ran using GROMACS, mdtraj can be used to convert GROMACS topology and trajectory in to an Amber file format.}
 In that way, a full Amber topology is available for GIST.
@@
 \end{enumerate}
+\todo{include: learn how to setup ad run a gist calculation}
 
@@
 Use the \software{equilibration.py} script (or your own) to perform short NVT and NpT equilibration runs on the starting structure.
+\todo {NPT}
 Note that the barostat might interfere with restrained equilibration of the complex, since different molecules are moved relative to each other when the volume changes.
@@
 Now, you can run restrained MD simulations based on the equilibrated structures for subsequent GIST analyses.
+\todo {explain why the protein needs to be restrained for an MD simulation that will be analyzed using GIST}
 We apply a harmonic restraint of \SI{100}{\kilo\calorie\per\mole\per\angstrom\squared} to all heavy atoms to keep the system in place.
@@
 \subsection{Running the GIST analysis}
-
+\todo{GIST analysis can be divided in three steps.
+\cite{Ramsey2016}}
+
+\todo{Step 1: Identify and define the region of interest
+In order to run a GIST analysis, the region where the solvent properties will be calculated must be first defined.
+This region is represented by a rectangular prism centered at (x,y,z) with a dimension of $L_x x L_y x L_z$.
+This region is further subdivided into three-dimesional grids based on the spacing that's specified.
+A smaller spacing indicates a higher resolution grids, but in return will slow down the calculation.
+A default value of 0.5 A is used that provides reasonable accuracy and speed of the calculation.
+}
+
+\todo{In ccptraj, the rectangular prism is defined by the command:}
+
+\todo {gist gridcntr <xcoordinate>  <ycordinated> <zcoordinate> griddim <xval> <yval> <zval> gridspacn <val>
+}
+
+\todo{The center of the grid is defined by gridcntr, the dimensions of the grid is defined by griddim, and the spacing is defined by gridspacn.}
+
+\todo{The region can be defined using a ligand in a specific binding pocket of the protein, a specific residue of the protein, or the whole protein itself.
+A helper script FindCentroid.py can be used to determine the centroid of the region of interest as well as the dimensions of the rectangular prism that will fully enclose the region.
+A pdb file is needed as an input and the rectangular prism will be centered based on the centroid of the given structure.}
+
+\todo {python FindCentroid.py -i structure.pdb  }
+
+\todo{Step 2: Run GIST using cpptraj}
+\todo{GIST is incorporated in cpptraj which requires the topology and the trajectory of the MD simulation to be analayzed.
+The commands needed to run GIST in cpptraj can be written in an input file (e.g gist.in) for convenience.
+The basic comands are:}
+
+\todo{parm topology.prmtop}
+\todo{trajin trajectory.nc 1 10000}
+\todo{gist gridcntr 10.5 20.1 30.1 griddim 30 30 30 gridspacn 0.50 out gist.dat}
+\todo{go}
+
+\todo{The first line specifies the amber topology file that was used to run MD.
+The second line species the trajectory file and what frames to be analyzed.
+In the above example, GIST will be run from frame 1 to 10000.
+The third line tells cpptraj to run a GIST calculation using the region defined and print the output of the calculation in a file called gist.dat.
+The last instructs cpptraj to run the calculation.
+The syntax for running cpptraj given this input file is:}
+
+\todo{cpptraj -i gist.in}
+
+\todo{Aside from the commands defining the region of interest, there are other options available to specify the GIST analysis.
+Here is the full list that can be added to the third line of the input file:}
+\begin{itemize}
+	\item{[doorder] Calculate the water order parameter [reference] for each voxel.}
+	\item{[doeij] Calculate the triangular matrix representing the water-water interactions between pairs of voxels (see below).}
+	\item{[skipE] Skip all energy calculations (cannot be specified with ’doeij’).}
+	\item{[skipS Skip all the entropy calculations.]}
+	\item{[refdens rdval>] Reference density of bulk water, used in computing $g_O, g_H$, and the translational entropy. Default is 0.0334 molecules/Å3.}
+	\item{[temp <tval>] Temperature of the input trajectory.}
+	\item{[noimage] Disable distance imaging in energy calculation.}
+	\item{[gridcntr <xval> <yval> <zval>] Coordinates (Å) of the center of the grid (default 0.0, 0.0, 0.0).}
+	\item{[griddim <xval> <yval> <zval>] Grid dimensions along each coordinate axis (default 40, 40, 40).}
+	\item{[gridspacn <spaceval>] Grid spacing (linear dimension of each voxel) in Angstroms. Values greater than 0.75 Å are not recommended (default 0.5 Å).}
+	\item{[prefix <filename prefix>] Output file name prefix (default “gist”).}
+	\item{[ext <grid extension>] Output grid file name extension (default “.dx”).}
+	\item{[out <output>] Name of the main GIST output file. If not specified set to ’<prefix>-output.dat’.}
+	\item{[info <info>] Name of main GIST info file. If not specified info is written to standard output.}
+
+\end{itemize}
+
+\todo{Step 3: Postprocessing and Analyzing GIST results}
+\todo{Cppraj outputs the quantities calculated in a data explorer "DX" file format which can be easily visualized in Pymol or VMD.
+Aside from the thermodynamic quantities, GIST also calculates the number densitiies of hydrogen and oxygen in each voxel.
+Here is a full list of quantities calculated by GIST:}
+
+\begin{itemize}
+	\item{[gO] Number density of oxygen centers found in the voxel, in units of the bulk density.}
+	\item{[gH] Number density of hydrogen centers found in the voxel in units of the reference bulk density.}
+	\item{[Esw] Mean solute-water interaction energy density.}
+	\item{[Eww] Mean water-water interaction energy density.}
+	\item{[dTStrans] First order translational entropy density.}
+	\item{[dTSorient] First order orientational entropy density .}
+	\item{[neighbor] Mean number of waters neighboring the water molecules found in this voxel multiplied by the voxel number density.}
+	\item{[dipole] Magnitude of mean dipole moment (polarization).}
+	\item{[order] Average Tetrahedral Order Parameter.}
+	\item{[dipolex] x-component of the mean water dipole moment density}
+	\item{[dipoley] y-component of the mean water dipole moment density}
+	\item{[dipolez] z-component of the mean water dipole moment density}
+	\item{[Eij] Water-water interaction matrix.}
+\end{itemize}
+
+\todo{There are default dx files that will be written as an output, such as the dx files for energies, entropies, number density of oxygen, and number density of hydrogen.
+All quantities mentioned above is written in the output file (output.dat) and can be used to generate DX files for other quantities.}
+
+\subsection{Running the GIST analysis for Streptavidin-Biotin Complex}
 For easier post-processing, we will use grids that contain the whole molecule.

### 2022-08-19 manuscript/manuscript.tex

--- 2022-06-14 manuscript/manuscript.tex
@@
 The main implementation of GIST is part of the MD analysis software \software{cpptraj} as part of AmberTools \cite{amber22}.
-A recent version of \software{cpptraj} should be used, such as the one in AmberTools22 or an installation from GitHub.
+A recent version of \software{cpptraj} should be used, such as the one in AmberTools22 or an installation from GitHub (\url{https://github.com/Amber-MD/cpptraj}).
 The closed-source Amber simulation engine is used in this tutorial to run the necessary MD simulations.
@@
 In that way, a full Amber topology is available for GIST. 
+\todo{ (vah) is it necessary to mention GROMACS specifically? I would remove the previous two lines}
 Also note that compressed trajectory formats such as \inlinecode{xtc} might bias the entropy calculation.
@@
 \begin{itemize}
-	\item \software{gisttools} (and its prerequisites)
-	\item \software{mdtraj}
+	\item \software{gisttools} (\url{https://github.com/liedllab/gisttools})
+	\item \software{mdtraj (\url{https://www.mdtraj.org/})}
 \end{itemize}
@@
 We also provide a Jupyter notebook \cite{Kluyver2016-jupyter,Granger2021-jupyter} with the presented analyses. 
+We recommend JupyterLab to work with the notebook. Jupyter is available at \url{https://jupyter.org/}.
 
+For the visualization of three-dimensional contributions we require molecular visualization software that can handle \software{OpenDX} files.
+We recommend using a current version of \software{PyMol}\cite{pymol} or \software{VMD}\cite{vmd}.
 \section{Theory}
 GIST is an implementation of Inhomogeneous Solvation Theory (IST) \cite{Lazaridis1998} that discretizes  the free energy of solvation $\dgsolv$ on a three-dimensional grid. 
-It was first devised by Nguyen et al.
-\cite{Nguyen2012} to overcome the limitation of IST which was commonly used for regions of high water occupancy and its implementation in \software{cpptraj} was thoroughly described in \cite{Ramsey2016}.
+It was first devised by Nguyen et al. \cite{Nguyen2012} to overcome the limitation of IST which was commonly used for regions of high water occupancy. 
+Its implementation in \software{cpptraj} was thoroughly described in \cite{Ramsey2016}.
 Here, we only provide a short overview of the theory behind GIST.
@@
 	\newcommand{\gro}{g_{sw}(\textbf{r},\omega)}
-	\equiv -k_\textit{B} \frac{\rho^\textit{0}}{8\pi^\textit{2}} \int \gro \ln\gro d\textbf{r}d\omega
+	\equiv -R \frac{\rho^\textit{0}}{8\pi^\textit{2}} \int \gro \ln\gro d\textbf{r}d\omega
 \end{equation}
-\todo{We are inconsistent regarding R/kB}
+\todo{We are inconsistent regarding R/kB; vah: Fixed. is it - tho? Papers seem to disagre }
 
@@
 estimate, a voxel is considered to be a neighboring voxel when it shares any vertices of the voxel of interest. 
+\todo{ vah: Is gamma divided by Nk or not? The formula in Franz' recent paper indicates otherwise.}
 \begin{equation}
@@
 \begin{equation}
-	S_{k}^\textit{trans} \approx \frac{-R}{N_\textit{k}} \left( \sum _{i=1}^{N_k} ln g_{NN, \textit{i}} \left( \textbf{r} \right) + \gamma \right)
+	S_{k}^\textit{trans} \approx \frac{-R}{N_\textit{k}} \left( \sum _{i=1}^{N_k} \ln g_{NN, \textit{i}} \left( \textbf{r} \right) + \gamma \right)
 \end{equation}
@@
 \begin{equation}
-	S_\textit{k} \approx \frac{-R}{N_\textit{k}} \left( \sum _{i=1}^{N_k} ln g_{NN, \textit{i}} \left( \textbf{r}, \omega \right) + \gamma \right)
+	S_\textit{k} \approx \frac{-R}{N_\textit{k}} \left( \sum _{i=1}^{N_k} \ln g_{NN, \textit{i}} \left( \textbf{r}, \omega \right) + \gamma \right)
 \end{equation}
@@
 \begin{equation}
-	E_\textit{lj} = E_\textit{lj\_short} +  E_\textit{corr}
+	E_\textit{lj} = E_\textit{lj,\ short} +  E_\textit{corr}
 \end{equation}
@@
 (Adapt for biotin as necessary.)
-
+\todo{vah: Maybe include a section on how to prepare a protein structure before equilibration.
+Protonation, force field and a large enough iso solvation box seem important to me.
+I will probably also at least mention what simulation settings we are running later on.}
 %We provide a script called \software{cpptraj\_remove\_mol.sh} that removes a single molecule from a topology and structure using \software{cpptraj} without altering the rest of the system.
@@
 While this is not true for $E_{ww}$, we will see below that the energy difference upon binding converges more readily than its individual contributions.
-%\todo{The reader is left hanging here - what should be done about the convergence issue? Is that bad or can it be ignored? -vah} - ok? -fwa
+\todo{vah: Is that shown somewhere yet?}
 

### 2022-08-19 manuscript/manuscript.tex

--- 2022-06-28 manuscript/manuscript.tex
@@
 The main implementation of GIST is part of the MD analysis software \software{cpptraj} as part of AmberTools \cite{amber22}.
-A recent version of \software{cpptraj} should be used, such as the one in AmberTools22 or an installation from GitHub.
+
+A recent version of \software{cpptraj} should be used, such as the one in AmberTools22 or an installation from GitHub (\url{https://github.com/Amber-MD/cpptraj}).
+The closed-source Amber simulation engine is used in this tutorial to run the necessary MD simulations.
 
@@
 We also provide a Jupyter notebook \cite{Kluyver2016-jupyter,Granger2021-jupyter} with the presented analyses. 
-
+We recommend JupyterLab to work with the notebook. Jupyter is available at \url{https://jupyter.org/}.
+
+For the visualization of three-dimensional contributions we require molecular visualization software that can handle \software{OpenDX} files.
+We recommend using a current version of \software{PyMol}\cite{pymol} or \software{VMD}\cite{vmd}.
 \section{Theory}
+
 GIST is an implementation of inhomogeneous fluid solvation theory (IST) \cite{Lazaridis1998} that discretizes the free energy of solvation $\dgsolv$ on a three-dimensional grid. 
@@
 \subsection{Solvation Entropy}
+
 %The solvation entropy of a system is totality of the contribution of the  solute-water entropy and the water-water entropy. -- this is inconsistent with IST theory (fwa)
@@
 %\textbf{STILLNEEDS REVISON + ADD CITATIONS} The quantity $k_\textit{b}$represents the Boltzman constant, 
+%$\rho^\textit{o}$ is the bulk number density, $g_\textit{sw}\left(\textbf{r},\omega \right)$ is the solute-water 
+%pair correlation function in the solute's frame of reference, \textbf{r} is the coordinates of the oxygen atom of the water molecule,  
+%$\omega$ represents the Euler angles, and $\frac{1}{8\pi^2}$ is the volume of the orientational space, which acts as a normalization constant.
+%In bulk and in regions where the orientational 
+%entropy is uniform, the quantity $g_\textit{sw}\left(\textbf{r},\omega \right)$ is unity which leads to the first-order solvation entropy of 
+%bulk to be zero. Also, this quantity reaches unity as the distance from the solute molecule increases which leads for the calculation of 
+%the solvation entropy to be an approximation of the integral around the solute atom.
+
 IST expresses the free energy of solvation in terms of the solvent density in a coordinate system defined by the solute.
@@
 	\newcommand{\gro}{g_{sw}(\textbf{r},\omega)}
-	\equiv -k_\textit{B} \frac{\rho^\textit{0}}{8\pi^\textit{2}} \int \gro \ln\gro d\textbf{r}d\omega
-\end{equation}
-\todo{We are inconsistent regarding R/kB}
+	\equiv -R \frac{\rho^\textit{0}}{8\pi^\textit{2}} \int \gro \ln\gro d\textbf{r}d\omega
+\end{equation}
+\todo{We are inconsistent regarding R/kB; vah: Fixed. is it - tho? Papers seem to disagre }
 
@@
 The average entropy contributions per solvent molecule in voxel $k$ are calculated as
+
 \begin{equation}
@@
 \begin{equation}
+
 	S_{k}^\textit{orient} = -R \left( \gamma + \frac{1}{N_k} \sum _{i=1}^{N_k} \ln g_{NN, i}(\omega) \right)
 	,
+
 \end{equation}
@@
 \begin{equation}
+
 	S_\textit{k} = -R \left( \gamma + \frac{1}{N_\textit{k}} \sum _{i=1}^{N_k} \ln g_{NN, \textit{i}}(\textbf{r},\omega) \right)
+
 \end{equation}

### 2022-08-19 manuscript/manuscript.tex

@@
 If using GROMACS, we propose to prepare the structures and topologies with AmberTools and converting them using \software{acpype.py}.
+
+\todo {If the simulation was ran using GROMACS, mdtraj can be used to convert GROMACS topology and trajectory in to an Amber file format.}
+\todo {vah: I would remove this mention of GROMACs and replace it with a more generall comment. Or we could test if this actually works...}
 In that way, a full Amber topology is available for GIST.
-\todo{ (vah) is it necessary to mention GROMACS specifically? I would remove the previous two lines}
+
 Also note that compressed trajectory formats such as \inlinecode{xtc} might bias the entropy calculation.

### 2022-08-25 manuscript/manuscript.tex
Continue the rework of the Tutorial part. Still unfinished.

@@
 
+% TODO: delete for final build
+\usepackage{pdfcomment}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \begin{equation}
-
 	S_{k}^\textit{orient} = -R \left( \gamma + \frac{1}{N_k} \sum _{i=1}^{N_k} \ln g_{NN, i}(\omega) \right)
 	,
-
 \end{equation}
@@
 \begin{equation}
-
 	S_\textit{k} = -R \left( \gamma + \frac{1}{N_\textit{k}} \sum _{i=1}^{N_k} \ln g_{NN, \textit{i}}(\textbf{r},\omega) \right)
-
 \end{equation}
@@
 \end{lstlisting}
+
 \subsection{Running the GIST analysis}
-\todo{GIST analysis can be divided in three steps.
-\cite{Ramsey2016}}
-
-\todo{Step 1: Identify and define the region of interest
+GIST analysis can be divided in three steps: defining a region of interest, running GIST using \software{cpptraj}, and post-processing and analysis.
+\cite{Ramsey2016}
+
+\paragraph{Identify and define the region of interest}
 In order to run a GIST analysis, the region where the solvent properties will be calculated must be first defined.
-This region is represented by a rectangular prism centered at (x,y,z) with a dimension of $L_x x L_y x L_z$.
-This region is further subdivided into three-dimesional grids based on the spacing that's specified.
-A smaller spacing indicates a higher resolution grids, but in return will slow down the calculation.
-A default value of 0.5 A is used that provides reasonable accuracy and speed of the calculation.
-}
-
-\todo{In ccptraj, the rectangular prism is defined by the command:}
-
-\todo {gist gridcntr <xcoordinate>  <ycordinated> <zcoordinate> griddim <xval> <yval> <zval> gridspacn <val>
-}
-
-\todo{The center of the grid is defined by gridcntr, the dimensions of the grid is defined by griddim, and the spacing is defined by gridspacn.}
-
-\todo{The region can be defined using a ligand in a specific binding pocket of the protein, a specific residue of the protein, or the whole protein itself.
-A helper script FindCentroid.py can be used to determine the centroid of the region of interest as well as the dimensions of the rectangular prism that will fully enclose the region.
-A pdb file is needed as an input and the rectangular prism will be centered based on the centroid of the given structure.}
-
-\todo {python FindCentroid.py -i structure.pdb  }
-
-\todo{Step 2: Run GIST using cpptraj}
+This region is represented by a rectangular prism centered at (x, y, z) with $N_x \times N_y \times N_z$ voxels, where each voxel is cubic with a sidelength of \inlinecode{gridspacn}.
+While integrals over GIST quantities are independent of the voxel size, individual voxel values converge slower at high resolutions.\pdfcomment{The calculation actually can actually get faster at higher resolution, because the same grid is used in the nearest-neighbor search.}
+A default value of 0.5 A is used that provides reasonable discretization for many use-cases.
+
+In \software{ccptraj}, the rectangular prism is defined by parameters of the \inlinecode{gist} command.
+The center of the grid is defined by \inlinecode{gridcntr}, the dimensions of the grid is defined by \inlinecode{griddim}, and the spacing is defined by \inlinecode{gridspacn}.
+
+\begin{lstlisting}
+gist gridcntr <x> <y> <z> \
+     griddim <Nx> <Ny> <Nz> gridspacn <val>
+\end{lstlisting}
+
+The region can be defined using a ligand in a specific binding pocket of the protein, a specific residue of the protein, or the whole protein itself.
+A helper script FindCentroid.py \todo{(upload FindCentroid.py!)} can be used to determine the centroid of the region of interest as well as the dimensions of the rectangular prism that will fully enclose the region.
+A pdb file is needed as an input and the rectangular prism will be centered based on the centroid of the given structure.
+
+\begin{lstlisting}
+python FindCentroid.py -i structure.pdb
+\end{lstlisting}
+
+\paragraph{Run GIST using cpptraj}
 \todo{GIST is incorporated in cpptraj which requires the topology and the trajectory of the MD simulation to be analayzed.
@@
 
-\todo{cpptraj -i gist.in}
-
-\todo{Aside from the commands defining the region of interest, there are other options available to specify the GIST analysis.
+\begin{lstlisting}
+cpptraj -i gist.in
+\end{lstlisting}
+
+\pdfcomment{Franz: I removed the list because it is very long and not complete. We can add it to the lists at the end, but then it should be complete.}
+%\todo{Aside from the commands defining the region of interest, there are other options available to specify the GIST analysis.
 Here is the full list that can be added to the third line of the input file:}
-\begin{itemize}
-	\item{[doorder] Calculate the water order parameter [reference] for each voxel.}
-	\item{[doeij] Calculate the triangular matrix representing the water-water interactions between pairs of voxels (see below).}
-	\item{[skipE] Skip all energy calculations (cannot be specified with ’doeij’).}
-	\item{[skipS Skip all the entropy calculations.]}
-	\item{[refdens rdval>] Reference density of bulk water, used in computing $g_O, g_H$, and the translational entropy. Default is 0.0334 molecules/Å3.}
-	\item{[temp <tval>] Temperature of the input trajectory.}
-	\item{[noimage] Disable distance imaging in energy calculation.}
-	\item{[gridcntr <xval> <yval> <zval>] Coordinates (Å) of the center of the grid (default 0.0, 0.0, 0.0).}
-	\item{[griddim <xval> <yval> <zval>] Grid dimensions along each coordinate axis (default 40, 40, 40).}
-	\item{[gridspacn <spaceval>] Grid spacing (linear dimension of each voxel) in Angstroms. Values greater than 0.75 Å are not recommended (default 0.5 Å).}
-	\item{[prefix <filename prefix>] Output file name prefix (default “gist”).}
-	\item{[ext <grid extension>] Output grid file name extension (default “.dx”).}
-	\item{[out <output>] Name of the main GIST output file. If not specified set to ’<prefix>-output.dat’.}
-	\item{[info <info>] Name of main GIST info file. If not specified info is written to standard output.}
-
-\end{itemize}
-
-\todo{Step 3: Postprocessing and Analyzing GIST results}
-\todo{Cppraj outputs the quantities calculated in a data explorer "DX" file format which can be easily visualized in Pymol or VMD.
+%\begin{itemize}
+%	\item{[doorder] Calculate the water order parameter [reference] for each voxel.}
+%	\item{[doeij] Calculate the triangular matrix representing the water-water interactions between pairs of voxels (see below).}
+%	\item{[skipE] Skip all energy calculations (cannot be specified with ’doeij’).}
+%	\item{[skipS Skip all the entropy calculations.]}
+%	\item{[refdens rdval>] Reference density of bulk water, used in computing $g_O, g_H$, and the translational entropy. Default is 0.0334 molecules/Å3.}
+%	\item{[temp <tval>] Temperature of the input trajectory.}
+%	\item{[noimage] Disable distance imaging in energy calculation.}
+%	\item{[gridcntr <xval> <yval> <zval>] Coordinates (Å) of the center of the grid (default 0.0, 0.0, 0.0).}
+%	\item{[griddim <xval> <yval> <zval>] Grid dimensions along each coordinate axis (default 40, 40, 40).}
+%	\item{[gridspacn <spaceval>] Grid spacing (linear dimension of each voxel) in Angstroms. Values greater than 0.75 Å are not recommended (default 0.5 Å).}
+%	\item{[prefix <filename prefix>] Output file name prefix (default “gist”).}
+%	\item{[ext <grid extension>] Output grid file name extension (default “.dx”).}
+%	\item{[out <output>] Name of the main GIST output file. If not specified set to ’<prefix>-output.dat’.}
+%	\item{[info <info>] Name of main GIST info file. If not specified info is written to standard output.}
+%\end{itemize}
+
+\paragraph{Postprocessing and Analyzing GIST results}
+Cppraj outputs the quantities calculated in a data explorer "DX" file format which can be easily visualized in Pymol or VMD.
 Aside from the thermodynamic quantities, GIST also calculates the number densitiies of hydrogen and oxygen in each voxel.
-Here is a full list of quantities calculated by GIST:}
+Here is a full list of quantities calculated by GIST \todo{This is incomplete!}:
 

### 2022-08-25 manuscript/manuscript.tex
Some text refactoring, suggest new section "Running GIST"

@@
 
+\section{Running GIST}
+\todo{We could put the examples for how to run GIST, how to visualize a dx file, how to create more dx files, etc., here and move them out.
+The actual tutorial could be only the streptavidin/biotin thing, and could be shortened a lot if we move the example inputs here.}
+
 \section{Tutorial}
@@
 \paragraph{Run GIST using cpptraj}
-\todo{GIST is incorporated in cpptraj which requires the topology and the trajectory of the MD simulation to be analayzed.
+GIST is incorporated in cpptraj which requires the topology and the trajectory of the MD simulation to be analayzed.
 The commands needed to run GIST in cpptraj can be written in an input file (e.g gist.in) for convenience.
-The basic comands are:}
-
-\todo{parm topology.prmtop}
-\todo{trajin trajectory.nc 1 10000}
-\todo{gist gridcntr 10.5 20.1 30.1 griddim 30 30 30 gridspacn 0.50 out gist.dat}
-\todo{go}
-
-\todo{The first line specifies the amber topology file that was used to run MD.
+The basic comands, assuming a given grid center and a grid size of $30\times30\times30$ voxels with \SI{0.5}{\angstrom} spacing, are:
+
+\begin{lstlisting}
+parm topology.prmtop
+trajin trajectory.nc
+gist gridcntr 10.5 20.1 30.1 griddim 30 30 30 gridspacn 0.50 out gist.dat
+go
+\end{lstlisting}
+
+The first line specifies the amber topology file that was used to run MD.
 The second line species the trajectory file and what frames to be analyzed.
-In the above example, GIST will be run from frame 1 to 10000.
 The third line tells cpptraj to run a GIST calculation using the region defined and print the output of the calculation in a file called gist.dat.
 The last instructs cpptraj to run the calculation.
-The syntax for running cpptraj given this input file is:}
+The syntax for running cpptraj given this input file is:
 
@@
 \begin{itemize}
-	\item{[gO] Number density of oxygen centers found in the voxel, in units of the bulk density.}
-	\item{[gH] Number density of hydrogen centers found in the voxel in units of the reference bulk density.}
+	\item{[gX] For every element in the main solvent, the number density of atoms found in the voxel, in units of the bulk density.
+If the same element occurs multiple times, the bulk density is scaled accordingly.
+For water, gO and gH are produced.}
+	%\item{[gH] Number density of hydrogen centers found in the voxel in units of the reference bulk density.}
 	\item{[Esw] Mean solute-water interaction energy density.}
 	\item{[Eww] Mean water-water interaction energy density.}
+	\item{[PME] (only if PME was used) water PME energy density.}
 	\item{[dTStrans] First order translational entropy density.}
@@
 	\item{[dipole] Magnitude of mean dipole moment (polarization).}
-	\item{[order] Average Tetrahedral Order Parameter.}
+	\item{[order] (if \inlinecode{doorder} was specified) Average Tetrahedral Order Parameter.}
 	\item{[dipolex] x-component of the mean water dipole moment density}
@@
 	\item{[dipolez] z-component of the mean water dipole moment density}
-	\item{[Eij] Water-water interaction matrix.}
+	\item{[Eij] (if \inlinecode{doeij} was specified) Water-water interaction matrix.}
 \end{itemize}
 
-\todo{There are default dx files that will be written as an output, such as the dx files for energies, entropies, number density of oxygen, and number density of hydrogen.
-All quantities mentioned above is written in the output file (output.dat) and can be used to generate DX files for other quantities.}
+Several dx files are written by default, such as the energies, entropies, number density of oxygen, and number density of hydrogen.
+All quantities mentioned above is written in the output file (output.dat) and can be used to generate DX files for other quantities.
+
+A minimal PyMOL input to visualize a dx file is shown here.
+For reference, an input structure is loaded from a file named \inlinecode{solute.pdb} (assuming that this file exists).
+The oxygen density is visualized at an isolevel of 2 (i.e., twice the reference density).
+
+% The current font makes the O look like a tofu / placeholder
+\begin{lstlisting}
+load solute.pdb
+as surface
+load gist-gO.dx, gO
+isosurface gO_high, gO, 2.
+\end{lstlisting}
+
+\todo{Should be put a picture here?}
 
@@
 \end{lstlisting}
-Then, find the extent (minimum and maximum) of the x-, y-, and z-coordinates for each system.
+Then, find the extent (minimum and maximum) of the x-, \mbox{y-,} and z-coordinates for each system.
 This can be done e.g., using the \software{mdtraj} package in Python.
@@
 Visualize the free energy, the total energy, and the entropy, at several iso-levels.
-A minimal PyMOL input might look like this:
-
-% The current font makes the O look like a tofu / placeholder
-\begin{lstlisting}
-load solute-centered.pdb
-show sticks, resname BTN
-as surface, not resname BTN
-load gist-gO.dx, gO
-isosurface gO_high, gO, 2.
-\end{lstlisting}
 

### 2022-08-28 manuscript/manuscript.tex
Move the info on running GIST into its own section.

@@
 \section{Running GIST}
-\todo{We could put the examples for how to run GIST, how to visualize a dx file, how to create more dx files, etc., here and move them out.
-The actual tutorial could be only the streptavidin/biotin thing, and could be shortened a lot if we move the example inputs here.}
-
-\section{Tutorial}
-Here, we aim to guide the reader through a GIST analysis of the streptavidin-biotin complex. The goal is to
-\begin{enumerate}
-	\item compute the binding free energy of the complex from the individual energy contributions combined with the contribution of hydration computed using GIST.
-	\item visualize local contributions to $\dgsolv$ in the streptavidin binding pocket as well as around biotin.
-\end{enumerate}
-\todo{include: learn how to setup ad run a gist calculation}
-
-\subsection{Streptavidin/Biotin}
-\subsection{Tutorial data}
-We will use the 1STP crystal structure to start our calculations.
-For the sake of simplicity, you can download prepared and solvated Amber topologies and structures based on 1STP by running
-\begin{lstlisting}
-git clone git@github.com:liedllab/gist-tutorial.git
-\end{lstlisting}
-This will download the tutorial files (including both the manuscript and the examples in the \software{code} folder).
-\subsection{System Preparation and Equilibration}
-If you want to skip this section, run 
-\begin{lstlisting}
-make equilibration-targets
-\end{lstlisting}
-in the code folder. This assumes that you have Python and Amber set up properly and that \software{pmemd.cuda} is in the PATH.
-
-When calculating solvation free energy differences between a complex and the corresponding monomers (the dissociated state), there are two options for dealing with configurational diversity in the structures.
-\todo{Maybe Valentin knows a citation for that?}
-One can choose to sample the complex and dissociated states individually, or to sample only one of them and assume that there is sufficient overlap in the sampled conformations.
-%One can choose to sample only the complex, assuming that all relevant conformations of the dissociated state are also found in the complex.
-%The probability distribution of the dissociated state can be obtained by reweighting the probabilities in the complex using the solvation free energies.
-%Alternatively, one can sample the complex and dissociated states individually.
-%This approach does not require an overlap between the conformational probabilities of the complex and dissociated states.
-%However, it is prone to statistical errors due to uncertainties in the ensembles, and requires the free energy contributions from the internal degrees of freedom of the monomers (in both the complex and the monomer states) to be known.
-\todo{This paragraph does not fit here well.
-Maybe put it somewhere else.}
-
-Here, we choose not to include any sampling of the conformational states. 
-It is therefore important that the conformations of the monomers in the complex and dissociated states match exactly. 
-Use \software{cpptraj} to produce biotin and streptavidin structures based on the complex by stripping the respective other molecule like this:
-\begin{lstlisting}
-parm complex/solvated.parm7
-trajin complex/equil/EQUIL-DONE.rst
-strip ^2 parmout streptavidin/solvated.parm7
-trajout streptavidin/pre-equil.ncrst
-\end{lstlisting}
-(Adapt for biotin as necessary.)
-\todo{vah: Maybe include a section on how to prepare a protein structure before equilibration.
-Protonation, force field and a large enough iso solvation box seem important to me.
-I will probably also at least mention what simulation settings we are running later on.}
-%We provide a script called \software{cpptraj\_remove\_mol.sh} that removes a single molecule from a topology and structure using \software{cpptraj} without altering the rest of the system.
-\todo{Explain cpptraj rather than our script.
-The script is for the automated workflow only.}
-
-Use the \software{equilibration.py} script (or your own) to perform short NVT and NpT equilibration runs on the starting structure.
-\todo {NPT}
-Note that the barostat might interfere with restrained equilibration of the complex, since different molecules are moved relative to each other when the volume changes.
-But since our equilibration script also contains unrestrained pressure equilibration, we should not run into any trouble.
-
-Then split the equilibrated structure into biotin and streptavidin parts using \software{cpptraj\_remove\_mol.sh}.
-After that, you can use \software{equilibration.py -R} to run short restrained equilibrations (where only the water is allowed to move) on the individual systems.
-
-\subsection{Running MD for GIST}
-
-Now, you can run restrained MD simulations based on the equilibrated structures for subsequent GIST analyses.
-\todo {explain why the protein needs to be restrained for an MD simulation that will be analyzed using GIST}
-We apply a harmonic restraint of \SI{100}{\kilo\calorie\per\mole\per\angstrom\squared} to all heavy atoms to keep the system in place.
-We run \SI{100}{\nano\second} of simulation and store the conformations every \SI{100}{\pico\second}.
-A simulation length of at least \SIrange{10}{20}{\nano\second} is recommended, to obtain statistically independent snapshots of the water.
-
-An example Amber input file for GIST might look like this:
-\begin{lstlisting}
-restrained 100 ns NpT
-&cntrl
-  ntx=5, irest=1,
-  ioutfm=1,
-  ntb=2, iwrap=1,
-  ntr=1, restraint_wt=100.0,
-    restraintmask="!@H=&!:WAT",
-  ntp=1, pres0=1.0, taup=1.0,
-  ntc=2, ntf=2,
-  ntt=3, tempi=300.0, temp0=300.0, gamma_ln=2,
-  nstlim=50000000, dt=0.002,
-  ntwr=50000, ntwx=5000,
-  ntpr=5000,
-/
-\end{lstlisting}
-Note that these are very standard MD settings except for the line that defines the restraints.
-
-Run the MD using \software{pmemd.cuda}:
-\begin{lstlisting}
-pmemd.cuda \
-	-O \
-	-i 100ns-npt-restraint.in \
-	-o md-01.out \
-	-p solvated.parm7 \
-	-c EQUIL-DONE.rst \
-	-x md-01.nc \
-	-r md-01.ncrst \
-	-ref EQUIL-DONE.rst
-\end{lstlisting}
-
-\subsection{Running the GIST analysis}
+%\subsection{Running the GIST analysis}
 GIST analysis can be divided in three steps: defining a region of interest, running GIST using \software{cpptraj}, and post-processing and analysis.
@@
 
+\section{Tutorial}
+Here, we aim to guide the reader through a GIST analysis of the streptavidin-biotin complex. The goal is to
+\begin{enumerate}
+	\item compute the binding free energy of the complex from the individual energy contributions combined with the contribution of hydration computed using GIST.
+	\item visualize local contributions to $\dgsolv$ in the streptavidin binding pocket as well as around biotin.
+\end{enumerate}
+\todo{include: learn how to setup ad run a gist calculation}
+
+\subsection{Streptavidin/Biotin}
+\subsection{Tutorial data}
+We will use the 1STP crystal structure to start our calculations.
+For the sake of simplicity, you can download prepared and solvated Amber topologies and structures based on 1STP by running
+\begin{lstlisting}
+git clone git@github.com:liedllab/gist-tutorial.git
+\end{lstlisting}
+This will download the tutorial files (including both the manuscript and the examples in the \software{code} folder).
+\subsection{System Preparation and Equilibration}
+If you want to skip this section, run 
+\begin{lstlisting}
+make equilibration-targets
+\end{lstlisting}
+in the code folder. This assumes that you have Python and Amber set up properly and that \software{pmemd.cuda} is in the PATH.
+
+When calculating solvation free energy differences between a complex and the corresponding monomers (the dissociated state), there are two options for dealing with configurational diversity in the structures.
+\todo{Maybe Valentin knows a citation for that?}
+One can choose to sample the complex and dissociated states individually, or to sample only one of them and assume that there is sufficient overlap in the sampled conformations.
+%One can choose to sample only the complex, assuming that all relevant conformations of the dissociated state are also found in the complex.
+%The probability distribution of the dissociated state can be obtained by reweighting the probabilities in the complex using the solvation free energies.
+%Alternatively, one can sample the complex and dissociated states individually.
+%This approach does not require an overlap between the conformational probabilities of the complex and dissociated states.
+%However, it is prone to statistical errors due to uncertainties in the ensembles, and requires the free energy contributions from the internal degrees of freedom of the monomers (in both the complex and the monomer states) to be known.
+\todo{This paragraph does not fit here well.
+Maybe put it somewhere else.}
+
+Here, we choose not to include any sampling of the conformational states. 
+It is therefore important that the conformations of the monomers in the complex and dissociated states match exactly. 
+Use \software{cpptraj} to produce biotin and streptavidin structures based on the complex by stripping the respective other molecule like this:
+\begin{lstlisting}
+parm complex/solvated.parm7
+trajin complex/equil/EQUIL-DONE.rst
+strip ^2 parmout streptavidin/solvated.parm7
+trajout streptavidin/pre-equil.ncrst
+\end{lstlisting}
+(Adapt for biotin as necessary.)
+\todo{vah: Maybe include a section on how to prepare a protein structure before equilibration.
+Protonation, force field and a large enough iso solvation box seem important to me.
+I will probably also at least mention what simulation settings we are running later on.}
+%We provide a script called \software{cpptraj\_remove\_mol.sh} that removes a single molecule from a topology and structure using \software{cpptraj} without altering the rest of the system.
+\todo{Explain cpptraj rather than our script.
+The script is for the automated workflow only.}
+
+Use the \software{equilibration.py} script (or your own) to perform short NVT and NpT equilibration runs on the starting structure.
+\todo {NPT}
+Note that the barostat might interfere with restrained equilibration of the complex, since different molecules are moved relative to each other when the volume changes.
+But since our equilibration script also contains unrestrained pressure equilibration, we should not run into any trouble.
+
+Then split the equilibrated structure into biotin and streptavidin parts using \software{cpptraj\_remove\_mol.sh}.
+After that, you can use \software{equilibration.py -R} to run short restrained equilibrations (where only the water is allowed to move) on the individual systems.
+
+\subsection{Running MD for GIST}
+
+Now, you can run restrained MD simulations based on the equilibrated structures for subsequent GIST analyses.
+\todo {explain why the protein needs to be restrained for an MD simulation that will be analyzed using GIST}
+We apply a harmonic restraint of \SI{100}{\kilo\calorie\per\mole\per\angstrom\squared} to all heavy atoms to keep the system in place.
+We run \SI{100}{\nano\second} of simulation and store the conformations every \SI{100}{\pico\second}.
+A simulation length of at least \SIrange{10}{20}{\nano\second} is recommended, to obtain statistically independent snapshots of the water.
+
+An example Amber input file for GIST might look like this:
+\begin{lstlisting}
+restrained 100 ns NpT
+&cntrl
+  ntx=5, irest=1,
+  ioutfm=1,
+  ntb=2, iwrap=1,
+  ntr=1, restraint_wt=100.0,
+    restraintmask="!@H=&!:WAT",
+  ntp=1, pres0=1.0, taup=1.0,
+  ntc=2, ntf=2,
+  ntt=3, tempi=300.0, temp0=300.0, gamma_ln=2,
+  nstlim=50000000, dt=0.002,
+  ntwr=50000, ntwx=5000,
+  ntpr=5000,
+/
+\end{lstlisting}
+Note that these are very standard MD settings except for the line that defines the restraints.
+
+Run the MD using \software{pmemd.cuda}:
+\begin{lstlisting}
+pmemd.cuda \
+	-O \
+	-i 100ns-npt-restraint.in \
+	-o md-01.out \
+	-p solvated.parm7 \
+	-c EQUIL-DONE.rst \
+	-x md-01.nc \
+	-r md-01.ncrst \
+	-ref EQUIL-DONE.rst
+\end{lstlisting}
+
 \subsection{Running the GIST analysis for Streptavidin-Biotin Complex}

### 2022-08-28 manuscript/manuscript.tex
Paragraph about Streptavidin, intro sentence for GIST usage

@@
 %\subsection{Running the GIST analysis}
+To run a gist analyis, the following is required:
+\begin{itemize}
+	\item A running installation of \software{cpptraj}.
+	\item A topology of the molecule of interest in a solvent box.
+	\item An MD trajectory of that system.
+\end{itemize}
 GIST analysis can be divided in three steps: defining a region of interest, running GIST using \software{cpptraj}, and post-processing and analysis.
@@
 \section{Tutorial}
-Here, we aim to guide the reader through a GIST analysis of the streptavidin-biotin complex. The goal is to
+Here, we aim to guide the reader through a GIST analysis of the streptavidin-biotin complex.
+In the process, the reader will learn how to set up and run a GIST calculation.
+During the tutorial, we will
 \begin{enumerate}
@@
 \end{enumerate}
-\todo{include: learn how to setup ad run a gist calculation}
 
 \subsection{Streptavidin/Biotin}
+The streptavidin-biotin complex is one of the strongest known non-covalent interactions involving a small molecule.
+Therefore, the combination of streptavidin and biotin is used routinely in biotechnology, and the interaction between them has been investigated in detail \todo{CITE}\cite{}.
+
+In nature, streptavidin occurs as a tetramer, and binds biotin with a binding constant of \todo{???}.
+Here, we will investigate the binding of biotin to a streptavidin monomer for simplicity.
+This interaction has been reported to have a binding constant of \todo{???}.
+
 \subsection{Tutorial data}

### 2022-08-31 code/scripts/findcentroid.py
Create findcentroid.py

@@ -1,0 +1,213 @@
+
+
+from os.path import basename
+#from more_itertools import unique_everseen #make sure more_itertools is installed (conda install more-itertools)
+from math import sqrt
+from math import floor
+
+
+
+import argparse
+
+
+parser = argparse.ArgumentParser()
+requiredNamed = parser.add_argument_group('required arguments')
+#requiredNamed.add_argument("-d", dest='threshold', help="distance threshold for atoms near waters", type=float, default=4.0, required=True)
+requiredNamed.add_argument("-i", dest='structure', help="input file, expected to be in .pdb format", type=str, required=True)
+parser.add_argument("-e", dest='Even', help="boolean to set the dimensions to even by default", default=False, type=bool)
+args = parser.parse_args()
+
+filenames = []
+
+
+filenames.append(basename(args.structure[:-4]))
+
+print(filenames)
+#print args.threshold
+#structure = sys.argv[1]
+#distance_cutoff = sys.argv[2]
+# HOLDS IN INFORMATION OF ATOM
+ATOM = []
+# HOLDS IN ALL NON HOH ATOMS THAT ARE LABELED HETAM
+HETAM = []
+# HOLDS ALL THE HOH ATOMS
+HETAMHOH = []
+DIS = []
+Unique = []
+newHOH =[]  
+Vectors = []
+Dictionary = {}
+
+
+
+
+#POINT CLASS USED TO HOLD POINTS OF ALL DIFFERENT ATOMS
+class Point:
+    #STORES THE X Y AND Z COORDINATES
+        def __init__(Self,x, y , z ):
+                Self.x = x
+                Self.y = y
+                Self.z = z
+        #FINDS DISTANCE BETWEEN TWO POINTS
+        def Distance(firstpoint, secondPoint):
+        	d = sqrt(((firstpoint.x-secondPoint.x)**2)+
+                    ((firstpoint.y-secondPoint.y)**2)+
+                    ((firstpoint.z-secondPoint.z)**2))
+        	return d
+        def length(Self):
+          return sqrt((float(Self.x)**2)+(float(Self.y)**2)+(float(Self.z)**2))
+
+		#SHOWS OUTPUT
+        def __str__(Self):
+                return(str(Self.x) + "," + str(Self.y)+ "," + str(Self.z))
+
+class Atom:
+  #ATOM CLASS HOLDS INFORMATION READ IN FROM FILE
+
[…]

### 2022-09-01 manuscript/manuscript.tex
Change some wording, add more todo tags and notes.

@@
 \begin{abstract}
-Grid inhomogeneous solvation theory (GIST) is a method to compute the free energy of hydration of a compound on a 3-dimensional grid.
+Grid inhomogeneous solvation theory (GIST) is a method to compute the free energy of solvation of a compound on a 3-dimensional grid based on molecular dynamics (MD) simulations.
 The high spatial resolution of the GIST output, as well as the decomposition into energy and entropy contributions, allow for highly detailed analyses on both proteins and small molecules.
@@
 Furthermore, we discuss the theory of GIST with a focus on practical aspects, and show several pitfalls and technical difficulties that may occur during a GIST study.
-We assume familiarity with molecular dynamics (MD) simulations as well as the AmberTools package.
+We assume familiarity with molecular dynamics simulations as well as the AmberTools package.
 
@@
 \section{Introduction}
-\todo{ (vah) Sources and better description of all these methods, more literature research, better flow }
+\todo{ (vah) Sources and better description of all these methods, more literature research, QSAR based methods? better flow }
 
@@
 Especially the interaction with water is of utmost importance, as many biological processes occur in a aqueous environment. 
-The free energy of solvation thereby strongly influences real-world problems of interest such as hydrophobic effects, the binding of ligands to a biomolecule, as well as the dynamics and folding of proteins.
+The free energy of solvation thereby strongly influences real-world problems such as hydrophobic effects, the binding of ligands to a biomolecule, as well as the dynamics and folding of proteins.
 
@@
 
-In this tutorial, GIST is applied to calculate solution thermodynamics properties of interest.
+In this tutorial, we aim to give the reader an introduction into GIST and its application to calculate solution thermodynamics properties of interest.
 
@@
 Furthermore, an implementation based on the particle mesh Ewald (PME) \cite{Darden1993-pme} method was presented, which improves the consistency between GIST and popular molecular dynamics engines. 
-Yet further changes allow the inclusion of other rigid solvents than water or allow for the treatment of solvent mixtures and ions in solution.
-
-While previous tutorials gave insightful pointers as to how to setup and run GIST calculations, the post-processing of such calculations was often treated less rigorously.
-While building on these tutorials, an updated guide to calculating solvation properties using GIST is therefore presented here to address the recent improvements to the method.
+Yet further changes allow the inclusion of other rigid solvents than water or allow for the treatment of solvent mixtures and ions in solution.\todo{Sources!}
+
+While previous tutorials gave insightful pointers how to setup and run GIST calculations, their post-processing was often treated less rigorously.
+While building on these tutorials, an updated guide to calculating solvation properties using GIST is therefore presented here to address recent improvements to the method as well as provide a basic template for the analysis and interpretation of GIST calculations.
 
@@
 
-A recent version of \software{cpptraj} should be used, such as the one in AmberTools22 or an installation from GitHub (\url{https://github.com/Amber-MD/cpptraj}).
+A recent version of \software{cpptraj} should be used, such as the one provided in AmberTools22 or a recent installation from GitHub (\url{https://github.com/Amber-MD/cpptraj}).
 The closed-source Amber simulation engine is used in this tutorial to run the necessary MD simulations.
@@
 \todo {If the simulation was ran using GROMACS, mdtraj can be used to convert GROMACS topology and trajectory in to an Amber file format.}
-\todo {vah: I would remove this mention of GROMACs and replace it with a more generall comment. Or we could test if this actually works...}
+\todo {vah: I would remove this mention of GROMACs and replace it with a more general comment. Or we could test if this actually works ( I doubt it does for ff19SB/OPC).}
 In that way, a full Amber topology is available for GIST.
 
-Also note that compressed trajectory formats such as \inlinecode{xtc} might bias the entropy calculation.
-
-Furthermore, a recent Python version (>3.6) should be available with the following packages.
+Also note that compressed trajectory formats such as \inlinecode{xtc} might bias the entropy calculation due to the loss of precision for the atom positions.
+
+Furthermore, a recent Python version (>3.6) should be available with the following packages installed:
 \begin{itemize}
@@
 If the reader prefers to skip the MD and GIST calculation, we also provide the GIST output files with this tutorial, such that the post-processing can be done without any expensive calculations.
-We also provide a Jupyter notebook \cite{Kluyver2016-jupyter,Granger2021-jupyter} with the presented analyses. 
+We additionally provide a Jupyter notebook \cite{Kluyver2016-jupyter,Granger2021-jupyter} with the presented analyses. 
 We recommend JupyterLab to work with the notebook. Jupyter is available at \url{https://jupyter.org/}.
@@
 Here, we only provide a short overview of the theory behind GIST.
-For more detailed information, we recommend one of the more recent publications on developments in GIST. \cite{Kraml2020,Chen2021}
+For more detailed information, we recommend one of the more recent publications on developments in GIST \cite{Kraml2020,Chen2021}.
 
@@
 \subsection{Entropy Calculations in Cpptraj}
-In \software{cpptraj} the calculation of  solvation entropy is handled into two methods.
+In \software{cpptraj} the calculation of  solvation entropy is handled by two methods.
 
@@
 \end{equation}
-where $N_\textit{k}$ is the number of water molecules found in voxel k, $\gamma$ is Euler's constant that accounts for the bias in naive entropy estimator, and $g_\textit{NN}$ is the nearest neighbor estimate.
+where $N_\textit{k}$ is the number of water molecules found in voxel k, $\gamma$ is Euler's constant accounting for the bias of the naive entropy estimator, and $g_\textit{NN}$ is the nearest neighbor estimate.
 
@@
 \subsection{Solvation Energy}
-The solvation energy is readily calculated by summing up the contributions of the water-water interaction and the water-solute interaction.
+The solvation energy is readily calculated by summing up the contributions of the water-water interaction and the water-solute interaction from the force field.
 \begin{equation}
@@
 When computing the total $E_{ww}$ of the system, double-counting of interactions must be avoided.
-The total solvent-solvent energy of the system is as follows.
+The total solvent-solvent energy of the system is as follows:
 
@@
 
-When water is replaced from a small region $R$ of interest, such as a single water molecule, almost all interactions are with water molecules outside of this region.
+However, when water is replaced from a small region $R$ of interest, such as a single water molecule, almost all interactions are with water molecules outside of this region.
 Therefore, there is no double counting, and the full solvent-solvent energy should be used.
@@
 This is not a problem when computing the total free energy of solvation $\dgsolv$, since no additional boundary is created.
-\todo{Do we need this paragraph?}
+\todo{This is a bit unclear to me, could we elaborate on this a bit more? What are we asking of the reader here, how would one consider this effect? -vah}
 
@@
 \end{equation}
+\todo{Is $E_{corr}$ the same in eq 13 and 14? How is it calculated? -vah}
 
@@
 trajin trajectory.nc
-gist gridcntr 10.5 20.1 30.1 griddim 30 30 30 gridspacn 0.50 out gist.dat
+gist gridcntr 10.5 20.1 30.1 griddim 30 30 30 \
+	gridspacn 0.50 out gist.dat
 go
@@
 \paragraph{Postprocessing and Analyzing GIST results}
-Cppraj outputs the quantities calculated in a data explorer "DX" file format which can be easily visualized in Pymol or VMD.
+Cppraj outputs the quantities calculated in a data explorer "OpenDX" (.dx) file format which can be easily visualized in Pymol or VMD.
 Aside from the thermodynamic quantities, GIST also calculates the number densitiies of hydrogen and oxygen in each voxel.
@@
 Several dx files are written by default, such as the energies, entropies, number density of oxygen, and number density of hydrogen.
-All quantities mentioned above is written in the output file (output.dat) and can be used to generate DX files for other quantities.
-
-A minimal PyMOL input to visualize a dx file is shown here.
+All quantities mentioned above are written to the output file (output.dat) and can be used to generate DX files for other quantities.
+
+A minimal PyMOL input to visualize a .dx file is shown here.
 For reference, an input structure is loaded from a file named \inlinecode{solute.pdb} (assuming that this file exists).
@@
 
-\todo{Should be put a picture here?}
+\todo{Should be put a picture here? | Yeah, I am in favour -vah}
 
@@
 But since our equilibration script also contains unrestrained pressure equilibration, we should not run into any trouble.
-
+\todo{Should we mention more sophisticated choices of structure here, eg clustering? Just taking the "equilibrated" structure is kinda random. -vah}
 Then split the equilibrated structure into biotin and streptavidin parts using \software{cpptraj\_remove\_mol.sh}.
@@
 Then, decide on grid dimensions for every system.
-For a system centered at the origin with a spacing of \SI{0.5}{\angstrom} and a buffer of, e.g., \SI{7}{\angstrom} between the solute and the grid wall, you need $4.0\times (d+7)$ bins in the respective direction, where $d$ is the larger of $-$min and $+$max in the respective direction.
-\todo{this does not work with the default gisttools values for the automatic detection of the energy reference. Better use a larger value. }
+For a system centered at the origin with a spacing of \SI{0.5}{\angstrom} and a buffer of, e.g., \SI{15}{\angstrom} between the solute and the grid wall, you need $4.0\times (d+15)$ bins in the respective direction, where $d$ is the larger of $-$min and $+$max in the respective direction.
+\todo{this does not work with the default gisttools values for the automatic detection of the energy reference. Better use a larger value. | changed from 7 to 15 -vah }
 We note that this centered grid is usually larger than it needs to be in some directions.
@@
 Now, you can finally run the actual GIST analysis. Center the molecule exactly in the same way as above, to make sure that its position matches the grid parameters that you decided on.
-We recommend using either the PME or GPU implementation of GIST, depending on whether you have a fast GPU.
-A further advantage of the PME implementation is that it matches the energy calculation of the MD engine more closely.
+We recommend using either the PME or GPU implementation of GIST, depending on whether you have a fast GPU. To use the GPU version of GIST, cpptraj needs to be compiled with the "-cuda" flag set.
+While the GPU version is generally faster, an advantage of the PME implementation is that it matches the energy calculation of the MD engine more closely.
 
@@
 
-The most accurate method to compute a reference energy is to run a separate GIST calculations with the same energy method (PME/GPU) and a similar box size, and then compute the average solvent-solvent energy per molecule.
+The most accurate method to compute a reference energy is to run a separate GIST calculations of the solvent without solute using the same energy method (PME/GPU) and a similar box size, and then compute the average solvent-solvent energy per molecule.
 
@@
 %\todo{Table 1 is placed quite akwardly at the moment for me, maybe enforce its position? -vah. Ok? -fwa}
+\todo{this table is super confusing to me. What's "total"? Doesn't seem to be E-TS. Diff is also not the sum of other total entries. Shouldn't it be TS instead of S?-vah}
 \begin{table}[h]
-	\caption{Free energy contributions for the monomers and the dimer.}\label{tab_dg_monomers_dimer}
+	\caption{Free energy contributions for the monomers and the dimer in kcal$\cdot$mol$^{-1}$.}\label{tab_dg_monomers_dimer}
 	\small

### 2022-09-08 manuscript/manuscript.tex
Change some wording, add more todo tags and notes.
changed the double-counting portion
changed the paragraphs related to double-counting and removed the last paragraph involving reorganization

--- 2022-08-28 manuscript/manuscript.tex
@@
 
-% TODO: delete for final build
-\usepackage{pdfcomment}
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-When computing the total $E_{ww}$ of the system, double-counting of interactions must be avoided.
-The total solvent-solvent energy of the system is as follows.
-
-\begin{equation}
-\Delta E_{ww} = \int \frac{\Delta E_{ww}(\textbf{r})}{2} d\textbf{r}
-\end{equation}
-
-When water is replaced from a small region $R$ of interest, such as a single water molecule, almost all interactions are with water molecules outside of this region.
-Therefore, there is no double counting, and the full solvent-solvent energy should be used.
-
-\begin{equation}
-\Delta E_{ww,R} = \int_R E_{ww}(\textbf{r}) d\textbf{r}
-\end{equation}
-
-When the region of interest $R$ comprises more than one molecule, interactions within this region are double-counted while interactions to the outside are not.
-To take this into account, the solvent-solvent energy between each pair of voxels $E_{ww,kl}$ is needed.
-While this is supported by the standard GIST implementation, it is rarely done due to the large size of the $E_{ww,kl}$ matrix.
-
-We note that replacement of a small or medium solvent region would lead to reorganization of the surrounding solvent.
-However, this depends on the local environment and on \emph{what} the molecule would be replaced by.
-Therefore, the effect of reorganization can not be directly included in the GIST method, but must be judged for each case individually.
-This is not a problem when computing the total free energy of solvation $\dgsolv$, since no additional boundary is created.
-\todo{Do we need this paragraph?}
+When calculating the total $E_{ww}$ of the system, double-counting of interactions must be avoided. For a given region $\bold{R}$, the contribution
+from the interaction of  water molecules outside $\bold{R}$ with the water molecules inside $\bold{R}$ is accounter for in $E_{ww}$. This means that
+the sum of $E_{ww}$ of two different regions, $\bold{R_1}$  and $\bold{R_2}$, taken from the same simulation box will not be equal 
+to the union of $E_{ww}$ of the two regions due to double-counting.
+
+Double-counting is accounted in two different ways.
+The first method uses the water-displacement method.
+This method is applicable when determining the energetic cost of kicking out a water molecule.
+For example, when a ligand binds to the protein, water molecules will be displaced from the binding pocket to
+the bulk.
+
+\begin{equation}
+	\Delta E^R_\textit{ww} = n^\textit{R} E^\textit{bulk}_\textit{ww} - E^\textit{R,corr}_\textit{ww}
+\end{equation}
+
+The second method uses normalized water properties to account the difference in water properties of two different regions. 
+The normalized water property is defined as the water property per water molecule.
+
+\begin{equation}
+	\Delta E^\textit{R,norm}_\textit{ww} = E^\textit{R,norm}_\textit{ww} - 2E^\textit{bulk}_\textit{ww}
+\end{equation}
 
@@
 \end{equation}
-
-\section{Running GIST}
-%\subsection{Running the GIST analysis}
-GIST analysis can be divided in three steps: defining a region of interest, running GIST using \software{cpptraj}, and post-processing and analysis.
-\cite{Ramsey2016}
-
-\paragraph{Identify and define the region of interest}
-In order to run a GIST analysis, the region where the solvent properties will be calculated must be first defined.
-This region is represented by a rectangular prism centered at (x, y, z) with $N_x \times N_y \times N_z$ voxels, where each voxel is cubic with a sidelength of \inlinecode{gridspacn}.
-While integrals over GIST quantities are independent of the voxel size, individual voxel values converge slower at high resolutions.\pdfcomment{The calculation actually can actually get faster at higher resolution, because the same grid is used in the nearest-neighbor search.}
-A default value of 0.5 A is used that provides reasonable discretization for many use-cases.
-
-In \software{ccptraj}, the rectangular prism is defined by parameters of the \inlinecode{gist} command.
-The center of the grid is defined by \inlinecode{gridcntr}, the dimensions of the grid is defined by \inlinecode{griddim}, and the spacing is defined by \inlinecode{gridspacn}.
-
-\begin{lstlisting}
-gist gridcntr <x> <y> <z> \
-griddim <Nx> <Ny> <Nz> gridspacn <val>
-\end{lstlisting}
-
-The region can be defined using a ligand in a specific binding pocket of the protein, a specific residue of the protein, or the whole protein itself.
-A helper script FindCentroid.py \todo{(upload FindCentroid.py!)} can be used to determine the centroid of the region of interest as well as the dimensions of the rectangular prism that will fully enclose the region.
-A pdb file is needed as an input and the rectangular prism will be centered based on the centroid of the given structure.
-
-\begin{lstlisting}
-python FindCentroid.py -i structure.pdb
-\end{lstlisting}
-
-\paragraph{Run GIST using cpptraj}
-GIST is incorporated in cpptraj which requires the topology and the trajectory of the MD simulation to be analayzed.
-The commands needed to run GIST in cpptraj can be written in an input file (e.g gist.in) for convenience.
-The basic comands, assuming a given grid center and a grid size of $30\times30\times30$ voxels with \SI{0.5}{\angstrom} spacing, are:
-
-\begin{lstlisting}
-parm topology.prmtop
-trajin trajectory.nc
-gist gridcntr 10.5 20.1 30.1 griddim 30 30 30 gridspacn 0.50 out gist.dat
-go
-\end{lstlisting}
-
-The first line specifies the amber topology file that was used to run MD.
-The second line species the trajectory file and what frames to be analyzed.
-The third line tells cpptraj to run a GIST calculation using the region defined and print the output of the calculation in a file called gist.dat.
-The last instructs cpptraj to run the calculation.
-The syntax for running cpptraj given this input file is:
-
-\begin{lstlisting}
-cpptraj -i gist.in
-\end{lstlisting}
-
-\pdfcomment{Franz: I removed the list because it is very long and not complete. We can add it to the lists at the end, but then it should be complete.}
-%\todo{Aside from the commands defining the region of interest, there are other options available to specify the GIST analysis.
-Here is the full list that can be added to the third line of the input file:}
-%\begin{itemize}
-%	\item{[doorder] Calculate the water order parameter [reference] for each voxel.}
-%	\item{[doeij] Calculate the triangular matrix representing the water-water interactions between pairs of voxels (see below).}
-%	\item{[skipE] Skip all energy calculations (cannot be specified with ’doeij’).}
-%	\item{[skipS Skip all the entropy calculations.]}
-%	\item{[refdens rdval>] Reference density of bulk water, used in computing $g_O, g_H$, and the translational entropy. Default is 0.0334 molecules/Å3.}
-%	\item{[temp <tval>] Temperature of the input trajectory.}
-%	\item{[noimage] Disable distance imaging in energy calculation.}
-%	\item{[gridcntr <xval> <yval> <zval>] Coordinates (Å) of the center of the grid (default 0.0, 0.0, 0.0).}
-%	\item{[griddim <xval> <yval> <zval>] Grid dimensions along each coordinate axis (default 40, 40, 40).}
-%	\item{[gridspacn <spaceval>] Grid spacing (linear dimension of each voxel) in Angstroms. Values greater than 0.75 Å are not recommended (default 0.5 Å).}
-%	\item{[prefix <filename prefix>] Output file name prefix (default “gist”).}
-%	\item{[ext <grid extension>] Output grid file name extension (default “.dx”).}
-%	\item{[out <output>] Name of the main GIST output file. If not specified set to ’<prefix>-output.dat’.}
-%	\item{[info <info>] Name of main GIST info file. If not specified info is written to standard output.}
-%\end{itemize}
-
-\paragraph{Postprocessing and Analyzing GIST results}
-Cppraj outputs the quantities calculated in a data explorer "DX" file format which can be easily visualized in Pymol or VMD.
-Aside from the thermodynamic quantities, GIST also calculates the number densitiies of hydrogen and oxygen in each voxel.
-Here is a full list of quantities calculated by GIST \todo{This is incomplete!}:
-
-\begin{itemize}
-	\item{[gX] For every element in the main solvent, the number density of atoms found in the voxel, in units of the bulk density.
-If the same element occurs multiple times, the bulk density is scaled accordingly.
-For water, gO and gH are produced.}
-	%\item{[gH] Number density of hydrogen centers found in the voxel in units of the reference bulk density.}
-	\item{[Esw] Mean solute-water interaction energy density.}
-	\item{[Eww] Mean water-water interaction energy density.}
-	\item{[PME] (only if PME was used) water PME energy density.}
-	\item{[dTStrans] First order translational entropy density.}
-	\item{[dTSorient] First order orientational entropy density .}
-	\item{[neighbor] Mean number of waters neighboring the water molecules found in this voxel multiplied by the voxel number density.}
-	\item{[dipole] Magnitude of mean dipole moment (polarization).}
-	\item{[order] (if \inlinecode{doorder} was specified) Average Tetrahedral Order Parameter.}
-	\item{[dipolex] x-component of the mean water dipole moment density}
-	\item{[dipoley] y-component of the mean water dipole moment density}
-	\item{[dipolez] z-component of the mean water dipole moment density}
-	\item{[Eij] (if \inlinecode{doeij} was specified) Water-water interaction matrix.}
-\end{itemize}
-
-Several dx files are written by default, such as the energies, entropies, number density of oxygen, and number density of hydrogen.
-All quantities mentioned above is written in the output file (output.dat) and can be used to generate DX files for other quantities.
-
-A minimal PyMOL input to visualize a dx file is shown here.
-For reference, an input structure is loaded from a file named \inlinecode{solute.pdb} (assuming that this file exists).
-The oxygen density is visualized at an isolevel of 2 (i.e., twice the reference density).
-
-% The current font makes the O look like a tofu / placeholder
-\begin{lstlisting}
-load solute.pdb
-as surface
-load gist-gO.dx, gO
-isosurface gO_high, gO, 2.
-\end{lstlisting}
-
-\todo{Should be put a picture here?}
 
@@
 \end{lstlisting}
+\subsection{Running the GIST analysis}
+\todo{GIST analysis can be divided in three steps.
+\cite{Ramsey2016}}
+
+\todo{Step 1: Identify and define the region of interest
+In order to run a GIST analysis, the region where the solvent properties will be calculated must be first defined.
+This region is represented by a rectangular prism centered at (x,y,z) with a dimension of $L_x x L_y x L_z$.
+This region is further subdivided into three-dimesional grids based on the spacing that's specified.
+A smaller spacing indicates a higher resolution grids, but in return will slow down the calculation.
+A default value of 0.5 A is used that provides reasonable accuracy and speed of the calculation.
+}
+
+\todo{In ccptraj, the rectangular prism is defined by the command:}
+
+\todo {gist gridcntr <xcoordinate>  <ycordinated> <zcoordinate> griddim <xval> <yval> <zval> gridspacn <val>
+}
+
+\todo{The center of the grid is defined by gridcntr, the dimensions of the grid is defined by griddim, and the spacing is defined by gridspacn.}
+
+\todo{The region can be defined using a ligand in a specific binding pocket of the protein, a specific residue of the protein, or the whole protein itself.
+A helper script FindCentroid.py can be used to determine the centroid of the region of interest as well as the dimensions of the rectangular prism that will fully enclose the region.
+A pdb file is needed as an input and the rectangular prism will be centered based on the centroid of the given structure.}
+
+\todo {python FindCentroid.py -i structure.pdb  }
+
+\todo{Step 2: Run GIST using cpptraj}
+\todo{GIST is incorporated in cpptraj which requires the topology and the trajectory of the MD simulation to be analayzed.
+The commands needed to run GIST in cpptraj can be written in an input file (e.g gist.in) for convenience.
+The basic comands are:}
+
+\todo{parm topology.prmtop}
+\todo{trajin trajectory.nc 1 10000}
+\todo{gist gridcntr 10.5 20.1 30.1 griddim 30 30 30 gridspacn 0.50 out gist.dat}
+\todo{go}
+
+\todo{The first line specifies the amber topology file that was used to run MD.
+The second line species the trajectory file and what frames to be analyzed.
+In the above example, GIST will be run from frame 1 to 10000.
+The third line tells cpptraj to run a GIST calculation using the region defined and print the output of the calculation in a file called gist.dat.
+The last instructs cpptraj to run the calculation.
+The syntax for running cpptraj given this input file is:}
+
+\todo{cpptraj -i gist.in}
+
+\todo{Aside from the commands defining the region of interest, there are other options available to specify the GIST analysis.
+Here is the full list that can be added to the third line of the input file:}
+\begin{itemize}
+	\item{[doorder] Calculate the water order parameter [reference] for each voxel.}
+	\item{[doeij] Calculate the triangular matrix representing the water-water interactions between pairs of voxels (see below).}
+	\item{[skipE] Skip all energy calculations (cannot be specified with ’doeij’).}
+	\item{[skipS Skip all the entropy calculations.]}
+	\item{[refdens rdval>] Reference density of bulk water, used in computing $g_O, g_H$, and the translational entropy. Default is 0.0334 molecules/Å3.}
+	\item{[temp <tval>] Temperature of the input trajectory.}
+	\item{[noimage] Disable distance imaging in energy calculation.}
+	\item{[gridcntr <xval> <yval> <zval>] Coordinates (Å) of the center of the grid (default 0.0, 0.0, 0.0).}
+	\item{[griddim <xval> <yval> <zval>] Grid dimensions along each coordinate axis (default 40, 40, 40).}
+	\item{[gridspacn <spaceval>] Grid spacing (linear dimension of each voxel) in Angstroms. Values greater than 0.75 Å are not recommended (default 0.5 Å).}
+	\item{[prefix <filename prefix>] Output file name prefix (default “gist”).}
+	\item{[ext <grid extension>] Output grid file name extension (default “.dx”).}
+	\item{[out <output>] Name of the main GIST output file. If not specified set to ’<prefix>-output.dat’.}
+	\item{[info <info>] Name of main GIST info file. If not specified info is written to standard output.}
+
+\end{itemize}
+
+\todo{Step 3: Postprocessing and Analyzing GIST results}
+\todo{Cppraj outputs the quantities calculated in a data explorer "DX" file format which can be easily visualized in Pymol or VMD.
+Aside from the thermodynamic quantities, GIST also calculates the number densitiies of hydrogen and oxygen in each voxel.
+Here is a full list of quantities calculated by GIST:}
+
+\begin{itemize}
+	\item{[gO] Number density of oxygen centers found in the voxel, in units of the bulk density.}
+	\item{[gH] Number density of hydrogen centers found in the voxel in units of the reference bulk density.}
+	\item{[Esw] Mean solute-water interaction energy density.}
+	\item{[Eww] Mean water-water interaction energy density.}
+	\item{[dTStrans] First order translational entropy density.}
+	\item{[dTSorient] First order orientational entropy density .}
+	\item{[neighbor] Mean number of waters neighboring the water molecules found in this voxel multiplied by the voxel number density.}
+	\item{[dipole] Magnitude of mean dipole moment (polarization).}
+	\item{[order] Average Tetrahedral Order Parameter.}
+	\item{[dipolex] x-component of the mean water dipole moment density}
+	\item{[dipoley] y-component of the mean water dipole moment density}
+	\item{[dipolez] z-component of the mean water dipole moment density}
+	\item{[Eij] Water-water interaction matrix.}
+\end{itemize}
+
+\todo{There are default dx files that will be written as an output, such as the dx files for energies, entropies, number density of oxygen, and number density of hydrogen.
+All quantities mentioned above is written in the output file (output.dat) and can be used to generate DX files for other quantities.}
 
@@
 \end{lstlisting}
-Then, find the extent (minimum and maximum) of the x-, \mbox{y-,} and z-coordinates for each system.
+Then, find the extent (minimum and maximum) of the x-, y-, and z-coordinates for each system.
 This can be done e.g., using the \software{mdtraj} package in Python.
@@
 Visualize the free energy, the total energy, and the entropy, at several iso-levels.
+A minimal PyMOL input might look like this:
+
+% The current font makes the O look like a tofu / placeholder
+\begin{lstlisting}
+load solute-centered.pdb
+show sticks, resname BTN
+as surface, not resname BTN
+load gist-gO.dx, gO
+isosurface gO_high, gO, 2.
+\end{lstlisting}
 

### 2022-09-08 manuscript/manuscript.tex

--- 2022-08-28 manuscript/manuscript.tex
@@
 		\midrule
-		complex      & -1303.4 & -361.2 & -233.1 & -139.9 & -192.0 \\
-		streptavidin & -1180.7 & -360.8 & -242.4 & -145.4 & -188.4 \\
-		biotin       & -24.0   &  -98.1 &  -35.1 &  -21.1 & -105.4 \\
+		complex      & -1303.4 & -361.2 & -233.1 & -139.9 & -1524.7 \\
+		streptavidin & -1180.7 & -360.8 & -242.4 & -145.4 & -1396.1 \\
+		biotin       & -24.0   &  -98.1 &  -35.1 &  -21.1 &  -101.0 \\
 		Diff         & -98.7   &   97.7 &   44.4 &   26.6 &   -27.6 \\

### 2022-09-08 manuscript/manuscript.tex
Update manuscript.tex

--- 2022-09-08 manuscript/manuscript.tex
@@
 
-Double-counting is accounted in two different ways.
-The first method uses the water-displacement method.
-This method is applicable when determining the energetic cost of kicking out a water molecule.
-For example, when a ligand binds to the protein, water molecules will be displaced from the binding pocket to
-the bulk.
-
-\begin{equation}
-	\Delta E^R_\textit{ww} = n^\textit{R} E^\textit{bulk}_\textit{ww} - E^\textit{R,corr}_\textit{ww}
-\end{equation}
+\todo The double counting is accounted for by divinding the Eww by 2. It's done in the code
+\todo postprocessing user can multiply it by 2
 
@@
 The normalized water property is defined as the water property per water molecule.
-
-\begin{equation}
-	\Delta E^\textit{R,norm}_\textit{ww} = E^\textit{R,norm}_\textit{ww} - 2E^\textit{bulk}_\textit{ww}
-\end{equation}
 

### 2022-09-12 manuscript/manuscript.tex
Add headers to lstlistings, add input example for GIST with ions.

--- 2022-09-08 manuscript/manuscript.tex
@@
 \lstdefinestyle{code}{}
+\lstdefinestyle{cpptraj}{title={\centering Cpptraj}}
+\lstdefinestyle{bash}{title={\centering Commandline}}
+\lstdefinestyle{amber-in}{title={\centering Amber input}}
+\lstdefinestyle{pymol}{title={\centering Pymol}}
 \lstdefinestyle{python}{
+	title={\centering Python},
 	language=Python,
@@
 
-When calculating the total $E_{ww}$ of the system, double-counting of interactions must be avoided. For a given region $\bold{R}$, the contribution
-from the interaction of  water molecules outside $\bold{R}$ with the water molecules inside $\bold{R}$ is accounter for in $E_{ww}$. This means that
-the sum of $E_{ww}$ of two different regions, $\bold{R_1}$  and $\bold{R_2}$, taken from the same simulation box will not be equal 
-to the union of $E_{ww}$ of the two regions due to double-counting.
-
-Double-counting is accounted in two different ways.
-The first method uses the water-displacement method.
-This method is applicable when determining the energetic cost of kicking out a water molecule.
-For example, when a ligand binds to the protein, water molecules will be displaced from the binding pocket to
-the bulk.
-
-\begin{equation}
-	\Delta E^R_\textit{ww} = n^\textit{R} E^\textit{bulk}_\textit{ww} - E^\textit{R,corr}_\textit{ww}
-\end{equation}
-
-The second method uses normalized water properties to account the difference in water properties of two different regions. 
-The normalized water property is defined as the water property per water molecule.
-
-\begin{equation}
-	\Delta E^\textit{R,norm}_\textit{ww} = E^\textit{R,norm}_\textit{ww} - 2E^\textit{bulk}_\textit{ww}
-\end{equation}
+When computing the total $E_{ww}$ of the system, double-counting of interactions must be avoided.
+The total solvent-solvent energy of the system is as follows.
+
+\begin{equation}
+\Delta E_{ww} = \int \frac{\Delta E_{ww}(\textbf{r})}{2} d\textbf{r}
+\end{equation}
+
+When water is replaced from a small region $R$ of interest, such as a single water molecule, almost all interactions are with water molecules outside of this region.
+Therefore, there is no double counting, and the full solvent-solvent energy should be used.
+
+\begin{equation}
+\Delta E_{ww,R} = \int_R E_{ww}(\textbf{r}) d\textbf{r}
+\end{equation}
+
+When the region of interest $R$ comprises more than one molecule, interactions within this region are double-counted while interactions to the outside are not.
+To take this into account, the solvent-solvent energy between each pair of voxels $E_{ww,kl}$ is needed.
+While this is supported by the standard GIST implementation, it is rarely done due to the large size of the $E_{ww,kl}$ matrix.
+
+We note that replacement of a small or medium solvent region would lead to reorganization of the surrounding solvent.
+However, this depends on the local environment and on \emph{what} the molecule would be replaced by.
+Therefore, the effect of reorganization can not be directly included in the GIST method, but must be judged for each case individually.
+This is not a problem when computing the total free energy of solvation $\dgsolv$, since no additional boundary is created.
+\todo{Do we need this paragraph?}
 
@@
 
-\begin{lstlisting}
+\begin{lstlisting}[style=cpptraj]
 gist gridcntr <x> <y> <z> \
@@
 
-\begin{lstlisting}
+\begin{lstlisting}[style=bash]
 python FindCentroid.py -i structure.pdb
@@
 
-\begin{lstlisting}
+\begin{lstlisting}[style=cpptraj]
 parm topology.prmtop
@@
 
-\begin{lstlisting}
+\begin{lstlisting}[style=bash]
 cpptraj -i gist.in
@@
 % The current font makes the O look like a tofu / placeholder
-\begin{lstlisting}
+\begin{lstlisting}[style=pymol]
 load solute.pdb
@@
 For the sake of simplicity, you can download prepared and solvated Amber topologies and structures based on 1STP by running
-\begin{lstlisting}
+\begin{lstlisting}[style=bash]
 git clone git@github.com:liedllab/gist-tutorial.git
@@
 If you want to skip this section, run 
-\begin{lstlisting}
+\begin{lstlisting}[style=bash]
 make equilibration-targets
@@
 Use \software{cpptraj} to produce biotin and streptavidin structures based on the complex by stripping the respective other molecule like this:
-\begin{lstlisting}
+\begin{lstlisting}[style=cpptraj]
 parm complex/solvated.parm7
@@
 An example Amber input file for GIST might look like this:
-\begin{lstlisting}
+\begin{lstlisting}[style=amber-in]
 restrained 100 ns NpT
@@
 Run the MD using \software{pmemd.cuda}:
-\begin{lstlisting}
+\begin{lstlisting}[style=bash]
 pmemd.cuda \
@@
 For each system (biotin, streptavidin, complex), produce a centered pdb file using \software{cpptraj}:
-\begin{lstlisting}
+\begin{lstlisting}[style=cpptraj]
 parm solvated.parm7
@@
 
-\begin{lstlisting}
+\begin{lstlisting}[style=cpptraj]
 parm solvated.parm7
@@
 
-\begin{lstlisting}
+\begin{lstlisting}[style=bash]
 rho0=0.03287465
@@
 An example \software{cpptraj} input might look like:
-\begin{lstlisting}
+\begin{lstlisting}[style=cpptraj]
 parm solvated.parm7
@@
 \subsection{GIST with salt-water mixtures}
+In \cite{Waibl2021-gist-salt}, an extension of GIST was presented that can use salt-water mixtures as a solvent.
+This allows treating salting-out effects, although it was shown that the salting-out coefficient is over-predicted by GIST unless higher-order entropies are incorporated.
+Generally, GIST should be able to treat arbitrary solvent mixtures as long as each molecule is sufficiently rigid.
+
+The \software{cpptraj} implementation of GIST allows the user to specify which components of the system to treat as a solute, using the \inlinecode{solute} keyword.
+Everything else will be treated as solvent.
+If the solvent contains more than one molecular species, a list of solvent molecule names (i.e., ``\inlinecode{solventmols WAT,NA,CL}'') must be specified.
+For each molecule in this list, densities (e.g., \inlinecode{g\_mol\_NA}) and energies (e.g., \inlinecode{Eww\_mol\_NA}) will be computed and written to .dx files.
+Note that, e.g., \inlinecode{Esw\_mol\_NA} contains the interactions of every ``\inlinecode{NA}'' molecule with the solute, and \inlinecode{Esw\_mol\_NA} contains the interactions among ``\inlinecode{NA}'' molecules as well as their interactions with all other solvents, divided by 2 to account for double counting.
+Furthermore, the first molecule in this list will be treated as the main solvent, for which first-order entropy values will be computed.
+
+For instance, a GIST calculation in salt water, including the first-order water entropy, can be run as follows:
+
+\begin{lstlisting}[style=cpptraj]
+gist gridcntr <x> <y> <z> \
+griddim <Nx> <Ny> <Nz> gridspacn <val> \
+solute !(:WAT,NA,CL) solventmols WAT,NA,CL
+\end{lstlisting}
+
+The current implementation of the entropy calculation in \software{cpptraj} requires at least 3 atoms in the main solvent.
+To compute the first-order entropy of the ions, as well as an approximate second order entropy, Python code is available on \url{https://github.com/liedllab/second-disorder}.
 

### 2022-09-13 manuscript/manuscript.tex
Update manuscript.tex
update on double counting
Added WIP cheatsheet

--- 2022-09-08 manuscript/manuscript.tex
@@
 
-When calculating the total $E_{ww}$ of the system, double-counting of interactions must be avoided. For a given region $\bold{R}$, the contribution
-from the interaction of  water molecules outside $\bold{R}$ with the water molecules inside $\bold{R}$ is accounter for in $E_{ww}$. This means that
-the sum of $E_{ww}$ of two different regions, $\bold{R_1}$  and $\bold{R_2}$, taken from the same simulation box will not be equal 
-to the union of $E_{ww}$ of the two regions due to double-counting.
-
-Double-counting is accounted in two different ways.
-The first method uses the water-displacement method.
-This method is applicable when determining the energetic cost of kicking out a water molecule.
-For example, when a ligand binds to the protein, water molecules will be displaced from the binding pocket to
-the bulk.
+When computing the total $E_{ww}$ of the system, double-counting of interactions must be avoided.
+The total solvent-solvent energy of the system is as follows:
 
@@
 
-The second method uses normalized water properties to account the difference in water properties of two different regions. 
-The normalized water property is defined as the water property per water molecule.
+However, when water is replaced from a small region $R$ of interest, such as a single water molecule, almost all interactions are with water molecules outside of this region.
+Therefore, there is no double counting, and the full solvent-solvent energy should be used.
 
@@
 \end{equation}
+
+When the region of interest $R$ comprises more than one molecule, interactions within this region are double-counted while interactions to the outside are not.
+To take this into account, the solvent-solvent energy between each pair of voxels $E_{ww,kl}$ is needed.
+While this is supported by the standard GIST implementation, it is rarely done due to the large size of the $E_{ww,kl}$ matrix.
+
+We note that replacement of a small or medium solvent region would lead to reorganization of the surrounding solvent.
+However, this depends on the local environment and on \emph{what} the molecule would be replaced by.
+Therefore, the effect of reorganization can not be directly included in the GIST method, but must be judged for each case individually.
+This is not a problem when computing the total free energy of solvation $\dgsolv$, since no additional boundary is created.
+\todo{This is a bit unclear to me, could we elaborate on this a bit more? What are we asking of the reader here, how would one consider this effect? -vah}
 
@@
 \paragraph{Postprocessing and Analyzing GIST results}
-Cppraj outputs the quantities calculated in a data explorer "DX" file format which can be easily visualized in Pymol or VMD.
-Aside from the thermodynamic quantities, GIST also calculates the number densitiies of hydrogen and oxygen in each voxel.
+Cpptraj outputs the quantities calculated as a "open data explorer" (.dx) file format which can be easily visualized in Pymol or VMD.
+Aside from the thermodynamic quantities, GIST also calculates the number densities of hydrogen and oxygen in each voxel.
 Here is a full list of quantities calculated by GIST \todo{This is incomplete!}:
@@
 \subsection{GIST with salt-water mixtures}
+\section{Cheatsheet}
+%This will be a cheatsheet containing all options for cpptrajs GIST implementation
+\subsection{Installation}
+GIST calculations in cpptraj can be sped up through various levels of parallelization.
+When cpptraj was compiled with the '-cuda' configure flag set, the GPU accelerated version of GIST is used automatically.
+
+\subsection{The CPPTRAJ action}
+\subsection{Output}
+GIST calculations put out a variety of data sorted by voxel into a file specified by the "out" keyword.
+Some of the output data is also automatically written to "open data explorer" (.dx) files for convenient visualisation in software such as VMD or PyMOL.
+The following columns can be found in the output file:
+\begin{tabular}{l l p{.3\textwidth}}
+\toprule
+gX & X & For every element in the main solvent, the number density of atoms found in the voxel, in units of the bulk density.
+If the same element occurs multiple times, the bulk density is scaled accordingly.
+\\
+Esw & X &  Mean solute-water interaction energy density. \\
+Eww & X & Mean water-water interaction energy density. \\
+PME & [pme] & Water PME energy density. \\
+U\_PME & [pme] & Water PME energy density. \\
+dTStrans & X & First order translational entropy density. \\
+dTSorient & X & First order orientational entropy density . \\
+neighbor & X & Mean number of waters neighboring the water molecules found in this voxel multiplied by the voxel number density. \\
+dipole & X & Magnitude of mean dipole moment (polarization). \\
+order & X & (if \inlinecode{doorder} was specified) Average Tetrahedral Order Parameter. \\
+dipolex & X & x-component of the mean water dipole moment density \\
+dipoley & X & y-component of the mean water dipole moment density \\
+dipolez & X & z-component of the mean water dipole moment density \\
+Eij & X & (if \inlinecode{doeij} was specified) Water-water interaction matrix. \\
+\bottomrule
+\end{tabular}
 

### 2022-09-13 manuscript/manuscript.tex
Add further text
Add cheatsheet
Add structure preperation
Remove cleared comments

@@
 Furthermore, an implementation based on the particle mesh Ewald (PME) \cite{Darden1993-pme} method was presented, which improves the consistency between GIST and popular molecular dynamics engines. 
-Yet further changes allow the inclusion of other rigid solvents than water or allow for the treatment of solvent mixtures and ions in solution.\todo{Sources!}
+Yet further changes allow the inclusion of other rigid solvents than water or 
+allow for the treatment of solvent mixtures and ions in solution. 
+\todo{Sources!}
 
@@
 The electrostatic and Lennard-Jones parts $E_\text{elec}$ and $E_\text{lj}$ are computed separately.
-\todo{E$_{LJ}$ seems more intuitive to me, need to check what the GIST literature is using -vah. They use $E_{lj}$, but LJ in the text -fwa.}
+
 \begin{equation}
@@
 In \software{ccptraj}, the rectangular prism is defined by parameters of the \inlinecode{gist} command.
-The center of the grid is defined by \inlinecode{gridcntr}, the dimensions of the grid is defined by \inlinecode{griddim}, and the spacing is defined by \inlinecode{gridspacn}.
+The center of the grid is defined by \inlinecode{gridcntr}, the dimensions of 
+the grid is defined by \inlinecode{griddim} (in voxel counts), and the spacing 
+is defined by \inlinecode{gridspacn} (in $\AA$).
 
@@
 
-The region can be defined using a ligand in a specific binding pocket of the protein, a specific residue of the protein, or the whole protein itself.
-A helper script FindCentroid.py \todo{(upload FindCentroid.py!)} can be used to determine the centroid of the region of interest as well as the dimensions of the rectangular prism that will fully enclose the region.
-A pdb file is needed as an input and the rectangular prism will be centered based on the centroid of the given structure.
+The region can be defined using a ligand in a specific binding pocket of the 
+protein, a specific residue of the protein, or the whole protein itself. A 
+helper script FindCentroid.py can be used to determine the centroid of the 
+region of interest as well as the dimensions of the rectangular prism that will 
+fully enclose the region. A pdb file is needed as an input and the rectangular 
+prism will be centered based on the centroid of the given structure.
 
@@
 \end{lstlisting}
-
-\pdfcomment{Franz: I removed the list because it is very long and not complete. We can add it to the lists at the end, but then it should be complete.}
-%\todo{Aside from the commands defining the region of interest, there are other options available to specify the GIST analysis.
-Here is the full list that can be added to the third line of the input file:}
-%\begin{itemize}
-%	\item{[doorder] Calculate the water order parameter [reference] for each voxel.}
-%	\item{[doeij] Calculate the triangular matrix representing the water-water interactions between pairs of voxels (see below).}
-%	\item{[skipE] Skip all energy calculations (cannot be specified with ’doeij’).}
-%	\item{[skipS Skip all the entropy calculations.]}
-%	\item{[refdens rdval>] Reference density of bulk water, used in computing $g_O, g_H$, and the translational entropy. Default is 0.0334 molecules/Å3.}
-%	\item{[temp <tval>] Temperature of the input trajectory.}
-%	\item{[noimage] Disable distance imaging in energy calculation.}
-%	\item{[gridcntr <xval> <yval> <zval>] Coordinates (Å) of the center of the grid (default 0.0, 0.0, 0.0).}
-%	\item{[griddim <xval> <yval> <zval>] Grid dimensions along each coordinate axis (default 40, 40, 40).}
-%	\item{[gridspacn <spaceval>] Grid spacing (linear dimension of each voxel) in Angstroms. Values greater than 0.75 Å are not recommended (default 0.5 Å).}
-%	\item{[prefix <filename prefix>] Output file name prefix (default “gist”).}
-%	\item{[ext <grid extension>] Output grid file name extension (default “.dx”).}
-%	\item{[out <output>] Name of the main GIST output file. If not specified set to ’<prefix>-output.dat’.}
-%	\item{[info <info>] Name of main GIST info file. If not specified info is written to standard output.}
-%\end{itemize}
 
@@
 
+The choice of the initial structure to be analyzed is crucial to the results 
+you can expect from the GIST calculation. Crystal structures may not be 
+representative of the dominant solution structure for a given analysis target. 
+In our example we use the final structure after a short equilibration run. A 
+more sophisticated approach would be identifying a dominant structure out of 
+the ensemble of sampled structures.  Depending on what properties you are 
+interested in, running a MD simulation of your structure beforehand and 
+choosing a dominant structure or even multiple structures through clustering 
+might therefore be beneficial.
+When starting with a crystal structure, special care has to be taken during the 
+preperation of the structure. Make sure to protonate your structure and take 
+care to choose a proper protonation state for the varying chemical moeties in 
+your structure. You will need to solvate your structure in a solvent box, which 
+should be chosen large enough so that the solvent around the region of interest 
+is not influenced through periodic images.
+
 When calculating solvation free energy differences between a complex and the corresponding monomers (the dissociated state), there are two options for dealing with configurational diversity in the structures.
@@
 (Adapt for biotin as necessary.)
-\todo{vah: Maybe include a section on how to prepare a protein structure before equilibration.
-Protonation, force field and a large enough iso solvation box seem important to me.
-I will probably also at least mention what simulation settings we are running later on.}
 %We provide a script called \software{cpptraj\_remove\_mol.sh} that removes a single molecule from a topology and structure using \software{cpptraj} without altering the rest of the system.
@@
 Note that the barostat might interfere with restrained equilibration of the complex, since different molecules are moved relative to each other when the volume changes.
-But since our equilibration script also contains unrestrained pressure equilibration, we should not run into any trouble.
-\todo{Should we mention more sophisticated choices of structure here, eg clustering? Just taking the "equilibrated" structure is kinda random. -vah}
+But since our equilibration script also contains unrestrained pressure 
+equilibration, we should not run into any trouble.
+
 Then split the equilibrated structure into biotin and streptavidin parts using \software{cpptraj\_remove\_mol.sh}.
-After that, you can use \software{equilibration.py -R} to run short restrained equilibrations (where only the water is allowed to move) on the individual systems.
+After that, you can use \software{equilibration.py -R} to run short restrained 
+equilibrations (where only the water is allowed to move) on the individual 
+systems.
 
@@
 For a system centered at the origin with a spacing of \SI{0.5}{\angstrom} and a buffer of, e.g., \SI{15}{\angstrom} between the solute and the grid wall, you need $4.0\times (d+15)$ bins in the respective direction, where $d$ is the larger of $-$min and $+$max in the respective direction.
-\todo{this does not work with the default gisttools values for the automatic detection of the energy reference. Better use a larger value. | changed from 7 to 15 -vah }
+\todo{ This is handled differently by FindCentroid.py: d = max - min; voxelnr = 
+d*5 for x,y,z each; we should handle it similiar or at least mention 
+FindCentroid.py here. 
+What does gisttools do if the grid is too small? -vah }
 We note that this centered grid is usually larger than it needs to be in some directions.
@@
 import matplotlib.pyplot as plt
-gist = gt.gist.load_gist_file(
+gist_obj = gt.gist.load_gist_file(
     "gist.dat", struct="solute-centered.pdb")
-bins, eww = gist.rdf(
+bins, eww = gist_obj.rdf(
     "Eww_unref_norm", bins=100, rmax=20., normalize="norm")
 plt.plot(bins, eww)
-eww_auto = gist.detect_reference_value().values[0]
+eww_auto = gist_obj.detect_reference_value().values[0]
 plt.axhline(eww_auto, color='k')
@@
 \end{lstlisting}
-\todo{While not massively important, I don't like that we call the gist object here "gist", as that's the same name as gisttools subpackage gist.
-"gt.gist.bla" and "gist.bla" therefore look quite similiar, but do very different things.
--vah}
+
 The expected output of this code is shown in Figure~\ref{fig_ewwref}.
@@
 \end{lstlisting}
-\todo{Maybe add more complete print calls? E.g. print('Biotin: Frames=\{biotin.n\_frames\}, Rho=\{biotin.rho0\} -vah)}
 Assign reference energies and check them for plausibility.
@@
 \subsection{Installation}
-GIST calculations in cpptraj can be sped up through various levels of parallelization.
-When cpptraj was compiled with the '-cuda' configure flag set, the GPU accelerated version of GIST is used automatically.
-
-\subsection{The CPPTRAJ action}
-\subsection{Output}
-GIST calculations put out a variety of data sorted by voxel into a file specified by the "out" keyword.
-Some of the output data is also automatically written to "open data explorer" (.dx) files for convenient visualisation in software such as VMD or PyMOL.
-The following columns can be found in the output file:
-\begin{tabular}{l l p{.3\textwidth}}
+GIST calculations in cpptraj can be sped up through various levels of 
+parallelization. When cpptraj was compiled with the '-cuda' configure flag set, 
+the GPU accelerated version of GIST is used automatically.
+
+\begin{Checklists*}
+
+\begin{checklist}{The CPPTRAJ action}
+
+\textbf{Options}\\
+Various flags and options can be provided when running a GIST calculation in 
+cpptraj. A list of possible and required options is provided here:
+
+\begin{tabular}{l p{.7\textwidth}}
 \toprule
+doorder & Calculate the water order parameter [reference] for each voxel\\
+doeij & Calculate the triangular matrix representing the water-water 
+interactions between pairs of voxels.\\
+skipE & Skip all energy calculations (cannot be specified with ’doeij’)\\
+skipS & Skip all the entropy calculations.\\
+refdens <rdval> & Reference density of bulk water, used in computing 
+$g_O, g_H$, and the translational entropy. Default is 0.0334 molecules/Å3. \\
+temp <tval> & Temperature of the input trajectory. \\
+noimage &  Disable distance imaging in energy calculation. \\
+gridcntr <xval> <yval> <zval> & Coordinates (Å) of the center of the grid 
+(default 0.0, 0.0, 0.0). \\
+griddim <xval> <yval> <zval> & Grid dimensions along each coordinate axis 
+(default 40, 40, 40).\\
+gridspacn <spaceval> & Grid spacing (linear dimension of each voxel) in 
+Angstroms. Values greater than 0.75 Å are not recommended (default 0.5 Å).\\
+prefix <filename prefix> & Output file name prefix (default “gist”).\\
+ext <grid extension> & Output grid file name extension (default “.dx”). \\
+out <output> & Name of the main GIST output file. If not specified set to 
+’<prefix>-output.dat’.\\
+info <info> & Name of main GIST info file. If not specified info is 
+written to standard output. \\
+\bottomrule
+\end{tabular}
+
+\textbf{Output}\\
+GIST calculations put out a variety of data sorted by voxel into a file 
+specified by the "out" keyword. Some of the output data is also automatically 
+written to "open data explorer" (.dx) files for convenient visualisation in 
+software such as VMD or PyMOL. The following columns can be found in the output 
+file:
+\begin{tabular}{l l p{.8\textwidth}}
+\toprule
+Name & Keyword & Description\\
+\midrule
 gX & X & For every element in the main solvent, the number density of atoms found in the voxel, in units of the bulk density.
@@
 \end{tabular}
-
+\end{checklist}
+\end{Checklists*}
 \section{Checklists}

### 2022-09-13 manuscript/manuscript.tex

@@
 
-A minimal PyMOL input to visualize a .dx file is shown here.
+A minimal PyMOL input to visualize a ..dx file is shown here.
 For reference, an input structure is loaded from a file named \inlinecode{solute.pdb} (assuming that this file exists).
@@
 \subsection{GIST with salt-water mixtures}
+In \cite{Waibl2021-gist-salt}, an extension of GIST was presented that can use salt-water mixtures as a solvent.
+This allows treating salting-out effects, although it was shown that the salting-out coefficient is over-predicted by GIST unless higher-order entropies are incorporated.
+Generally, GIST should be able to treat arbitrary solvent mixtures as long as each molecule is sufficiently rigid.
+
+The \software{cpptraj} implementation of GIST allows the user to specify which components of the system to treat as a solute, using the \inlinecode{solute} keyword.
+Everything else will be treated as solvent.
+If the solvent contains more than one molecular species, a list of solvent molecule names (i.e., ``\inlinecode{solventmols WAT,NA,CL}'') must be specified.
+For each molecule in this list, densities (e.g., \inlinecode{g\_mol\_NA}) and energies (e.g., \inlinecode{Eww\_mol\_NA}) will be computed and written to .dx files.
+Note that, e.g., \inlinecode{Esw\_mol\_NA} contains the interactions of every ``\inlinecode{NA}'' molecule with the solute, and \inlinecode{Esw\_mol\_NA} contains the interactions among ``\inlinecode{NA}'' molecules as well as their interactions with all other solvents, divided by 2 to account for double counting.
+Furthermore, the first molecule in this list will be treated as the main solvent, for which first-order entropy values will be computed.
+
+For instance, a GIST calculation in salt water, including the first-order water entropy, can be run as follows:
+
+\begin{lstlisting}[style=cpptraj]
+gist gridcntr <x> <y> <z> \
+griddim <Nx> <Ny> <Nz> gridspacn <val> \
+solute !(:WAT,NA,CL) solventmols WAT,NA,CL
+\end{lstlisting}
+
+The current implementation of the entropy calculation in \software{cpptraj} requires at least 3 atoms in the main solvent.
+To compute the first-order entropy of the ions, as well as an approximate second order entropy, Python code is available on \url{https://github.com/liedllab/second-disorder}.
+
 \section{Cheatsheet}
@@
 \end{Checklists*}
+
 \section{Checklists}

### 2022-09-13 manuscript/manuscript.tex
Add example visualizations, minor text updates

@@
 
+\todo{Remove the part about QM vs.\ MM. We are doing MD and cite ALL solvation methods. COSMO, other implicit solvents, information theory, QSAR, Pratt and coworkers, Mobley and coworkers}
 Molecular mechanics based approaches describe the underlying PES through empirically fit force-fields, which reduce the computational demand of these calculations tremendously.
@@
 the grid is defined by \inlinecode{griddim} (in voxel counts), and the spacing 
-is defined by \inlinecode{gridspacn} (in $\AA$).
+is defined by \inlinecode{gridspacn} (in Å).
 
@@
 \begin{table}[h]
-	\caption{Free energy contributions for the monomers and the dimer in kcal$\cdot$mol$^{-1}$.}\label{tab_dg_monomers_dimer}
+	\caption{Free energy contributions for the monomers and the dimer in kcal$\cdot$mol$^{-1}$.
+``total'' is calculated as $E^\text{internal} + \Delta E^\text{GIST} - \Delta S^\text{scaled}$.}\label{tab_dg_monomers_dimer}
 	\small
-	\begin{tabular}{lrrrrr}
+	\begin{tabular}{@{}lrrrrr@{}}
 		\toprule
@@
 
-\begin{tabular}{l p{.7\textwidth}}
+\begin{tabular}{@{}l p{.7\textwidth}@{}}
 \toprule
@@
 file:
-\begin{tabular}{l l p{.8\textwidth}}
+\begin{tabular}{@{}l l p{.8\textwidth}@{}}
 \toprule

### 2022-09-15 manuscript/manuscript.tex
Make Intro more solvation-focused, add some references.

@@
 
-The computational calculation of thermodynamical solvation properties is therefore of particular interest and has been tackled by a wide array of methods.
-Quantum mechanical methods provide the most rigorous description of the underlying potential energy surface (PES) but suffer from high computational cost, allowing only the treatment of small systems and few solvent molecules.
+%The computational calculation of thermodynamical solvation properties is therefore of particular interest and has been tackled by a wide array of methods.
+%Quantum mechanical methods provide the most rigorous description of the underlying potential energy surface (PES) but suffer from high computational cost, allowing only the treatment of small systems and few solvent molecules.
 %More recently, QM/MM approaches and potentials based on deep learning \cite{Smith2017-ani} are trying to bridge the gap to computationally less demanding methods while keeping the accuracy of DFT methods.
 
-\todo{Remove the part about QM vs.\ MM. We are doing MD and cite ALL solvation methods. COSMO, other implicit solvents, information theory, QSAR, Pratt and coworkers, Mobley and coworkers}
-Molecular mechanics based approaches describe the underlying PES through empirically fit force-fields, which reduce the computational demand of these calculations tremendously.
-This allows for the treatment of large biological systems such as proteins in aqueous solution.
-Implicit solvent methods such as MM/PBSA or MM/GBSA \cite{Sitkoff1994-pbsa,Kollman2000-mmpbsa} can quickly provide results for a large number of structures, but yield inaccurate results when compared to more rigorous explicit solvent methods \cite{Genheden2015-mmpbsa-review}.
-On the other hand, the free energy of solvation can be calculated in a statistically rigorous way using explicit solvent simulations combined with alchemical methods such as free energy perturbation (FEP) \cite{Zwanzig1954-reweighting} or thermodynamic integration(TI) \cite{Kirkwood1935-ti}.
+A wide range of methods have been devised to compute the free energy of hydration.
+Data-driven methods such as QSPR or UNIFAC can deliver good results \cite{Borhani2019-qspr,Fredenslund1975-unifac}, but might be limited to molecules that are similar to those in the training set.
+Polarizable continuum models (PCMs) \cite{Miertus1981-pcm,Klamt1993-cosmo} such as treat the solvent phase as a homogeneous phase that reacts to the electrostatic potential of the solute \cite{Mennucci2010-pcm}.
+%Molecular mechanics based approaches describe the underlying PES through empirically fit force-fields, which reduce the computational demand of these calculations tremendously.
+%This allows for the treatment of large biological systems such as proteins in aqueous solution.
+Methods such as MM/PBSA or MM/GBSA \cite{Sitkoff1994-pbsa,Kollman2000-mmpbsa} combine an implicit solvent treatment of electrostatics with structural sampling of the solute.
+They can quickly provide results for a large number of structures, but yield inaccurate results when compared to more rigorous explicit solvent methods \cite{Genheden2015-mmpbsa-review}.
+
+On the other hand, explicit solvation methods model the solvent using individual molecules.
+They are generally more accurate than implicit methods, but require a statistical sampling of the solvent conformations \cite{Liu2016-md-solubility,Swails2014-cphmd}.
+Molecular dynamics (MD) simulations provide an accurate way of modeling solvent packing in confined areas \cite{}\todo{ask Tom for a citation!} and general hydrophobic effects \cite{Pratt2016-hydrophobicity}.
+
+The free energy of solvation can be calculated in a statistically rigorous way using alchemical methods \cite{Liu2016-md-solubility,Mobley2009-dgsolv,Mobley2014-freesolv} such as free energy perturbation (FEP) \cite{Zwanzig1954-reweighting} or thermodynamic integration (TI) \cite{Kirkwood1935-ti}.
 However, they are unable to provide a spatial interpretation of the results.

### 2022-09-20 manuscript/manuscript.tex
add text about GPU, PME, and 1 sentence about other solvents

@@
 \section{Recent developments}
+In recent years, several updates to the original GIST implementation have been published.
+The basic functionality of the program, however, has not been changed.
+In this chapter, we will shortly present each of those updates.
+\subsection{GPU implementation}
+In Reference~\cite{Kraml2020}, a GPU implementation of the energy calculation was implemented, which typically increases the speed by 1--2 orders of magnitude.
+
+If \software{cpptraj} is compiled with GPU support, the energy calculation uses the GPU automatically, without any additional input.
+However, PME-GIST is not supported on the GPU.
+
+\subsection{PME implementation}
+In Reference~\cite{Chen2021}, a PME implementation of the GIST energy was presented.
+For typical systems, this implementation is slightly slower than GPU-GIST, but much faster than the original CPU code.
+Furthermore, PME-GIST offers the best agreement between GIST energies and those observed in an MD simulation.
+
+To run a PME-GIST calculation, the \inlinecode{pme} flag to the \inlinecode{gist} action in \software{cpptraj} can be used.
+The output will contain two additional columns: \inlinecode{PME\_dens} and \inlinecode{PME\_norm}.
+They contain the potential of solute and solvent evaluated at the solvent positions and divided by two.
+Furthermore, the \inlinecode{Eww} and \inlinecode{Esw} columns are also computed using PME, and are reported as usual.
+For typical use cases, it is recommended to use \inlinecode{Eww} and \inlinecode{Esw} over the \inlinecode{PME} column.
+
 \subsection{GIST with non-water solvents}
+In References~\cite{Kraml2020,Kamenik2020-gist-macrocycles,Waibl2022-gist-solvents}, an extension of GIST towards solvents other than water is described.
+
 \subsection{GIST with salt-water mixtures}

### 2022-09-22 manuscript/manuscript.tex

--- 2022-09-15 manuscript/manuscript.tex
@@
 All quantities mentioned above are written to the output file (output.dat) and can be used to generate DX files for other quantities.
-
-A minimal PyMOL input to visualize a ..dx file is shown here.
-For reference, an input structure is loaded from a file named \inlinecode{solute.pdb} (assuming that this file exists).
-The oxygen density is visualized at an isolevel of 2 (i.e., twice the reference density).
-
-% The current font makes the O look like a tofu / placeholder
-\begin{lstlisting}[style=pymol]
-load solute.pdb
-as surface
-load gist-gO.dx, gO
-isosurface gO_high, gO, 2.
-\end{lstlisting}
-
-\todo{Should be put a picture here? | Yeah, I am in favour -vah}
 
@@
 To compute the first-order entropy of the ions, as well as an approximate second order entropy, Python code is available on \url{https://github.com/liedllab/second-disorder}.
+
+\section{Visualization}
+
+\subsection{Loading a DX file in Pymol }
+A minimal PyMOL input to visualize a dx file is shown here.
+For reference, an input structure is loaded from a file named \inlinecode{streptavidin.pdb}.
+The oxygen density is visualized at an isolevel of 2 (i.e., twice the reference density).
+
+% The current font makes the O look like a tofu / placeholder
+\begin{lstlisting}[style=pymol]
+	load streptavidin.pdb
+	as surface
+	color gray70, streptavidin
+	load gist-gO.dx, gO
+	isomesh gO_high, gO, 2
+	color slate, gO_high
+\end{lstlisting}
+
+\begin{figure}
+	\centering
+	\includegraphics[width=1.0\linewidth]{figures/streptavidin_gO_high_surf.png}
+	\caption{Oxygen density around strepativin at twice the reference bulk density}\label{streptavidin_gO}
+\end{figure}
+
+\subsection{Solvent Accessible Surface Area (SASA)}
+
+The solvent accessible surface area (SASA) represents  the voxels that are close to the protein surface and occupied by oxygen water.
+It represents the surface where water can go, and potentially surfaces that a ligand atom can occupy.
+
+The surface is generated from the oxygen density (gO.dx) file using \software{gistpp}. 
+A voxel is deemed to be  part of the surface if it is occupied by water (gO > 0.3) and has at least one neighboring empty voxel (gO < 0.1).
+
+\begin{lstlisting}[style=bash]
+	gistpp -op sasa -i gO.dx -o sasa.dx
+\end{lstlisting}
+
+\begin{figure}
+	\centering
+	\includegraphics[width=1.0\linewidth]{streptavidin_sasa_close.png}
+	\caption{SASA for streptavidin focused on the binding pocket}\label{streptavidin_sasa}
+\end{figure}
+
+In the above figure, the SASA was generated from a simulation of the protein in the absence of the ligand.
+When the ligand is placed in the binding pocked, all atoms are within the SASA.
+
+\subsection{Region around the binding pocket}
+Specific regions in a protein can be defined to calculate the thermodynamic quantities for that region.
+This is significant specially when calculating the properties of water that will be expelled when a ligand binds to the protein.
+
+For example, voxels that are part of the binding pocket can be defined using a ligand molecule and a distance criteria.
+In \software{gistpp}, voxels that are within the distance criteria from any ligand heavy atom is considered as part of the region.
+The resulting file is a binary file and can be visualized using \software{pymol}.
+
+\begin{lstlisting}[style=bash]
+	gistpp -op defbp -i gO.dx -i2 ligand.pdb
+
+	-opt const 3.5 -o gO_3.5.dx
+\end{lstlisting}
+
+In this example, the region defined in the binding pocket includes all voxels that are within 3.5 A of the heavy atoms of biotin.
+
+\begin{figure}
+	\centering
+	\includegraphics[width=1.0\linewidth]{streptavidin_bp.png}
+	\caption{Volume of strepatividin defined around 3.5 A of biotin. }\label{streptavidin_sasa}
+\end{figure}
 

### 2022-10-05 manuscript/manuscript.tex
Update 6-dimensional density formula
I think it is incorrect to say that the 6D density g(r,w) is computed
from the individual *voxel* densities g_vox(r) and g_vox(w). Therefore,
I deleted the subscripts.
Update text about different solvents

--- 2022-09-20 manuscript/manuscript.tex
@@
 \begin{equation}
-g_\textit{vox} \left( \textbf{r}, \omega \right) \approx g_\textit{vox}(\textbf{r}) g_\textit{vox}(\omega)
+g_\textit{vox} ( \textbf{r}, \omega ) \approx g(\textbf{r}) g(\omega)
 \end{equation}
@@
 
+To run a GIST calculation in a solvent other than water, use the \inlinecode{solute} flag together with a \software{cpptraj} selection mask.
+Everything that is not in the \inlinecode{solute} mask is treated as solvent.
+If there are multiple solvent species, use the \inlinecode{solventmols} option as described in Section~\ref{sec-salt-water}.
+The code will automatically choose three atoms which are used to define the solvent orientation, and print this selection to the output.
+
+Sometimes, \software{cpptraj} does not automatically choose the optimal atoms.
+In methanol, for instance, one can either incorporate the C--H bonds or the O--H bond in the selection.
+The latter is probably more relevant since it can incorporate hydrogen bonding effects.
+Assuming that the alcohol hydrogen is called H1, this could be specified as follows:
+
+\begin{lstlisting}[style=cpptraj]
+gist gridcntr <x> <y> <z> \
+griddim <Nx> <Ny> <Nz> gridspacn <val> \
+solute ^1 rigidatoms O1 H1 C1
+\end{lstlisting}
+
+Note that the central atom of the \inlinecode{rigidatoms} selection goes first.
+
 \subsection{GIST with salt-water mixtures}
+\label{sec-salt-water}
 In \cite{Waibl2021-gist-salt}, an extension of GIST was presented that can use salt-water mixtures as a solvent.

### 2022-10-05 manuscript/manuscript.tex
Start of the restructuring process
Changed some wording to reflect the split into a bsaic and advanced
tutorial
Added some tables to the Running GIST section for more information
concerning cpptrajs gist option.

--- 2022-09-22 manuscript/manuscript.tex
@@
 \usepackage{listings}
+\usepackage{tabularx}
 
@@
 In addition to a short introduction of the method and the theory behind it, we present an overview of the various different flavors of GIST methods.
-
-Furthermore, a GIST study of the streptavidin-biotin complex is presented, which reflects on many common aspects of GIST studies.
-Using this test system, we present solvation analysis of a small molecule (biotin) as well as a biomolecule (streptavidin).
+We ease the user into the basic usage of GIST by presenting a calculation for a small molecule.
+Building on the basics, a more advanced GIST study of the streptavidin-biotin complex is presented, which reflects on many common aspects of GIST studies.
+Using these test systems, we present solvation analysis of a small molecule (biotin) as well as a biomolecule (streptavidin).
 We show how to interpret the three-dimensional contributions of solvation free energy in a binding pocket and around the ligand.
@@
 Previous experience of GIST is not necessary.
-However, we only provide a short explanation of the theory behind GIST and point the reader towards the primary sources for more detailed information.
+However, we only provide a short explanation of the theory behind GIST in section \ref{sec:theory} at the end of this tutorial.
+For more detailed information, we point the reader towards the primary sources for more detailed information.
 
@@
 Tutorials requiring specific software packages must provide instructions and files for the referenced version of the software.
-The main implementation of GIST is part of the MD analysis software \software{cpptraj} as part of AmberTools \cite{amber22}.
+The main implementation of GIST is part of the MD analysis software \software{cpptraj} released with AmberTools \cite{amber22}.
 
 A recent version of \software{cpptraj} should be used, such as the one provided in AmberTools22 or a recent installation from GitHub (\url{https://github.com/Amber-MD/cpptraj}).
-The closed-source Amber simulation engine is used in this tutorial to run the necessary MD simulations.
 
@@
 We recommend using a current version of \software{PyMol}\cite{pymol} or \software{VMD}\cite{vmd}.
-\section{Theory}
-
-GIST is an implementation of inhomogeneous fluid solvation theory (IST) \cite{Lazaridis1998} that discretizes the free energy of solvation $\dgsolv$ on a three-dimensional grid. 
-It was first devised by Nguyen et al.
-\cite{Nguyen2012} and its implementation in \software{cpptraj} was thoroughly described by Ramsey et al.
-\cite{Ramsey2016}.
-% to overcome the limitation of IST which was commonly used for regions of high water occupancy and its implementation in \software{cpptraj} was thoroughly described in \cite{Ramsey2016}.
-Here, we only provide a short overview of the theory behind GIST.
-For more detailed information, we recommend one of the more recent publications on developments in GIST \cite{Kraml2020,Chen2021}.
-
-\subsection{Solvation Entropy}
-
-%The solvation entropy of a system is totality of the contribution of the  solute-water entropy and the water-water entropy. -- this is inconsistent with IST theory (fwa)
-
-%\begin{equation}
-%	\Delta S_\textit{solv} = \Delta S_\textit{sw} + \Delta S_\textit{ww}
-%\end{equation}
-%Here the solvation entropy is approximated from the contribution of solute-water interaction. 
-%\textbf{STILLNEEDS REVISON + ADD CITATIONS} The quantity $k_\textit{b}$represents the Boltzman constant, 
-%$\rho^\textit{o}$ is the bulk number density, $g_\textit{sw}\left(\textbf{r},\omega \right)$ is the solute-water 
-%pair correlation function in the solute's frame of reference, \textbf{r} is the coordinates of the oxygen atom of the water molecule,  
-%$\omega$ represents the Euler angles, and $\frac{1}{8\pi^2}$ is the volume of the orientational space, which acts as a normalization constant.
-%In bulk and in regions where the orientational 
-%entropy is uniform, the quantity $g_\textit{sw}\left(\textbf{r},\omega \right)$ is unity which leads to the first-order solvation entropy of 
-%bulk to be zero. Also, this quantity reaches unity as the distance from the solute molecule increases which leads for the calculation of 
-%the solvation entropy to be an approximation of the integral around the solute atom.
-
-IST expresses the free energy of solvation in terms of the solvent density in a coordinate system defined by the solute.
-The solute is kept fixed in space, following the standard solvation process established by Ben-Naim \cite{ben-naim-book}.
-
-For the entropy, this leads to an infinite expansion of correlations of increasing order.
-In a slightly simplified view, the first order is the solvent density $g_\textit{sw}\left(\textbf{r},\omega \right)$ at each position $\textbf{r}$ and orientation $\omega$.
-The first-order entropy omits all solvent-solvent (and higher) correlations, while the solute-solvent correlation is taken into account because the coordinates are relative to the solute.
-Thus, it is called the solute-water entropy $S_{sw}$.
-%The second order is the pair density, i.e., the probability of finding two solvent molecules in positions $\textbf{r}$ and $\textbf{r}^\prime$.
-
-The solvent density is expressed relative to the bulk density $\rho^\textit{o}$. %$g_\textit{sw}\left(\textbf{r},\omega \right)$ is the solute-water 
-%pair correlation function in the solute's frame of reference, \textbf{r} is the coordinates of the oxygen atom of the water molecule,  
-%$\omega$ represents the Euler angles, and 
-The entropy integral is also normalized by $8\pi^2$, which is the volume of the orientational space.
-In bulk or at a high distance to the solute, the quantity $g_\textit{sw}\left(\textbf{r},\omega \right)$ is unity which leads to the first-order entropy of 
-bulk to be zero. 
-Therefore, no reference entropy needs to be computed unless there is a numeric bias, e.g., when the MD frames are not statistically independent due to insufficient spacing.
-%Also, this quantity reaches unity as the distance from the solute molecule increases which leads for the calculation of 
-%the solvation entropy to be an approximation of the integral around the solute atom. 
-\begin{equation}
-	\Delta S_\textit{solv}
-	\approx \Delta S_\textit{sw}
-	\newcommand{\gro}{g_{sw}(\textbf{r},\omega)}
-	\equiv -R \frac{\rho^\textit{0}}{8\pi^\textit{2}} \int \gro \ln\gro d\textbf{r}d\omega
-\end{equation}
-\todo{We are inconsistent regarding R/kB; vah: Fixed. is it - tho? Papers seem to disagre }
-
-The solvation entropy is further broken down in terms of the contribution from translational and orientational terms.
-\begin{equation}
-	\Delta S_\textit{solv} = \Delta S_\textit{trans} + \Delta S_\textit{orient}
-\end{equation}
-
-\subsection{Entropy Calculations in Cpptraj}
-In \software{cpptraj} the calculation of  solvation entropy is handled by two methods.
-
-The first method assumes that the distribution of the orientation $\omega$ is constant within a voxel $k$.
-This allows for the splitting of the contributions of the translational and the orientational entropy. 
-A first nearest neighbor (NN) approach is used to evaluate the density.
-%The translational density $g_{NN}(\textbf{r})$ is expressed in reference to a homogeneous bulk solvent at the reference density, while the orientational density $g_{NN}(\omega)$ is expressed to a random placement of the number of solvents found in this voxel.
-Both the translational ($g_{NN}(\textbf{r})$) and orientational ($g_{NN}(\omega)$) density are computed relative to a bulk solvent at the reference density $\rho^0$.
-
-The average entropy contributions per solvent molecule in voxel $k$ are calculated as
-
-\begin{equation}
-	S_{k}^\textit{trans} = -R \left( \gamma + \frac{1}{N_\textit{k}} \sum _{i=1}^{N_k} \ln g_{NN, \textit{i}}(\textbf{r}) \right)
-\end{equation}
-and
-\begin{equation}
-	S_{k}^\textit{orient} = -R \left( \gamma + \frac{1}{N_k} \sum _{i=1}^{N_k} \ln g_{NN, i}(\omega) \right)
-	,
-\end{equation}
-where $N_\textit{k}$ is the number of water molecules found in voxel k, $\gamma$ is Euler's constant accounting for the bias of the naive entropy estimator, and $g_\textit{NN}$ is the nearest neighbor estimate.
-
-The second method directly calculates the solvation entropy by evaluating the six-dimensional integral (3 for position and 3 for orientation) 
-using the nearest-neighbor approach. 
-% Unlike the first method, the nearest neighbor estimate for the sixth-order entropy considers all the water found within the same and neighboring voxels.
--- misleading because this is also done for S_trans (fwa)
-
-\begin{equation}
-g_\textit{vox} \left( \textbf{r}, \omega \right) \approx g_\textit{vox}(\textbf{r}) g_\textit{vox}(\omega)
-\end{equation}
-
-\begin{equation}
-	S_\textit{k} = -R \left( \gamma + \frac{1}{N_\textit{k}} \sum _{i=1}^{N_k} \ln g_{NN, \textit{i}}(\textbf{r},\omega) \right)
-\end{equation}
-
-\subsection{Solvation Energy}
-The solvation energy is readily calculated by summing up the contributions of the water-water interaction and the water-solute interaction from the force field.
-\begin{equation}
-	\Delta E_\textit{solv} = \Delta E_\textit{sw} + \Delta E_\textit{ww}
-\end{equation}
-
-The solute-solvent energy $E_{sw}$ can be expressed in terms of the solvent density and the potential $U_{sw}(\textbf{r},\omega)$ induced by the solute.
-In practice, it is evaluated as the expectation value $\langle\cdot\rangle$ of the pairwise force field energy $U_{ij}$ between all solvent molecules in voxel $k$ and all solute atoms.
-
-\begin{equation}
-\begin{aligned}
-\label{eq-esw}
-	\Delta E_{sw}(\textbf{r}) \equiv& \frac{1}{8\pi^2} \int g_\textit{sw}\left(\omega|\textbf{r}\right) U_\textit{sw}\left(\textbf{r}, \omega\right) d\omega \\
-	\Delta E_{sw,k}=& \left\langle \sum_i^{\textit{solvent},k} \; \sum_j^\textit{solute} U_{ij}\right\rangle
-\end{aligned}
-\end{equation}
-
-To localize $E_{sw}$ on the three-dimensional grid, every energy term is assigned to the voxel $k$ holding the solvent, and the expectation value in Equation~\ref{eq-esw} is computed per voxel.
-Similar to the entropy integrals, the solute-water solvation energy integral also decays with increasing distance from the solute.
-Hence, the solvation energy can be approximated by local spatial integrals.
-
-The solvent-solvent energy $E_{ww}$ is computed in a similar manner.
-Again, it can be expressed in terms of density functions, but is practically computed as a sum over solvent-solvent interactions where one solvent is in voxel $k$.
-In contrast to $E_{sw}$, $E_{ww}$ does not tend to zero in bulk.
-Therefore, a reference energy corresponding to the average energy of a water molecule in bulk needs to be subtracted.
-
-%However, there is a complication arising from this approximation.
-Water molecules outside the region of interest will also be accounted for in the pairwise water term.
-For example, given two different regions $R_a$ and $R_b$, will not sum up to the total energy of the two regions due to double counting.
-
-\begin{equation}
-\begin{aligned}
-	\Delta E_{ww}(\textbf{r}) \equiv & \left(\frac{1}{8\pi^2} \right)^2 \rho^o \int g_\textit{sw}(\omega|\textbf{r}) \\
-	& \times \left[g_\textit{sw}(\textbf{r}^\prime , \omega^\prime) - g^o_\textit{ww} (\textbf{r}, \omega, \textbf{r}^\prime, \omega^\prime )\right] \\
-	& \times U_\textit{ww}(\textbf{r}, \omega, \textbf{r}^\prime, \omega^\prime)d\omega d\textbf{r}^\prime d\omega^\prime \\
-	\Delta E_{ww,k} =& \frac{1}{N_k} \left\langle \sum_i^{\text{solvent},k} \; \sum_{j \neq i}^\text{solvent} U_{ij}\right\rangle - \left\langle E_{ww}^\text{bulk}\right\rangle
-\end{aligned}
-\end{equation}
-
-%AFAIK, this was never done, since the pair interaction matrix Eij would have to be written out.
-%There are two ways in which this double counting was accounted for in the original paper.
-First, using the water-displacement method.
-This method is applicable when a determining the energetic cost of kicking out a water molecule.
-%\begin{equation}
-%	\Delta E^R_\textit{ww} = n^\textit{R} E^\textit{bulk}_\textit{ww} - E^\textit{R,corr}_\textit{ww}
-%\end{equation}
-
-%The second method uses normalized-water properties to account the difference in water properties of two different regions.
-The normalized water property is defined as the water property per water molecule.
-%\begin{equation}
-%	\Delta E^\textit{R,norm}_\textit{ww} = E^\textit{R,norm}_\textit{ww} - 2E^\textit{bulk}_\textit{ww}
-%\end{equation}
-
-When computing the total $E_{ww}$ of the system, double-counting of interactions must be avoided.
-The total solvent-solvent energy of the system is as follows:
-
-\begin{equation}
-	\Delta E^R_\textit{ww} = n^\textit{R} E^\textit{bulk}_\textit{ww} - E^\textit{R,corr}_\textit{ww}
-\end{equation}
-
-However, when water is replaced from a small region $R$ of interest, such as a single water molecule, almost all interactions are with water molecules outside of this region.
-Therefore, there is no double counting, and the full solvent-solvent energy should be used.
-
-\begin{equation}
-	\Delta E^\textit{R,norm}_\textit{ww} = E^\textit{R,norm}_\textit{ww} - 2E^\textit{bulk}_\textit{ww}
-\end{equation}
-
-When the region of interest $R$ comprises more than one molecule, interactions within this region are double-counted while interactions to the outside are not.
-To take this into account, the solvent-solvent energy between each pair of voxels $E_{ww,kl}$ is needed.
-While this is supported by the standard GIST implementation, it is rarely done due to the large size of the $E_{ww,kl}$ matrix.
-
-We note that replacement of a small or medium solvent region would lead to reorganization of the surrounding solvent.
-However, this depends on the local environment and on \emph{what} the molecule would be replaced by.
-Therefore, the effect of reorganization can not be directly included in the GIST method, but must be judged for each case individually.
-This is not a problem when computing the total free energy of solvation $\dgsolv$, since no additional boundary is created.
-\todo{This is a bit unclear to me, could we elaborate on this a bit more? What are we asking of the reader here, how would one consider this effect? -vah}
-
-\subsection{PME implementation}
-
-In the original version of GIST, the energies are calculated based on the the minimum image convention.
-In PME-GIST, the electrostatics are calculated using the particle mesh Ewald (PME) method, which yields energies that are highly consistent with the treatment in the Amber MD engine \cite{Chen2021}.
-The electrostatic and Lennard-Jones parts $E_\text{elec}$ and $E_\text{lj}$ are computed separately.
-
-\begin{equation}
-	E_\text{total} = E_\text{elec} + E_\text{lj}
-\end{equation}
-
-In PME-GIST, the system is treated as periodic, and the electrostatic energy is split into a short-range term $E_\text{dir}$, which is calculated in direct space using a distance cutoff, and a long-range term $E_\text{rec}$, which is calculated in reciprocal space.
-Additionally, there is a correction term $E_\text{corr}$, which corrects for the self-interaction of each solvent molecule in the reciprocal term.
-
-\begin{equation}
-	E_\textit{elec} = E_\textit{dir} + E_\textit{rec} + E_\textit{corr}
-\end{equation}
-%\todo{How is E$_{dir}$ treated? Is there a cutoff as well (PME GIST paper seems to indicate so)? If so, we should probably note that. } - ok? -fwa
-The Lennard-Jones contribution is computed in the direct space, using a distance cutoff.
-The short-range contribution is the LJ energy within the set cutoff and the long-range correction term accounts for the contributions above this cutoff.
-
-\begin{equation}
-	E_\textit{lj} = E_\textit{lj,\ short} +  E_\textit{corr}
-\end{equation}
-\todo{Is $E_{corr}$ the same in eq 13 and 14? How is it calculated? -vah}
 
@@
 %\subsection{Running the GIST analysis}
-To run a gist analyis, the following is required:
+To run a GIST analyis, the following is required:
 \begin{itemize}
@@
 
+Cpptrajs' gist action has a large number of options to modify the GIST calculation or its' output.
+A full list is provided at the end of this tutorial or through cpptrajs manual.
+The major options are summarized in the following table:
+
+\begin{table}[h]
+	\caption{The major options available in cpptrajs' GIST implementation.}\label{tab:GIST_options}
+	\small
+	\begin{tabularx}{\columnwidth}{@{}llXl@{}}
+		\toprule
+		Command       & Options & Explanation & Default  \\
+		\midrule
+		gridcntr    & <y> <y> <z> & Coordinates of the grid center in \AA & 0 0 0  \\
+		griddim     & <nx> <ny> <nz> & Number of voxels along each axis defining the grid size & 40 40 40\\
+		gridspacn & <dval> &Grid spacing / voxel dimensions in  \AA & 0.5    \\
+		refdens         & <rdval> & Reference density of the bulk water in molecules/\AA$^3$ . & 0.0334    \\
+		\bottomrule
+	\end{tabularx}
+\end{table}
+
+Setting the reference density according to the water model used in your simulation is important to get correct results.
+The reference density can be calculated from a GIST calculation for a pure water simulation or estimated from bulk-like water sufficiently far away from a solvent.
+\todo{How much information on the choice of reference density should we include?}
+
 \paragraph{Postprocessing and Analyzing GIST results}
 Cpptraj outputs the quantities calculated as a "open data explorer" (.dx) file format which can be easily visualized in Pymol or VMD.
-Aside from the thermodynamic quantities, GIST also calculates the number densities of hydrogen and oxygen in each voxel.
-Here is a full list of quantities calculated by GIST \todo{This is incomplete!}:
+Aside from the thermodynamic quantities, GIST also calculates the number densities of hydrogen and oxygen in each voxel. Here is a list of the most important quantities calculated by GIST:
 
@@
 For water, gO and gH are produced.}
-	%\item{[gH] Number density of hydrogen centers found in the voxel in units of the reference bulk density.}
 	\item{[Esw] Mean solute-water interaction energy density.}
@@
 	\item{[dTSorient] First order orientational entropy density .}
+	\item{[dTSsix] First order entropy density for the six-dimensional combined orientation and translation space.}
 	\item{[neighbor] Mean number of waters neighboring the water molecules found in this voxel multiplied by the voxel number density.}
 	\item{[dipole] Magnitude of mean dipole moment (polarization).}
-	\item{[order] (if \inlinecode{doorder} was specified) Average Tetrahedral Order Parameter.}
 	\item{[dipolex] x-component of the mean water dipole moment density}
@@
 	\item{[dipolez] z-component of the mean water dipole moment density}
-	\item{[Eij] (if \inlinecode{doeij} was specified) Water-water interaction matrix.}
 \end{itemize}
@@
 Several dx files are written by default, such as the energies, entropies, number density of oxygen, and number density of hydrogen.
-All quantities mentioned above are written to the output file (output.dat) and can be used to generate DX files for other quantities.
+All calculated quantities are written to the output file (output.dat) and can be used to generate DX files for other quantities.
+A full list of all calculated properties found in the output in addition to the ones mentioned above is provided at the end of the tutorial.
 
@@
 Here, we aim to guide the reader through a GIST analysis of the streptavidin-biotin complex.
-In the process, the reader will learn how to set up and run a GIST calculation.
 During the tutorial, we will
 \begin{enumerate}
+	\item set up and run GIST calculations for a small molecule as well as a protein binding pocket.
+	\item compute the free energy of solvation for a small molecule using GIST.
+	\item visualize local contributions to $\dgsolv$ in the streptavidin binding pocket as well as around biotin.
 	\item compute the binding free energy of the complex from the individual energy contributions combined with the contribution of hydration computed using GIST.
-	\item visualize local contributions to $\dgsolv$ in the streptavidin binding pocket as well as around biotin.
+	\item explore advantages and disadvantages of the GIST method when used to calculate  $\dgsolv$
 \end{enumerate}
+The tutorial is divided into a beginners part and an advanced part.
+In the beginners part, we will take a look at how to run GIST calculations for small molecules and interpret the output for a small molecule.
+In the advanced part, we will apply this knowledge to a protein binding pocket, visualize $\dgsolv$ hotspots and estimate a free energy of binding for a small molecule into this pocket using an end-state approach.
+In addition, while the beginners part focuses only on GIST calculation run, we will discuss all steps along the way in detail in the advanced section.
 
@@
 We will use the 1STP crystal structure to start our calculations.
+For the beginners section, we will use the biotin ligand as extracted from the crystal structure.
+For the second part, we will use the whole crystal structure.
 For the sake of simplicity, you can download prepared and solvated Amber topologies and structures based on 1STP by running
@@
 This will download the tutorial files (including both the manuscript and the examples in the \software{code} folder).
-\subsection{System Preparation and Equilibration}
+\
+\subsection{Running GIST for Biotin}
+
+\subsection{Applying GIST for the Streptavidin-Biotin complex}
+\subsubsection{System Preparation and Equilibration}
 If you want to skip this section, run 
@@
 
-\subsection{Running MD for GIST}
+\subsubsection{Running MD for GIST}
 
@@
 
-\begin{figure}
-	\centering
-	\includegraphics[width=1.0\linewidth]{streptavidin_bp.png}
-	\caption{Volume of strepatividin defined around 3.5 A of biotin. }\label{streptavidin_sasa}
-\end{figure}
+%\begin{figure}
+%	\centering
+%	\includegraphics[width=1.0\linewidth]{streptavidin_bp.png}
+%	\caption{Volume of strepatividin defined around 3.5 A of biotin. }\label{streptavidin_sasa}
+%\end{figure}
+
+\section{Theory}
+\label{sec:theory}
+GIST is an implementation of inhomogeneous fluid solvation theory (IST) \cite{Lazaridis1998} that discretizes the free energy of solvation $\dgsolv$ on a three-dimensional grid. 
+It was first devised by Nguyen et al.
+\cite{Nguyen2012} and its implementation in \software{cpptraj} was thoroughly described by Ramsey et al.
+\cite{Ramsey2016}.
+% to overcome the limitation of IST which was commonly used for regions of high water occupancy and its implementation in \software{cpptraj} was thoroughly described in \cite{Ramsey2016}.
+Here, we only provide a short overview of the theory behind GIST.
+For more detailed information, we recommend one of the more recent publications on developments in GIST \cite{Kraml2020,Chen2021}.
+
+\subsection{Solvation Entropy}
+
+%The solvation entropy of a system is totality of the contribution of the  solute-water entropy and the water-water entropy. -- this is inconsistent with IST theory (fwa)
+
+%\begin{equation}
+%	\Delta S_\textit{solv} = \Delta S_\textit{sw} + \Delta S_\textit{ww}
+%\end{equation}
+%Here the solvation entropy is approximated from the contribution of solute-water interaction. 
+%\textbf{STILLNEEDS REVISON + ADD CITATIONS} The quantity $k_\textit{b}$represents the Boltzman constant, 
+%$\rho^\textit{o}$ is the bulk number density, $g_\textit{sw}\left(\textbf{r},\omega \right)$ is the solute-water 
+%pair correlation function in the solute's frame of reference, \textbf{r} is the coordinates of the oxygen atom of the water molecule,  
+%$\omega$ represents the Euler angles, and $\frac{1}{8\pi^2}$ is the volume of the orientational space, which acts as a normalization constant.
+%In bulk and in regions where the orientational 
+%entropy is uniform, the quantity $g_\textit{sw}\left(\textbf{r},\omega \right)$ is unity which leads to the first-order solvation entropy of 
+%bulk to be zero. Also, this quantity reaches unity as the distance from the solute molecule increases which leads for the calculation of 
+%the solvation entropy to be an approximation of the integral around the solute atom.
+
+IST expresses the free energy of solvation in terms of the solvent density in a coordinate system defined by the solute.
+The solute is kept fixed in space, following the standard solvation process established by Ben-Naim \cite{ben-naim-book}.
+
+For the entropy, this leads to an infinite expansion of correlations of increasing order.
+In a slightly simplified view, the first order is the solvent density $g_\textit{sw}\left(\textbf{r},\omega \right)$ at each position $\textbf{r}$ and orientation $\omega$.
+The first-order entropy omits all solvent-solvent (and higher) correlations, while the solute-solvent correlation is taken into account because the coordinates are relative to the solute.
+Thus, it is called the solute-water entropy $S_{sw}$.
+%The second order is the pair density, i.e., the probability of finding two solvent molecules in positions $\textbf{r}$ and $\textbf{r}^\prime$.
+
+The solvent density is expressed relative to the bulk density $\rho^\textit{o}$. %$g_\textit{sw}\left(\textbf{r},\omega \right)$ is the solute-water 
+%pair correlation function in the solute's frame of reference, \textbf{r} is the coordinates of the oxygen atom of the water molecule,  
+%$\omega$ represents the Euler angles, and 
+The entropy integral is also normalized by $8\pi^2$, which is the volume of the orientational space.
+In bulk or at a high distance to the solute, the quantity $g_\textit{sw}\left(\textbf{r},\omega \right)$ is unity which leads to the first-order entropy of 
+bulk to be zero. 
+Therefore, no reference entropy needs to be computed unless there is a numeric bias, e.g., when the MD frames are not statistically independent due to insufficient spacing.
+%Also, this quantity reaches unity as the distance from the solute molecule increases which leads for the calculation of 
+%the solvation entropy to be an approximation of the integral around the solute atom. 
+\begin{equation}
+	\Delta S_\textit{solv}
+	\approx \Delta S_\textit{sw}
+	\newcommand{\gro}{g_{sw}(\textbf{r},\omega)}
+	\equiv -R \frac{\rho^\textit{0}}{8\pi^\textit{2}} \int \gro \ln\gro d\textbf{r}d\omega
+\end{equation}
+\todo{We are inconsistent regarding R/kB; vah: Fixed. is it - tho? Papers seem to disagre }
+
+The solvation entropy is further broken down in terms of the contribution from translational and orientational terms.
+\begin{equation}
+	\Delta S_\textit{solv} = \Delta S_\textit{trans} + \Delta S_\textit{orient}
+\end{equation}
+
+\subsection{Entropy Calculations in Cpptraj}
+In \software{cpptraj} the calculation of  solvation entropy is handled by two methods.
+
+The first method assumes that the distribution of the orientation $\omega$ is constant within a voxel $k$.
+This allows for the splitting of the contributions of the translational and the orientational entropy. 
+A first nearest neighbor (NN) approach is used to evaluate the density.
+%The translational density $g_{NN}(\textbf{r})$ is expressed in reference to a homogeneous bulk solvent at the reference density, while the orientational density $g_{NN}(\omega)$ is expressed to a random placement of the number of solvents found in this voxel.
+Both the translational ($g_{NN}(\textbf{r})$) and orientational ($g_{NN}(\omega)$) density are computed relative to a bulk solvent at the reference density $\rho^0$.
+
+The average entropy contributions per solvent molecule in voxel $k$ are calculated as
+
+\begin{equation}
+	S_{k}^\textit{trans} = -R \left( \gamma + \frac{1}{N_\textit{k}} \sum _{i=1}^{N_k} \ln g_{NN, \textit{i}}(\textbf{r}) \right)
+\end{equation}
+and
+\begin{equation}
+	S_{k}^\textit{orient} = -R \left( \gamma + \frac{1}{N_k} \sum _{i=1}^{N_k} \ln g_{NN, i}(\omega) \right)
+	,
+\end{equation}
+where $N_\textit{k}$ is the number of water molecules found in voxel k, $\gamma$ is Euler's constant accounting for the bias of the naive entropy estimator, and $g_\textit{NN}$ is the nearest neighbor estimate.
+
+The second method directly calculates the solvation entropy by evaluating the six-dimensional integral (3 for position and 3 for orientation) 
+using the nearest-neighbor approach. 
+% Unlike the first method, the nearest neighbor estimate for the sixth-order entropy considers all the water found within the same and neighboring voxels.
+-- misleading because this is also done for S_trans (fwa)
+
+\begin{equation}
+	g_\textit{vox} \left( \textbf{r}, \omega \right) \approx g_\textit{vox}(\textbf{r}) g_\textit{vox}(\omega)
+\end{equation}
+
+\begin{equation}
+	S_\textit{k} = -R \left( \gamma + \frac{1}{N_\textit{k}} \sum _{i=1}^{N_k} \ln g_{NN, \textit{i}}(\textbf{r},\omega) \right)
+\end{equation}
+
+\subsection{Solvation Energy}
+The solvation energy is readily calculated by summing up the contributions of the water-water interaction and the water-solute interaction from the force field.
+\begin{equation}
+	\Delta E_\textit{solv} = \Delta E_\textit{sw} + \Delta E_\textit{ww}
+\end{equation}
+
+The solute-solvent energy $E_{sw}$ can be expressed in terms of the solvent density and the potential $U_{sw}(\textbf{r},\omega)$ induced by the solute.
+In practice, it is evaluated as the expectation value $\langle\cdot\rangle$ of the pairwise force field energy $U_{ij}$ between all solvent molecules in voxel $k$ and all solute atoms.
+
+\begin{equation}
+	\begin{aligned}
+		\label{eq-esw}
+		\Delta E_{sw}(\textbf{r}) \equiv& \frac{1}{8\pi^2} \int g_\textit{sw}\left(\omega|\textbf{r}\right) U_\textit{sw}\left(\textbf{r}, \omega\right) d\omega \\
+		\Delta E_{sw,k}=& \left\langle \sum_i^{\textit{solvent},k} \; \sum_j^\textit{solute} U_{ij}\right\rangle
+	\end{aligned}
+\end{equation}
+
+To localize $E_{sw}$ on the three-dimensional grid, every energy term is assigned to the voxel $k$ holding the solvent, and the expectation value in Equation~\ref{eq-esw} is computed per voxel.
+Similar to the entropy integrals, the solute-water solvation energy integral also decays with increasing distance from the solute.
+Hence, the solvation energy can be approximated by local spatial integrals.
+
+The solvent-solvent energy $E_{ww}$ is computed in a similar manner.
+Again, it can be expressed in terms of density functions, but is practically computed as a sum over solvent-solvent interactions where one solvent is in voxel $k$.
+In contrast to $E_{sw}$, $E_{ww}$ does not tend to zero in bulk.
+Therefore, a reference energy corresponding to the average energy of a water molecule in bulk needs to be subtracted.
+
+%However, there is a complication arising from this approximation.
+Water molecules outside the region of interest will also be accounted for in the pairwise water term.
+For example, given two different regions $R_a$ and $R_b$, will not sum up to the total energy of the two regions due to double counting.
+
+\begin{equation}
+	\begin{aligned}
+		\Delta E_{ww}(\textbf{r}) \equiv & \left(\frac{1}{8\pi^2} \right)^2 \rho^o \int g_\textit{sw}(\omega|\textbf{r}) \\
+		& \times \left[g_\textit{sw}(\textbf{r}^\prime , \omega^\prime) - g^o_\textit{ww} (\textbf{r}, \omega, \textbf{r}^\prime, \omega^\prime )\right] \\
+		& \times U_\textit{ww}(\textbf{r}, \omega, \textbf{r}^\prime, \omega^\prime)d\omega d\textbf{r}^\prime d\omega^\prime \\
+		\Delta E_{ww,k} =& \frac{1}{N_k} \left\langle \sum_i^{\text{solvent},k} \; \sum_{j \neq i}^\text{solvent} U_{ij}\right\rangle - \left\langle E_{ww}^\text{bulk}\right\rangle
+	\end{aligned}
+\end{equation}
+
+%AFAIK, this was never done, since the pair interaction matrix Eij would have to be written out.
+%There are two ways in which this double counting was accounted for in the original paper.
+First, using the water-displacement method.
+This method is applicable when a determining the energetic cost of kicking out a water molecule.
+%\begin{equation}
+%	\Delta E^R_\textit{ww} = n^\textit{R} E^\textit{bulk}_\textit{ww} - E^\textit{R,corr}_\textit{ww}
+%\end{equation}
+
+%The second method uses normalized-water properties to account the difference in water properties of two different regions.
+The normalized water property is defined as the water property per water molecule.
+%\begin{equation}
+%	\Delta E^\textit{R,norm}_\textit{ww} = E^\textit{R,norm}_\textit{ww} - 2E^\textit{bulk}_\textit{ww}
+%\end{equation}
+
+When computing the total $E_{ww}$ of the system, double-counting of interactions must be avoided.
+The total solvent-solvent energy of the system is as follows:
+
+\begin{equation}
+	\Delta E^R_\textit{ww} = n^\textit{R} E^\textit{bulk}_\textit{ww} - E^\textit{R,corr}_\textit{ww}
+\end{equation}
+
+However, when water is replaced from a small region $R$ of interest, such as a single water molecule, almost all interactions are with water molecules outside of this region.
+Therefore, there is no double counting, and the full solvent-solvent energy should be used.
+
+\begin{equation}
+	\Delta E^\textit{R,norm}_\textit{ww} = E^\textit{R,norm}_\textit{ww} - 2E^\textit{bulk}_\textit{ww}
+\end{equation}
+
+When the region of interest $R$ comprises more than one molecule, interactions within this region are double-counted while interactions to the outside are not.
+To take this into account, the solvent-solvent energy between each pair of voxels $E_{ww,kl}$ is needed.
+While this is supported by the standard GIST implementation, it is rarely done due to the large size of the $E_{ww,kl}$ matrix.
+
+We note that replacement of a small or medium solvent region would lead to reorganization of the surrounding solvent.
+However, this depends on the local environment and on \emph{what} the molecule would be replaced by.
+Therefore, the effect of reorganization can not be directly included in the GIST method, but must be judged for each case individually.
+This is not a problem when computing the total free energy of solvation $\dgsolv$, since no additional boundary is created.
+\todo{This is a bit unclear to me, could we elaborate on this a bit more? What are we asking of the reader here, how would one consider this effect? -vah}
+
+\subsection{PME implementation}
+
+In the original version of GIST, the energies are calculated based on the the minimum image convention.
+In PME-GIST, the electrostatics are calculated using the particle mesh Ewald (PME) method, which yields energies that are highly consistent with the treatment in the Amber MD engine \cite{Chen2021}.
+The electrostatic and Lennard-Jones parts $E_\text{elec}$ and $E_\text{lj}$ are computed separately.
+
+\begin{equation}
+	E_\text{total} = E_\text{elec} + E_\text{lj}
+\end{equation}
+
+In PME-GIST, the system is treated as periodic, and the electrostatic energy is split into a short-range term $E_\text{dir}$, which is calculated in direct space using a distance cutoff, and a long-range term $E_\text{rec}$, which is calculated in reciprocal space.
+Additionally, there is a correction term $E_\text{corr}$, which corrects for the self-interaction of each solvent molecule in the reciprocal term.
+
+\begin{equation}
+	E_\textit{elec} = E_\textit{dir} + E_\textit{rec} + E_\textit{corr}
+\end{equation}
+%\todo{How is E$_{dir}$ treated? Is there a cutoff as well (PME GIST paper seems to indicate so)? If so, we should probably note that. } - ok? -fwa
+The Lennard-Jones contribution is computed in the direct space, using a distance cutoff.
+The short-range contribution is the LJ energy within the set cutoff and the long-range correction term accounts for the contributions above this cutoff.
+
+\begin{equation}
+	E_\textit{lj} = E_\textit{lj,\ short} +  E_\textit{corr}
+\end{equation}
+\todo{Is $E_{corr}$ the same in eq 13 and 14? How is it calculated? -vah}
 

### 2022-10-14 manuscript/manuscript.tex
More restructuring. Add first reworks to intoduce a split into a beginners part (only biotin, how to run GIST) and an advanced part (biotin-streptavidin; pitfalls, tipps and discussion of settings)
Change SASA method. Pymol instead of gistpp

@@
 
-The solvent accessible surface area (SASA) represents  the voxels that are close to the protein surface and occupied by oxygen water.
-It represents the surface where water can go, and potentially surfaces that a ligand atom can occupy.
-
-The surface is generated from the oxygen density (gO.dx) file using \software{gistpp}. 
-A voxel is deemed to be  part of the surface if it is occupied by water (gO > 0.3) and has at least one neighboring empty voxel (gO < 0.1).
-
-\begin{lstlisting}[style=bash]
-	gistpp -op sasa -i gO.dx -o sasa.dx
+The solvent accessible surface (SASA) represents the interface between regions that are occupied by the solute, and those which can be accessed by the solvent.
+GIST provides a very natural way of creating a SASA using an isosurface of the oxygen density gO.
+It can be visualized in PyMOL as follows:
+
+\begin{lstlisting}[style=pymol]
+load ../gist.pdb
+load ../gist-gO.dx
+isomesh sas, gist-gO, 0.1
 \end{lstlisting}
@@
 	\centering
-	\includegraphics[width=1.0\linewidth]{streptavidin_sasa_close.png}
+	\includegraphics[width=1.0\linewidth]{figures/sasa-simple.png}
 	\caption{SASA for streptavidin focused on the binding pocket}\label{streptavidin_sasa}
@@
 
-In this example, the region defined in the binding pocket includes all voxels that are within 3.5 A of the heavy atoms of biotin.
+In this example, the region defined in the binding pocket includes all voxels that are within \SI{3.5}{\angstrom} of the heavy atoms of biotin.
 
@@
 	\centering
-	\includegraphics[width=1.0\linewidth]{streptavidin_bp.png}
-	\caption{Volume of strepatividin defined around 3.5 A of biotin. }\label{streptavidin_sasa}
+	\includegraphics[width=1.0\linewidth]{streptavidin_bp_close.png}
+	\caption{Volume of streptavidin defined around \SI{3.5}{\angstrom} of biotin. }\label{streptavidin_volume}
 \end{figure}

### 2022-10-14 manuscript/manuscript.tex

--- 2022-10-05 manuscript/manuscript.tex
@@
 \begin{equation}
-g_\textit{vox} ( \textbf{r}, \omega ) \approx g(\textbf{r}) g(\omega)
+g_\textit{vox} \left( \textbf{r}, \omega \right) \approx g_\textit{vox}(\textbf{r}) g_\textit{vox}(\omega)
 \end{equation}
@@
 \section{Recent developments}
-In recent years, several updates to the original GIST implementation have been published.
-The basic functionality of the program, however, has not been changed.
-In this chapter, we will shortly present each of those updates.
-\subsection{GPU implementation}
-In Reference~\cite{Kraml2020}, a GPU implementation of the energy calculation was implemented, which typically increases the speed by 1--2 orders of magnitude.
-
-If \software{cpptraj} is compiled with GPU support, the energy calculation uses the GPU automatically, without any additional input.
-However, PME-GIST is not supported on the GPU.
-
-\subsection{PME implementation}
-In Reference~\cite{Chen2021}, a PME implementation of the GIST energy was presented.
-For typical systems, this implementation is slightly slower than GPU-GIST, but much faster than the original CPU code.
-Furthermore, PME-GIST offers the best agreement between GIST energies and those observed in an MD simulation.
-
-To run a PME-GIST calculation, the \inlinecode{pme} flag to the \inlinecode{gist} action in \software{cpptraj} can be used.
-The output will contain two additional columns: \inlinecode{PME\_dens} and \inlinecode{PME\_norm}.
-They contain the potential of solute and solvent evaluated at the solvent positions and divided by two.
-Furthermore, the \inlinecode{Eww} and \inlinecode{Esw} columns are also computed using PME, and are reported as usual.
-For typical use cases, it is recommended to use \inlinecode{Eww} and \inlinecode{Esw} over the \inlinecode{PME} column.
-
 \subsection{GIST with non-water solvents}
-In References~\cite{Kraml2020,Kamenik2020-gist-macrocycles,Waibl2022-gist-solvents}, an extension of GIST towards solvents other than water is described.
-
-To run a GIST calculation in a solvent other than water, use the \inlinecode{solute} flag together with a \software{cpptraj} selection mask.
-Everything that is not in the \inlinecode{solute} mask is treated as solvent.
-If there are multiple solvent species, use the \inlinecode{solventmols} option as described in Section~\ref{sec-salt-water}.
-The code will automatically choose three atoms which are used to define the solvent orientation, and print this selection to the output.
-
-Sometimes, \software{cpptraj} does not automatically choose the optimal atoms.
-In methanol, for instance, one can either incorporate the C--H bonds or the O--H bond in the selection.
-The latter is probably more relevant since it can incorporate hydrogen bonding effects.
-Assuming that the alcohol hydrogen is called H1, this could be specified as follows:
-
-\begin{lstlisting}[style=cpptraj]
-gist gridcntr <x> <y> <z> \
-griddim <Nx> <Ny> <Nz> gridspacn <val> \
-solute ^1 rigidatoms O1 H1 C1
-\end{lstlisting}
-
-Note that the central atom of the \inlinecode{rigidatoms} selection goes first.
-
 \subsection{GIST with salt-water mixtures}
-\label{sec-salt-water}
 In \cite{Waibl2021-gist-salt}, an extension of GIST was presented that can use salt-water mixtures as a solvent.
@@
 	\centering
-	\includegraphics[width=1.0\linewidth]{streptavidin_bp.png}
-	\caption{Volume of strepatividin defined around 3.5 A of biotin. }\label{streptavidin_sasa}
+	\includegraphics[width=1.0\linewidth]{streptavidin_bp_close.png}
+	\caption{Volume of strepatividin defined around 3.5 A of biotin. }\label{streptavidin_bp_close}
 \end{figure}

### 2022-10-28 manuscript/manuscript.tex
Fix left-over merging conflict

@@
 %Solutes are constantly interacting with solvents in their environment, necessitating a good description of these interactions for accurate computational predictions. 
+The free energy of solvation is defined as the work needed to insert a solute into a solvent.
 Especially the interaction with water is of utmost importance, as many biological processes occur in a aqueous environment.
-The free energy of solvation thereby strongly influences real-world problems such as hydrophobic effects, the binding of ligands to a biomolecule, as well as the dynamics and folding of proteins.
+It thereby strongly influences real-world problems such as hydrophobic effects, the binding of ligands to a biomolecule, as well as the dynamics and folding of proteins.
 
@@
 In order to run a GIST analysis, the region where the solvent properties will be calculated must be first defined.
-This region is represented by a rectangular prism centered at (x, y, z) with $N_x \times N_y \times N_z$ voxels, where each voxel is cubic with a sidelength of \inlinecode{gridspacn}.
+This region is represented by a rectangular grid centered at (x, y, z) with $N_x \times N_y \times N_z$ voxels, where each voxel is cubic with a sidelength of \inlinecode{gridspacn}.
 While integrals over GIST quantities are independent of the voxel size, individual voxel values converge slower at high resolutions.\pdfcomment{The calculation actually can actually get faster at higher resolution, because the same grid is used in the nearest-neighbor search.}
-A default value of 0.5 A is used that provides reasonable discretization for many use-cases.
-
-In \software{ccptraj}, the rectangular prism is defined by parameters of the \inlinecode{gist} command.
-The center of the grid is defined by \inlinecode{gridcntr}, the dimensions of 
-the grid is defined by \inlinecode{griddim} (in voxel counts), and the spacing 
-is defined by \inlinecode{gridspacn} (in Å).
+The default value of \SI{0.5}{\angstrom} provides reasonable discretization for many use-cases.
+
+The grid is defined by parameters of the \inlinecode{gist} command in \software{cpptraj}.
+The grid center is defined by \inlinecode{gridcntr}, the voxel number in x, y, and z direction by \inlinecode{griddim}, and the spacing in \AA by \inlinecode{gridspacn}.
 
@@
 
-The region can be defined using a ligand in a specific binding pocket of the 
-protein, a specific residue of the protein, or the whole protein itself. A 
-helper script FindCentroid.py can be used to determine the centroid of the 
-region of interest as well as the dimensions of the rectangular prism that will 
-fully enclose the region. A pdb file is needed as an input and the rectangular 
-prism will be centered based on the centroid of the given structure.
+Often, the region of interest will be defined by the surrounding of a molecule, which might either be a ligand in the binding pocket or also the whole protein. 
+A helper script called \inlinecode{FindCentroid.py} for this purpose is supplied with this tutorial, and can be used as follows:
 
@@
 \paragraph{Run GIST using cpptraj}
-GIST is incorporated in cpptraj which requires the topology and the trajectory of the MD simulation to be analayzed.
+GIST is implemented in \software{cpptraj} which requires the topology and the trajectory of the MD simulation to be analyzed.
 The commands needed to run GIST in cpptraj can be written in an input file (e.g gist.in) for convenience.
-The basic comands, assuming a given grid center and a grid size of $30\times30\times30$ voxels with \SI{0.5}{\angstrom} spacing, are:
+The basic commands, assuming a given grid center and a grid size of $30\times30\times30$ voxels with \SI{0.5}{\angstrom} spacing, are:
 
@@
 
-The first line specifies the amber topology file that was used to run MD.
-The second line species the trajectory file and what frames to be analyzed.
-The third line tells cpptraj to run a GIST calculation using the region defined and print the output of the calculation in a file called gist.dat.
+The first line specifies the amber topology file that was used to run MD. The second line species the trajectory file.
+Note that the solute should not move significantly for a GIST calculation.
+This is typically achieved by using restraints on the solute coordinates.
+The third line tells \software{cpptraj} to run a GIST calculation using the region defined and print the output of the calculation in a file called \inlinecode{gist.dat}.
 The last instructs cpptraj to run the calculation.
-The syntax for running cpptraj given this input file is:
+The syntax for running \software{cpptraj} given this input file is:
 
@@
 
+The \inlinecode{gist} action has a large number of options to modify the GIST calculation or its' output.
+A full list is provided at the end of this tutorial or through cpptrajs manual.
+The major options are summarized in the following table:
+
+\begin{table}[h]
+	\caption{The major options available in cpptrajs' GIST implementation.}\label{tab:GIST_options}
+	\small
+	\begin{tabularx}{\columnwidth}{@{}llXl@{}}
+		\toprule
+		Command       & Options & Explanation & Default  \\
+		\midrule
+		gridcntr    & <y> <y> <z> & Coordinates of the grid center in \AA & 0 0 0  \\
+		griddim     & <nx> <ny> <nz> & Number of voxels along each axis defining the grid size & 40 40 40\\
+		gridspacn & <dval> &Grid spacing / voxel dimensions in  \AA & 0.5    \\
+		refdens         & <rdval> & Reference density of the bulk water in molecules/\AA$^3$ . & 0.0334    \\
+		\bottomrule
+	\end{tabularx}
+\end{table}
+
+Setting the reference density according to the water model used in your simulation is important to get correct results.
+Densities of typical water models can be found in the \software{cpptraj} manual.
+For other models, the density should be computed from a pure water simulation.
+
 \paragraph{Postprocessing and Analyzing GIST results}
 Cpptraj outputs the quantities calculated as a "open data explorer" (.dx) file format which can be easily visualized in Pymol or VMD.
-Aside from the thermodynamic quantities, GIST also calculates the number densities of hydrogen and oxygen in each voxel.
-Here is a full list of quantities calculated by GIST \todo{This is incomplete!}:
+Aside from the thermodynamic quantities, GIST also calculates the number densities of hydrogen and oxygen in each voxel. Here is a list of the most important quantities calculated by GIST:
 
@@
 	\item{[dTSorient] First order orientational entropy density .}
-	\item{[neighbor] Mean number of waters neighboring the water molecules found in this voxel multiplied by the voxel number density.}
+	\item{[dTSsix] First order entropy density for the six-dimensional combined orientation and translation space.}
 	\item{[dipole] Magnitude of mean dipole moment (polarization).}
-	\item{[order] (if \inlinecode{doorder} was specified) Average Tetrahedral Order Parameter.}
 	\item{[dipolex] x-component of the mean water dipole moment density}
@@
 This will download the tutorial files (including both the manuscript and the examples in the \software{code} folder).
-\subsection{System Preparation and Equilibration}
+\
+\subsection{Running GIST for Biotin}
+In this example, we will be running a simple GIST calculation for a small molecule to get familiar with the usage of the GIST command in \software{cpptraj}.
+\subsubsection{System preparation}
+First, extract biotin from the crystal structure and generate a PDB file containing only biotin.
+Prepare your structure by adding hydrogens, making sure to choose the correct protonation state.
+Parametrize biotin with the forcefield of your choice.
+Solvate the parametrized biotin, making sur to choose a water box of large enough size so that the water on the box edges behaves bulk-like (a wall distance of at least 15\AA is recommended).
+A good starting point (though a bit dated) for the parametrization and solvation of small molecules is AMBER's tutorial 4b (\url{http://ambermd.org/tutorials/basic/tutorial4b/}).
+\subsubsection{Equilibration}
+Equilibrate your solvated biotin to the conditions of your choice.
+Even when running a NVT simulation as the production run for the GIST analysis, it is advisible to include an NPT equilibration step to adjust the box size before hand.
+Allow enough equilibration time in the ensemble of your choice to allow the water to relax around your biotin structure.
+
+\subsubsection{MD simulation}
+\todo{Move simulation part from complex to here}
+
+\subsection{Applying GIST for the Streptavidin-Biotin complex}
+\subsubsection{System Preparation and Equilibration}
 If you want to skip this section, run 
@@
 
+\section{Theory}
+\label{sec:theory}
+GIST is an implementation of inhomogeneous fluid solvation theory (IST) \cite{Lazaridis1998} that discretizes the free energy of solvation $\dgsolv$ on a three-dimensional grid. 
+It was first devised by Nguyen et al.
+\cite{Nguyen2012} and its implementation in \software{cpptraj} was thoroughly described by Ramsey et al.
+\cite{Ramsey2016}.
+% to overcome the limitation of IST which was commonly used for regions of high water occupancy and its implementation in \software{cpptraj} was thoroughly described in \cite{Ramsey2016}.
+Here, we only provide a short overview of the theory behind GIST.
+For more detailed information, we recommend one of the more recent publications on developments in GIST \cite{Kraml2020,Chen2021}.
+
+\subsection{Solvation Entropy}
+
+%The solvation entropy of a system is totality of the contribution of the  solute-water entropy and the water-water entropy. -- this is inconsistent with IST theory (fwa)
+
+%\begin{equation}
+%	\Delta S_\textit{solv} = \Delta S_\textit{sw} + \Delta S_\textit{ww}
+%\end{equation}
+%Here the solvation entropy is approximated from the contribution of solute-water interaction. 
+%\textbf{STILLNEEDS REVISON + ADD CITATIONS} The quantity $k_\textit{b}$represents the Boltzman constant, 
+%$\rho^\textit{o}$ is the bulk number density, $g_\textit{sw}\left(\textbf{r},\omega \right)$ is the solute-water 
+%pair correlation function in the solute's frame of reference, \textbf{r} is the coordinates of the oxygen atom of the water molecule,  
+%$\omega$ represents the Euler angles, and $\frac{1}{8\pi^2}$ is the volume of the orientational space, which acts as a normalization constant.
+%In bulk and in regions where the orientational 
+%entropy is uniform, the quantity $g_\textit{sw}\left(\textbf{r},\omega \right)$ is unity which leads to the first-order solvation entropy of 
+%bulk to be zero. Also, this quantity reaches unity as the distance from the solute molecule increases which leads for the calculation of 
+%the solvation entropy to be an approximation of the integral around the solute atom.
+
+IST expresses the free energy of solvation in terms of the solvent density in a coordinate system defined by the solute.
+The solute is kept fixed in space, following the standard solvation process established by Ben-Naim \cite{ben-naim-book}.
+
+For the entropy, this leads to an infinite expansion of correlations of increasing order.
+In a slightly simplified view, the first order is the solvent density $g_\textit{sw}\left(\textbf{r},\omega \right)$ at each position $\textbf{r}$ and orientation $\omega$.
+The first-order entropy omits all solvent-solvent (and higher) correlations, while the solute-solvent correlation is taken into account because the coordinates are relative to the solute.
+Thus, it is called the solute-water entropy $S_{sw}$.
+%The second order is the pair density, i.e., the probability of finding two solvent molecules in positions $\textbf{r}$ and $\textbf{r}^\prime$.
+
+The solvent density is expressed relative to the bulk density $\rho^\textit{o}$. %$g_\textit{sw}\left(\textbf{r},\omega \right)$ is the solute-water 
+%pair correlation function in the solute's frame of reference, \textbf{r} is the coordinates of the oxygen atom of the water molecule,  
+%$\omega$ represents the Euler angles, and 
+The entropy integral is also normalized by $8\pi^2$, which is the volume of the orientational space.
+In bulk or at a high distance to the solute, the quantity $g_\textit{sw}\left(\textbf{r},\omega \right)$ is unity which leads to the first-order entropy of 
+bulk to be zero. 
+Therefore, no reference entropy needs to be computed unless there is a numeric bias, e.g., when the MD frames are not statistically independent due to insufficient spacing.
+%Also, this quantity reaches unity as the distance from the solute molecule increases which leads for the calculation of 
+%the solvation entropy to be an approximation of the integral around the solute atom. 
+\begin{equation}
+	\Delta S_\textit{solv}
+	\approx \Delta S_\textit{sw}
+	\newcommand{\gro}{g_{sw}(\textbf{r},\omega)}
+	\equiv -R \frac{\rho^\textit{0}}{8\pi^\textit{2}} \int \gro \ln\gro d\textbf{r}d\omega
+\end{equation}
+\todo{We are inconsistent regarding R/kB; vah: Fixed. is it - tho? Papers seem to disagre }
+
+The solvation entropy is further broken down in terms of the contribution from translational and orientational terms.
+\begin{equation}
+	\Delta S_\textit{solv} = \Delta S_\textit{trans} + \Delta S_\textit{orient}
+\end{equation}
+
+\subsection{Entropy Calculations in Cpptraj}
+In \software{cpptraj} the calculation of  solvation entropy is handled by two methods.
+
+The first method assumes that the distribution of the orientation $\omega$ is constant within a voxel $k$.
+This allows for the splitting of the contributions of the translational and the orientational entropy. 
+A first nearest neighbor (NN) approach is used to evaluate the density.
+%The translational density $g_{NN}(\textbf{r})$ is expressed in reference to a homogeneous bulk solvent at the reference density, while the orientational density $g_{NN}(\omega)$ is expressed to a random placement of the number of solvents found in this voxel.
+Both the translational ($g_{NN}(\textbf{r})$) and orientational ($g_{NN}(\omega)$) density are computed relative to a bulk solvent at the reference density $\rho^0$.
+
+The average entropy contributions per solvent molecule in voxel $k$ are calculated as
+
+\begin{equation}
+	S_{k}^\textit{trans} = -R \left( \gamma + \frac{1}{N_\textit{k}} \sum _{i=1}^{N_k} \ln g_{NN, \textit{i}}(\textbf{r}) \right)
+\end{equation}
+and
+\begin{equation}
+	S_{k}^\textit{orient} = -R \left( \gamma + \frac{1}{N_k} \sum _{i=1}^{N_k} \ln g_{NN, i}(\omega) \right)
+	,
+\end{equation}
+where $N_\textit{k}$ is the number of water molecules found in voxel k, $\gamma$ is Euler's constant accounting for the bias of the naive entropy estimator, and $g_\textit{NN}$ is the nearest neighbor estimate.
+
+The second method directly calculates the solvation entropy by evaluating the six-dimensional integral (3 for position and 3 for orientation) 
+using the nearest-neighbor approach. 
+% Unlike the first method, the nearest neighbor estimate for the sixth-order entropy considers all the water found within the same and neighboring voxels.
+-- misleading because this is also done for S_trans (fwa)
+
+\begin{equation}
+	g_\textit{vox} \left( \textbf{r}, \omega \right) \approx g_\textit{vox}(\textbf{r}) g_\textit{vox}(\omega)
+\end{equation}
+
+\begin{equation}
+	S_\textit{k} = -R \left( \gamma + \frac{1}{N_\textit{k}} \sum _{i=1}^{N_k} \ln g_{NN, \textit{i}}(\textbf{r},\omega) \right)
+\end{equation}
+
+\subsection{Solvation Energy}
+The solvation energy is readily calculated by summing up the contributions of the water-water interaction and the water-solute interaction from the force field.
+\begin{equation}
+	\Delta E_\textit{solv} = \Delta E_\textit{sw} + \Delta E_\textit{ww}
+\end{equation}
+
+The solute-solvent energy $E_{sw}$ can be expressed in terms of the solvent density and the potential $U_{sw}(\textbf{r},\omega)$ induced by the solute.
+In practice, it is evaluated as the expectation value $\langle\cdot\rangle$ of the pairwise force field energy $U_{ij}$ between all solvent molecules in voxel $k$ and all solute atoms.
+
+\begin{equation}
+	\begin{aligned}
+		\label{eq-esw}
+		\Delta E_{sw}(\textbf{r}) \equiv& \frac{1}{8\pi^2} \int g_\textit{sw}\left(\omega|\textbf{r}\right) U_\textit{sw}\left(\textbf{r}, \omega\right) d\omega \\
+		\Delta E_{sw,k}=& \left\langle \sum_i^{\textit{solvent},k} \; \sum_j^\textit{solute} U_{ij}\right\rangle
+	\end{aligned}
+\end{equation}
+
+To localize $E_{sw}$ on the three-dimensional grid, every energy term is assigned to the voxel $k$ holding the solvent, and the expectation value in Equation~\ref{eq-esw} is computed per voxel.
+Similar to the entropy integrals, the solute-water solvation energy integral also decays with increasing distance from the solute.
+Hence, the solvation energy can be approximated by local spatial integrals.
+
+The solvent-solvent energy $E_{ww}$ is computed in a similar manner.
+Again, it can be expressed in terms of density functions, but is practically computed as a sum over solvent-solvent interactions where one solvent is in voxel $k$.
+In contrast to $E_{sw}$, $E_{ww}$ does not tend to zero in bulk.
+Therefore, a reference energy corresponding to the average energy of a water molecule in bulk needs to be subtracted.
+
+%However, there is a complication arising from this approximation.
+Water molecules outside the region of interest will also be accounted for in the pairwise water term.
+For example, given two different regions $R_a$ and $R_b$, will not sum up to the total energy of the two regions due to double counting.
+
+\begin{equation}
+	\begin{aligned}
+		\Delta E_{ww}(\textbf{r}) \equiv & \left(\frac{1}{8\pi^2} \right)^2 \rho^o \int g_\textit{sw}(\omega|\textbf{r}) \\
+		& \times \left[g_\textit{sw}(\textbf{r}^\prime , \omega^\prime) - g^o_\textit{ww} (\textbf{r}, \omega, \textbf{r}^\prime, \omega^\prime )\right] \\
+		& \times U_\textit{ww}(\textbf{r}, \omega, \textbf{r}^\prime, \omega^\prime)d\omega d\textbf{r}^\prime d\omega^\prime \\
+		\Delta E_{ww,k} =& \frac{1}{N_k} \left\langle \sum_i^{\text{solvent},k} \; \sum_{j \neq i}^\text{solvent} U_{ij}\right\rangle - \left\langle E_{ww}^\text{bulk}\right\rangle
+	\end{aligned}
+\end{equation}
+
+%AFAIK, this was never done, since the pair interaction matrix Eij would have to be written out.
+%There are two ways in which this double counting was accounted for in the original paper.
+First, using the water-displacement method.
+This method is applicable when a determining the energetic cost of kicking out a water molecule.
+%\begin{equation}
+%	\Delta E^R_\textit{ww} = n^\textit{R} E^\textit{bulk}_\textit{ww} - E^\textit{R,corr}_\textit{ww}
+%\end{equation}
+
+%The second method uses normalized-water properties to account the difference in water properties of two different regions.
+The normalized water property is defined as the water property per water molecule.
+%\begin{equation}
+%	\Delta E^\textit{R,norm}_\textit{ww} = E^\textit{R,norm}_\textit{ww} - 2E^\textit{bulk}_\textit{ww}
+%\end{equation}
+
+When computing the total $E_{ww}$ of the system, double-counting of interactions must be avoided.
+The total solvent-solvent energy of the system is as follows:
+
+\begin{equation}
+	\Delta E^R_\textit{ww} = n^\textit{R} E^\textit{bulk}_\textit{ww} - E^\textit{R,corr}_\textit{ww}
+\end{equation}
+
+However, when water is replaced from a small region $R$ of interest, such as a single water molecule, almost all interactions are with water molecules outside of this region.
+Therefore, there is no double counting, and the full solvent-solvent energy should be used.
+
+\begin{equation}
+	\Delta E^\textit{R,norm}_\textit{ww} = E^\textit{R,norm}_\textit{ww} - 2E^\textit{bulk}_\textit{ww}
+\end{equation}
+
+When the region of interest $R$ comprises more than one molecule, interactions within this region are double-counted while interactions to the outside are not.
+To take this into account, the solvent-solvent energy between each pair of voxels $E_{ww,kl}$ is needed.
+While this is supported by the standard GIST implementation, it is rarely done due to the large size of the $E_{ww,kl}$ matrix.
+
+We note that replacement of a small or medium solvent region would lead to reorganization of the surrounding solvent.
+However, this depends on the local environment and on \emph{what} the molecule would be replaced by.
+Therefore, the effect of reorganization can not be directly included in the GIST method, but must be judged for each case individually.
+This is not a problem when computing the total free energy of solvation $\dgsolv$, since no additional boundary is created.
+\todo{This is a bit unclear to me, could we elaborate on this a bit more? What are we asking of the reader here, how would one consider this effect? -vah}
+
+\subsection{PME implementation}
+
+In the original version of GIST, the energies are calculated based on the the minimum image convention.
+In PME-GIST, the electrostatics are calculated using the particle mesh Ewald (PME) method, which yields energies that are highly consistent with the treatment in the Amber MD engine \cite{Chen2021}.
+The electrostatic and Lennard-Jones parts $E_\text{elec}$ and $E_\text{lj}$ are computed separately.
+
+\begin{equation}
+	E_\text{total} = E_\text{elec} + E_\text{lj}
+\end{equation}
+
+In PME-GIST, the system is treated as periodic, and the electrostatic energy is split into a short-range term $E_\text{dir}$, which is calculated in direct space using a distance cutoff, and a long-range term $E_\text{rec}$, which is calculated in reciprocal space.
+Additionally, there is a correction term $E_\text{corr}$, which corrects for the self-interaction of each solvent molecule in the reciprocal term.
+
+\begin{equation}
+	E_\textit{elec} = E_\textit{dir} + E_\textit{rec} + E_\textit{corr}
+\end{equation}
+%\todo{How is E$_{dir}$ treated? Is there a cutoff as well (PME GIST paper seems to indicate so)? If so, we should probably note that. } - ok? -fwa
+The Lennard-Jones contribution is computed in the direct space, using a distance cutoff.
+The short-range contribution is the LJ energy within the set cutoff and the long-range correction term accounts for the contributions above this cutoff.
+
+\begin{equation}
+	E_\textit{lj} = E_\textit{lj,\ short} +  E_\textit{corr}
+\end{equation}
+\todo{Is $E_{corr}$ the same in eq 13 and 14? How is it calculated? -vah}
+
 \section{Cheatsheet}
@@
 
-\begin{checklist}{The CPPTRAJ action}
+\begin{checklist}{The CPPTRAJ GIST action}
 

### 2022-10-31 manuscript/manuscript.tex

@@
 \DeclareSIUnit{\kcalPerMolASqr}{\kilo\calorie\per\mole\per\angstrom\squared}
+\DeclareSIUnit{\Molar}{M}
 
@@
 The closed-source Amber simulation engine is used in this tutorial to run the necessary MD simulations.
-If no Amber installation is available, the MD can be performed other simulation engines compatible with \software{cpptraj}.
-If using GROMACS, we propose to prepare the structures and topologies with AmberTools and converting them using \software{acpype.py}.
-
-\todo {If the simulation was ran using GROMACS, mdtraj can be used to convert GROMACS topology and trajectory in to an Amber file format.}
-\todo {vah: I would remove this mention of GROMACs and replace it with a more general comment. Or we could test if this actually works ( I doubt it does for ff19SB/OPC).}
-In that way, a full Amber topology is available for GIST.
-
+If no Amber installation is available, the MD can be performed other simulation engines, as long as the topology and trajectory can be produced in a format supported by \software{cpptraj}.
+If using GROMACS \cite{Abraham2015-gromacs,Pronk2013-gromacs}, we propose to prepare the topology with AmberTools and convert it using \software{acpype.py} \cite{Sousa_da_Silva2012-acpype}.
+%We should actually test that!
 Also note that compressed trajectory formats such as \inlinecode{xtc} might bias the entropy calculation due to the loss of precision for the atom positions.
@@
 We additionally provide a Jupyter notebook \cite{Kluyver2016-jupyter,Granger2021-jupyter} with the presented analyses. 
-We recommend JupyterLab to work with the notebook. Jupyter is available at \url{https://jupyter.org/}.
+%We recommend JupyterLab to work with the notebook. Jupyter is available at \url{https://jupyter.org/}.
 
@@
 The streptavidin-biotin complex is one of the strongest known non-covalent interactions involving a small molecule.
-Therefore, the combination of streptavidin and biotin is used routinely in biotechnology, and the interaction between them has been investigated in detail \todo{CITE}\cite{}.
-
-In nature, streptavidin occurs as a tetramer, and binds biotin with a binding constant of \todo{???}.
-Here, we will investigate the binding of biotin to a streptavidin monomer for simplicity.
-This interaction has been reported to have a binding constant of \todo{???}.
+Therefore, the combination of streptavidin and biotin is used routinely in biotechnology \cite{Dundas2013-streptavidin-review}, and the interaction between them has been investigated in detail \cite{McConnell2021-biotin}.
+
+\newcommand{\appr}{{\mathord{\sim}}}
+In nature, streptavidin occurs as a tetramer, and binds biotin with a binding constant of \SI{\sim e-14}{\per\Molar} \cite{Dundas2013-streptavidin-review}.
+The binding to the closed conformation of the tetramer has been estimated to have a free energy of \SI{26.6}{\kilo\calorie\per\mol} \cite{Bansal2018-biotin}.
+Here, we will investigate the binding of biotin to a closed streptavidin monomer for simplicity.
 

### 2022-11-02 code/scripts/get-dx-files.py
Add dx files with energy scaled by 2, and the script to make them

@@ -1,0 +1,40 @@
+#!/usr/bin/env python
+# -*- coding: utf-8 -*-
+
+import gisttools as gt
+
+
+def main():
+    import argparse
+
+    parser = argparse.ArgumentParser()
+    parser.add_argument("gistfile")
+    parser.add_argument("pdb")
+    parser.add_argument("out_base", help="Base name for output files, e.g. OUT_BASE-Eall_dens.dx")
+    parser.add_argument("--columns", nargs="+", default=["A_dens", "Eall_dens", "Eall2_dens", "A2_dens"])
+    args = parser.parse_args()
+    gistfile = gt.gist.load_gist_file(args.gistfile, struct=args.pdb)
+    gistfile.struct = remove_solvent(gistfile.struct)
+    assert 0.03 < gistfile.rho0 < 0.04, "Unusual reference density!"
+    gistfile.eww_ref = gistfile.detect_reference_value()
+    gistfile["Eww2_dens"] = gistfile["Eww_dens"] * 2
+    gistfile["Eall2_dens"] = gistfile["Eww2_dens"] + gistfile["Esw_dens"]
+    gistfile["A2_dens"] = gistfile["Eall2_dens"] - gistfile["dTSsix_dens"]
+    print("Settings")
+    print(f"Input GIST file: {args.gistfile}")
+    print(f"Input PDB file: {args.pdb}")
+    print(f"Detected rho0: {gistfile.rho0}")
+    print(f"Detected reference energy: {gistfile.eww_ref}")
+    for col in args.columns:
+        col_fname = col.replace("_", "-")
+        fname = f"{args.out_base}-{col_fname}.dx"
+        gistfile.save_dx(col, fname)
+
+
+def remove_solvent(traj):
+    not_water = traj.top.select("not resname WAT HOH NA CL")
+    return traj.atom_slice(not_water)
+
+
+if __name__ == "__main__":
+    main()

### 2022-11-02 manuscript/manuscript.tex

@@
 \section{Recent developments}
+In recent years, several updates to the original GIST implementation have been published.
+The basic functionality of the program, however, has not been changed.
+In this chapter, we will shortly present each of those updates.
+\subsection{GPU implementation}
+In Reference~\cite{Kraml2020}, a GPU implementation of the energy calculation was implemented, which typically increases the speed by 1--2 orders of magnitude.
+
+If \software{cpptraj} is compiled with GPU support, the energy calculation uses the GPU automatically, without any additional input.
+However, PME-GIST is not supported on the GPU.
+
+\subsection{PME implementation}
+In Reference~\cite{Chen2021}, a PME implementation of the GIST energy was presented.
+For typical systems, this implementation is slightly slower than GPU-GIST, but much faster than the original CPU code.
+Furthermore, PME-GIST offers the best agreement between GIST energies and those observed in an MD simulation.
+
+To run a PME-GIST calculation, the \inlinecode{pme} flag to the \inlinecode{gist} action in \software{cpptraj} can be used.
+The output will contain two additional columns: \inlinecode{PME\_dens} and \inlinecode{PME\_norm}.
+They contain the potential of solute and solvent evaluated at the solvent positions and divided by two.
+Furthermore, the \inlinecode{Eww} and \inlinecode{Esw} columns are also computed using PME, and are reported as usual.
+For typical use cases, it is recommended to use \inlinecode{Eww} and \inlinecode{Esw} over the \inlinecode{PME} column.
+
 \subsection{GIST with non-water solvents}
+In References~\cite{Kraml2020,Kamenik2020-gist-macrocycles,Waibl2022-gist-solvents}, an extension of GIST towards solvents other than water is described.
+
+To run a GIST calculation in a solvent other than water, use the \inlinecode{solute} flag together with a \software{cpptraj} selection mask.
+Everything that is not in the \inlinecode{solute} mask is treated as solvent.
+If there are multiple solvent species, use the \inlinecode{solventmols} option as described in Section~\ref{sec-salt-water}.
+The code will automatically choose three atoms which are used to define the solvent orientation, and print this selection to the output.
+
+Sometimes, \software{cpptraj} does not automatically choose the optimal atoms.
+In methanol, for instance, one can either incorporate the C--H bonds or the O--H bond in the selection.
+The latter is probably more relevant since it can incorporate hydrogen bonding effects.
+Assuming that the alcohol hydrogen is called H1, this could be specified as follows:
+
+\begin{lstlisting}[style=cpptraj]
+gist gridcntr <x> <y> <z> \
+griddim <Nx> <Ny> <Nz> gridspacn <val> \
+solute ^1 rigidatoms O1 H1 C1
+\end{lstlisting}
+
+Note that the central atom of the \inlinecode{rigidatoms} selection goes first.
+
 \subsection{GIST with salt-water mixtures}
+\label{sec-salt-water}
 In \cite{Waibl2021-gist-salt}, an extension of GIST was presented that can use salt-water mixtures as a solvent.
@@
 	\includegraphics[width=1.0\linewidth]{streptavidin_bp_close.png}
-	\caption{Volume of strepatividin defined around 3.5 A of biotin. }\label{streptavidin_bp_close}
+	\caption{Volume of streptavidin defined around \SI{3.5}{\angstrom} of biotin. }\label{streptavidin_volume}
 \end{figure}

### 2022-11-02 manuscript/manuscript.tex

@@
 In PME-GIST, the system is treated as periodic, and the electrostatic energy is split into a short-range term $E_\text{dir}$, which is calculated in direct space using a distance cutoff, and a long-range term $E_\text{rec}$, which is calculated in reciprocal space.
-Additionally, there is a correction term $E_\text{corr}$, which corrects for the self-interaction of each solvent molecule in the reciprocal term.
-
-\begin{equation}
-	E_\textit{elec} = E_\textit{dir} + E_\textit{rec} + E_\textit{corr}
+Additionally, there is a correction term $E_\text{self}$ (called $E_\text{corr}$ in the original publication \cite{Chen2021}), which corrects for the self-interaction of each solvent molecule in the reciprocal term.
+
+\begin{equation}
+	E_\textit{elec} = E_\textit{dir} + E_\textit{rec} + E_\textit{self}
 \end{equation}
 %\todo{How is E$_{dir}$ treated? Is there a cutoff as well (PME GIST paper seems to indicate so)? If so, we should probably note that. } - ok? -fwa
-The Lennard-Jones contribution is computed in the direct space, using a distance cutoff.
-The short-range contribution is the LJ energy within the set cutoff and the long-range correction term accounts for the contributions above this cutoff.
-
-\begin{equation}
-	E_\textit{lj} = E_\textit{lj,\ short} +  E_\textit{corr}
-\end{equation}
-\todo{Is $E_{corr}$ the same in eq 13 and 14? How is it calculated? -vah}
+The short-range Lennard-Jones contribution is computed in the direct space, using a distance cutoff.
+Furthermore, a long-range correction term is computed that accounts for the contributions above this cutoff assuming a homogeneous distribution of particles.
+
+\begin{equation}
+	E_\textit{lj} = E_\textit{lj,\ short} +  E_\textit{lj\_corr}
+\end{equation}
 

### 2022-11-22 manuscript/manuscript.tex
Proof-read text, many minor comments until Strept/Biotin section.

@@
 
-\newcommand{\versionnumber}{1.3}  % you should update the minor version number in preprints and major version number of submissions.
-\newcommand{\githubrepository}{\url{https://github.com/myaccount/homegithubrepository}}  %this should be the main github repository for this article
+\newcommand{\versionnumber}{1.0}  % you should update the minor version number in preprints and major version number of submissions.
+\newcommand{\githubrepository}{\url{https://github.com/liedllab/gist-tutorial}}  %this should be the main github repository for this article
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\title{This is the title [Article v\versionnumber]}
-
-\author[1*]{Firstname Middlename Surname}
-\author[1,2\authfn{1}\authfn{3}]{Firstname Middlename Familyname}
-\author[2\authfn{1}\authfn{4}]{Firstname Initials Surname}
-\author[2*]{Firstname Surname}
-\affil[1]{Institution 1}
+\title{Quantifying Spatially Resolved Hydration Thermodynamics Using Grid Inhomogeneous Solvation Theory [Article v\versionnumber]}
+
+\author[1\authfn{1}]{Franz Waibl}
+\author[1\authfn{1}]{Valentin J Hoerschinger}
+\author[2]{Vjay Molino}
+\author[1]{Monica L Fern{\'a}ndez-Quintero}
+%\author[1,2]{Firstname Middlename Familyname}
+%\author[2]{Firstname Initials Surname}
+\author[4]{Daniel R Roe}
+\author[1*]{Klaus R Liedl}
+\author[2*]{Michael K Gilson}
+\author[2*]{Tom Kurtzman}
+\affil[1]{Department of General, Inorganic and Theoretical Chemistry, University of Innsbruck, Austria}
 \affil[2]{Institution 2}
@@
 \contrib[\authfn{1}]{These authors contributed equally to this work}
-\contrib[\authfn{2}]{These authors also contributed equally to this work}
-
-\presentadd[\authfn{3}]{Department, Institute, Country}
-\presentadd[\authfn{4}]{Department, Institute, Country}
+%\contrib[\authfn{2}]{These authors also contributed equally to this work}
+
+%\presentadd[\authfn{3}]{Department of General, Inorganic and Theoretical Chemistry, University of Innsbruck, Austria}
+%\presentadd[\authfn{4}]{Department, Institute, Country}
 
@@
 The high spatial resolution of the GIST output, as well as the decomposition into energy and entropy contributions, allow for highly detailed analyses on both proteins and small molecules.
-However, this versatility also comes with a higher entry barrier for new users.
+However, this versatility also comes with a significant entry barrier for new users.
 
 In this tutorial, we aim to guide the reader through the most common steps involved in a GIST analysis at the example of the streptavidin-biotin complex.
-Furthermore, we discuss the theory of GIST with a focus on practical aspects, and show several pitfalls and technical difficulties that may occur during a GIST study.
+Furthermore, we discuss the theory of GIST with a focus on practical aspects.
+We highlight potential pitfalls and show how to avoid technical difficulties.
 We assume familiarity with molecular dynamics simulations as well as the AmberTools package.
@@
 \todo{IMO, our article looks too dry. We should have something nice (e.g., an illustration of the GIST method) on the 1st or 2nd page -fwa.}
+\todo{Potentially relevant: 10.1021/acs.jcim.7b00520 }
 
 Solvation thermodynamics govern any process involving solutes in a solvent.
+%Cite "Theory of Simple Liquids"?
 %Solutes are constantly interacting with solvents in their environment, necessitating a good description of these interactions for accurate computational predictions. 
-The free energy of solvation is defined as the work needed to insert a solute into a solvent.
-Especially the interaction with water is of utmost importance, as many biological processes occur in a aqueous environment.
-It thereby strongly influences real-world problems such as hydrophobic effects, the binding of ligands to a biomolecule, as well as the dynamics and folding of proteins.
+The free energy of solvation is defined as the reversible work needed to insert a solute into the solvent \cite{ben-naim-book}.
+Especially the interaction with water is of utmost importance, as many biological processes occur in aqueous environment \cite{Privalov2017-water-review}.
+It drives real-world effects such as hydrophobic interactions, binding of ligands to a biomolecule, as well as dynamics and folding of proteins.
 
@@
 Data-driven methods such as QSPR or UNIFAC can deliver good results \cite{Borhani2019-qspr,Fredenslund1975-unifac}, but might be limited to molecules that are similar to those in the training set.
-Polarizable continuum models (PCMs) \cite{Miertus1981-pcm,Klamt1993-cosmo} such as treat the solvent phase as a homogeneous phase that reacts to the electrostatic potential of the solute \cite{Mennucci2010-pcm}.
+Polarizable continuum models (PCMs) \cite{Miertus1981-pcm} such as COSMO \cite{Klamt1993-cosmo} treat the solvent as a homogeneous phase that reacts to the electrostatic potential of the solute \cite{Mennucci2010-pcm}.
 %Molecular mechanics based approaches describe the underlying PES through empirically fit force-fields, which reduce the computational demand of these calculations tremendously.
 %This allows for the treatment of large biological systems such as proteins in aqueous solution.
-Methods such as MM/PBSA or MM/GBSA \cite{Sitkoff1994-pbsa,Kollman2000-mmpbsa} combine an implicit solvent treatment of electrostatics with structural sampling of the solute.
-They can quickly provide results for a large number of structures, but yield inaccurate results when compared to more rigorous explicit solvent methods \cite{Genheden2015-mmpbsa-review}.
+Methods such as MM/PBSA or MM/GBSA \cite{Sitkoff1994-pbsa,Kollman2000-mmpbsa} combine an implicit treatment of solvent electrostatics with structural sampling of the solute.
+They can quickly provide results for a large number of structures, but can yield inaccurate results when compared to more rigorous explicit solvent methods \cite{Genheden2015-mmpbsa-review}.
 
@@
 They are generally more accurate than implicit methods, but require a statistical sampling of the solvent conformations \cite{Liu2016-md-solubility,Swails2014-cphmd}.
-Molecular dynamics (MD) simulations provide an accurate way of modeling solvent packing in confined areas \cite{}\todo{ask Tom for a citation!} and general hydrophobic effects \cite{Pratt2016-hydrophobicity}.
+Molecular dynamics (MD) simulations provide an accurate way of modeling solvent packing in confined areas \cite{Haider2016-water-on-surfaces} and general hydrophobic effects \cite{Pratt2016-hydrophobicity}.
 
 The free energy of solvation can be calculated in a statistically rigorous way using alchemical methods \cite{Liu2016-md-solubility,Mobley2009-dgsolv,Mobley2014-freesolv} such as free energy perturbation (FEP) \cite{Zwanzig1954-reweighting} or thermodynamic integration (TI) \cite{Kirkwood1935-ti}.
-However, they are unable to provide a spatial interpretation of the results.
-Furthermore, splitting the resulting free energy into contributions for the enthalpy and entropy is challenging \cite{Peter2004-alchemical-entropy}.
+However, those methods are unable to provide a spatial interpretation of the results.
+Furthermore, splitting the free energy into enthalpy and entropy contributions is challenging \cite{Peter2004-alchemical-entropy}.
 %Assuming sufficient sampling, they can accurately predict the free energy of solvation in the boundaries of the assumptions underlying their forcefield. 
@@
 Methods derived from statistical mechanics bridge this gap by providing three-dimensional resolution.
-Two major groups of methods that fall into this group are those based on inhomogeneous solvation theory (IST) \cite{Lazaridis1998} and those based on the Ornstein--Zernike (OZ) equation \cite{Hansen2013-simple-liquids}.
+Two major flavors of methods in this group are those based on inhomogeneous fluid solvation theory (IST) \cite{Lazaridis1998} and those based on the Ornstein--Zernike (OZ) equation \cite{Hansen2013-simple-liquids}.
 The most important example of OZ-based methods is the reference interaction site model (RISM) \cite{Chandler1972-rism} as well as its extension to three dimensions (3D-RISM) \cite{Kovalenko1998-3drism}.
-Those methods compute the solvent distributions by self-consistent solution of the OZ integral equation to predict solvation thermodynamics.
+Those methods compute the atomic solvent distributions by self-consistently solving the OZ integral equation to predict solvation thermodynamics.
 %The 3D-RISM approach allows for a decomposition into entropic and enthalpic terms as well as a full localization of the underlying distributions to the solute structure.
@@
 On the other hand, methods based on IST generally use a molecular solvent distribution obtained from an MD simulation to compute the free energy of hydration from its enthalpic and entropic contributions.
-Examples of those methods include grid inhomogeneous solvation theory (GIST) \cite{Nguyen2012,Ramsey2016}, WaterMap \cite{Young2007-watermap,Abel2008-watermap}, SSTMap \cite{Haider2018-sstmap}, and STOW \cite{Li2012-stow}.
+Examples of those methods include grid inhomogeneous solvation theory (GIST) \cite{Nguyen2012,Ramsey2016}, WaterMap \cite{Young2007-watermap,Abel2008-watermap}, SSTMap \cite{Haider2018-sstmap}, STOW \cite{Li2012-stow} and Solvaware \cite{Huggins2016-solvaware}.
 The main limitation of those methods is that they rely on an infinite expansion of the entropy in terms of the density distribution, which is truncated after the first or, sometimes, the second \cite{Nguyen2016-gist-second-order,Waibl2022-gist-solvents} term.
@@
 GIST is based on IST, solving the integral equations on a grid. 
-This approach is more general than other IST-based methods because it accounts for all solvent regions rather than focusing on high-occopancy solvation sites.
+This approach is more general than other IST-based methods because it accounts for all solvent regions rather than focusing on high-occupancy solvation sites.
 %By simulating the dynamics of a solvent around the solute, GIST calculates thermodynamic properties for a standard solvation process as described by Ben-Naim \cite{ben-naim-book}.
-It estimates the solvent density from a molecular dynamics (MD) trajectory and allows for a full localization of entropic and enthalpic contributions on the grid around the solute.
-It has been shown that GIST can produce comparable accuracy to TI by approximating the higher-order entropies as a fraction of the first-order entropy (using a factor of \SI{-0.4} for water) \cite{Chen2021,Waibl2022-gist-solvents}.
+It estimates the solvent density from an MD trajectory and allows for a full localization of entropic and enthalpic contributions on the grid around the solute.
+For small molecules, GIST has been shown to reach comparable accuracy to TI by approximating the higher-order entropies as a fraction of the first-order entropy, using a factor of \SI{-0.4} for water \cite{Chen2021,Waibl2022-gist-solvents}.
 %applying a linear correction factor (0.6 for water) to the first-order entropy as follows \cite{Chen2021,Waibl2022-gist-solvents}:
@@
 
-Recent work has improved the calculation speed tremendously, allowing for the treatment of large biological systems such as proteases or antibodies.
+Recently, GIST was implemented on the GPU, tremendously improving the calculation speed, and allowing for the treatment of large biological systems such as proteases or antibodies \cite{Kraml2019-gigist}.
 Furthermore, an implementation based on the particle mesh Ewald (PME) \cite{Darden1993-pme} method was presented, which improves the consistency between GIST and popular molecular dynamics engines. 
-Yet further changes allow the inclusion of other rigid solvents than water or 
-allow for the treatment of solvent mixtures and ions in solution. 
-\todo{Sources!}
-
-While previous tutorials gave insightful pointers how to setup and run GIST calculations, their post-processing was often treated less rigorously.
-While building on these tutorials, an updated guide to calculating solvation properties using GIST is therefore presented here to address recent improvements to the method as well as provide a basic template for the analysis and interpretation of GIST calculations.
+Yet further changes allow the inclusion of other rigid solvents than water \cite{Kraml2020,Waibl2022-gist-solvents} or 
+allow for the treatment of salt-water mixtures as solvents \cite{Waibl2021-gist-salt}.
+
+While previous publications \cite{Ramsey2016} and tutorials\footnote{See \url{http://ambermd.org/tutorials/advanced/tutorial25/}} gave insightful pointers how to setup and run GIST calculations, the post-processing was often treated less rigorously.
+While building on these tutorials, we provide an updated guide to calculating solvation properties using GIST to address recent improvements.
+Furthermore, we aim to establish a template for the analysis and interpretation of GIST calculations.
 
@@
 
-This tutorial provides an example of a GIST workflow that can be easily adapted towards different systems. 
-In addition to a short introduction of the method and the theory behind it, we present an overview of the various different flavors of GIST methods.
-We ease the user into the basic usage of GIST by presenting a calculation for a small molecule.
-Building on the basics, a more advanced GIST study of the streptavidin-biotin complex is presented, which reflects on many common aspects of GIST studies.
-Using these test systems, we present solvation analysis of a small molecule (biotin) as well as a biomolecule (streptavidin).
+This tutorial provides an example of a GIST workflow that can be easily adapted to different systems. 
+In addition to a short introduction of the method and the theory behind it, we present an overview of the different implementations of GIST.
+We ease the user into basic GIST analysis at the example of biotin.
+Building on the basics, we present a more advanced GIST study of the streptavidin-biotin complex, reflecting on many common aspects of GIST studies.
 We show how to interpret the three-dimensional contributions of solvation free energy in a binding pocket and around the ligand.
-
-Additionally, we compute the binding contribution of the free energy of solvation, which requires accurate post-processing of the GIST outputs to avoid unfavorable summation of bias in the GIST calculation.
-We believe that this example study covers a wide range of applications of the GIST method.
+Additionally, we compute the binding contribution of the free energy of solvation, which requires accurate post-processing of the GIST outputs to avoid unfavorable summation of bias.
+We believe that this example study covers a wide range of applications.
 
 Furthermore, we discuss several technical aspects and typical pitfalls of GIST analyses, such as the normalization of voxel values and how to deal with double counting of energy contributions. 
-Additionally, we provide a python library to unify the analysis of GIST data produced by various versions of GIST and make the post-processing of GIST results more accessible.
+Additionally, we provide a Python library (\software{gisttools}) to unify the analysis of GIST data produced by various versions of GIST and make the post-processing of GIST results more accessible.
 
@@
 This tutorial is aimed at users with a solid knowledge of molecular dynamics (MD) simulations.
-The user should be able to run MD simulations and work with their output data.
+The user should be able to run MD simulations and analyze the resulting trajectories.
 Previous experience of GIST is not necessary.
-However, we only provide a short explanation of the theory behind GIST in section \ref{sec:theory} at the end of this tutorial.
-For more detailed information, we point the reader towards the primary sources for more detailed information.
+However, we only provide a short summary of the theory behind GIST in section~\ref{sec:theory} at the end of this tutorial.
+For more detailed information, we point the reader towards the primary sources.
 
 The presented analyses are run in Python.
-While no programming experience should be necessary to follow the tutorial, a basic understanding of Python is advantageous to adjust the code towards different requirements and use-cases. 
-We provide a Jupyter notebook which contains the presented analyses. 
-Therefore, an understanding of Jupyter notebooks is beneficial to reproduce the analyses.
+While no programming experience is necessary to follow the tutorial, a basic understanding of Python is advantageous to adjust the code towards different requirements and use-cases.
 
@@
 
-The closed-source Amber simulation engine is used in this tutorial to run the necessary MD simulations.
-If no Amber installation is available, the MD can be performed other simulation engines, as long as the topology and trajectory can be produced in a format supported by \software{cpptraj}.
+The closed-source Amber simulation engine is used in this tutorial to run the MD simulations.
+If Amber is not available, other MD engines can be used as long as topologies and trajectories can be produced in a format supported by \software{cpptraj}.
 If using GROMACS \cite{Abraham2015-gromacs,Pronk2013-gromacs}, we propose to prepare the topology with AmberTools and convert it using \software{acpype.py} \cite{Sousa_da_Silva2012-acpype}.
 %We should actually test that!
-Also note that compressed trajectory formats such as \inlinecode{xtc} might bias the entropy calculation due to the loss of precision for the atom positions.
+Also note that compressed trajectory formats such as \inlinecode{xtc} might bias the entropy calculation due to the loss of precision for the atom positions. \todo{This might be in the technical part}
 
@@
 	\item \software{gisttools} (\url{https://github.com/liedllab/gisttools})
-	\item \software{mdtraj (\url{https://www.mdtraj.org/})}
+	\item \software{mdtraj (\url{https://www.mdtraj.org/}) \cite{McGibbon2015-mdtraj}}
 \end{itemize}
@@
 
-For the visualization of three-dimensional contributions we require molecular visualization software that can handle \software{OpenDX} files.
-We recommend using a current version of \software{PyMol}\cite{pymol} or \software{VMD}\cite{vmd}.
+For the visualization of three-dimensional contributions we require molecular visualization software that can handle \software{OpenDX} files, such as \software{PyMol} \cite{pymol} or \software{VMD} \cite{vmd}.
 
@@
 \begin{itemize}
-	\item A running installation of \software{cpptraj}.
+	\item \software{cpptraj}.
 	\item A topology of the molecule of interest in a solvent box.
@@
 \paragraph{Identify and define the region of interest}
-In order to run a GIST analysis, the region where the solvent properties will be calculated must be first defined.
-This region is represented by a rectangular grid centered at (x, y, z) with $N_x \times N_y \times N_z$ voxels, where each voxel is cubic with a sidelength of \inlinecode{gridspacn}.
-While integrals over GIST quantities are independent of the voxel size, individual voxel values converge slower at high resolutions.\pdfcomment{The calculation actually can actually get faster at higher resolution, because the same grid is used in the nearest-neighbor search.}
+In order to run a GIST analysis, one must define the region where solvent properties will be calculated.
+This region is represented by a rectangular grid with given position, size, and voxel spacing.
 The default value of \SI{0.5}{\angstrom} provides reasonable discretization for many use-cases.
+While integrals of GIST quantities over the grid are independent of the voxel size, individual values converge slower at high resolutions.
 
 The grid is defined by parameters of the \inlinecode{gist} command in \software{cpptraj}.
-The grid center is defined by \inlinecode{gridcntr}, the voxel number in x, y, and z direction by \inlinecode{griddim}, and the spacing in \AA by \inlinecode{gridspacn}.
+The grid center is defined by \inlinecode{gridcntr}, the voxel number in x, y, and z direction by \inlinecode{griddim}, and the spacing (voxel side-length) in \AA by \inlinecode{gridspacn}.
 
@@
 
-Often, the region of interest will be defined by the surrounding of a molecule, which might either be a ligand in the binding pocket or also the whole protein. 
+Often, the region of interest will be defined as the surrounding of a molecule, which might for example be a ligand in the binding pocket or the whole protein.
 A helper script called \inlinecode{FindCentroid.py} for this purpose is supplied with this tutorial, and can be used as follows:
@@
 
-The first line specifies the amber topology file that was used to run MD. The second line species the trajectory file.
-Note that the solute should not move significantly for a GIST calculation.
-This is typically achieved by using restraints on the solute coordinates.
-The third line tells \software{cpptraj} to run a GIST calculation using the region defined and print the output of the calculation in a file called \inlinecode{gist.dat}.
-The last instructs cpptraj to run the calculation.
+The first line specifies the Amber topology file that was used to run MD. The second line species the trajectory file.
+Note that the solute should not move significantly for a GIST calculation, since IST does not take the solute degrees of freedom into account.
+This is typically achieved using restraints on the solute heavy atoms.
+The third line tells \software{cpptraj} to run a GIST calculation using the specified region and print the output in a file called \inlinecode{gist.dat}.
+The last line instructs \software{cpptraj} to run the calculation.
+
 The syntax for running \software{cpptraj} given this input file is:
-
 \begin{lstlisting}[style=bash]
@@
 
-The \inlinecode{gist} action has a large number of options to modify the GIST calculation or its' output.
-A full list is provided at the end of this tutorial or through cpptrajs manual.
-The major options are summarized in the following table:
+The \inlinecode{gist} action has a large number of options to modify the GIST calculation or its output.
+A full list is provided at the end of this tutorial or through the \software{cpptraj} manual.
+The most important options are summarized in the following table:
 
 \begin{table}[h]
-	\caption{The major options available in cpptrajs' GIST implementation.}\label{tab:GIST_options}
+	\caption{The main options for GIST in \software{cpptraj}.}\label{tab:GIST_options}
 	\small
@@
 		gridcntr    & <y> <y> <z> & Coordinates of the grid center in \AA & 0 0 0  \\
-		griddim     & <nx> <ny> <nz> & Number of voxels along each axis defining the grid size & 40 40 40\\
+		griddim     & <nx> <ny> <nz> & Number of voxels per axis & 40 40 40\\
 		gridspacn & <dval> &Grid spacing / voxel dimensions in  \AA & 0.5    \\
 		refdens         & <rdval> & Reference density of the bulk water in molecules/\AA$^3$ . & 0.0334    \\
+		temp & <tval> & System temperature in Kelvin & 300 \\
 		\bottomrule
@@
 \paragraph{Postprocessing and Analyzing GIST results}
-Cpptraj outputs the quantities calculated as a "open data explorer" (.dx) file format which can be easily visualized in Pymol or VMD.
-Aside from the thermodynamic quantities, GIST also calculates the number densities of hydrogen and oxygen in each voxel. Here is a list of the most important quantities calculated by GIST:
+Cpptraj outputs quantities in OpenDX (.dx) file format which can be visualized in Pymol or VMD.
+Aside from thermodynamic quantities, GIST also calculates the number density of each atomic element in the solvent.
+Here is a list of the most important quantities calculated by GIST:
 
@@
 
-Several dx files are written by default, such as the energies, entropies, number density of oxygen, and number density of hydrogen.
-All calculated quantities are written to the output file (output.dat) and can be used to generate DX files for other quantities.
-A full list of all calculated properties found in the output in addition to the ones mentioned above is provided at the end of the tutorial.
+Several dx files are written by default, such as the energy, entropy, and atomic number densities.
+All calculated quantities are written to the output file (default \inlinecode{gist-output.dat}) and can be used to generate DX files for other quantities.
+A full list of all properties in the output is provided at the end of the tutorial.
 
@@
 Here, we aim to guide the reader through a GIST analysis of the streptavidin-biotin complex.
-During the tutorial, we will
+We will
 \begin{enumerate}
-	\item set up and run GIST calculations for a small molecule as well as a protein binding pocket.
-	\item compute the free energy of solvation for a small molecule using GIST.
-	\item visualize local contributions to $\dgsolv$ in the streptavidin binding pocket as well as around biotin.
-	\item compute the binding free energy of the complex from the individual energy contributions combined with the contribution of hydration computed using GIST.
+	\item set up and run GIST calculations for a small molecule (biotin) and a protein (streptavidin) binding pocket.
+	\item compute the free energy of solvation for biotin using GIST.
+	\item visualize local contributions to $\dgsolv$ in the streptavidin binding pocket and around biotin.
+	\item compute the hydration contribution to the binding. Combined with the biotin--streptavidin interaction energy, the binding free energy can be estimated.
 	\item explore advantages and disadvantages of the GIST method when used to calculate  $\dgsolv$
 \end{enumerate}
-The tutorial is divided into a beginners part and an advanced part.
-In the beginners part, we will take a look at how to run GIST calculations for small molecules and interpret the output for a small molecule.
-In the advanced part, we will apply this knowledge to a protein binding pocket, visualize $\dgsolv$ hotspots and estimate a free energy of binding for a small molecule into this pocket using an end-state approach.
-In addition, while the beginners part focuses only on GIST calculation run, we will discuss all steps along the way in detail in the advanced section.
+
+In the first part of the tutorial, we will look at how to run GIST calculations for small molecules and interpret the output.
+Afterwards, we will apply this knowledge to a protein binding pocket, visualize $\dgsolv$ hotspots and estimate a free energy of ligand binding using an end-state approach based on GIST.
+In addition, we will discuss all steps in further detail at the end.
+\todo{TODO}
 
@@
 In nature, streptavidin occurs as a tetramer, and binds biotin with a binding constant of \SI{\sim e-14}{\per\Molar} \cite{Dundas2013-streptavidin-review}.
-The binding to the closed conformation of the tetramer has been estimated to have a free energy of \SI{26.6}{\kilo\calorie\per\mol} \cite{Bansal2018-biotin}.
+The binding to the closed conformation of the tetramer has been estimated to have a free energy of \SI{-26.6}{\kilo\calorie\per\mol} \cite{Bansal2018-biotin}.
 Here, we will investigate the binding of biotin to a closed streptavidin monomer for simplicity.

### 2022-11-23 manuscript/manuscript.tex
Continue the proof-reading process, crop white space from an image

@@
 	extendedchars=true,
-	literate={Å}{{\AA}}1
+	literate={Å}{{\AA}}1,
+	backgroundcolor = \color[rgb]{0.9, 0.9, 0.9},
 }
@@
 \lstdefinestyle{cpptraj}{title={\centering Cpptraj}}
-\lstdefinestyle{bash}{title={\centering Commandline}}
+\lstdefinestyle{bash}{title={\centering Command-line}}
 \lstdefinestyle{amber-in}{title={\centering Amber input}}
@@
 \subsection{Streptavidin/Biotin}
-The streptavidin-biotin complex is one of the strongest known non-covalent interactions involving a small molecule.
+The streptavidin-biotin complex binding is one of the strongest known non-covalent interactions involving a small molecule.
 Therefore, the combination of streptavidin and biotin is used routinely in biotechnology \cite{Dundas2013-streptavidin-review}, and the interaction between them has been investigated in detail \cite{McConnell2021-biotin}.
@@
 In nature, streptavidin occurs as a tetramer, and binds biotin with a binding constant of \SI{\sim e-14}{\per\Molar} \cite{Dundas2013-streptavidin-review}.
+Streptavidin has a so-called flap region, which closes the binding pocket after biotin binding.
+Here, we will investigate the binding of biotin to a closed streptavidin monomer for simplicity.
 The binding to the closed conformation of the tetramer has been estimated to have a free energy of \SI{-26.6}{\kilo\calorie\per\mol} \cite{Bansal2018-biotin}.
-Here, we will investigate the binding of biotin to a closed streptavidin monomer for simplicity.
 
 \subsection{Tutorial data}
-We will use the 1STP crystal structure to start our calculations.
-For the beginners section, we will use the biotin ligand as extracted from the crystal structure.
+We will use the crystal structure of streptavidin (PDB code: 1STP \cite{Weber1989-streptavidin-structure}) to start our calculations.
+For the beginners section, we will use the biotin ligand extracted from the crystal structure.
 For the second part, we will use the whole crystal structure.
@@
 \end{lstlisting}
-This will download the tutorial files (including both the manuscript and the examples in the \software{code} folder).
-\
+This will download the tutorial files, including both the manuscript and the examples in the \inlinecode{code} folder.
+
 \subsection{Running GIST for Biotin}
-In this example, we will be running a simple GIST calculation for a small molecule to get familiar with the usage of the GIST command in \software{cpptraj}.
+In this example, we will be running a simple GIST calculation of the biotin molecule to get familiar with the usage of GIST in \software{cpptraj}.
 \subsubsection{System preparation}
-First, extract biotin from the crystal structure and generate a PDB file containing only biotin.
+First, extract biotin from the crystal structure and generate a topology and coordinate file compatible with Amber or a simulation package of your choice.
+The molecule should be solvated with a sufficiently large solvent box.
+Carefully check the employed force field, protonation state and partial charges.
+%a PDB file containing only biotin.
 Prepare your structure by adding hydrogens, making sure to choose the correct protonation state.
 Parametrize biotin with the forcefield of your choice.
-Solvate the parametrized biotin, making sur to choose a water box of large enough size so that the water on the box edges behaves bulk-like (a wall distance of at least 15\AA is recommended).
-A good starting point (though a bit dated) for the parametrization and solvation of small molecules is AMBER's tutorial 4b (\url{http://ambermd.org/tutorials/basic/tutorial4b/}).
+Solvate the parametrized biotin, making sure to choose a water box of large enough size so that the water on the box edges behaves bulk-like (a wall distance of at least 15\AA is recommended).
+A good starting point (though a bit dated) for the parametrization and solvation of small molecules is the Amber tutorial 4b (\url{http://ambermd.org/tutorials/basic/tutorial4b/}).
+
 \subsubsection{Equilibration}
-Equilibrate your solvated biotin to the conditions of your choice.
-Even when running a NVT simulation as the production run for the GIST analysis, it is advisible to include an NPT equilibration step to adjust the box size before hand.
-Allow enough equilibration time in the ensemble of your choice to allow the water to relax around your biotin structure.
+Equilibrate your solvated biotin in an NVT or NpT ensemble.
+Even when running the simulation in NVT, it is advisable to include an NpT equilibration step to adjust the box size beforehand.
+Use enough equilibration time to allow the water to relax around your biotin structure.
 
@@
 
-The choice of the initial structure to be analyzed is crucial to the results 
-you can expect from the GIST calculation. Crystal structures may not be 
-representative of the dominant solution structure for a given analysis target. 
-In our example we use the final structure after a short equilibration run. A 
-more sophisticated approach would be identifying a dominant structure out of 
-the ensemble of sampled structures.  Depending on what properties you are 
-interested in, running a MD simulation of your structure beforehand and 
-choosing a dominant structure or even multiple structures through clustering 
-might therefore be beneficial.
-When starting with a crystal structure, special care has to be taken during the 
-preperation of the structure. Make sure to protonate your structure and take 
-care to choose a proper protonation state for the varying chemical moeties in 
-your structure. You will need to solvate your structure in a solvent box, which 
-should be chosen large enough so that the solvent around the region of interest 
-is not influenced through periodic images.
+The choice of initial structure is crucial to the GIST results.
+Crystal structures may not represent the dominant solution structure for a given target.
+In our example, we only use a single equilibrated structure.
+A more sophisticated approach would be to identify one or several dominant structures out of the ensemble.
+Depending on what properties you are interested in, it might be beneficial to run an MD simulation choose dominant structures through clustering.
+
+When starting from a crystal structure, care has to be taken during the preparation step.
+Make sure to choose a proper protonation state for all titratable moeties.
+You will need to solvate your structure in a solvent box, which should be large enough to obtain unbiased long-range electrostatics.
+A good criterion for validating the box size is to check whether the water on the outside of the box behaves bulk-like according to the GIST output.
 
 When calculating solvation free energy differences between a complex and the corresponding monomers (the dissociated state), there are two options for dealing with configurational diversity in the structures.
-\todo{Maybe Valentin knows a citation for that?}
-One can choose to sample the complex and dissociated states individually, or to sample only one of them and assume that there is sufficient overlap in the sampled conformations.
-%One can choose to sample only the complex, assuming that all relevant conformations of the dissociated state are also found in the complex.
-%The probability distribution of the dissociated state can be obtained by reweighting the probabilities in the complex using the solvation free energies.
-%Alternatively, one can sample the complex and dissociated states individually.
-%This approach does not require an overlap between the conformational probabilities of the complex and dissociated states.
-%However, it is prone to statistical errors due to uncertainties in the ensembles, and requires the free energy contributions from the internal degrees of freedom of the monomers (in both the complex and the monomer states) to be known.
-\todo{This paragraph does not fit here well.
-Maybe put it somewhere else.}
-
-Here, we choose not to include any sampling of the conformational states. 
-It is therefore important that the conformations of the monomers in the complex and dissociated states match exactly. 
+One can choose to sample the complex and dissociated states individually, or to sample only one of them and assume that there is sufficient overlap in the conformations.
+
+Here, we do not include any sampling of the conformational states.
+It is therefore important that the molecular conformations in the complex and dissociated states match exactly.
 Use \software{cpptraj} to produce biotin and streptavidin structures based on the complex by stripping the respective other molecule like this:
@@
 
-Use the \software{equilibration.py} script (or your own) to perform short NVT and NpT equilibration runs on the starting structure.
-\todo {NPT}
-Note that the barostat might interfere with restrained equilibration of the complex, since different molecules are moved relative to each other when the volume changes.
-But since our equilibration script also contains unrestrained pressure 
-equilibration, we should not run into any trouble.
+Use the \software{equilibration.py} script (or your own) to perform short NVT and NpT equilibration runs on the complex structure.
+Note that the equilibration script contains unrestrained NpT steps, because restraining two molecules at the same time might interfere with the pressure equilibration.
 
@@
 Now, you can run restrained MD simulations based on the equilibrated structures for subsequent GIST analyses.
-\todo {explain why the protein needs to be restrained for an MD simulation that will be analyzed using GIST}
+Restraints should be used during the MD because IST theory is not defined for moving solutes.
 We apply a harmonic restraint of \SI{100}{\kilo\calorie\per\mole\per\angstrom\squared} to all heavy atoms to keep the system in place.
 We run \SI{100}{\nano\second} of simulation and store the conformations every \SI{100}{\pico\second}.
-A simulation length of at least \SIrange{10}{20}{\nano\second} is recommended, to obtain statistically independent snapshots of the water.
+A simulation length of at least \SI{20}{\nano\second} is recommended to obtain statistically independent snapshots of the water molecules.
 
@@
   nstlim=50000000, dt=0.002,
-  ntwr=50000, ntwx=5000,
-  ntpr=5000,
+  ntwr=50000, ntwx=5000, ntpr=5000,
 /
 \end{lstlisting}
-Note that these are very standard MD settings except for the line that defines the restraints.
+Note that these are very standard MD settings except for the restraints.
 
@@
 
-\subsection{Running the GIST analysis for Streptavidin-Biotin Complex}
-For easier post-processing, we will use grids that contain the whole molecule.
-This could in theory be avoided for streptavidin and the complex, since only the surrounding of the biotin molecule (i.e., the binding pocket) is required for the analysis.
-We also recommend that you center the molecule in space.
+\subsection{GIST analysis of Streptavidin-Biotin}
+For easier post-processing, here we use grids that contain the whole molecule.
+Smaller grids encompassing only the binding site could be used for streptavidin and the complex, but this would complicate the analysis.
+%This could in theory be avoided for streptavidin and the complex, since only the surrounding of the biotin molecule (i.e., the binding pocket) is required for the analysis.
+Furthermore, we recommend that you center the molecule in space.
 For each system (biotin, streptavidin, complex), produce a centered pdb file using \software{cpptraj}:
@@
 \end{lstlisting}
+
 Then, find the extent (minimum and maximum) of the x-, \mbox{y-,} and z-coordinates for each system.
@@
 What does gisttools do if the grid is too small? -vah }
-We note that this centered grid is usually larger than it needs to be in some directions.
-If you want to save some extra memory and calculation time, feel free to align the grid more precisely to the minimum and maximum coordinates, or use a smaller grid that only covers the binding pocket.
-You can also rotate the molecule by its principal axes to fit the cuboid grid more exactly.
-However, this needs to be done \emph{before} the MD run, since rotating the trajectory damages the periodic box information.
-
-Now, you can finally run the actual GIST analysis. Center the molecule exactly in the same way as above, to make sure that its position matches the grid parameters that you decided on.
-We recommend using either the PME or GPU implementation of GIST, depending on whether you have a fast GPU. To use the GPU version of GIST, cpptraj needs to be compiled with the "-cuda" flag set.
-While the GPU version is generally faster, an advantage of the PME implementation is that it matches the energy calculation of the MD engine more closely.
+A buffer distance of \SI{15.0}{\angstrom} is reasonable for biotin, but smaller values are sufficient for the larger streptavidin and complex systems.
+
+Now, you can finally run the actual GIST analysis.
+Center the molecule as above, to make sure that its position matches the grid parameters you decided on.
+We recommend using either the PME or GPU implementation of GIST.
+For the GPU version, cpptraj needs to be compiled with the \inlinecode{-cuda} flag.
+While the GPU version is generally faster, the PME implementation matches the energy of the Amber MD engine more closely.
 
@@
 autoimage !(:WAT) origin
-# remove the "pme" option if you want to use the GPU
+# optionally remove the "pme" option
 gist griddim 100 100 100 out gist.dat \
   refdens 0.03287 pme
-# other settings you might want to change
-# (those are the defaults)
-# gridcntr 0 0 0 gridspacn 0.5 temp 300
-\end{lstlisting}
+\end{lstlisting}
+
 \subsection{Reference Values and Radial Convergence}
 All GIST quantities should be expressed relative to bulk.
-In practice, this is not problematic for the entropies, since they are computed relative to a distribution of randomly oriented molecules at the reference density.
-The solute-solvent energy $E_{sw}$ is also no problem, because it naturally tends to zero in bulk. 
+This automatically done for the entropy, using the bulk density specified by \inlinecode{refdens}.
+The solute-solvent energy $E_{sw}$ naturally tends towards zero in bulk.
 However, the solvent-solvent energy $E_{ww}$ needs to be referenced.
 
-The Amber manual provides reference energies for several solvents.
-However, the exact reference value is different when using PME-GIST, and also depends on the box size.
-For exact analyses such as computing the total $\dgsolv$ of a compound, it is important to compute an exact reference energy.
-
-The most accurate method to compute a reference energy is to run a separate GIST calculations of the solvent without solute using the same energy method (PME/GPU) and a similar box size, and then compute the average solvent-solvent energy per molecule.
+The Amber manual provides reference densities and energies for several solvents.
+However, the exact energy values are different when using PME-GIST, and also depend on the box size.
+For quantitative analyses, it is recommended to compute an exact reference energy.
+
+The most accurate method to compute a reference energy is to run a GIST calculation of the pure solvent using the same energy method (PME/GPU) and a similar box size, and then compute the average solvent-solvent energy per molecule.
 
@@
 
-Alternatively, one can note that the solvent-solvent energy in a sufficiently large GIST grid converges to the correct bulk value at large distance to the solute.
-This means that the reference energy can be obtained from a large GIST grid by binning the voxels by their distance to the solute, and evaluate Equation~\ref{eq_ewwref} within each bin.
-If this value converges to a constant, this can be used as $E_{ww}^\textit{ref}$.
+Alternatively, the solvent-solvent energy in a large GIST grid converges to the bulk value at sufficient distance to the solute.
+Therefore, the reference energy can be obtained by binning the voxels by their distance to the solute, and evaluate Equation~\ref{eq_ewwref} within each bin.
+If this value converges, this can be used as $E_{ww}^\textit{ref}$.
 
 \software{gisttools} contains functionality to perform this analysis easily.
-There is also a method \inlinecode{detect\_reference\_value} that automatically tries to find the converged value from the rdf.
-Although it has a simple convergence check built-in, it is always recommended to check the convergence by hand as well.
+There is also a method \inlinecode{detect\_reference\_value} that automatically tries to find the converged value.
+Although it has a simple convergence check built-in, it is always recommended to check the convergence by hand.
 
 \begin{lstlisting}[style=python]
-import gisttools as gt
+from gisttools.gist import load_gist_file
 import matplotlib.pyplot as plt
-gist_obj = gt.gist.load_gist_file(
+gist = load_gist_file(
     "gist.dat", struct="solute-centered.pdb")
-bins, eww = gist_obj.rdf(
-    "Eww_unref_norm", bins=100, rmax=20., normalize="norm")
+bins, eww = gist.rdf(
+    "Eww_unref_norm", bins=100, rmax=20.,
+    normalize="norm")
 plt.plot(bins, eww)
-eww_auto = gist_obj.detect_reference_value().values[0]
+eww_auto = gist.detect_reference_value()
 plt.axhline(eww_auto, color='k')
@@
 	\includegraphics[width=0.8\linewidth]{figures/Eww_convergence.eps}
-	\caption{Convergence of $E_{ww}$ with increasing distance to the solute.
-The horizontal line shows the automatically computed reference energy.
-Computed from the complex calculation.}\label{fig_ewwref}
+	\caption{Convergence of $E_{ww}$ in the complex calculation with increasing distance to the solute (streptavidin-biotin).
+The horizontal line shows the automatically computed reference energy.}\label{fig_ewwref}
 \end{figure}
@@
 In \software{gisttools}, this is done simply by setting \inlinecode{gist.eww\_ref}.
-In \software{gistpp}, this can be done as follows: 
-\todo{Is there some official documentation for gistpp? We should probably link that somewhere, I feel that the code presented here is not really helpful/legible without further documentation. -vah}
-\todo{This might become invalid with future GIST versions, depending on whether we adopt the COM as molecular center.} 
-\todo{Better use the population.}
-
-\begin{lstlisting}[style=bash]
-rho0=0.03287465
-neg_eww_ref=9.533
-gistpp -op multconst -i gist-gO.dx -opt const $rho0 \
-    -o gist-gO-abs.dx 
-gistpp -op div -i gist-Eww-dens.dx -i2 gist-gO-abs.dx \
-    -o gist-Eww-norm.dx
-gistpp -op addconst -i gist-Eww-norm.dx \
-    -opt const $neg_eww_ref -o gist-Eww-ref-norm.dx
-gistpp -op mult -i gist-Eww-ref-norm.dx \
-    -i2 gist-gO-abs.dx -o gist-Eww-ref-dens.dx
-\end{lstlisting}
+%In \software{gistpp}, this can be done as follows: 
+%\todo{Is there some official documentation for gistpp? We should probably link that somewhere, I feel that the code presented here is not really helpful/legible without further documentation. -vah}
+%\todo{This might become invalid with future GIST versions, depending on whether we adopt the COM as molecular center.} 
+%\todo{Better use the population.}
+
+%\begin{lstlisting}[style=bash]
+%rho0=0.03287465
+%neg_eww_ref=9.533
+%gistpp -op multconst -i gist-gO.dx -opt const $rho0 \
+%    -o gist-gO-abs.dx 
+%gistpp -op div -i gist-Eww-dens.dx -i2 gist-gO-abs.dx \
+%    -o gist-Eww-norm.dx
+%gistpp -op addconst -i gist-Eww-norm.dx \
+%    -opt const $neg_eww_ref -o gist-Eww-ref-norm.dx
+%gistpp -op mult -i gist-Eww-ref-norm.dx \
+%    -i2 gist-gO-abs.dx -o gist-Eww-ref-dens.dx
+%\end{lstlisting}
 
 Now, add all free energy contributions to obtain the spatially resolved $\dgsolv$.
-In \software{gisttools}, this is done automatically (simply use the \inlinecode{A\_dens} and \inlinecode{A\_norm} columns).
-In \software{gistpp}, use the \inlinecode{add} command, with the same syntax as the \inlinecode{div} command used above.
-After that, check whether your free energy contributions become negligible at high distances to the solute.
+In \software{gisttools}, this is done automatically, and can be accessed using the \inlinecode{A\_dens} and \inlinecode{A\_norm} data rows. \todo{norm and dens not defined. }
+
+\newcommand{\coordinate}{\mathbf{r}}
+\begin{equation}
+	\dgsolv(\coordinate) = \Delta E_{ww}(\coordinate) + \Delta E_{sw}(\coordinate) - T\Delta S^\text{six}(\coordinate)
+\end{equation}
+
+%In \software{gistpp}, use the \inlinecode{add} command, with the same syntax as the \inlinecode{div} command used above.
+After that, check whether your free energy contributions become negligible far away from the solute.
 In a plot like Figure~\ref{fig_ewwref} based on the \inlinecode{A\_dens} column, the values should tend to zero.
@@
 It the curve flattens out, the converged value is your final $\dgsolv$\@.
-If it does not, you might need to tweak the $E_{ww}$ reference, or introduce a reference value for the entropy (especially with older versions of \software{cpptraj}).
-% \todo{Eww or E$_{ww}$? Doesn't seem consistent so far. -vah} - let's go with all math mode $E_{ww}$ -fwa
+Otherwise, you might need to tweak the $E_{ww}$ reference, or introduce a reference value for the entropy.
 
 \begin{lstlisting}[style=python]
-import gisttools as gt
+from gisttools.gist import load_gist_file
 import matplotlib.pyplot as plt
 # adapt eww_ref!
-gist = gt.gist.load_gist_file("gist.dat",
+gist = load_gist_file("gist.dat",
     struct="solute-centered.pdb", eww_ref=-9.533)
@@
 The expected output of this code is shown in Figure~\ref{fig_radial_convergence}.
-Both the solvent-solvent energy and the entropy are not perfectly flat, indicating that the reference value is not optimal.
-The slow energy convergence indicates that a non-optimal reference value was used, but is also complicated by the fact that a charged solute is used.
-In this case, both the entropy and $E_{sw}$ can be considered converged around \SI{10}{\angstrom}.
+Especially the solvent-solvent energy is not perfectly flat, indicating that the reference value is not optimal.
+%but is also complicated because the solute is charged.
+Both the entropy and $E_{sw}$ can be considered converged around \SI{10}{\angstrom}.
 While this is not true for $E_{ww}$, we will see below that the energy difference upon binding converges more readily than its individual contributions.
@@
 To include the effect of the lid closing transition in the binding process, free energy calculation methods such as umbrella sampling can be used.
+
+Smaller GIST grids could be used by focusing on the binding pocket, or by rotating the molecule by its principal axes to fit the cuboid grid more exactly.
+However, this should be done \emph{before} the MD, since rotating the trajectory damages the periodic box information.
 %\section{Content and links}
@@
 
+\begin{checklist}{Handling convergence problems}
+\begin{itemize}
+	\item Insufficient sampling leads to high noise in the results. Use higher numbers of frames.
+	\item Autocorrelation between frames leads to negative entropy in bulk. Increase simulation time and/or reduce the number of frames.
+\end{itemize}
+\end{checklist}
+
 %\begin{checklist}{A list}

### 2022-12-05 manuscript/manuscript.tex
Fleshed out biotin chapter a bit

@@
 		gridcntr    & <y> <y> <z> & Coordinates of the grid center in \AA & 0 0 0  \\
-		griddim     & <nx> <ny> <nz> & Number of voxels per axis & 40 40 40\\
+		griddim     & <nx> <ny> <nz> & Number of voxels per axis & 40 40 40\pdfcomment{Overfull hbox?}\\
 		gridspacn & <dval> &Grid spacing / voxel dimensions in  \AA & 0.5    \\
@@
 
+Energy and entropy quantities are reported in two different ways in the GIST output.
+The \inlinecode{\_dens} quantities are normalized by the voxel volume and can be used to compute integrals.
+The \inlinecode{\_norm} quantities are instead normalized by the average number of solvents in the respective voxel, and are useful to visualize local water properties.
+
 \section{Tutorial}
@@
 Even when running the simulation in NVT, it is advisable to include an NpT equilibration step to adjust the box size beforehand.
-Use enough equilibration time to allow the water to relax around your biotin structure.
+Use enough equilibration time to allow the solvent to relax. 
+For water, \SI{0.5}{\nano\second} are usually sufficient to equilibrate the density, but narrow binding pockets can take longer.
 
@@
 \todo{Move simulation part from complex to here}
-
-\subsection{Applying GIST for the Streptavidin-Biotin complex}
-\subsubsection{System Preparation and Equilibration}
-If you want to skip this section, run 
-\begin{lstlisting}[style=bash]
-make equilibration-targets
-\end{lstlisting}
-in the code folder. This assumes that you have Python and Amber set up properly and that \software{pmemd.cuda} is in the PATH.
-
-The choice of initial structure is crucial to the GIST results.
-Crystal structures may not represent the dominant solution structure for a given target.
-In our example, we only use a single equilibrated structure.
-A more sophisticated approach would be to identify one or several dominant structures out of the ensemble.
-Depending on what properties you are interested in, it might be beneficial to run an MD simulation choose dominant structures through clustering.
-
-When starting from a crystal structure, care has to be taken during the preparation step.
-Make sure to choose a proper protonation state for all titratable moeties.
-You will need to solvate your structure in a solvent box, which should be large enough to obtain unbiased long-range electrostatics.
-A good criterion for validating the box size is to check whether the water on the outside of the box behaves bulk-like according to the GIST output.
-
-When calculating solvation free energy differences between a complex and the corresponding monomers (the dissociated state), there are two options for dealing with configurational diversity in the structures.
-One can choose to sample the complex and dissociated states individually, or to sample only one of them and assume that there is sufficient overlap in the conformations.
-
-Here, we do not include any sampling of the conformational states.
-It is therefore important that the molecular conformations in the complex and dissociated states match exactly.
-Use \software{cpptraj} to produce biotin and streptavidin structures based on the complex by stripping the respective other molecule like this:
-\begin{lstlisting}[style=cpptraj]
-parm complex/solvated.parm7
-trajin complex/equil/EQUIL-DONE.rst
-strip ^2 parmout streptavidin/solvated.parm7
-trajout streptavidin/pre-equil.ncrst
-\end{lstlisting}
-(Adapt for biotin as necessary.)
-%We provide a script called \software{cpptraj\_remove\_mol.sh} that removes a single molecule from a topology and structure using \software{cpptraj} without altering the rest of the system.
-\todo{Explain cpptraj rather than our script.
-The script is for the automated workflow only.}
-
-Use the \software{equilibration.py} script (or your own) to perform short NVT and NpT equilibration runs on the complex structure.
-Note that the equilibration script contains unrestrained NpT steps, because restraining two molecules at the same time might interfere with the pressure equilibration.
-
-Then split the equilibrated structure into biotin and streptavidin parts using \software{cpptraj\_remove\_mol.sh}.
-After that, you can use \software{equilibration.py -R} to run short restrained 
-equilibrations (where only the water is allowed to move) on the individual 
-systems.
-
-\subsubsection{Running MD for GIST}
-
 Now, you can run restrained MD simulations based on the equilibrated structures for subsequent GIST analyses.
@@
 \end{lstlisting}
-Note that these are very standard MD settings except for the restraints.
+This file will run a 100 ns NpT MD simulation while keeping all atoms except for hydrogens and waters restrained. Note that these are very standard MD settings except for the restraints.
 
@@
 
-\subsection{GIST analysis of Streptavidin-Biotin}
+\subsection{Applying GIST for the Streptavidin-Biotin complex}
+\subsubsection{System Preparation and Equilibration}
+If you want to skip this section, run 
+\begin{lstlisting}[style=bash]
+make equilibration-targets
+\end{lstlisting}
+in the code folder. This assumes that you have Python and Amber set up properly and that \software{pmemd.cuda} is in the PATH.
+
+The choice of initial structure is crucial to the GIST results.
+Crystal structures may not represent the dominant solution structure for a given target.
+In our example, we only use a single equilibrated structure.
+A more sophisticated approach would be to identify one or several dominant structures out of the ensemble.
+Depending on what properties you are interested in, it might be beneficial to run an MD simulation choose dominant structures through clustering.
+
+When starting from a crystal structure, care has to be taken during the preparation.
+Make sure to choose a proper protonation state for all titratable moeties.
+You will need to solvate your structure in a solvent box, which should be large enough to obtain unbiased long-range electrostatics.
+A good criterion for validating the box size is to check whether the water on the outside of the box behaves bulk-like according to the GIST output.
+
+When calculating solvation free energy differences between a complex and the corresponding monomers (the dissociated state), there are two options for dealing with configurational diversity in the structures.
+One can choose to sample the complex and dissociated states individually, or to sample only one of them and assume that there is sufficient overlap in the conformations.
+
+Here, we do not include any sampling of the conformational states.
+It is therefore important that the molecular conformations in the complex and dissociated states match exactly.
+Use \software{cpptraj} to produce biotin and streptavidin structures based on the complex by stripping the respective other molecule like this:
+\begin{lstlisting}[style=cpptraj]
+parm complex/solvated.parm7
+trajin complex/equil/EQUIL-DONE.rst
+strip ^2 parmout streptavidin/solvated.parm7
+trajout streptavidin/pre-equil.ncrst
+\end{lstlisting}
+(Adapt for biotin as necessary.)
+%We provide a script called \software{cpptraj\_remove\_mol.sh} that removes a single molecule from a topology and structure using \software{cpptraj} without altering the rest of the system.
+\todo{Explain cpptraj rather than our script.
+The script is for the automated workflow only.}
+
+Use the \software{equilibration.py} script (or your own) to perform short NVT and NpT equilibration runs on the complex structure.
+It is important to include unrestrained NpT steps, because restraining two molecules at the same time might interfere with the pressure equilibration.
+
+Then split the equilibrated structure into biotin and streptavidin parts using \software{cpptraj\_remove\_mol.sh}.
+After that, use \software{equilibration.py -R} to run short restrained 
+equilibrations (where only the water is allowed to move) on the individual 
+systems.
+By using restraints, we can make sure that the monomer conformations match the complex.
+
+\subsubsection{MD Simulation}
+The MD can be run as described above for each monomer and the complex.
+
+\subsubsection{GIST analysis of Streptavidin-Biotin}
 For easier post-processing, here we use grids that contain the whole molecule.
@@
 
-Now, you can finally run the actual GIST analysis.
+Now, you can finally run the GIST analysis.
 Center the molecule as above, to make sure that its position matches the grid parameters you decided on.
@@
 
-\subsection{Reference Values and Radial Convergence}
+\subsubsection{Reference Values and Radial Convergence}
 All GIST quantities should be expressed relative to bulk.
-This automatically done for the entropy, using the bulk density specified by \inlinecode{refdens}.
+This is automatically done by \software{cpptraj} for the entropy, using the bulk density specified by \inlinecode{refdens}.
 The solute-solvent energy $E_{sw}$ naturally tends towards zero in bulk.
@@
 
-\software{gisttools} contains functionality to perform this analysis easily.
+\software{gisttools} contains functions to facilitate this analysis.
 There is also a method \inlinecode{detect\_reference\_value} that automatically tries to find the converged value.
@@
 
-Now, add all free energy contributions to obtain the spatially resolved $\dgsolv$.
-In \software{gisttools}, this is done automatically, and can be accessed using the \inlinecode{A\_dens} and \inlinecode{A\_norm} data rows. \todo{norm and dens not defined. }
+Now, sum all free energy contributions to obtain the spatially resolved $\dgsolv$.
+In \software{gisttools}, this is done automatically and can be accessed using the \inlinecode{A\_dens} and \inlinecode{A\_norm} data rows.
 
@@
 While this is not true for $E_{ww}$, we will see below that the energy difference upon binding converges more readily than its individual contributions.
-\todo{vah: Is that shown somewhere yet?}
+\todo{vah: Is that shown somewhere yet -> fwa?}
 
@@
 
-\subsection{Visualizing $\dgsolv$}
+\subsubsection{Visualizing $\dgsolv$}
 Next, we want to visualize the 3-dimensional contributions to the free energy of hydration, as well as the entropy and enthalpy contributions.
@@
 It might also be interesting to visualize the average energy of a water molecule at each grid voxel.
-This quantity is given by 2*Eww + Esw.
+This quantity is given by $2E_{ww} + E_{sw}$.
 The energy referencing leads to non-zero normalized values where the population is zero.
@@
 
-Visualize the free energy, the total energy, and the entropy, at several iso-levels.
+Visualize the free energy, the energy, and the entropy, at several isolevels.
 
@@
 
-You should find that there are several regions in the binding pocket that feature a high solvent energy and/or low solvent entropy.
+You should find that there are several regions in the binding pocket that feature a high solvent energy and/or low entropy.
 However, such regions also exist outside of the binding pocket, and there are also low-energy regions within the binding pocket.
-Therefore, the strong affinity to biotin can not be explained solely by the hydration contribution, and the direct interaction energy also needs to be taken into account.
-
-\subsection{Contribution of Hydration to Binding}
-In the next step, calculate the $\dgsolv$ contributions around the biotin ligand in each of the three systems (biotin, streptavidin, and complex).
-Note that our integration region does not fully reach into bulk, but also comprises regions that are close to the streptavidin protein. 
+Therefore, the strong affinity to biotin cannot be explained only by hydration, and the direct interaction energy also needs to be taken into account.
+
+\subsubsection{Contribution of Hydration to Binding}
+In the next step, calculate the $\dgsolv$ contributions around biotin in each system (biotin, streptavidin, and complex).
+Note that our integration region does not fully reach into bulk, and some of the boundary regions are close to streptavidin. 
 Therefore, our results for each system will depend on the exact position of the integration boundary. 
@@
 Although we kept the systems rigid, the center of mass might have been shifted during pressure equilibration.
-Therefore, we align the complex structure to the streptavidin system and use the shifted biotin coordinates to define the integration region.
-
-We recommend using a \software{Jupyter Notebook} for the following analyses, although they can also be done in a script or an interactive Python prompt.
+Therefore, we align the complex structure to streptavidin and use the shifted biotin coordinates to define the integration region.
+
+We recommend using a \software{Jupyter Notebook} for the following analyses.
 The code is not explained in detail here, but should be clear after reading the previous sections.
@@
 
-Subtract a reference entropy from the dTSsix columns.
-\todo{Is there actually a reason why this is not in gisttools itself implemented? -vah}
+Subtract a reference entropy from the \inlinecode{dTSsix} columns.
+\todo{Update with newest gisttools -> fwa}
 \begin{lstlisting}[style=python]
@@
 For the streptavidin integral, we align the complex structure onto streptavidin and then use the biotin positions as centers.
-Note that we use \software{mdtraj} here, since \software{gisttools} stores the reference structure as a \inlinecode{mdtraj.Trajectory} object, but any other library for post-processing of trajectories works as well.
+Note that we use \software{mdtraj} here, since \software{gisttools} stores the reference structure as an \software{mdtraj} \inlinecode{Trajectory} object.
 \begin{lstlisting}[style=python]
@@
 compl_x = select(compl.struct, biotin_mask).xyz[0] * 10.
-biotin_x = select(biotin.struct, biotin_mask).xyz[0] * 10.
+biotin_x = select(biotin.struct, biotin_mask).xyz[0]*10.
 aligned = compl.struct[:].superpose(strept.struct, \
@@
 Compute the energy (\inlinecode{Eall}) and entropy (\inlinecode{dTSsix}) contributions separately.
-You will find that the energy is strongly unfavorable to binding, because we do not yet include the interaction energy between biotin and streptavidin.
-
-Using the \inlinecode{energy} command in \software{cpptraj}, you can compute this energy difference.
+You will find that the energy disfavors binding, because we do not yet include the interaction energy between biotin and streptavidin.
+
+You can also plot the difference between the complex and monomer contributions against the distance to biotin, using the complex coordinates for the holo structure.
+The expected result is shown in Figure~\ref{fig-dg-sums}.
+
+\begin{figure}
+	\centering
+	\includegraphics[width=1.0\linewidth]{figures/deltaG-difference.png}
+	\caption{TODO caption}\label{fig-dg-sums}
+\end{figure}
+
+Using the \inlinecode{energy} command in \software{cpptraj}, you can compute this interaction energy.
 We recommend using PME in combination with PME-GIST, but not with GPU-GIST\@.
@@
 
-It has been shown in literature \cite{Chen2021,Waibl2022-gist-solvents} that the total solvation entropy in water is best approximated by 0.6 times the first order entropy.
+It has been shown \cite{Chen2021,Waibl2022-gist-solvents} that the solvation entropy in water is best approximated by 0.6 times the first order entropy.
 Also compute the scaled entropy and check the effect on the binding affinity.
@@
 %\todo{Table 1 is placed quite akwardly at the moment for me, maybe enforce its position? -vah. Ok? -fwa}
-\todo{this table is super confusing to me. What's "total"? Doesn't seem to be E-TS. Diff is also not the sum of other total entries. Shouldn't it be TS instead of S?-vah}
+
 \begin{table}[h]
 	\caption{Free energy contributions for the monomers and the dimer in kcal$\cdot$mol$^{-1}$.
-``total'' is calculated as $E^\text{internal} + \Delta E^\text{GIST} - \Delta S^\text{scaled}$.}\label{tab_dg_monomers_dimer}
+``Diff'' is calculated as $E^\text{internal} + \Delta E^\text{GIST} - \Delta S^\text{scaled}$.}\label{tab_dg_monomers_dimer}
 	\small
@@
 		\toprule
-		System       & E(internal) & $\Delta E^\textit{GIST}$ & $\Delta S^\textit{GIST}$ & $\Delta S^\textit{scaled}$ & total \\
+		System       & E(internal) & $\Delta E^\textit{GIST}$ & $T\Delta S^\textit{GIST}$ & $T\Delta S^\textit{scaled}$ & Diff \\
 		\midrule
@@
 However, we do not take the conformational transition of the streptavidin binding pocket into account.
+
 Prior computational studies suggest a $\Delta G$ of \SI{-26.6}{\kilo\calorie\per\mole} for the binding of biotin into the closed conformation of streptavidin \cite{Bansal2018-biotin}.
 This indicates that our result is in good agreement with literature.
-However, the conformational changes of the binding pocket would need to be described to improve the agreement with experiments.
-\todo{Should we include a short introduction to other solvents with GIST? What about solvent/ion mixtures? Can we somehow distinguish python code from input files optically? -vah}
-\subsection{Further steps}
-In this tutorial, we obtained a value for the binding of biotin into the closed conformation of the streptavidin binging pocket.
+To improve the agreement with ITC measurements, the conformational changes of the binding pocket should be included.
+
+\subsubsection{Further steps}
+In this tutorial, we obtained a value for the binding of biotin into the closed conformation of the streptavidin binding pocket.
 To investigate the effect of different conformations on the binding affinity, you can perform molecular dynamics simulations of the complex and/or monomers and perform GIST on multiple cluster representatives.
-To include the effect of the lid closing transition in the binding process, free energy calculation methods such as umbrella sampling can be used.
-
-Smaller GIST grids could be used by focusing on the binding pocket, or by rotating the molecule by its principal axes to fit the cuboid grid more exactly.
-However, this should be done \emph{before} the MD, since rotating the trajectory damages the periodic box information.
-%\section{Content and links}
-%
-%A tutorial will normally draw on additional files and materials; clearly indicate where and how these are available, with links, and how they are being archived for the long-term and maintained so they stay current.
-%You will likely want to reference your GitHub repository as a central point to access all of this information, and then the GitHub repository may link out to other content as needed.
+To include the effect of lid closing in the binding process, free energy calculation methods such as umbrella sampling can be used.
+
+To speed up the calculation, smaller GIST grids could be used by focusing on the binding pocket, or by rotating the molecule by its principal axes to fit the cuboid grid more exactly.
+However, this should be done \emph{before} the MD, since rotating the trajectory damages the periodic box information \todo{@Dan: is this still the case?}.
 
@@
 \subsection{GPU implementation}
-In Reference~\cite{Kraml2020}, a GPU implementation of the energy calculation was implemented, which typically increases the speed by 1--2 orders of magnitude.
+In Reference~\cite{Kraml2020}, a GPU implementation of the energy calculation was presented, which typically increases the speed by 1--2 orders of magnitude.
 
 If \software{cpptraj} is compiled with GPU support, the energy calculation uses the GPU automatically, without any additional input.
-However, PME-GIST is not supported on the GPU.
+However, PME-GIST is not supported on the GPU, and no GPU code will be used when specifying \inlinecode{pme}.
 
@@
 
-To run a PME-GIST calculation, the \inlinecode{pme} flag to the \inlinecode{gist} action in \software{cpptraj} can be used.
+To run PME-GIST, the \inlinecode{pme} flag to \inlinecode{gist} can be used in \software{cpptraj}.
 The output will contain two additional columns: \inlinecode{PME\_dens} and \inlinecode{PME\_norm}.
@@
 Furthermore, the \inlinecode{Eww} and \inlinecode{Esw} columns are also computed using PME, and are reported as usual.
-For typical use cases, it is recommended to use \inlinecode{Eww} and \inlinecode{Esw} over the \inlinecode{PME} column.
+For typical use cases, we recommend using \inlinecode{Eww} and \inlinecode{Esw} over the \inlinecode{PME} column.
 
@@
 
-To run a GIST calculation in a solvent other than water, use the \inlinecode{solute} flag together with a \software{cpptraj} selection mask.
-Everything that is not in the \inlinecode{solute} mask is treated as solvent.
-If there are multiple solvent species, use the \inlinecode{solventmols} option as described in Section~\ref{sec-salt-water}.
-The code will automatically choose three atoms which are used to define the solvent orientation, and print this selection to the output.
-
-Sometimes, \software{cpptraj} does not automatically choose the optimal atoms.
-In methanol, for instance, one can either incorporate the C--H bonds or the O--H bond in the selection.
+To run a GIST calculation in a solvent other than water, use \inlinecode{solute} with a \software{cpptraj} selection mask.
+All other molecules will be treated as solvent.
+If there are multiple solvent species, also use the \inlinecode{solventmols} flag as described in Section~\ref{sec-salt-water}.
+
+The code will automatically choose three atoms per molecule to define the solvent orientation, and print this selection to the output.
+Sometimes, however, the automatic selection is not optimal.
+For instance, in methanol one can either incorporate the C--H bonds or the O--H bond in the selection.
 The latter is probably more relevant since it can incorporate hydrogen bonding effects.
@@
 In \cite{Waibl2021-gist-salt}, an extension of GIST was presented that can use salt-water mixtures as a solvent.
-This allows treating salting-out effects, although it was shown that the salting-out coefficient is over-predicted by GIST unless higher-order entropies are incorporated.
+This allows treating salting-out effects, although it was shown that the salting-out coefficient is over-predicted by GIST because of the first-order entropy approximation.
 Generally, GIST should be able to treat arbitrary solvent mixtures as long as each molecule is sufficiently rigid.
 
-The \software{cpptraj} implementation of GIST allows the user to specify which components of the system to treat as a solute, using the \inlinecode{solute} keyword.
+In the \software{cpptraj} implementation of GIST, the \inlinecode{solute} keyword specifies which components of the system are treated as a solute.
 Everything else will be treated as solvent.
@@
 
-\subsection{Loading a DX file in Pymol }
+\subsection{Visualizing DX files}
 A minimal PyMOL input to visualize a dx file is shown here.
@@
 The oxygen density is visualized at an isolevel of 2 (i.e., twice the reference density).
+The expected output is shown in Figure~\ref{fig-streptavidin_gO}.
 
@@
 	\includegraphics[width=1.0\linewidth]{figures/streptavidin_gO_high_surf.png}
-	\caption{Oxygen density around strepativin at twice the reference bulk density}\label{streptavidin_gO}
+	\caption{Oxygen density around streptavidin at an isolevel of twice the reference density}\label{fig-streptavidin_gO}
 \end{figure}
@@
 GIST provides a very natural way of creating a SASA using an isosurface of the oxygen density gO.
-It can be visualized in PyMOL as follows:
+It can be visualized in PyMOL.
+The expected output is shown in Figure~\ref{fig-streptavidin_sasa}.
 
@@
 	\includegraphics[width=1.0\linewidth]{figures/sasa-simple.png}
-	\caption{SASA for streptavidin focused on the binding pocket}\label{streptavidin_sasa}
+	\caption{SASA for streptavidin focused on the binding pocket}\label{fig-streptavidin_sasa}
 \end{figure}
 
-In the above figure, the SASA was generated from a simulation of the protein in the absence of the ligand.
+Here, the SASA was generated from a simulation of the holo protein structure.
 When the ligand is placed in the binding pocked, all atoms are within the SASA.
 
-\subsection{Region around the binding pocket}
+\subsection{Defining the binding pocket}
 Specific regions in a protein can be defined to calculate the thermodynamic quantities for that region.
@@
 The resulting file is a binary file and can be visualized using \software{pymol}.
+\todo{Not GISTPP}
 
@@
 
-In this example, the region defined in the binding pocket includes all voxels that are within \SI{3.5}{\angstrom} of the heavy atoms of biotin.
+Here, the binding pocket includes all voxels within \SI{3.5}{\angstrom} of the heavy atoms of biotin.
 
@@
 GIST is an implementation of inhomogeneous fluid solvation theory (IST) \cite{Lazaridis1998} that discretizes the free energy of solvation $\dgsolv$ on a three-dimensional grid. 
-It was first devised by Nguyen et al.
-\cite{Nguyen2012} and its implementation in \software{cpptraj} was thoroughly described by Ramsey et al.
-\cite{Ramsey2016}.
-% to overcome the limitation of IST which was commonly used for regions of high water occupancy and its implementation in \software{cpptraj} was thoroughly described in \cite{Ramsey2016}.
+It was first devised by Nguyen et al.\ \cite{Nguyen2012} and its implementation in \software{cpptraj} was thoroughly described by Ramsey et al.\ \cite{Ramsey2016}.
 Here, we only provide a short overview of the theory behind GIST.
@@
 \subsection{Solvation Entropy}
-
-%The solvation entropy of a system is totality of the contribution of the  solute-water entropy and the water-water entropy. -- this is inconsistent with IST theory (fwa)
 
@@
 
-The solvent density is expressed relative to the bulk density $\rho^\textit{o}$. %$g_\textit{sw}\left(\textbf{r},\omega \right)$ is the solute-water 
+The solvent density is expressed relative to the bulk density $\rho^0$. %$g_\textit{sw}\left(\textbf{r},\omega \right)$ is the solute-water 
 %pair correlation function in the solute's frame of reference, \textbf{r} is the coordinates of the oxygen atom of the water molecule,  
@@
 The entropy integral is also normalized by $8\pi^2$, which is the volume of the orientational space.
-In bulk or at a high distance to the solute, the quantity $g_\textit{sw}\left(\textbf{r},\omega \right)$ is unity which leads to the first-order entropy of 
-bulk to be zero. 
-Therefore, no reference entropy needs to be computed unless there is a numeric bias, e.g., when the MD frames are not statistically independent due to insufficient spacing.
-%Also, this quantity reaches unity as the distance from the solute molecule increases which leads for the calculation of 
-%the solvation entropy to be an approximation of the integral around the solute atom. 
-\begin{equation}
-	\Delta S_\textit{solv}
+In bulk or at a high distance to the solute, the quantity $g_\textit{sw}\left(\textbf{r},\omega \right)$ approaches one, which leads to zero first-order entropy. 
+Therefore, no reference entropy is needed unless there is a numeric bias, e.g., when the MD frames are statistically dependent because the time between them is insufficient.
+\begin{equation}
+	\Delta S_\text{solv}
 	\approx \Delta S_\textit{sw}
 	\newcommand{\gro}{g_{sw}(\textbf{r},\omega)}
-	\equiv -R \frac{\rho^\textit{0}}{8\pi^\textit{2}} \int \gro \ln\gro d\textbf{r}d\omega
-\end{equation}
-\todo{We are inconsistent regarding R/kB; vah: Fixed. is it - tho? Papers seem to disagre }
-
-The solvation entropy is further broken down in terms of the contribution from translational and orientational terms.
+	\equiv -R \frac{\rho^0}{8\pi^\textit{2}} \int \gro \ln\gro d\textbf{r}d\omega
+\end{equation}
+
+\subsubsection{Entropy Calculations in Cpptraj}
+In \software{cpptraj}, the calculation of solvation entropy is handled by two methods.
+
+In the first method, the solvation entropy is broken down into translational and orientational contributions.
 \begin{equation}
@@
 \end{equation}
-
-\subsection{Entropy Calculations in Cpptraj}
-In \software{cpptraj} the calculation of  solvation entropy is handled by two methods.
-
-The first method assumes that the distribution of the orientation $\omega$ is constant within a voxel $k$.
-This allows for the splitting of the contributions of the translational and the orientational entropy. 
+This is exact assuming that the distribution of the orientation $\omega$ is constant within a voxel $k$, but might converge slower than computing the entropy from the combined space (see below).
+
+The second method directly calculates the solvation entropy by evaluating the six-dimensional integral (3 for position and 3 for orientation).
+
+\begin{equation}
+g_\textit{vox} \left( \textbf{r}, \omega \right) \approx g_\textit{vox}(\textbf{r}) g_\textit{vox}(\omega)
+\end{equation}
+
 A first nearest neighbor (NN) approach is used to evaluate the density.
-%The translational density $g_{NN}(\textbf{r})$ is expressed in reference to a homogeneous bulk solvent at the reference density, while the orientational density $g_{NN}(\omega)$ is expressed to a random placement of the number of solvents found in this voxel.
-Both the translational ($g_{NN}(\textbf{r})$) and orientational ($g_{NN}(\omega)$) density are computed relative to a bulk solvent at the reference density $\rho^0$.
+The densities are computed relative to a homogeneous distribution of solvent molecules, given a bulk density $\rho^0$.
 
@@
 \begin{equation}
-	S_{k}^\textit{trans} = -R \left( \gamma + \frac{1}{N_\textit{k}} \sum _{i=1}^{N_k} \ln g_{NN, \textit{i}}(\textbf{r}) \right)
-\end{equation}
-and
+	S_{k}^\textit{trans} = -R \left( \gamma + \frac{1}{N_\textit{k}} \sum _{i=1}^{N_k} \ln g_{NN, \textit{i}}(\textbf{r}) \right),
+\end{equation}
 \begin{equation}
@@
 \end{equation}
-where $N_\textit{k}$ is the number of water molecules found in voxel k, $\gamma$ is Euler's constant accounting for the bias of the naive entropy estimator, and $g_\textit{NN}$ is the nearest neighbor estimate.
-
-The second method directly calculates the solvation entropy by evaluating the six-dimensional integral (3 for position and 3 for orientation) 
-using the nearest-neighbor approach. 
-% Unlike the first method, the nearest neighbor estimate for the sixth-order entropy considers all the water found within the same and neighboring voxels.
--- misleading because this is also done for S_trans (fwa)
-
-\begin{equation}
-	g_\textit{vox} \left( \textbf{r}, \omega \right) \approx g_\textit{vox}(\textbf{r}) g_\textit{vox}(\omega)
-\end{equation}
-
-\begin{equation}
-	S_\textit{k} = -R \left( \gamma + \frac{1}{N_\textit{k}} \sum _{i=1}^{N_k} \ln g_{NN, \textit{i}}(\textbf{r},\omega) \right)
-\end{equation}
+and
+\begin{equation}
+S_\textit{k}^\text{six} = -R \left( \gamma + \frac{1}{N_\textit{k}} \sum _{i=1}^{N_k} \ln g_{NN, \textit{i}}(\textbf{r},\omega) \right),
+\end{equation}
+where $N_\textit{k}$ is the number of water molecules found in voxel k, $\gamma$ is Euler's constant accounting for the bias of the naive entropy estimator, and $g_\textit{NN}$ is the nearest neighbor estimate of the density.
 
 \subsection{Solvation Energy}
-The solvation energy is readily calculated by summing up the contributions of the water-water interaction and the water-solute interaction from the force field.
+The solvation energy is calculated from the water-water and water-solute interactions from the force field.
 \begin{equation}
@@
 To localize $E_{sw}$ on the three-dimensional grid, every energy term is assigned to the voxel $k$ holding the solvent, and the expectation value in Equation~\ref{eq-esw} is computed per voxel.
-Similar to the entropy integrals, the solute-water solvation energy integral also decays with increasing distance from the solute.
-Hence, the solvation energy can be approximated by local spatial integrals.
-
-The solvent-solvent energy $E_{ww}$ is computed in a similar manner.
-Again, it can be expressed in terms of density functions, but is practically computed as a sum over solvent-solvent interactions where one solvent is in voxel $k$.
+Similar to the entropy integrals, the solute-water solvation energy decays with increasing distance to the solute.
+Hence, it can be approximated by local spatial integrals.
+
+The solvent-solvent energy $E_{ww}$ is computed similarly.
+It can also be expressed in terms of density functions, but is practically computed as a sum over solvent-solvent interactions per voxel $k$.
 In contrast to $E_{sw}$, $E_{ww}$ does not tend to zero in bulk.
-Therefore, a reference energy corresponding to the average energy of a water molecule in bulk needs to be subtracted.
-
-%However, there is a complication arising from this approximation.
-Water molecules outside the region of interest will also be accounted for in the pairwise water term.
-For example, given two different regions $R_a$ and $R_b$, will not sum up to the total energy of the two regions due to double counting.
+Therefore, a reference corresponding to the average energy of a bulk water molecule needs to be subtracted.
+The referenced solvent-solvent energy will be denoted as $E_{ww}^\text{corr}$.
 
@@
 	\begin{aligned}
-		\Delta E_{ww}(\textbf{r}) \equiv & \left(\frac{1}{8\pi^2} \right)^2 \rho^o \int g_\textit{sw}(\omega|\textbf{r}) \\
-		& \times \left[g_\textit{sw}(\textbf{r}^\prime , \omega^\prime) - g^o_\textit{ww} (\textbf{r}, \omega, \textbf{r}^\prime, \omega^\prime )\right] \\
-		& \times U_\textit{ww}(\textbf{r}, \omega, \textbf{r}^\prime, \omega^\prime)d\omega d\textbf{r}^\prime d\omega^\prime \\
-		\Delta E_{ww,k} =& \frac{1}{N_k} \left\langle \sum_i^{\text{solvent},k} \; \sum_{j \neq i}^\text{solvent} U_{ij}\right\rangle - \left\langle E_{ww}^\text{bulk}\right\rangle
+		\Delta E_{ww}^\text{corr}(\textbf{r}) \equiv & \left(\frac{1}{8\pi^2} \right)^2 \rho^o \int g_{sw}(\omega|\textbf{r}) \\
+			& \times \left[g_{sw}(\textbf{r}^\prime , \omega^\prime) - g^0_{ww} (\textbf{r}, \omega, \textbf{r}^\prime, \omega^\prime )\right] \\
+			& \times U_{ww}(\textbf{r}, \omega, \textbf{r}^\prime, \omega^\prime)d\omega d\textbf{r}^\prime d\omega^\prime \\
+		\Delta E_{ww,k}^\text{corr} =& \frac{1}{N_k} \left\langle \sum_i^{\text{solvent},k} \; \sum_{j \neq i}^\text{solvent} U_{ij}\right\rangle - \left\langle E_{ww}^\text{bulk}\right\rangle
 	\end{aligned}
@@
 
-%AFAIK, this was never done, since the pair interaction matrix Eij would have to be written out.
-%There are two ways in which this double counting was accounted for in the original paper.
-First, using the water-displacement method.
-This method is applicable when a determining the energetic cost of kicking out a water molecule.
-%\begin{equation}
-%	\Delta E^R_\textit{ww} = n^\textit{R} E^\textit{bulk}_\textit{ww} - E^\textit{R,corr}_\textit{ww}
-%\end{equation}
-
-%The second method uses normalized-water properties to account the difference in water properties of two different regions.
-The normalized water property is defined as the water property per water molecule.
-%\begin{equation}
-%	\Delta E^\textit{R,norm}_\textit{ww} = E^\textit{R,norm}_\textit{ww} - 2E^\textit{bulk}_\textit{ww}
-%\end{equation}
-
+\subsubsection{Interpreting the energy values}
 When computing the total $E_{ww}$ of the system, double-counting of interactions must be avoided.
@@
 \begin{equation}
-	\Delta E^R_\textit{ww} = n^\textit{R} E^\textit{bulk}_\textit{ww} - E^\textit{R,corr}_\textit{ww}
+	\Delta E^\text{total}_{ww} = \sum_{k}^\text{voxels} \frac{E_{ww,k}^\text{corr}}{2}
+%	\Delta E^R_{ww} = n^R E^\text{bulk}_{ww} - E^\textit{R,corr}_{ww}
 \end{equation}
@@
 \begin{equation}
-	\Delta E^\textit{R,norm}_\textit{ww} = E^\textit{R,norm}_\textit{ww} - 2E^\textit{bulk}_\textit{ww}
-\end{equation}
-
-When the region of interest $R$ comprises more than one molecule, interactions within this region are double-counted while interactions to the outside are not.
-To take this into account, the solvent-solvent energy between each pair of voxels $E_{ww,kl}$ is needed.
+	\Delta E^\text{total}_{ww} = \sum_{k}^{\text{voxels in }R} E_{ww,k}^\text{corr}
+%	\Delta E^\textit{R,norm}_\textit{ww} = E^\textit{R,norm}_\textit{ww} - 2E^\textit{bulk}_\textit{ww}
+\end{equation}
+
+When the region $R$ comprises more than one molecule, interactions within this region are double-counted while interactions to the outside are not.
+This could be solved using the solvent-solvent energy between each pair of voxels $E_{ww,kl}$.
 While this is supported by the standard GIST implementation, it is rarely done due to the large size of the $E_{ww,kl}$ matrix.
 
-We note that replacement of a small or medium solvent region would lead to reorganization of the surrounding solvent.
-However, this depends on the local environment and on \emph{what} the molecule would be replaced by.
+Integrating GIST values over the whole grid corresponds to a process where all the solvent is removed into bulk.
+However, when integrating over a small region, this process is less well-defined, since it is unclear to what extent the remaining solvent would reorganize.
+However, this depends on the local environment and on \emph{what} the solvent would be replaced by.
 Therefore, the effect of reorganization can not be directly included in the GIST method, but must be judged for each case individually.
-This is not a problem when computing the total free energy of solvation $\dgsolv$, since no additional boundary is created.
-\todo{This is a bit unclear to me, could we elaborate on this a bit more? What are we asking of the reader here, how would one consider this effect? -vah}
-
-\subsection{PME implementation}
+
+\subsubsection{PME energy}
 
 In the original version of GIST, the energies are calculated based on the the minimum image convention.
-In PME-GIST, the electrostatics are calculated using the particle mesh Ewald (PME) method, which yields energies that are highly consistent with the treatment in the Amber MD engine \cite{Chen2021}.
-The electrostatic and Lennard-Jones parts $E_\text{elec}$ and $E_\text{lj}$ are computed separately.
+In PME-GIST, the electrostatic energy $E_\text{elec}$ is calculated using the particle mesh Ewald (PME) method, which yields energies that are highly consistent with the Amber MD engine \cite{Chen2021}.
+The Lennard-Jones part $E_\text{lj}$ is computed separately in direct space.
 
@@
 
-In PME-GIST, the system is treated as periodic, and the electrostatic energy is split into a short-range term $E_\text{dir}$, which is calculated in direct space using a distance cutoff, and a long-range term $E_\text{rec}$, which is calculated in reciprocal space.
+During the electrostatics calculation, the system is treated as periodic and the energy is split into a short-range term $E_\text{dir}$, which is calculated in direct space using a distance cutoff, and a long-range term $E_\text{rec}$, which is calculated in reciprocal space.
 Additionally, there is a correction term $E_\text{self}$ (called $E_\text{corr}$ in the original publication \cite{Chen2021}), which corrects for the self-interaction of each solvent molecule in the reciprocal term.
-
-\begin{equation}
-	E_\textit{elec} = E_\textit{dir} + E_\textit{rec} + E_\textit{self}
-\end{equation}
-%\todo{How is E$_{dir}$ treated? Is there a cutoff as well (PME GIST paper seems to indicate so)? If so, we should probably note that. } - ok? -fwa
-The short-range Lennard-Jones contribution is computed in the direct space, using a distance cutoff.
+\begin{equation}
+	E_\textit{elec} = E_\text{dir} + E_\text{rec} + E_\text{self}
+\end{equation}
+The short-range Lennard-Jones contribution is computed in the direct space using a distance cutoff.
 Furthermore, a long-range correction term is computed that accounts for the contributions above this cutoff assuming a homogeneous distribution of particles.
@@
 \begin{equation}
-	E_\textit{lj} = E_\textit{lj,\ short} +  E_\textit{lj\_corr}
-\end{equation}
-
-\section{Cheatsheet}
+	E_\text{lj} = E_\text{lj,\ short} +  E_\text{lj,\ corr}
+\end{equation}
+
+\section{Cheat sheet}
 %This will be a cheatsheet containing all options for cpptrajs GIST implementation
+The checklist ``The CPPTRAJ GIST action'' provides a cheat sheet summarizing the important options to the GIST action and its output, as well as frequently used equations.
+
 \subsection{Installation}
 GIST calculations in cpptraj can be sped up through various levels of 
-parallelization. When cpptraj was compiled with the '-cuda' configure flag set, 
-the GPU accelerated version of GIST is used automatically.
+parallelization.
+If cpptraj was compiled with the '-cuda' configure flag set, the GPU accelerated version of GIST is used automatically. 
+Furthermore, if cpptraj was compiled with OpenMP support, the entropy calculation will be parallelized.
+
+For detailed instructions on how to install \software{cpptraj}, we refer the reader to the AmberMD GitHub page \url{https://github.com/Amber-MD/cpptraj}.
 
@@
 
-\begin{checklist}{The CPPTRAJ GIST action}
+\begin{checklist}{The CPPTRAJ GIST action}\label{cheatsheet}
 

### 2022-12-05 manuscript/manuscript.tex
Fleshed out biotin chapter a bit

--- 2022-11-23 manuscript/manuscript.tex
@@
 \todo{Move simulation part from complex to here}
+Now, you can run restrained MD simulations based on the equilibrated structures for subsequent GIST analyses.
+Restraints should be used during the MD because IST theory is not defined for moving solutes.
+We apply a harmonic restraint of \SI{100}{\kilo\calorie\per\mole\per\angstrom\squared} to all heavy atoms to keep the system in place.
+We run \SI{100}{\nano\second} of simulation and store the conformations every \SI{100}{\pico\second}.
+A simulation length of at least \SI{20}{\nano\second} is recommended to obtain statistically independent snapshots of the water molecules.
+
+An example Amber input file for GIST might look like this:
+\begin{lstlisting}[style=amber-in]
+	restrained 100 ns NpT
+	&cntrl
+	ntx=5, irest=1,
+	ioutfm=1,
+	ntb=2, iwrap=1,
+	ntr=1, restraint_wt=100.0,
+	restraintmask="!@H=&!:WAT",
+	ntp=1, pres0=1.0, taup=1.0,
+	ntc=2, ntf=2,
+	ntt=3, tempi=300.0, temp0=300.0, gamma_ln=2,
+	nstlim=50000000, dt=0.002,
+	ntwr=50000, ntwx=5000, ntpr=5000,
+	/
+\end{lstlisting}
+This file would run a 100 ns NpT MD simulation while keeping all atoms except for hydrogens and waters restrained. Note that these are very standard MD settings except for the restraints.
+
+Run the MD using \software{pmemd.cuda}:
+\begin{lstlisting}[style=bash]
+	pmemd.cuda \
+	-O \
+	-i 100ns-npt-restraint.in \
+	-o md-01.out \
+	-p solvated.parm7 \
+	-c EQUIL-DONE.rst \
+	-x md-01.nc \
+	-r md-01.ncrst \
+	-ref EQUIL-DONE.rst
+\end{lstlisting}
 

### 2022-12-07 manuscript/manuscript.tex
Move MD part to basic tutorial, other minor edits

--- 2022-11-23 manuscript/manuscript.tex
@@
 Even when running the simulation in NVT, it is advisable to include an NpT equilibration step to adjust the box size beforehand.
-Use enough equilibration time to allow the water to relax around your biotin structure.
+Use enough equilibration time to allow the solvent to relax. 
+For water, \SI{0.5}{\nano\second} are usually sufficient to equilibrate the density, but narrow binding pockets can take longer.
 
 \subsubsection{MD simulation}
-\todo{Move simulation part from complex to here}
-
-\subsection{Applying GIST for the Streptavidin-Biotin complex}
-\subsubsection{System Preparation and Equilibration}
-If you want to skip this section, run 
-\begin{lstlisting}[style=bash]
-make equilibration-targets
-\end{lstlisting}
-in the code folder. This assumes that you have Python and Amber set up properly and that \software{pmemd.cuda} is in the PATH.
-
-The choice of initial structure is crucial to the GIST results.
-Crystal structures may not represent the dominant solution structure for a given target.
-In our example, we only use a single equilibrated structure.
-A more sophisticated approach would be to identify one or several dominant structures out of the ensemble.
-Depending on what properties you are interested in, it might be beneficial to run an MD simulation choose dominant structures through clustering.
-
-When starting from a crystal structure, care has to be taken during the preparation step.
-Make sure to choose a proper protonation state for all titratable moeties.
-You will need to solvate your structure in a solvent box, which should be large enough to obtain unbiased long-range electrostatics.
-A good criterion for validating the box size is to check whether the water on the outside of the box behaves bulk-like according to the GIST output.
-
-When calculating solvation free energy differences between a complex and the corresponding monomers (the dissociated state), there are two options for dealing with configurational diversity in the structures.
-One can choose to sample the complex and dissociated states individually, or to sample only one of them and assume that there is sufficient overlap in the conformations.
-
-Here, we do not include any sampling of the conformational states.
-It is therefore important that the molecular conformations in the complex and dissociated states match exactly.
-Use \software{cpptraj} to produce biotin and streptavidin structures based on the complex by stripping the respective other molecule like this:
-\begin{lstlisting}[style=cpptraj]
-parm complex/solvated.parm7
-trajin complex/equil/EQUIL-DONE.rst
-strip ^2 parmout streptavidin/solvated.parm7
-trajout streptavidin/pre-equil.ncrst
-\end{lstlisting}
-(Adapt for biotin as necessary.)
-%We provide a script called \software{cpptraj\_remove\_mol.sh} that removes a single molecule from a topology and structure using \software{cpptraj} without altering the rest of the system.
-\todo{Explain cpptraj rather than our script.
-The script is for the automated workflow only.}
-
-Use the \software{equilibration.py} script (or your own) to perform short NVT and NpT equilibration runs on the complex structure.
-Note that the equilibration script contains unrestrained NpT steps, because restraining two molecules at the same time might interfere with the pressure equilibration.
-
-Then split the equilibrated structure into biotin and streptavidin parts using \software{cpptraj\_remove\_mol.sh}.
-After that, you can use \software{equilibration.py -R} to run short restrained 
-equilibrations (where only the water is allowed to move) on the individual 
-systems.
-
-\subsubsection{Running MD for GIST}
-
-Now, you can run restrained MD simulations based on the equilibrated structures for subsequent GIST analyses.
-Restraints should be used during the MD because IST theory is not defined for moving solutes.
+Now, you can run restrained MD simulations based on the equilibrated structures.
+Restraints should be used because IST theory is not defined for moving solutes.
 We apply a harmonic restraint of \SI{100}{\kilo\calorie\per\mole\per\angstrom\squared} to all heavy atoms to keep the system in place.
@@
 Compute the energy (\inlinecode{Eall}) and entropy (\inlinecode{dTSsix}) contributions separately.
-You will find that the energy is strongly unfavorable to binding, because we do not yet include the interaction energy between biotin and streptavidin.
-
-Using the \inlinecode{energy} command in \software{cpptraj}, you can compute this energy difference.
+You will find that the energy disfavors binding, because we do not yet include the interaction energy between biotin and streptavidin.
+
+Using the \inlinecode{energy} command in \software{cpptraj}, you can compute this interaction energy.
 We recommend using PME in combination with PME-GIST, but not with GPU-GIST\@.
@@
 	\caption{Free energy contributions for the monomers and the dimer in kcal$\cdot$mol$^{-1}$.
-``total'' is calculated as $E^\text{internal} + \Delta E^\text{GIST} - \Delta S^\text{scaled}$.}\label{tab_dg_monomers_dimer}
+``Diff'' is calculated as $E^\text{internal} + \Delta E^\text{GIST} - \Delta S^\text{scaled}$.}\label{tab_dg_monomers_dimer}
 	\small

### 2022-12-07 manuscript/manuscript.tex
Proof-reading from GIST in multiple solvents downwards

--- 2022-12-05 manuscript/manuscript.tex
@@
 		gridcntr    & <y> <y> <z> & Coordinates of the grid center in \AA & 0 0 0  \\
-		griddim     & <nx> <ny> <nz> & Number of voxels per axis & 40 40 40\pdfcomment{Overfull hbox?}\\
+		griddim     & <nx> <ny> <nz> & Number of voxels per axis & 40 40 40\\
 		gridspacn & <dval> &Grid spacing / voxel dimensions in  \AA & 0.5    \\
@@
 \subsubsection{MD simulation}
-\todo{Move simulation part from complex to here}
-Now, you can run restrained MD simulations based on the equilibrated structures for subsequent GIST analyses.
-Restraints should be used during the MD because IST theory is not defined for moving solutes.
+Now, you can run restrained MD simulations based on the equilibrated structures.
+Restraints should be used because IST theory is not defined for moving solutes.
 We apply a harmonic restraint of \SI{100}{\kilo\calorie\per\mole\per\angstrom\squared} to all heavy atoms to keep the system in place.
@@
 \end{lstlisting}
-This file will run a 100 ns NpT MD simulation while keeping all atoms except for hydrogens and waters restrained. Note that these are very standard MD settings except for the restraints.
+Note that these are very standard MD settings except for the restraints.
 
@@
 You will find that the energy disfavors binding, because we do not yet include the interaction energy between biotin and streptavidin.
-
-You can also plot the difference between the complex and monomer contributions against the distance to biotin, using the complex coordinates for the holo structure.
-The expected result is shown in Figure~\ref{fig-dg-sums}.
-
-\begin{figure}
-	\centering
-	\includegraphics[width=1.0\linewidth]{figures/deltaG-difference.png}
-	\caption{TODO caption}\label{fig-dg-sums}
-\end{figure}
 
@@
 %\todo{Table 1 is placed quite akwardly at the moment for me, maybe enforce its position? -vah. Ok? -fwa}
-
+\todo{this table is super confusing to me. What's "total"? Doesn't seem to be E-TS. Diff is also not the sum of other total entries. Shouldn't it be TS instead of S?-vah}
 \begin{table}[h]
@@
 Additionally, there is a correction term $E_\text{self}$ (called $E_\text{corr}$ in the original publication \cite{Chen2021}), which corrects for the self-interaction of each solvent molecule in the reciprocal term.
-\begin{equation}
-	E_\textit{elec} = E_\text{dir} + E_\text{rec} + E_\text{self}
+
+\begin{equation}
+	E_\textit{elec} = E_\textit{dir} + E_\textit{rec} + E_\textit{self}
 \end{equation}
@@
 \begin{equation}
-	E_\text{lj} = E_\text{lj,\ short} +  E_\text{lj,\ corr}
+	E_\textit{lj} = E_\textit{lj,\ short} +  E_\textit{lj,\ corr}
 \end{equation}

### 2022-12-12 manuscript/manuscript.tex
Add delta-G difference plot

--- 2022-12-07 manuscript/manuscript.tex
@@
 You will find that the energy disfavors binding, because we do not yet include the interaction energy between biotin and streptavidin.
+
+You can also plot the difference between the complex and monomer contributions against the distance to biotin, using the complex coordinates for the holo structure.
+The expected result is shown in Figure~\ref{fig-dg-sums}.
+
+\begin{figure}
+	\centering
+	\includegraphics[width=1.0\linewidth]{figures/deltaG-difference.png}
+	\caption{TODO caption}\label{fig-dg-sums}
+\end{figure}
 
@@
 %\todo{Table 1 is placed quite akwardly at the moment for me, maybe enforce its position? -vah. Ok? -fwa}
-\todo{this table is super confusing to me. What's "total"? Doesn't seem to be E-TS. Diff is also not the sum of other total entries. Shouldn't it be TS instead of S?-vah}
+
 \begin{table}[h]
@@
 Additionally, there is a correction term $E_\text{self}$ (called $E_\text{corr}$ in the original publication \cite{Chen2021}), which corrects for the self-interaction of each solvent molecule in the reciprocal term.
-
-\begin{equation}
-	E_\textit{elec} = E_\textit{dir} + E_\textit{rec} + E_\textit{self}
-\end{equation}
+\begin{equation}
+	E_\textit{elec} = E_\text{dir} + E_\text{rec} + E_\text{self}
+\end{equation}
+
 %\todo{How is E$_{dir}$ treated? Is there a cutoff as well (PME GIST paper seems to indicate so)? If so, we should probably note that. } - ok? -fwa
@@
 \begin{equation}
-	E_\textit{lj} = E_\textit{lj,\ short} +  E_\textit{lj\_corr}
+	E_\text{lj} = E_\text{lj,\ short} +  E_\text{lj,\ corr}
 \end{equation}

### 2022-12-12 manuscript/manuscript.tex

--- 2022-12-07 manuscript/manuscript.tex
@@
 Additionally, there is a correction term $E_\text{self}$ (called $E_\text{corr}$ in the original publication \cite{Chen2021}), which corrects for the self-interaction of each solvent molecule in the reciprocal term.
-
-\begin{equation}
-	E_\textit{elec} = E_\textit{dir} + E_\textit{rec} + E_\textit{self}
+\begin{equation}
+	E_\textit{elec} = E_\text{dir} + E_\text{rec} + E_\text{self}
 \end{equation}
@@
 \begin{equation}
-	E_\textit{lj} = E_\textit{lj,\ short} +  E_\textit{lj,\ corr}
+	E_\text{lj} = E_\text{lj,\ short} +  E_\text{lj,\ corr}
 \end{equation}

### 2022-12-12 manuscript/manuscript.tex
Include the isolevels figure, and add a caption

--- 2022-12-12 manuscript/manuscript.tex
@@
 
-\newcommand{\versionnumber}{1.0}  % you should update the minor version number in preprints and major version number of submissions.
-\newcommand{\githubrepository}{\url{https://github.com/liedllab/gist-tutorial}}  %this should be the main github repository for this article
+\newcommand{\versionnumber}{1.3}  % you should update the minor version number in preprints and major version number of submissions.
+\newcommand{\githubrepository}{\url{https://github.com/myaccount/homegithubrepository}}  %this should be the main github repository for this article
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\title{Quantifying Spatially Resolved Hydration Thermodynamics Using Grid Inhomogeneous Solvation Theory [Article v\versionnumber]}
-
-\author[1\authfn{1}]{Franz Waibl}
-\author[1\authfn{1}]{Valentin J Hoerschinger}
-\author[2]{Vjay Molino}
-\author[1]{Monica L Fern{\'a}ndez-Quintero}
-%\author[1,2]{Firstname Middlename Familyname}
-%\author[2]{Firstname Initials Surname}
-\author[4]{Daniel R Roe}
-\author[1*]{Klaus R Liedl}
-\author[2*]{Michael K Gilson}
-\author[2*]{Tom Kurtzman}
-\affil[1]{Department of General, Inorganic and Theoretical Chemistry, University of Innsbruck, Austria}
+\title{This is the title [Article v\versionnumber]}
+
+\author[1*]{Firstname Middlename Surname}
+\author[1,2\authfn{1}\authfn{3}]{Firstname Middlename Familyname}
+\author[2\authfn{1}\authfn{4}]{Firstname Initials Surname}
+\author[2*]{Firstname Surname}
+\affil[1]{Institution 1}
 \affil[2]{Institution 2}
@@
 \contrib[\authfn{1}]{These authors contributed equally to this work}
-%\contrib[\authfn{2}]{These authors also contributed equally to this work}
-
-%\presentadd[\authfn{3}]{Department of General, Inorganic and Theoretical Chemistry, University of Innsbruck, Austria}
-%\presentadd[\authfn{4}]{Department, Institute, Country}
+\contrib[\authfn{2}]{These authors also contributed equally to this work}
+
+\presentadd[\authfn{3}]{Department, Institute, Country}
+\presentadd[\authfn{4}]{Department, Institute, Country}
 
@@
 	extendedchars=true,
-	literate={Å}{{\AA}}1,
-	backgroundcolor = \color[rgb]{0.9, 0.9, 0.9},
+	literate={Å}{{\AA}}1
 }
@@
 \lstdefinestyle{cpptraj}{title={\centering Cpptraj}}
-\lstdefinestyle{bash}{title={\centering Command-line}}
+\lstdefinestyle{bash}{title={\centering Commandline}}
 \lstdefinestyle{amber-in}{title={\centering Amber input}}
@@
 The high spatial resolution of the GIST output, as well as the decomposition into energy and entropy contributions, allow for highly detailed analyses on both proteins and small molecules.
-However, this versatility also comes with a significant entry barrier for new users.
+However, this versatility also comes with a higher entry barrier for new users.
 
 In this tutorial, we aim to guide the reader through the most common steps involved in a GIST analysis at the example of the streptavidin-biotin complex.
-Furthermore, we discuss the theory of GIST with a focus on practical aspects.
-We highlight potential pitfalls and show how to avoid technical difficulties.
+Furthermore, we discuss the theory of GIST with a focus on practical aspects, and show several pitfalls and technical difficulties that may occur during a GIST study.
 We assume familiarity with molecular dynamics simulations as well as the AmberTools package.
@@
 \todo{IMO, our article looks too dry. We should have something nice (e.g., an illustration of the GIST method) on the 1st or 2nd page -fwa.}
-\todo{Potentially relevant: 10.1021/acs.jcim.7b00520 }
 
 Solvation thermodynamics govern any process involving solutes in a solvent.
-%Cite "Theory of Simple Liquids"?
 %Solutes are constantly interacting with solvents in their environment, necessitating a good description of these interactions for accurate computational predictions. 
-The free energy of solvation is defined as the reversible work needed to insert a solute into the solvent \cite{ben-naim-book}.
-Especially the interaction with water is of utmost importance, as many biological processes occur in aqueous environment \cite{Privalov2017-water-review}.
-It drives real-world effects such as hydrophobic interactions, binding of ligands to a biomolecule, as well as dynamics and folding of proteins.
+The free energy of solvation is defined as the work needed to insert a solute into a solvent.
+Especially the interaction with water is of utmost importance, as many biological processes occur in a aqueous environment.
+It thereby strongly influences real-world problems such as hydrophobic effects, the binding of ligands to a biomolecule, as well as the dynamics and folding of proteins.
 
@@
 Data-driven methods such as QSPR or UNIFAC can deliver good results \cite{Borhani2019-qspr,Fredenslund1975-unifac}, but might be limited to molecules that are similar to those in the training set.
-Polarizable continuum models (PCMs) \cite{Miertus1981-pcm} such as COSMO \cite{Klamt1993-cosmo} treat the solvent as a homogeneous phase that reacts to the electrostatic potential of the solute \cite{Mennucci2010-pcm}.
+Polarizable continuum models (PCMs) \cite{Miertus1981-pcm,Klamt1993-cosmo} such as treat the solvent phase as a homogeneous phase that reacts to the electrostatic potential of the solute \cite{Mennucci2010-pcm}.
 %Molecular mechanics based approaches describe the underlying PES through empirically fit force-fields, which reduce the computational demand of these calculations tremendously.
 %This allows for the treatment of large biological systems such as proteins in aqueous solution.
-Methods such as MM/PBSA or MM/GBSA \cite{Sitkoff1994-pbsa,Kollman2000-mmpbsa} combine an implicit treatment of solvent electrostatics with structural sampling of the solute.
-They can quickly provide results for a large number of structures, but can yield inaccurate results when compared to more rigorous explicit solvent methods \cite{Genheden2015-mmpbsa-review}.
+Methods such as MM/PBSA or MM/GBSA \cite{Sitkoff1994-pbsa,Kollman2000-mmpbsa} combine an implicit solvent treatment of electrostatics with structural sampling of the solute.
+They can quickly provide results for a large number of structures, but yield inaccurate results when compared to more rigorous explicit solvent methods \cite{Genheden2015-mmpbsa-review}.
 
@@
 They are generally more accurate than implicit methods, but require a statistical sampling of the solvent conformations \cite{Liu2016-md-solubility,Swails2014-cphmd}.
-Molecular dynamics (MD) simulations provide an accurate way of modeling solvent packing in confined areas \cite{Haider2016-water-on-surfaces} and general hydrophobic effects \cite{Pratt2016-hydrophobicity}.
+Molecular dynamics (MD) simulations provide an accurate way of modeling solvent packing in confined areas \cite{}\todo{ask Tom for a citation!} and general hydrophobic effects \cite{Pratt2016-hydrophobicity}.
 
 The free energy of solvation can be calculated in a statistically rigorous way using alchemical methods \cite{Liu2016-md-solubility,Mobley2009-dgsolv,Mobley2014-freesolv} such as free energy perturbation (FEP) \cite{Zwanzig1954-reweighting} or thermodynamic integration (TI) \cite{Kirkwood1935-ti}.
-However, those methods are unable to provide a spatial interpretation of the results.
-Furthermore, splitting the free energy into enthalpy and entropy contributions is challenging \cite{Peter2004-alchemical-entropy}.
+However, they are unable to provide a spatial interpretation of the results.
+Furthermore, splitting the resulting free energy into contributions for the enthalpy and entropy is challenging \cite{Peter2004-alchemical-entropy}.
 %Assuming sufficient sampling, they can accurately predict the free energy of solvation in the boundaries of the assumptions underlying their forcefield. 
@@
 Methods derived from statistical mechanics bridge this gap by providing three-dimensional resolution.
-Two major flavors of methods in this group are those based on inhomogeneous fluid solvation theory (IST) \cite{Lazaridis1998} and those based on the Ornstein--Zernike (OZ) equation \cite{Hansen2013-simple-liquids}.
+Two major groups of methods that fall into this group are those based on inhomogeneous solvation theory (IST) \cite{Lazaridis1998} and those based on the Ornstein--Zernike (OZ) equation \cite{Hansen2013-simple-liquids}.
 The most important example of OZ-based methods is the reference interaction site model (RISM) \cite{Chandler1972-rism} as well as its extension to three dimensions (3D-RISM) \cite{Kovalenko1998-3drism}.
-Those methods compute the atomic solvent distributions by self-consistently solving the OZ integral equation to predict solvation thermodynamics.
+Those methods compute the solvent distributions by self-consistent solution of the OZ integral equation to predict solvation thermodynamics.
 %The 3D-RISM approach allows for a decomposition into entropic and enthalpic terms as well as a full localization of the underlying distributions to the solute structure.
@@
 On the other hand, methods based on IST generally use a molecular solvent distribution obtained from an MD simulation to compute the free energy of hydration from its enthalpic and entropic contributions.
-Examples of those methods include grid inhomogeneous solvation theory (GIST) \cite{Nguyen2012,Ramsey2016}, WaterMap \cite{Young2007-watermap,Abel2008-watermap}, SSTMap \cite{Haider2018-sstmap}, STOW \cite{Li2012-stow} and Solvaware \cite{Huggins2016-solvaware}.
+Examples of those methods include grid inhomogeneous solvation theory (GIST) \cite{Nguyen2012,Ramsey2016}, WaterMap \cite{Young2007-watermap,Abel2008-watermap}, SSTMap \cite{Haider2018-sstmap}, and STOW \cite{Li2012-stow}.
 The main limitation of those methods is that they rely on an infinite expansion of the entropy in terms of the density distribution, which is truncated after the first or, sometimes, the second \cite{Nguyen2016-gist-second-order,Waibl2022-gist-solvents} term.
@@
 GIST is based on IST, solving the integral equations on a grid. 
-This approach is more general than other IST-based methods because it accounts for all solvent regions rather than focusing on high-occupancy solvation sites.
+This approach is more general than other IST-based methods because it accounts for all solvent regions rather than focusing on high-occopancy solvation sites.
 %By simulating the dynamics of a solvent around the solute, GIST calculates thermodynamic properties for a standard solvation process as described by Ben-Naim \cite{ben-naim-book}.
-It estimates the solvent density from an MD trajectory and allows for a full localization of entropic and enthalpic contributions on the grid around the solute.
-For small molecules, GIST has been shown to reach comparable accuracy to TI by approximating the higher-order entropies as a fraction of the first-order entropy, using a factor of \SI{-0.4} for water \cite{Chen2021,Waibl2022-gist-solvents}.
+It estimates the solvent density from a molecular dynamics (MD) trajectory and allows for a full localization of entropic and enthalpic contributions on the grid around the solute.
+It has been shown that GIST can produce comparable accuracy to TI by approximating the higher-order entropies as a fraction of the first-order entropy (using a factor of \SI{-0.4} for water) \cite{Chen2021,Waibl2022-gist-solvents}.
 %applying a linear correction factor (0.6 for water) to the first-order entropy as follows \cite{Chen2021,Waibl2022-gist-solvents}:
@@
 
-Recently, GIST was implemented on the GPU, tremendously improving the calculation speed, and allowing for the treatment of large biological systems such as proteases or antibodies \cite{Kraml2019-gigist}.
+Recent work has improved the calculation speed tremendously, allowing for the treatment of large biological systems such as proteases or antibodies.
 Furthermore, an implementation based on the particle mesh Ewald (PME) \cite{Darden1993-pme} method was presented, which improves the consistency between GIST and popular molecular dynamics engines. 
-Yet further changes allow the inclusion of other rigid solvents than water \cite{Kraml2020,Waibl2022-gist-solvents} or 
-allow for the treatment of salt-water mixtures as solvents \cite{Waibl2021-gist-salt}.
-
-While previous publications \cite{Ramsey2016} and tutorials\footnote{See \url{http://ambermd.org/tutorials/advanced/tutorial25/}} gave insightful pointers how to setup and run GIST calculations, the post-processing was often treated less rigorously.
-While building on these tutorials, we provide an updated guide to calculating solvation properties using GIST to address recent improvements.
-Furthermore, we aim to establish a template for the analysis and interpretation of GIST calculations.
+Yet further changes allow the inclusion of other rigid solvents than water or 
+allow for the treatment of solvent mixtures and ions in solution. 
+\todo{Sources!}
+
+While previous tutorials gave insightful pointers how to setup and run GIST calculations, their post-processing was often treated less rigorously.
+While building on these tutorials, an updated guide to calculating solvation properties using GIST is therefore presented here to address recent improvements to the method as well as provide a basic template for the analysis and interpretation of GIST calculations.
 
@@
 
-This tutorial provides an example of a GIST workflow that can be easily adapted to different systems. 
-In addition to a short introduction of the method and the theory behind it, we present an overview of the different implementations of GIST.
-We ease the user into basic GIST analysis at the example of biotin.
-Building on the basics, we present a more advanced GIST study of the streptavidin-biotin complex, reflecting on many common aspects of GIST studies.
+This tutorial provides an example of a GIST workflow that can be easily adapted towards different systems. 
+In addition to a short introduction of the method and the theory behind it, we present an overview of the various different flavors of GIST methods.
+We ease the user into the basic usage of GIST by presenting a calculation for a small molecule.
+Building on the basics, a more advanced GIST study of the streptavidin-biotin complex is presented, which reflects on many common aspects of GIST studies.
+Using these test systems, we present solvation analysis of a small molecule (biotin) as well as a biomolecule (streptavidin).
 We show how to interpret the three-dimensional contributions of solvation free energy in a binding pocket and around the ligand.
-Additionally, we compute the binding contribution of the free energy of solvation, which requires accurate post-processing of the GIST outputs to avoid unfavorable summation of bias.
-We believe that this example study covers a wide range of applications.
+
+Additionally, we compute the binding contribution of the free energy of solvation, which requires accurate post-processing of the GIST outputs to avoid unfavorable summation of bias in the GIST calculation.
+We believe that this example study covers a wide range of applications of the GIST method.
 
 Furthermore, we discuss several technical aspects  and typical pitfalls of GIST analyses, such as the normalization of voxel values and how to deal with double counting of energy contributions. 
-Additionally, we provide a Python library (\software{gisttools}) to unify the analysis of GIST data produced by various versions of GIST and make the post-processing of GIST results more accessible.
+Additionally, we provide a python library to unify the analysis of GIST data produced by various versions of GIST and make the post-processing of GIST results more accessible.
 
@@
 This tutorial is aimed at users with a solid knowledge of molecular dynamics (MD) simulations.
-The user should be able to run MD simulations and analyze the resulting trajectories.
+The user should be able to run MD simulations and work with their output data.
 Previous experience of GIST is not necessary.
-However, we only provide a short summary of the theory behind GIST in section~\ref{sec:theory} at the end of this tutorial.
-For more detailed information, we point the reader towards the primary sources.
+However, we only provide a short explanation of the theory behind GIST in section \ref{sec:theory} at the end of this tutorial.
+For more detailed information, we point the reader towards the primary sources for more detailed information.
 
 The presented analyses are run in Python.
-While no programming experience is necessary to follow the tutorial, a basic understanding of Python is advantageous to adjust the code towards different requirements and use-cases.
+While no programming experience should be necessary to follow the tutorial, a basic understanding of Python is advantageous to adjust the code towards different requirements and use-cases. 
+We provide a Jupyter notebook which contains the presented analyses. 
+Therefore, an understanding of Jupyter notebooks is beneficial to reproduce the analyses.
 
@@
 
-The closed-source Amber simulation engine is used in this tutorial to run the MD simulations.
-If Amber is not available, other MD engines can be used as long as topologies and trajectories can be produced in a format supported by \software{cpptraj}.
+The closed-source Amber simulation engine is used in this tutorial to run the necessary MD simulations.
+If no Amber installation is available, the MD can be performed other simulation engines, as long as the topology and trajectory can be produced in a format supported by \software{cpptraj}.
 If using GROMACS \cite{Abraham2015-gromacs,Pronk2013-gromacs}, we propose to prepare the topology with AmberTools and convert it using \software{acpype.py} \cite{Sousa_da_Silva2012-acpype}.
 %We should actually test that!
-Also note that compressed trajectory formats such as \inlinecode{xtc} might bias the entropy calculation due to the loss of precision for the atom positions. \todo{This might be in the technical part}
+Also note that compressed trajectory formats such as \inlinecode{xtc} might bias the entropy calculation due to the loss of precision for the atom positions.
 
@@
 	\item \software{gisttools} (\url{https://github.com/liedllab/gisttools})
-	\item \software{mdtraj (\url{https://www.mdtraj.org/}) \cite{McGibbon2015-mdtraj}}
+	\item \software{mdtraj (\url{https://www.mdtraj.org/})}
 \end{itemize}
@@
 
-For the visualization of three-dimensional contributions we require molecular visualization software that can handle \software{OpenDX} files, such as \software{PyMol} \cite{pymol} or \software{VMD} \cite{vmd}.
+For the visualization of three-dimensional contributions we require molecular visualization software that can handle \software{OpenDX} files.
+We recommend using a current version of \software{PyMol}\cite{pymol} or \software{VMD}\cite{vmd}.
 
@@
 \begin{itemize}
-	\item \software{cpptraj}.
+	\item A running installation of \software{cpptraj}.
 	\item A topology of the molecule of interest in a solvent box.
@@
 \paragraph{Identify and define the region of interest}
-In order to run a GIST analysis, one must define the region where solvent properties will be calculated.
-This region is represented by a rectangular grid with given position, size, and voxel spacing.
+In order to run a GIST analysis, the region where the solvent properties will be calculated must be first defined.
+This region is represented by a rectangular grid centered at (x, y, z) with $N_x \times N_y \times N_z$ voxels, where each voxel is cubic with a sidelength of \inlinecode{gridspacn}.
+While integrals over GIST quantities are independent of the voxel size, individual voxel values converge slower at high resolutions.\pdfcomment{The calculation actually can actually get faster at higher resolution, because the same grid is used in the nearest-neighbor search.}
 The default value of \SI{0.5}{\angstrom} provides reasonable discretization for many use-cases.
-While integrals of GIST quantities over the grid are independent of the voxel size, individual values converge slower at high resolutions.
 
 The grid is defined by parameters of the \inlinecode{gist} command in \software{cpptraj}.
-The grid center is defined by \inlinecode{gridcntr}, the voxel number in x, y, and z direction by \inlinecode{griddim}, and the spacing (voxel side-length) in \AA by \inlinecode{gridspacn}.
+The grid center is defined by \inlinecode{gridcntr}, the voxel number in x, y, and z direction by \inlinecode{griddim}, and the spacing in \AA by \inlinecode{gridspacn}.
 
@@
 
-Often, the region of interest will be defined as the surrounding of a molecule, which might for example be a ligand in the binding pocket or the whole protein.
+Often, the region of interest will be defined by the surrounding of a molecule, which might either be a ligand in the binding pocket or also the whole protein. 
 A helper script called \inlinecode{FindCentroid.py} for this purpose is supplied with this tutorial, and can be used as follows:
@@
 
-The first line specifies the Amber topology file that was used to run MD. The second line species the trajectory file.
-Note that the solute should not move significantly for a GIST calculation, since IST does not take the solute degrees of freedom into account.
-This is typically achieved using restraints on the solute heavy atoms.
-The third line tells \software{cpptraj} to run a GIST calculation using the specified region and print the output in a file called \inlinecode{gist.dat}.
-The last line instructs \software{cpptraj} to run the calculation.
-
+The first line specifies the amber topology file that was used to run MD. The second line species the trajectory file.
+Note that the solute should not move significantly for a GIST calculation.
+This is typically achieved by using restraints on the solute coordinates.
+The third line tells \software{cpptraj} to run a GIST calculation using the region defined and print the output of the calculation in a file called \inlinecode{gist.dat}.
+The last instructs cpptraj to run the calculation.
 The syntax for running \software{cpptraj} given this input file is:
+
 \begin{lstlisting}[style=bash]
@@
 
-The \inlinecode{gist} action has a large number of options to modify the GIST calculation or its output.
-A full list is provided at the end of this tutorial or through the \software{cpptraj} manual.
-The most important options are summarized in the following table:
+The \inlinecode{gist} action has a large number of options to modify the GIST calculation or its' output.
+A full list is provided at the end of this tutorial or through cpptrajs manual.
+The major options are summarized in the following table:
 
 \begin{table}[h]
-	\caption{The main options for GIST in \software{cpptraj}.}\label{tab:GIST_options}
+	\caption{The major options available in cpptrajs' GIST implementation.}\label{tab:GIST_options}
 	\small
@@
 		gridcntr    & <y> <y> <z> & Coordinates of the grid center in \AA & 0 0 0  \\
-		griddim     & <nx> <ny> <nz> & Number of voxels per axis & 40 40 40\pdfcomment{Overfull hbox?}\\
+		griddim     & <nx> <ny> <nz> & Number of voxels along each axis defining the grid size & 40 40 40\\
 		gridspacn & <dval> &Grid spacing / voxel dimensions in  \AA & 0.5    \\
 		refdens         & <rdval> & Reference density of the bulk water in molecules/\AA$^3$ . & 0.0334    \\
-		temp & <tval> & System temperature in Kelvin & 300 \\
 		\bottomrule
@@
 \paragraph{Postprocessing and Analyzing GIST results}
-Cpptraj outputs quantities in OpenDX (.dx) file format which can be visualized in Pymol or VMD.
-Aside from thermodynamic quantities, GIST also calculates the number density of each atomic element in the solvent.
-Here is a list of the most important quantities calculated by GIST:
+Cpptraj outputs the quantities calculated as a "open data explorer" (.dx) file format which can be easily visualized in Pymol or VMD.
+Aside from the thermodynamic quantities, GIST also calculates the number densities of hydrogen and oxygen in each voxel. Here is a list of the most important quantities calculated by GIST:
 
@@
 
-Several dx files are written by default, such as the energy, entropy, and atomic number densities.
-All calculated quantities are written to the output file (default \inlinecode{gist-output.dat}) and can be used to generate DX files for other quantities.
-A full list of all properties in the output is provided at the end of the tutorial.
-
-Energy and entropy quantities are reported in two different ways in the GIST output.
-The \inlinecode{\_dens} quantities are normalized by the voxel volume and can be used to compute integrals.
-The \inlinecode{\_norm} quantities are instead normalized by the average number of solvents in the respective voxel, and are useful to visualize local water properties.
+Several dx files are written by default, such as the energies, entropies, number density of oxygen, and number density of hydrogen.
+All calculated quantities are written to the output file (output.dat) and can be used to generate DX files for other quantities.
+A full list of all calculated properties found in the output in addition to the ones mentioned above is provided at the end of the tutorial.
 
@@
 Here, we aim to guide the reader through a GIST analysis of the streptavidin-biotin complex.
-We will
+During the tutorial, we will
 \begin{enumerate}
-	\item set up and run GIST calculations for a small molecule (biotin) and a protein (streptavidin) binding pocket.
-	\item compute the free energy of solvation for biotin using GIST.
-	\item visualize local contributions to $\dgsolv$ in the streptavidin binding pocket and around biotin.
-	\item compute the hydration contribution to the binding. Combined with the biotin--streptavidin interaction energy, the binding free energy can be estimated.
+	\item set up and run GIST calculations for a small molecule as well as a protein binding pocket.
+	\item compute the free energy of solvation for a small molecule using GIST.
+	\item visualize local contributions to $\dgsolv$ in the streptavidin binding pocket as well as around biotin.
+	\item compute the binding free energy of the complex from the individual energy contributions combined with the contribution of hydration computed using GIST.
 	\item explore advantages and disadvantages of the GIST method when used to calculate  $\dgsolv$
 \end{enumerate}
-
-In the first part of the tutorial, we will look at how to run GIST calculations for small molecules and interpret the output.
-Afterwards, we will apply this knowledge to a protein binding pocket, visualize $\dgsolv$ hotspots and estimate a free energy of ligand binding using an end-state approach based on GIST.
-In addition, we will discuss all steps in further detail at the end.
-\todo{TODO}
+The tutorial is divided into a beginners part and an advanced part.
+In the beginners part, we will take a look at how to run GIST calculations for small molecules and interpret the output for a small molecule.
+In the advanced part, we will apply this knowledge to a protein binding pocket, visualize $\dgsolv$ hotspots and estimate a free energy of binding for a small molecule into this pocket using an end-state approach.
+In addition, while the beginners part focuses only on GIST calculation run, we will discuss all steps along the way in detail in the advanced section.
 
 \subsection{Streptavidin/Biotin}
-The streptavidin-biotin complex binding is one of the strongest known non-covalent interactions involving a small molecule.
+The streptavidin-biotin complex is one of the strongest known non-covalent interactions involving a small molecule.
 Therefore, the combination of streptavidin and biotin is used routinely in biotechnology \cite{Dundas2013-streptavidin-review}, and the interaction between them has been investigated in detail \cite{McConnell2021-biotin}.
@@
 In nature, streptavidin occurs as a tetramer, and binds biotin with a binding constant of \SI{\sim e-14}{\per\Molar} \cite{Dundas2013-streptavidin-review}.
-Streptavidin has a so-called flap region, which closes the binding pocket after biotin binding.
+The binding to the closed conformation of the tetramer has been estimated to have a free energy of \SI{26.6}{\kilo\calorie\per\mol} \cite{Bansal2018-biotin}.
 Here, we will investigate the binding of biotin to a closed streptavidin monomer for simplicity.
-The binding to the closed conformation of the tetramer has been estimated to have a free energy of \SI{-26.6}{\kilo\calorie\per\mol} \cite{Bansal2018-biotin}.
 
 \subsection{Tutorial data}
-We will use the crystal structure of streptavidin (PDB code: 1STP \cite{Weber1989-streptavidin-structure}) to start our calculations.
-For the beginners section, we will use the biotin ligand extracted from the crystal structure.
+We will use the 1STP crystal structure to start our calculations.
+For the beginners section, we will use the biotin ligand as extracted from the crystal structure.
 For the second part, we will use the whole crystal structure.
@@
 \end{lstlisting}
-This will download the tutorial files, including both the manuscript and the examples in the \inlinecode{code} folder.
-
+This will download the tutorial files (including both the manuscript and the examples in the \software{code} folder).
+\
 \subsection{Running GIST for Biotin}
-In this example, we will be running a simple GIST calculation of the biotin molecule to get familiar with the usage of GIST in \software{cpptraj}.
+In this example, we will be running a simple GIST calculation for a small molecule to get familiar with the usage of the GIST command in \software{cpptraj}.
 \subsubsection{System preparation}
-First, extract biotin from the crystal structure and generate a topology and coordinate file compatible with Amber or a simulation package of your choice.
-The molecule should be solvated with a sufficiently large solvent box.
-Carefully check the employed force field, protonation state and partial charges.
-%a PDB file containing only biotin.
+First, extract biotin from the crystal structure and generate a PDB file containing only biotin.
 Prepare your structure by adding hydrogens, making sure to choose the correct protonation state.
 Parametrize biotin with the forcefield of your choice.
-Solvate the parametrized biotin, making sure to choose a water box of large enough size so that the water on the box edges behaves bulk-like (a wall distance of at least 15\AA is recommended).
-A good starting point (though a bit dated) for the parametrization and solvation of small molecules is the Amber tutorial 4b (\url{http://ambermd.org/tutorials/basic/tutorial4b/}).
-
+Solvate the parametrized biotin, making sur to choose a water box of large enough size so that the water on the box edges behaves bulk-like (a wall distance of at least 15\AA is recommended).
+A good starting point (though a bit dated) for the parametrization and solvation of small molecules is AMBER's tutorial 4b (\url{http://ambermd.org/tutorials/basic/tutorial4b/}).
 \subsubsection{Equilibration}
-Equilibrate your solvated biotin in an NVT or NpT ensemble.
-Even when running the simulation in NVT, it is advisable to include an NpT equilibration step to adjust the box size beforehand.
-Use enough equilibration time to allow the solvent to relax. 
-For water, \SI{0.5}{\nano\second} are usually sufficient to equilibrate the density, but narrow binding pockets can take longer.
+Equilibrate your solvated biotin to the conditions of your choice.
+Even when running a NVT simulation as the production run for the GIST analysis, it is advisible to include an NPT equilibration step to adjust the box size before hand.
+Allow enough equilibration time in the ensemble of your choice to allow the water to relax around your biotin structure.
 
 \subsubsection{MD simulation}
-Now, you can run restrained MD simulations based on the equilibrated structures.
-Restraints should be used because IST theory is not defined for moving solutes.
+\todo{Move simulation part from complex to here}
+
+\subsection{Applying GIST for the Streptavidin-Biotin complex}
+\subsubsection{System Preparation and Equilibration}
+If you want to skip this section, run 
+\begin{lstlisting}[style=bash]
+make equilibration-targets
+\end{lstlisting}
+in the code folder. This assumes that you have Python and Amber set up properly and that \software{pmemd.cuda} is in the PATH.
+
+The choice of the initial structure to be analyzed is crucial to the results 
+you can expect from the GIST calculation. Crystal structures may not be 
+representative of the dominant solution structure for a given analysis target. 
+In our example we use the final structure after a short equilibration run. A 
+more sophisticated approach would be identifying a dominant structure out of 
+the ensemble of sampled structures.  Depending on what properties you are 
+interested in, running a MD simulation of your structure beforehand and 
+choosing a dominant structure or even multiple structures through clustering 
+might therefore be beneficial.
+When starting with a crystal structure, special care has to be taken during the 
+preperation of the structure. Make sure to protonate your structure and take 
+care to choose a proper protonation state for the varying chemical moeties in 
+your structure. You will need to solvate your structure in a solvent box, which 
+should be chosen large enough so that the solvent around the region of interest 
+is not influenced through periodic images.
+
+When calculating solvation free energy differences between a complex and the corresponding monomers (the dissociated state), there are two options for dealing with configurational diversity in the structures.
+\todo{Maybe Valentin knows a citation for that?}
+One can choose to sample the complex and dissociated states individually, or to sample only one of them and assume that there is sufficient overlap in the sampled conformations.
+%One can choose to sample only the complex, assuming that all relevant conformations of the dissociated state are also found in the complex.
+%The probability distribution of the dissociated state can be obtained by reweighting the probabilities in the complex using the solvation free energies.
+%Alternatively, one can sample the complex and dissociated states individually.
+%This approach does not require an overlap between the conformational probabilities of the complex and dissociated states.
+%However, it is prone to statistical errors due to uncertainties in the ensembles, and requires the free energy contributions from the internal degrees of freedom of the monomers (in both the complex and the monomer states) to be known.
+\todo{This paragraph does not fit here well.
+Maybe put it somewhere else.}
+
+Here, we choose not to include any sampling of the conformational states. 
+It is therefore important that the conformations of the monomers in the complex and dissociated states match exactly. 
+Use \software{cpptraj} to produce biotin and streptavidin structures based on the complex by stripping the respective other molecule like this:
+\begin{lstlisting}[style=cpptraj]
+parm complex/solvated.parm7
+trajin complex/equil/EQUIL-DONE.rst
+strip ^2 parmout streptavidin/solvated.parm7
+trajout streptavidin/pre-equil.ncrst
+\end{lstlisting}
+(Adapt for biotin as necessary.)
+%We provide a script called \software{cpptraj\_remove\_mol.sh} that removes a single molecule from a topology and structure using \software{cpptraj} without altering the rest of the system.
+\todo{Explain cpptraj rather than our script.
+The script is for the automated workflow only.}
+
+Use the \software{equilibration.py} script (or your own) to perform short NVT and NpT equilibration runs on the starting structure.
+\todo {NPT}
+Note that the barostat might interfere with restrained equilibration of the complex, since different molecules are moved relative to each other when the volume changes.
+But since our equilibration script also contains unrestrained pressure 
+equilibration, we should not run into any trouble.
+
+Then split the equilibrated structure into biotin and streptavidin parts using \software{cpptraj\_remove\_mol.sh}.
+After that, you can use \software{equilibration.py -R} to run short restrained 
+equilibrations (where only the water is allowed to move) on the individual 
+systems.
+
+\subsubsection{Running MD for GIST}
+
+Now, you can run restrained MD simulations based on the equilibrated structures for subsequent GIST analyses.
+\todo {explain why the protein needs to be restrained for an MD simulation that will be analyzed using GIST}
 We apply a harmonic restraint of \SI{100}{\kilo\calorie\per\mole\per\angstrom\squared} to all heavy atoms to keep the system in place.
 We run \SI{100}{\nano\second} of simulation and store the conformations every \SI{100}{\pico\second}.
-A simulation length of at least \SI{20}{\nano\second} is recommended to obtain statistically independent snapshots of the water molecules.
+A simulation length of at least \SIrange{10}{20}{\nano\second} is recommended, to obtain statistically independent snapshots of the water.
 
@@
   nstlim=50000000, dt=0.002,
-ntwr=50000, ntwx=5000, ntpr=5000,
+  ntwr=50000, ntwx=5000,
+  ntpr=5000,
 /
 \end{lstlisting}
-Note that these are very standard MD settings except for the restraints.
+Note that these are very standard MD settings except for the line that defines the restraints.
 
@@
 
-\subsection{Applying GIST for the Streptavidin-Biotin complex}
-\subsubsection{System Preparation and Equilibration}
-If you want to skip this section, run 
-\begin{lstlisting}[style=bash]
-make equilibration-targets
-\end{lstlisting}
-in the code folder. This assumes that you have Python and Amber set up properly and that \software{pmemd.cuda} is in the PATH.
-
-The choice of initial structure is crucial to the GIST results.
-Crystal structures may not represent the dominant solution structure for a given target.
-In our example, we only use a single equilibrated structure.
-A more sophisticated approach would be to identify one or several dominant structures out of the ensemble.
-Depending on what properties you are interested in, it might be beneficial to run an MD simulation choose dominant structures through clustering.
-
-When starting from a crystal structure, care has to be taken during the preparation.
-Make sure to choose a proper protonation state for all titratable moeties.
-You will need to solvate your structure in a solvent box, which should be large enough to obtain unbiased long-range electrostatics.
-A good criterion for validating the box size is to check whether the water on the outside of the box behaves bulk-like according to the GIST output.
-
-When calculating solvation free energy differences between a complex and the corresponding monomers (the dissociated state), there are two options for dealing with configurational diversity in the structures.
-One can choose to sample the complex and dissociated states individually, or to sample only one of them and assume that there is sufficient overlap in the conformations.
-
-Here, we do not include any sampling of the conformational states.
-It is therefore important that the molecular conformations in the complex and dissociated states match exactly.
-Use \software{cpptraj} to produce biotin and streptavidin structures based on the complex by stripping the respective other molecule like this:
-\begin{lstlisting}[style=cpptraj]
-parm complex/solvated.parm7
-trajin complex/equil/EQUIL-DONE.rst
-strip ^2 parmout streptavidin/solvated.parm7
-trajout streptavidin/pre-equil.ncrst
-\end{lstlisting}
-(Adapt for biotin as necessary.)
-%We provide a script called \software{cpptraj\_remove\_mol.sh} that removes a single molecule from a topology and structure using \software{cpptraj} without altering the rest of the system.
-\todo{Explain cpptraj rather than our script.
-The script is for the automated workflow only.}
-
-Use the \software{equilibration.py} script (or your own) to perform short NVT and NpT equilibration runs on the complex structure.
-It is important to include unrestrained NpT steps, because restraining two molecules at the same time might interfere with the pressure equilibration.
-
-Then split the equilibrated structure into biotin and streptavidin parts using \software{cpptraj\_remove\_mol.sh}.
-After that, use \software{equilibration.py -R} to run short restrained 
-equilibrations (where only the water is allowed to move) on the individual 
-systems.
-By using restraints, we can make sure that the monomer conformations match the complex.
-
-\subsubsection{MD Simulation}
-The MD can be run as described above for each monomer and the complex.
-
-\subsubsection{GIST analysis of Streptavidin-Biotin}
-For easier post-processing, here we use grids that contain the whole molecule.
-Smaller grids encompassing only the binding site could be used for streptavidin and the complex, but this would complicate the analysis.
-%This could in theory be avoided for streptavidin and the complex, since only the surrounding of the biotin molecule (i.e., the binding pocket) is required for the analysis.
-Furthermore, we recommend that you center the molecule in space.
+\subsection{Running the GIST analysis for Streptavidin-Biotin Complex}
+For easier post-processing, we will use grids that contain the whole molecule.
+This could in theory be avoided for streptavidin and the complex, since only the surrounding of the biotin molecule (i.e., the binding pocket) is required for the analysis.
+We also recommend that you center the molecule in space.
 For each system (biotin, streptavidin, complex), produce a centered pdb file using \software{cpptraj}:
@@
 \end{lstlisting}
-
 Then, find the extent (minimum and maximum) of the x-, \mbox{y-,} and z-coordinates for each system.
@@
 What does gisttools do if the grid is too small? -vah }
-A buffer distance of \SI{15.0}{\angstrom} is reasonable for biotin, but smaller values are sufficient for the larger streptavidin and complex systems.
-
-Now, you can finally run the GIST analysis.
-Center the molecule as above, to make sure that its position matches the grid parameters you decided on.
-We recommend using either the PME or GPU implementation of GIST.
-For the GPU version, cpptraj needs to be compiled with the \inlinecode{-cuda} flag.
-While the GPU version is generally faster, the PME implementation matches the energy of the Amber MD engine more closely.
+We note that this centered grid is usually larger than it needs to be in some directions.
+If you want to save some extra memory and calculation time, feel free to align the grid more precisely to the minimum and maximum coordinates, or use a smaller grid that only covers the binding pocket.
+You can also rotate the molecule by its principal axes to fit the cuboid grid more exactly.
+However, this needs to be done \emph{before} the MD run, since rotating the trajectory damages the periodic box information.
+
+Now, you can finally run the actual GIST analysis. Center the molecule exactly in the same way as above, to make sure that its position matches the grid parameters that you decided on.
+We recommend using either the PME or GPU implementation of GIST, depending on whether you have a fast GPU. To use the GPU version of GIST, cpptraj needs to be compiled with the "-cuda" flag set.
+While the GPU version is generally faster, an advantage of the PME implementation is that it matches the energy calculation of the MD engine more closely.
 
@@
 autoimage !(:WAT) origin
-# optionally remove the "pme" option
+# remove the "pme" option if you want to use the GPU
 gist griddim 100 100 100 out gist.dat \
   refdens 0.03287 pme
-\end{lstlisting}
-
-\subsubsection{Reference Values and Radial Convergence}
+# other settings you might want to change
+# (those are the defaults)
+# gridcntr 0 0 0 gridspacn 0.5 temp 300
+\end{lstlisting}
+\subsection{Reference Values and Radial Convergence}
 All GIST quantities should be expressed relative to bulk.
-This is automatically done by \software{cpptraj} for the entropy, using the bulk density specified by \inlinecode{refdens}.
-The solute-solvent energy $E_{sw}$ naturally tends towards zero in bulk.
+In practice, this is not problematic for the entropies, since they are computed relative to a distribution of randomly oriented molecules at the reference density.
+The solute-solvent energy $E_{sw}$ is also no problem, because it naturally tends to zero in bulk. 
 However, the solvent-solvent energy $E_{ww}$ needs to be referenced.
 
-The Amber manual provides reference densities and energies for several solvents.
-However, the exact energy values are different when using PME-GIST, and also depend on the box size.
-For quantitative analyses, it is recommended to compute an exact reference energy.
-
-The most accurate method to compute a reference energy is to run a GIST calculation of the pure solvent using the same energy method (PME/GPU) and a similar box size, and then compute the average solvent-solvent energy per molecule.
+The Amber manual provides reference energies for several solvents.
+However, the exact reference value is different when using PME-GIST, and also depends on the box size.
+For exact analyses such as computing the total $\dgsolv$ of a compound, it is important to compute an exact reference energy.
+
+The most accurate method to compute a reference energy is to run a separate GIST calculations of the solvent without solute using the same energy method (PME/GPU) and a similar box size, and then compute the average solvent-solvent energy per molecule.
 
@@
 
-Alternatively, the solvent-solvent energy in a large GIST grid converges to the bulk value at sufficient distance to the solute.
-Therefore, the reference energy can be obtained by binning the voxels by their distance to the solute, and evaluate Equation~\ref{eq_ewwref} within each bin.
-If this value converges, this can be used as $E_{ww}^\textit{ref}$.
-
-\software{gisttools} contains functions to facilitate this analysis.
-There is also a method \inlinecode{detect\_reference\_value} that automatically tries to find the converged value.
-Although it has a simple convergence check built-in, it is always recommended to check the convergence by hand.
+Alternatively, one can note that the solvent-solvent energy in a sufficiently large GIST grid converges to the correct bulk value at large distance to the solute.
+This means that the reference energy can be obtained from a large GIST grid by binning the voxels by their distance to the solute, and evaluate Equation~\ref{eq_ewwref} within each bin.
+If this value converges to a constant, this can be used as $E_{ww}^\textit{ref}$.
+
+\software{gisttools} contains functionality to perform this analysis easily.
+There is also a method \inlinecode{detect\_reference\_value} that automatically tries to find the converged value from the rdf.
+Although it has a simple convergence check built-in, it is always recommended to check the convergence by hand as well.
 
 \begin{lstlisting}[style=python]
-from gisttools.gist import load_gist_file
+import gisttools as gt
 import matplotlib.pyplot as plt
-gist = load_gist_file(
+gist_obj = gt.gist.load_gist_file(
     "gist.dat", struct="solute-centered.pdb")
-bins, eww = gist.rdf(
-    "Eww_unref_norm", bins=100, rmax=20.,
-    normalize="norm")
+bins, eww = gist_obj.rdf(
+    "Eww_unref_norm", bins=100, rmax=20., normalize="norm")
 plt.plot(bins, eww)
-eww_auto = gist.detect_reference_value()
+eww_auto = gist_obj.detect_reference_value().values[0]
 plt.axhline(eww_auto, color='k')
@@
 	\includegraphics[width=0.8\linewidth]{figures/Eww_convergence.eps}
-	\caption{Convergence of $E_{ww}$ in the complex calculation with increasing distance to the solute (streptavidin-biotin).
-The horizontal line shows the automatically computed reference energy.}\label{fig_ewwref}
+	\caption{Convergence of $E_{ww}$ with increasing distance to the solute.
+The horizontal line shows the automatically computed reference energy.
+Computed from the complex calculation.}\label{fig_ewwref}
 \end{figure}
@@
 In \software{gisttools}, this is done simply by setting \inlinecode{gist.eww\_ref}.
-%In \software{gistpp}, this can be done as follows: 
-%\todo{Is there some official documentation for gistpp? We should probably link that somewhere, I feel that the code presented here is not really helpful/legible without further documentation. -vah}
-%\todo{This might become invalid with future GIST versions, depending on whether we adopt the COM as molecular center.} 
-%\todo{Better use the population.}
-
-%\begin{lstlisting}[style=bash]
-%rho0=0.03287465
-%neg_eww_ref=9.533
-%gistpp -op multconst -i gist-gO.dx -opt const $rho0 \
-%    -o gist-gO-abs.dx 
-%gistpp -op div -i gist-Eww-dens.dx -i2 gist-gO-abs.dx \
-%    -o gist-Eww-norm.dx
-%gistpp -op addconst -i gist-Eww-norm.dx \
-%    -opt const $neg_eww_ref -o gist-Eww-ref-norm.dx
-%gistpp -op mult -i gist-Eww-ref-norm.dx \
-%    -i2 gist-gO-abs.dx -o gist-Eww-ref-dens.dx
-%\end{lstlisting}
-
-Now, sum all free energy contributions to obtain the spatially resolved $\dgsolv$.
-In \software{gisttools}, this is done automatically and can be accessed using the \inlinecode{A\_dens} and \inlinecode{A\_norm} data rows.
-
-\newcommand{\coordinate}{\mathbf{r}}
-\begin{equation}
-	\dgsolv(\coordinate) = \Delta E_{ww}(\coordinate) + \Delta E_{sw}(\coordinate) - T\Delta S^\text{six}(\coordinate)
-\end{equation}
-
-%In \software{gistpp}, use the \inlinecode{add} command, with the same syntax as the \inlinecode{div} command used above.
-After that, check whether your free energy contributions become negligible far away from the solute.
+In \software{gistpp}, this can be done as follows: 
+\todo{Is there some official documentation for gistpp? We should probably link that somewhere, I feel that the code presented here is not really helpful/legible without further documentation. -vah}
+\todo{This might become invalid with future GIST versions, depending on whether we adopt the COM as molecular center.} 
+\todo{Better use the population.}
+
+\begin{lstlisting}[style=bash]
+rho0=0.03287465
+neg_eww_ref=9.533
+gistpp -op multconst -i gist-gO.dx -opt const $rho0 \
+    -o gist-gO-abs.dx 
+gistpp -op div -i gist-Eww-dens.dx -i2 gist-gO-abs.dx \
+    -o gist-Eww-norm.dx
+gistpp -op addconst -i gist-Eww-norm.dx \
+    -opt const $neg_eww_ref -o gist-Eww-ref-norm.dx
+gistpp -op mult -i gist-Eww-ref-norm.dx \
+    -i2 gist-gO-abs.dx -o gist-Eww-ref-dens.dx
+\end{lstlisting}
+
+Now, add all free energy contributions to obtain the spatially resolved $\dgsolv$.
+In \software{gisttools}, this is done automatically (simply use the \inlinecode{A\_dens} and \inlinecode{A\_norm} columns).
+In \software{gistpp}, use the \inlinecode{add} command, with the same syntax as the \inlinecode{div} command used above.
+After that, check whether your free energy contributions become negligible at high distances to the solute.
 In a plot like Figure~\ref{fig_ewwref} based on the \inlinecode{A\_dens} column, the values should tend to zero.
@@
 It the curve flattens out, the converged value is your final $\dgsolv$\@.
-Otherwise, you might need to tweak the $E_{ww}$ reference, or introduce a reference value for the entropy.
+If it does not, you might need to tweak the $E_{ww}$ reference, or introduce a reference value for the entropy (especially with older versions of \software{cpptraj}).
+% \todo{Eww or E$_{ww}$? Doesn't seem consistent so far. -vah} - let's go with all math mode $E_{ww}$ -fwa
 
 \begin{lstlisting}[style=python]
-from gisttools.gist import load_gist_file
+import gisttools as gt
 import matplotlib.pyplot as plt
 # adapt eww_ref!
-gist = load_gist_file("gist.dat",
+gist = gt.gist.load_gist_file("gist.dat",
     struct="solute-centered.pdb", eww_ref=-9.533)
@@
 The expected output of this code is shown in Figure~\ref{fig_radial_convergence}.
-Especially the solvent-solvent energy is not perfectly flat, indicating that the reference value is not optimal.
-%but is also complicated because the solute is charged.
-Both the entropy and $E_{sw}$ can be considered converged around \SI{10}{\angstrom}.
+Both the solvent-solvent energy and the entropy are not perfectly flat, indicating that the reference value is not optimal.
+The slow energy convergence indicates that a non-optimal reference value was used, but is also complicated by the fact that a charged solute is used.
+In this case, both the entropy and $E_{sw}$ can be considered converged around \SI{10}{\angstrom}.
 While this is not true for $E_{ww}$, we will see below that the energy difference upon binding converges more readily than its individual contributions.
-\todo{vah: Is that shown somewhere yet -> fwa?}
+\todo{vah: Is that shown somewhere yet?}
 
@@
 
-\subsubsection{Visualizing $\dgsolv$}
+\subsection{Visualizing $\dgsolv$}
 Next, we want to visualize the 3-dimensional contributions to the free energy of hydration, as well as the entropy and enthalpy contributions.
@@
 It might also be interesting to visualize the average energy of a water molecule at each grid voxel.
-This quantity is given by $2E_{ww} + E_{sw}$.
+This quantity is given by 2*Eww + Esw.
 The energy referencing leads to non-zero normalized values where the population is zero.
@@
 
-Visualize the free energy, the energy, and the entropy, at several isolevels.
+Visualize the free energy, the total energy, and the entropy, at several iso-levels.
 
@@
 
-You should find that there are several regions in the binding pocket that feature a high solvent energy and/or low entropy.
+You should find that there are several regions in the binding pocket that feature a high solvent energy and/or low solvent entropy.
 However, such regions also exist outside of the binding pocket, and there are also low-energy regions within the binding pocket.
-Therefore, the strong affinity to biotin cannot be explained only by hydration, and the direct interaction energy also needs to be taken into account.
-
-\subsubsection{Contribution of Hydration to Binding}
-In the next step, calculate the $\dgsolv$ contributions around biotin in each system (biotin, streptavidin, and complex).
-Note that our integration region does not fully reach into bulk, and some of the boundary regions are close to streptavidin. 
+Therefore, the strong affinity to biotin can not be explained solely by the hydration contribution, and the direct interaction energy also needs to be taken into account.
+
+\subsection{Contribution of Hydration to Binding}
+In the next step, calculate the $\dgsolv$ contributions around the biotin ligand in each of the three systems (biotin, streptavidin, and complex).
+Note that our integration region does not fully reach into bulk, but also comprises regions that are close to the streptavidin protein. 
 Therefore, our results for each system will depend on the exact position of the integration boundary. 
@@
 Although we kept the systems rigid, the center of mass might have been shifted during pressure equilibration.
-Therefore, we align the complex structure to streptavidin and use the shifted biotin coordinates to define the integration region.
-
-We recommend using a \software{Jupyter Notebook} for the following analyses.
+Therefore, we align the complex structure to the streptavidin system and use the shifted biotin coordinates to define the integration region.
+
+We recommend using a \software{Jupyter Notebook} for the following analyses, although they can also be done in a script or an interactive Python prompt.
 The code is not explained in detail here, but should be clear after reading the previous sections.
@@
 
-Subtract a reference entropy from the \inlinecode{dTSsix} columns.
-\todo{Update with newest gisttools -> fwa}
+Subtract a reference entropy from the dTSsix columns.
+\todo{Is there actually a reason why this is not in gisttools itself implemented? -vah}
 \begin{lstlisting}[style=python]
@@
 For the streptavidin integral, we align the complex structure onto streptavidin and then use the biotin positions as centers.
-Note that we use \software{mdtraj} here, since \software{gisttools} stores the reference structure as an \software{mdtraj} \inlinecode{Trajectory} object.
+Note that we use \software{mdtraj} here, since \software{gisttools} stores the reference structure as a \inlinecode{mdtraj.Trajectory} object, but any other library for post-processing of trajectories works as well.
 \begin{lstlisting}[style=python]
@@
 compl_x = select(compl.struct, biotin_mask).xyz[0] * 10.
-biotin_x = select(biotin.struct, biotin_mask).xyz[0]*10.
+biotin_x = select(biotin.struct, biotin_mask).xyz[0] * 10.
 aligned = compl.struct[:].superpose(strept.struct, \
@@
 Compute the energy (\inlinecode{Eall}) and entropy (\inlinecode{dTSsix}) contributions separately.
-You will find that the energy disfavors binding, because we do not yet include the interaction energy between biotin and streptavidin.
-
-You can also plot the difference between the complex and monomer contributions against the distance to biotin, using the complex coordinates for the holo structure.
-The expected result is shown in Figure~\ref{fig-dg-sums}.
-
-\begin{figure}
-	\centering
-	\includegraphics[width=1.0\linewidth]{figures/deltaG-difference.png}
-	\caption{TODO caption}\label{fig-dg-sums}
-\end{figure}
-
-Using the \inlinecode{energy} command in \software{cpptraj}, you can compute this interaction energy.
+You will find that the energy is strongly unfavorable to binding, because we do not yet include the interaction energy between biotin and streptavidin.
+
+Using the \inlinecode{energy} command in \software{cpptraj}, you can compute this energy difference.
 We recommend using PME in combination with PME-GIST, but not with GPU-GIST\@.
@@
 
-It has been shown \cite{Chen2021,Waibl2022-gist-solvents} that the solvation entropy in water is best approximated by 0.6 times the first order entropy.
+It has been shown in literature \cite{Chen2021,Waibl2022-gist-solvents} that the total solvation entropy in water is best approximated by 0.6 times the first order entropy.
 Also compute the scaled entropy and check the effect on the binding affinity.
@@
 %\todo{Table 1 is placed quite akwardly at the moment for me, maybe enforce its position? -vah. Ok? -fwa}
-
+\todo{this table is super confusing to me. What's "total"? Doesn't seem to be E-TS. Diff is also not the sum of other total entries. Shouldn't it be TS instead of S?-vah}
 \begin{table}[h]
 	\caption{Free energy contributions for the monomers and the dimer in kcal$\cdot$mol$^{-1}$.
-``Diff'' is calculated as $E^\text{internal} + \Delta E^\text{GIST} - \Delta S^\text{scaled}$.}\label{tab_dg_monomers_dimer}
+``total'' is calculated as $E^\text{internal} + \Delta E^\text{GIST} - \Delta S^\text{scaled}$.}\label{tab_dg_monomers_dimer}
 	\small
@@
 		\toprule
-		System       & E(internal) & $\Delta E^\textit{GIST}$ & $T\Delta S^\textit{GIST}$ & $T\Delta S^\textit{scaled}$ & Diff \\
+		System       & E(internal) & $\Delta E^\textit{GIST}$ & $\Delta S^\textit{GIST}$ & $\Delta S^\textit{scaled}$ & total \\
 		\midrule
@@
 However, we do not take the conformational transition of the streptavidin binding pocket into account.
-
 Prior computational studies suggest a $\Delta G$ of \SI{-26.6}{\kilo\calorie\per\mole} for the binding of biotin into the closed conformation of streptavidin \cite{Bansal2018-biotin}.
 This indicates that our result is in good agreement with literature.
-To improve the agreement with ITC measurements, the conformational changes of the binding pocket should be included.
-
-\subsubsection{Further steps}
-In this tutorial, we obtained a value for the binding of biotin into the closed conformation of the streptavidin binding pocket.
+However, the conformational changes of the binding pocket would need to be described to improve the agreement with experiments.
+\todo{Should we include a short introduction to other solvents with GIST? What about solvent/ion mixtures? Can we somehow distinguish python code from input files optically? -vah}
+\subsection{Further steps}
+In this tutorial, we obtained a value for the binding of biotin into the closed conformation of the streptavidin binging pocket.
 To investigate the effect of different conformations on the binding affinity, you can perform molecular dynamics simulations of the complex and/or monomers and perform GIST on multiple cluster representatives.
-To include the effect of lid closing in the binding process, free energy calculation methods such as umbrella sampling can be used.
-
-To speed up the calculation, smaller GIST grids could be used by focusing on the binding pocket, or by rotating the molecule by its principal axes to fit the cuboid grid more exactly.
-However, this should be done \emph{before} the MD, since rotating the trajectory damages the periodic box information \todo{@Dan: is this still the case?}.
+To include the effect of the lid closing transition in the binding process, free energy calculation methods such as umbrella sampling can be used.
+%\section{Content and links}
+%
+%A tutorial will normally draw on additional files and materials; clearly indicate where and how these are available, with links, and how they are being archived for the long-term and maintained so they stay current.
+%You will likely want to reference your GitHub repository as a central point to access all of this information, and then the GitHub repository may link out to other content as needed.
 
@@
 \subsection{GPU implementation}
-In Reference~\cite{Kraml2020}, a GPU implementation of the energy calculation was presented, which typically increases the speed by 1--2 orders of magnitude.
+In Reference~\cite{Kraml2020}, a GPU implementation of the energy calculation was implemented, which typically increases the speed by 1--2 orders of magnitude.
 
 If \software{cpptraj} is compiled with GPU support, the energy calculation uses the GPU automatically, without any additional input.
-However, PME-GIST is not supported on the GPU, and no GPU code will be used when specifying \inlinecode{pme}.
+However, PME-GIST is not supported on the GPU.
 
@@
 
-To run PME-GIST, the \inlinecode{pme} flag to \inlinecode{gist} can be used in \software{cpptraj}.
+To run a PME-GIST calculation, the \inlinecode{pme} flag to the \inlinecode{gist} action in \software{cpptraj} can be used.
 The output will contain two additional columns: \inlinecode{PME\_dens} and \inlinecode{PME\_norm}.
@@
 Furthermore, the \inlinecode{Eww} and \inlinecode{Esw} columns are also computed using PME, and are reported as usual.
-For typical use cases, we recommend using \inlinecode{Eww} and \inlinecode{Esw} over the \inlinecode{PME} column.
+For typical use cases, it is recommended to use \inlinecode{Eww} and \inlinecode{Esw} over the \inlinecode{PME} column.
 
@@
 
-To run a GIST calculation in a solvent other than water, use \inlinecode{solute} with a \software{cpptraj} selection mask.
-All other molecules will be treated as solvent.
-If there are multiple solvent species, also use the \inlinecode{solventmols} flag as described in Section~\ref{sec-salt-water}.
-
-The code will automatically choose three atoms per molecule to define the solvent orientation, and print this selection to the output.
-Sometimes, however, the automatic selection is not optimal.
-For instance, in methanol one can either incorporate the C--H bonds or the O--H bond in the selection.
+To run a GIST calculation in a solvent other than water, use the \inlinecode{solute} flag together with a \software{cpptraj} selection mask.
+Everything that is not in the \inlinecode{solute} mask is treated as solvent.
+If there are multiple solvent species, use the \inlinecode{solventmols} option as described in Section~\ref{sec-salt-water}.
+The code will automatically choose three atoms which are used to define the solvent orientation, and print this selection to the output.
+
+Sometimes, \software{cpptraj} does not automatically choose the optimal atoms.
+In methanol, for instance, one can either incorporate the C--H bonds or the O--H bond in the selection.
 The latter is probably more relevant since it can incorporate hydrogen bonding effects.
@@
 In \cite{Waibl2021-gist-salt}, an extension of GIST was presented that can use salt-water mixtures as a solvent.
-This allows treating salting-out effects, although it was shown that the salting-out coefficient is over-predicted by GIST because of the first-order entropy approximation.
+This allows treating salting-out effects, although it was shown that the salting-out coefficient is over-predicted by GIST unless higher-order entropies are incorporated.
 Generally, GIST should be able to treat arbitrary solvent mixtures as long as each molecule is sufficiently rigid.
 
-In the \software{cpptraj} implementation of GIST, the \inlinecode{solute} keyword specifies which components of the system are treated as a solute.
+The \software{cpptraj} implementation of GIST allows the user to specify which components of the system to treat as a solute, using the \inlinecode{solute} keyword.
 Everything else will be treated as solvent.
@@
 \end{lstlisting}
+
+A more sophisticated visualization, including the energy, entropy, and free energy of removing a single water molecule, is found in Figure~\ref{fig-binding-pocket.png}
+
+\begin{figure}
+	\centering
+	\includegraphics[width=\linewidth]{figures/binding-pocket-energy-levels.png}
+	\caption{Biotin in the streptavidin binding pocket, with isosurfaces based on the GIST calculation on the holo structure.
+Green: negative entropy ($T\Delta S$).
+Blue: negative energy (based on $E_{sw} + 2E_{ww}$).
+Red: positive energy.
+Isolevels are ($\pm$) 0.25, 0.5, 1.0, 1.5, 2.0, 2.5, and 3.0, except for the positive energy, where they are divided by two.}\label{fig-binding-pocket.png}
+\end{figure}
 
@@
 Additionally, there is a correction term $E_\text{self}$ (called $E_\text{corr}$ in the original publication \cite{Chen2021}), which corrects for the self-interaction of each solvent molecule in the reciprocal term.
+
 \begin{equation}
-	E_\textit{elec} = E_\text{dir} + E_\text{rec} + E_\text{self}
+	E_\textit{elec} = E_\textit{dir} + E_\textit{rec} + E_\textit{self}
 \end{equation}
-
 %\todo{How is E$_{dir}$ treated? Is there a cutoff as well (PME GIST paper seems to indicate so)? If so, we should probably note that. } - ok? -fwa
@@
 \begin{equation}
-	E_\text{lj} = E_\text{lj,\ short} +  E_\text{lj,\ corr}
+	E_\textit{lj} = E_\textit{lj,\ short} +  E_\textit{lj\_corr}
 \end{equation}
@@
 
-\begin{checklist}{Handling convergence problems}
-\begin{itemize}
-	\item Insufficient sampling leads to high noise in the results. Use higher numbers of frames.
-	\item Autocorrelation between frames leads to negative entropy in bulk. Increase simulation time and/or reduce the number of frames.
-\end{itemize}
-\end{checklist}
-
 %\begin{checklist}{A list}

### 2022-12-15 manuscript/manuscript.tex
Small changes

--- 2022-12-12 manuscript/manuscript.tex
@@
 \subsubsection{MD simulation}
-Now, you can run restrained MD simulations based on the equilibrated structures.
-Restraints should be used because IST theory is not defined for moving solutes.
+\todo{Move simulation part from complex to here}
+Now, you can run restrained MD simulations based on the equilibrated structures for subsequent GIST analyses.
+Restraints should be used during the MD because IST theory is not defined for moving solutes.
 We apply a harmonic restraint of \SI{100}{\kilo\calorie\per\mole\per\angstrom\squared} to all heavy atoms to keep the system in place.
@@
 \end{lstlisting}
-
+\subsubsection{GIST analysis}
+To start our GIST analysis of biotin, we'll first have to decide on the parameters defining the GIST grid during the analyis.
+We recommend that you center the biotin molecule in space:
+\begin{lstlisting}[style=cpptraj]
+	parm solvated.parm7
+	trajin md-01.nc 1 1 1
+	autoimage !(:WAT) origin
+	strip :WAT
+	trajout biotin-centered.pdb
+\end{lstlisting}
+
+To decide on a grid size, find the extent (minimum and maximum) of the x-, \mbox{y-,} and z-coordinates of the molecule.
+This can be done e.g., using the \software{mdtraj} package in Python.
+\begin{lstlisting}[style=python]
+	import mdtraj as md
+	pdb = md.load("biotin-centered.pdb")
+	# mdtraj converts to nm.
+	# *10 to convert back to Angstrom.
+	min = pdb.xyz[0].min(0) * 10.
+	max = pdb.xyz[0].max(0) * 10.
+	print(*list(zip("xyz", min, max)), sep="\n")
+\end{lstlisting}
+For a system centered at the origin with a spacing of \SI{0.5}{\angstrom} and a buffer of, e.g., \SI{15}{\angstrom} between the solute and the grid wall, you need $4.0\times (d+15)$ bins in the respective direction, where $d$ is the larger of $-$min and $+$max in the respective direction.
+To get bulk-like water properties at the grid border, a buffer distance of \SI{15.0}{\angstrom} is reasonable for biotin.
+
+Now, you can finally run the actual GIST analysis.
+Center the molecule as above, to make sure that its position matches the grid parameters you decided on.
+We recommend using either the PME or GPU implementation of GIST.
+For the GPU version, cpptraj needs to be compiled with the \inlinecode{-cuda} flag.
+While the GPU version is generally faster, the PME implementation matches the energy of the Amber MD engine more closely.
+
+\begin{lstlisting}[style=cpptraj]
+	parm solvated.parm7
+	trajin md-01.nc
+	autoimage !(:WAT) origin
+	# optionally remove the "pme" option
+	gist griddim 100 100 100 out gist.dat \
+	refdens 0.03287 pme
+\end{lstlisting}
+\subsubsection{Interpreting and visualizing the results}
+\todo{write this, show the various GIST properties here: Eww, Esw, S. Show how to calculate A and E with gisttools}
 \subsection{Applying GIST for the Streptavidin-Biotin complex}
@@
 In our example, we only use a single equilibrated structure.
-A more sophisticated approach would be to identify one or several dominant structures out of the ensemble.
-Depending on what properties you are interested in, it might be beneficial to run an MD simulation choose dominant structures through clustering.
-
-When starting from a crystal structure, care has to be taken during the preparation.
+A more sophisticated approach would identify one or several dominant structures out of the ensemble - either gathered from experimental data or by clustering an MD simulation.
+
+When starting from an experimental structure, care has to be taken during the preparation step.
 Make sure to choose a proper protonation state for all titratable moeties.
@@
 
-Use the \software{equilibration.py} script (or your own) to perform short NVT and NpT equilibration runs on the complex structure.
-It is important to include unrestrained NpT steps, because restraining two molecules at the same time might interfere with the pressure equilibration.
+Use the \software{equilibration.py} script (or your own procedure) to perform short NVT and NpT equilibration runs on the complex structure.
+Note that the equilibration script contains unrestrained NpT steps, because restraining two molecules at the same time might interfere with the pressure equilibration.
 

### 2022-12-15 manuscript/manuscript.tex
Small changes

@@
 
-Energy and entropy quantities are reported in two different ways in the GIST output.
-The \inlinecode{\_dens} quantities are normalized by the voxel volume and can be used to compute integrals.
-The \inlinecode{\_norm} quantities are instead normalized by the average number of solvents in the respective voxel, and are useful to visualize local water properties.
-
 \section{Tutorial}
@@
 Equilibrate your solvated biotin in an NVT or NpT ensemble.
-Even when running the simulation in NVT, it is advisable to include an NpT equilibration step to adjust the box size beforehand.
-Use enough equilibration time to allow the solvent to relax. 
-For water, \SI{0.5}{\nano\second} are usually sufficient to equilibrate the density, but narrow binding pockets can take longer.
+Even when running the production simulation in NVT, it is advisable to include an NpT equilibration step to adjust the box size beforehand.
+Use enough equilibration time to allow the water to relax around your biotin structure.
 
@@
 \end{lstlisting}
-This file will run a 100 ns NpT MD simulation while keeping all atoms except for hydrogens and waters restrained. Note that these are very standard MD settings except for the restraints.
+This file would run a 100 ns NpT MD simulation while keeping all atoms except for hydrogens and waters restrained. Note that these are very standard MD settings except for the restraints.
 
@@
 Then split the equilibrated structure into biotin and streptavidin parts using \software{cpptraj\_remove\_mol.sh}.
-After that, use \software{equilibration.py -R} to run short restrained 
+After that, you can use \software{equilibration.py -R} to run short restrained 
 equilibrations (where only the water is allowed to move) on the individual 
 systems.
-By using restraints, we can make sure that the monomer conformations match the complex.
-
-\subsubsection{MD Simulation}
-The MD can be run as described above for each monomer and the complex.
+
+\subsubsection{Running MD for GIST}
+
+Now, you can run restrained MD simulations based on the equilibrated structures for subsequent GIST analyses.
+Restraints should be used during the MD because IST theory is not defined for moving solutes.
+We apply a harmonic restraint of \SI{100}{\kilo\calorie\per\mole\per\angstrom\squared} to all heavy atoms to keep the system in place.
+We run \SI{100}{\nano\second} of simulation and store the conformations every \SI{100}{\pico\second}.
+A simulation length of at least \SI{20}{\nano\second} is recommended to obtain statistically independent snapshots of the water molecules.
+
+An example Amber input file for GIST might look like this:
+\begin{lstlisting}[style=amber-in]
+restrained 100 ns NpT
+&cntrl
+  ntx=5, irest=1,
+  ioutfm=1,
+  ntb=2, iwrap=1,
+  ntr=1, restraint_wt=100.0,
+  restraintmask="!@H=&!:WAT",
+  ntp=1, pres0=1.0, taup=1.0,
+  ntc=2, ntf=2,
+  ntt=3, tempi=300.0, temp0=300.0, gamma_ln=2,
+  nstlim=50000000, dt=0.002,
+  ntwr=50000, ntwx=5000, ntpr=5000,
+/
+\end{lstlisting}
+Note that these are very standard MD settings except for the restraints.
+
+Run the MD using \software{pmemd.cuda}:
+\begin{lstlisting}[style=bash]
+pmemd.cuda \
+	-O \
+	-i 100ns-npt-restraint.in \
+	-o md-01.out \
+	-p solvated.parm7 \
+	-c EQUIL-DONE.rst \
+	-x md-01.nc \
+	-r md-01.ncrst \
+	-ref EQUIL-DONE.rst
+\end{lstlisting}
 
@@
 
-Now, you can finally run the GIST analysis.
+Now, you can finally run the actual GIST analysis.
 Center the molecule as above, to make sure that its position matches the grid parameters you decided on.
@@
 
-\subsubsection{Reference Values and Radial Convergence}
+\subsection{Reference Values and Radial Convergence}
 All GIST quantities should be expressed relative to bulk.
-This is automatically done by \software{cpptraj} for the entropy, using the bulk density specified by \inlinecode{refdens}.
+This automatically done for the entropy, using the bulk density specified by \inlinecode{refdens}.
 The solute-solvent energy $E_{sw}$ naturally tends towards zero in bulk.
@@
 
-\software{gisttools} contains functions to facilitate this analysis.
+\software{gisttools} contains functionality to perform this analysis easily.
 There is also a method \inlinecode{detect\_reference\_value} that automatically tries to find the converged value.
@@
 
-Now, sum all free energy contributions to obtain the spatially resolved $\dgsolv$.
-In \software{gisttools}, this is done automatically and can be accessed using the \inlinecode{A\_dens} and \inlinecode{A\_norm} data rows.
+Now, add all free energy contributions to obtain the spatially resolved $\dgsolv$.
+In \software{gisttools}, this is done automatically, and can be accessed using the \inlinecode{A\_dens} and \inlinecode{A\_norm} data rows. \todo{norm and dens not defined. }
 
@@
 While this is not true for $E_{ww}$, we will see below that the energy difference upon binding converges more readily than its individual contributions.
-\todo{vah: Is that shown somewhere yet -> fwa?}
+\todo{vah: Is that shown somewhere yet?}
 
@@
 However, we do not take the conformational transition of the streptavidin binding pocket into account.
-
 Prior computational studies suggest a $\Delta G$ of \SI{-26.6}{\kilo\calorie\per\mole} for the binding of biotin into the closed conformation of streptavidin \cite{Bansal2018-biotin}.
 This indicates that our result is in good agreement with literature.
-To improve the agreement with ITC measurements, the conformational changes of the binding pocket should be included.
-
-\subsubsection{Further steps}
-In this tutorial, we obtained a value for the binding of biotin into the closed conformation of the streptavidin binding pocket.
+However, the conformational changes of the binding pocket would need to be described to improve the agreement with experiments.
+\todo{Should we include a short introduction to other solvents with GIST? What about solvent/ion mixtures? Can we somehow distinguish python code from input files optically? -vah}
+\subsection{Further steps}
+In this tutorial, we obtained a value for the binding of biotin into the closed conformation of the streptavidin binging pocket.
 To investigate the effect of different conformations on the binding affinity, you can perform molecular dynamics simulations of the complex and/or monomers and perform GIST on multiple cluster representatives.
-To include the effect of lid closing in the binding process, free energy calculation methods such as umbrella sampling can be used.
-
-To speed up the calculation, smaller GIST grids could be used by focusing on the binding pocket, or by rotating the molecule by its principal axes to fit the cuboid grid more exactly.
-However, this should be done \emph{before} the MD, since rotating the trajectory damages the periodic box information \todo{@Dan: is this still the case?}.
+To include the effect of the lid closing transition in the binding process, free energy calculation methods such as umbrella sampling can be used.
+
+Smaller GIST grids could be used by focusing on the binding pocket, or by rotating the molecule by its principal axes to fit the cuboid grid more exactly.
+However, this should be done \emph{before} the MD, since rotating the trajectory damages the periodic box information.
+%\section{Content and links}
+%
+%A tutorial will normally draw on additional files and materials; clearly indicate where and how these are available, with links, and how they are being archived for the long-term and maintained so they stay current.
+%You will likely want to reference your GitHub repository as a central point to access all of this information, and then the GitHub repository may link out to other content as needed.
 
@@
 
-\subsection{Visualizing DX files}
+\subsection{Loading a DX file in Pymol }
 A minimal PyMOL input to visualize a dx file is shown here.
@@
 The oxygen density is visualized at an isolevel of 2 (i.e., twice the reference density).
-The expected output is shown in Figure~\ref{fig-streptavidin_gO}.
 
@@
 	\includegraphics[width=1.0\linewidth]{figures/streptavidin_gO_high_surf.png}
-	\caption{Oxygen density around streptavidin at an isolevel of twice the reference density}\label{fig-streptavidin_gO}
+	\caption{Oxygen density around strepativin at twice the reference bulk density}\label{streptavidin_gO}
 \end{figure}
@@
 GIST provides a very natural way of creating a SASA using an isosurface of the oxygen density gO.
-It can be visualized in PyMOL.
-The expected output is shown in Figure~\ref{fig-streptavidin_sasa}.
+It can be visualized in PyMOL as follows:
 
@@
 	\includegraphics[width=1.0\linewidth]{figures/sasa-simple.png}
-	\caption{SASA for streptavidin focused on the binding pocket}\label{fig-streptavidin_sasa}
+	\caption{SASA for streptavidin focused on the binding pocket}\label{streptavidin_sasa}
 \end{figure}
 
-Here, the SASA was generated from a simulation of the holo protein structure.
+In the above figure, the SASA was generated from a simulation of the protein in the absence of the ligand.
 When the ligand is placed in the binding pocked, all atoms are within the SASA.
 
-\subsection{Defining the binding pocket}
+\subsection{Region around the binding pocket}
 Specific regions in a protein can be defined to calculate the thermodynamic quantities for that region.
@@
 The resulting file is a binary file and can be visualized using \software{pymol}.
-\todo{Not GISTPP}
 
@@
 
-Here, the binding pocket includes all voxels within \SI{3.5}{\angstrom} of the heavy atoms of biotin.
+In this example, the region defined in the binding pocket includes all voxels that are within \SI{3.5}{\angstrom} of the heavy atoms of biotin.
 
@@
 GIST is an implementation of inhomogeneous fluid solvation theory (IST) \cite{Lazaridis1998} that discretizes the free energy of solvation $\dgsolv$ on a three-dimensional grid. 
-It was first devised by Nguyen et al.\ \cite{Nguyen2012} and its implementation in \software{cpptraj} was thoroughly described by Ramsey et al.\ \cite{Ramsey2016}.
+It was first devised by Nguyen et al.
+\cite{Nguyen2012} and its implementation in \software{cpptraj} was thoroughly described by Ramsey et al.
+\cite{Ramsey2016}.
+% to overcome the limitation of IST which was commonly used for regions of high water occupancy and its implementation in \software{cpptraj} was thoroughly described in \cite{Ramsey2016}.
 Here, we only provide a short overview of the theory behind GIST.
@@
 \subsection{Solvation Entropy}
+
+%The solvation entropy of a system is totality of the contribution of the  solute-water entropy and the water-water entropy. -- this is inconsistent with IST theory (fwa)
 
@@
 
-The solvent density is expressed relative to the bulk density $\rho^0$. %$g_\textit{sw}\left(\textbf{r},\omega \right)$ is the solute-water 
+The solvent density is expressed relative to the bulk density $\rho^\textit{o}$. %$g_\textit{sw}\left(\textbf{r},\omega \right)$ is the solute-water 
 %pair correlation function in the solute's frame of reference, \textbf{r} is the coordinates of the oxygen atom of the water molecule,  
@@
 The entropy integral is also normalized by $8\pi^2$, which is the volume of the orientational space.
-In bulk or at a high distance to the solute, the quantity $g_\textit{sw}\left(\textbf{r},\omega \right)$ approaches one, which leads to zero first-order entropy. 
-Therefore, no reference entropy is needed unless there is a numeric bias, e.g., when the MD frames are statistically dependent because the time between them is insufficient.
-\begin{equation}
-	\Delta S_\text{solv}
+In bulk or at a high distance to the solute, the quantity $g_\textit{sw}\left(\textbf{r},\omega \right)$ is unity which leads to the first-order entropy of 
+bulk to be zero. 
+Therefore, no reference entropy needs to be computed unless there is a numeric bias, e.g., when the MD frames are not statistically independent due to insufficient spacing.
+%Also, this quantity reaches unity as the distance from the solute molecule increases which leads for the calculation of 
+%the solvation entropy to be an approximation of the integral around the solute atom. 
+\begin{equation}
+	\Delta S_\textit{solv}
 	\approx \Delta S_\textit{sw}
 	\newcommand{\gro}{g_{sw}(\textbf{r},\omega)}
-	\equiv -R \frac{\rho^0}{8\pi^\textit{2}} \int \gro \ln\gro d\textbf{r}d\omega
-\end{equation}
-
-\subsubsection{Entropy Calculations in Cpptraj}
-In \software{cpptraj}, the calculation of solvation entropy is handled by two methods.
-
-In the first method, the solvation entropy is broken down into translational and orientational contributions.
+	\equiv -R \frac{\rho^\textit{0}}{8\pi^\textit{2}} \int \gro \ln\gro d\textbf{r}d\omega
+\end{equation}
+\todo{We are inconsistent regarding R/kB; vah: Fixed. is it - tho? Papers seem to disagre }
+
+The solvation entropy is further broken down in terms of the contribution from translational and orientational terms.
 \begin{equation}
@@
 \end{equation}
-This is exact assuming that the distribution of the orientation $\omega$ is constant within a voxel $k$, but might converge slower than computing the entropy from the combined space (see below).
-
-The second method directly calculates the solvation entropy by evaluating the six-dimensional integral (3 for position and 3 for orientation).
-
-\begin{equation}
-g_\textit{vox} \left( \textbf{r}, \omega \right) \approx g_\textit{vox}(\textbf{r}) g_\textit{vox}(\omega)
-\end{equation}
-
+
+\subsection{Entropy Calculations in Cpptraj}
+In \software{cpptraj} the calculation of  solvation entropy is handled by two methods.
+
+The first method assumes that the distribution of the orientation $\omega$ is constant within a voxel $k$.
+This allows for the splitting of the contributions of the translational and the orientational entropy. 
 A first nearest neighbor (NN) approach is used to evaluate the density.
-The densities are computed relative to a homogeneous distribution of solvent molecules, given a bulk density $\rho^0$.
+%The translational density $g_{NN}(\textbf{r})$ is expressed in reference to a homogeneous bulk solvent at the reference density, while the orientational density $g_{NN}(\omega)$ is expressed to a random placement of the number of solvents found in this voxel.
+Both the translational ($g_{NN}(\textbf{r})$) and orientational ($g_{NN}(\omega)$) density are computed relative to a bulk solvent at the reference density $\rho^0$.
 
@@
 \begin{equation}
-	S_{k}^\textit{trans} = -R \left( \gamma + \frac{1}{N_\textit{k}} \sum _{i=1}^{N_k} \ln g_{NN, \textit{i}}(\textbf{r}) \right),
-\end{equation}
+	S_{k}^\textit{trans} = -R \left( \gamma + \frac{1}{N_\textit{k}} \sum _{i=1}^{N_k} \ln g_{NN, \textit{i}}(\textbf{r}) \right)
+\end{equation}
+and
 \begin{equation}
@@
 \end{equation}
-and
-\begin{equation}
-S_\textit{k}^\text{six} = -R \left( \gamma + \frac{1}{N_\textit{k}} \sum _{i=1}^{N_k} \ln g_{NN, \textit{i}}(\textbf{r},\omega) \right),
-\end{equation}
-where $N_\textit{k}$ is the number of water molecules found in voxel k, $\gamma$ is Euler's constant accounting for the bias of the naive entropy estimator, and $g_\textit{NN}$ is the nearest neighbor estimate of the density.
+where $N_\textit{k}$ is the number of water molecules found in voxel k, $\gamma$ is Euler's constant accounting for the bias of the naive entropy estimator, and $g_\textit{NN}$ is the nearest neighbor estimate.
+
+The second method directly calculates the solvation entropy by evaluating the six-dimensional integral (3 for position and 3 for orientation) 
+using the nearest-neighbor approach. 
+% Unlike the first method, the nearest neighbor estimate for the sixth-order entropy considers all the water found within the same and neighboring voxels.
+-- misleading because this is also done for S_trans (fwa)
+
+\begin{equation}
+	g_\textit{vox} \left( \textbf{r}, \omega \right) \approx g_\textit{vox}(\textbf{r}) g_\textit{vox}(\omega)
+\end{equation}
+
+\begin{equation}
+	S_\textit{k} = -R \left( \gamma + \frac{1}{N_\textit{k}} \sum _{i=1}^{N_k} \ln g_{NN, \textit{i}}(\textbf{r},\omega) \right)
+\end{equation}
 
 \subsection{Solvation Energy}
-The solvation energy is calculated from the water-water and water-solute interactions from the force field.
+The solvation energy is readily calculated by summing up the contributions of the water-water interaction and the water-solute interaction from the force field.
 \begin{equation}
@@
 To localize $E_{sw}$ on the three-dimensional grid, every energy term is assigned to the voxel $k$ holding the solvent, and the expectation value in Equation~\ref{eq-esw} is computed per voxel.
-Similar to the entropy integrals, the solute-water solvation energy decays with increasing distance to the solute.
-Hence, it can be approximated by local spatial integrals.
-
-The solvent-solvent energy $E_{ww}$ is computed similarly.
-It can also be expressed in terms of density functions, but is practically computed as a sum over solvent-solvent interactions per voxel $k$.
+Similar to the entropy integrals, the solute-water solvation energy integral also decays with increasing distance from the solute.
+Hence, the solvation energy can be approximated by local spatial integrals.
+
+The solvent-solvent energy $E_{ww}$ is computed in a similar manner.
+Again, it can be expressed in terms of density functions, but is practically computed as a sum over solvent-solvent interactions where one solvent is in voxel $k$.
 In contrast to $E_{sw}$, $E_{ww}$ does not tend to zero in bulk.
-Therefore, a reference corresponding to the average energy of a bulk water molecule needs to be subtracted.
-The referenced solvent-solvent energy will be denoted as $E_{ww}^\text{corr}$.
+Therefore, a reference energy corresponding to the average energy of a water molecule in bulk needs to be subtracted.
+
+%However, there is a complication arising from this approximation.
+Water molecules outside the region of interest will also be accounted for in the pairwise water term.
+For example, given two different regions $R_a$ and $R_b$, will not sum up to the total energy of the two regions due to double counting.
 
@@
 	\begin{aligned}
-		\Delta E_{ww}^\text{corr}(\textbf{r}) \equiv & \left(\frac{1}{8\pi^2} \right)^2 \rho^o \int g_{sw}(\omega|\textbf{r}) \\
-			& \times \left[g_{sw}(\textbf{r}^\prime , \omega^\prime) - g^0_{ww} (\textbf{r}, \omega, \textbf{r}^\prime, \omega^\prime )\right] \\
-			& \times U_{ww}(\textbf{r}, \omega, \textbf{r}^\prime, \omega^\prime)d\omega d\textbf{r}^\prime d\omega^\prime \\
-		\Delta E_{ww,k}^\text{corr} =& \frac{1}{N_k} \left\langle \sum_i^{\text{solvent},k} \; \sum_{j \neq i}^\text{solvent} U_{ij}\right\rangle - \left\langle E_{ww}^\text{bulk}\right\rangle
+		\Delta E_{ww}(\textbf{r}) \equiv & \left(\frac{1}{8\pi^2} \right)^2 \rho^o \int g_\textit{sw}(\omega|\textbf{r}) \\
+		& \times \left[g_\textit{sw}(\textbf{r}^\prime , \omega^\prime) - g^o_\textit{ww} (\textbf{r}, \omega, \textbf{r}^\prime, \omega^\prime )\right] \\
+		& \times U_\textit{ww}(\textbf{r}, \omega, \textbf{r}^\prime, \omega^\prime)d\omega d\textbf{r}^\prime d\omega^\prime \\
+		\Delta E_{ww,k} =& \frac{1}{N_k} \left\langle \sum_i^{\text{solvent},k} \; \sum_{j \neq i}^\text{solvent} U_{ij}\right\rangle - \left\langle E_{ww}^\text{bulk}\right\rangle
 	\end{aligned}
@@
 
-\subsubsection{Interpreting the energy values}
+%AFAIK, this was never done, since the pair interaction matrix Eij would have to be written out.
+%There are two ways in which this double counting was accounted for in the original paper.
+First, using the water-displacement method.
+This method is applicable when a determining the energetic cost of kicking out a water molecule.
+%\begin{equation}
+%	\Delta E^R_\textit{ww} = n^\textit{R} E^\textit{bulk}_\textit{ww} - E^\textit{R,corr}_\textit{ww}
+%\end{equation}
+
+%The second method uses normalized-water properties to account the difference in water properties of two different regions.
+The normalized water property is defined as the water property per water molecule.
+%\begin{equation}
+%	\Delta E^\textit{R,norm}_\textit{ww} = E^\textit{R,norm}_\textit{ww} - 2E^\textit{bulk}_\textit{ww}
+%\end{equation}
+
 When computing the total $E_{ww}$ of the system, double-counting of interactions must be avoided.
@@
 \begin{equation}
-	\Delta E^\text{total}_{ww} = \sum_{k}^\text{voxels} \frac{E_{ww,k}^\text{corr}}{2}
-%	\Delta E^R_{ww} = n^R E^\text{bulk}_{ww} - E^\textit{R,corr}_{ww}
+	\Delta E^R_\textit{ww} = n^\textit{R} E^\textit{bulk}_\textit{ww} - E^\textit{R,corr}_\textit{ww}
 \end{equation}
@@
 \begin{equation}
-	\Delta E^\text{total}_{ww} = \sum_{k}^{\text{voxels in }R} E_{ww,k}^\text{corr}
-%	\Delta E^\textit{R,norm}_\textit{ww} = E^\textit{R,norm}_\textit{ww} - 2E^\textit{bulk}_\textit{ww}
-\end{equation}
-
-When the region $R$ comprises more than one molecule, interactions within this region are double-counted while interactions to the outside are not.
-This could be solved using the solvent-solvent energy between each pair of voxels $E_{ww,kl}$.
+	\Delta E^\textit{R,norm}_\textit{ww} = E^\textit{R,norm}_\textit{ww} - 2E^\textit{bulk}_\textit{ww}
+\end{equation}
+
+When the region of interest $R$ comprises more than one molecule, interactions within this region are double-counted while interactions to the outside are not.
+To take this into account, the solvent-solvent energy between each pair of voxels $E_{ww,kl}$ is needed.
 While this is supported by the standard GIST implementation, it is rarely done due to the large size of the $E_{ww,kl}$ matrix.
 
-Integrating GIST values over the whole grid corresponds to a process where all the solvent is removed into bulk.
-However, when integrating over a small region, this process is less well-defined, since it is unclear to what extent the remaining solvent would reorganize.
-However, this depends on the local environment and on \emph{what} the solvent would be replaced by.
+We note that replacement of a small or medium solvent region would lead to reorganization of the surrounding solvent.
+However, this depends on the local environment and on \emph{what} the molecule would be replaced by.
 Therefore, the effect of reorganization can not be directly included in the GIST method, but must be judged for each case individually.
-
-\subsubsection{PME energy}
+This is not a problem when computing the total free energy of solvation $\dgsolv$, since no additional boundary is created.
+\todo{This is a bit unclear to me, could we elaborate on this a bit more? What are we asking of the reader here, how would one consider this effect? -vah}
+
+\subsection{PME implementation}
 
 In the original version of GIST, the energies are calculated based on the the minimum image convention.
-In PME-GIST, the electrostatic energy $E_\text{elec}$ is calculated using the particle mesh Ewald (PME) method, which yields energies that are highly consistent with the Amber MD engine \cite{Chen2021}.
-The Lennard-Jones part $E_\text{lj}$ is computed separately in direct space.
+In PME-GIST, the electrostatics are calculated using the particle mesh Ewald (PME) method, which yields energies that are highly consistent with the treatment in the Amber MD engine \cite{Chen2021}.
+The electrostatic and Lennard-Jones parts $E_\text{elec}$ and $E_\text{lj}$ are computed separately.
 
@@
 
-During the electrostatics calculation, the system is treated as periodic and the energy is split into a short-range term $E_\text{dir}$, which is calculated in direct space using a distance cutoff, and a long-range term $E_\text{rec}$, which is calculated in reciprocal space.
+In PME-GIST, the system is treated as periodic, and the electrostatic energy is split into a short-range term $E_\text{dir}$, which is calculated in direct space using a distance cutoff, and a long-range term $E_\text{rec}$, which is calculated in reciprocal space.
 Additionally, there is a correction term $E_\text{self}$ (called $E_\text{corr}$ in the original publication \cite{Chen2021}), which corrects for the self-interaction of each solvent molecule in the reciprocal term.
-\begin{equation}
-	E_\textit{elec} = E_\text{dir} + E_\text{rec} + E_\text{self}
-\end{equation}
-The short-range Lennard-Jones contribution is computed in the direct space using a distance cutoff.
+
+\begin{equation}
+	E_\textit{elec} = E_\textit{dir} + E_\textit{rec} + E_\textit{self}
+\end{equation}
+%\todo{How is E$_{dir}$ treated? Is there a cutoff as well (PME GIST paper seems to indicate so)? If so, we should probably note that. } - ok? -fwa
+The short-range Lennard-Jones contribution is computed in the direct space, using a distance cutoff.
 Furthermore, a long-range correction term is computed that accounts for the contributions above this cutoff assuming a homogeneous distribution of particles.
@@
 \begin{equation}
-	E_\text{lj} = E_\text{lj,\ short} +  E_\text{lj,\ corr}
-\end{equation}
-
-\section{Cheat sheet}
+	E_\textit{lj} = E_\textit{lj,\ short} +  E_\textit{lj\_corr}
+\end{equation}
+
+\section{Cheatsheet}
 %This will be a cheatsheet containing all options for cpptrajs GIST implementation
-The checklist ``The CPPTRAJ GIST action'' provides a cheat sheet summarizing the important options to the GIST action and its output, as well as frequently used equations.
-
 \subsection{Installation}
 GIST calculations in cpptraj can be sped up through various levels of 
-parallelization.
-If cpptraj was compiled with the '-cuda' configure flag set, the GPU accelerated version of GIST is used automatically. 
-Furthermore, if cpptraj was compiled with OpenMP support, the entropy calculation will be parallelized.
-
-For detailed instructions on how to install \software{cpptraj}, we refer the reader to the AmberMD GitHub page \url{https://github.com/Amber-MD/cpptraj}.
+parallelization. When cpptraj was compiled with the '-cuda' configure flag set, 
+the GPU accelerated version of GIST is used automatically.
 
@@
 
-\begin{checklist}{The CPPTRAJ GIST action}\label{cheatsheet}
+\begin{checklist}{The CPPTRAJ GIST action}
 

### 2022-12-19 manuscript/manuscript.tex
Ongoing restructuring
Add some clarifications concerning the output
Change some code to not overwrite preexisting python functions
Modify complex section to be inline with biotin

--- 2022-12-15 manuscript/manuscript.tex
@@
 
-Energy and entropy quantities are reported in two different ways in the GIST output.
+The output is organized by voxel and sorted according to their x,y, and z coordinates, describing the center point of the voxel.
+Energy and entropy quantities are reported in two different ways ´
 The \inlinecode{\_dens} quantities are normalized by the voxel volume and can be used to compute integrals.
-The \inlinecode{\_norm} quantities are instead normalized by the average number of solvents in the respective voxel, and are useful to visualize local water properties.
-
+The \inlinecode{\_norm} quantities are instead normalized by the number of solvents in the respective voxel averaged over all considered frames, and are useful to visualize local water properties.
+Therefore, to convert from a \inlinecode{\_norm} quantity $X^k_{norm}$ in voxel k to a \inlinecode{\_dens} quantity $X^k_{dens}$ the following equation is applied:
+\begin{equation}
+	X^k_{dens} = \frac{X^k_{norm}}{V^k} \cdot \frac{N^k_{solvent}}{N_{frames}} 
+\end{equation}
+with $V^k$, the volume of voxel $k$, $N^k_{solvent}$, the solvent population in voxel $k$, and $N_{frames}$, the number of simulation frames considered in the GIST analysis.
+Acoordingly, converting $X^k_{dens}$ to  $X^k_{norm}$ is possible using the following equation:
+\begin{equation}
+	X^k_{norm} = X^k_{dens} \cdot V^k \cdot \frac{N_{frames}}{N_{frames}N^k_{solvent}} 
+\end{equation}
 \section{Tutorial}
@@
 	# *10 to convert back to Angstrom.
-	min = pdb.xyz[0].min(0) * 10.
-	max = pdb.xyz[0].max(0) * 10.
-	print(*list(zip("xyz", min, max)), sep="\n")
+	min_xyz = pdb.xyz[0].min(0) * 10.
+	max_xyz = pdb.xyz[0].max(0) * 10.
+	print(*list(zip("xyz", min_xyz, max_xyz)), sep="\n")
 \end{lstlisting}
@@
 \subsubsection{Interpreting and visualizing the results}
-\todo{write this, show the various GIST properties here: Eww, Esw, S. Show how to calculate A and E with gisttools}
+\todo{write this, show the various GIST properties here: Eww, Esw, S.}
 \subsection{Applying GIST for the Streptavidin-Biotin complex}
+In this section we will apply GIST to the full complex and discuss each step of the process in more detail to outline pitfalls.
+We will afterwards introduce advanced analysis workflows which allow you to modify and better visualize the resulting data.
 \subsubsection{System Preparation and Equilibration}
@@
 Smaller grids encompassing only the binding site could be used for streptavidin and the complex, but this would complicate the analysis.
-%This could in theory be avoided for streptavidin and the complex, since only the surrounding of the biotin molecule (i.e., the binding pocket) is required for the analysis.
 Furthermore, we recommend that you center the molecule in space.
-For each system (biotin, streptavidin, complex), produce a centered pdb file using \software{cpptraj}:
-\begin{lstlisting}[style=cpptraj]
-parm solvated.parm7
-trajin md-01.nc 1 1 1
-autoimage !(:WAT) origin
-strip :WAT
-trajout solute-centered.pdb
-\end{lstlisting}
-
-Then, find the extent (minimum and maximum) of the x-, \mbox{y-,} and z-coordinates for each system.
-This can be done e.g., using the \software{mdtraj} package in Python.
-\begin{lstlisting}[style=python]
-import mdtraj as md
-pdb = md.load("solute-centered.pdb")
-# mdtraj converts to nm.
-# *10 to convert back to Angstrom.
-min = pdb.xyz[0].min(0) * 10.
-max = pdb.xyz[0].max(0) * 10.
-print(*list(zip("xyz", min, max)), sep="\n")
-\end{lstlisting}
-Then, decide on grid dimensions for every system.
-For a system centered at the origin with a spacing of \SI{0.5}{\angstrom} and a buffer of, e.g., \SI{15}{\angstrom} between the solute and the grid wall, you need $4.0\times (d+15)$ bins in the respective direction, where $d$ is the larger of $-$min and $+$max in the respective direction.
-\todo{ This is handled differently by FindCentroid.py: d = max - min; voxelnr = 
-d*5 for x,y,z each; we should handle it similiar or at least mention 
-FindCentroid.py here. 
-What does gisttools do if the grid is too small? -vah }
-A buffer distance of \SI{15.0}{\angstrom} is reasonable for biotin, but smaller values are sufficient for the larger streptavidin and complex systems.
-
-Now, you can finally run the GIST analysis.
-Center the molecule as above, to make sure that its position matches the grid parameters you decided on.
-We recommend using either the PME or GPU implementation of GIST.
+Calculate the GIST grid parameters as outlined in the biotin section for the complex and streptavidin and run the GIST analysis.
+For larger systems like whole proteins you can usually get away with a smaller buffer region compared to small molecules.
+To test whether your grid is large enough look at solvent properties close to the grid edges, e.g. the solvent-solvent energies.
+If these correspond to the respective bulk properties, it can be assumed that your grid is large enough to capture all perturbations of the solute on the solvent.
+Note that a different box-size or number of solvent molecules can skew the bulk properties slightly compared to the reference values provided above.
+In that case, checking for convergence of the solvent properties near the grid borders is sufficient but you might want to modify your reference values according to your results then as described in the next section.
+
+For the GIST analysis, we recommend using either the PME or GPU implementation of GIST.
 For the GPU version, cpptraj needs to be compiled with the \inlinecode{-cuda} flag.
 While the GPU version is generally faster, the PME implementation matches the energy of the Amber MD engine more closely.
-
-\begin{lstlisting}[style=cpptraj]
-parm solvated.parm7
-trajin md-01.nc
-autoimage !(:WAT) origin
-# optionally remove the "pme" option
-gist griddim 100 100 100 out gist.dat \
-  refdens 0.03287 pme
-\end{lstlisting}
 
@@
 \end{lstlisting}
-% \todo{\$\textbackslash AA\$ should produce an angstrom sign in matplotlib (what a pain to have this printed out in \LaTeX\ tho) -vah} -- Å produces an angstrom sign ;-)
+
 Compute the energy (\inlinecode{Eall}) and entropy (\inlinecode{dTSsix}) contributions separately.
@@
 The expected result is shown in Figure~\ref{fig-dg-sums}.
-
-\begin{figure}
-	\centering
-	\includegraphics[width=1.0\linewidth]{figures/deltaG-difference.png}
-	\caption{TODO caption}\label{fig-dg-sums}
-\end{figure}
+\todo{This figure is missing -vah}
+%\begin{figure}
+%	\centering
+%	\includegraphics[width=1.0\linewidth]{figures/deltaG-difference.png}
+%	\caption{TODO caption}\label{fig-dg-sums}
+%\end{figure}
 

### 2022-12-19 manuscript/manuscript.tex

@@
 \end{lstlisting}
+
 This file will run a 100 ns NpT MD simulation while keeping all atoms except for hydrogens and waters restrained. Note that these are very standard MD settings except for the restraints.
@@
 \end{lstlisting}
-
 To decide on a grid size, find the extent (minimum and maximum) of the x-, \mbox{y-,} and z-coordinates of the molecule.
@@
 \subsubsection{Interpreting and visualizing the results}
-\todo{write this, show the various GIST properties here: Eww, Esw, S.}
+
+\todo{write this, show the various GIST properties here: Eww, Esw, S. Show how to calculate A and E with gisttools}
+
 \subsection{Applying GIST for the Streptavidin-Biotin complex}

### 2022-12-20 manuscript/manuscript.tex
Add crystal water pitfalls. Start writing gisttools and pymol intro.

@@
 \section{Running GIST}
+\label{sec:running_GIST}
 %\subsection{Running the GIST analysis}
@@
 \todo{Move simulation part from complex to here}
-Now, you can run restrained MD simulations based on the equilibrated structures for subsequent GIST analyses.
+Now, you can run a restrained MD simulation based on the equilibrated structure of biotin for the subsequent GIST analysis.
 Restraints should be used during the MD because IST theory is not defined for moving solutes.
@@
 \subsubsection{Interpreting and visualizing the results}
+As outlined in section \ref{sec:running_GIST}
 
@@
 Make sure to choose a proper protonation state for all titratable moeties.
+
 You will need to solvate your structure in a solvent box, which should be large enough to obtain unbiased long-range electrostatics.
 A good criterion for validating the box size is to check whether the water on the outside of the box behaves bulk-like according to the GIST output.
+Note that due to the solute restraints, some important solvent sites might be unavailable to exchange with the bulk solvent.
+Make sure to explicitly place water at such sites, if they are important to the process of interest e.g. waters in confined binding pockets, as they might be missed by automated solvation procedures.
+It is good practice to check your restrained MD simulation or solvent populations obtained by GIST afterwards to confirm wether all important solvent sites were properly solvated.
 

### 2022-12-20 manuscript/manuscript.tex
Add gisttools intro.

@@
 \subsubsection{Interpreting and visualizing the results}
-As outlined in section \ref{sec:running_GIST}
-
-\todo{write this, show the various GIST properties here: Eww, Esw, S. Show how to calculate A and E with gisttools}
+Running the GIST analysis above, an output data file (gist.dat) containing all calculated quantities for each voxel in the system is produced. 
+Additionally as outlined in section \ref{sec:running_GIST}, various quantities are written to OPenDX files for accessible visualisation.
+
+To faciliate an accessible way to interact with your GIST data, we recommend using the gisttools library for python3. Open the GIST output with gisttools:
+\begin{lstlisting}[style=python]
+	from gisttools.gist import load_gist_file
+	import matplotlib.pyplot as plt
+	gist = load_gist_file(
+	"gist.dat", struct="solute-centered.pdb",  eww_ref=-9.539)
+	print(f'n_frames = {gist.n_frames}, rho0 = {gist.rho0}')
+\end{lstlisting}
+All information from the output file is now accessible through the gist object. 
+If the number of frames of the simulation ($n_frames$) and the reference density (rho0) are not provided, they are automatically detected from the GIST data. 
+We will discuss a way to similarly detect the reference energy of the solvent-solvent energy Eww in the advanced section.
+
+The GIST data can be accessed and modified from the gist object in various ways.
+Gisttools provides access to single or multiple voxels through a .loc property similiar to the pandas library. 
+GIST quantity data for all voxels can be accessed through directly indexing the gist object, either by providing a single name or a list of names.
+Indexing with .loc works by providing a voxel index (either a index number or list of numbers) and optionally a column index (column name or list of names).
+Indices start from 0 and indexing will return a single entry, pandas series or pandas dataframe depending on the query:
+\begin{lstlisting}[style=python]
+	display(gist.loc[0, 'population'])
+	display(gist.loc[0])
+	display(gist.loc[[0,1], ['x', 'y', 'z']])
+	display(gist['dTSsix_norm'].head())
+	display(gist[['Eww_norm', 'Esw_norm']].head())
+\end{lstlisting}
+
+Gisttools provides access to all quantities listed in the GIST output file.
+In addition it calculates the total enthalpy per voxel (Eall) as a sum of Eww and Esw and the free energy of solvation (A) per voxel as a sum of Eall and -dTSsix:
+\begin{lstlisting}[style=python]
+	display(gist[['A_norm', 'Eall_norm']].head())
+\end{lstlisting}
+
+Information about the grid is stored in the grid property of the gist object, structural data is stored as mdtraj trajectory in the struct property:
+\begin{lstlisting}[style=python]
+	print(gist.grid)
+	print(gist.struct)
+\end{lstlisting}
+Gisttools also provides minimal OpenDX writing capabilities to visualize the calculated or modified GIST data.
+For visualization, you'll most likely want to write out the density-normalized GIST quantities ($X_dens$):
+
+\begin{lstlisting}[style=python]
+	gist.save_dx('A_dens', fname)
+	with open(fname) as f:
+		for l in f.readlines(300):
+			print(l, end='')
+\end{lstlisting}
 

### 2022-12-20 manuscript/manuscript.tex
Remove some mentions of gistpp, small changes to code sections

@@
 
-We also show examples using the \software{gistpp} program, which can be obtained from \url{https://github.com/KurtzmanLab/Gist-Post-Processing} (code and documentation).
-
 If the reader prefers to skip the MD and GIST calculation, we also provide the GIST output files with this tutorial, such that the post-processing can be done without any expensive calculations.
@@
 For the visualization of three-dimensional contributions we require molecular visualization software that can handle \software{OpenDX} files, such as \software{PyMol} \cite{pymol} or \software{VMD} \cite{vmd}.
+
+Most of the visualisations and results in this tutorial can also be obtained using the \software{gistpp} program, which is available from \url{https://github.com/KurtzmanLab/Gist-Post-Processing} (code and documentation).
+\software{Gistpp} can directly manipulate \software{OpenDX} files from the GIST output.
+Its usage and some examples were shown in a previous tutorial on GIST\cite{Ramsey2016}.
 
@@
 \end{lstlisting}
-\subsubsection{Interpreting and visualizing the results}
+\subsubsection{Working with the GIST results}
 Running the GIST analysis above, an output data file (gist.dat) containing all calculated quantities for each voxel in the system is produced. 
@@
 	"gist.dat", struct="solute-centered.pdb",  eww_ref=-9.539)
-	print(f'n_frames = {gist.n_frames}, rho0 = {gist.rho0}')
+	print(f'n_frames = {gist.n_frames}')
+	print(f'rho0 = {gist.rho0}')
 \end{lstlisting}
@@
 Gisttools also provides minimal OpenDX writing capabilities to visualize the calculated or modified GIST data.
-For visualization, you'll most likely want to write out the density-normalized GIST quantities ($X_dens$):
+For visualization, you'll most likely want to write out the density-normalized GIST quantities (X\_dens):
 
 \begin{lstlisting}[style=python]
-	gist.save_dx('A_dens', fname)
+	gist.save_dx('A_dens', ')
 	with open(fname) as f:
-		for l in f.readlines(300):
-			print(l, end='')
-\end{lstlisting}
-
+		print([next(f) for x in range(10)])
+\end{lstlisting}
+
+\subsubsection{Visualization the GIST results}
+Visualizing the GIST quantities is possible through the OpenDX (.dx) files, either provided directly from the GIST calculation or output by the gisttools library. 
 \subsection{Applying GIST for the Streptavidin-Biotin complex}
@@
 You can use PyMol\cite{pymol} or VMD\cite{vmd} to visualize the .dx files from GIST\@.
-With \software{gistpp}, you can use the file from the convergence check. Additional OpenDX files can be produced using \inlinecode{gistpp -op makedx}.
+%With \software{gistpp}, you can use the file from the convergence check. Additional OpenDX files can be produced using \inlinecode{gistpp -op makedx}.
 With \software{gisttools}, you can create one using e.g., \inlinecode{gist.save\_dx("A\_dens", "A\_dens.dx")}.

### 2022-12-21 manuscript/manuscript.tex
Add reference energy/density table

@@
 
-Setting the reference density according to the water model used in your simulation is important to get correct results.
+Setting the reference density $\rho_0$ according to the solvent model used in your simulation is important to get correct results.
+You will also want to reference the bulk mean solvent-solvent enthalpy $E_{ww}$ correctly, though we will show ways to do this differently in a later section of the tutorial.
+
 Densities of typical water models can be found in the \software{cpptraj} manual.
-For other models, the density should be computed from a pure water simulation.
+We also provide the densities for water models commonly used with AMBER forcefields in table \ref{tab:ref_densities}.
+These densities were calculated from 1$\mu$s MD simulations.
+Pressure was kept constant at 1 bar with the Berendsen barostat and the water box was prepared with tleap to a size of 100$\cdot$100$\cdot$100\AA before pressure equilibration.
+The temperature was kept constant at 300 K with the Langevin thermostat.
+\begin{table}[h]
+	\caption{Reference density $\rho_0$ and mean water-water energy $E_{ww, norm}$ for various water models from NpT simulations at 300K/1bar.}\label{tab:ref_densities}
+	\small
+	\begin{tabularx}{\columnwidth}{@{}lrr@{}}
+		\toprule
+		Water Model       & $\rho_0$ /\AA$^{-3}$ & $E_{ww, norm}$ /$kcal\cdot mol^{-1} \cdot water^{-1}$  \\
+		\midrule
+		TIP3P    & 0.0329 & -9.533   \\
+		TIP3PFB  & 0.0334 & -11.369  \\
+		TIP4P    & 0.0332 & -9.856   \\
+		TIP4P/Ew & 0.0332 & -11.036  \\
+		TIP5P    & 0.0329 & -9.596   \\
+		SPC/E    & 0.0333 & -11.123  \\
+		OPC      & XXXX   & XXXX     \\
+		OPC3     & XXXX   & XXXX     \\
+		POL3     & XXXX   & XXXX     \\
+		\bottomrule
+	\end{tabularx}
+\end{table}
+If your simulation settings diverge significantly from the above setting or for other solvent models not listed either here or in the \software{cpptraj} manual, the density should be computed from a pure solvent simulation beforehand.
 
@@
 All information from the output file is now accessible through the gist object. 
-If the number of frames of the simulation ($n_frames$) and the reference density (rho0) are not provided, they are automatically detected from the GIST data. 
+If the number of frames of the simulation (n\_frames) and the reference density (rho0) are not provided, they are automatically detected from the GIST data. 
 We will discuss a way to similarly detect the reference energy of the solvent-solvent energy Eww in the advanced section.
@@
 \begin{lstlisting}[style=python]
-	gist.save_dx('A_dens', ')
-	with open(fname) as f:
+	gist.save_dx('A_dens','A.dx')
+	with open('A.dx') as f:
 		print([next(f) for x in range(10)])

### 2022-12-21 manuscript/manuscript.tex
Add formula to double counting section, small changes

@@
 These densities were calculated from 1$\mu$s MD simulations.
-Pressure was kept constant at 1 bar with the Berendsen barostat and the water box was prepared with tleap to a size of 100$\cdot$100$\cdot$100\AA before pressure equilibration.
+Pressure was kept constant at 1 bar with the Berendsen barostat and the water box was prepared with tleap to a size of 100$\cdot$100$\cdot$100\AA\ before pressure equilibration.
 The temperature was kept constant at 300 K with the Langevin thermostat.
+\todo{vah -> Finish these calculations}
 \begin{table}[h]
@@
 \subsubsection{MD simulation}
-\todo{Move simulation part from complex to here}
+
 Now, you can run a restrained MD simulation based on the equilibrated structure of biotin for the subsequent GIST analysis.
@@
 For typical use cases, we recommend using \inlinecode{Eww} and \inlinecode{Esw} over the \inlinecode{PME} column.
-
+\todo{vah: The cpptraj manual says that GIST PME energies cannot be decomposed into Esw and Eww -> what's true? Seems to conflict with what is written in this section.}
 \subsection{GIST with non-water solvents}
@@
 \begin{equation}
-	\Delta E^\text{total}_{ww} = \sum_{k}^\text{voxels} \frac{E_{ww,k}^\text{corr}}{2}
+	\Delta E^\text{R}_{ww} = \sum_{k}^\text{voxels} \frac{E_{ww,k}^\text{corr}}{2}
 %	\Delta E^R_{ww} = n^R E^\text{bulk}_{ww} - E^\textit{R,corr}_{ww}
@@
 \begin{equation}
-	\Delta E^\text{total}_{ww} = \sum_{k}^{\text{voxels in }R} E_{ww,k}^\text{corr}
+	\Delta E^\text{R}_{ww} = \sum_{k}^{\text{voxels in }R} E_{ww,k}^\text{corr}
 %	\Delta E^\textit{R,norm}_\textit{ww} = E^\textit{R,norm}_\textit{ww} - 2E^\textit{bulk}_\textit{ww}
@@
 When the region $R$ comprises more than one molecule, interactions within this region are double-counted while interactions to the outside are not.
-This could be solved using the solvent-solvent energy between each pair of voxels $E_{ww,kl}$.
-While this is supported by the standard GIST implementation, it is rarely done due to the large size of the $E_{ww,kl}$ matrix.
+This could be solved using the solvent-solvent energy between each pair of voxels $E_{ww,k\;l}$:
+\begin{equation}
+	\Delta E^\text{R}_{ww} = 2\left(\sum_{k}^\text{voxels in R} E_{ww,k}^\text{corr}-\sum_{k}^\text{voxels in R}\sum_{l}^\text{l>k} E_{ww, k\;l}\right)
+\end{equation}
+While this is supported by the standard GIST implementation, it is rarely done due to the large size of the $E_{ww,k\;l}$ matrix.
 

### 2022-12-22 manuscript/manuscript.tex
Add Franzs new figure, code for the new figure and integration. Replace GISTPP with python for the binding pocket. Add some TODOs, small fixes and changes all over the place. Restructuring is now complete, some details are not done though.

@@
 
-\subsubsection{Visualization the GIST results}
-Visualizing the GIST quantities is possible through the OpenDX (.dx) files, either provided directly from the GIST calculation or output by the gisttools library. 
+This tutorial discusses the visualisation of OpenDX files in section \ref{sec:visualisation.}
 \subsection{Applying GIST for the Streptavidin-Biotin complex}
@@
 \subsubsection{Contribution of Hydration to Binding}
+\label{sec:binding_contributions}
 In the next step, calculate the $\dgsolv$ contributions around biotin in each system (biotin, streptavidin, and complex).
@@
 You will find that the energy disfavors binding, because we do not yet include the interaction energy between biotin and streptavidin.
-
+To make this easier, gisttool includes a function to integrate in a radius around atom centers:
+\begin{lstlisting}[style=python]
+centers = {biotin: biotin_x, 
+	   strept: strept_x, 
+	   compl: compl_x}
+names = {biotin: 'Biotin', 
+	   strept: 'Streptavidin', 
+	   compl: 'Complex'}
+cols = ['Eall_dens','dTSsix_dens']
+rmax = 12
+results = { names[gist] : 
+	   gist.integrate_around(
+	   		cols, rmax, centers[gist]) 
+	   for gist in [biotin, strept, compl]}
+display( pd.DataFrame( results ))
+
+\end{lstlisting}
+You can also plot the difference between the complex and monomer contributions against the distance to biotin, using the complex coordinates for the holo structure:
+\begin{lstlisting}[style=python]
+fig, (ax1, ax2) = plt.subplots(1, 2, figsize=(8, 4))
+cutoff = 12
+ax1.plot(bins, np.cumsum(biotin_rdf), label='biotin')
+ax1.plot(bins, np.cumsum(strept_rdf), label='streptavidin')
+ax1.plot(bins, np.cumsum(compl_rdf), label='complex')
+ax1.legend()
+ax1.axvline(cutoff, color='k', linestyle='--')
+ax1.set_xlabel('distance to biotin [Å]')
+ax1.set_ylabel('$\Delta G$ contributions [Å]')
+ax1.grid()
+
+difference = compl_rdf - biotin_rdf - strept_rdf
+ax2.plot(bins, np.cumsum(difference))
+ax2.axvline(cutoff, color='k', linestyle='--')
+ax2.set_xlabel('distance to biotin [Å]')
+ax2.set_ylabel('$\Delta \Delta G$ [kcal/mol]')
+ax2.grid()
+plt.show()
+\end{lstlisting}
 You can also plot the difference between the complex and monomer contributions against the distance to biotin, using the complex coordinates for the holo structure.
 The expected result is shown in Figure~\ref{fig-dg-sums}.
-\todo{This figure is missing -vah}
-%\begin{figure}
-%	\centering
-%	\includegraphics[width=1.0\linewidth]{figures/deltaG-difference.png}
-%	\caption{TODO caption}\label{fig-dg-sums}
-%\end{figure}
+\todo{Missing caption}
+\begin{figure}
+	\centering
+	\includegraphics[width=1.0\linewidth]{figures/deltaG-difference.png}
+	\caption{TODO caption}\label{fig-dg-sums}
+\end{figure}
 
@@
 \section{Visualization}
-
+\label{sec:visualization}
 \subsection{Visualizing DX files}
@@
 The expected output is shown in Figure~\ref{fig-streptavidin_gO}.
-
+\todo{VJ -> include the PyMOL settings somewhere, this is not the default output}
 % The current font makes the O look like a tofu / placeholder
 \begin{lstlisting}[style=pymol]
-load streptavidin.pdb
+load output/streptavidin/gist.pdb, streptavidin
 as surface
 color gray70, streptavidin
-load gist-gO.dx, gO
+load output/streptavidin/gist-gO.dx, gO
 isomesh gO_high, gO, 2
@@
 When the ligand is placed in the binding pocked, all atoms are within the SASA.
+\todo{I doubt this is the holo/complex, should be the apo streptavidin, right? -vah´}
 
@@
 For example, voxels that are part of the binding pocket can be defined using a ligand molecule and a distance criteria. 
-In \software{gistpp}, voxels that are within the distance criteria from any ligand heavy atom is considered as part of the region.
+%In \software{gistpp}, voxels that are within the distance criteria from any ligand heavy atom is considered as part of the region.
 The resulting file is a binary file and can be visualized using \software{pymol}.
 \todo{Not GISTPP}
-
-\begin{lstlisting}[style=bash]
-gistpp -op defbp -i gO.dx -i2 ligand.pdb
-
--opt const 3.5 -o gO_3.5.dx
-\end{lstlisting}
-
-Here, the binding pocket includes all voxels within \SI{3.5}{\angstrom} of the heavy atoms of biotin.
+In section~\ref{sec:binding_contributions} we show how to integrate over regions such as the binding pocket.
+To visualize the considered region, we can create a OpenDX file where we set the considered voxels to 1 and all others to 0.
+Here, we show what this would look like for a radius of \SI{3.5}{\angstrom} around the heavy atoms of biotin.
+%\begin{lstlisting}[style=bash]
+%gistpp -op defbp -i gO.dx -i2 ligand.pdb 
+%
+%-opt const 3.5 -o gO_3.5.dx
+%\end{lstlisting}
+\begin{lstlisting}[style=python]
+def select(traj, sel):
+	"Slice a Trajectory by selection mask."
+	return traj.atom_slice(traj.top.select(sel))
+### Load GIST data
+compl = gt.gist.load_gist_file(
+	'complex/gist/gistoutlong.dat', 
+	struct='complex/gist/gist_nowat.pdb')
+### Define ligand atoms and find voxels around them
+rmax = 3.5
+biotin_mask = 'resname BTN and not element H'
+btn = select(compl.struct, biotin_mask).xyz[0] * 10.
+ind, _, _ = compl.distance_to_spheres(btn, rmax)
+
+### Set voxels within 3.5A to 1, write out dx file
+compl.loc[ind, 'BP'] = 1
+compl.save_dx('BP', 'binding_pocket.dx')
+\end{lstlisting}
 

### 2022-12-22 manuscript/manuscript.tex
Rewrite some sentences, more siunitx, fix angstrom sign, fix indentation

@@
 \DeclareSIUnit\Molar{M}
+\DeclareSIUnit\calorie{cal}
+\DeclareSIUnit\kcalPerMolASqr{\kilo\calorie\per\mole\per\angstrom\squared}
+%\DeclareSIUnit{\Molar}{M}
+\DeclareSIUnit\angstrom{\text{\AA}}
+%\DeclareSIPrefix\micro{\text{\textmu}}{0}
+\sisetup{
+	%math-rm=\mathsf,
+	text-rm=\sffamily
+}
+
 \usepackage[italic,LGRgreek]{mathastext}
@@
 
-\DeclareSIUnit{\calorie}{cal}
-\DeclareSIUnit{\kcalPerMolASqr}{\kilo\calorie\per\mole\per\angstrom\squared}
-\DeclareSIUnit{\Molar}{M}
-
 % TODO: delete for final build
@@
 
-Setting the reference density $\rho_0$ according to the solvent model used in your simulation is important to get correct results.
-You will also want to reference the bulk mean solvent-solvent enthalpy $E_{ww}$ correctly, though we will show ways to do this differently in a later section of the tutorial.
+It is important to set the reference density $\rho_0$ according to the solvent model used in your simulation.
+The solvent-solvent enthalpy $E_{ww}$ also needs to be referenced, but this is usually done in the post-processing stage.
 
@@
 We also provide the densities for water models commonly used with AMBER forcefields in table \ref{tab:ref_densities}.
-These densities were calculated from 1$\mu$s MD simulations.
-Pressure was kept constant at 1 bar with the Berendsen barostat and the water box was prepared with tleap to a size of 100$\cdot$100$\cdot$100\AA\ before pressure equilibration.
-The temperature was kept constant at 300 K with the Langevin thermostat.
+These densities were calculated from \SI{1}{\micro\second}\pdfcomment{weird small mu symbol} MD simulations at \SI{1}{\bar} and \SI{300}{\kelvin} using the Berendsen barostat and the Langevin thermostat.
+The water boxes were prepared with \software{tleap} to a size of \SI{100 x 100 x 100}{\angstrom} before pressure equilibration.
 \todo{vah -> Finish these calculations}
@@
 
-The output is organized by voxel and sorted according to their x,y, and z coordinates, describing the center point of the voxel.
-Energy and entropy quantities are reported in two different ways ´
+The output is organized by voxel and sorted according to their x, y, and z coordinates, describing the center point of the voxel.
+Energy and entropy quantities are reported in two different ways.
 The \inlinecode{\_dens} quantities are normalized by the voxel volume and can be used to compute integrals.
 The \inlinecode{\_norm} quantities are instead normalized by the number of solvents in the respective voxel averaged over all considered frames, and are useful to visualize local water properties.
-Therefore, to convert from a \inlinecode{\_norm} quantity $X^k_{norm}$ in voxel k to a \inlinecode{\_dens} quantity $X^k_{dens}$ the following equation is applied:
+To convert from a \inlinecode{\_norm} quantity $X^k_{norm}$ in voxel $k$ to a \inlinecode{\_dens} quantity $X^k_{dens}$, the following equation can be applied:
 \begin{equation}
@@
 with $V^k$, the volume of voxel $k$, $N^k_{solvent}$, the solvent population in voxel $k$, and $N_{frames}$, the number of simulation frames considered in the GIST analysis.
-Acoordingly, converting $X^k_{dens}$ to  $X^k_{norm}$ is possible using the following equation:
-\begin{equation}
-	X^k_{norm} = X^k_{dens} \cdot V^k \cdot \frac{N_{frames}}{N_{frames}N^k_{solvent}} 
-\end{equation}
+
 \section{Tutorial}
@@
 
-This file will run a 100 ns NpT MD simulation while keeping all atoms except for hydrogens and waters restrained. Note that these are very standard MD settings except for the restraints.
+This file will run a \SI{100}{\nano\second} NpT MD simulation while keeping all atoms except for hydrogens and waters restrained.
+Note that these are very standard MD settings except for the restraints.
 
@@
 Running the GIST analysis above, an output data file (gist.dat) containing all calculated quantities for each voxel in the system is produced. 
-Additionally as outlined in section \ref{sec:running_GIST}, various quantities are written to OPenDX files for accessible visualisation.
-
-To faciliate an accessible way to interact with your GIST data, we recommend using the gisttools library for python3. Open the GIST output with gisttools:
+Additionally as outlined in section \ref{sec:running_GIST}, various quantities are written to OpenDX files.
+
+To interact with your GIST data, we recommend the gisttools library for python3. Open the GIST output with gisttools:
 \begin{lstlisting}[style=python]
@@
 \begin{lstlisting}[style=python]
-	display(gist.loc[0, 'population'])
-	display(gist.loc[0])
-	display(gist.loc[[0,1], ['x', 'y', 'z']])
-	display(gist['dTSsix_norm'].head())
-	display(gist[['Eww_norm', 'Esw_norm']].head())
+gist.loc[0, 'population'] # single value
+gist.loc[0] # Series
+gist.loc[[0,1], ['x', 'y', 'z']] # DataFrame
+gist['dTSsix_norm'].head() # Series
+gist[['Eww_norm', 'Esw_norm']].head() # DataFrame
 \end{lstlisting}
@@
 Gisttools provides access to all quantities listed in the GIST output file.
-In addition it calculates the total enthalpy per voxel (Eall) as a sum of Eww and Esw and the free energy of solvation (A) per voxel as a sum of Eall and -dTSsix:
-\begin{lstlisting}[style=python]
-	display(gist[['A_norm', 'Eall_norm']].head())
-\end{lstlisting}
-
-Information about the grid is stored in the grid property of the gist object, structural data is stored as mdtraj trajectory in the struct property:
-\begin{lstlisting}[style=python]
-	print(gist.grid)
-	print(gist.struct)
-\end{lstlisting}
+Additionally, it calculates the total enthalpy per voxel (Eall) as a sum of Eww and Esw, and the free energy of solvation (A) by subtracting dTSsix from Eall.
+
+Information about the grid is stored in the grid property of the gist object, structural data is stored as an \software{mdtraj} trajectory in the struct property:
+
 Gisttools also provides minimal OpenDX writing capabilities to visualize the calculated or modified GIST data.
@@
 gist.save_dx('A_dens','A.dx')
-	with open('A.dx') as f:
-		print([next(f) for x in range(10)])
-\end{lstlisting}
-
-This tutorial discusses the visualisation of OpenDX files in section \ref{sec:visualisation.}
+\end{lstlisting}
+
+This tutorial discusses the visualization of OpenDX files in Section~\ref{sec:visualization}.
 \subsection{Applying GIST for the Streptavidin-Biotin complex}
@@
 In our example, we only use a single equilibrated structure.
-A more sophisticated approach would identify one or several dominant structures out of the ensemble - either gathered from experimental data or by clustering an MD simulation.
+A more sophisticated approach would be to identify several dominant structures in the ensemble -- either from experimental ensembles or by clustering an MD simulation.
 
@@
 A good criterion for validating the box size is to check whether the water on the outside of the box behaves bulk-like according to the GIST output.
-Note that due to the solute restraints, some important solvent sites might be unavailable to exchange with the bulk solvent.
-Make sure to explicitly place water at such sites, if they are important to the process of interest e.g. waters in confined binding pockets, as they might be missed by automated solvation procedures.
-It is good practice to check your restrained MD simulation or solvent populations obtained by GIST afterwards to confirm wether all important solvent sites were properly solvated.
+Note that, due to the restraints, some solvent sites might be inaccessible to the bulk solvent.
+To investigate such positions, make sure to explicitly place water molecules there.
+It is good practice to check your restrained MD simulation or solvent populations obtained by GIST to confirm whether all relevant sites were properly solvated.
 

### 2023-01-16 manuscript/manuscript.tex
Fix bibliography with betterbib.
Updates from Overleaf

--- 2022-12-21 manuscript/manuscript.tex
@@
 \usepackage{listings}
-\usepackage{tabularx}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\title{Quantifying Spatially Resolved Hydration Thermodynamics Using Grid Inhomogeneous Solvation Theory [Article v\versionnumber]}
-
-\author[1\authfn{1}]{Franz Waibl}
-\author[1\authfn{1}]{Valentin J Hoerschinger}
-\author[2]{Vjay Molino}
-\author[1]{Monica L Fern{\'a}ndez-Quintero}
-%\author[1,2]{Firstname Middlename Familyname}
-%\author[2]{Firstname Initials Surname}
-\author[4]{Daniel R Roe}
-\author[1*]{Klaus R Liedl}
-\author[2*]{Michael K Gilson}
-\author[2*]{Tom Kurtzman}
-\affil[1]{Department of General, Inorganic and Theoretical Chemistry, University of Innsbruck, Austria}
+\title{This is the title test [Article v\versionnumber]}
+
+\author[1*]{Firstname Middlename Surname}
+\author[1,2\authfn{1}\authfn{3}]{Firstname Middlename Familyname}
+\author[2\authfn{1}\authfn{4}]{Firstname Initials Surname}
+\author[2*]{Firstname Surname}
+\affil[1]{Institution 1}
 \affil[2]{Institution 2}
@@
 \DeclareSIUnit{\kcalPerMolASqr}{\kilo\calorie\per\mole\per\angstrom\squared}
-\DeclareSIUnit{\Molar}{M}
 
@@
 \todo{IMO, our article looks too dry. We should have something nice (e.g., an illustration of the GIST method) on the 1st or 2nd page -fwa.}
-\todo{Potentially relevant: 10.1021/acs.jcim.7b00520 }
 
 Solvation thermodynamics govern any process involving solutes in a solvent.
-%Cite "Theory of Simple Liquids"?
 %Solutes are constantly interacting with solvents in their environment, necessitating a good description of these interactions for accurate computational predictions. 
-The free energy of solvation is defined as the reversible work needed to insert a solute into the solvent \cite{ben-naim-book}.
-Especially the interaction with water is of utmost importance, as many biological processes occur in aqueous environment \cite{Privalov2017-water-review}.
-It drives real-world effects such as hydrophobic interactions, binding of ligands to a biomolecule, as well as dynamics and folding of proteins.
+Especially the interaction with water is of utmost importance, as many biological processes occur in a aqueous environment. 
+The free energy of solvation thereby strongly influences real-world problems such as hydrophobic effects, the binding of ligands to a biomolecule, as well as the dynamics and folding of proteins.
 
@@
 
-This tutorial provides an example of a GIST workflow that can be easily adapted to different systems. 
-In addition to a short introduction of the method and the theory behind it, we present an overview of the different implementations of GIST.
-We ease the user into basic GIST analysis at the example of biotin.
-Building on the basics, we present a more advanced GIST study of the streptavidin-biotin complex, reflecting on many common aspects of GIST studies.
+This tutorial provides an example of a GIST workflow that can be easily adapted towards different systems. 
+In addition to a short introduction of the method and the theory behind it, we present an overview of the various different flavors of GIST methods.
+
+Furthermore, a GIST study of the streptavidin-biotin complex is presented, which reflects on many common aspects of GIST studies.
+Using this test system, we present solvation analysis of a small molecule (biotin) as well as a biomolecule (streptavidin).
 We show how to interpret the three-dimensional contributions of solvation free energy in a binding pocket and around the ligand.
-Additionally, we compute the binding contribution of the free energy of solvation, which requires accurate post-processing of the GIST outputs to avoid unfavorable summation of bias.
-We believe that this example study covers a wide range of applications.
+
+Additionally, we compute the binding contribution of the free energy of solvation, which requires accurate post-processing of the GIST outputs to avoid unfavorable summation of bias in the GIST calculation.
+We believe that this example study covers a wide range of applications of the GIST method.
 
 Furthermore, we discuss several technical aspects  and typical pitfalls of GIST analyses, such as the normalization of voxel values and how to deal with double counting of energy contributions. 
-Additionally, we provide a Python library (\software{gisttools}) to unify the analysis of GIST data produced by various versions of GIST and make the post-processing of GIST results more accessible.
+Additionally, we provide a python library to unify the analysis of GIST data produced by various versions of GIST and make the post-processing of GIST results more accessible.
 
@@
 This tutorial is aimed at users with a solid knowledge of molecular dynamics (MD) simulations.
-The user should be able to run MD simulations and analyze the resulting trajectories.
+The user should be able to run MD simulations and work with their output data.
 Previous experience of GIST is not necessary.
-However, we only provide a short summary of the theory behind GIST in section~\ref{sec:theory} at the end of this tutorial.
-For more detailed information, we point the reader towards the primary sources.
+However, we only provide a short explanation of the theory behind GIST and point the reader towards the primary sources for more detailed information.
 
 The presented analyses are run in Python.
-While no programming experience is necessary to follow the tutorial, a basic understanding of Python is advantageous to adjust the code towards different requirements and use-cases.
+While no programming experience should be necessary to follow the tutorial, a basic understanding of Python is advantageous to adjust the code towards different requirements and use-cases. 
+We provide a Jupyter notebook which contains the presented analyses. 
+Therefore, an understanding of Jupyter notebooks is beneficial to reproduce the analyses.
 
@@
 Tutorials requiring specific software packages must provide instructions and files for the referenced version of the software.
-The main implementation of GIST is part of the MD analysis software \software{cpptraj} released with AmberTools \cite{amber22}.
+The main implementation of GIST is part of the MD analysis software \software{cpptraj} as part of AmberTools \cite{amber22}.
 
 A recent version of \software{cpptraj} should be used, such as the one provided in AmberTools22 or a recent installation from GitHub (\url{https://github.com/Amber-MD/cpptraj}).
-
-The closed-source Amber simulation engine is used in this tutorial to run the MD simulations.
-If Amber is not available, other MD engines can be used as long as topologies and trajectories can be produced in a format supported by \software{cpptraj}.
-If using GROMACS \cite{Abraham2015-gromacs,Pronk2013-gromacs}, we propose to prepare the topology with AmberTools and convert it using \software{acpype.py} \cite{Sousa_da_Silva2012-acpype}.
-%We should actually test that!
-Also note that compressed trajectory formats such as \inlinecode{xtc} might bias the entropy calculation due to the loss of precision for the atom positions. \todo{This might be in the technical part}
+The closed-source Amber simulation engine is used in this tutorial to run the necessary MD simulations.
+
+The closed-source Amber simulation engine is used in this tutorial to run the necessary MD simulations.
+If no Amber installation is available, the MD can be performed other simulation engines compatible with \software{cpptraj}.
+If using GROMACS, we propose to prepare the structures and topologies with AmberTools and converting them using \software{acpype.py}.
+
+\todo {If the simulation was ran using GROMACS, mdtraj can be used to convert GROMACS topology and trajectory in to an Amber file format.}
+\todo {vah: I would remove this mention of GROMACs and replace it with a more general comment. Or we could test if this actually works ( I doubt it does for ff19SB/OPC).}
+In that way, a full Amber topology is available for GIST.
+
+Also note that compressed trajectory formats such as \inlinecode{xtc} might bias the entropy calculation due to the loss of precision for the atom positions.
 
@@
 	\item \software{gisttools} (\url{https://github.com/liedllab/gisttools})
-	\item \software{mdtraj (\url{https://www.mdtraj.org/}) \cite{McGibbon2015-mdtraj}}
+	\item \software{mdtraj (\url{https://www.mdtraj.org/})}
 \end{itemize}
+
+We also show examples using the \software{gistpp} program, which can be obtained from \url{https://github.com/KurtzmanLab/Gist-Post-Processing} (code and documentation).
 
@@
 We additionally provide a Jupyter notebook \cite{Kluyver2016-jupyter,Granger2021-jupyter} with the presented analyses. 
-%We recommend JupyterLab to work with the notebook. Jupyter is available at \url{https://jupyter.org/}.
-
-For the visualization of three-dimensional contributions we require molecular visualization software that can handle \software{OpenDX} files, such as \software{PyMol} \cite{pymol} or \software{VMD} \cite{vmd}.
-
-Most of the visualisations and results in this tutorial can also be obtained using the \software{gistpp} program, which is available from \url{https://github.com/KurtzmanLab/Gist-Post-Processing} (code and documentation).
-\software{Gistpp} can directly manipulate \software{OpenDX} files from the GIST output.
-Its usage and some examples were shown in a previous tutorial on GIST\cite{Ramsey2016}.
+We recommend JupyterLab to work with the notebook. Jupyter is available at \url{https://jupyter.org/}.
+
+For the visualization of three-dimensional contributions we require molecular visualization software that can handle \software{OpenDX} files.
+We recommend using a current version of \software{PyMol}\cite{pymol} or \software{VMD}\cite{vmd}.
+\section{Theory}
+
+GIST is an implementation of inhomogeneous fluid solvation theory (IST) \cite{Lazaridis1998} that discretizes the free energy of solvation $\dgsolv$ on a three-dimensional grid. 
+It was first devised by Nguyen et al.
+\cite{Nguyen2012} and its implementation in \software{cpptraj} was thoroughly described by Ramsey et al.
+\cite{Ramsey2016}.
+% to overcome the limitation of IST which was commonly used for regions of high water occupancy and its implementation in \software{cpptraj} was thoroughly described in \cite{Ramsey2016}.
+Here, we only provide a short overview of the theory behind GIST.
+For more detailed information, we recommend one of the more recent publications on developments in GIST \cite{Kraml2020,Chen2021}.
+
+\subsection{Solvation Entropy}
+
+%The solvation entropy of a system is totality of the contribution of the  solute-water entropy and the water-water entropy. -- this is inconsistent with IST theory (fwa)
+
+%\begin{equation}
+%	\Delta S_\textit{solv} = \Delta S_\textit{sw} + \Delta S_\textit{ww}
+%\end{equation}
+%Here the solvation entropy is approximated from the contribution of solute-water interaction. 
+%\textbf{STILLNEEDS REVISON + ADD CITATIONS} The quantity $k_\textit{b}$represents the Boltzman constant, 
+%$\rho^\textit{o}$ is the bulk number density, $g_\textit{sw}\left(\textbf{r},\omega \right)$ is the solute-water 
+%pair correlation function in the solute's frame of reference, \textbf{r} is the coordinates of the oxygen atom of the water molecule,  
+%$\omega$ represents the Euler angles, and $\frac{1}{8\pi^2}$ is the volume of the orientational space, which acts as a normalization constant.
+%In bulk and in regions where the orientational 
+%entropy is uniform, the quantity $g_\textit{sw}\left(\textbf{r},\omega \right)$ is unity which leads to the first-order solvation entropy of 
+%bulk to be zero. Also, this quantity reaches unity as the distance from the solute molecule increases which leads for the calculation of 
+%the solvation entropy to be an approximation of the integral around the solute atom.
+
+IST expresses the free energy of solvation in terms of the solvent density in a coordinate system defined by the solute.
+The solute is kept fixed in space, following the standard solvation process established by Ben-Naim \cite{ben-naim-book}.
+
+For the entropy, this leads to an infinite expansion of correlations of increasing order.
+In a slightly simplified view, the first order is the solvent density $g_\textit{sw}\left(\textbf{r},\omega \right)$ at each position $\textbf{r}$ and orientation $\omega$.
+The first-order entropy omits all solvent-solvent (and higher) correlations, while the solute-solvent correlation is taken into account because the coordinates are relative to the solute.
+Thus, it is called the solute-water entropy $S_{sw}$.
+%The second order is the pair density, i.e., the probability of finding two solvent molecules in positions $\textbf{r}$ and $\textbf{r}^\prime$.
+
+The solvent density is expressed relative to the bulk density $\rho^\textit{o}$. %$g_\textit{sw}\left(\textbf{r},\omega \right)$ is the solute-water 
+%pair correlation function in the solute's frame of reference, \textbf{r} is the coordinates of the oxygen atom of the water molecule,  
+%$\omega$ represents the Euler angles, and 
+The entropy integral is also normalized by $8\pi^2$, which is the volume of the orientational space.
+In bulk or at a high distance to the solute, the quantity $g_\textit{sw}\left(\textbf{r},\omega \right)$ is unity which leads to the first-order entropy of 
+bulk to be zero. 
+Therefore, no reference entropy needs to be computed unless there is a numeric bias, e.g., when the MD frames are not statistically independent due to insufficient spacing.
+%Also, this quantity reaches unity as the distance from the solute molecule increases which leads for the calculation of 
+%the solvation entropy to be an approximation of the integral around the solute atom. 
+\begin{equation}
+	\Delta S_\textit{solv}
+	\approx \Delta S_\textit{sw}
+	\newcommand{\gro}{g_{sw}(\textbf{r},\omega)}
+	\equiv -R \frac{\rho^\textit{0}}{8\pi^\textit{2}} \int \gro \ln\gro d\textbf{r}d\omega
+\end{equation}
+\todo{We are inconsistent regarding R/kB; vah: Fixed. is it - tho? Papers seem to disagre }
+
+The solvation entropy is further broken down in terms of the contribution from translational and orientational terms.
+\begin{equation}
+	\Delta S_\textit{solv} = \Delta S_\textit{trans} + \Delta S_\textit{orient}
+\end{equation}
+
+\subsection{Entropy Calculations in Cpptraj}
+In \software{cpptraj} the calculation of  solvation entropy is handled by two methods.
+
+The first method assumes that the distribution of the orientation $\omega$ is constant within a voxel $k$.
+This allows for the splitting of the contributions of the translational and the orientational entropy. 
+A first nearest neighbor (NN) approach is used to evaluate the density.
+%The translational density $g_{NN}(\textbf{r})$ is expressed in reference to a homogeneous bulk solvent at the reference density, while the orientational density $g_{NN}(\omega)$ is expressed to a random placement of the number of solvents found in this voxel.
+Both the translational ($g_{NN}(\textbf{r})$) and orientational ($g_{NN}(\omega)$) density are computed relative to a bulk solvent at the reference density $\rho^0$.
+
+The average entropy contributions per solvent molecule in voxel $k$ are calculated as
+
+\begin{equation}
+	S_{k}^\textit{trans} = -R \left( \gamma + \frac{1}{N_\textit{k}} \sum _{i=1}^{N_k} \ln g_{NN, \textit{i}}(\textbf{r}) \right)
+\end{equation}
+and
+\begin{equation}
+	S_{k}^\textit{orient} = -R \left( \gamma + \frac{1}{N_k} \sum _{i=1}^{N_k} \ln g_{NN, i}(\omega) \right)
+	,
+\end{equation}
+where $N_\textit{k}$ is the number of water molecules found in voxel k, $\gamma$ is Euler's constant accounting for the bias of the naive entropy estimator, and $g_\textit{NN}$ is the nearest neighbor estimate.
+
+The second method directly calculates the solvation entropy by evaluating the six-dimensional integral (3 for position and 3 for orientation) 
+using the nearest-neighbor approach. 
+% Unlike the first method, the nearest neighbor estimate for the sixth-order entropy considers all the water found within the same and neighboring voxels.
+-- misleading because this is also done for S_trans (fwa)
+
+\begin{equation}
+g_\textit{vox} \left( \textbf{r}, \omega \right) \approx g_\textit{vox}(\textbf{r}) g_\textit{vox}(\omega)
+\end{equation}
+
+\begin{equation}
+	S_\textit{k} = -R \left( \gamma + \frac{1}{N_\textit{k}} \sum _{i=1}^{N_k} \ln g_{NN, \textit{i}}(\textbf{r},\omega) \right)
+\end{equation}
+
+\subsection{Solvation Energy}
+The solvation energy is readily calculated by summing up the contributions of the water-water interaction and the water-solute interaction from the force field.
+\begin{equation}
+	\Delta E_\textit{solv} = \Delta E_\textit{sw} + \Delta E_\textit{ww}
+\end{equation}
+
+The solute-solvent energy $E_{sw}$ can be expressed in terms of the solvent density and the potential $U_{sw}(\textbf{r},\omega)$ induced by the solute.
+In practice, it is evaluated as the expectation value $\langle\cdot\rangle$ of the pairwise force field energy $U_{ij}$ between all solvent molecules in voxel $k$ and all solute atoms.
+
+\begin{equation}
+\begin{aligned}
+\label{eq-esw}
+	\Delta E_{sw}(\textbf{r}) \equiv& \frac{1}{8\pi^2} \int g_\textit{sw}\left(\omega|\textbf{r}\right) U_\textit{sw}\left(\textbf{r}, \omega\right) d\omega \\
+	\Delta E_{sw,k}=& \left\langle \sum_i^{\textit{solvent},k} \; \sum_j^\textit{solute} U_{ij}\right\rangle
+\end{aligned}
+\end{equation}
+
+To localize $E_{sw}$ on the three-dimensional grid, every energy term is assigned to the voxel $k$ holding the solvent, and the expectation value in Equation~\ref{eq-esw} is computed per voxel.
+Similar to the entropy integrals, the solute-water solvation energy integral also decays with increasing distance from the solute.
+Hence, the solvation energy can be approximated by local spatial integrals.
+
+The solvent-solvent energy $E_{ww}$ is computed in a similar manner.
+Again, it can be expressed in terms of density functions, but is practically computed as a sum over solvent-solvent interactions where one solvent is in voxel $k$.
+In contrast to $E_{sw}$, $E_{ww}$ does not tend to zero in bulk.
+Therefore, a reference energy corresponding to the average energy of a water molecule in bulk needs to be subtracted.
+
+%However, there is a complication arising from this approximation.
+Water molecules outside the region of interest will also be accounted for in the pairwise water term.
+For example, given two different regions $R_a$ and $R_b$, will not sum up to the total energy of the two regions due to double counting.
+
+\begin{equation}
+\begin{aligned}
+	\Delta E_{ww}(\textbf{r}) \equiv & \left(\frac{1}{8\pi^2} \right)^2 \rho^o \int g_\textit{sw}(\omega|\textbf{r}) \\
+	& \times \left[g_\textit{sw}(\textbf{r}^\prime , \omega^\prime) - g^o_\textit{ww} (\textbf{r}, \omega, \textbf{r}^\prime, \omega^\prime )\right] \\
+	& \times U_\textit{ww}(\textbf{r}, \omega, \textbf{r}^\prime, \omega^\prime)d\omega d\textbf{r}^\prime d\omega^\prime \\
+	\Delta E_{ww,k} =& \frac{1}{N_k} \left\langle \sum_i^{\text{solvent},k} \; \sum_{j \neq i}^\text{solvent} U_{ij}\right\rangle - \left\langle E_{ww}^\text{bulk}\right\rangle
+\end{aligned}
+\end{equation}
+
+%AFAIK, this was never done, since the pair interaction matrix Eij would have to be written out.
+%There are two ways in which this double counting was accounted for in the original paper.
+First, using the water-displacement method.
+This method is applicable when a determining the energetic cost of kicking out a water molecule.
+%\begin{equation}
+%	\Delta E^R_\textit{ww} = n^\textit{R} E^\textit{bulk}_\textit{ww} - E^\textit{R,corr}_\textit{ww}
+%\end{equation}
+
+%The second method uses normalized-water properties to account the difference in water properties of two different regions.
+The normalized water property is defined as the water property per water molecule.
+%\begin{equation}
+%	\Delta E^\textit{R,norm}_\textit{ww} = E^\textit{R,norm}_\textit{ww} - 2E^\textit{bulk}_\textit{ww}
+%\end{equation}
+
+When computing the total $E_{ww}$ of the system, double-counting of interactions must be avoided.
+The total solvent-solvent energy of the system is as follows:
+
+\begin{equation}
+	\Delta E^R_\textit{ww} = n^\textit{R} E^\textit{bulk}_\textit{ww} - E^\textit{R,corr}_\textit{ww}
+\end{equation}
+
+However, when water is replaced from a small region $R$ of interest, such as a single water molecule, almost all interactions are with water molecules outside of this region.
+Therefore, there is no double counting, and the full solvent-solvent energy should be used.
+
+\begin{equation}
+	\Delta E^\textit{R,norm}_\textit{ww} = E^\textit{R,norm}_\textit{ww} - 2E^\textit{bulk}_\textit{ww}
+\end{equation}
+
+When the region of interest $R$ comprises more than one molecule, interactions within this region are double-counted while interactions to the outside are not.
+To take this into account, the solvent-solvent energy between each pair of voxels $E_{ww,kl}$ is needed.
+While this is supported by the standard GIST implementation, it is rarely done due to the large size of the $E_{ww,kl}$ matrix.
+
+We note that replacement of a small or medium solvent region would lead to reorganization of the surrounding solvent.
+However, this depends on the local environment and on \emph{what} the molecule would be replaced by.
+Therefore, the effect of reorganization can not be directly included in the GIST method, but must be judged for each case individually.
+This is not a problem when computing the total free energy of solvation $\dgsolv$, since no additional boundary is created.
+\todo{This is a bit unclear to me, could we elaborate on this a bit more? What are we asking of the reader here, how would one consider this effect? -vah}
+
+\subsection{PME implementation}
+
+In the original version of GIST, the energies are calculated based on the the minimum image convention.
+In PME-GIST, the electrostatics are calculated using the particle mesh Ewald (PME) method, which yields energies that are highly consistent with the treatment in the Amber MD engine \cite{Chen2021}.
+The electrostatic and Lennard-Jones parts $E_\text{elec}$ and $E_\text{lj}$ are computed separately.
+
+\begin{equation}
+	E_\text{total} = E_\text{elec} + E_\text{lj}
+\end{equation}
+
+In PME-GIST, the system is treated as periodic, and the electrostatic energy is split into a short-range term $E_\text{dir}$, which is calculated in direct space using a distance cutoff, and a long-range term $E_\text{rec}$, which is calculated in reciprocal space.
+Additionally, there is a correction term $E_\text{corr}$, which corrects for the self-interaction of each solvent molecule in the reciprocal term.
+
+\begin{equation}
+	E_\textit{elec} = E_\textit{dir} + E_\textit{rec} + E_\textit{corr}
+\end{equation}
+%\todo{How is E$_{dir}$ treated? Is there a cutoff as well (PME GIST paper seems to indicate so)? If so, we should probably note that. } - ok? -fwa
+The Lennard-Jones contribution is computed in the direct space, using a distance cutoff.
+The short-range contribution is the LJ energy within the set cutoff and the long-range correction term accounts for the contributions above this cutoff.
+
+\begin{equation}
+	E_\textit{lj} = E_\textit{lj,\ short} +  E_\textit{corr}
+\end{equation}
+\todo{Is $E_{corr}$ the same in eq 13 and 14? How is it calculated? -vah}
 
 \section{Running GIST}
-\label{sec:running_GIST}
 %\subsection{Running the GIST analysis}
-To run a GIST analyis, the following is required:
+To run a gist analyis, the following is required:
 \begin{itemize}
-	\item \software{cpptraj}.
+	\item A running installation of \software{cpptraj}.
 	\item A topology of the molecule of interest in a solvent box.
@@
 \paragraph{Identify and define the region of interest}
-In order to run a GIST analysis, one must define the region where solvent properties will be calculated.
-This region is represented by a rectangular grid with given position, size, and voxel spacing.
-The default value of \SI{0.5}{\angstrom} provides reasonable discretization for many use-cases.
-While integrals of GIST quantities over the grid are independent of the voxel size, individual values converge slower at high resolutions.
-
-The grid is defined by parameters of the \inlinecode{gist} command in \software{cpptraj}.
-The grid center is defined by \inlinecode{gridcntr}, the voxel number in x, y, and z direction by \inlinecode{griddim}, and the spacing (voxel side-length) in \AA by \inlinecode{gridspacn}.
+In order to run a GIST analysis, the region where the solvent properties will be calculated must be first defined.
+This region is represented by a rectangular prism centered at (x, y, z) with $N_x \times N_y \times N_z$ voxels, where each voxel is cubic with a sidelength of \inlinecode{gridspacn}.
+While integrals over GIST quantities are independent of the voxel size, individual voxel values converge slower at high resolutions.\pdfcomment{The calculation actually can actually get faster at higher resolution, because the same grid is used in the nearest-neighbor search.}
+A default value of 0.5 A is used that provides reasonable discretization for many use-cases.
+
+In \software{ccptraj}, the rectangular prism is defined by parameters of the \inlinecode{gist} command.
+The center of the grid is defined by \inlinecode{gridcntr}, the dimensions of 
+the grid is defined by \inlinecode{griddim} (in voxel counts), and the spacing 
+is defined by \inlinecode{gridspacn} (in Å).
 
@@
 
-Often, the region of interest will be defined as the surrounding of a molecule, which might for example be a ligand in the binding pocket or the whole protein.
-A helper script called \inlinecode{FindCentroid.py} for this purpose is supplied with this tutorial, and can be used as follows:
+The region can be defined using a ligand in a specific binding pocket of the 
+protein, a specific residue of the protein, or the whole protein itself. A 
+helper script FindCentroid.py can be used to determine the centroid of the 
+region of interest as well as the dimensions of the rectangular prism that will 
+fully enclose the region. A pdb file is needed as an input and the rectangular 
+prism will be centered based on the centroid of the given structure.
 
@@
 \paragraph{Run GIST using cpptraj}
-GIST is implemented in \software{cpptraj} which requires the topology and the trajectory of the MD simulation to be analyzed.
+GIST is incorporated in cpptraj which requires the topology and the trajectory of the MD simulation to be analayzed.
 The commands needed to run GIST in cpptraj can be written in an input file (e.g gist.in) for convenience.
-The basic commands, assuming a given grid center and a grid size of $30\times30\times30$ voxels with \SI{0.5}{\angstrom} spacing, are:
+The basic comands, assuming a given grid center and a grid size of $30\times30\times30$ voxels with \SI{0.5}{\angstrom} spacing, are:
 
@@
 
-The first line specifies the Amber topology file that was used to run MD. The second line species the trajectory file.
-Note that the solute should not move significantly for a GIST calculation, since IST does not take the solute degrees of freedom into account.
-This is typically achieved using restraints on the solute heavy atoms.
-The third line tells \software{cpptraj} to run a GIST calculation using the specified region and print the output in a file called \inlinecode{gist.dat}.
-The last line instructs \software{cpptraj} to run the calculation.
-
-The syntax for running \software{cpptraj} given this input file is:
+The first line specifies the amber topology file that was used to run MD.
+The second line species the trajectory file and what frames to be analyzed.
+The third line tells cpptraj to run a GIST calculation using the region defined and print the output of the calculation in a file called gist.dat.
+The last instructs cpptraj to run the calculation.
+The syntax for running cpptraj given this input file is:
+
 \begin{lstlisting}[style=bash]
@@
 
-The \inlinecode{gist} action has a large number of options to modify the GIST calculation or its output.
-A full list is provided at the end of this tutorial or through the \software{cpptraj} manual.
-The most important options are summarized in the following table:
-
-\begin{table}[h]
-	\caption{The main options for GIST in \software{cpptraj}.}\label{tab:GIST_options}
-	\small
-	\begin{tabularx}{\columnwidth}{@{}llXl@{}}
-		\toprule
-		Command       & Options & Explanation & Default  \\
-		\midrule
-		gridcntr    & <y> <y> <z> & Coordinates of the grid center in \AA & 0 0 0  \\
-		griddim     & <nx> <ny> <nz> & Number of voxels per axis & 40 40 40\pdfcomment{Overfull hbox?}\\
-		gridspacn & <dval> &Grid spacing / voxel dimensions in  \AA & 0.5    \\
-		refdens         & <rdval> & Reference density of the bulk water in molecules/\AA$^3$ . & 0.0334    \\
-		temp & <tval> & System temperature in Kelvin & 300 \\
-		\bottomrule
-	\end{tabularx}
-\end{table}
-
-Setting the reference density $\rho_0$ according to the solvent model used in your simulation is important to get correct results.
-You will also want to reference the bulk mean solvent-solvent enthalpy $E_{ww}$ correctly, though we will show ways to do this differently in a later section of the tutorial.
-
-Densities of typical water models can be found in the \software{cpptraj} manual.
-We also provide the densities for water models commonly used with AMBER forcefields in table \ref{tab:ref_densities}.
-These densities were calculated from 1$\mu$s MD simulations.
-Pressure was kept constant at 1 bar with the Berendsen barostat and the water box was prepared with tleap to a size of 100$\cdot$100$\cdot$100\AA\ before pressure equilibration.
-The temperature was kept constant at 300 K with the Langevin thermostat.
-\todo{vah -> Finish these calculations}
-\begin{table}[h]
-	\caption{Reference density $\rho_0$ and mean water-water energy $E_{ww, norm}$ for various water models from NpT simulations at 300K/1bar.}\label{tab:ref_densities}
-	\small
-	\begin{tabularx}{\columnwidth}{@{}lrr@{}}
-		\toprule
-		Water Model       & $\rho_0$ /\AA$^{-3}$ & $E_{ww, norm}$ /$kcal\cdot mol^{-1} \cdot water^{-1}$  \\
-		\midrule
-		TIP3P    & 0.0329 & -9.533   \\
-		TIP3PFB  & 0.0334 & -11.369  \\
-		TIP4P    & 0.0332 & -9.856   \\
-		TIP4P/Ew & 0.0332 & -11.036  \\
-		TIP5P    & 0.0329 & -9.596   \\
-		SPC/E    & 0.0333 & -11.123  \\
-		OPC      & XXXX   & XXXX     \\
-		OPC3     & XXXX   & XXXX     \\
-		POL3     & XXXX   & XXXX     \\
-		\bottomrule
-	\end{tabularx}
-\end{table}
-If your simulation settings diverge significantly from the above setting or for other solvent models not listed either here or in the \software{cpptraj} manual, the density should be computed from a pure solvent simulation beforehand.
-
 \paragraph{Postprocessing and Analyzing GIST results}
-Cpptraj outputs quantities in OpenDX (.dx) file format which can be visualized in Pymol or VMD.
-Aside from thermodynamic quantities, GIST also calculates the number density of each atomic element in the solvent.
-Here is a list of the most important quantities calculated by GIST:
+Cpptraj outputs the quantities calculated as a "open data explorer" (.dx) file format which can be easily visualized in Pymol or VMD.
+Aside from the thermodynamic quantities, GIST also calculates the number densities of hydrogen and oxygen in each voxel.
+Here is a full list of quantities calculated by GIST \todo{This is incomplete!}:
 
@@
 For water, gO and gH are produced.}
+	%\item{[gH] Number density of hydrogen centers found in the voxel in units of the reference bulk density.}
 	\item{[Esw] Mean solute-water interaction energy density.}
@@
 	\item{[dTSorient] First order orientational entropy density .}
-	\item{[dTSsix] First order entropy density for the six-dimensional combined orientation and translation space.}
+	\item{[neighbor] Mean number of waters neighboring the water molecules found in this voxel multiplied by the voxel number density.}
 	\item{[dipole] Magnitude of mean dipole moment (polarization).}
+	\item{[order] (if \inlinecode{doorder} was specified) Average Tetrahedral Order Parameter.}
 	\item{[dipolex] x-component of the mean water dipole moment density}
@@
 	\item{[dipolez] z-component of the mean water dipole moment density}
+	\item{[Eij] (if \inlinecode{doeij} was specified) Water-water interaction matrix.}
 \end{itemize}
 
-Several dx files are written by default, such as the energy, entropy, and atomic number densities.
-All calculated quantities are written to the output file (default \inlinecode{gist-output.dat}) and can be used to generate DX files for other quantities. 
-A full list of all properties in the output is provided at the end of the tutorial.
-
-The output is organized by voxel and sorted according to their x,y, and z coordinates, describing the center point of the voxel.
-Energy and entropy quantities are reported in two different ways ´
-The \inlinecode{\_dens} quantities are normalized by the voxel volume and can be used to compute integrals.
-The \inlinecode{\_norm} quantities are instead normalized by the number of solvents in the respective voxel averaged over all considered frames, and are useful to visualize local water properties.
-Therefore, to convert from a \inlinecode{\_norm} quantity $X^k_{norm}$ in voxel k to a \inlinecode{\_dens} quantity $X^k_{dens}$ the following equation is applied:
-\begin{equation}
-	X^k_{dens} = \frac{X^k_{norm}}{V^k} \cdot \frac{N^k_{solvent}}{N_{frames}} 
-\end{equation}
-with $V^k$, the volume of voxel $k$, $N^k_{solvent}$, the solvent population in voxel $k$, and $N_{frames}$, the number of simulation frames considered in the GIST analysis.
-Acoordingly, converting $X^k_{dens}$ to  $X^k_{norm}$ is possible using the following equation:
-\begin{equation}
-	X^k_{norm} = X^k_{dens} \cdot V^k \cdot \frac{N_{frames}}{N_{frames}N^k_{solvent}} 
-\end{equation}
+Several dx files are written by default, such as the energies, entropies, number density of oxygen, and number density of hydrogen.
+All quantities mentioned above are written to the output file (output.dat) and can be used to generate DX files for other quantities.
+
 \section{Tutorial}
 Here, we aim to guide the reader through a GIST analysis of the streptavidin-biotin complex.
-We will
+In the process, the reader will learn how to set up and run a GIST calculation.
+During the tutorial, we will
 \begin{enumerate}
-	\item set up and run GIST calculations for a small molecule (biotin) and a protein (streptavidin) binding pocket.
-	\item compute the free energy of solvation for biotin using GIST.
-	\item visualize local contributions to $\dgsolv$ in the streptavidin binding pocket and around biotin.
-	\item compute the hydration contribution to the binding. Combined with the biotin--streptavidin interaction energy, the binding free energy can be estimated.
-	\item explore advantages and disadvantages of the GIST method when used to calculate  $\dgsolv$
+	\item compute the binding free energy of the complex from the individual energy contributions combined with the contribution of hydration computed using GIST.
+	\item visualize local contributions to $\dgsolv$ in the streptavidin binding pocket as well as around biotin.
 \end{enumerate}
 
-In the first part of the tutorial, we will look at how to run GIST calculations for small molecules and interpret the output.
-Afterwards, we will apply this knowledge to a protein binding pocket, visualize $\dgsolv$ hotspots and estimate a free energy of ligand binding using an end-state approach based on GIST.
-In addition, we will discuss all steps in further detail at the end.
-\todo{TODO}
-
 \subsection{Streptavidin/Biotin}
-The streptavidin-biotin complex binding is one of the strongest known non-covalent interactions involving a small molecule.
-Therefore, the combination of streptavidin and biotin is used routinely in biotechnology \cite{Dundas2013-streptavidin-review}, and the interaction between them has been investigated in detail \cite{McConnell2021-biotin}.
-
-\newcommand{\appr}{{\mathord{\sim}}}
-In nature, streptavidin occurs as a tetramer, and binds biotin with a binding constant of \SI{\sim e-14}{\per\Molar} \cite{Dundas2013-streptavidin-review}.
-Streptavidin has a so-called flap region, which closes the binding pocket after biotin binding.
-Here, we will investigate the binding of biotin to a closed streptavidin monomer for simplicity.
-The binding to the closed conformation of the tetramer has been estimated to have a free energy of \SI{-26.6}{\kilo\calorie\per\mol} \cite{Bansal2018-biotin}.
+The streptavidin-biotin complex is one of the strongest known non-covalent interactions involving a small molecule.
+Therefore, the combination of streptavidin and biotin is used routinely in biotechnology, and the interaction between them has been investigated in detail \todo{CITE}\cite{}.
+
+In nature, streptavidin occurs as a tetramer, and binds biotin with a binding constant of \todo{???}.
+Here, we will investigate the binding of biotin to a streptavidin monomer for simplicity.
+This interaction has been reported to have a binding constant of \todo{???}.
 
 \subsection{Tutorial data}
-We will use the crystal structure of streptavidin (PDB code: 1STP \cite{Weber1989-streptavidin-structure}) to start our calculations.
-For the beginners section, we will use the biotin ligand extracted from the crystal structure.
-For the second part, we will use the whole crystal structure.
+We will use the 1STP crystal structure to start our calculations.
 For the sake of simplicity, you can download prepared and solvated Amber topologies and structures based on 1STP by running
@@
 \end{lstlisting}
-This will download the tutorial files, including both the manuscript and the examples in the \inlinecode{code} folder.
-
-\subsection{Running GIST for Biotin}
-In this example, we will be running a simple GIST calculation of the biotin molecule to get familiar with the usage of GIST in \software{cpptraj}.
-\subsubsection{System preparation}
-First, extract biotin from the crystal structure and generate a topology and coordinate file compatible with Amber or a simulation package of your choice.
-The molecule should be solvated with a sufficiently large solvent box.
-Carefully check the employed force field, protonation state and partial charges.
-%a PDB file containing only biotin.
-Prepare your structure by adding hydrogens, making sure to choose the correct protonation state.
-Parametrize biotin with the forcefield of your choice.
-Solvate the parametrized biotin, making sure to choose a water box of large enough size so that the water on the box edges behaves bulk-like (a wall distance of at least 15\AA is recommended).
-A good starting point (though a bit dated) for the parametrization and solvation of small molecules is the Amber tutorial 4b (\url{http://ambermd.org/tutorials/basic/tutorial4b/}).
-
-\subsubsection{Equilibration}
-Equilibrate your solvated biotin in an NVT or NpT ensemble.
-Even when running the simulation in NVT, it is advisable to include an NpT equilibration step to adjust the box size beforehand.
-Use enough equilibration time to allow the solvent to relax. 
-For water, \SI{0.5}{\nano\second} are usually sufficient to equilibrate the density, but narrow binding pockets can take longer.
-
-\subsubsection{MD simulation}
-
-Now, you can run a restrained MD simulation based on the equilibrated structure of biotin for the subsequent GIST analysis.
-Restraints should be used during the MD because IST theory is not defined for moving solutes.
+This will download the tutorial files (including both the manuscript and the examples in the \software{code} folder).
+\subsection{System Preparation and Equilibration}
+If you want to skip this section, run 
+\begin{lstlisting}[style=bash]
+make equilibration-targets
+\end{lstlisting}
+in the code folder. This assumes that you have Python and Amber set up properly and that \software{pmemd.cuda} is in the PATH.
+
+The choice of the initial structure to be analyzed is crucial to the results 
+you can expect from the GIST calculation. Crystal structures may not be 
+representative of the dominant solution structure for a given analysis target. 
+In our example we use the final structure after a short equilibration run. A 
+more sophisticated approach would be identifying a dominant structure out of 
+the ensemble of sampled structures.  Depending on what properties you are 
+interested in, running a MD simulation of your structure beforehand and 
+choosing a dominant structure or even multiple structures through clustering 
+might therefore be beneficial.
+When starting with a crystal structure, special care has to be taken during the 
+preperation of the structure. Make sure to protonate your structure and take 
+care to choose a proper protonation state for the varying chemical moeties in 
+your structure. You will need to solvate your structure in a solvent box, which 
+should be chosen large enough so that the solvent around the region of interest 
+is not influenced through periodic images.
+
+When calculating solvation free energy differences between a complex and the corresponding monomers (the dissociated state), there are two options for dealing with configurational diversity in the structures.
+\todo{Maybe Valentin knows a citation for that?}
+One can choose to sample the complex and dissociated states individually, or to sample only one of them and assume that there is sufficient overlap in the sampled conformations.
+%One can choose to sample only the complex, assuming that all relevant conformations of the dissociated state are also found in the complex.
+%The probability distribution of the dissociated state can be obtained by reweighting the probabilities in the complex using the solvation free energies.
+%Alternatively, one can sample the complex and dissociated states individually.
+%This approach does not require an overlap between the conformational probabilities of the complex and dissociated states.
+%However, it is prone to statistical errors due to uncertainties in the ensembles, and requires the free energy contributions from the internal degrees of freedom of the monomers (in both the complex and the monomer states) to be known.
+\todo{This paragraph does not fit here well.
+Maybe put it somewhere else.}
+
+Here, we choose not to include any sampling of the conformational states. 
+It is therefore important that the conformations of the monomers in the complex and dissociated states match exactly. 
+Use \software{cpptraj} to produce biotin and streptavidin structures based on the complex by stripping the respective other molecule like this:
+\begin{lstlisting}[style=cpptraj]
+parm complex/solvated.parm7
+trajin complex/equil/EQUIL-DONE.rst
+strip ^2 parmout streptavidin/solvated.parm7
+trajout streptavidin/pre-equil.ncrst
+\end{lstlisting}
+(Adapt for biotin as necessary.)
+%We provide a script called \software{cpptraj\_remove\_mol.sh} that removes a single molecule from a topology and structure using \software{cpptraj} without altering the rest of the system.
+\todo{Explain cpptraj rather than our script.
+The script is for the automated workflow only.}
+
+Use the \software{equilibration.py} script (or your own) to perform short NVT and NpT equilibration runs on the starting structure.
+\todo {NPT}
+Note that the barostat might interfere with restrained equilibration of the complex, since different molecules are moved relative to each other when the volume changes.
+But since our equilibration script also contains unrestrained pressure 
+equilibration, we should not run into any trouble.
+
+Then split the equilibrated structure into biotin and streptavidin parts using \software{cpptraj\_remove\_mol.sh}.
+After that, you can use \software{equilibration.py -R} to run short restrained 
+equilibrations (where only the water is allowed to move) on the individual 
+systems.
+
+\subsection{Running MD for GIST}
+
+Now, you can run restrained MD simulations based on the equilibrated structures for subsequent GIST analyses.
+\todo {explain why the protein needs to be restrained for an MD simulation that will be analyzed using GIST}
 We apply a harmonic restraint of \SI{100}{\kilo\calorie\per\mole\per\angstrom\squared} to all heavy atoms to keep the system in place.
 We run \SI{100}{\nano\second} of simulation and store the conformations every \SI{100}{\pico\second}.
-A simulation length of at least \SI{20}{\nano\second} is recommended to obtain statistically independent snapshots of the water molecules.
+A simulation length of at least \SIrange{10}{20}{\nano\second} is recommended, to obtain statistically independent snapshots of the water.
 
@@
   nstlim=50000000, dt=0.002,
-	ntwr=50000, ntwx=5000, ntpr=5000,
+  ntwr=50000, ntwx=5000,
+  ntpr=5000,
 /
 \end{lstlisting}
-
-This file will run a 100 ns NpT MD simulation while keeping all atoms except for hydrogens and waters restrained. Note that these are very standard MD settings except for the restraints.
+Note that these are very standard MD settings except for the line that defines the restraints.
 
@@
 \end{lstlisting}
-\subsubsection{GIST analysis}
-To start our GIST analysis of biotin, we'll first have to decide on the parameters defining the GIST grid during the analyis.
-We recommend that you center the biotin molecule in space:
+
+\subsection{Running the GIST analysis for Streptavidin-Biotin Complex}
+For easier post-processing, we will use grids that contain the whole molecule.
+This could in theory be avoided for streptavidin and the complex, since only the surrounding of the biotin molecule (i.e., the binding pocket) is required for the analysis.
+We also recommend that you center the molecule in space.
+For each system (biotin, streptavidin, complex), produce a centered pdb file using \software{cpptraj}:
 \begin{lstlisting}[style=cpptraj]
@@
 strip :WAT
-	trajout biotin-centered.pdb
-\end{lstlisting}
-To decide on a grid size, find the extent (minimum and maximum) of the x-, \mbox{y-,} and z-coordinates of the molecule.
+trajout solute-centered.pdb
+\end{lstlisting}
+Then, find the extent (minimum and maximum) of the x-, \mbox{y-,} and z-coordinates for each system.
 This can be done e.g., using the \software{mdtraj} package in Python.
@@
 import mdtraj as md
-	pdb = md.load("biotin-centered.pdb")
+pdb = md.load("solute-centered.pdb")
 # mdtraj converts to nm.
 # *10 to convert back to Angstrom.
-	min_xyz = pdb.xyz[0].min(0) * 10.
-	max_xyz = pdb.xyz[0].max(0) * 10.
-	print(*list(zip("xyz", min_xyz, max_xyz)), sep="\n")
-\end{lstlisting}
+min = pdb.xyz[0].min(0) * 10.
+max = pdb.xyz[0].max(0) * 10.
+print(*list(zip("xyz", min, max)), sep="\n")
+\end{lstlisting}
+Then, decide on grid dimensions for every system.
 For a system centered at the origin with a spacing of \SI{0.5}{\angstrom} and a buffer of, e.g., \SI{15}{\angstrom} between the solute and the grid wall, you need $4.0\times (d+15)$ bins in the respective direction, where $d$ is the larger of $-$min and $+$max in the respective direction.
-To get bulk-like water properties at the grid border, a buffer distance of \SI{15.0}{\angstrom} is reasonable for biotin.
-
-Now, you can finally run the actual GIST analysis.
-Center the molecule as above, to make sure that its position matches the grid parameters you decided on.
-We recommend using either the PME or GPU implementation of GIST.
-For the GPU version, cpptraj needs to be compiled with the \inlinecode{-cuda} flag.
-While the GPU version is generally faster, the PME implementation matches the energy of the Amber MD engine more closely.
+\todo{ This is handled differently by FindCentroid.py: d = max - min; voxelnr = 
+d*5 for x,y,z each; we should handle it similiar or at least mention 
+FindCentroid.py here. 
+What does gisttools do if the grid is too small? -vah }
+We note that this centered grid is usually larger than it needs to be in some directions.
+If you want to save some extra memory and calculation time, feel free to align the grid more precisely to the minimum and maximum coordinates, or use a smaller grid that only covers the binding pocket.
+You can also rotate the molecule by its principal axes to fit the cuboid grid more exactly.
+However, this needs to be done \emph{before} the MD run, since rotating the trajectory damages the periodic box information.
+
+Now, you can finally run the actual GIST analysis. Center the molecule exactly in the same way as above, to make sure that its position matches the grid parameters that you decided on.
+We recommend using either the PME or GPU implementation of GIST, depending on whether you have a fast GPU. To use the GPU version of GIST, cpptraj needs to be compiled with the "-cuda" flag set.
+While the GPU version is generally faster, an advantage of the PME implementation is that it matches the energy calculation of the MD engine more closely.
 
@@
 autoimage !(:WAT) origin
-	# optionally remove the "pme" option
+# remove the "pme" option if you want to use the GPU
 gist griddim 100 100 100 out gist.dat \
   refdens 0.03287 pme
-\end{lstlisting}
-\subsubsection{Working with the GIST results}
-Running the GIST analysis above, an output data file (gist.dat) containing all calculated quantities for each voxel in the system is produced. 
-Additionally as outlined in section \ref{sec:running_GIST}, various quantities are written to OPenDX files for accessible visualisation.
-
-To faciliate an accessible way to interact with your GIST data, we recommend using the gisttools library for python3. Open the GIST output with gisttools:
-\begin{lstlisting}[style=python]
-	from gisttools.gist import load_gist_file
-	import matplotlib.pyplot as plt
-	gist = load_gist_file(
-	"gist.dat", struct="solute-centered.pdb",  eww_ref=-9.539)
-	print(f'n_frames = {gist.n_frames}')
-	print(f'rho0 = {gist.rho0}')
-\end{lstlisting}
-All information from the output file is now accessible through the gist object. 
-If the number of frames of the simulation (n\_frames) and the reference density (rho0) are not provided, they are automatically detected from the GIST data. 
-We will discuss a way to similarly detect the reference energy of the solvent-solvent energy Eww in the advanced section.
-
-The GIST data can be accessed and modified from the gist object in various ways.
-Gisttools provides access to single or multiple voxels through a .loc property similiar to the pandas library. 
-GIST quantity data for all voxels can be accessed through directly indexing the gist object, either by providing a single name or a list of names.
-Indexing with .loc works by providing a voxel index (either a index number or list of numbers) and optionally a column index (column name or list of names).
-Indices start from 0 and indexing will return a single entry, pandas series or pandas dataframe depending on the query:
-\begin{lstlisting}[style=python]
-	display(gist.loc[0, 'population'])
-	display(gist.loc[0])
-	display(gist.loc[[0,1], ['x', 'y', 'z']])
-	display(gist['dTSsix_norm'].head())
-	display(gist[['Eww_norm', 'Esw_norm']].head())
-\end{lstlisting}
-
-Gisttools provides access to all quantities listed in the GIST output file.
-In addition it calculates the total enthalpy per voxel (Eall) as a sum of Eww and Esw and the free energy of solvation (A) per voxel as a sum of Eall and -dTSsix:
-\begin{lstlisting}[style=python]
-	display(gist[['A_norm', 'Eall_norm']].head())
-\end{lstlisting}
-
-Information about the grid is stored in the grid property of the gist object, structural data is stored as mdtraj trajectory in the struct property:
-\begin{lstlisting}[style=python]
-	print(gist.grid)
-	print(gist.struct)
-\end{lstlisting}
-Gisttools also provides minimal OpenDX writing capabilities to visualize the calculated or modified GIST data.
-For visualization, you'll most likely want to write out the density-normalized GIST quantities (X\_dens):
-
-\begin{lstlisting}[style=python]
-	gist.save_dx('A_dens','A.dx')
-	with open('A.dx') as f:
-		print([next(f) for x in range(10)])
-\end{lstlisting}
-
-\subsubsection{Visualization the GIST results}
-Visualizing the GIST quantities is possible through the OpenDX (.dx) files, either provided directly from the GIST calculation or output by the gisttools library. 
-\subsection{Applying GIST for the Streptavidin-Biotin complex}
-In this section we will apply GIST to the full complex and discuss each step of the process in more detail to outline pitfalls.
-We will afterwards introduce advanced analysis workflows which allow you to modify and better visualize the resulting data.
-\subsubsection{System Preparation and Equilibration}
-If you want to skip this section, run 
-\begin{lstlisting}[style=bash]
-make equilibration-targets
-\end{lstlisting}
-in the code folder. This assumes that you have Python and Amber set up properly and that \software{pmemd.cuda} is in the PATH.
-
-The choice of initial structure is crucial to the GIST results.
-Crystal structures may not represent the dominant solution structure for a given target.
-In our example, we only use a single equilibrated structure.
-A more sophisticated approach would identify one or several dominant structures out of the ensemble - either gathered from experimental data or by clustering an MD simulation.
-
-When starting from an experimental structure, care has to be taken during the preparation step.
-Make sure to choose a proper protonation state for all titratable moeties.
-
-You will need to solvate your structure in a solvent box, which should be large enough to obtain unbiased long-range electrostatics.
-A good criterion for validating the box size is to check whether the water on the outside of the box behaves bulk-like according to the GIST output.
-Note that due to the solute restraints, some important solvent sites might be unavailable to exchange with the bulk solvent.
-Make sure to explicitly place water at such sites, if they are important to the process of interest e.g. waters in confined binding pockets, as they might be missed by automated solvation procedures.
-It is good practice to check your restrained MD simulation or solvent populations obtained by GIST afterwards to confirm wether all important solvent sites were properly solvated.
-
-When calculating solvation free energy differences between a complex and the corresponding monomers (the dissociated state), there are two options for dealing with configurational diversity in the structures.
-One can choose to sample the complex and dissociated states individually, or to sample only one of them and assume that there is sufficient overlap in the conformations.
-
-Here, we do not include any sampling of the conformational states.
-It is therefore important that the molecular conformations in the complex and dissociated states match exactly.
-Use \software{cpptraj} to produce biotin and streptavidin structures based on the complex by stripping the respective other molecule like this:
-\begin{lstlisting}[style=cpptraj]
-parm complex/solvated.parm7
-trajin complex/equil/EQUIL-DONE.rst
-strip ^2 parmout streptavidin/solvated.parm7
-trajout streptavidin/pre-equil.ncrst
-\end{lstlisting}
-(Adapt for biotin as necessary.)
-%We provide a script called \software{cpptraj\_remove\_mol.sh} that removes a single molecule from a topology and structure using \software{cpptraj} without altering the rest of the system.
-\todo{Explain cpptraj rather than our script.
-The script is for the automated workflow only.}
-
-Use the \software{equilibration.py} script (or your own procedure) to perform short NVT and NpT equilibration runs on the complex structure.
-Note that the equilibration script contains unrestrained NpT steps, because restraining two molecules at the same time might interfere with the pressure equilibration.
-
-Then split the equilibrated structure into biotin and streptavidin parts using \software{cpptraj\_remove\_mol.sh}.
-After that, use \software{equilibration.py -R} to run short restrained 
-equilibrations (where only the water is allowed to move) on the individual 
-systems.
-By using restraints, we can make sure that the monomer conformations match the complex.
-
-\subsubsection{MD Simulation}
-The MD can be run as described above for each monomer and the complex.
-
-\subsubsection{GIST analysis of Streptavidin-Biotin}
-For easier post-processing, here we use grids that contain the whole molecule.
-Smaller grids encompassing only the binding site could be used for streptavidin and the complex, but this would complicate the analysis.
-Furthermore, we recommend that you center the molecule in space.
-Calculate the GIST grid parameters as outlined in the biotin section for the complex and streptavidin and run the GIST analysis.
-For larger systems like whole proteins you can usually get away with a smaller buffer region compared to small molecules.
-To test whether your grid is large enough look at solvent properties close to the grid edges, e.g. the solvent-solvent energies.
-If these correspond to the respective bulk properties, it can be assumed that your grid is large enough to capture all perturbations of the solute on the solvent.
-Note that a different box-size or number of solvent molecules can skew the bulk properties slightly compared to the reference values provided above.
-In that case, checking for convergence of the solvent properties near the grid borders is sufficient but you might want to modify your reference values according to your results then as described in the next section.
-
-For the GIST analysis, we recommend using either the PME or GPU implementation of GIST.
-For the GPU version, cpptraj needs to be compiled with the \inlinecode{-cuda} flag.
-While the GPU version is generally faster, the PME implementation matches the energy of the Amber MD engine more closely.
-
-\subsubsection{Reference Values and Radial Convergence}
+# other settings you might want to change
+# (those are the defaults)
+# gridcntr 0 0 0 gridspacn 0.5 temp 300
+\end{lstlisting}
+\subsection{Reference Values and Radial Convergence}
 All GIST quantities should be expressed relative to bulk.
-This is automatically done by \software{cpptraj} for the entropy, using the bulk density specified by \inlinecode{refdens}.
-The solute-solvent energy $E_{sw}$ naturally tends towards zero in bulk.
+In practice, this is not problematic for the entropies, since they are computed relative to a distribution of randomly oriented molecules at the reference density.
+The solute-solvent energy $E_{sw}$ is also no problem, because it naturally tends to zero in bulk. 
 However, the solvent-solvent energy $E_{ww}$ needs to be referenced.
 
-The Amber manual provides reference densities and energies for several solvents.
-However, the exact energy values are different when using PME-GIST, and also depend on the box size.
-For quantitative analyses, it is recommended to compute an exact reference energy.
-
-The most accurate method to compute a reference energy is to run a GIST calculation of the pure solvent using the same energy method (PME/GPU) and a similar box size, and then compute the average solvent-solvent energy per molecule.
+The Amber manual provides reference energies for several solvents.
+However, the exact reference value is different when using PME-GIST, and also depends on the box size.
+For exact analyses such as computing the total $\dgsolv$ of a compound, it is important to compute an exact reference energy.
+
+The most accurate method to compute a reference energy is to run a separate GIST calculations of the solvent without solute using the same energy method (PME/GPU) and a similar box size, and then compute the average solvent-solvent energy per molecule.
 
@@
 
-\subsubsection{Visualizing $\dgsolv$}
+\subsection{Visualizing $\dgsolv$}
 Next, we want to visualize the 3-dimensional contributions to the free energy of hydration, as well as the entropy and enthalpy contributions.
 You can use PyMol\cite{pymol} or VMD\cite{vmd} to visualize the .dx files from GIST\@.
-%With \software{gistpp}, you can use the file from the convergence check. Additional OpenDX files can be produced using \inlinecode{gistpp -op makedx}.
+With \software{gistpp}, you can use the file from the convergence check. Additional OpenDX files can be produced using \inlinecode{gistpp -op makedx}.
 With \software{gisttools}, you can create one using e.g., \inlinecode{gist.save\_dx("A\_dens", "A\_dens.dx")}.
@@
 \end{lstlisting}
-
+% \todo{\$\textbackslash AA\$ should produce an angstrom sign in matplotlib (what a pain to have this printed out in \LaTeX\ tho) -vah} -- Å produces an angstrom sign ;-)
 Compute the energy (\inlinecode{Eall}) and entropy (\inlinecode{dTSsix}) contributions separately.
-You will find that the energy disfavors binding, because we do not yet include the interaction energy between biotin and streptavidin.
-
-You can also plot the difference between the complex and monomer contributions against the distance to biotin, using the complex coordinates for the holo structure.
-The expected result is shown in Figure~\ref{fig-dg-sums}.
-\todo{This figure is missing -vah}
-%\begin{figure}
-%	\centering
-%	\includegraphics[width=1.0\linewidth]{figures/deltaG-difference.png}
-%	\caption{TODO caption}\label{fig-dg-sums}
-%\end{figure}
-
-Using the \inlinecode{energy} command in \software{cpptraj}, you can compute this interaction energy.
+You will find that the energy is strongly unfavorable to binding, because we do not yet include the interaction energy between biotin and streptavidin.
+
+Using the \inlinecode{energy} command in \software{cpptraj}, you can compute this energy difference.
 We recommend using PME in combination with PME-GIST, but not with GPU-GIST\@.
@@
 However, we do not take the conformational transition of the streptavidin binding pocket into account.
-
 Prior computational studies suggest a $\Delta G$ of \SI{-26.6}{\kilo\calorie\per\mole} for the binding of biotin into the closed conformation of streptavidin \cite{Bansal2018-biotin}.
 This indicates that our result is in good agreement with literature.
-To improve the agreement with ITC measurements, the conformational changes of the binding pocket should be included.
-
-\subsubsection{Further steps}
-In this tutorial, we obtained a value for the binding of biotin into the closed conformation of the streptavidin binding pocket.
+However, the conformational changes of the binding pocket would need to be described to improve the agreement with experiments.
+\todo{Should we include a short introduction to other solvents with GIST? What about solvent/ion mixtures? Can we somehow distinguish python code from input files optically? -vah}
+\subsection{Further steps}
+In this tutorial, we obtained a value for the binding of biotin into the closed conformation of the streptavidin binging pocket.
 To investigate the effect of different conformations on the binding affinity, you can perform molecular dynamics simulations of the complex and/or monomers and perform GIST on multiple cluster representatives.
-To include the effect of lid closing in the binding process, free energy calculation methods such as umbrella sampling can be used.
-
-To speed up the calculation, smaller GIST grids could be used by focusing on the binding pocket, or by rotating the molecule by its principal axes to fit the cuboid grid more exactly.
-However, this should be done \emph{before} the MD, since rotating the trajectory damages the periodic box information \todo{@Dan: is this still the case?}.
+To include the effect of the lid closing transition in the binding process, free energy calculation methods such as umbrella sampling can be used.
+%\section{Content and links}
+%
+%A tutorial will normally draw on additional files and materials; clearly indicate where and how these are available, with links, and how they are being archived for the long-term and maintained so they stay current.
+%You will likely want to reference your GitHub repository as a central point to access all of this information, and then the GitHub repository may link out to other content as needed.
 
 \section{Recent developments}
-In recent years, several updates to the original GIST implementation have been published.
-The basic functionality of the program, however, has not been changed.
-In this chapter, we will shortly present each of those updates.
-\subsection{GPU implementation}
-In Reference~\cite{Kraml2020}, a GPU implementation of the energy calculation was presented, which typically increases the speed by 1--2 orders of magnitude.
-
-If \software{cpptraj} is compiled with GPU support, the energy calculation uses the GPU automatically, without any additional input.
-However, PME-GIST is not supported on the GPU, and no GPU code will be used when specifying \inlinecode{pme}.
-
-\subsection{PME implementation}
-In Reference~\cite{Chen2021}, a PME implementation of the GIST energy was presented.
-For typical systems, this implementation is slightly slower than GPU-GIST, but much faster than the original CPU code.
-Furthermore, PME-GIST offers the best agreement between GIST energies and those observed in an MD simulation.
-
-To run PME-GIST, the \inlinecode{pme} flag to \inlinecode{gist} can be used in \software{cpptraj}.
-The output will contain two additional columns: \inlinecode{PME\_dens} and \inlinecode{PME\_norm}.
-They contain the potential of solute and solvent evaluated at the solvent positions and divided by two.
-Furthermore, the \inlinecode{Eww} and \inlinecode{Esw} columns are also computed using PME, and are reported as usual.
-For typical use cases, we recommend using \inlinecode{Eww} and \inlinecode{Esw} over the \inlinecode{PME} column.
-\todo{vah: The cpptraj manual says that GIST PME energies cannot be decomposed into Esw and Eww -> what's true? Seems to conflict with what is written in this section.}
 \subsection{GIST with non-water solvents}
-In References~\cite{Kraml2020,Kamenik2020-gist-macrocycles,Waibl2022-gist-solvents}, an extension of GIST towards solvents other than water is described.
-
-To run a GIST calculation in a solvent other than water, use \inlinecode{solute} with a \software{cpptraj} selection mask.
-All other molecules will be treated as solvent.
-If there are multiple solvent species, also use the \inlinecode{solventmols} flag as described in Section~\ref{sec-salt-water}.
-
-The code will automatically choose three atoms per molecule to define the solvent orientation, and print this selection to the output.
-Sometimes, however, the automatic selection is not optimal.
-For instance, in methanol one can either incorporate the C--H bonds or the O--H bond in the selection.
-The latter is probably more relevant since it can incorporate hydrogen bonding effects.
-Assuming that the alcohol hydrogen is called H1, this could be specified as follows:
-
-\begin{lstlisting}[style=cpptraj]
-gist gridcntr <x> <y> <z> \
-griddim <Nx> <Ny> <Nz> gridspacn <val> \
-solute ^1 rigidatoms O1 H1 C1
-\end{lstlisting}
-
-Note that the central atom of the \inlinecode{rigidatoms} selection goes first.
-
 \subsection{GIST with salt-water mixtures}
-\label{sec-salt-water}
 In \cite{Waibl2021-gist-salt}, an extension of GIST was presented that can use salt-water mixtures as a solvent.
-This allows treating salting-out effects, although it was shown that the salting-out coefficient is over-predicted by GIST because of the first-order entropy approximation.
+This allows treating salting-out effects, although it was shown that the salting-out coefficient is over-predicted by GIST unless higher-order entropies are incorporated.
 Generally, GIST should be able to treat arbitrary solvent mixtures as long as each molecule is sufficiently rigid.
 
-In the \software{cpptraj} implementation of GIST, the \inlinecode{solute} keyword specifies which components of the system are treated as a solute.
+The \software{cpptraj} implementation of GIST allows the user to specify which components of the system to treat as a solute, using the \inlinecode{solute} keyword.
 Everything else will be treated as solvent.
@@
 
-The solvent accessible surface (SASA) represents the interface between regions that are occupied by the solute, and those which can be accessed by the solvent.
-GIST provides a very natural way of creating a SASA using an isosurface of the oxygen density gO.
-It can be visualized in PyMOL.
-The expected output is shown in Figure~\ref{fig-streptavidin_sasa}.
-
-\begin{lstlisting}[style=pymol]
-load ../gist.pdb
-load ../gist-gO.dx
-isomesh sas, gist-gO, 0.1
+The solvent accessible surface area (SASA) represents  the voxels that are close to the protein surface and occupied by oxygen water.
+It represents the surface where water can go, and potentially surfaces that a ligand atom can occupy.
+
+The surface is generated from the oxygen density (gO.dx) file using \software{gistpp}. 
+A voxel is deemed to be  part of the surface if it is occupied by water (gO > 0.3) and has at least one neighboring empty voxel (gO < 0.1).
+
+\begin{lstlisting}[style=bash]
+	gistpp -op sasa -i gO.dx -o sasa.dx
 \end{lstlisting}
@@
 	\centering
-	\includegraphics[width=1.0\linewidth]{figures/sasa-simple.png}
-	\caption{SASA for streptavidin focused on the binding pocket}\label{fig-streptavidin_sasa}
+	\includegraphics[width=1.0\linewidth]{streptavidin_sasa_close.png}
+	\caption{SASA for streptavidin focused on the binding pocket}\label{streptavidin_sasa}
 \end{figure}
 
-Here, the SASA was generated from a simulation of the holo protein structure.
+In the above figure, the SASA was generated from a simulation of the protein in the absence of the ligand.
 When the ligand is placed in the binding pocked, all atoms are within the SASA.
 
-\subsection{Defining the binding pocket}
+\subsection{Region around the binding pocket}
 Specific regions in a protein can be defined to calculate the thermodynamic quantities for that region.
@@
 
-Here, the binding pocket includes all voxels within \SI{3.5}{\angstrom} of the heavy atoms of biotin.
+In this example, the region defined in the binding pocket includes all voxels that are within 3.5 A of the heavy atoms of biotin.
 
@@
 	\centering
-	\includegraphics[width=1.0\linewidth]{streptavidin_bp_close.png}
-	\caption{Volume of streptavidin defined around \SI{3.5}{\angstrom} of biotin. }\label{streptavidin_volume}
+	\includegraphics[width=1.0\linewidth]{streptavidin_bp.png}
+	\caption{Volume of strepatividin defined around 3.5 A of biotin. }\label{streptavidin_sasa}
 \end{figure}
 
-\section{Theory}
-\label{sec:theory}
-GIST is an implementation of inhomogeneous fluid solvation theory (IST) \cite{Lazaridis1998} that discretizes the free energy of solvation $\dgsolv$ on a three-dimensional grid. 
-It was first devised by Nguyen et al.\ \cite{Nguyen2012} and its implementation in \software{cpptraj} was thoroughly described by Ramsey et al.\ \cite{Ramsey2016}.
-Here, we only provide a short overview of the theory behind GIST.
-For more detailed information, we recommend one of the more recent publications on developments in GIST \cite{Kraml2020,Chen2021}.
-
-\subsection{Solvation Entropy}
-
-%\begin{equation}
-%	\Delta S_\textit{solv} = \Delta S_\textit{sw} + \Delta S_\textit{ww}
-%\end{equation}
-%Here the solvation entropy is approximated from the contribution of solute-water interaction. 
-%\textbf{STILLNEEDS REVISON + ADD CITATIONS} The quantity $k_\textit{b}$represents the Boltzman constant, 
-%$\rho^\textit{o}$ is the bulk number density, $g_\textit{sw}\left(\textbf{r},\omega \right)$ is the solute-water 
-%pair correlation function in the solute's frame of reference, \textbf{r} is the coordinates of the oxygen atom of the water molecule,  
-%$\omega$ represents the Euler angles, and $\frac{1}{8\pi^2}$ is the volume of the orientational space, which acts as a normalization constant.
-%In bulk and in regions where the orientational 
-%entropy is uniform, the quantity $g_\textit{sw}\left(\textbf{r},\omega \right)$ is unity which leads to the first-order solvation entropy of 
-%bulk to be zero. Also, this quantity reaches unity as the distance from the solute molecule increases which leads for the calculation of 
-%the solvation entropy to be an approximation of the integral around the solute atom.
-
-IST expresses the free energy of solvation in terms of the solvent density in a coordinate system defined by the solute.
-The solute is kept fixed in space, following the standard solvation process established by Ben-Naim \cite{ben-naim-book}.
-
-For the entropy, this leads to an infinite expansion of correlations of increasing order.
-In a slightly simplified view, the first order is the solvent density $g_\textit{sw}\left(\textbf{r},\omega \right)$ at each position $\textbf{r}$ and orientation $\omega$.
-The first-order entropy omits all solvent-solvent (and higher) correlations, while the solute-solvent correlation is taken into account because the coordinates are relative to the solute.
-Thus, it is called the solute-water entropy $S_{sw}$.
-%The second order is the pair density, i.e., the probability of finding two solvent molecules in positions $\textbf{r}$ and $\textbf{r}^\prime$.
-
-The solvent density is expressed relative to the bulk density $\rho^0$. %$g_\textit{sw}\left(\textbf{r},\omega \right)$ is the solute-water 
-%pair correlation function in the solute's frame of reference, \textbf{r} is the coordinates of the oxygen atom of the water molecule,  
-%$\omega$ represents the Euler angles, and 
-The entropy integral is also normalized by $8\pi^2$, which is the volume of the orientational space.
-In bulk or at a high distance to the solute, the quantity $g_\textit{sw}\left(\textbf{r},\omega \right)$ approaches one, which leads to zero first-order entropy. 
-Therefore, no reference entropy is needed unless there is a numeric bias, e.g., when the MD frames are statistically dependent because the time between them is insufficient.
-\begin{equation}
-	\Delta S_\text{solv}
-	\approx \Delta S_\textit{sw}
-	\newcommand{\gro}{g_{sw}(\textbf{r},\omega)}
-	\equiv -R \frac{\rho^0}{8\pi^\textit{2}} \int \gro \ln\gro d\textbf{r}d\omega
-\end{equation}
-
-\subsubsection{Entropy Calculations in Cpptraj}
-In \software{cpptraj}, the calculation of solvation entropy is handled by two methods.
-
-In the first method, the solvation entropy is broken down into translational and orientational contributions.
-\begin{equation}
-\Delta S_\textit{solv} = \Delta S_\textit{trans} + \Delta S_\textit{orient}
-\end{equation}
-This is exact assuming that the distribution of the orientation $\omega$ is constant within a voxel $k$, but might converge slower than computing the entropy from the combined space (see below).
-
-The second method directly calculates the solvation entropy by evaluating the six-dimensional integral (3 for position and 3 for orientation).
-
-\begin{equation}
-g_\textit{vox} \left( \textbf{r}, \omega \right) \approx g_\textit{vox}(\textbf{r}) g_\textit{vox}(\omega)
-\end{equation}
-
-A first nearest neighbor (NN) approach is used to evaluate the density.
-The densities are computed relative to a homogeneous distribution of solvent molecules, given a bulk density $\rho^0$.
-
-The average entropy contributions per solvent molecule in voxel $k$ are calculated as
-
-\begin{equation}
-	S_{k}^\textit{trans} = -R \left( \gamma + \frac{1}{N_\textit{k}} \sum _{i=1}^{N_k} \ln g_{NN, \textit{i}}(\textbf{r}) \right),
-\end{equation}
-\begin{equation}
-S_{k}^\textit{orient} = -R \left( \gamma + \frac{1}{N_k} \sum _{i=1}^{N_k} \ln g_{NN, i}(\omega) \right)
-,
-\end{equation}
-and
-\begin{equation}
-S_\textit{k}^\text{six} = -R \left( \gamma + \frac{1}{N_\textit{k}} \sum _{i=1}^{N_k} \ln g_{NN, \textit{i}}(\textbf{r},\omega) \right),
-\end{equation}
-where $N_\textit{k}$ is the number of water molecules found in voxel k, $\gamma$ is Euler's constant accounting for the bias of the naive entropy estimator, and $g_\textit{NN}$ is the nearest neighbor estimate of the density.
-
-\subsection{Solvation Energy}
-The solvation energy is calculated from the water-water and water-solute interactions from the force field.
-\begin{equation}
-	\Delta E_\textit{solv} = \Delta E_\textit{sw} + \Delta E_\textit{ww}
-\end{equation}
-
-The solute-solvent energy $E_{sw}$ can be expressed in terms of the solvent density and the potential $U_{sw}(\textbf{r},\omega)$ induced by the solute.
-In practice, it is evaluated as the expectation value $\langle\cdot\rangle$ of the pairwise force field energy $U_{ij}$ between all solvent molecules in voxel $k$ and all solute atoms.
-
-\begin{equation}
-	\begin{aligned}
-		\label{eq-esw}
-		\Delta E_{sw}(\textbf{r}) \equiv& \frac{1}{8\pi^2} \int g_\textit{sw}\left(\omega|\textbf{r}\right) U_\textit{sw}\left(\textbf{r}, \omega\right) d\omega \\
-		\Delta E_{sw,k}=& \left\langle \sum_i^{\textit{solvent},k} \; \sum_j^\textit{solute} U_{ij}\right\rangle
-	\end{aligned}
-\end{equation}
-
-To localize $E_{sw}$ on the three-dimensional grid, every energy term is assigned to the voxel $k$ holding the solvent, and the expectation value in Equation~\ref{eq-esw} is computed per voxel.
-Similar to the entropy integrals, the solute-water solvation energy decays with increasing distance to the solute.
-Hence, it can be approximated by local spatial integrals.
-
-The solvent-solvent energy $E_{ww}$ is computed similarly.
-It can also be expressed in terms of density functions, but is practically computed as a sum over solvent-solvent interactions per voxel $k$.
-In contrast to $E_{sw}$, $E_{ww}$ does not tend to zero in bulk.
-Therefore, a reference corresponding to the average energy of a bulk water molecule needs to be subtracted.
-The referenced solvent-solvent energy will be denoted as $E_{ww}^\text{corr}$.
-
-\begin{equation}
-	\begin{aligned}
-		\Delta E_{ww}^\text{corr}(\textbf{r}) \equiv & \left(\frac{1}{8\pi^2} \right)^2 \rho^o \int g_{sw}(\omega|\textbf{r}) \\
-			& \times \left[g_{sw}(\textbf{r}^\prime , \omega^\prime) - g^0_{ww} (\textbf{r}, \omega, \textbf{r}^\prime, \omega^\prime )\right] \\
-			& \times U_{ww}(\textbf{r}, \omega, \textbf{r}^\prime, \omega^\prime)d\omega d\textbf{r}^\prime d\omega^\prime \\
-		\Delta E_{ww,k}^\text{corr} =& \frac{1}{N_k} \left\langle \sum_i^{\text{solvent},k} \; \sum_{j \neq i}^\text{solvent} U_{ij}\right\rangle - \left\langle E_{ww}^\text{bulk}\right\rangle
-	\end{aligned}
-\end{equation}
-
-\subsubsection{Interpreting the energy values}
-When computing the total $E_{ww}$ of the system, double-counting of interactions must be avoided.
-The total solvent-solvent energy of the system is as follows:
-
-\begin{equation}
-	\Delta E^\text{R}_{ww} = \sum_{k}^\text{voxels} \frac{E_{ww,k}^\text{corr}}{2}
-%	\Delta E^R_{ww} = n^R E^\text{bulk}_{ww} - E^\textit{R,corr}_{ww}
-\end{equation}
-
-However, when water is replaced from a small region $R$ of interest, such as a single water molecule, almost all interactions are with water molecules outside of this region.
-Therefore, there is no double counting, and the full solvent-solvent energy should be used.
-
-\begin{equation}
-	\Delta E^\text{R}_{ww} = \sum_{k}^{\text{voxels in }R} E_{ww,k}^\text{corr}
-%	\Delta E^\textit{R,norm}_\textit{ww} = E^\textit{R,norm}_\textit{ww} - 2E^\textit{bulk}_\textit{ww}
-\end{equation}
-
-When the region $R$ comprises more than one molecule, interactions within this region are double-counted while interactions to the outside are not.
-This could be solved using the solvent-solvent energy between each pair of voxels $E_{ww,k\;l}$:
-\begin{equation}
-	\Delta E^\text{R}_{ww} = 2\left(\sum_{k}^\text{voxels in R} E_{ww,k}^\text{corr}-\sum_{k}^\text{voxels in R}\sum_{l}^\text{l>k} E_{ww, k\;l}\right)
-\end{equation}
-While this is supported by the standard GIST implementation, it is rarely done due to the large size of the $E_{ww,k\;l}$ matrix.
-
-Integrating GIST values over the whole grid corresponds to a process where all the solvent is removed into bulk.
-However, when integrating over a small region, this process is less well-defined, since it is unclear to what extent the remaining solvent would reorganize.
-However, this depends on the local environment and on \emph{what} the solvent would be replaced by.
-Therefore, the effect of reorganization can not be directly included in the GIST method, but must be judged for each case individually.
-
-\subsubsection{PME energy}
-
-In the original version of GIST, the energies are calculated based on the the minimum image convention.
-In PME-GIST, the electrostatic energy $E_\text{elec}$ is calculated using the particle mesh Ewald (PME) method, which yields energies that are highly consistent with the Amber MD engine \cite{Chen2021}.
-The Lennard-Jones part $E_\text{lj}$ is computed separately in direct space.
-
-\begin{equation}
-	E_\text{total} = E_\text{elec} + E_\text{lj}
-\end{equation}
-
-During the electrostatics calculation, the system is treated as periodic and the energy is split into a short-range term $E_\text{dir}$, which is calculated in direct space using a distance cutoff, and a long-range term $E_\text{rec}$, which is calculated in reciprocal space.
-Additionally, there is a correction term $E_\text{self}$ (called $E_\text{corr}$ in the original publication \cite{Chen2021}), which corrects for the self-interaction of each solvent molecule in the reciprocal term.
-\begin{equation}
-	E_\textit{elec} = E_\text{dir} + E_\text{rec} + E_\text{self}
-\end{equation}
-The short-range Lennard-Jones contribution is computed in the direct space using a distance cutoff.
-Furthermore, a long-range correction term is computed that accounts for the contributions above this cutoff assuming a homogeneous distribution of particles.
-
-\begin{equation}
-	E_\text{lj} = E_\text{lj,\ short} +  E_\text{lj,\ corr}
-\end{equation}
-
-\section{Cheat sheet}
+\section{Cheatsheet}
 %This will be a cheatsheet containing all options for cpptrajs GIST implementation
-The checklist ``The CPPTRAJ GIST action'' provides a cheat sheet summarizing the important options to the GIST action and its output, as well as frequently used equations.
-
 \subsection{Installation}
 GIST calculations in cpptraj can be sped up through various levels of 
-parallelization.
-If cpptraj was compiled with the '-cuda' configure flag set, the GPU accelerated version of GIST is used automatically. 
-Furthermore, if cpptraj was compiled with OpenMP support, the entropy calculation will be parallelized.
-
-For detailed instructions on how to install \software{cpptraj}, we refer the reader to the AmberMD GitHub page \url{https://github.com/Amber-MD/cpptraj}.
+parallelization. When cpptraj was compiled with the '-cuda' configure flag set, 
+the GPU accelerated version of GIST is used automatically.
 
@@
 
-\begin{checklist}{The CPPTRAJ GIST action}\label{cheatsheet}
+\begin{checklist}{The CPPTRAJ action}
 

### 2023-01-16 manuscript/manuscript.tex
Updates from Overleaf

--- 2022-12-22 manuscript/manuscript.tex
@@
 \DeclareSIUnit\Molar{M}
-\DeclareSIUnit\calorie{cal}
-\DeclareSIUnit\kcalPerMolASqr{\kilo\calorie\per\mole\per\angstrom\squared}
-%\DeclareSIUnit{\Molar}{M}
-\DeclareSIUnit\angstrom{\text{\AA}}
-%\DeclareSIPrefix\micro{\text{\textmu}}{0}
-\sisetup{
-	%math-rm=\mathsf,
-	text-rm=\sffamily
-}
-
 \usepackage[italic,LGRgreek]{mathastext}
@@
 
+\DeclareSIUnit{\calorie}{cal}
+\DeclareSIUnit{\kcalPerMolASqr}{\kilo\calorie\per\mole\per\angstrom\squared}
+\DeclareSIUnit{\Molar}{M}
+
 % TODO: delete for final build
@@
 
+We also show examples using the \software{gistpp} program, which can be obtained from \url{https://github.com/KurtzmanLab/Gist-Post-Processing} (code and documentation).
+
 If the reader prefers to skip the MD and GIST calculation, we also provide the GIST output files with this tutorial, such that the post-processing can be done without any expensive calculations.
@@
 
-Most of the visualisations and results in this tutorial can also be obtained using the \software{gistpp} program, which is available from \url{https://github.com/KurtzmanLab/Gist-Post-Processing} (code and documentation).
-\software{Gistpp} can directly manipulate \software{OpenDX} files from the GIST output.
-Its usage and some examples were shown in a previous tutorial on GIST\cite{Ramsey2016}.
-
 \section{Running GIST}
-\label{sec:running_GIST}
 %\subsection{Running the GIST analysis}
@@
 
-It is important to set the reference density $\rho_0$ according to the solvent model used in your simulation.
-The solvent-solvent enthalpy $E_{ww}$ also needs to be referenced, but this is usually done in the post-processing stage.
-
+Setting the reference density according to the water model used in your simulation is important to get correct results.
 Densities of typical water models can be found in the \software{cpptraj} manual.
-We also provide the densities for water models commonly used with AMBER forcefields in table \ref{tab:ref_densities}.
-These densities were calculated from \SI{1}{\micro\second}\pdfcomment{weird small mu symbol} MD simulations at \SI{1}{\bar} and \SI{300}{\kelvin} using the Berendsen barostat and the Langevin thermostat.
-The water boxes were prepared with \software{tleap} to a size of \SI{100 x 100 x 100}{\angstrom} before pressure equilibration.
-\todo{vah -> Finish these calculations}
-\begin{table}[h]
-	\caption{Reference density $\rho_0$ and mean water-water energy $E_{ww, norm}$ for various water models from NpT simulations at 300K/1bar.}\label{tab:ref_densities}
-	\small
-	\begin{tabularx}{\columnwidth}{@{}lrr@{}}
-		\toprule
-		Water Model       & $\rho_0$ /\AA$^{-3}$ & $E_{ww, norm}$ /$kcal\cdot mol^{-1} \cdot water^{-1}$  \\
-		\midrule
-		TIP3P    & 0.0329 & -9.533   \\
-		TIP3PFB  & 0.0334 & -11.369  \\
-		TIP4P    & 0.0332 & -9.856   \\
-		TIP4P/Ew & 0.0332 & -11.036  \\
-		TIP5P    & 0.0329 & -9.596   \\
-		SPC/E    & 0.0333 & -11.123  \\
-		OPC      & XXXX   & XXXX     \\
-		OPC3     & XXXX   & XXXX     \\
-		POL3     & XXXX   & XXXX     \\
-		\bottomrule
-	\end{tabularx}
-\end{table}
-If your simulation settings diverge significantly from the above setting or for other solvent models not listed either here or in the \software{cpptraj} manual, the density should be computed from a pure solvent simulation beforehand.
+For other models, the density should be computed from a pure water simulation.
 
@@
 
-The output is organized by voxel and sorted according to their x, y, and z coordinates, describing the center point of the voxel.
-Energy and entropy quantities are reported in two different ways.
+Energy and entropy quantities are reported in two different ways in the GIST output.
 The \inlinecode{\_dens} quantities are normalized by the voxel volume and can be used to compute integrals.
-The \inlinecode{\_norm} quantities are instead normalized by the number of solvents in the respective voxel averaged over all considered frames, and are useful to visualize local water properties.
-To convert from a \inlinecode{\_norm} quantity $X^k_{norm}$ in voxel $k$ to a \inlinecode{\_dens} quantity $X^k_{dens}$, the following equation can be applied:
-\begin{equation}
-	X^k_{dens} = \frac{X^k_{norm}}{V^k} \cdot \frac{N^k_{solvent}}{N_{frames}}
-\end{equation}
-with $V^k$, the volume of voxel $k$, $N^k_{solvent}$, the solvent population in voxel $k$, and $N_{frames}$, the number of simulation frames considered in the GIST analysis.
+The \inlinecode{\_norm} quantities are instead normalized by the average number of solvents in the respective voxel, and are useful to visualize local water properties.
 
@@
 \subsubsection{MD simulation}
-
-Now, you can run a restrained MD simulation based on the equilibrated structure of biotin for the subsequent GIST analysis.
-Restraints should be used during the MD because IST theory is not defined for moving solutes.
+Now, you can run restrained MD simulations based on the equilibrated structures.
+Restraints should be used because IST theory is not defined for moving solutes.
 We apply a harmonic restraint of \SI{100}{\kilo\calorie\per\mole\per\angstrom\squared} to all heavy atoms to keep the system in place.
@@
 \end{lstlisting}
-
-This file will run a \SI{100}{\nano\second} NpT MD simulation while keeping all atoms except for hydrogens and waters restrained.
 Note that these are very standard MD settings except for the restraints.
@@
 \end{lstlisting}
-\subsubsection{GIST analysis}
-To start our GIST analysis of biotin, we'll first have to decide on the parameters defining the GIST grid during the analyis.
-We recommend that you center the biotin molecule in space:
-\begin{lstlisting}[style=cpptraj]
-parm solvated.parm7
-trajin md-01.nc 1 1 1
-autoimage !(:WAT) origin
-strip :WAT
-trajout biotin-centered.pdb
-\end{lstlisting}
-To decide on a grid size, find the extent (minimum and maximum) of the x-, \mbox{y-,} and z-coordinates of the molecule.
-This can be done e.g., using the \software{mdtraj} package in Python.
-\begin{lstlisting}[style=python]
-import mdtraj as md
-pdb = md.load("biotin-centered.pdb")
-# mdtraj converts to nm.
-# *10 to convert back to Angstrom.
-min_xyz = pdb.xyz[0].min(0) * 10.
-max_xyz = pdb.xyz[0].max(0) * 10.
-print(*list(zip("xyz", min_xyz, max_xyz)), sep="\n")
-\end{lstlisting}
-For a system centered at the origin with a spacing of \SI{0.5}{\angstrom} and a buffer of, e.g., \SI{15}{\angstrom} between the solute and the grid wall, you need $4.0\times (d+15)$ bins in the respective direction, where $d$ is the larger of $-$min and $+$max in the respective direction.
-To get bulk-like water properties at the grid border, a buffer distance of \SI{15.0}{\angstrom} is reasonable for biotin.
-
-Now, you can finally run the actual GIST analysis.
-Center the molecule as above, to make sure that its position matches the grid parameters you decided on.
-We recommend using either the PME or GPU implementation of GIST.
-For the GPU version, cpptraj needs to be compiled with the \inlinecode{-cuda} flag.
-While the GPU version is generally faster, the PME implementation matches the energy of the Amber MD engine more closely.
-
-\begin{lstlisting}[style=cpptraj]
-parm solvated.parm7
-trajin md-01.nc
-autoimage !(:WAT) origin
-# optionally remove the "pme" option
-gist griddim 100 100 100 out gist.dat \
-refdens 0.03287 pme
-\end{lstlisting}
-\subsubsection{Working with the GIST results}
-Running the GIST analysis above, an output data file (gist.dat) containing all calculated quantities for each voxel in the system is produced. 
-Additionally as outlined in section \ref{sec:running_GIST}, various quantities are written to OpenDX files.
-
-To interact with your GIST data, we recommend the gisttools library for python3. Open the GIST output with gisttools:
-\begin{lstlisting}[style=python]
-from gisttools.gist import load_gist_file
-import matplotlib.pyplot as plt
-gist = load_gist_file(
-"gist.dat", struct="solute-centered.pdb",  eww_ref=-9.539)
-print(f'n_frames = {gist.n_frames}')
-print(f'rho0 = {gist.rho0}')
-\end{lstlisting}
-All information from the output file is now accessible through the gist object. 
-If the number of frames of the simulation (n\_frames) and the reference density (rho0) are not provided, they are automatically detected from the GIST data. 
-We will discuss a way to similarly detect the reference energy of the solvent-solvent energy Eww in the advanced section.
-
-The GIST data can be accessed and modified from the gist object in various ways.
-Gisttools provides access to single or multiple voxels through a .loc property similiar to the pandas library. 
-GIST quantity data for all voxels can be accessed through directly indexing the gist object, either by providing a single name or a list of names.
-Indexing with .loc works by providing a voxel index (either a index number or list of numbers) and optionally a column index (column name or list of names).
-Indices start from 0 and indexing will return a single entry, pandas series or pandas dataframe depending on the query:
-\begin{lstlisting}[style=python]
-gist.loc[0, 'population'] # single value
-gist.loc[0] # Series
-gist.loc[[0,1], ['x', 'y', 'z']] # DataFrame
-gist['dTSsix_norm'].head() # Series
-gist[['Eww_norm', 'Esw_norm']].head() # DataFrame
-\end{lstlisting}
-
-Gisttools provides access to all quantities listed in the GIST output file.
-Additionally, it calculates the total enthalpy per voxel (Eall) as a sum of Eww and Esw, and the free energy of solvation (A) by subtracting dTSsix from Eall.
-
-Information about the grid is stored in the grid property of the gist object, structural data is stored as an \software{mdtraj} trajectory in the struct property:
-
-Gisttools also provides minimal OpenDX writing capabilities to visualize the calculated or modified GIST data.
-For visualization, you'll most likely want to write out the density-normalized GIST quantities (X\_dens):
-
-\begin{lstlisting}[style=python]
-gist.save_dx('A_dens','A.dx')
-\end{lstlisting}
-
-This tutorial discusses the visualization of OpenDX files in Section~\ref{sec:visualization}.
+
 \subsection{Applying GIST for the Streptavidin-Biotin complex}
-In this section we will apply GIST to the full complex and discuss each step of the process in more detail to outline pitfalls.
-We will afterwards introduce advanced analysis workflows which allow you to modify and better visualize the resulting data.
 \subsubsection{System Preparation and Equilibration}
@@
 In our example, we only use a single equilibrated structure.
-A more sophisticated approach would be to identify several dominant structures in the ensemble -- either from experimental ensembles or by clustering an MD simulation.
-
-When starting from an experimental structure, care has to be taken during the preparation step.
+A more sophisticated approach would be to identify one or several dominant structures out of the ensemble.
+Depending on what properties you are interested in, it might be beneficial to run an MD simulation choose dominant structures through clustering.
+
+When starting from a crystal structure, care has to be taken during the preparation.
 Make sure to choose a proper protonation state for all titratable moeties.
-
 You will need to solvate your structure in a solvent box, which should be large enough to obtain unbiased long-range electrostatics.
 A good criterion for validating the box size is to check whether the water on the outside of the box behaves bulk-like according to the GIST output.
-Note that, due to the restraints, some solvent sites might be inaccessible to the bulk solvent.
-To investigate such positions, make sure to explicitly place water molecules there.
-It is good practice to check your restrained MD simulation or solvent populations obtained by GIST to confirm whether all relevant sites were properly solvated.
 
@@
 
-Use the \software{equilibration.py} script (or your own procedure) to perform short NVT and NpT equilibration runs on the complex structure.
-Note that the equilibration script contains unrestrained NpT steps, because restraining two molecules at the same time might interfere with the pressure equilibration.
+Use the \software{equilibration.py} script (or your own) to perform short NVT and NpT equilibration runs on the complex structure.
+It is important to include unrestrained NpT steps, because restraining two molecules at the same time might interfere with the pressure equilibration.
 
@@
 Smaller grids encompassing only the binding site could be used for streptavidin and the complex, but this would complicate the analysis.
+%This could in theory be avoided for streptavidin and the complex, since only the surrounding of the biotin molecule (i.e., the binding pocket) is required for the analysis.
 Furthermore, we recommend that you center the molecule in space.
-Calculate the GIST grid parameters as outlined in the biotin section for the complex and streptavidin and run the GIST analysis.
-For larger systems like whole proteins you can usually get away with a smaller buffer region compared to small molecules.
-To test whether your grid is large enough look at solvent properties close to the grid edges, e.g. the solvent-solvent energies.
-If these correspond to the respective bulk properties, it can be assumed that your grid is large enough to capture all perturbations of the solute on the solvent.
-Note that a different box-size or number of solvent molecules can skew the bulk properties slightly compared to the reference values provided above.
-In that case, checking for convergence of the solvent properties near the grid borders is sufficient but you might want to modify your reference values according to your results then as described in the next section.
-
-For the GIST analysis, we recommend using either the PME or GPU implementation of GIST.
+For each system (biotin, streptavidin, complex), produce a centered pdb file using \software{cpptraj}:
+\begin{lstlisting}[style=cpptraj]
+parm solvated.parm7
+trajin md-01.nc 1 1 1
+autoimage !(:WAT) origin
+strip :WAT
+trajout solute-centered.pdb
+\end{lstlisting}
+
+Then, find the extent (minimum and maximum) of the x-, \mbox{y-,} and z-coordinates for each system.
+This can be done e.g., using the \software{mdtraj} package in Python.
+\begin{lstlisting}[style=python]
+import mdtraj as md
+pdb = md.load("solute-centered.pdb")
+# mdtraj converts to nm.
+# *10 to convert back to Angstrom.
+min = pdb.xyz[0].min(0) * 10.
+max = pdb.xyz[0].max(0) * 10.
+print(*list(zip("xyz", min, max)), sep="\n")
+\end{lstlisting}
+Then, decide on grid dimensions for every system.
+For a system centered at the origin with a spacing of \SI{0.5}{\angstrom} and a buffer of, e.g., \SI{15}{\angstrom} between the solute and the grid wall, you need $4.0\times (d+15)$ bins in the respective direction, where $d$ is the larger of $-$min and $+$max in the respective direction.
+\todo{ This is handled differently by FindCentroid.py: d = max - min; voxelnr = 
+d*5 for x,y,z each; we should handle it similiar or at least mention 
+FindCentroid.py here. 
+What does gisttools do if the grid is too small? -vah }
+A buffer distance of \SI{15.0}{\angstrom} is reasonable for biotin, but smaller values are sufficient for the larger streptavidin and complex systems.
+
+Now, you can finally run the GIST analysis.
+Center the molecule as above, to make sure that its position matches the grid parameters you decided on.
+We recommend using either the PME or GPU implementation of GIST.
 For the GPU version, cpptraj needs to be compiled with the \inlinecode{-cuda} flag.
 While the GPU version is generally faster, the PME implementation matches the energy of the Amber MD engine more closely.
+
+\begin{lstlisting}[style=cpptraj]
+parm solvated.parm7
+trajin md-01.nc
+autoimage !(:WAT) origin
+# optionally remove the "pme" option
+gist griddim 100 100 100 out gist.dat \
+  refdens 0.03287 pme
+\end{lstlisting}
 
@@
 You can use PyMol\cite{pymol} or VMD\cite{vmd} to visualize the .dx files from GIST\@.
-%With \software{gistpp}, you can use the file from the convergence check. Additional OpenDX files can be produced using \inlinecode{gistpp -op makedx}.
+With \software{gistpp}, you can use the file from the convergence check. Additional OpenDX files can be produced using \inlinecode{gistpp -op makedx}.
 With \software{gisttools}, you can create one using e.g., \inlinecode{gist.save\_dx("A\_dens", "A\_dens.dx")}.
@@
 \subsubsection{Contribution of Hydration to Binding}
-\label{sec:binding_contributions}
 In the next step, calculate the $\dgsolv$ contributions around biotin in each system (biotin, streptavidin, and complex).
@@
 \end{lstlisting}
-
+% \todo{\$\textbackslash AA\$ should produce an angstrom sign in matplotlib (what a pain to have this printed out in \LaTeX\ tho) -vah} -- Å produces an angstrom sign ;-)
 Compute the energy (\inlinecode{Eall}) and entropy (\inlinecode{dTSsix}) contributions separately.
 You will find that the energy disfavors binding, because we do not yet include the interaction energy between biotin and streptavidin.
-To make this easier, gisttool includes a function to integrate in a radius around atom centers:
-\begin{lstlisting}[style=python]
-centers = {biotin: biotin_x, 
-	   strept: strept_x, 
-	   compl: compl_x}
-names = {biotin: 'Biotin', 
-	   strept: 'Streptavidin', 
-	   compl: 'Complex'}
-cols = ['Eall_dens','dTSsix_dens']
-rmax = 12
-results = { names[gist] : 
-	   gist.integrate_around(
-	   		cols, rmax, centers[gist]) 
-	   for gist in [biotin, strept, compl]}
-display( pd.DataFrame( results ))
-
-\end{lstlisting}
-You can also plot the difference between the complex and monomer contributions against the distance to biotin, using the complex coordinates for the holo structure:
-\begin{lstlisting}[style=python]
-fig, (ax1, ax2) = plt.subplots(1, 2, figsize=(8, 4))
-cutoff = 12
-ax1.plot(bins, np.cumsum(biotin_rdf), label='biotin')
-ax1.plot(bins, np.cumsum(strept_rdf), label='streptavidin')
-ax1.plot(bins, np.cumsum(compl_rdf), label='complex')
-ax1.legend()
-ax1.axvline(cutoff, color='k', linestyle='--')
-ax1.set_xlabel('distance to biotin [Å]')
-ax1.set_ylabel('$\Delta G$ contributions [Å]')
-ax1.grid()
-
-difference = compl_rdf - biotin_rdf - strept_rdf
-ax2.plot(bins, np.cumsum(difference))
-ax2.axvline(cutoff, color='k', linestyle='--')
-ax2.set_xlabel('distance to biotin [Å]')
-ax2.set_ylabel('$\Delta \Delta G$ [kcal/mol]')
-ax2.grid()
-plt.show()
-\end{lstlisting}
+
 You can also plot the difference between the complex and monomer contributions against the distance to biotin, using the complex coordinates for the holo structure.
 The expected result is shown in Figure~\ref{fig-dg-sums}.
-\todo{Missing caption}
+
 \begin{figure}
@@
 For typical use cases, we recommend using \inlinecode{Eww} and \inlinecode{Esw} over the \inlinecode{PME} column.
-\todo{vah: The cpptraj manual says that GIST PME energies cannot be decomposed into Esw and Eww -> what's true? Seems to conflict with what is written in this section.}
+
 \subsection{GIST with non-water solvents}
@@
 \section{Visualization}
-\label{sec:visualization}
+
 \subsection{Visualizing DX files}
@@
 The expected output is shown in Figure~\ref{fig-streptavidin_gO}.
-\todo{VJ -> include the PyMOL settings somewhere, this is not the default output}
+
 % The current font makes the O look like a tofu / placeholder
 \begin{lstlisting}[style=pymol]
-load output/streptavidin/gist.pdb, streptavidin
+load streptavidin.pdb
 as surface
 color gray70, streptavidin
-load output/streptavidin/gist-gO.dx, gO
+load gist-gO.dx, gO
 isomesh gO_high, gO, 2
@@
 When the ligand is placed in the binding pocked, all atoms are within the SASA.
-\todo{I doubt this is the holo/complex, should be the apo streptavidin, right? -vah´}
 
@@
 For example, voxels that are part of the binding pocket can be defined using a ligand molecule and a distance criteria.
-%In \software{gistpp}, voxels that are within the distance criteria from any ligand heavy atom is considered as part of the region.
+In \software{gistpp}, voxels that are within the distance criteria from any ligand heavy atom is considered as part of the region.
 The resulting file is a binary file and can be visualized using \software{pymol}.
 \todo{Not GISTPP}
-In section~\ref{sec:binding_contributions} we show how to integrate over regions such as the binding pocket.
-To visualize the considered region, we can create a OpenDX file where we set the considered voxels to 1 and all others to 0.
-Here, we show what this would look like for a radius of \SI{3.5}{\angstrom} around the heavy atoms of biotin.
-%\begin{lstlisting}[style=bash]
-%gistpp -op defbp -i gO.dx -i2 ligand.pdb 
-%
-%-opt const 3.5 -o gO_3.5.dx
-%\end{lstlisting}
-\begin{lstlisting}[style=python]
-def select(traj, sel):
-	"Slice a Trajectory by selection mask."
-	return traj.atom_slice(traj.top.select(sel))
-### Load GIST data
-compl = gt.gist.load_gist_file(
-	'complex/gist/gistoutlong.dat', 
-	struct='complex/gist/gist_nowat.pdb')
-### Define ligand atoms and find voxels around them
-rmax = 3.5
-biotin_mask = 'resname BTN and not element H'
-btn = select(compl.struct, biotin_mask).xyz[0] * 10.
-ind, _, _ = compl.distance_to_spheres(btn, rmax)
-
-### Set voxels within 3.5A to 1, write out dx file
-compl.loc[ind, 'BP'] = 1
-compl.save_dx('BP', 'binding_pocket.dx')
-\end{lstlisting}
+
+\begin{lstlisting}[style=bash]
+gistpp -op defbp -i gO.dx -i2 ligand.pdb
+
+-opt const 3.5 -o gO_3.5.dx
+\end{lstlisting}
+
+Here, the binding pocket includes all voxels within \SI{3.5}{\angstrom} of the heavy atoms of biotin.
 
@@
 \begin{equation}
-	\Delta E^\text{R}_{ww} = \sum_{k}^\text{voxels} \frac{E_{ww,k}^\text{corr}}{2}
+	\Delta E^\text{total}_{ww} = \sum_{k}^\text{voxels} \frac{E_{ww,k}^\text{corr}}{2}
 %	\Delta E^R_{ww} = n^R E^\text{bulk}_{ww} - E^\textit{R,corr}_{ww}
@@
 \begin{equation}
-	\Delta E^\text{R}_{ww} = \sum_{k}^{\text{voxels in }R} E_{ww,k}^\text{corr}
+	\Delta E^\text{total}_{ww} = \sum_{k}^{\text{voxels in }R} E_{ww,k}^\text{corr}
 %	\Delta E^\textit{R,norm}_\textit{ww} = E^\textit{R,norm}_\textit{ww} - 2E^\textit{bulk}_\textit{ww}
@@
 When the region $R$ comprises more than one molecule, interactions within this region are double-counted while interactions to the outside are not.
-This could be solved using the solvent-solvent energy between each pair of voxels $E_{ww,k\;l}$:
-\begin{equation}
-	\Delta E^\text{R}_{ww} = 2\left(\sum_{k}^\text{voxels in R} E_{ww,k}^\text{corr}-\sum_{k}^\text{voxels in R}\sum_{l}^\text{l>k} E_{ww, k\;l}\right)
-\end{equation}
-While this is supported by the standard GIST implementation, it is rarely done due to the large size of the $E_{ww,k\;l}$ matrix.
+This could be solved using the solvent-solvent energy between each pair of voxels $E_{ww,kl}$.
+While this is supported by the standard GIST implementation, it is rarely done due to the large size of the $E_{ww,kl}$ matrix.
 

### 2023-02-10 manuscript/manuscript.tex
test overleaf commit. only a test comment, feel free to undo this

--- 2022-12-22 manuscript/manuscript.tex
@@
 These densities were calculated from \SI{1}{\micro\second}\pdfcomment{weird small mu symbol} MD simulations at \SI{1}{\bar} and \SI{300}{\kelvin} using the Berendsen barostat and the Langevin thermostat.
-The water boxes were prepared with \software{tleap} to a size of \SI{100 x 100 x 100}{\angstrom} before pressure equilibration.
+The water boxes were prepared with \software{tleap} to a size of \SI{100}{\angstrom} x \SI{100}{\angstrom} x \SI{100}{\angstrom} before pressure equilibration.
 \todo{vah -> Finish these calculations}
@@
 \end{lstlisting}
+% test edit
 

### 2023-02-10 manuscript/manuscript.tex
Very minor text changes, remove type = Journal Article from some references.

@@
 \end{equation}
-with $V^k$, the volume of voxel $k$, $N^k_{solvent}$, the solvent population in voxel $k$, and $N_{frames}$, the number of simulation frames considered in the GIST analysis.
+with $V^k$, the volume of voxel $k$, $N^k_{solvent}$, the number of solvent molecules in $k$ over all simulation frames, and $N_{frames}$, the number of frames.
 
@@
 For typical use cases, we recommend using \inlinecode{Eww} and \inlinecode{Esw} over the \inlinecode{PME} column.
-\todo{vah: The cpptraj manual says that GIST PME energies cannot be decomposed into Esw and Eww -> what's true? Seems to conflict with what is written in this section.}
 \subsection{GIST with non-water solvents}

### 2023-03-21 manuscript/manuscript.tex
Minor text updates, add caption of Fig 4

@@
 The expected result is shown in Figure~\ref{fig-dg-sums}.
-\todo{Missing caption}
 \begin{figure}
@@
 	\includegraphics[width=1.0\linewidth]{figures/deltaG-difference.png}
-	\caption{TODO caption}\label{fig-dg-sums}
+	\caption{Left: $\dgsolv$ contributions around the location of biotin in the binding pocket, evaluated in the biotin, streptavidin, and complex systems.
+Each line represents a cumulative sum plotted against the distance to biotin.
+Right: The hydration contribution to binding, evaluated as the difference between the lines in the left panel.
+The vertical dashed line is at 12 \AA\ and represents the chosen integration cutoff.}\label{fig-dg-sums}
 \end{figure}
@@
 The expected output is shown in Figure~\ref{fig-streptavidin_gO}.
-\todo{VJ -> include the PyMOL settings somewhere, this is not the default output}
+\todo{vah -> include the PyMOL settings somewhere, this is not the default output}
 % The current font makes the O look like a tofu / placeholder
@@
 
-Here, the SASA was generated from a simulation of the holo protein structure.
-When the ligand is placed in the binding pocked, all atoms are within the SASA.
-\todo{I doubt this is the holo/complex, should be the apo streptavidin, right? -vah´}
+Here, the SASA was generated from a simulation of water around streptavidin without biotin, leaving the binding site solvent-accessible.
 

### 2023-06-26 manuscript/manuscript.tex
update according to comments by Dan and Mike

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\title{Quantifying Spatially Resolved Hydration Thermodynamics Using Grid Inhomogeneous Solvation Theory [Article v\versionnumber]}
+\title{Tutorial: Quantifying Spatially Resolved Hydration Thermodynamics Using Grid Inhomogeneous Solvation Theory [Article v\versionnumber]}
 
@@
 \author[1*]{Klaus R Liedl}
-\author[2*]{Michael K Gilson}
+\author[3*]{Michael K Gilson}
 \author[2*]{Tom Kurtzman}
+
 \affil[1]{Department of General, Inorganic and Theoretical Chemistry, University of Innsbruck, Austria}
-\affil[2]{Institution 2}
-
-\corr{email1@example.com}{FMS}  % Correspondence emails.  FMS and FS are the appropriate authors initials.
-\corr{email2@example.com}{FS}
+\affil[2]{Department of Chemistry, Lehman College, The City University of New York, Bronx, New York, USA}
+\affil[3]{Skaggs School of Pharmacy and Pharmaceutical Sciences, University of California, San Diego, USA}
+\affil[4]{Laboratory of Computational Biology, National Heart, Lung, and Blood Institute, National Institutes of Health, Bethesda, Maryland, USA}
+
+\corr{klaus.liedl@uibk.ac.at}{KRL}  % Correspondence emails.  FMS and FS are the appropriate authors initials.
+\corr{mgilson@health.ucsd.edu}{MKG}
+\corr{thomas.kurtzman@lehman.cuny.edu}{TK}
 
@@
 
-Solvation thermodynamics govern any process involving solutes in a solvent.
-%Cite "Theory of Simple Liquids"?
-%Solutes are constantly interacting with solvents in their environment, necessitating a good description of these interactions for accurate computational predictions. 
+Solvation thermodynamics govern many biological and chemical processes involving, e.g., aqueous solutions, hydrophobic effects, or partitioning between different environments.
+Interaction with water are of special importance as many biological processes occur in aqueous environment \cite{Privalov2017-water-review}.
 The free energy of solvation is defined as the reversible work needed to insert a solute into the solvent \cite{ben-naim-book}.
-Especially the interaction with water is of utmost importance, as many biological processes occur in aqueous environment \cite{Privalov2017-water-review}.
-It drives real-world effects such as hydrophobic interactions, binding of ligands to a biomolecule, as well as dynamics and folding of proteins.
-
-%The computational calculation of thermodynamical solvation properties is therefore of particular interest and has been tackled by a wide array of methods.
-%Quantum mechanical methods provide the most rigorous description of the underlying potential energy surface (PES) but suffer from high computational cost, allowing only the treatment of small systems and few solvent molecules.
-%More recently, QM/MM approaches and potentials based on deep learning \cite{Smith2017-ani} are trying to bridge the gap to computationally less demanding methods while keeping the accuracy of DFT methods.
-
+It is directly linked to the solubility of gases and drives many real-world effects such as hydrophobicity, ligand binding, and protein folding.
+
+Grid inhomogeneous solvation theory (GIST) is a method to compute the free energy of solvation, with applications in the context of drug design and biophysical property description.
+It estimates the solvent density from an MD trajectory and computes localized entropic and enthalpic contributions on a grid around the solute.
+The entropy is computed from the density distribution of the solvent around the solute, in a first-order approximation.
+A schematic of this is shown in Figure~\ref{fig:carbazole}.
+Nevertheless, solvation free energies of small molecules can be computed at high accuracy using linear scaling factors for the higher-order terms, such as $\Delta S = 0.6\, \Delta S^\text{1st-order}$ for water \cite{Chen2021,Waibl2022-gist-solvents}.
+
+\begin{figure}
+    \centering
+    \includegraphics[width=\linewidth]{manuscript/figures/carbazole-figure.png}
+    \caption{Schematic showing regions of high water density around a carbazole molecule.
+The energy is computed from the intermolecular interactions of a force field, while the entropy is approximated from the 3-dimensional solvent distribution around the solute.}
+    \label{fig:carbazole}
+\end{figure}
+
+Since its initial development, GIST has been used to describe the thermodynamics of water displacement in protein binding pockets \cite{Nguyen2016-gist-second-order,Ramsey2016,Balius2017-gist-ligand-discovery,Hufner-Wulsdorf2020-gist-drug-design}.
+It has also been deployed in systematically mapping the hydration properties of proteins found in SARS-CoV-2 \cite{Olson2020-covid-gist}.
+Furthermore, recent works used GIST to describe biophysical properties of serine proteases \cite{Kraml2019-gigist}, macrocycles \cite{Kamenik2020-gist-macrocycles}, or antibody fragements \cite{Waibl2021-gist-antibodies}.
+
+GIST has been implemented on the GPU, substantially improving its performance on large systems \cite{Kraml2019-gigist}.
+Furthermore, an implementation based on the particle mesh Ewald (PME) \cite{Darden1993-pme} method improves both the efficiency and the agreement with popular molecular dynamics engines \cite{Chen2021}.
+Yet further changes allow the inclusion of other rigid solvents than water \cite{Kraml2020,Waibl2022-gist-solvents} or the use of salt-water mixtures as solvents \cite{Waibl2021-gist-salt}.
+
+In this tutorial, we aim to give the reader an introduction into GIST and its application to calculate solution thermodynamics properties of interest. 
+While previous publications \cite{Ramsey2016} and tutorials\footnote{See \url{http://ambermd.org/tutorials/advanced/tutorial25/}} gave insightful pointers how to setup and run GIST calculations, the post-processing was often treated less rigorously.
+While building on these tutorials, we provide an updated guide to calculating solvation properties using GIST to address recent improvements.
+Furthermore, we aim to establish a template for the analysis and interpretation of GIST calculations.
+
+\subsection{Comparison to other methods}
 A wide range of methods have been devised to compute the free energy of hydration.
-Data-driven methods such as QSPR or UNIFAC can deliver good results \cite{Borhani2019-qspr,Fredenslund1975-unifac}, but might be limited to molecules that are similar to those in the training set.
-Polarizable continuum models (PCMs) \cite{Miertus1981-pcm} such as COSMO \cite{Klamt1993-cosmo} treat the solvent as a homogeneous phase that reacts to the electrostatic potential of the solute \cite{Mennucci2010-pcm}.
+Data-driven methods include QSPR and UNIFAC \cite{Borhani2019-qspr,Fredenslund1975-unifac}.
+Polarizable continuum models (PCMs) \cite{Miertus1981-pcm,Klamt1993-cosmo} treat the solvent as a homogeneous phase that reacts to the electrostatic potential of the solute \cite{Mennucci2010-pcm}.
 %Molecular mechanics based approaches describe the underlying PES through empirically fit force-fields, which reduce the computational demand of these calculations tremendously.
@@
 Methods such as MM/PBSA or MM/GBSA \cite{Sitkoff1994-pbsa,Kollman2000-mmpbsa} combine an implicit treatment of solvent electrostatics with structural sampling of the solute.
-They can quickly provide results for a large number of structures, but can yield inaccurate results when compared to more rigorous explicit solvent methods \cite{Genheden2015-mmpbsa-review}.
+They can quickly provide results for a large number of structures \cite{Genheden2015-mmpbsa-review}.\pdfcomment{Is there a rule of thumb when (not) to use implicit solvent methods?}
 
@@
 Methods derived from statistical mechanics bridge this gap by providing three-dimensional resolution.
-Two major flavors of methods in this group are those based on inhomogeneous fluid solvation theory (IST) \cite{Lazaridis1998} and those based on the Ornstein--Zernike (OZ) equation \cite{Hansen2013-simple-liquids}.
+Two major flavors of methods in this group are those based on the Ornstein--Zernike (OZ) equation \cite{Hansen2013-simple-liquids} and those based on inhomogeneous fluid solvation theory (IST) \cite{Lazaridis1998}.
 The most important example of OZ-based methods is the reference interaction site model (RISM) \cite{Chandler1972-rism} as well as its extension to three dimensions (3D-RISM) \cite{Kovalenko1998-3drism}.
 Those methods compute the atomic solvent distributions by self-consistently solving the OZ integral equation to predict solvation thermodynamics.
+
+On the other hand, methods based on IST generally use a molecular solvent distribution obtained from an MD simulation to compute the free energy of hydration from its enthalpic and entropic contributions.
+Examples include grid inhomogeneous solvation theory (GIST) \cite{Nguyen2012,Ramsey2016}, WaterMap \cite{Young2007-watermap,Abel2008-watermap}, SSTMap \cite{Haider2018-sstmap}, STOW \cite{Li2012-stow} and Solvaware \cite{Huggins2016-solvaware}.
+The main limitation of those methods is that they rely on an infinite expansion of the entropy in terms of the density distribution, which is truncated after the first or, sometimes, the second \cite{Nguyen2016-gist-second-order,Waibl2022-gist-solvents} term.
+Furthermore, IST does not consider the internal degrees of freedom of the solvent.
+Therefore, IST-based approaches can only be used with sufficiently rigid solvents.
+Nevertheless, they have been applied on a wide variety of problems involving both small molecules and biological systems.
+
+GIST is an implementation of IST on a three-dimensional grid.
+It can be seen as a more general version of WaterMap \cite{Abel2008-watermap,Young2007-watermap} that provides solvation thermodynamics on the entire grid, rather than high-density clusters in a binding pocket.
+This implies a higher versatility, but also puts more responsibility on the user in terms of correct post-processing.
+
+Since the density is computed from an MD simulation, GIST can accurately treat confined water regions, which are challenging for implicit-solvent methods such as GBSA and self-consistent theories such as 3D-RISM \cite{Kovalenko1998-3drism}.
+While the need for MD simulations comes at a large computational cost, GIST is still more efficient than end-state approaches such as TI \cite{Kirkwood1935-ti} or FEP \cite{Zwanzig1954-reweighting}, with applications to large systems such as serine proteases \cite{Kraml2019-gigist} or antibody fragments \cite{Waibl2021-gist-antibodies}.
+On the other hand, the first-order approximation means that the solvation entropy is less accurately described than in TI or FEP \cite{Chen2021,Waibl2021-gist-salt}.
 %The 3D-RISM approach allows for a decomposition into entropic and enthalpic terms as well as a full localization of the underlying distributions to the solute structure.
-
-On the other hand, methods based on IST generally use a molecular solvent distribution obtained from an MD simulation to compute the free energy of hydration from its enthalpic and entropic contributions.
-Examples of those methods include grid inhomogeneous solvation theory (GIST) \cite{Nguyen2012,Ramsey2016}, WaterMap \cite{Young2007-watermap,Abel2008-watermap}, SSTMap \cite{Haider2018-sstmap}, STOW \cite{Li2012-stow} and Solvaware \cite{Huggins2016-solvaware}.
-The main limitation of those methods is that they rely on an infinite expansion of the entropy in terms of the density distribution, which is truncated after the first or, sometimes, the second \cite{Nguyen2016-gist-second-order,Waibl2022-gist-solvents} term.
-Nevertheless, they have been applied on a wide variety of problems involving both small molecules and biological systems.
-
-In this tutorial, we aim to give the reader an introduction into GIST and its application to calculate solution thermodynamics properties of interest.
-
-GIST is based on IST, solving the integral equations on a grid. 
-This approach is more general than other IST-based methods because it accounts for all solvent regions rather than focusing on high-occupancy solvation sites.
-%By simulating the dynamics of a solvent around the solute, GIST calculates thermodynamic properties for a standard solvation process as described by Ben-Naim \cite{ben-naim-book}.
-It estimates the solvent density from an MD trajectory and allows for a full localization of entropic and enthalpic contributions on the grid around the solute.
-For small molecules, GIST has been shown to reach comparable accuracy to TI by approximating the higher-order entropies as a fraction of the first-order entropy, using a factor of \SI{-0.4} for water \cite{Chen2021,Waibl2022-gist-solvents}.
-%applying a linear correction factor (0.6 for water) to the first-order entropy as follows \cite{Chen2021,Waibl2022-gist-solvents}:
-%\begin{equation}
-%S^\textit{total} = S^\textit{1st-order} \times f
-%\end{equation}
-
-Recently, GIST was implemented on the GPU, tremendously improving the calculation speed, and allowing for the treatment of large biological systems such as proteases or antibodies \cite{Kraml2019-gigist}.
-Furthermore, an implementation based on the particle mesh Ewald (PME) \cite{Darden1993-pme} method was presented, which improves the consistency between GIST and popular molecular dynamics engines. 
-Yet further changes allow the inclusion of other rigid solvents than water \cite{Kraml2020,Waibl2022-gist-solvents} or 
-allow for the treatment of salt-water mixtures as solvents \cite{Waibl2021-gist-salt}.
-
-While previous publications \cite{Ramsey2016} and tutorials\footnote{See \url{http://ambermd.org/tutorials/advanced/tutorial25/}} gave insightful pointers how to setup and run GIST calculations, the post-processing was often treated less rigorously.
-While building on these tutorials, we provide an updated guide to calculating solvation properties using GIST to address recent improvements.
-Furthermore, we aim to establish a template for the analysis and interpretation of GIST calculations.
-
 %Here you would explain what problem you are tackling and briefly motivate your work.
@@
 \subsection{Scope}
-
-This tutorial provides an example of a GIST workflow that can be easily adapted to different systems. 
-In addition to a short introduction of the method and the theory behind it, we present an overview of the different implementations of GIST.
-We ease the user into basic GIST analysis at the example of biotin.
+The purpose of this tutorial is to help users run GIST calculations and interpret the results.
+The tutorial is structured around two examples: a GIST calculation for the small molecule biotin, and a study of biotin-streptavidin binding.
+We provide a GIST workflow that can be easily adapted to different systems, an introduction to the method and the theory behind it, and an overview of the different implementations of GIST.
+
+We start with a basic GIST analysis at the example of biotin.
 Building on the basics, we present a more advanced GIST study of the streptavidin-biotin complex, reflecting on many common aspects of GIST studies.
@@
 
-Furthermore, we discuss several technical aspects and typical pitfalls of GIST analyses, such as the normalization of voxel values and how to deal with double counting of energy contributions. 
+Furthermore, we discuss several technical aspects of GIST studies, such as the normalization of voxel values and how to avoid double counting of energy contributions. 
 Additionally, we provide a Python library (\software{gisttools}) to unify the analysis of GIST data produced by various versions of GIST and make the post-processing of GIST results more accessible.
@@
 The user should be able to run MD simulations and analyze the resulting trajectories.
-Previous experience of GIST is not necessary.
-However, we only provide a short summary of the theory behind GIST in section~\ref{sec:theory} at the end of this tutorial.
-For more detailed information, we point the reader towards the primary sources.
+Previous experience of GIST is not necessary, although we recommend reading the works referenced in section~\ref{sec:theory}.
 
@@
 
-A recent version of \software{cpptraj} should be used, such as the one provided in AmberTools22 or a recent installation from GitHub (\url{https://github.com/Amber-MD/cpptraj}).
-
-The closed-source Amber simulation engine is used in this tutorial to run the MD simulations.
-If Amber is not available, other MD engines can be used as long as topologies and trajectories can be produced in a format supported by \software{cpptraj}.
+A recent version of \software{cpptraj} should be used. We recommend using at least version 6.15.2, which can be obtained from GitHub (\url{https://github.com/Amber-MD/cpptraj}).
+
+The Amber simulation engine is used in this tutorial to run the MD simulations.
+While Amber is free for non-commercial use, it might not always be available in industry settings.
+In that case, other MD engines can be used as long as topologies and trajectories can be produced in a format supported by \software{cpptraj}.
 If using GROMACS \cite{Abraham2015-gromacs,Pronk2013-gromacs}, we propose to prepare the topology with AmberTools and convert it using \software{acpype.py} \cite{Sousa_da_Silva2012-acpype}.
 %We should actually test that!
-Also note that compressed trajectory formats such as \inlinecode{xtc} might bias the entropy calculation due to the loss of precision for the atom positions. \todo{This might be in the technical part}
-
-Furthermore, a recent Python version (>3.6) should be available with the following packages installed:
+Also note that compressed trajectory formats such as \inlinecode{xtc} might bias the entropy calculation due to the loss of precision for the atom positions.
+For further reference, the influence of compression on energy calculations was discussed in a recent publication \cite{Roe2022-compression}.
+
+Furthermore, a recent Python version (3.6 or newer) should be available with the following packages installed:
 \begin{itemize}
-	\item \software{gisttools} (\url{https://github.com/liedllab/gisttools})
-	\item \software{mdtraj (\url{https://www.mdtraj.org/}) \cite{McGibbon2015-mdtraj}}
+	\item \software{gisttools} (\url{https://github.com/liedllab/gisttools}, $\geq$ 0.2)
+	\item \software{mdtraj} (\url{https://www.mdtraj.org/}, $\geq$ 1.9.7) \cite{McGibbon2015-mdtraj}
 \end{itemize}
@@
 %\subsection{Running the GIST analysis}
-To run a GIST analyis, the following is required:
+To run a GIST analysis, the following is required:
 \begin{itemize}
@@
 The grid is defined by parameters of the \inlinecode{gist} command in \software{cpptraj}.
-The grid center is defined by \inlinecode{gridcntr}, the voxel number in x, y, and z direction by \inlinecode{griddim}, and the spacing (voxel side-length) in \AA by \inlinecode{gridspacn}.
+The grid center is defined by \inlinecode{gridcntr}, the voxel number in x, y, and z direction by \inlinecode{griddim}, and the spacing (voxel side-length) in \AA{} by \inlinecode{gridspacn}.
 
@@
 Often, the region of interest will be defined as the surrounding of a molecule, which might for example be a ligand in the binding pocket or the whole protein.
-A helper script called \inlinecode{FindCentroid.py} for this purpose is supplied with this tutorial, and can be used as follows:
+Using \inlinecode{cpptraj's } \inlinecode{bound} command, the region of interest can be output to a file.
+For example, here is the command for a region of 10\AA{} around the ligand LIG:
+\begin{lstlisting}[style=cpptraj]
+bounds :LIG dx 0.5 offset 20 name Grid out bounds.dat
+\end{lstlisting}
+Note that any loading and preprocessing, such as imaging, needs to be run before this command and needs to be the exact same as is used for the \inlinecode{gist} command.
+Additionally, a helper script called \inlinecode{FindCentroid.py} for the same purpose is supplied with this tutorial, and can be used as follows:
 
@@
 GIST is implemented in \software{cpptraj} which requires the topology and the trajectory of the MD simulation to be analyzed.
-The commands needed to run GIST in cpptraj can be written in an input file (e.g gist.in) for convenience.
-The basic commands, assuming a given grid center and a grid size of $30\times30\times30$ voxels with \SI{0.5}{\angstrom} spacing, are:
+The command needed to run GIST in cpptraj can be written in an input file (e.g gist.in) for convenience.
+The basic command, assuming a given grid center and a grid size of $30\times30\times30$ voxels with \SI{0.5}{\angstrom} spacing, is:
 
 \begin{lstlisting}[style=cpptraj]
-parm topology.prmtop
-trajin trajectory.nc
 gist gridcntr 10.5 20.1 30.1 griddim 30 30 30 \
 	gridspacn 0.50 out gist.dat
-go
-\end{lstlisting}
-
-The first line specifies the Amber topology file that was used to run MD. The second line species the trajectory file.
+\end{lstlisting}
+
 Note that the solute should not move significantly for a GIST calculation, since IST does not take the solute degrees of freedom into account.
 This is typically achieved using restraints on the solute heavy atoms.
-The third line tells \software{cpptraj} to run a GIST calculation using the specified region and print the output in a file called \inlinecode{gist.dat}.
-The last line instructs \software{cpptraj} to run the calculation.
-
-The syntax for running \software{cpptraj} given this input file is:
-\begin{lstlisting}[style=bash]
-cpptraj -i gist.in
-\end{lstlisting}
-
+This command tells \software{cpptraj} to run a GIST calculation using the specified region and print the output in a file called \inlinecode{gist.dat}.
+You will need to load and image your trajectory before running this command.
+An overview of how to work with \software{cpptraj} is provided in the program's documentation \cite{cpptraj_doc}, AMBER tutorials \cite{amber_tut_cpptraj} or on the AMBER-hub website \cite{amber_hub}.
+
+To use OpenMP or GPU-accelerated (using CUDA) versions of cpptraj, a different executable name such as \inlinecode{cpptraj.OMP} or \inlinecode{cpptraj.cuda} is often used.
+Therefore, make sure to use the right compilation and executable, to get full benefit of the various improvements to  GIST.
 The \inlinecode{gist} action has a large number of options to modify the GIST calculation or its output.
@@
 	\small
-	\begin{tabularx}{\columnwidth}{@{}llXl@{}}
+    \newcolumntype{L}[1]{>{\raggedright\arraybackslash}p{#1}}
+	\begin{tabularx}{\columnwidth}{@{}llL{3cm}l@{}}
 		\toprule
@@
 It is important to set the reference density $\rho_0$ according to the solvent model used in your simulation.
-The solvent-solvent enthalpy $E_{ww}$ also needs to be referenced, but this is usually done in the post-processing stage.
-
-Densities of typical water models can be found in the \software{cpptraj} manual.
-We also provide the densities for water models commonly used with AMBER forcefields in table \ref{tab:ref_densities}.
-These densities were calculated from \SI{1}{\micro\second}\pdfcomment{weird small mu symbol} MD simulations at \SI{1}{\bar} and \SI{300}{\kelvin} using the Berendsen barostat and the Langevin thermostat.
+The solvent-solvent enthalpy $E_{ww}$ also needs to be referenced, but this can be done in the post-processing stage.
+
+Densities and reference energies of typical water models can be found in the \software{cpptraj} manual.
+We also provide the densities for water models commonly used with AMBER force fields in table \ref{tab:ref_densities}.
+These densities were calculated from \SI{100}{\nano\second} MD simulations at \SI{1}{\bar} and \SI{300}{\kelvin} using the Berendsen barostat and the Langevin thermostat.
 The water boxes were prepared with \software{tleap} to a size of \SI{100}{\angstrom} x \SI{100}{\angstrom} x \SI{100}{\angstrom} before pressure equilibration.
-\todo{vah -> Finish these calculations}
+
+Due to the varying sizes of the boxes, the reference energies used in this context may not align with those specified in the Amber manual.
+Generally, it is recommended that the size of the reference water box closely matches that of the production system.
+Therefore, it is advisable to conduct your own reference calculations, especially for systems that are either very large or very small.
+While the impact on density is not as pronounced, deviations from the specified settings can lead to differences of up to 0.005 kcal/mol in the $E_{ww, norm}$ value for water boxes.
+These discrepancies, when summing over large grid regions, have the potential to introduce significant errors due to incorrect referencing.
+
 \begin{table}[h]
-	\caption{Reference density $\rho_0$ and mean water-water energy $E_{ww, norm}$ for various water models from NpT simulations at 300K/1bar.}\label{tab:ref_densities}
+	\caption{Reference density $\rho_0$ and mean water-water energy $E_{ww, norm}$ for various water models from NpT simulations of N$_{WAT}$ water molecules at \SI{300}{\kelvin} and 1bar.}\label{tab:ref_densities}
 	\small
-	\begin{tabularx}{\columnwidth}{@{}lrr@{}}
+	\begin{tabularx}{\columnwidth}{@{}lrrr@{}}
 		\toprule
-		Water Model       & $\rho_0$ /\AA$^{-3}$ & $E_{ww, norm}$ /$kcal\cdot mol^{-1} \cdot water^{-1}$  \\
+		Water Model       & $\rho_0$ & $E_{ww, norm}$ & $N_\text{WAT}$ \\
+        & $/\text{\AA}^{-3}$ & $/\text{kcal}\cdot \text{mol}^{-1}$ & \\
 		\midrule
-		TIP3P    & 0.0329 & -9.533   \\
-		TIP3PFB  & 0.0334 & -11.369  \\
-		TIP4P    & 0.0332 & -9.856   \\
-		TIP4P/Ew & 0.0332 & -11.036  \\
-		TIP5P    & 0.0329 & -9.596   \\
-		SPC/E    & 0.0333 & -11.123  \\
-		OPC      & XXXX   & XXXX     \\
-		OPC3     & XXXX   & XXXX     \\
-		POL3     & XXXX   & XXXX     \\
+		TIP3P \cite{Jorgensen1983-tip3p}    & 0.03287 & -9.5398   & 31795 \\
+		TIP3PFB \cite{Wang2014-tip3p-force-balance}  & 0.03321 & -11.7214  & 32502 \\
+		TIP4P \cite{Jorgensen1985-tip4p}    & 0.03316 & -9.8583   & 31218 \\
+       TIP4PFB \cite{Wang2014-tip3p-force-balance} & 0.03330 & -11.8724   & 31653 \\
+		TIP4P/Ew \cite{Horn2004-tip4pew} & 0.03323 & -11.0541  & 31218 \\
+		TIP5P \cite{Mahoney2000-tip5p}    & 0.03285 & -9.5989   & 31382 \\
+		SPC/E \cite{Berendsen1987-spce}    & 0.03333 & -11.1334  & 31795 \\
+       SPC/Eb \cite{Takemura2012-spceb}   & 0.03358 & -11.6966 & 31795 \\
+		OPC \cite{Izadi2014-opc}      & 0.03330 & -12.2456 & 31842 \\
+		OPC3 \cite{Izadi2016-opc3}     & 0.03323 & -11.6994 & 32251 \\
 		\bottomrule
@@
 \end{table}
-If your simulation settings diverge significantly from the above setting or for other solvent models not listed either here or in the \software{cpptraj} manual, the density should be computed from a pure solvent simulation beforehand.
 
@@
 Here, we aim to guide the reader through a GIST analysis of the streptavidin-biotin complex.
-We will
+The tutorial is structured as follows:
 \begin{enumerate}
-	\item set up and run GIST calculations for a small molecule (biotin) and a protein (streptavidin) binding pocket.
-	\item compute the free energy of solvation for biotin using GIST.
-	\item visualize local contributions to $\dgsolv$ in the streptavidin binding pocket and around biotin.
-	\item compute the hydration contribution to the binding. Combined with the biotin--streptavidin interaction energy, the binding free energy can be estimated.
-	\item explore advantages and disadvantages of the GIST method when used to calculate  $\dgsolv$
+	\item Set up and run GIST calculations for a small molecule (biotin) and a protein (streptavidin) binding pocket.
+	\item Compute the free energy of solvation for biotin using GIST.
+	\item Visualize local contributions to $\dgsolv$ in the streptavidin binding pocket and around biotin.
+	\item Compute the hydration contribution to the binding. Combined with the biotin--streptavidin interaction energy, the binding free energy can be estimated.
+	\item Explore advantages and disadvantages of the GIST method when used to calculate  $\dgsolv$
 \end{enumerate}
 
-In the first part of the tutorial, we will look at how to run GIST calculations for small molecules and interpret the output.
-Afterwards, we will apply this knowledge to a protein binding pocket, visualize $\dgsolv$ hotspots and estimate a free energy of ligand binding using an end-state approach based on GIST.
-In addition, we will discuss all steps in further detail at the end.
-\todo{TODO}
+In the first part, we show how to run GIST calculations for small molecules and interpret the output.
+Afterwards, we apply this knowledge to a protein binding pocket, visualize $\dgsolv$ hotspots and estimate a free energy of ligand binding using an end-state approach based on GIST.
+In addition, we discuss all steps in further detail at the end.
 
@@
 For water, \SI{0.5}{\nano\second} are usually sufficient to equilibrate the density, but narrow binding pockets can take longer.
+
+A well-tested equilibration protocol can be found in the literature \cite{Roe2020-equilibration}.
 
@@
 To start our GIST analysis of biotin, we'll first have to decide on the parameters defining the GIST grid during the analyis.
-We recommend that you center the biotin molecule in space:
+We recommend that you center the biotin molecule to the coordinate origin:
 \begin{lstlisting}[style=cpptraj]
@@
 To decide on a grid size, find the extent (minimum and maximum) of the x-, \mbox{y-,} and z-coordinates of the molecule.
-This can be done e.g., using the \software{mdtraj} package in Python.
-\begin{lstlisting}[style=python]
-import mdtraj as md
-pdb = md.load("biotin-centered.pdb")
-# mdtraj converts to nm.
-# *10 to convert back to Angstrom.
-min_xyz = pdb.xyz[0].min(0) * 10.
-max_xyz = pdb.xyz[0].max(0) * 10.
-print(*list(zip("xyz", min_xyz, max_xyz)), sep="\n")
-\end{lstlisting}
-For a system centered at the origin with a spacing of \SI{0.5}{\angstrom} and a buffer of, e.g., \SI{15}{\angstrom} between the solute and the grid wall, you need $4.0\times (d+15)$ bins in the respective direction, where $d$ is the larger of $-$min and $+$max in the respective direction.
 To get bulk-like water properties at the grid border, a buffer distance of \SI{15.0}{\angstrom} is reasonable for biotin.
+The grid parameters can be directly calculated using \software{Cpptraj's} \inlinecode{bound} command:
+\begin{lstlisting}[style=cpptraj]
+parm solvated.parm7
+trajin md-01.nc 1 1 1
+autoimage !(:WAT) origin
+bounds !(:WAT) dx 0.5 offset 30 name MyGrid out bounds.dat
+\end{lstlisting}
+Here, we calculate the bounds around the whole protein for a grid with \SI{0.5}{\angstrom} side-lenght voxels and a buffer zone of \SI{15.0}{\angstrom}.
+Note that the offset is given in terms of number of additional voxels, not voxel dimension.
+By varying the atom mask in the \inlinecode{bound} command, it is possible to center the grid on a region of interest, similiar to the \software{FindCentroid.py} script.
 
@@
 Smaller grids encompassing only the binding site could be used for streptavidin and the complex, but this would complicate the analysis.
-Furthermore, we recommend that you center the molecule in space.
+Furthermore, we recommend that you center the molecule to the coordinate origin.
 Calculate the GIST grid parameters as outlined in the biotin section for the complex and streptavidin and run the GIST analysis.
@@
 \begin{lstlisting}[style=python]
-centers = {biotin: biotin_x, 
-	   strept: strept_x, 
-	   compl: compl_x}
-names = {biotin: 'Biotin', 
-	   strept: 'Streptavidin', 
-	   compl: 'Complex'}
+centers = {"biotin": biotin_x, 
+	   "strept": strept_x, 
+	   "compl": compl_x}
+gist_objs = {"biotin": biotin, 
+             "strept": strept, 
+             "compl": compl}
 cols = ['Eall_dens','dTSsix_dens']
 rmax = 12
-results = { names[gist] : 
-	   gist.integrate_around(
-	   		cols, rmax, centers[gist]) 
-	   for gist in [biotin, strept, compl]}
-display( pd.DataFrame( results ))
+results = {
+    name: gist.integrate_around(
+        cols, rmax, centers[name])
+    for name, gist in gist_objs.items()}
+display(pd.DataFrame(results))
 
@@
 ax1.plot(bins, np.cumsum(biotin_rdf), label='biotin')
-ax1.plot(bins, np.cumsum(strept_rdf), label='streptavidin')
+ax1.plot(
+    bins, np.cumsum(strept_rdf), label='streptavidin')
 ax1.plot(bins, np.cumsum(compl_rdf), label='complex')
@@
 \end{equation}
-While this is supported by the standard GIST implementation, it is rarely done due to the large size of the $E_{ww,k\;l}$ matrix.
+While this is supported by the standard GIST implementation (using the \inlinecode{doeij} flag), it is rarely done due to the large size of the $E_{ww,k\;l}$ matrix.
 
@@
 However, when integrating over a small region, this process is less well-defined, since it is unclear to what extent the remaining solvent would reorganize.
-However, this depends on the local environment and on \emph{what} the solvent would be replaced by.
+This depends on the local environment and on \emph{what} the solvent would be replaced by.
 Therefore, the effect of reorganization can not be directly included in the GIST method, but must be judged for each case individually.

### 2023-06-26 manuscript/manuscript.tex
Fix figure 1 path. Remove most comments and todos

@@
 \section{Introduction}
-\todo{ (vah) Sources and better description of all these methods, more literature research, QSAR based methods? better flow }
-
-\todo{IMO, our article looks too dry. We should have something nice (e.g., an illustration of the GIST method) on the 1st or 2nd page -fwa.}
-\todo{Potentially relevant: 10.1021/acs.jcim.7b00520 }
-
 Solvation thermodynamics govern many biological and chemical processes involving, e.g., aqueous solutions, hydrophobic effects, or partitioning between different environments.
@@
     \centering
-    \includegraphics[width=\linewidth]{manuscript/figures/carbazole-figure.png}
+    \includegraphics[width=\linewidth]{figures/carbazole-figure.png}
     \caption{Schematic showing regions of high water density around a carbazole molecule.
@@
 Methods such as MM/PBSA or MM/GBSA \cite{Sitkoff1994-pbsa,Kollman2000-mmpbsa} combine an implicit treatment of solvent electrostatics with structural sampling of the solute.
-They can quickly provide results for a large number of structures \cite{Genheden2015-mmpbsa-review}.\pdfcomment{Is there a rule of thumb when (not) to use implicit solvent methods?}
+They can quickly provide results for a large number of structures \cite{Genheden2015-mmpbsa-review}.
 
@@
 (Adapt for biotin as necessary.)
-%We provide a script called \software{cpptraj\_remove\_mol.sh} that removes a single molecule from a topology and structure using \software{cpptraj} without altering the rest of the system.
-\todo{Explain cpptraj rather than our script.
-The script is for the automated workflow only.}
 
@@
 In \software{gisttools}, this is done simply by setting \inlinecode{gist.eww\_ref}.
-%In \software{gistpp}, this can be done as follows: 
-%\todo{Is there some official documentation for gistpp? We should probably link that somewhere, I feel that the code presented here is not really helpful/legible without further documentation. -vah}
-%\todo{This might become invalid with future GIST versions, depending on whether we adopt the COM as molecular center.} 
-%\todo{Better use the population.}
-
-%\begin{lstlisting}[style=bash]
-%rho0=0.03287465
-%neg_eww_ref=9.533
-%gistpp -op multconst -i gist-gO.dx -opt const $rho0 \
-%    -o gist-gO-abs.dx 
-%gistpp -op div -i gist-Eww-dens.dx -i2 gist-gO-abs.dx \
-%    -o gist-Eww-norm.dx
-%gistpp -op addconst -i gist-Eww-norm.dx \
-%    -opt const $neg_eww_ref -o gist-Eww-ref-norm.dx
-%gistpp -op mult -i gist-Eww-ref-norm.dx \
-%    -i2 gist-gO-abs.dx -o gist-Eww-ref-dens.dx
-%\end{lstlisting}
 
@@
 While this is not true for $E_{ww}$, we will see below that the energy difference upon binding converges more readily than its individual contributions.
-\todo{vah: Is that shown somewhere yet -> fwa?}
 
@@
 Subtract a reference entropy from the \inlinecode{dTSsix} columns.
-\todo{Update with newest gisttools -> fwa}
+
 \begin{lstlisting}[style=python]
@@
 The expected results are summarized in Table~\ref{tab_dg_monomers_dimer}.
-%\todo{Table 1 is placed quite akwardly at the moment for me, maybe enforce its position? -vah. Ok? -fwa}
 
@@
 To speed up the calculation, smaller GIST grids could be used by focusing on the binding pocket, or by rotating the molecule by its principal axes to fit the cuboid grid more exactly.
-However, this should be done \emph{before} the MD, since rotating the trajectory damages the periodic box information \todo{@Dan: is this still the case?}.
+However, this should be done \emph{before} the MD, since rotating the trajectory damages the periodic box information.
 
@@
 The expected output is shown in Figure~\ref{fig-streptavidin_gO}.
-\todo{vah -> include the PyMOL settings somewhere, this is not the default output}
-% The current font makes the O look like a tofu / placeholder
+
 \begin{lstlisting}[style=pymol]
@@
 For example, voxels that are part of the binding pocket can be defined using a ligand molecule and a distance criteria.
-%In \software{gistpp}, voxels that are within the distance criteria from any ligand heavy atom is considered as part of the region.
-The resulting file is a binary file and can be visualized using \software{pymol}.
-\todo{Not GISTPP}
+
 In section~\ref{sec:binding_contributions} we show how to integrate over regions such as the binding pocket.
@@
 Here, we show what this would look like for a radius of \SI{3.5}{\angstrom} around the heavy atoms of biotin.
-%\begin{lstlisting}[style=bash]
-%gistpp -op defbp -i gO.dx -i2 ligand.pdb 
-%
-%-opt const 3.5 -o gO_3.5.dx
-%\end{lstlisting}
+
 \begin{lstlisting}[style=python]

### 2023-08-16 manuscript/manuscript.tex
Address most of the comments raised by MKG in issue #45.
Change various text passages and adds some more citations.

Fix various issues with the code listings: code is now copyable,
correct single quotes are used and spaces are correctly displayed.

Restructure some text in Section 4.4.5, though this is still WIP

@@
 \usepackage{listings}
+\usepackage{upquote}
 \usepackage{tabularx}
@@
 	backgroundcolor = \color[rgb]{0.9, 0.9, 0.9},
+	showstringspaces=false,
+	columns=fullflexible,
 }
@@
 Solvation thermodynamics govern many biological and chemical processes involving, e.g., aqueous solutions, hydrophobic effects, or partitioning between different environments.
-Interaction with water are of special importance as many biological processes occur in aqueous environment \cite{Privalov2017-water-review}.
+Interactions with water are of special importance as many biological processes occur in aqueous environment \cite{Privalov2017-water-review}.
 The free energy of solvation is defined as the reversible work needed to insert a solute into the solvent \cite{ben-naim-book}.
@@
 
-Grid inhomogeneous solvation theory (GIST) is a method to compute the free energy of solvation, with applications in the context of drug design and biophysical property description.
+Grid inhomogeneous solvation theory (GIST) \cite{Nguyen2012} is a method to compute the free energy of solvation, with applications in the context of drug design and biophysical property description. 
 It estimates the solvent density from an MD trajectory and computes localized entropic and enthalpic contributions on a grid around the solute.
+Different from most other methods, these contributions are therefore spatially resolved, improving interpretability and allowing better insights into the nature of solvation around the investigated solute.
 The entropy is computed from the density distribution of the solvent around the solute, in a first-order approximation.
 A schematic of this is shown in Figure~\ref{fig:carbazole}.
-Nevertheless, solvation free energies of small molecules can be computed at high accuracy using linear scaling factors for the higher-order terms, such as $\Delta S = 0.6\, \Delta S^\text{1st-order}$ for water \cite{Chen2021,Waibl2022-gist-solvents}.
+The calculation of higher-order entropy terms is computationally demanding but was realized in recent works \cite{Nguyen2016-gist-second-order, Waibl2021-gist-salt}.
+Nevertheless, solvation free energies of small molecules can be computed at high accuracy using linear scaling factors as an approximation of the higher-order terms, such as $\Delta S = 0.6\, \Delta S^\text{1st-order}$ for water \cite{Chen2021,Waibl2022-gist-solvents}.
 
@@
 Since its initial development, GIST has been used to describe the thermodynamics of water displacement in protein binding pockets \cite{Nguyen2016-gist-second-order,Ramsey2016,Balius2017-gist-ligand-discovery,Hufner-Wulsdorf2020-gist-drug-design}.
-It has also been deployed in systematically mapping the hydration properties of proteins found in SARS-CoV-2 \cite{Olson2020-covid-gist}.
+As such, it has been deployed in systematically mapping the hydration properties of proteins found in SARS-CoV-2 \cite{Olson2020-covid-gist}.
 Furthermore, recent works used GIST to describe biophysical properties of serine proteases \cite{Kraml2019-gigist}, macrocycles \cite{Kamenik2020-gist-macrocycles}, or antibody fragements \cite{Waibl2021-gist-antibodies}.
@@
 
-On the other hand, methods based on IST generally use a molecular solvent distribution obtained from an MD simulation to compute the free energy of hydration from its enthalpic and entropic contributions.
+On the other hand, methods based on IST generally use a molecular solvent distribution obtained from an MD simulation to compute the free energy of hydration from its enthalpic and entropic contributions, providing easier interpretability of the resulting distributions.
 Examples include grid inhomogeneous solvation theory (GIST) \cite{Nguyen2012,Ramsey2016}, WaterMap \cite{Young2007-watermap,Abel2008-watermap}, SSTMap \cite{Haider2018-sstmap}, STOW \cite{Li2012-stow} and Solvaware \cite{Huggins2016-solvaware}.
-The main limitation of those methods is that they rely on an infinite expansion of the entropy in terms of the density distribution, which is truncated after the first or, sometimes, the second \cite{Nguyen2016-gist-second-order,Waibl2022-gist-solvents} term.
+The main limitation of those methods is their reliance on approximating the entropy through an infinite expansion in terms of the density distribution. 
+Due to the high computational effort required to converge the higher order terms, the entropy expansion is usually truncated after the first or, sometimes, the second \cite{Nguyen2016-gist-second-order,Waibl2022-gist-solvents} term, thereby introducing an error.
 Furthermore, IST does not consider the internal degrees of freedom of the solvent.
@@
 GIST is an implementation of IST on a three-dimensional grid.
-It can be seen as a more general version of WaterMap \cite{Abel2008-watermap,Young2007-watermap} that provides solvation thermodynamics on the entire grid, rather than high-density clusters in a binding pocket.
+Unlike the related method WaterMap \cite{Abel2008-watermap,Young2007-watermap}, it provides solvation thermodynamics on an entire grid, rather than only high-density clusters in a binding pocket.
 This implies a higher versatility, but also puts more responsibility on the user in terms of correct post-processing.
 
-Since the density is computed from an MD simulation, GIST can accurately treat confined water regions, which are challenging for implicit-solvent methods such as GBSA and self-consistent theories such as 3D-RISM \cite{Kovalenko1998-3drism}.
+Since the density is computed from an explicit solvent MD simulation, GIST provides an accurate picture of solvation at an atomistic level, even for demanding systems such as drug binding sites in proteins.
 While the need for MD simulations comes at a large computational cost, GIST is still more efficient than end-state approaches such as TI \cite{Kirkwood1935-ti} or FEP \cite{Zwanzig1954-reweighting}, with applications to large systems such as serine proteases \cite{Kraml2019-gigist} or antibody fragments \cite{Waibl2021-gist-antibodies}.
 On the other hand, the first-order approximation means that the solvation entropy is less accurately described than in TI or FEP \cite{Chen2021,Waibl2021-gist-salt}.
-%The 3D-RISM approach allows for a decomposition into entropic and enthalpic terms as well as a full localization of the underlying distributions to the solute structure.
-%Here you would explain what problem you are tackling and briefly motivate your work.
-%
-%In this particular template, we have removed most of the usage examples which occur in \texttt{sample-document.tex} to provide a minimal template you can modify; however, we retain a couple of examples illustrating more unusual features of our templates/article class, such as the checklists, and information on algorithms and pseudocode.
-%
-%Keep in mind, as you prepare your manuscript, that you should plan for a representative image  which will be used to highlight your article on the journal website and publications.
-Usually, this would be one of your figures, but it must also be uploaded separately upon article submission.
-We give specific guidelines for this image on the journal website in the section on article submission (see \url{https://livecomsjournal.github.io/authors/policies/index.html#article-submission}).
-%
-%Additionally, for well-formatted manuscripts, we recommend that you let LaTeX handle figure/table placement for you as much as possible, so please avoid specifying strenuous float instructions like `[h!]` and `[H]` as much as possible.
+Similar to the self-consistent integral theories, however, GIST only considers a single solute conformation and  multiple computations of representative structures are necessary to describe ensemble properties.
 
@@
 The purpose of this tutorial is to help users run GIST calculations and interpret the results.
-The tutorial is structured around two examples: a GIST calculation for the small molecule biotin, and a study of biotin-streptavidin binding.
-We provide a GIST workflow that can be easily adapted to different systems, an introduction to the method and the theory behind it, and an overview of the different implementations of GIST.
-
-We start with a basic GIST analysis at the example of biotin.
+The tutorial is structured around two examples: a GIST calculation for the small molecule biotin, and a GIST study of biotin-streptavidin binding.
+We provide a workflow that can be easily adapted to different systems, an introduction to the method and the theory behind it, and an overview of the different implementations of GIST.
+
+We start with a basic GIST analysis of biotin.
 Building on the basics, we present a more advanced GIST study of the streptavidin-biotin complex, reflecting on many common aspects of GIST studies.
@@
 The user should be able to run MD simulations and analyze the resulting trajectories. 
+The MD simulations presented in this tutorial are run with the Amber simulation engine and analyzed using AmberTools \software{cpptraj}. 
+For a conceptual overview of running MD simulations with Amber, we recommend reading the Amber manual (\url{https://ambermd.org/Manuals.php}) \cite{amber22} and the Amber tutorials (\url{https://ambermd.org/Manuals.php}).
 Previous experience of GIST is not necessary, although we recommend reading the works referenced in section~\ref{sec:theory}.
@@
 \software{Gistpp} can directly manipulate \software{OpenDX} files from the GIST output.
-Its usage and some examples were shown in a previous tutorial on GIST\cite{Ramsey2016}.
+Its usage and some examples were shown in a previous tutorial on GIST \cite{Ramsey2016}.
 
@@
 \end{itemize}
+Note that the solute should not move significantly for a GIST calculation, since IST does not take the solute degrees of freedom into account.
+This is typically achieved by applying restraints on the solute heavy atoms during the MD simulation.
 GIST analysis can be divided in three steps: defining a region of interest, running GIST using \software{cpptraj}, and post-processing and analysis.
@@
 In order to run a GIST analysis, one must define the region where solvent properties will be calculated.
-This region is represented by a rectangular grid with given position, size, and voxel spacing.
-The default value of \SI{0.5}{\angstrom} provides reasonable discretization for many use-cases.
-While integrals of GIST quantities over the grid are independent of the voxel size, individual values converge slower at high resolutions.
-
+This region is occupied by a rectangular grid with given position, size, and voxel spacing.
+The default spacing of \SI{0.5}{\angstrom} provides reasonable discretization for many use-cases.
+While integrals of GIST quantities over the grid are independent of the voxel size, provided sufficient sampling, individual values converge slower at high resolutions.
 The grid is defined by parameters of the \inlinecode{gist} command in \software{cpptraj}.
-The grid center is defined by \inlinecode{gridcntr}, the voxel number in x, y, and z direction by \inlinecode{griddim}, and the spacing (voxel side-length) in \AA{} by \inlinecode{gridspacn}.
+The grid center is defined by \inlinecode{gridcntr}, the voxel number in x, y, and z direction by \inlinecode{griddim}, and the spacing (voxel side-length) in \AA{} by \inlinecode{gridspacn}:
 
@@
 \end{lstlisting}
-
+Additionally, the GIST calculation will usually be preceded by preprocessing steps, such as centering the trajectory or reimaging to account for periodic boundary conditions.
 Often, the region of interest will be defined as the surrounding of a molecule, which might for example be a ligand in the binding pocket or the whole protein.
-Using \inlinecode{cpptraj's } \inlinecode{bound} command, the region of interest can be output to a file.
+Using \inlinecode{cpptraj's } \inlinecode{bounds} command, the region of interest can be output to a file.
 For example, here is the command for a region of 10\AA{} around the ligand LIG:
@@
 \end{lstlisting}
-Note that any loading and preprocessing, such as imaging, needs to be run before this command and needs to be the exact same as is used for the \inlinecode{gist} command.
+Note that the \inlinecode{bounds} command needs to be supplied with the exact same trajectory as used in the GIST calculation to generate correct boundaries.
+Therefore, any preprocessing steps need to match between the \inlinecode{bounds} command and the following \inlinecode{gist} calculation.
+
 Additionally, a helper script called \inlinecode{FindCentroid.py} for the same purpose is supplied with this tutorial, and can be used as follows:
@@
 The command needed to run GIST in cpptraj can be written in an input file (e.g gist.in) for convenience.
-The basic command, assuming a given grid center and a grid size of $30\times30\times30$ voxels with \SI{0.5}{\angstrom} spacing, is:
+The basic command, assuming a grid centered at (10.5 20.1 30.1) and a grid size of $30\times30\times30$ voxels with \SI{0.5}{\angstrom} spacing, is:
 
@@
 
-Note that the solute should not move significantly for a GIST calculation, since IST does not take the solute degrees of freedom into account.
-This is typically achieved using restraints on the solute heavy atoms.
 This command tells \software{cpptraj} to run a GIST calculation using the specified region and print the output in a file called \inlinecode{gist.dat}.
-You will need to load and image your trajectory before running this command.
-An overview of how to work with \software{cpptraj} is provided in the program's documentation \cite{cpptraj_doc}, AMBER tutorials \cite{amber_tut_cpptraj} or on the AMBER-hub website \cite{amber_hub}.
+An overview of how to work with \software{cpptraj} is provided in the program's documentation \cite{cpptraj_doc}, AMBER tutorials \cite{amber_tut_cpptraj} and on the AMBER-hub website \cite{amber_hub}.
 
@@
 \end{table}
-
-It is important to set the reference density $\rho_0$ according to the solvent model used in your simulation.
+In GIST, the calculated properties are derived by referencing against the bulk solvent.
+It is therefore important to set the reference number density $\rho_0$ according to the solvent model used in your simulation.
 The solvent-solvent enthalpy $E_{ww}$ also needs to be referenced, but this can be done in the post-processing stage.
+The calculation of these properties and the need for referencing is discussed in-depth in section /ref{sec:theory}.
 
@@
 \begin{table}[h]
-	\caption{Reference density $\rho_0$ and mean water-water energy $E_{ww, norm}$ for various water models from NpT simulations of N$_{WAT}$ water molecules at \SI{300}{\kelvin} and 1bar.}\label{tab:ref_densities}
+	\label{tab:ref_values}
+	\caption{Reference number density $\rho_0$ and mean water-water energy $E_{ww, norm}$ for various water models from NPT simulations of N$_{WAT}$ water molecules at \SI{300}{\kelvin} and 1bar.}\label{tab:ref_densities}
 	\small
@@
 	\item{[gX] For every element in the main solvent, the number density of atoms found in the voxel, in units of the bulk density.
-If the same element occurs multiple times, the bulk density is scaled accordingly.
+If the same element occurs multiple times, the bulk density is scaled such that that the expectation value for the bulk is unity.
 For water, gO and gH are produced.}
+
 	\item{[Esw] Mean solute-water interaction energy density.}
@@
 A full list of all properties in the output is provided at the end of the tutorial. 
-
-The output is organized by voxel and sorted according to their x, y, and z coordinates, describing the center point of the voxel.
+The output file is organized by voxel and sorted according to their x, y, and z coordinates, describing the center point of the voxel.
 Energy and entropy quantities are reported in two different ways.
@@
 The molecule should be solvated with a sufficiently large solvent box. 
+For periodic boundary conditions, the size of the solvent box should be large enough for the solvent to behave bulk-like so as to reduce artifact from solvent-mediated interactions between the periodic images.
+This corresponds to a distance between solute and box boundary of at least 12\AA{} for water, though we recommend 15\AA{} or higher.
 Carefully check the employed force field, protonation state and partial charges.
@@
 Solvate the parametrized biotin, making sure to choose a water box of large enough size so that the water on the box edges behaves bulk-like (a wall distance of at least 15\AA is recommended).
-A good starting point (though a bit dated) for the parametrization and solvation of small molecules is the Amber tutorial 4b (\url{http://ambermd.org/tutorials/basic/tutorial4b/}).
+A good starting point for the parametrization and solvation of small molecules is the Amber tutorial 4b (\url{http://ambermd.org/tutorials/basic/tutorial4b/}).
 
 \subsubsection{Equilibration}
-Equilibrate your solvated biotin in an NVT or NpT ensemble.
-Even when running the simulation in NVT, it is advisable to include an NpT equilibration step to adjust the box size beforehand.
+Equilibrate your solvated biotin in an NVT or NPT ensemble.
+Even when running the simulation in NVT, it is advisable to include an NPT equilibration step to adjust the box size beforehand.
 Use enough equilibration time to allow the solvent to relax. 
-For water, \SI{0.5}{\nano\second} are usually sufficient to equilibrate the density, but narrow binding pockets can take longer.
+For water, \SI{0.5}{\nano\second} is usually sufficient to equilibrate the density, but narrow binding pockets can take longer.
 
@@
 \subsubsection{MD simulation}
-
+\label{sec:md_simulation}
 Now, you can run a restrained MD simulation based on the equilibrated structure of biotin for the subsequent GIST analysis.
@@
 \begin{lstlisting}[style=amber-in]
-restrained 100 ns NpT
+restrained 100 ns NPT
 &cntrl
@@
 	ntr=1, restraint_wt=100.0,
-	restraintmask="!@H=&!:WAT",
+	restraintmask='!@H=&!:WAT',
 	ntp=1, pres0=1.0, taup=1.0,
@@
 
-This file will run a \SI{100}{\nano\second} NpT MD simulation while keeping all atoms except for hydrogens and waters restrained.
+This file will run a \SI{100}{\nano\second} NPT MD simulation while keeping all atoms except for hydrogens and waters restrained.
 Note that these are very standard MD settings except for the restraints.
@@
 \subsubsection{GIST analysis}
-To start our GIST analysis of biotin, we'll first have to decide on the parameters defining the GIST grid during the analyis.
-We recommend that you center the biotin molecule to the coordinate origin:
+To start our GIST analysis of biotin, we'll first have to decide on the parameters defining the GIST grid during the analysis.
+We will need a solute structure without water for the analysis of our future GIST results.
+As we ran our MD simulation with periodic boundaries, we always need to reimage all molecules to the original box using cpptrajs \inlinecode{autoimage} function.
+We also recommend that you center the biotin molecule to the coordinate origin, as this simplifies the definition of our grid origin later on.
+The following code snippets are
 \begin{lstlisting}[style=cpptraj]
@@
 trajout biotin-centered.pdb 
-\end{lstlisting}
+run
+\end{lstlisting}
+In this code snippet, the first two lines load the topology and coordinates of the first frame of our MD simulation.
+In the third line, we reimage the coordinates to the first box and center the biotin at the origin.
+Then, we strip all water molecules in the fourth line and write out a pdb file in the fith line.
+Finally, we start the analysis with the "run" command.
+
 To decide on a grid size, find the extent (minimum and maximum) of the x-, \mbox{y-,} and z-coordinates of the molecule.
@@
 bounds !(:WAT) dx 0.5 offset 30 name MyGrid out bounds.dat
-\end{lstlisting}
-Here, we calculate the bounds around the whole protein for a grid with \SI{0.5}{\angstrom} side-lenght voxels and a buffer zone of \SI{15.0}{\angstrom}.
-Note that the offset is given in terms of number of additional voxels, not voxel dimension.
-By varying the atom mask in the \inlinecode{bound} command, it is possible to center the grid on a region of interest, similiar to the \software{FindCentroid.py} script.
-
-Now, you can finally run the actual GIST analysis.
-Center the molecule as above, to make sure that its position matches the grid parameters you decided on.
+run
+\end{lstlisting}
+In the above snippet, we calculate the bounds around the whole protein for a grid with \SI{0.5}{\angstrom} side-length voxels and a buffer zone of \SI{15.0}{\angstrom}.
+Note that the offset is given in terms of number of additional voxels, not length dimension.
+By varying the atom mask in the \inlinecode{bound} command, it is possible to center the grid on a region of interest, similar to the \software{FindCentroid.py} script.
+
+Now, you can run the actual GIST analysis.
+Reimage and center the molecule as above, to make sure that its position matches the grid parameters you decided on.
 We recommend using either the PME or GPU implementation of GIST.
@@
 autoimage !(:WAT) origin
-# optionally remove the "pme" option
+# optionally remove the 'pme' option
 gist griddim 100 100 100 out gist.dat \
@@
 \subsubsection{Working with the GIST results}
-Running the GIST analysis above, an output data file (gist.dat) containing all calculated quantities for each voxel in the system is produced. 
+Running the GIST analysis above generates an output data file (gist.dat) containing all calculated quantities for each voxel in the system. 
 Additionally as outlined in section \ref{sec:running_GIST}, various quantities are written to OpenDX files.
@@
 gist = load_gist_file(
-"gist.dat", struct="solute-centered.pdb",  eww_ref=-9.539)
+'gist.dat', struct='solute-centered.pdb',  
+eww_ref=-9.539)
 print(f'n_frames = {gist.n_frames}')
@@
 Gisttools provides access to all quantities listed in the GIST output file.
-Additionally, it calculates the total enthalpy per voxel (Eall) as a sum of Eww and Esw, and the free energy of solvation (A) by subtracting dTSsix from Eall.
+Additionally, it calculates the total enthalpy per voxel (Eall) as a sum of Eww and Esw, and the free energy of solvation (A) by subtracting $\Delta$TSsix from Eall.
 
@@
 
-Gisttools also provides minimal OpenDX writing capabilities to visualize the calculated or modified GIST data.
+Gisttools also provides OpenDX writing capabilities to visualize the calculated or modified GIST data.
 For visualization, you'll most likely want to write out the density-normalized GIST quantities (X\_dens):
@@
 
-This tutorial discusses the visualization of OpenDX files in Section~\ref{sec:visualization}.
+Section~\ref{sec:visualization} discusses the visualization of OpenDX files.
+
 \subsection{Applying GIST for the Streptavidin-Biotin complex}
@@
 In our example, we only use a single equilibrated structure.
-A more sophisticated approach would be to identify several dominant structures in the ensemble -- either from experimental ensembles or by clustering an MD simulation.
+A more sophisticated approach would be to identify several dominant structures in the ensemble and run a separate MD simulation and GIST calculation for each structure.
+Such dominant structures could either be identified from experimental ensembles or by clustering an unrestrained MD simulation.
 
@@
 Note that, due to the restraints, some solvent sites might be inaccessible to the bulk solvent.
-To investigate such positions, make sure to explicitly place water molecules there.
+To investigate such positions, make sure to explicitly place water molecules there. %TODO add MC/MD or mupt and sources
 It is good practice to check your restrained MD simulation or solvent populations obtained by GIST to confirm whether all relevant sites were properly solvated.
 
-When calculating solvation free energy differences between a complex and the corresponding monomers (the dissociated state), there are two options for dealing with configurational diversity in the structures.
+When calculating solvation free energy differences between a complex and the corresponding monomers (the dissociated state), there are generally two approaches for dealing with configurational diversity.
+In the first approach, the ensemble properties of each individual part and the complex are accounted for by running individual MD simulations for each.
+The second approach assumes sufficient overlap in the conformational space of the dissociated states and the complex.
+Then, only configurations of the complex are sampled in a single MD simulation of the complex.
+Calculations for the disassociated states are then run on the extracted structures from the complex configurations.
+%TODO source MMPBSA
 One can choose to sample the complex and dissociated states individually, or to sample only one of them and assume that there is sufficient overlap in the conformations.
@@
 It is therefore important that the molecular conformations in the complex and dissociated states match exactly.
-Use \software{cpptraj} to produce biotin and streptavidin structures based on the complex by stripping the respective other molecule like this:
+Use the \software{equilibration.py} script (or your own procedure) to perform short NVT and NPT equilibration runs on the complex structure.
+Note that the equilibration script contains unrestrained NPT steps, because restraining two molecules at the same time might interfere with the pressure equilibration.
+Then, use \software{cpptraj} to produce biotin and streptavidin structures based on the complex by stripping the respective other molecule like this:
 \begin{lstlisting}[style=cpptraj]
@@
 \end{lstlisting}
-(Adapt for biotin as necessary.)
-
-Use the \software{equilibration.py} script (or your own procedure) to perform short NVT and NpT equilibration runs on the complex structure.
-Note that the equilibration script contains unrestrained NpT steps, because restraining two molecules at the same time might interfere with the pressure equilibration.
-
-Then split the equilibrated structure into biotin and streptavidin parts using \software{cpptraj\_remove\_mol.sh}.
+Adapt the above input file for biotin by striping molecule 1 instead of 2 and replacing the path names in the \inlinecode{strip} and  \inlinecode{trajout} commands.
+Alternatively, we provide a small script named \software{cpptraj\_remove\_mol.sh} for this procedure.
+
 After that, use \software{equilibration.py -R} to run short restrained 
@@
 
-\subsubsection{MD Simulation}
-The MD can be run as described above for each monomer and the complex.
-
 \subsubsection{GIST analysis of Streptavidin-Biotin}
+After running the MD simulations as described in section \ref{sec:md_simulation}, proceed with the GIST analysis.
 For easier post-processing, here we use grids that contain the whole molecule.
@@
 Calculate the GIST grid parameters as outlined in the biotin section for the complex and streptavidin and run the GIST analysis.
-For larger systems like whole proteins you can usually get away with a smaller buffer region compared to small molecules.
-To test whether your grid is large enough look at solvent properties close to the grid edges, e.g. the solvent-solvent energies.
+For larger systems like whole proteins you can usually get away with a smaller buffer region for the GIST grid compared to small molecules.
+To test whether your grid is large enough, look at solvent properties close to the grid edges, e.g. the solvent-solvent energies.
 If these correspond to the respective bulk properties, it can be assumed that your grid is large enough to capture all perturbations of the solute on the solvent.
 Note that a different box-size or number of solvent molecules can skew the bulk properties slightly compared to the reference values provided above.
-In that case, checking for convergence of the solvent properties near the grid borders is sufficient but you might want to modify your reference values according to your results then as described in the next section.
+In that case, checking for convergence of the solvent properties near the grid borders is sufficient.
+You might want to modify your reference values according to your results, as described in the next section.
 
@@
 \subsubsection{Reference Values and Radial Convergence}
-All GIST quantities should be expressed relative to bulk.
+All GIST quantities should be expressed relative to bulk as outlined in section \ref{sec:theory}.
 This is automatically done by \software{cpptraj} for the entropy, using the bulk density specified by \inlinecode{refdens}.
@@
 The Amber manual provides reference densities and energies for several solvents.
+Furthermore, a large variety of reference values calculated with GPU-GIST are listed in table \ref{tab:ref_values}.
 However, the exact energy values are different when using PME-GIST, and also depend on the box size.
@@
 Therefore, the reference energy can be obtained by binning the voxels by their distance to the solute, and evaluate Equation~\ref{eq_ewwref} within each bin.
-If this value converges, this can be used as $E_{ww}^\textit{ref}$.
+If this value converges with increasing distance from the solute, this can be used as $E_{ww}^\textit{ref}$.
 
@@
 Although it has a simple convergence check built-in, it is always recommended to check the convergence by hand.
+A handy way to check this is to calculate the radial distribution of this GIST property, i.e.
+a histogram in relation to the distance to the solute.
+Due to the conceptual similarity with classical radial distribution functions, describing the density of distances between two particle types, this function is exposed to the user as the \inlinecode{rdf} method.
 
@@
 gist = load_gist_file(
-    "gist.dat", struct="solute-centered.pdb")
+    'gist.dat', struct='solute-centered.pdb')
 bins, eww = gist.rdf(
-    "Eww_unref_norm", bins=100, rmax=20.,
-    normalize="norm")
+    'Eww_unref_norm', bins=100, rmax=20.,
+    normalize='norm')
 plt.plot(bins, eww)
@@
 The expected output of this code is shown in Figure~\ref{fig_ewwref}.
-
+%TODO redo figure caption
 \begin{figure}
@@
 In a plot like Figure~\ref{fig_ewwref} based on the \inlinecode{A\_dens} column, the values should tend to zero.
-However, it is more informative to plot the cumulative free energy contribution against the distance to the solute, to check the convergence.
+However, it is more informative to plot the cumulative free energy contribution against the distance to the solute, to check the convergence (see Figure~\ref{fig_radial_convergence}).
 It the curve flattens out, the converged value is your final $\dgsolv$\@.
-Otherwise, you might need to tweak the $E_{ww}$ reference, or introduce a reference value for the entropy.
+If the curve diverges even for large integration distances you might need to tweak the $E_{ww}$ reference, or introduce a reference value for the entropy.
 
@@
 import matplotlib.pyplot as plt
-# adapt eww_ref!
-gist = load_gist_file("gist.dat",
-    struct="solute-centered.pdb", eww_ref=-9.533)
+# adapt eww_ref according to the used solvent model!
+gist = load_gist_file('gist.dat',
+    struct='solute-centered.pdb', eww_ref=-9.533)
 bins, (dg, esw, eww, s) = gist.rdf(
-    ["A_dens", "Esw_dens", "Eww_dens", "dTSsix_dens"],
-    bins=100, rmax=20., normalize="none")
-plt.plot(bins, np.cumsum(dg), label="dG")
-plt.plot(bins, np.cumsum(eww), label="Eww")
-plt.plot(bins, np.cumsum(esw), label="Esw")
-plt.plot(bins, np.cumsum(s), label="dS")
+    ['A_dens', 'Esw_dens', 'Eww_dens', 'dTSsix_dens'],
+    bins=100, rmax=20., normalize='none')
+plt.plot(bins, np.cumsum(dg), label='dG')
+plt.plot(bins, np.cumsum(eww), label='Eww')
+plt.plot(bins, np.cumsum(esw), label='Esw')
+plt.plot(bins, np.cumsum(s), label='dS')
 \end{lstlisting}
@@
 %With \software{gistpp}, you can use the file from the convergence check. Additional OpenDX files can be produced using \inlinecode{gistpp -op makedx}.
-With \software{gisttools}, you can create one using e.g., \inlinecode{gist.save\_dx("A\_dens", "A\_dens.dx")}.
+With \software{gisttools}, you can create one using e.g., \inlinecode{gist.save\_dx('A\_dens', 'A\_dens.dx')}.
 
@@
 import gisttools as gt
-gist = gt.gist.load_gist_file("gist.dat",
-    struct="solute-centered.pdb", eww_ref=-9.533)
+gist = gt.gist.load_gist_file('gist.dat',
+    struct='solute-centered.pdb', eww_ref=-9.533)
 gist['E_norm'] = gist['Eww_norm'] * 2 + gist['Esw_norm']
@@
 
-Visualize the free energy, the energy, and the entropy, at several isolevels.
-
-Adapt the above PyMOL script to visualize the average energy per solvent at a positive and a negative isolevel, as well as regions of negative entropy per solvent.
-The expected representation of the streptavidin binding pocket is shown in Figure~\ref{fig_binding_pocket_pymol}.
+The OpenDX files can then be imported into your molecular viewer of choice.
+%TODO we are missing a pymol script here
+%Adapt the above PyMOL script to visualize the average energy per solvent at a positive and a negative isolevel, as well as regions of negative entropy per solvent.
+A visualization of entropy and energy contributions to the free energy of solvation of the streptavidin binding pocket is shown in Figure~\ref{fig_binding_pocket_pymol}.
+PyMOL input scripts to generate such visualizations are available in the GitHub repository for this tutorial.
 
@@
 
-You should find that there are several regions in the binding pocket that feature a high solvent energy and/or low entropy.
-However, such regions also exist outside of the binding pocket, and there are also low-energy regions within the binding pocket.
-Therefore, the strong affinity to biotin cannot be explained only by hydration, and the direct interaction energy also needs to be taken into account.
+%You should find that there are several regions in the binding pocket that feature a high solvent energy and/or low entropy.
+%However, such regions also exist outside of the binding pocket, and there are also low-energy regions within the binding pocket.
+%Therefore, the strong affinity to biotin cannot be explained only by hydration, and the direct interaction energy also needs to be taken into account.
 
@@
 \end{lstlisting}
+
 Assign reference energies and check them for plausibility.
+
 \begin{lstlisting}[style=python]
 biotin.eww_ref = biotin.detect_reference_value()
-print("Biotin:", biotin.eww_ref)
+print('Biotin:', biotin.eww_ref)
 strept.eww_ref = strept.detect_reference_value()
-print("Streptavidin:", strept.eww_ref)
+print('Streptavidin:', strept.eww_ref)
 compl.eww_ref = compl.detect_reference_value()
-print("Complex:", compl.eww_ref)
+print('Complex:', compl.eww_ref)
 \end{lstlisting}
@@
 Note that we use \software{mdtraj} here, since \software{gisttools} stores the reference structure as an \software{mdtraj} \inlinecode{Trajectory} object.
+Here, we compute the histogram of the contributions of single properties to assess the convergence.
+
 \begin{lstlisting}[style=python]
@@
 def select(traj, sel):
-    "Slice a Trajectory by selection mask."
+    'Slice a Trajectory by selection mask.'
     return traj.atom_slice(traj.top.select(sel))
@@
 
-Now, subtract the monomer rdfs from the complex, and compute the integrals.
-If you also visualize the individual rdfs, you will notice that the difference converges much better with increasing radius than the contributions.
+Now, subtract the monomer histograms from the complex, and compute the sum of your property within some cutoff distance to the solute.
+If you also visualize the individual histograms, you will notice that the difference converges much better with increasing radius than the single contributions.
 This is because the cutoff contains regions that are close to the atom, since the integration region does not comprise the whole protein.
+%TODO this sentence doesn't make sense
 
@@
 integral = difference[bins < cutoff].sum()
-print("Integral = {}".format(integral))
+print('Integral = {}'.format(integral))
 plt.plot(bins, np.cumsum(difference))
@@
 
-Compute the energy (\inlinecode{Eall}) and entropy (\inlinecode{dTSsix}) contributions separately.
-You will find that the energy disfavors binding, because we do not yet include the interaction energy between biotin and streptavidin.
-To make this easier, gisttool includes a function to integrate in a radius around atom centers:
-\begin{lstlisting}[style=python]
-centers = {"biotin": biotin_x, 
-	   "strept": strept_x, 
-	   "compl": compl_x}
-gist_objs = {"biotin": biotin, 
-             "strept": strept, 
-             "compl": compl}
-cols = ['Eall_dens','dTSsix_dens']
-rmax = 12
-results = {
-    name: gist.integrate_around(
-        cols, rmax, centers[name])
-    for name, gist in gist_objs.items()}
-display(pd.DataFrame(results))
-
-\end{lstlisting}
 You can also plot the difference between the complex and monomer contributions against the distance to biotin, using the complex coordinates for the holo structure:
@@
 \end{lstlisting}
-You can also plot the difference between the complex and monomer contributions against the distance to biotin, using the complex coordinates for the holo structure.
+
 The expected result is shown in Figure~\ref{fig-dg-sums}.
@@
 
+Compute the energy (\inlinecode{Eall}) and entropy (\inlinecode{dTSsix}) contributions separately.
+ To make this easier, gisttool includes a function to integrate in a radius around atom centers:
+\begin{lstlisting}[style=python]
+	centers = {'biotin': biotin_x, 
+		'strept': strept_x, 
+		'compl': compl_x}
+	gist_objs = {'biotin': biotin, 
+		'strept': strept, 
+		'compl': compl}
+	cols = ['Eall_dens','dTSsix_dens']
+	rmax = 12
+	results = {
+		name: gist.integrate_around(
+		cols, rmax, centers[name])
+		for name, gist in gist_objs.items()}
+	display(pd.DataFrame(results))
+
+\end{lstlisting}
+You will find that the energy disfavors binding, because we do not yet include the interaction energy between biotin and streptavidin.
 Using the \inlinecode{energy} command in \software{cpptraj}, you can compute this interaction energy.
-We recommend using PME in combination with PME-GIST, but not with GPU-GIST\@.
+We recommend using PME in combination with PME-GIST, but not with GPU-GIST to be in line with the different ways the energy is calculated in the two GIST methods.
 An example \software{cpptraj} input might look like:
@@
 Even though streptavidin-biotin is known to be a very stable complex, the solvation energy favors the dissociation.
-This is expected: if the molecules were not able to form any interactions, but interacted fully with the water, they would not form a complex.
+This is expected: the binding of biotin to streptavidin is faciliated over a large number of H-bonds and over polar interactions.
+As such, water is favourably solvates both biotin and streptavidin and an energetic penalty is paid when replacing it during the binding process.
 
@@
 def select(traj, sel):
-	"Slice a Trajectory by selection mask."
+	'Slice a Trajectory by selection mask.'
 	return traj.atom_slice(traj.top.select(sel))
@@
 
-For the entropy, this leads to an infinite expansion of correlations of increasing order.
-In a slightly simplified view, the first order is the solvent density $g_\textit{sw}\left(\textbf{r},\omega \right)$ at each position $\textbf{r}$ and orientation $\omega$.
+The entropy can be approximated as an infinite expansion of correlations of increasing order.
+In a slightly simplified view, the first order is the log solvent density $ln\left(g_\textit{sw}\left(\textbf{r},\omega \right)\right)$ at each position $\textbf{r}$ and orientation $\omega$.
 The first-order entropy omits all solvent-solvent (and higher) correlations, while the solute-solvent correlation is taken into account because the coordinates are relative to the solute.

### 2023-08-16 code/scripts/cpptraj_remove_mol.sh

@@ -1,4 +1,4 @@
-#!/bin/bash
+#!/usr/bin/env bash
 
 if [ -z "$5" ]; then
 	echo Usage: $0 IN_PARM IN_RST N_TO_REMOVE OUT_PARM OUT_RST

### 2023-08-29 code/scripts/equilibration.py
Made version agnostic

@@ -1,4 +1,4 @@
-#!/usr/bin/env python2
+#!/usr/bin/env python
 # import sys
 from __future__ import print_function
 import os

### 2023-08-29 manuscript/manuscript.tex
Fixed English error
Add numpy import to second code block in 4.4.3

@@
 import matplotlib.pyplot as plt
+import numpy as np
 # adapt eww_ref according to the used solvent model!
@@
 This is expected: the binding of biotin to streptavidin is faciliated over a large number of H-bonds and over polar interactions.
-As such, water is favourably solvates both biotin and streptavidin and an energetic penalty is paid when replacing it during the binding process.
+As such, water favourably solvates both biotin and streptavidin and an energetic penalty is paid when displacing it during the binding process.
 

### 2023-08-29 manuscript/manuscript.tex
Reordered plt.plot() calls to match figure 3

@@
     bins=100, rmax=20., normalize='none')
-plt.plot(bins, np.cumsum(dg), label='dG')
 plt.plot(bins, np.cumsum(eww), label='Eww')
 plt.plot(bins, np.cumsum(esw), label='Esw')
+plt.plot(bins, np.cumsum(dg), label='dG')
 plt.plot(bins, np.cumsum(s), label='dS')

### 2023-08-29 manuscript/manuscript.tex
Added legend, xlabel, and ylabel calls to match figure 3

@@
 plt.plot(bins, np.cumsum(s), label='dS')
+plt.legend()
+plt.xlabel("distance cutoff (Angstrom)")
+plt.ylabel("free energy contribution [kcal/mol]")
 \end{lstlisting}

### 2023-08-29 manuscript/manuscript.tex
Changed xlabel unit to match style of other figures

--- 2023-08-29 manuscript/manuscript.tex
@@
 plt.plot(bins, np.cumsum(s), label='dS')
+plt.legend()
+plt.xlabel("distance cutoff [Å]")
+plt.ylabel("free energy contribution [kcal/mol]")
 \end{lstlisting}

### 2023-08-29 manuscript/manuscript.tex
Formatted some codeblock comments to use python format

@@
 def select(traj, sel):
-    'Slice a Trajectory by selection mask.'
+    # Slice a Trajectory by selection mask.
     return traj.atom_slice(traj.top.select(sel))
@@
 def select(traj, sel):
-	'Slice a Trajectory by selection mask.'
+	# Slice a Trajectory by selection mask.
 	return traj.atom_slice(traj.top.select(sel))

### 2023-08-29 manuscript/manuscript.tex
Added masking strings to codeblock from Jupyter notebook
Added mdtraj import to codeblock

@@
 \begin{lstlisting}[style=python]
+import mdtraj as md
 col = 'dTSsix_dens'
@@
 # we multiply by 10 to convert nm to Angstrom.
+biotin_mask = 'resname BTN and not element H'
+strept_mask = 'not resname BTN and not resname WAT and not element H'
 compl_x = select(compl.struct, biotin_mask).xyz[0] * 10.

### 2023-08-29 manuscript/manuscript.tex
Removed erroneous indents from codeblock
Added pandas import to codeblock

@@
 \begin{lstlisting}[style=python]
+import pandas as pd
 centers = {'biotin': biotin_x, 

### 2023-08-29 manuscript/manuscript.tex
Changed display() to print() in codeblock

--- 2023-08-29 manuscript/manuscript.tex
@@
 	for name, gist in gist_objs.items()}
-	display(pd.DataFrame(results))
+print(pd.DataFrame(results))
 

### 2023-08-29 manuscript/manuscript.tex
Added gisttools import to codeblock
Added matplotlib to list of required python packages

@@
 	\item \software{mdtraj} (\url{https://www.mdtraj.org/}, $\geq$ 1.9.7) \cite{McGibbon2015-mdtraj}
+	\item \software{matplotlib} (\url{https://matplotlib.org/}, $\geq$ 3.7.0) \cite{Hunter2007-matplotlib}
 \end{itemize}
@@
 \begin{lstlisting}[style=python]
+import gisttools as gt
 def select(traj, sel):

### 2023-08-29 manuscript/manuscript.tex
Added numpy to list of required python packages
Added pandas to list of required python packages
Added apostrophe to signify possession in cpptrajs

@@
 	\item \software{matplotlib} (\url{https://matplotlib.org/}, $\geq$ 3.7.0) \cite{Hunter2007-matplotlib}
+	\item \software{numpy} (\url{https://numpy.org/}, $\geq$ 1.23.5) \cite{harris2020-numpy}
+	\item \software{pandas} (\url{https://pandas.pydata.org/}, $\geq$ 1.5.3) \cite{pandas-2023_8239932}
 \end{itemize}
@@
 We will need a solute structure without water for the analysis of our future GIST results.
-As we ran our MD simulation with periodic boundaries, we always need to reimage all molecules to the original box using cpptrajs \inlinecode{autoimage} function.
+As we ran our MD simulation with periodic boundaries, we always need to reimage all molecules to the original box using cpptraj's \inlinecode{autoimage} function.
 We also recommend that you center the biotin molecule to the coordinate origin, as this simplifies the definition of our grid origin later on.

### 2024-06-04 manuscript/manuscript.tex
Update manuscript.tex

--- 2023-08-29 manuscript/manuscript.tex
@@
 \usepackage{listings}
-\usepackage{upquote}
 \usepackage{tabularx}
@@
 %\author[2]{Firstname Initials Surname}
+\author[2]{Steven Ramsey}
 \author[4]{Daniel R Roe}
@@
 	backgroundcolor = \color[rgb]{0.9, 0.9, 0.9},
-	showstringspaces=false,
-	columns=fullflexible,
 }
@@
 
-In this tutorial, we aim to guide the reader through the most common steps involved in a GIST analysis at the example of the streptavidin-biotin complex.
+In this tutorial, we aim to guide the reader through the most common steps involved in a GIST analysis using the streptavidin-biotin complex as an demonstrative system.
 Furthermore, we discuss the theory of GIST with a focus on practical aspects.
@@
 Solvation thermodynamics govern many biological and chemical processes involving, e.g., aqueous solutions, hydrophobic effects, or partitioning between different environments.
-Interactions with water are of special importance as many biological processes occur in aqueous environment \cite{Privalov2017-water-review}.
+Interaction with water are of special importance as many biological processes occur in aqueous environment \cite{Privalov2017-water-review}.
 The free energy of solvation is defined as the reversible work needed to insert a solute into the solvent \cite{ben-naim-book}.
@@
 
-Grid inhomogeneous solvation theory (GIST) \cite{Nguyen2012} is a method to compute the free energy of solvation, with applications in the context of drug design and biophysical property description. 
+Grid inhomogeneous solvation theory (GIST) is a method to compute the free energy of solvation, with applications in the context of drug design and biophysical property description.
 It estimates the solvent density from an MD trajectory and computes localized entropic and enthalpic contributions on a grid around the solute.
-Different from most other methods, these contributions are therefore spatially resolved, improving interpretability and allowing better insights into the nature of solvation around the investigated solute.
 The entropy is computed from the density distribution of the solvent around the solute, in a first-order approximation.
 A schematic of this is shown in Figure~\ref{fig:carbazole}.
-The calculation of higher-order entropy terms is computationally demanding but was realized in recent works \cite{Nguyen2016-gist-second-order, Waibl2021-gist-salt}.
-Nevertheless, solvation free energies of small molecules can be computed at high accuracy using linear scaling factors as an approximation of the higher-order terms, such as $\Delta S = 0.6\, \Delta S^\text{1st-order}$ for water \cite{Chen2021,Waibl2022-gist-solvents}.
+Nevertheless, solvation free energies of small molecules can be computed at high accuracy using linear scaling factors for the higher-order terms, such as $\Delta S = 0.6\, \Delta S^\text{1st-order}$ for water \cite{Chen2021,Waibl2022-gist-solvents}.
 
@@
 Since its initial development, GIST has been used to describe the thermodynamics of water displacement in protein binding pockets \cite{Nguyen2016-gist-second-order,Ramsey2016,Balius2017-gist-ligand-discovery,Hufner-Wulsdorf2020-gist-drug-design}.
-As such, it has been deployed in systematically mapping the hydration properties of proteins found in SARS-CoV-2 \cite{Olson2020-covid-gist}.
+It has also been deployed in systematically mapping the hydration properties of proteins found in SARS-CoV-2 \cite{Olson2020-covid-gist}.
 Furthermore, recent works used GIST to describe biophysical properties of serine proteases \cite{Kraml2019-gigist}, macrocycles \cite{Kamenik2020-gist-macrocycles}, or antibody fragements \cite{Waibl2021-gist-antibodies}.
@@
 
-On the other hand, methods based on IST generally use a molecular solvent distribution obtained from an MD simulation to compute the free energy of hydration from its enthalpic and entropic contributions, providing easier interpretability of the resulting distributions.
+On the other hand, methods based on IST generally use a molecular solvent distribution obtained from an MD simulation to compute the free energy of hydration from its enthalpic and entropic contributions.
 Examples include grid inhomogeneous solvation theory (GIST) \cite{Nguyen2012,Ramsey2016}, WaterMap \cite{Young2007-watermap,Abel2008-watermap}, SSTMap \cite{Haider2018-sstmap}, STOW \cite{Li2012-stow} and Solvaware \cite{Huggins2016-solvaware}.
-The main limitation of those methods is their reliance on approximating the entropy through an infinite expansion in terms of the density distribution. 
-Due to the high computational effort required to converge the higher order terms, the entropy expansion is usually truncated after the first or, sometimes, the second \cite{Nguyen2016-gist-second-order,Waibl2022-gist-solvents} term, thereby introducing an error.
+The main limitation of those methods is that they rely on an infinite expansion of the entropy in terms of the density distribution, which is truncated after the first or, sometimes, the second \cite{Nguyen2016-gist-second-order,Waibl2022-gist-solvents} term.
 Furthermore, IST does not consider the internal degrees of freedom of the solvent.
@@
 GIST is an implementation of IST on a three-dimensional grid.
-Unlike the related method WaterMap \cite{Abel2008-watermap,Young2007-watermap}, it provides solvation thermodynamics on an entire grid, rather than only high-density clusters in a binding pocket.
+It can be seen as a more general version of WaterMap \cite{Abel2008-watermap,Young2007-watermap} that provides solvation thermodynamics on a space filling grid, rather than high-density clusters in a binding pocket.
 This implies a higher versatility, but also puts more responsibility on the user in terms of correct post-processing.
 
-Since the density is computed from an explicit solvent MD simulation, GIST provides an accurate picture of solvation at an atomistic level, even for demanding systems such as drug binding sites in proteins.
+Since the density is computed from an MD simulation, GIST can accurately treat confined water regions, which are challenging for implicit-solvent methods such as GBSA and self-consistent theories such as 3D-RISM \cite{Kovalenko1998-3drism}.
 While the need for MD simulations comes at a large computational cost, GIST is still more efficient than end-state approaches such as TI \cite{Kirkwood1935-ti} or FEP \cite{Zwanzig1954-reweighting}, with applications to large systems such as serine proteases \cite{Kraml2019-gigist} or antibody fragments \cite{Waibl2021-gist-antibodies}.
 On the other hand, the first-order approximation means that the solvation entropy is less accurately described than in TI or FEP \cite{Chen2021,Waibl2021-gist-salt}.
-Similar to the self-consistent integral theories, however, GIST only considers a single solute conformation and  multiple computations of representative structures are necessary to describe ensemble properties.
+%The 3D-RISM approach allows for a decomposition into entropic and enthalpic terms as well as a full localization of the underlying distributions to the solute structure.
+%Here you would explain what problem you are tackling and briefly motivate your work.
+%
+%In this particular template, we have removed most of the usage examples which occur in \texttt{sample-document.tex} to provide a minimal template you can modify; however, we retain a couple of examples illustrating more unusual features of our templates/article class, such as the checklists, and information on algorithms and pseudocode.
+%
+%Keep in mind, as you prepare your manuscript, that you should plan for a representative image  which will be used to highlight your article on the journal website and publications.
+Usually, this would be one of your figures, but it must also be uploaded separately upon article submission.
+We give specific guidelines for this image on the journal website in the section on article submission (see \url{https://livecomsjournal.github.io/authors/policies/index.html#article-submission}).
+%
+%Additionally, for well-formatted manuscripts, we recommend that you let LaTeX handle figure/table placement for you as much as possible, so please avoid specifying strenuous float instructions like `[h!]` and `[H]` as much as possible.
 
@@
 The purpose of this tutorial is to help users run GIST calculations and interpret the results.
-The tutorial is structured around two examples: a GIST calculation for the small molecule biotin, and a GIST study of biotin-streptavidin binding.
-We provide a workflow that can be easily adapted to different systems, an introduction to the method and the theory behind it, and an overview of the different implementations of GIST.
-
-We start with a basic GIST analysis of biotin.
+The tutorial is structured around two examples: a GIST calculation for the small molecule biotin, and a study of biotin-streptavidin binding.
+We provide a GIST workflow that can be easily adapted to different systems, an introduction to the method and the theory behind it, and an overview of the different implementations of GIST.
+
+We start with a basic GIST analysis at the example of biotin.
 Building on the basics, we present a more advanced GIST study of the streptavidin-biotin complex, reflecting on many common aspects of GIST studies.
@@
 The user should be able to run MD simulations and analyze the resulting trajectories.
-The MD simulations presented in this tutorial are run with the Amber simulation engine and analyzed using AmberTools \software{cpptraj}. 
-For a conceptual overview of running MD simulations with Amber, we recommend reading the Amber manual (\url{https://ambermd.org/Manuals.php}) \cite{amber22} and the Amber tutorials (\url{https://ambermd.org/Manuals.php}).
-Previous experience of GIST is not necessary, although we recommend reading the works referenced in section~\ref{sec:theory}.
+Previous experience of GIST is not necessary, although we recommend reading the articles referenced in section~\ref{sec:theory}.
 
@@
 The Amber simulation engine is used in this tutorial to run the MD simulations.
-While Amber is free for non-commercial use, it might not always be available in industry settings.
+While Amber is free for non-commercial use, it might not always be available in industrial settings.
 In that case, other MD engines can be used as long as topologies and trajectories can be produced in a format supported by \software{cpptraj}.
@@
 \software{Gistpp} can directly manipulate \software{OpenDX} files from the GIST output.
-Its usage and some examples were shown in a previous tutorial on GIST \cite{Ramsey2016}.
+Its usage and some examples were shown in a previous tutorial on GIST\cite{Ramsey2016}.
 
@@
 \end{itemize}
-Note that the solute should not move significantly for a GIST calculation, since IST does not take the solute degrees of freedom into account.
-This is typically achieved by applying restraints on the solute heavy atoms during the MD simulation.
 GIST analysis can be divided in three steps: defining a region of interest, running GIST using \software{cpptraj}, and post-processing and analysis.
@@
 In order to run a GIST analysis, one must define the region where solvent properties will be calculated.
-This region is occupied by a rectangular grid with given position, size, and voxel spacing.
-The default spacing of \SI{0.5}{\angstrom} provides reasonable discretization for many use-cases.
-While integrals of GIST quantities over the grid are independent of the voxel size, provided sufficient sampling, individual values converge slower at high resolutions.
+This region is represented by a rectangular grid with given position, size, and voxel spacing.
+The default value of \SI{0.5}{\angstrom} provides reasonable discretization for many use-cases.
+While integrals of GIST quantities over the grid are independent of the voxel size, values for each voxel converge slower at high resolutions.
+
 The grid is defined by parameters of the \inlinecode{gist} command in \software{cpptraj}.
-The grid center is defined by \inlinecode{gridcntr}, the voxel number in x, y, and z direction by \inlinecode{griddim}, and the spacing (voxel side-length) in \AA{} by \inlinecode{gridspacn}:
+The grid center is defined by \inlinecode{gridcntr}, the voxel number in x, y, and z direction by \inlinecode{griddim}, and the spacing (voxel side-length) in \AA{} by \inlinecode{gridspacn}.
 
@@
 \end{lstlisting}
-Additionally, the GIST calculation will usually be preceded by preprocessing steps, such as centering the trajectory or reimaging to account for periodic boundary conditions.
+
 Often, the region of interest will be defined as the surrounding of a molecule, which might for example be a ligand in the binding pocket or the whole protein.
-Using \inlinecode{cpptraj's } \inlinecode{bounds} command, the region of interest can be output to a file.
+Using \inlinecode{cpptraj's } \inlinecode{bound} command, the region of interest can be output to a file.
 For example, here is the command for a region of 10\AA{} around the ligand LIG:
@@
 \end{lstlisting}
-Note that the \inlinecode{bounds} command needs to be supplied with the exact same trajectory as used in the GIST calculation to generate correct boundaries.
-Therefore, any preprocessing steps need to match between the \inlinecode{bounds} command and the following \inlinecode{gist} calculation.
-
+Note that any loading and preprocessing, such as imaging, needs to be run before this command and needs to be the exact same as is used for the \inlinecode{gist} command.
 Additionally, a helper script called \inlinecode{FindCentroid.py} for the same purpose is supplied with this tutorial, and can be used as follows:
@@
 The command needed to run GIST in cpptraj can be written in an input file (e.g gist.in) for convenience.
-The basic command, assuming a grid centered at (10.5 20.1 30.1) and a grid size of $30\times30\times30$ voxels with \SI{0.5}{\angstrom} spacing, is:
+The basic command, assuming a given grid center and a grid size of $30\times30\times30$ voxels with \SI{0.5}{\angstrom} spacing, is:
 
@@
 
+Note that the solute should not move significantly for a GIST calculation, since IST does not take the solute degrees of freedom into account.
+This is typically achieved using restraints on the solute heavy atoms.
 This command tells \software{cpptraj} to run a GIST calculation using the specified region and print the output in a file called \inlinecode{gist.dat}.
-An overview of how to work with \software{cpptraj} is provided in the program's documentation \cite{cpptraj_doc}, AMBER tutorials \cite{amber_tut_cpptraj} and on the AMBER-hub website \cite{amber_hub}.
+You will need to load and image your trajectory before running this command.
+An overview of how to work with \software{cpptraj} is provided in the program's documentation \cite{cpptraj_doc}, AMBER tutorials \cite{amber_tut_cpptraj} or on the AMBER-hub website \cite{amber_hub}.
 
@@
 \end{table}
-In GIST, the calculated properties are derived by referencing against the bulk solvent.
-It is therefore important to set the reference number density $\rho_0$ according to the solvent model used in your simulation.
+
+It is important to set the reference density $\rho_0$ according to the solvent model used in your simulation.
 The solvent-solvent enthalpy $E_{ww}$ also needs to be referenced, but this can be done in the post-processing stage.
-The calculation of these properties and the need for referencing is discussed in-depth in section /ref{sec:theory}.
 
@@
 \begin{table}[h]
-	\label{tab:ref_values}
-	\caption{Reference number density $\rho_0$ and mean water-water energy $E_{ww, norm}$ for various water models from NPT simulations of N$_{WAT}$ water molecules at \SI{300}{\kelvin} and 1bar.}\label{tab:ref_densities}
+	\caption{Reference density $\rho_0$ and mean water-water energy $E_{ww, norm}$ for various water models from NpT simulations of N$_{WAT}$ water molecules at \SI{300}{\kelvin} and 1bar.}\label{tab:ref_densities}
 	\small
@@
 	\item{[gX] For every element in the main solvent, the number density of atoms found in the voxel, in units of the bulk density.
-If the same element occurs multiple times, the bulk density is scaled such that that the expectation value for the bulk is unity.
+If the same element occurs multiple times, the bulk density is scaled accordingly.
 For water, gO and gH are produced.}
-
 	\item{[Esw] Mean solute-water interaction energy density.}
@@
 A full list of all properties in the output is provided at the end of the tutorial.
-The output file is organized by voxel and sorted according to their x, y, and z coordinates, describing the center point of the voxel.
+
+The output is organized by voxel and sorted according to their x, y, and z coordinates, describing the center point of the voxel.
 Energy and entropy quantities are reported in two different ways.
@@
 The molecule should be solvated with a sufficiently large solvent box.
-For periodic boundary conditions, the size of the solvent box should be large enough for the solvent to behave bulk-like so as to reduce artifact from solvent-mediated interactions between the periodic images.
-This corresponds to a distance between solute and box boundary of at least 12\AA{} for water, though we recommend 15\AA{} or higher.
 Carefully check the employed force field, protonation state and partial charges.
@@
 Solvate the parametrized biotin, making sure to choose a water box of large enough size so that the water on the box edges behaves bulk-like (a wall distance of at least 15\AA is recommended).
-A good starting point for the parametrization and solvation of small molecules is the Amber tutorial 4b (\url{http://ambermd.org/tutorials/basic/tutorial4b/}).
+A good starting point (though a bit dated) for the parametrization and solvation of small molecules is the Amber tutorial 4b (\url{http://ambermd.org/tutorials/basic/tutorial4b/}).
 
 \subsubsection{Equilibration}
-Equilibrate your solvated biotin in an NVT or NPT ensemble.
-Even when running the simulation in NVT, it is advisable to include an NPT equilibration step to adjust the box size beforehand.
+Equilibrate your solvated biotin in an NVT or NpT ensemble.
+Even when running the simulation in NVT, it is advisable to include an NpT equilibration step to adjust the box size beforehand.
 Use enough equilibration time to allow the solvent to relax. 
-For water, \SI{0.5}{\nano\second} is usually sufficient to equilibrate the density, but narrow binding pockets can take longer.
+For water, \SI{0.5}{\nano\second} are usually sufficient to equilibrate the density, but narrow binding pockets can take longer.
 
@@
 \subsubsection{MD simulation}
-\label{sec:md_simulation}
+
 Now, you can run a restrained MD simulation based on the equilibrated structure of biotin for the subsequent GIST analysis.
@@
 \begin{lstlisting}[style=amber-in]
-restrained 100 ns NPT
+restrained 100 ns NpT
 &cntrl
@@
 	ntr=1, restraint_wt=100.0,
-	restraintmask='!@H=&!:WAT',
+	restraintmask="!@H=&!:WAT",
 	ntp=1, pres0=1.0, taup=1.0,
@@
 
-This file will run a \SI{100}{\nano\second} NPT MD simulation while keeping all atoms except for hydrogens and waters restrained.
+This file will run a \SI{100}{\nano\second} NpT MD simulation while keeping all atoms except for hydrogens and waters restrained.
 Note that these are very standard MD settings except for the restraints.
@@
 \subsubsection{GIST analysis}
-To start our GIST analysis of biotin, we'll first have to decide on the parameters defining the GIST grid during the analysis.
-We will need a solute structure without water for the analysis of our future GIST results.
-As we ran our MD simulation with periodic boundaries, we always need to reimage all molecules to the original box using cpptrajs \inlinecode{autoimage} function.
-We also recommend that you center the biotin molecule to the coordinate origin, as this simplifies the definition of our grid origin later on.
-The following code snippets are
+To start our GIST analysis of biotin, we'll first have to decide on the parameters defining the GIST grid during the analyis.
+We recommend that you center the biotin molecule to the coordinate origin:
 \begin{lstlisting}[style=cpptraj]
@@
 trajout biotin-centered.pdb
-run
-\end{lstlisting}
-In this code snippet, the first two lines load the topology and coordinates of the first frame of our MD simulation.
-In the third line, we reimage the coordinates to the first box and center the biotin at the origin.
-Then, we strip all water molecules in the fourth line and write out a pdb file in the fith line.
-Finally, we start the analysis with the "run" command.
-
+\end{lstlisting}
 To decide on a grid size, find the extent (minimum and maximum) of the x-, \mbox{y-,} and z-coordinates of the molecule.
@@
 bounds !(:WAT) dx 0.5 offset 30 name MyGrid out bounds.dat
-run
-\end{lstlisting}
-In the above snippet, we calculate the bounds around the whole protein for a grid with \SI{0.5}{\angstrom} side-length voxels and a buffer zone of \SI{15.0}{\angstrom}.
-Note that the offset is given in terms of number of additional voxels, not length dimension.
-By varying the atom mask in the \inlinecode{bound} command, it is possible to center the grid on a region of interest, similar to the \software{FindCentroid.py} script.
-
-Now, you can run the actual GIST analysis.
-Reimage and center the molecule as above, to make sure that its position matches the grid parameters you decided on.
+\end{lstlisting}
+Here, we calculate the bounds around the whole protein for a grid with \SI{0.5}{\angstrom} side-lenght voxels and a buffer zone of \SI{15.0}{\angstrom}.
+Note that the offset is given in terms of number of additional voxels, not voxel dimension.
+By varying the atom mask in the \inlinecode{bound} command, it is possible to center the grid on a region of interest, similiar to the \software{FindCentroid.py} script.
+
+Now, you can finally run the actual GIST analysis.
+Center the molecule as above, to make sure that its position matches the grid parameters you decided on.
 We recommend using either the PME or GPU implementation of GIST.
@@
 autoimage !(:WAT) origin
-# optionally remove the 'pme' option
+# optionally remove the "pme" option
 gist griddim 100 100 100 out gist.dat \
@@
 \subsubsection{Working with the GIST results}
-Running the GIST analysis above generates an output data file (gist.dat) containing all calculated quantities for each voxel in the system. 
+Running the GIST analysis above, an output data file (gist.dat) containing all calculated quantities for each voxel in the system is produced. 
 Additionally as outlined in section \ref{sec:running_GIST}, various quantities are written to OpenDX files.
@@
 gist = load_gist_file(
-'gist.dat', struct='solute-centered.pdb',  
-eww_ref=-9.539)
+"gist.dat", struct="solute-centered.pdb",  eww_ref=-9.539)
 print(f'n_frames = {gist.n_frames}')
@@
 Gisttools provides access to all quantities listed in the GIST output file.
-Additionally, it calculates the total enthalpy per voxel (Eall) as a sum of Eww and Esw, and the free energy of solvation (A) by subtracting $\Delta$TSsix from Eall.
+Additionally, it calculates the total enthalpy per voxel (Eall) as a sum of Eww and Esw, and the free energy of solvation (A) by subtracting dTSsix from Eall.
 
@@
 
-Gisttools also provides OpenDX writing capabilities to visualize the calculated or modified GIST data.
+Gisttools also provides minimal OpenDX writing capabilities to visualize the calculated or modified GIST data.
 For visualization, you'll most likely want to write out the density-normalized GIST quantities (X\_dens):
@@
 
-Section~\ref{sec:visualization} discusses the visualization of OpenDX files.
-
+This tutorial discusses the visualization of OpenDX files in Section~\ref{sec:visualization}.
 \subsection{Applying GIST for the Streptavidin-Biotin complex}
@@
 In our example, we only use a single equilibrated structure.
-A more sophisticated approach would be to identify several dominant structures in the ensemble and run a separate MD simulation and GIST calculation for each structure.
-Such dominant structures could either be identified from experimental ensembles or by clustering an unrestrained MD simulation.
+A more sophisticated approach would be to identify several dominant structures in the ensemble -- either from experimental ensembles or by clustering an MD simulation.
 
@@
 Note that, due to the restraints, some solvent sites might be inaccessible to the bulk solvent.
-To investigate such positions, make sure to explicitly place water molecules there. %TODO add MC/MD or mupt and sources
+To investigate such positions, make sure to explicitly place water molecules there.
 It is good practice to check your restrained MD simulation or solvent populations obtained by GIST to confirm whether all relevant sites were properly solvated.
 
-When calculating solvation free energy differences between a complex and the corresponding monomers (the dissociated state), there are generally two approaches for dealing with configurational diversity.
-In the first approach, the ensemble properties of each individual part and the complex are accounted for by running individual MD simulations for each.
-The second approach assumes sufficient overlap in the conformational space of the dissociated states and the complex.
-Then, only configurations of the complex are sampled in a single MD simulation of the complex.
-Calculations for the disassociated states are then run on the extracted structures from the complex configurations.
-%TODO source MMPBSA
+When calculating solvation free energy differences between a complex and the corresponding monomers (the dissociated state), there are two options for dealing with configurational diversity in the structures.
 One can choose to sample the complex and dissociated states individually, or to sample only one of them and assume that there is sufficient overlap in the conformations.
@@
 It is therefore important that the molecular conformations in the complex and dissociated states match exactly.
-Use the \software{equilibration.py} script (or your own procedure) to perform short NVT and NPT equilibration runs on the complex structure.
-Note that the equilibration script contains unrestrained NPT steps, because restraining two molecules at the same time might interfere with the pressure equilibration.
-Then, use \software{cpptraj} to produce biotin and streptavidin structures based on the complex by stripping the respective other molecule like this:
+Use \software{cpptraj} to produce biotin and streptavidin structures based on the complex by stripping the respective other molecule like this:
 \begin{lstlisting}[style=cpptraj]
@@
 \end{lstlisting}
-Adapt the above input file for biotin by striping molecule 1 instead of 2 and replacing the path names in the \inlinecode{strip} and  \inlinecode{trajout} commands.
-Alternatively, we provide a small script named \software{cpptraj\_remove\_mol.sh} for this procedure.
-
+(Adapt for biotin as necessary.)
+
+Use the \software{equilibration.py} script (or your own procedure) to perform short NVT and NpT equilibration runs on the complex structure.
+Note that the equilibration script contains unrestrained NpT steps, because restraining two molecules at the same time might interfere with the pressure equilibration.
+
+Then split the equilibrated structure into biotin and streptavidin parts using \software{cpptraj\_remove\_mol.sh}.
 After that, use \software{equilibration.py -R} to run short restrained 
@@
 
+\subsubsection{MD Simulation}
+The MD can be run as described above for each monomer and the complex.
+
 \subsubsection{GIST analysis of Streptavidin-Biotin}
-After running the MD simulations as described in section \ref{sec:md_simulation}, proceed with the GIST analysis.
 For easier post-processing, here we use grids that contain the whole molecule.
@@
 Calculate the GIST grid parameters as outlined in the biotin section for the complex and streptavidin and run the GIST analysis.
-For larger systems like whole proteins you can usually get away with a smaller buffer region for the GIST grid compared to small molecules.
-To test whether your grid is large enough, look at solvent properties close to the grid edges, e.g. the solvent-solvent energies.
+For larger systems like whole proteins you can usually get away with a smaller buffer region compared to small molecules.
+To test whether your grid is large enough look at solvent properties close to the grid edges, e.g. the solvent-solvent energies.
 If these correspond to the respective bulk properties, it can be assumed that your grid is large enough to capture all perturbations of the solute on the solvent.
 Note that a different box-size or number of solvent molecules can skew the bulk properties slightly compared to the reference values provided above.
-In that case, checking for convergence of the solvent properties near the grid borders is sufficient.
-You might want to modify your reference values according to your results, as described in the next section.
+In that case, checking for convergence of the solvent properties near the grid borders is sufficient but you might want to modify your reference values according to your results then as described in the next section.
 
@@
 \subsubsection{Reference Values and Radial Convergence}
-All GIST quantities should be expressed relative to bulk as outlined in section \ref{sec:theory}.
+All GIST quantities should be expressed relative to bulk.
 This is automatically done by \software{cpptraj} for the entropy, using the bulk density specified by \inlinecode{refdens}.
@@
 The Amber manual provides reference densities and energies for several solvents.
-Furthermore, a large variety of reference values calculated with GPU-GIST are listed in table \ref{tab:ref_values}.
 However, the exact energy values are different when using PME-GIST, and also depend on the box size.
@@
 Therefore, the reference energy can be obtained by binning the voxels by their distance to the solute, and evaluate Equation~\ref{eq_ewwref} within each bin.
-If this value converges with increasing distance from the solute, this can be used as $E_{ww}^\textit{ref}$.
+If this value converges, this can be used as $E_{ww}^\textit{ref}$.
 
@@
 Although it has a simple convergence check built-in, it is always recommended to check the convergence by hand.
-A handy way to check this is to calculate the radial distribution of this GIST property, i.e.
-a histogram in relation to the distance to the solute.
-Due to the conceptual similarity with classical radial distribution functions, describing the density of distances between two particle types, this function is exposed to the user as the \inlinecode{rdf} method.
 
@@
 gist = load_gist_file(
-    'gist.dat', struct='solute-centered.pdb')
+    "gist.dat", struct="solute-centered.pdb")
 bins, eww = gist.rdf(
-    'Eww_unref_norm', bins=100, rmax=20.,
-    normalize='norm')
+    "Eww_unref_norm", bins=100, rmax=20.,
+    normalize="norm")
 plt.plot(bins, eww)
@@
 The expected output of this code is shown in Figure~\ref{fig_ewwref}.
-%TODO redo figure caption
+
 \begin{figure}
@@
 In a plot like Figure~\ref{fig_ewwref} based on the \inlinecode{A\_dens} column, the values should tend to zero.
-However, it is more informative to plot the cumulative free energy contribution against the distance to the solute, to check the convergence (see Figure~\ref{fig_radial_convergence}).
+However, it is more informative to plot the cumulative free energy contribution against the distance to the solute, to check the convergence.
 It the curve flattens out, the converged value is your final $\dgsolv$\@.
-If the curve diverges even for large integration distances you might need to tweak the $E_{ww}$ reference, or introduce a reference value for the entropy.
+Otherwise, you might need to tweak the $E_{ww}$ reference, or introduce a reference value for the entropy.
 
@@
 import matplotlib.pyplot as plt
-import numpy as np
-# adapt eww_ref according to the used solvent model! 
-gist = load_gist_file('gist.dat',
-    struct='solute-centered.pdb', eww_ref=-9.533)
+# adapt eww_ref!
+gist = load_gist_file("gist.dat",
+    struct="solute-centered.pdb", eww_ref=-9.533)
 bins, (dg, esw, eww, s) = gist.rdf(
-    ['A_dens', 'Esw_dens', 'Eww_dens', 'dTSsix_dens'],
-    bins=100, rmax=20., normalize='none')
-plt.plot(bins, np.cumsum(eww), label='Eww')
-plt.plot(bins, np.cumsum(esw), label='Esw')
-plt.plot(bins, np.cumsum(dg), label='dG')
-plt.plot(bins, np.cumsum(s), label='dS')
-plt.legend()
-plt.xlabel("distance cutoff [Å]")
-plt.ylabel("free energy contribution [kcal/mol]")
+    ["A_dens", "Esw_dens", "Eww_dens", "dTSsix_dens"],
+    bins=100, rmax=20., normalize="none")
+plt.plot(bins, np.cumsum(dg), label="dG")
+plt.plot(bins, np.cumsum(eww), label="Eww")
+plt.plot(bins, np.cumsum(esw), label="Esw")
+plt.plot(bins, np.cumsum(s), label="dS")
 \end{lstlisting}
@@
 %With \software{gistpp}, you can use the file from the convergence check. Additional OpenDX files can be produced using \inlinecode{gistpp -op makedx}.
-With \software{gisttools}, you can create one using e.g., \inlinecode{gist.save\_dx('A\_dens', 'A\_dens.dx')}.
+With \software{gisttools}, you can create one using e.g., \inlinecode{gist.save\_dx("A\_dens", "A\_dens.dx")}.
 
@@
 import gisttools as gt
-gist = gt.gist.load_gist_file('gist.dat',
-    struct='solute-centered.pdb', eww_ref=-9.533)
+gist = gt.gist.load_gist_file("gist.dat",
+    struct="solute-centered.pdb", eww_ref=-9.533)
 gist['E_norm'] = gist['Eww_norm'] * 2 + gist['Esw_norm']
@@
 
-The OpenDX files can then be imported into your molecular viewer of choice.
-%TODO we are missing a pymol script here
-%Adapt the above PyMOL script to visualize the average energy per solvent at a positive and a negative isolevel, as well as regions of negative entropy per solvent.
-A visualization of entropy and energy contributions to the free energy of solvation of the streptavidin binding pocket is shown in Figure~\ref{fig_binding_pocket_pymol}.
-PyMOL input scripts to generate such visualizations are available in the GitHub repository for this tutorial.
+Visualize the free energy, the energy, and the entropy, at several isolevels.
+
+Adapt the above PyMOL script to visualize the average energy per solvent at a positive and a negative isolevel, as well as regions of negative entropy per solvent.
+The expected representation of the streptavidin binding pocket is shown in Figure~\ref{fig_binding_pocket_pymol}.
 
@@
 
-%You should find that there are several regions in the binding pocket that feature a high solvent energy and/or low entropy.
-%However, such regions also exist outside of the binding pocket, and there are also low-energy regions within the binding pocket.
-%Therefore, the strong affinity to biotin cannot be explained only by hydration, and the direct interaction energy also needs to be taken into account.
+You should find that there are several regions in the binding pocket that feature a high solvent energy and/or low entropy.
+However, such regions also exist outside of the binding pocket, and there are also low-energy regions within the binding pocket.
+Therefore, the strong affinity to biotin cannot be explained only by hydration, and the direct interaction energy also needs to be taken into account.
 
@@
 \end{lstlisting}
-
 Assign reference energies and check them for plausibility.
-
 \begin{lstlisting}[style=python]
 biotin.eww_ref = biotin.detect_reference_value()
-print('Biotin:', biotin.eww_ref)
+print("Biotin:", biotin.eww_ref)
 strept.eww_ref = strept.detect_reference_value()
-print('Streptavidin:', strept.eww_ref)
+print("Streptavidin:", strept.eww_ref)
 compl.eww_ref = compl.detect_reference_value()
-print('Complex:', compl.eww_ref)
+print("Complex:", compl.eww_ref)
 \end{lstlisting}
@@
 Note that we use \software{mdtraj} here, since \software{gisttools} stores the reference structure as an \software{mdtraj} \inlinecode{Trajectory} object.
-Here, we compute the histogram of the contributions of single properties to assess the convergence.
-
 \begin{lstlisting}[style=python]
-import mdtraj as md
 col = 'dTSsix_dens'
 def select(traj, sel):
-    # Slice a Trajectory by selection mask.
+    "Slice a Trajectory by selection mask."
     return traj.atom_slice(traj.top.select(sel))
 # we multiply by 10 to convert nm to Angstrom.
-biotin_mask = 'resname BTN and not element H'
-strept_mask = 'not resname BTN and not resname WAT and not element H'
 compl_x = select(compl.struct, biotin_mask).xyz[0] * 10.
@@
 
-Now, subtract the monomer histograms from the complex, and compute the sum of your property within some cutoff distance to the solute.
-If you also visualize the individual histograms, you will notice that the difference converges much better with increasing radius than the single contributions.
+Now, subtract the monomer rdfs from the complex, and compute the integrals.
+If you also visualize the individual rdfs, you will notice that the difference converges much better with increasing radius than the contributions.
 This is because the cutoff contains regions that are close to the atom, since the integration region does not comprise the whole protein.
-%TODO this sentence doesn't make sense
 
@@
 integral = difference[bins < cutoff].sum()
-print('Integral = {}'.format(integral))
+print("Integral = {}".format(integral))
 plt.plot(bins, np.cumsum(difference))
@@
 
+Compute the energy (\inlinecode{Eall}) and entropy (\inlinecode{dTSsix}) contributions separately.
+You will find that the energy disfavors binding, because we do not yet include the interaction energy between biotin and streptavidin.
+To make this easier, gisttool includes a function to integrate in a radius around atom centers:
+\begin{lstlisting}[style=python]
+centers = {"biotin": biotin_x, 
+	   "strept": strept_x, 
+	   "compl": compl_x}
+gist_objs = {"biotin": biotin, 
+             "strept": strept, 
+             "compl": compl}
+cols = ['Eall_dens','dTSsix_dens']
+rmax = 12
+results = {
+    name: gist.integrate_around(
+        cols, rmax, centers[name])
+    for name, gist in gist_objs.items()}
+display(pd.DataFrame(results))
+
+\end{lstlisting}
 You can also plot the difference between the complex and monomer contributions against the distance to biotin, using the complex coordinates for the holo structure:
@@
 \end{lstlisting}
-
+You can also plot the difference between the complex and monomer contributions against the distance to biotin, using the complex coordinates for the holo structure.
 The expected result is shown in Figure~\ref{fig-dg-sums}.
@@
 
-Compute the energy (\inlinecode{Eall}) and entropy (\inlinecode{dTSsix}) contributions separately.
- To make this easier, gisttool includes a function to integrate in a radius around atom centers:
-\begin{lstlisting}[style=python]
-import pandas as pd
-centers = {'biotin': biotin_x, 
-	'strept': strept_x, 
-	'compl': compl_x}
-gist_objs = {'biotin': biotin, 
-	'strept': strept, 
-	'compl': compl}
-cols = ['Eall_dens','dTSsix_dens']
-rmax = 12
-results = {
-	name: gist.integrate_around(
-	cols, rmax, centers[name])
-	for name, gist in gist_objs.items()}
-print(pd.DataFrame(results))
-
-\end{lstlisting}
-You will find that the energy disfavors binding, because we do not yet include the interaction energy between biotin and streptavidin.
 Using the \inlinecode{energy} command in \software{cpptraj}, you can compute this interaction energy.
-We recommend using PME in combination with PME-GIST, but not with GPU-GIST to be in line with the different ways the energy is calculated in the two GIST methods.
+We recommend using PME in combination with PME-GIST, but not with GPU-GIST\@.
 An example \software{cpptraj} input might look like:
@@
 Even though streptavidin-biotin is known to be a very stable complex, the solvation energy favors the dissociation.
-This is expected: the binding of biotin to streptavidin is faciliated over a large number of H-bonds and over polar interactions.
-As such, water favourably solvates both biotin and streptavidin and an energetic penalty is paid when displacing it during the binding process.
+This is expected: if the molecules were not able to form any interactions, but interacted fully with the water, they would not form a complex.
 
@@
 def select(traj, sel):
-	# Slice a Trajectory by selection mask.
+	"Slice a Trajectory by selection mask."
 	return traj.atom_slice(traj.top.select(sel))
@@
 
-The entropy can be approximated as an infinite expansion of correlations of increasing order.
-In a slightly simplified view, the first order is the log solvent density $ln\left(g_\textit{sw}\left(\textbf{r},\omega \right)\right)$ at each position $\textbf{r}$ and orientation $\omega$.
+For the entropy, this leads to an infinite expansion of correlations of increasing order.
+In a slightly simplified view, the first order is the solvent density $g_\textit{sw}\left(\textbf{r},\omega \right)$ at each position $\textbf{r}$ and orientation $\omega$.
 The first-order entropy omits all solvent-solvent (and higher) correlations, while the solute-solvent correlation is taken into account because the coordinates are relative to the solute.

### 2024-06-04 manuscript/manuscript.tex
Update manuscript.tex

@@
 Therefore, it is advisable to conduct your own reference calculations, especially for systems that are either very large or very small.
-While the impact on density is not as pronounced, deviations from the specified settings can lead to differences of up to 0.005 kcal/mol in the $E_{ww, norm}$ value for water boxes.
+While the impact on density is not as pronounced, deviations from the specified settings can lead to differences of up to 0.005 kcal/mol in the energy per water molecule ($E_{ww, norm}$) for water boxes.
 These discrepancies, when summing over large grid regions, have the potential to introduce significant errors due to incorrect referencing.

### 2024-06-04 manuscript/manuscript.tex
Update manuscript.tex

@@
 The \inlinecode{\_dens} quantities are normalized by the voxel volume and can be used to compute integrals.
-The \inlinecode{\_norm} quantities are instead normalized by the number of solvents in the respective voxel averaged over all considered frames, and are useful to visualize local water properties.
+The \inlinecode{\_norm} quantities are instead normalized by the solvent number density in the respective voxel averaged over all considered frames, and are useful to visualize local water properties.
 To convert from a \inlinecode{\_norm} quantity $X^k_{norm}$ in voxel $k$ to a \inlinecode{\_dens} quantity $X^k_{dens}$, the following equation can be applied:

### 2024-06-04 manuscript/manuscript.tex
Update manuscript.tex

@@
 Since the density is computed from an MD simulation, GIST can accurately treat confined water regions, which are challenging for implicit-solvent methods such as GBSA and self-consistent theories such as 3D-RISM \cite{Kovalenko1998-3drism}.
-While the need for MD simulations comes at a large computational cost, GIST is still more efficient than end-state approaches such as TI \cite{Kirkwood1935-ti} or FEP \cite{Zwanzig1954-reweighting}, with applications to large systems such as serine proteases \cite{Kraml2019-gigist} or antibody fragments \cite{Waibl2021-gist-antibodies}.
+While the need for MD simulations comes at a large computational cost, GIST is still more efficient than end-states approaches such as TI \cite{Kirkwood1935-ti} or FEP \cite{Zwanzig1954-reweighting}, with applications to large systems such as serine proteases \cite{Kraml2019-gigist} or antibody fragments \cite{Waibl2021-gist-antibodies}.
 On the other hand, the first-order approximation means that the solvation entropy is less accurately described than in TI or FEP \cite{Chen2021,Waibl2021-gist-salt}.
@@
 In the first part, we show how to run GIST calculations for small molecules and interpret the output.
-Afterwards, we apply this knowledge to a protein binding pocket, visualize $\dgsolv$ hotspots and estimate a free energy of ligand binding using an end-state approach based on GIST.
+Afterwards, we apply this knowledge to a protein binding pocket, visualize $\dgsolv$ hotspots and estimate a free energy of ligand binding using an end-states approach based on GIST.
 In addition, we discuss all steps in further detail at the end.
@@
 Now, you can finally run the actual GIST analysis.
-Center the molecule as above, to make sure that its position matches the grid parameters you decided on.
+Center the molecule as above, to verify that its position matches the grid parameters you decided on.
 We recommend using either the PME or GPU implementation of GIST.
@@
 \subsubsection{Working with the GIST results}
-Running the GIST analysis above, an output data file (gist.dat) containing all calculated quantities for each voxel in the system is produced. 
+Running the GIST analysis above, an output data file (gist.dat) containing all calculated quantities for each voxel in the grid is produced. 
 Additionally as outlined in section \ref{sec:running_GIST}, various quantities are written to OpenDX files.
@@
 Gisttools provides access to single or multiple voxels through a .loc property similiar to the pandas library. 
-GIST quantity data for all voxels can be accessed through directly indexing the gist object, either by providing a single name or a list of names.
+GIST quantity data for all voxels can be accessed by directly indexing the gist object, either by providing a single name or a list of names.
 Indexing with .loc works by providing a voxel index (either a index number or list of numbers) and optionally a column index (column name or list of names).
@@
 \subsection{Applying GIST for the Streptavidin-Biotin complex}
-In this section we will apply GIST to the full complex and discuss each step of the process in more detail to outline pitfalls.
+In this section we will apply GIST to the full complex and discuss each step of the process in more detail to outline potential pitfalls.
 We will afterwards introduce advanced analysis workflows which allow you to modify and better visualize the resulting data.
@@
 systems.
-By using restraints, we can make sure that the monomer conformations match the complex.
+By using restraints, we can ensure that the monomer conformations match the complex.
 

### 2024-06-04 manuscript/manuscript.tex
Update manuscript.tex

@@
 To test whether your grid is large enough look at solvent properties close to the grid edges, e.g. the solvent-solvent energies.
-If these correspond to the respective bulk properties, it can be assumed that your grid is large enough to capture all perturbations of the solute on the solvent.
+If these correspond to the respective bulk properties, it can be assumed that your grid is large enough to fully capture the perturbation of the solvent by the solute.
 Note that a different box-size or number of solvent molecules can skew the bulk properties slightly compared to the reference values provided above.

### 2024-06-04 manuscript/manuscript.tex
Update manuscript.tex

@@
 
-The most accurate method to compute a reference energy is to run a GIST calculation of the pure solvent using the same energy method (PME/GPU) and a similar box size, and then compute the average solvent-solvent energy per molecule.
+The most accurate method to compute a reference energy is to run a GIST calculation of the pure solvent using the same energy method (PME/GPU), a similar box size, at the same temperature and pressure and then compute the average solvent-solvent energy per molecule.
 

### 2024-06-10 manuscript/manuscript.tex

@@
 GIST is an implementation of IST on a three-dimensional grid.
-It can be seen as a more general version of WaterMap \cite{Abel2008-watermap,Young2007-watermap} that provides solvation thermodynamics on a space filling grid, rather than high-density clusters in a binding pocket.
+Unlike the related method WaterMap \cite{Abel2008-watermap,Young2007-watermap}, it provides solvation thermodynamics on an entire grid, rather than only high-density clusters in a binding pocket.
 This implies a higher versatility, but also puts more responsibility on the user in terms of correct post-processing.
 
-Since the density is computed from an MD simulation, GIST can accurately treat confined water regions, which are challenging for implicit-solvent methods such as GBSA and self-consistent theories such as 3D-RISM \cite{Kovalenko1998-3drism}.
-While the need for MD simulations comes at a large computational cost, GIST is still more efficient than end-states approaches such as TI \cite{Kirkwood1935-ti} or FEP \cite{Zwanzig1954-reweighting}, with applications to large systems such as serine proteases \cite{Kraml2019-gigist} or antibody fragments \cite{Waibl2021-gist-antibodies}.
+Since the density is computed from an explicit solvent MD simulation, GIST provides an accurate picture of solvation at an atomistic level, even for demanding systems such as drug binding sites in proteins.
+While the need for MD simulations comes at a large computational cost, GIST is still more efficient than end-state approaches such as TI \cite{Kirkwood1935-ti} or FEP \cite{Zwanzig1954-reweighting}, with applications to large systems such as serine proteases \cite{Kraml2019-gigist} or antibody fragments \cite{Waibl2021-gist-antibodies}.
 On the other hand, the first-order approximation means that the solvation entropy is less accurately described than in TI or FEP \cite{Chen2021,Waibl2021-gist-salt}.
-%The 3D-RISM approach allows for a decomposition into entropic and enthalpic terms as well as a full localization of the underlying distributions to the solute structure.
-%Here you would explain what problem you are tackling and briefly motivate your work.
-%
-%In this particular template, we have removed most of the usage examples which occur in \texttt{sample-document.tex} to provide a minimal template you can modify; however, we retain a couple of examples illustrating more unusual features of our templates/article class, such as the checklists, and information on algorithms and pseudocode.
-%
-%Keep in mind, as you prepare your manuscript, that you should plan for a representative image  which will be used to highlight your article on the journal website and publications.
-Usually, this would be one of your figures, but it must also be uploaded separately upon article submission.
-We give specific guidelines for this image on the journal website in the section on article submission (see \url{https://livecomsjournal.github.io/authors/policies/index.html#article-submission}).
-%
-%Additionally, for well-formatted manuscripts, we recommend that you let LaTeX handle figure/table placement for you as much as possible, so please avoid specifying strenuous float instructions like `[h!]` and `[H]` as much as possible.
+Similar to the self-consistent integral theories, however, GIST only considers a single solute conformation and  multiple computations of representative structures are necessary to describe ensemble properties.
 
@@
 This tutorial is aimed at users with a solid knowledge of molecular dynamics (MD) simulations.
+
 The user should be able to run MD simulations and analyze the resulting trajectories. 
-Previous experience of GIST is not necessary, although we recommend reading the articles referenced in section~\ref{sec:theory}.
+The MD simulations presented in this tutorial are run with the Amber simulation engine and analyzed using AmberTools \software{cpptraj}. 
+For a conceptual overview of running MD simulations with Amber, we recommend reading the Amber manual (\url{https://ambermd.org/Manuals.php}) \cite{amber22} and the Amber tutorials (\url{https://ambermd.org/Manuals.php}).
+Previous experience of GIST is not necessary, although we recommend reading the works referenced in section~\ref{sec:theory}.
 
@@
 	\item \software{mdtraj} (\url{https://www.mdtraj.org/}, $\geq$ 1.9.7) \cite{McGibbon2015-mdtraj}
+	\item \software{matplotlib} (\url{https://matplotlib.org/}, $\geq$ 3.7.0) \cite{Hunter2007-matplotlib}
+	\item \software{numpy} (\url{https://numpy.org/}, $\geq$ 1.23.5) \cite{harris2020-numpy}
+	\item \software{pandas} (\url{https://pandas.pydata.org/}, $\geq$ 1.5.3) \cite{pandas-2023_8239932}
 \end{itemize}
@@
 In order to run a GIST analysis, one must define the region where solvent properties will be calculated.
-This region is represented by a rectangular grid with given position, size, and voxel spacing.
-The default value of \SI{0.5}{\angstrom} provides reasonable discretization for many use-cases.
-While integrals of GIST quantities over the grid are independent of the voxel size, values for each voxel converge slower at high resolutions.
-
+
+This region is occupied by a rectangular grid with given position, size, and voxel spacing.
+The default spacing of \SI{0.5}{\angstrom} provides reasonable discretization for many use-cases.
+While integrals of GIST quantities over the grid are independent of the voxel size, provided sufficient sampling, individual values converge slower at high resolutions.
 The grid is defined by parameters of the \inlinecode{gist} command in \software{cpptraj}.
-The grid center is defined by \inlinecode{gridcntr}, the voxel number in x, y, and z direction by \inlinecode{griddim}, and the spacing (voxel side-length) in \AA{} by \inlinecode{gridspacn}.
+The grid center is defined by \inlinecode{gridcntr}, the voxel number in x, y, and z direction by \inlinecode{griddim}, and the spacing (voxel side-length) in \AA{} by \inlinecode{gridspacn}:
 
@@
 \subsubsection{GIST analysis}
-To start our GIST analysis of biotin, we'll first have to decide on the parameters defining the GIST grid during the analyis.
-We recommend that you center the biotin molecule to the coordinate origin:
+To start our GIST analysis of biotin, we'll first have to decide on the parameters defining the GIST grid during the analysis.
+We will need a solute structure without water for the analysis of our future GIST results.
+As we ran our MD simulation with periodic boundaries, we always need to reimage all molecules to the original box using cpptraj's \inlinecode{autoimage} function.
+We also recommend that you center the biotin molecule to the coordinate origin, as this simplifies the definition of our grid origin later on.
+The following code snippets are
 \begin{lstlisting}[style=cpptraj]
@@
 bounds !(:WAT) dx 0.5 offset 30 name MyGrid out bounds.dat
-\end{lstlisting}
-Here, we calculate the bounds around the whole protein for a grid with \SI{0.5}{\angstrom} side-lenght voxels and a buffer zone of \SI{15.0}{\angstrom}.
-Note that the offset is given in terms of number of additional voxels, not voxel dimension.
-By varying the atom mask in the \inlinecode{bound} command, it is possible to center the grid on a region of interest, similiar to the \software{FindCentroid.py} script.
-
-Now, you can finally run the actual GIST analysis.
-Center the molecule as above, to verify that its position matches the grid parameters you decided on.
+run
+\end{lstlisting}
+In the above snippet, we calculate the bounds around the whole protein for a grid with \SI{0.5}{\angstrom} side-length voxels and a buffer zone of \SI{15.0}{\angstrom}.
+Note that the offset is given in terms of number of additional voxels, not length dimension.
+By varying the atom mask in the \inlinecode{bound} command, it is possible to center the grid on a region of interest, similar to the \software{FindCentroid.py} script.
+
+Now, you can run the actual GIST analysis.
+Reimage and center the molecule as above, to verify that its position matches the grid parameters you decided on.
+
 We recommend using either the PME or GPU implementation of GIST.
@@
 \subsubsection{Working with the GIST results}
-Running the GIST analysis above, an output data file (gist.dat) containing all calculated quantities for each voxel in the grid is produced. 
+Running the GIST analysis above generates an output data file (gist.dat) containing all calculated quantities for each voxel in the grid. 
 Additionally as outlined in section \ref{sec:running_GIST}, various quantities are written to OpenDX files.
@@
 Calculate the GIST grid parameters as outlined in the biotin section for the complex and streptavidin and run the GIST analysis.
-For larger systems like whole proteins you can usually get away with a smaller buffer region compared to small molecules.
-To test whether your grid is large enough look at solvent properties close to the grid edges, e.g. the solvent-solvent energies.
-If these correspond to the respective bulk properties, it can be assumed that your grid is large enough to fully capture the perturbation of the solvent by the solute.
+
+For larger systems like whole proteins you can usually get away with a smaller buffer region for the GIST grid compared to small molecules.
+To test whether your grid is large enough, look at solvent properties close to the grid edges, e.g. the solvent-solvent energies.
+If these correspond to the respective bulk properties, it can be assumed that your grid is large enough to capture all perturbations of the solute on the solvent.
+
 Note that a different box-size or number of solvent molecules can skew the bulk properties slightly compared to the reference values provided above.
-In that case, checking for convergence of the solvent properties near the grid borders is sufficient but you might want to modify your reference values according to your results then as described in the next section.
+In that case, checking for convergence of the solvent properties near the grid borders is sufficient.
+You might want to modify your reference values according to your results, as described in the next section.
 
@@
 import matplotlib.pyplot as plt
-# adapt eww_ref!
-gist = load_gist_file("gist.dat",
-    struct="solute-centered.pdb", eww_ref=-9.533)
+import numpy as np
+# adapt eww_ref according to the used solvent model! 
+gist = load_gist_file('gist.dat',
+    struct='solute-centered.pdb', eww_ref=-9.533)
 bins, (dg, esw, eww, s) = gist.rdf(
-    ["A_dens", "Esw_dens", "Eww_dens", "dTSsix_dens"],
-    bins=100, rmax=20., normalize="none")
-plt.plot(bins, np.cumsum(dg), label="dG")
-plt.plot(bins, np.cumsum(eww), label="Eww")
-plt.plot(bins, np.cumsum(esw), label="Esw")
-plt.plot(bins, np.cumsum(s), label="dS")
+    ['A_dens', 'Esw_dens', 'Eww_dens', 'dTSsix_dens'],
+    bins=100, rmax=20., normalize='none')
+plt.plot(bins, np.cumsum(eww), label='Eww')
+plt.plot(bins, np.cumsum(esw), label='Esw')
+plt.plot(bins, np.cumsum(dg), label='dG')
+plt.plot(bins, np.cumsum(s), label='dS')
+plt.legend()
+plt.xlabel("distance cutoff [Å]")
+plt.ylabel("free energy contribution [kcal/mol]")
 \end{lstlisting}
@@
 Note that we use \software{mdtraj} here, since \software{gisttools} stores the reference structure as an \software{mdtraj} \inlinecode{Trajectory} object.
+Here, we compute the histogram of the contributions of single properties to assess the convergence.
+
 \begin{lstlisting}[style=python]
+import mdtraj as md
 col = 'dTSsix_dens'
 def select(traj, sel):
-    "Slice a Trajectory by selection mask."
+    # Slice a Trajectory by selection mask.
     return traj.atom_slice(traj.top.select(sel))
 # we multiply by 10 to convert nm to Angstrom.
+biotin_mask = 'resname BTN and not element H'
+strept_mask = 'not resname BTN and not resname WAT and not element H'
 compl_x = select(compl.struct, biotin_mask).xyz[0] * 10.
@@
 
+Compute the energy (\inlinecode{Eall}) and entropy (\inlinecode{dTSsix}) contributions separately.
+ To make this easier, gisttool includes a function to integrate in a radius around atom centers:
+\begin{lstlisting}[style=python]
+import pandas as pd
+centers = {'biotin': biotin_x, 
+	'strept': strept_x, 
+	'compl': compl_x}
+gist_objs = {'biotin': biotin, 
+	'strept': strept, 
+	'compl': compl}
+cols = ['Eall_dens','dTSsix_dens']
+rmax = 12
+results = {
+	name: gist.integrate_around(
+	cols, rmax, centers[name])
+	for name, gist in gist_objs.items()}
+print(pd.DataFrame(results))
+
+\end{lstlisting}
+You will find that the energy disfavors binding, because we do not yet include the interaction energy between biotin and streptavidin.
 Using the \inlinecode{energy} command in \software{cpptraj}, you can compute this interaction energy.
-We recommend using PME in combination with PME-GIST, but not with GPU-GIST\@.
+We recommend using PME in combination with PME-GIST, but not with GPU-GIST to be in line with the different ways the energy is calculated in the two GIST methods.
 An example \software{cpptraj} input might look like:
@@
 Even though streptavidin-biotin is known to be a very stable complex, the solvation energy favors the dissociation.
-This is expected: if the molecules were not able to form any interactions, but interacted fully with the water, they would not form a complex.
+This is expected: the binding of biotin to streptavidin is faciliated over a large number of H-bonds and over polar interactions.
+As such, water favourably solvates both biotin and streptavidin and an energetic penalty is paid when displacing it during the binding process.
 
@@
 \begin{lstlisting}[style=python]
+import gisttools as gt
 def select(traj, sel):
-	"Slice a Trajectory by selection mask."
+	# Slice a Trajectory by selection mask.
 	return traj.atom_slice(traj.top.select(sel))

### 2024-06-11 manuscript/manuscript.tex
Add Helmut to author list
Add Dan's recent MPI parallelization to tutorial and bibliography

--- 2024-06-04 manuscript/manuscript.tex
@@
 %\author[2]{Firstname Initials Surname}
-\author[2]{Steven Ramsey}
 \author[4]{Daniel R Roe}
@@
 
-In this tutorial, we aim to guide the reader through the most common steps involved in a GIST analysis using the streptavidin-biotin complex as an demonstrative system.
+In this tutorial, we aim to guide the reader through the most common steps involved in a GIST analysis at the example of the streptavidin-biotin complex.
 Furthermore, we discuss the theory of GIST with a focus on practical aspects.
@@
 GIST is an implementation of IST on a three-dimensional grid.
-It can be seen as a more general version of WaterMap \cite{Abel2008-watermap,Young2007-watermap} that provides solvation thermodynamics on a space filling grid, rather than high-density clusters in a binding pocket.
+It can be seen as a more general version of WaterMap \cite{Abel2008-watermap,Young2007-watermap} that provides solvation thermodynamics on the entire grid, rather than high-density clusters in a binding pocket.
 This implies a higher versatility, but also puts more responsibility on the user in terms of correct post-processing.
@@
 Since the density is computed from an MD simulation, GIST can accurately treat confined water regions, which are challenging for implicit-solvent methods such as GBSA and self-consistent theories such as 3D-RISM \cite{Kovalenko1998-3drism}.
-While the need for MD simulations comes at a large computational cost, GIST is still more efficient than end-states approaches such as TI \cite{Kirkwood1935-ti} or FEP \cite{Zwanzig1954-reweighting}, with applications to large systems such as serine proteases \cite{Kraml2019-gigist} or antibody fragments \cite{Waibl2021-gist-antibodies}.
+While the need for MD simulations comes at a large computational cost, GIST is still more efficient than end-state approaches such as TI \cite{Kirkwood1935-ti} or FEP \cite{Zwanzig1954-reweighting}, with applications to large systems such as serine proteases \cite{Kraml2019-gigist} or antibody fragments \cite{Waibl2021-gist-antibodies}.
 On the other hand, the first-order approximation means that the solvation entropy is less accurately described than in TI or FEP \cite{Chen2021,Waibl2021-gist-salt}.
@@
 The user should be able to run MD simulations and analyze the resulting trajectories.
-Previous experience of GIST is not necessary, although we recommend reading the articles referenced in section~\ref{sec:theory}.
+Previous experience of GIST is not necessary, although we recommend reading the works referenced in section~\ref{sec:theory}.
 
@@
 The Amber simulation engine is used in this tutorial to run the MD simulations.
-While Amber is free for non-commercial use, it might not always be available in industrial settings.
+While Amber is free for non-commercial use, it might not always be available in industry settings.
 In that case, other MD engines can be used as long as topologies and trajectories can be produced in a format supported by \software{cpptraj}.
@@
 The default value of \SI{0.5}{\angstrom} provides reasonable discretization for many use-cases.
-While integrals of GIST quantities over the grid are independent of the voxel size, values for each voxel converge slower at high resolutions.
+While integrals of GIST quantities over the grid are independent of the voxel size, individual values converge slower at high resolutions.
 
@@
 
-To use OpenMP or GPU-accelerated (using CUDA) versions of cpptraj, a different executable name such as \inlinecode{cpptraj.OMP} or \inlinecode{cpptraj.cuda} is often used.
+To use OpenMP, MPI or GPU-accelerated (using CUDA) versions of cpptraj, a different executable name such as \inlinecode{cpptraj.OMP}, \inlinecode{cpptraj.MPI} or \inlinecode{cpptraj.cuda} is often used.
 Therefore, make sure to use the right compilation and executable, to get full benefit of the various improvements to GIST.
@@
 Therefore, it is advisable to conduct your own reference calculations, especially for systems that are either very large or very small.
-While the impact on density is not as pronounced, deviations from the specified settings can lead to differences of up to 0.005 kcal/mol in the energy per water molecule ($E_{ww, norm}$) for water boxes.
+While the impact on density is not as pronounced, deviations from the specified settings can lead to differences of up to 0.005 kcal/mol in the $E_{ww, norm}$ value for water boxes.
 These discrepancies, when summing over large grid regions, have the potential to introduce significant errors due to incorrect referencing.
@@
 The \inlinecode{\_dens} quantities are normalized by the voxel volume and can be used to compute integrals.
-The \inlinecode{\_norm} quantities are instead normalized by the solvent number density in the respective voxel averaged over all considered frames, and are useful to visualize local water properties.
+The \inlinecode{\_norm} quantities are instead normalized by the number of solvents in the respective voxel averaged over all considered frames, and are useful to visualize local water properties.
 To convert from a \inlinecode{\_norm} quantity $X^k_{norm}$ in voxel $k$ to a \inlinecode{\_dens} quantity $X^k_{dens}$, the following equation can be applied:
@@
 In the first part, we show how to run GIST calculations for small molecules and interpret the output.
-Afterwards, we apply this knowledge to a protein binding pocket, visualize $\dgsolv$ hotspots and estimate a free energy of ligand binding using an end-states approach based on GIST.
+Afterwards, we apply this knowledge to a protein binding pocket, visualize $\dgsolv$ hotspots and estimate a free energy of ligand binding using an end-state approach based on GIST.
 In addition, we discuss all steps in further detail at the end.
@@
 Now, you can finally run the actual GIST analysis.
-Center the molecule as above, to verify that its position matches the grid parameters you decided on.
-We recommend using either the PME or GPU implementation of GIST.
-For the GPU version, cpptraj needs to be compiled with the \inlinecode{-cuda} flag.
-While the GPU version is generally faster, the PME implementation matches the energy of the Amber MD engine more closely.
+Center the molecule as above, to make sure that its position matches the grid parameters you decided on.
+We recommend using either the GPU or MPI-accelerated PME implementation of GIST.
+While the GPU version is faster when limited CPU resources are available, the PME implementation matches the energy of the Amber MD engine more closely.
+More recently, MPI support was added for both the energy and entropy calculation\cite{Roe2023-mpi-gist}. 
+The MPI implementation scales well over multiple CPU cores and can be combined with either the GPU or PME implementations for maximum efficiency. 
+A speed comparison of the various parallelization implementations is shown in \cite{Roe2023-mpi-gist}.
 
@@
 \subsubsection{Working with the GIST results}
-Running the GIST analysis above, an output data file (gist.dat) containing all calculated quantities for each voxel in the grid is produced. 
+Running the GIST analysis above, an output data file (gist.dat) containing all calculated quantities for each voxel in the system is produced. 
 Additionally as outlined in section \ref{sec:running_GIST}, various quantities are written to OpenDX files.
@@
 Gisttools provides access to single or multiple voxels through a .loc property similiar to the pandas library. 
-GIST quantity data for all voxels can be accessed by directly indexing the gist object, either by providing a single name or a list of names.
+GIST quantity data for all voxels can be accessed through directly indexing the gist object, either by providing a single name or a list of names.
 Indexing with .loc works by providing a voxel index (either a index number or list of numbers) and optionally a column index (column name or list of names).
@@
 \subsection{Applying GIST for the Streptavidin-Biotin complex}
-In this section we will apply GIST to the full complex and discuss each step of the process in more detail to outline potential pitfalls.
+In this section we will apply GIST to the full complex and discuss each step of the process in more detail to outline pitfalls.
 We will afterwards introduce advanced analysis workflows which allow you to modify and better visualize the resulting data.
@@
 systems.
-By using restraints, we can ensure that the monomer conformations match the complex.
+By using restraints, we can make sure that the monomer conformations match the complex.
 
@@
 To test whether your grid is large enough look at solvent properties close to the grid edges, e.g. the solvent-solvent energies.
-If these correspond to the respective bulk properties, it can be assumed that your grid is large enough to fully capture the perturbation of the solvent by the solute.
+If these correspond to the respective bulk properties, it can be assumed that your grid is large enough to capture all perturbations of the solute on the solvent.
 Note that a different box-size or number of solvent molecules can skew the bulk properties slightly compared to the reference values provided above.
 In that case, checking for convergence of the solvent properties near the grid borders is sufficient but you might want to modify your reference values according to your results then as described in the next section.
-
-For the GIST analysis, we recommend using either the PME or GPU implementation of GIST.
-For the GPU version, cpptraj needs to be compiled with the \inlinecode{-cuda} flag.
-While the GPU version is generally faster, the PME implementation matches the energy of the Amber MD engine more closely.
 
@@
 For typical systems, this implementation is slightly slower than GPU-GIST, but much faster than the original CPU code.
-Furthermore, PME-GIST offers the best agreement between GIST energies and those observed in an MD simulation.
+Furthermore, PME-GIST offers the best agreement between GIST energies and those observed in an classical MD simulation.
 
@@
 For typical use cases, we recommend using \inlinecode{Eww} and \inlinecode{Esw} over the \inlinecode{PME} column.
+
+\subsection{MPI parallelization}
+The most recent addition to GIST is the MPI parallelization of both energy and entropy calculations\cite{Roe2023-mpi-gist}.
+Since the MPI parallelization is orthogonal to the other improvements, it can be used with both PME or GPU accelerated GIST.
 \subsection{GIST with non-water solvents}
-In References~\cite{Kraml2020,Kamenik2020-gist-macrocycles,Waibl2022-gist-solvents}, an extension of GIST towards solvents other than water is described.
+In references~\cite{Kraml2020,Kamenik2020-gist-macrocycles,Waibl2022-gist-solvents}, an extension of GIST towards solvents other than water is described.
 
@@
 The current implementation of the entropy calculation in \software{cpptraj} requires at least 3 atoms in the main solvent.
-To compute the first-order entropy of the ions, as well as an approximate second order entropy, Python code is available on \url{https://github.com/liedllab/second-disorder}.
+Note that the calculation of the reference value is complicated by the introduction of additional solvents.
+For an overview of how to correctly handle such calculations, refer to reference \cite{Waibl2021-gist-salt}.
+Additionally, to compute the first-order entropy of the ions, as well as an approximate second order entropy, Python code is available on \url{https://github.com/liedllab/second-disorder}.
 
@@
 If cpptraj was compiled with the '-cuda' configure flag set, the GPU accelerated version of GIST is used automatically. 
-Furthermore, if cpptraj was compiled with OpenMP support, the entropy calculation will be parallelized.
+In addition, OpenMP and MPI parallelization are available if cpptraj was compiled the necessary options.
 

### 2024-06-11 manuscript/manuscript.tex
Add AmberTools citations, Update Amber citation to 2024

@@
 We highlight potential pitfalls and show how to avoid technical difficulties.
-We assume familiarity with molecular dynamics simulations as well as the AmberTools package.
+We assume familiarity with molecular dynamics simulations as well as the AmberTools  package \cite{Case2023-ambertools}.
 
@@
 Tutorials requiring specific software packages must provide instructions and files for the referenced version of the software.
-The main implementation of GIST is part of the MD analysis software \software{cpptraj} released with AmberTools \cite{amber22}.
+The main implementation of GIST is part of the MD analysis software \software{cpptraj} released with AmberTools \cite{amber24, Case2023-ambertools}.
 
@@
 In that case, other MD engines can be used as long as topologies and trajectories can be produced in a format supported by \software{cpptraj}.
-If using GROMACS \cite{Abraham2015-gromacs,Pronk2013-gromacs}, we propose to prepare the topology with AmberTools and convert it using \software{acpype.py} \cite{Sousa_da_Silva2012-acpype}.
+If using GROMACS \cite{Abraham2015-gromacs,Pronk2013-gromacs}, we propose to prepare the topology with AmberTools \cite{Case2023-ambertools} and convert it using \software{acpype.py} \cite{Sousa_da_Silva2012-acpype}.
 %We should actually test that!

### 2024-06-11 manuscript/manuscript.tex
Unify name of 'solute molecule' (from molecule, solute, compound). Small molecule is changed to small molecule solutes to keep in line with medchem naming conventions.

@@
 \begin{abstract}
-Grid inhomogeneous solvation theory (GIST) is a method to compute the free energy of solvation of a compound on a 3-dimensional grid based on molecular dynamics (MD) simulations.
+Grid inhomogeneous solvation theory (GIST) is a method to compute the free energy of solvation of a solute molecule on a 3-dimensional grid based on molecular dynamics (MD) simulations.
 The high spatial resolution of the GIST output, as well as the decomposition into energy and entropy contributions, allow for highly detailed analyses on both proteins and small molecules.
@@
 Interaction with water are of special importance as many biological processes occur in aqueous environment \cite{Privalov2017-water-review}.
-The free energy of solvation is defined as the reversible work needed to insert a solute into the solvent \cite{ben-naim-book}.
+The free energy of solvation is defined as the reversible work needed to insert a solute molecule into the solvent \cite{ben-naim-book}.
 It is directly linked to the solubility of gases and drives many real-world effects such as hydrophobicity, ligand binding, and protein folding.
@@
 Grid inhomogeneous solvation theory (GIST) is a method to compute the free energy of solvation, with applications in the context of drug design and biophysical property description.
-It estimates the solvent density from an MD trajectory and computes localized entropic and enthalpic contributions on a grid around the solute.
-The entropy is computed from the density distribution of the solvent around the solute, in a first-order approximation.
+It estimates the solvent density from an MD trajectory and computes localized entropic and enthalpic contributions on a grid around the solute molecule.
+The entropy is computed from the density distribution of the solvent around the solute molecule, in a first-order approximation.
 A schematic of this is shown in Figure~\ref{fig:carbazole}.
-Nevertheless, solvation free energies of small molecules can be computed at high accuracy using linear scaling factors for the higher-order terms, such as $\Delta S = 0.6\, \Delta S^\text{1st-order}$ for water \cite{Chen2021,Waibl2022-gist-solvents}.
+Nevertheless, solvation free energies of small molecule solutes can be computed at high accuracy using linear scaling factors for the higher-order terms, such as $\Delta S = 0.6\, \Delta S^\text{1st-order}$ for water \cite{Chen2021,Waibl2022-gist-solvents}.
 
@@
     \includegraphics[width=\linewidth]{figures/carbazole-figure.png}
-    \caption{Schematic showing regions of high water density around a carbazole molecule.
-The energy is computed from the intermolecular interactions of a force field, while the entropy is approximated from the 3-dimensional solvent distribution around the solute.}
+    \caption{Schematic showing regions of high water density around a carbazole solute molecule.
+The energy is computed from the intermolecular interactions of a force field, while the entropy is approximated from the 3-dimensional solvent distribution around the solute molecule.}
     \label{fig:carbazole}
@@
 Data-driven methods include QSPR and UNIFAC \cite{Borhani2019-qspr,Fredenslund1975-unifac}.
-Polarizable continuum models (PCMs) \cite{Miertus1981-pcm,Klamt1993-cosmo} treat the solvent as a homogeneous phase that reacts to the electrostatic potential of the solute \cite{Mennucci2010-pcm}.
+Polarizable continuum models (PCMs) \cite{Miertus1981-pcm,Klamt1993-cosmo} treat the solvent as a homogeneous phase that reacts to the electrostatic potential of the solute molecule \cite{Mennucci2010-pcm}.
 %Molecular mechanics based approaches describe the underlying PES through empirically fit force-fields, which reduce the computational demand of these calculations tremendously.
 %This allows for the treatment of large biological systems such as proteins in aqueous solution.
-Methods such as MM/PBSA or MM/GBSA \cite{Sitkoff1994-pbsa,Kollman2000-mmpbsa} combine an implicit treatment of solvent electrostatics with structural sampling of the solute.
+Methods such as MM/PBSA or MM/GBSA \cite{Sitkoff1994-pbsa,Kollman2000-mmpbsa} combine an implicit treatment of solvent electrostatics with structural sampling of the solute molecule.
 They can quickly provide results for a large number of structures \cite{Genheden2015-mmpbsa-review}.
@@
 Therefore, IST-based approaches can only be used with sufficiently rigid solvents.
-Nevertheless, they have been applied on a wide variety of problems involving both small molecules and biological systems.
+Nevertheless, they have been applied on a wide variety of problems involving both small molecules and biological systems as solute molecules.
 
@@
 On the other hand, the first-order approximation means that the solvation entropy is less accurately described than in TI or FEP \cite{Chen2021,Waibl2021-gist-salt}.
-%The 3D-RISM approach allows for a decomposition into entropic and enthalpic terms as well as a full localization of the underlying distributions to the solute structure.
+%The 3D-RISM approach allows for a decomposition into entropic and enthalpic terms as well as a full localization of the underlying distributions to the solute molecule structure.
 %Here you would explain what problem you are tackling and briefly motivate your work.
@@
 The purpose of this tutorial is to help users run GIST calculations and interpret the results.
-The tutorial is structured around two examples: a GIST calculation for the small molecule biotin, and a study of biotin-streptavidin binding.
+The tutorial is structured around two examples: a GIST calculation for the small molecule solute biotin, and a study of biotin-streptavidin binding.
 We provide a GIST workflow that can be easily adapted to different systems, an introduction to the method and the theory behind it, and an overview of the different implementations of GIST.
@@
 	\item \software{cpptraj}.
-	\item A topology of the molecule of interest in a solvent box.
+	\item A topology of the solute molecule of interest in a solvent box.
 	\item An MD trajectory of that system.
@@
 
-Often, the region of interest will be defined as the surrounding of a molecule, which might for example be a ligand in the binding pocket or the whole protein.
+Often, the region of interest will be defined as the surrounding of a solute molecule, which might for example be a ligand in the binding pocket or the whole protein.
 Using \inlinecode{cpptraj's } \inlinecode{bound} command, the region of interest can be output to a file.
@@
 
-Note that the solute should not move significantly for a GIST calculation, since IST does not take the solute degrees of freedom into account.
+Note that the solute molecule should not move significantly for a GIST calculation, since IST does not take the solute molecule's degrees of freedom into account.
 This is typically achieved using restraints on the solute heavy atoms.
@@
 \begin{table}[h]
-	\caption{Reference density $\rho_0$ and mean water-water energy $E_{ww, norm}$ for various water models from NpT simulations of N$_{WAT}$ water molecules at \SI{300}{\kelvin} and 1bar.}\label{tab:ref_densities}
+	\caption{Reference density $\rho_0$ and mean water-water energy $E_{ww, norm}$ for various water models from NpT simulations of N$_{WAT}$ water solvent molecules at \SI{300}{\kelvin} and 1bar.}\label{tab:ref_densities}
 	\small
@@
 \begin{enumerate}
-	\item Set up and run GIST calculations for a small molecule (biotin) and a protein (streptavidin) binding pocket.
+	\item Set up and run GIST calculations for a small molecule solute (biotin) and a protein (streptavidin) binding pocket.
 	\item Compute the free energy of solvation for biotin using GIST.
@@
 
-In the first part, we show how to run GIST calculations for small molecules and interpret the output.
+In the first part, we show how to run GIST calculations for small molecule solutes and interpret the output.
 Afterwards, we apply this knowledge to a protein binding pocket, visualize $\dgsolv$ hotspots and estimate a free energy of ligand binding using an end-state approach based on GIST.
@@
 \subsection{Running GIST for Biotin}
-In this example, we will be running a simple GIST calculation of the biotin molecule to get familiar with the usage of GIST in \software{cpptraj}.
+In this example, we will be running a simple GIST calculation of the biotin solute molecule to get familiar with the usage of GIST in \software{cpptraj}.
 \subsubsection{System preparation}
 First, extract biotin from the crystal structure and generate a topology and coordinate file compatible with Amber or a simulation package of your choice.
-The molecule should be solvated with a sufficiently large solvent box.
+The solute molecule should be solvated with a sufficiently large solvent box.
 Carefully check the employed force field, protonation state and partial charges.
@@
 Solvate the parametrized biotin, making sure to choose a water box of large enough size so that the water on the box edges behaves bulk-like (a wall distance of at least 15\AA is recommended).
-A good starting point (though a bit dated) for the parametrization and solvation of small molecules is the Amber tutorial 4b (\url{http://ambermd.org/tutorials/basic/tutorial4b/}).
+A good starting point (though a bit dated) for the parametrization and solvation of small molecule solutes is the Amber tutorial 4b (\url{http://ambermd.org/tutorials/basic/tutorial4b/}).
 
@@
 Now, you can run a restrained MD simulation based on the equilibrated structure of biotin for the subsequent GIST analysis.
-Restraints should be used during the MD because IST theory is not defined for moving solutes.
+Restraints should be used during the MD because IST theory is not defined for moving solute molecules.
 We apply a harmonic restraint of \SI{100}{\kilo\calorie\per\mole\per\angstrom\squared} to all heavy atoms to keep the system in place.
 We run \SI{100}{\nano\second} of simulation and store the conformations every \SI{100}{\pico\second}.
-A simulation length of at least \SI{20}{\nano\second} is recommended to obtain statistically independent snapshots of the water molecules.
+A simulation length of at least \SI{20}{\nano\second} is recommended to obtain statistically independent snapshots of the water solvent molecules.
 
@@
 To start our GIST analysis of biotin, we'll first have to decide on the parameters defining the GIST grid during the analyis.
-We recommend that you center the biotin molecule to the coordinate origin:
+We recommend that you center the biotin solute molecule to the coordinate origin:
 \begin{lstlisting}[style=cpptraj]
@@
 \end{lstlisting}
-To decide on a grid size, find the extent (minimum and maximum) of the x-, \mbox{y-,} and z-coordinates of the molecule.
+To decide on a grid size, find the extent (minimum and maximum) of the x-, \mbox{y-,} and z-coordinates of the solute molecule.
 To get bulk-like water properties at the grid border, a buffer distance of \SI{15.0}{\angstrom} is reasonable for biotin.
@@
 Now, you can finally run the actual GIST analysis.
-Center the molecule as above, to make sure that its position matches the grid parameters you decided on.
+Center the solute molecule as above, to make sure that its position matches the grid parameters you decided on.
 We recommend using either the GPU or MPI-accelerated PME implementation of GIST.
@@
 Note that, due to the restraints, some solvent sites might be inaccessible to the bulk solvent.
-To investigate such positions, make sure to explicitly place water molecules there.
+To investigate such positions, make sure to explicitly place water solvent molecules there.
 It is good practice to check your restrained MD simulation or solvent populations obtained by GIST to confirm whether all relevant sites were properly solvated.
@@
 Use the \software{equilibration.py} script (or your own procedure) to perform short NVT and NpT equilibration runs on the complex structure.
-Note that the equilibration script contains unrestrained NpT steps, because restraining two molecules at the same time might interfere with the pressure equilibration.
+Note that the equilibration script contains unrestrained NpT steps, because restraining two solute molecules at the same time might interfere with the pressure equilibration.
 
@@
 \subsubsection{GIST analysis of Streptavidin-Biotin}
-For easier post-processing, here we use grids that contain the whole molecule.
+For easier post-processing, here we use grids that contain the whole solute molecule.
 Smaller grids encompassing only the binding site could be used for streptavidin and the complex, but this would complicate the analysis.
-Furthermore, we recommend that you center the molecule to the coordinate origin.
+Furthermore, we recommend that you center the solute molecule to the coordinate origin.
 Calculate the GIST grid parameters as outlined in the biotin section for the complex and streptavidin and run the GIST analysis.
-For larger systems like whole proteins you can usually get away with a smaller buffer region compared to small molecules.
+For larger systems like whole proteins you can usually get away with a smaller buffer region compared to small molecule solutes.
 To test whether your grid is large enough look at solvent properties close to the grid edges, e.g. the solvent-solvent energies.
-If these correspond to the respective bulk properties, it can be assumed that your grid is large enough to capture all perturbations of the solute on the solvent.
+If these correspond to the respective bulk properties, it can be assumed that your grid is large enough to capture all perturbations of the solute molecule on the solvent.
 Note that a different box-size or number of solvent molecules can skew the bulk properties slightly compared to the reference values provided above.
@@
 
-The most accurate method to compute a reference energy is to run a GIST calculation of the pure solvent using the same energy method (PME/GPU) and a similar box size, and then compute the average solvent-solvent energy per molecule.
+The most accurate method to compute a reference energy is to run a GIST calculation of the pure solvent using the same energy method (PME/GPU) and a similar box size, and then compute the average solvent-solvent energy per solvent molecule.
 
@@
 
-Alternatively, the solvent-solvent energy in a large GIST grid converges to the bulk value at sufficient distance to the solute.
-Therefore, the reference energy can be obtained by binning the voxels by their distance to the solute, and evaluate Equation~\ref{eq_ewwref} within each bin.
+Alternatively, the solvent-solvent energy in a large GIST grid converges to the bulk value at sufficient distance to the solute molecule.
+Therefore, the reference energy can be obtained by binning the voxels by their distance to the solute molecule, and evaluate Equation~\ref{eq_ewwref} within each bin.
 If this value converges, this can be used as $E_{ww}^\textit{ref}$.
@@
 	\includegraphics[width=0.8\linewidth]{figures/Eww_convergence.eps}
-	\caption{Convergence of $E_{ww}$ in the complex calculation with increasing distance to the solute (streptavidin-biotin).
+	\caption{Convergence of $E_{ww}$ in the complex calculation with increasing distance to the solute molecule (streptavidin-biotin).
 The horizontal line shows the automatically computed reference energy.}\label{fig_ewwref}
@@
 %In \software{gistpp}, use the \inlinecode{add} command, with the same syntax as the \inlinecode{div} command used above.
-After that, check whether your free energy contributions become negligible far away from the solute.
+After that, check whether your free energy contributions become negligible far away from the solute molecule.
 In a plot like Figure~\ref{fig_ewwref} based on the \inlinecode{A\_dens} column, the values should tend to zero.
-However, it is more informative to plot the cumulative free energy contribution against the distance to the solute, to check the convergence.
+However, it is more informative to plot the cumulative free energy contribution against the distance to the solute molecule, to check the convergence.
 It the curve flattens out, the converged value is your final $\dgsolv$\@.
@@
 	\includegraphics[width=0.8\linewidth]{figures/A_E_S_convergence.eps}
-	\caption{Convergence of $\dgsolv$ and its contributions with increasing distance to the solute.
+	\caption{Convergence of $\dgsolv$ and its contributions with increasing distance to the solute molecule.
 All quantities cumulative, i.e., summed up to the respective radius.
@@
 
-It might also be interesting to visualize the average energy of a water molecule at each grid voxel.
+It might also be interesting to visualize the average energy of a water solvent molecule at each grid voxel.
 This quantity is given by $2E_{ww} + E_{sw}$.
@@
 Even though streptavidin-biotin is known to be a very stable complex, the solvation energy favors the dissociation.
-This is expected: if the molecules were not able to form any interactions, but interacted fully with the water, they would not form a complex.
+This is expected: if the solute molecules were not able to form any strong interactions, but interacted favorably with the water instead, they would not form a complex.
 
@@
 
-To speed up the calculation, smaller GIST grids could be used by focusing on the binding pocket, or by rotating the molecule by its principal axes to fit the cuboid grid more exactly.
+To speed up the calculation, smaller GIST grids could be used by focusing on the binding pocket, or by rotating the solute molecule by its principal axes to fit the cuboid grid more exactly.
 However, this should be done \emph{before} the MD, since rotating the trajectory damages the periodic box information.
@@
 The output will contain two additional columns: \inlinecode{PME\_dens} and \inlinecode{PME\_norm}.
-They contain the potential of solute and solvent evaluated at the solvent positions and divided by two.
+They contain the potential of solute molecule and solvent evaluated at the solvent positions and divided by two.
 Furthermore, the \inlinecode{Eww} and \inlinecode{Esw} columns are also computed using PME, and are reported as usual.
@@
 
-To run a GIST calculation in a solvent other than water, use \inlinecode{solute} with a \software{cpptraj} selection mask.
+To run a GIST calculation in a solvent other than water, use \inlinecode{solute} with a \software{cpptraj} selection mask to select the solute molecule.
 All other molecules will be treated as solvent.
@@
 
-The code will automatically choose three atoms per molecule to define the solvent orientation, and print this selection to the output.
+The code will automatically choose three atoms per solvent species to define the solvent orientation, and print this selection to the output.
 Sometimes, however, the automatic selection is not optimal.
@@
 This allows treating salting-out effects, although it was shown that the salting-out coefficient is over-predicted by GIST because of the first-order entropy approximation.
-Generally, GIST should be able to treat arbitrary solvent mixtures as long as each molecule is sufficiently rigid.
-
-In the \software{cpptraj} implementation of GIST, the \inlinecode{solute} keyword specifies which components of the system are treated as a solute.
+Generally, GIST should be able to treat arbitrary solvent mixtures as long as each solvent molecule is sufficiently rigid.
+
+In the \software{cpptraj} implementation of GIST, the \inlinecode{solute} keyword specifies which components of the system are treated as a solute molecule.
 Everything else will be treated as solvent.
 If the solvent contains more than one molecular species, a list of solvent molecule names (i.e., ``\inlinecode{solventmols WAT,NA,CL}'') must be specified.
-For each molecule in this list, densities (e.g., \inlinecode{g\_mol\_NA}) and energies (e.g., \inlinecode{Eww\_mol\_NA}) will be computed and written to .dx files.
-Note that, e.g., \inlinecode{Esw\_mol\_NA} contains the interactions of every ``\inlinecode{NA}'' molecule with the solute, and \inlinecode{Esw\_mol\_NA} contains the interactions among ``\inlinecode{NA}'' molecules as well as their interactions with all other solvents, divided by 2 to account for double counting.
-Furthermore, the first molecule in this list will be treated as the main solvent, for which first-order entropy values will be computed.
+For each solvent molecule in this list, densities (e.g., \inlinecode{g\_mol\_NA}) and energies (e.g., \inlinecode{Eww\_mol\_NA}) will be computed and written to .dx files.
+Note that, e.g., \inlinecode{Esw\_mol\_NA} contains the interactions of every ``\inlinecode{NA}'' solvent molecule with the solute molecule, and \inlinecode{Esw\_mol\_NA} contains the interactions among ``\inlinecode{NA}'' solvent molecules as well as their interactions with all other solvents, divided by 2 to account for double counting.
+Furthermore, the first solvent molecule in this list will be treated as the main solvent, for which first-order entropy values will be computed.
 
@@
 
-A more sophisticated visualization, including the energy, entropy, and free energy of removing a single water molecule, is found in Figure~\ref{fig-binding-pocket.png}
+A more sophisticated visualization, including the energy, entropy, and free energy of removing a single water solvent molecule, is found in Figure~\ref{fig-binding-pocket.png}
 
@@
 
-The solvent accessible surface (SASA) represents the interface between regions that are occupied by the solute, and those which can be accessed by the solvent.
+The solvent accessible surface (SASA) represents the interface between regions that are occupied by the solute molecule, and those which can be accessed by the solvent.
 GIST provides a very natural way of creating a SASA using an isosurface of the oxygen density gO.
@@
 
-For example, voxels that are part of the binding pocket can be defined using a ligand molecule and a distance criteria.
+For example, voxels that are part of the binding pocket can be defined using a ligand solute molecule and a distance criteria.
 
@@
 %$\rho^\textit{o}$ is the bulk number density, $g_\textit{sw}\left(\textbf{r},\omega \right)$ is the solute-water 
-%pair correlation function in the solute's frame of reference, \textbf{r} is the coordinates of the oxygen atom of the water molecule,  
+%pair correlation function in the solute's frame of reference, \textbf{r} is the coordinates of the oxygen atom of the water solvent molecule,  
 %$\omega$ represents the Euler angles, and $\frac{1}{8\pi^2}$ is the volume of the orientational space, which acts as a normalization constant.
@@
 
-IST expresses the free energy of solvation in terms of the solvent density in a coordinate system defined by the solute.
-The solute is kept fixed in space, following the standard solvation process established by Ben-Naim \cite{ben-naim-book}.
+IST expresses the free energy of solvation in terms of the solvent density in a coordinate system defined by the solute molecule.
+The solute molecule is kept fixed in space, following the standard solvation process established by Ben-Naim \cite{ben-naim-book}.
 
@@
 In a slightly simplified view, the first order is the solvent density $g_\textit{sw}\left(\textbf{r},\omega \right)$ at each position $\textbf{r}$ and orientation $\omega$.
-The first-order entropy omits all solvent-solvent (and higher) correlations, while the solute-solvent correlation is taken into account because the coordinates are relative to the solute.
+The first-order entropy omits all solvent-solvent (and higher) correlations, while the solute-solvent correlation is taken into account because the coordinates are relative to the solute molecule.
 Thus, it is called the solute-water entropy $S_{sw}$.
@@
 The solvent density is expressed relative to the bulk density $\rho^0$. %$g_\textit{sw}\left(\textbf{r},\omega \right)$ is the solute-water 
-%pair correlation function in the solute's frame of reference, \textbf{r} is the coordinates of the oxygen atom of the water molecule,  
+%pair correlation function in the solute's frame of reference, \textbf{r} is the coordinates of the oxygen atom of the water solvent molecule,  
 %$\omega$ represents the Euler angles, and 
 The entropy integral is also normalized by $8\pi^2$, which is the volume of the orientational space.
-In bulk or at a high distance to the solute, the quantity $g_\textit{sw}\left(\textbf{r},\omega \right)$ approaches one, which leads to zero first-order entropy. 
+In bulk or at a high distance to the solute molecule, the quantity $g_\textit{sw}\left(\textbf{r},\omega \right)$ approaches one, which leads to zero first-order entropy. 
 Therefore, no reference entropy is needed unless there is a numeric bias, e.g., when the MD frames are statistically dependent because the time between them is insufficient.
@@
 \end{equation}
-where $N_\textit{k}$ is the number of water molecules found in voxel k, $\gamma$ is Euler's constant accounting for the bias of the naive entropy estimator, and $g_\textit{NN}$ is the nearest neighbor estimate of the density.
+where $N_\textit{k}$ is the number of water solvent molecules found in voxel k, $\gamma$ is Euler's constant accounting for the bias of the naive entropy estimator, and $g_\textit{NN}$ is the nearest neighbor estimate of the density.
 
@@
 
-The solute-solvent energy $E_{sw}$ can be expressed in terms of the solvent density and the potential $U_{sw}(\textbf{r},\omega)$ induced by the solute.
+The solute-solvent energy $E_{sw}$ can be expressed in terms of the solvent density and the potential $U_{sw}(\textbf{r},\omega)$ induced by the solute molecule.
 In practice, it is evaluated as the expectation value $\langle\cdot\rangle$ of the pairwise force field energy $U_{ij}$ between all solvent molecules in voxel $k$ and all solute atoms.
@@
 To localize $E_{sw}$ on the three-dimensional grid, every energy term is assigned to the voxel $k$ holding the solvent, and the expectation value in Equation~\ref{eq-esw} is computed per voxel.
-Similar to the entropy integrals, the solute-water solvation energy decays with increasing distance to the solute.
+Similar to the entropy integrals, the solute-water solvation energy decays with increasing distance to the solute molecule.
 Hence, it can be approximated by local spatial integrals.
@@
 In contrast to $E_{sw}$, $E_{ww}$ does not tend to zero in bulk.
-Therefore, a reference corresponding to the average energy of a bulk water molecule needs to be subtracted.
+Therefore, a reference corresponding to the average energy of a bulk water solvent molecule needs to be subtracted.
 The referenced solvent-solvent energy will be denoted as $E_{ww}^\text{corr}$.
@@
 
-However, when water is replaced from a small region $R$ of interest, such as a single water molecule, almost all interactions are with water molecules outside of this region.
+However, when water is replaced from a small region $R$ of interest, such as a single water solvent molecule, almost all interactions are with water solvent molecules outside of this region.
 Therefore, there is no double counting, and the full solvent-solvent energy should be used.
@@
 
-When the region $R$ comprises more than one molecule, interactions within this region are double-counted while interactions to the outside are not.
+When the region $R$ comprises more than one solvent molecule, interactions within this region are double-counted while interactions to the outside are not.
 This could be solved using the solvent-solvent energy between each pair of voxels $E_{ww,k\;l}$:
@@
 dTSorient & X & First order orientational entropy density . \\
-neighbor & X & Mean number of waters neighboring the water molecules found in this voxel multiplied by the voxel number density. \\
+neighbor & X & Mean number of waters neighboring the water solvent molecules found in this voxel multiplied by the voxel number density. \\
 dipole & X & Magnitude of mean dipole moment (polarization). \\

### 2024-06-11 manuscript/manuscript.tex

@@
 GIST is an implementation of IST on a three-dimensional grid.
-It can be seen as a more general version of WaterMap \cite{Abel2008-watermap,Young2007-watermap} that provides solvation thermodynamics on the entire grid, rather than high-density clusters in a binding pocket.
+It can be seen as a more general version of WaterMap \cite{Abel2008-watermap,Young2007-watermap} that provides solvation thermodynamics on a space filling grid, rather than high-density clusters in a binding pocket.
 This implies a higher versatility, but also puts more responsibility on the user in terms of correct post-processing.
@@
 The user should be able to run MD simulations and analyze the resulting trajectories.
-Previous experience of GIST is not necessary, although we recommend reading the works referenced in section~\ref{sec:theory}.
+Previous experience of GIST is not necessary, although we recommend reading the articles referenced in section~\ref{sec:theory}.
 
@@
 The default value of \SI{0.5}{\angstrom} provides reasonable discretization for many use-cases.
-While integrals of GIST quantities over the grid are independent of the voxel size, individual values converge slower at high resolutions.
+While integrals of GIST quantities over the grid are independent of the voxel size, values for each voxel converge slower at high resolutions.
 
@@
 In the first part, we show how to run GIST calculations for small molecule solutes and interpret the output.
-Afterwards, we apply this knowledge to a protein binding pocket, visualize $\dgsolv$ hotspots and estimate a free energy of ligand binding using an end-state approach based on GIST.
+Afterwards, we apply this knowledge to a protein binding pocket, visualize $\dgsolv$ hotspots and estimate a free energy of ligand binding using an end-states approach based on GIST.
 In addition, we discuss all steps in further detail at the end.
@@
 Now, you can finally run the actual GIST analysis.
-Center the solute molecule as above, to make sure that its position matches the grid parameters you decided on.
+
+Center the solute molecule as above, to verify that its position matches the grid parameters you decided on.
 We recommend using either the GPU or MPI-accelerated PME implementation of GIST.
@@
 To test whether your grid is large enough look at solvent properties close to the grid edges, e.g. the solvent-solvent energies.
-If these correspond to the respective bulk properties, it can be assumed that your grid is large enough to capture all perturbations of the solute molecule on the solvent.
+
+If these correspond to the respective bulk properties, it can be assumed that your grid is large enough to fully capture the perturbation of the solvent by the solute molecule.
+
 Note that a different box-size or number of solvent molecules can skew the bulk properties slightly compared to the reference values provided above.
@@
 
-The most accurate method to compute a reference energy is to run a GIST calculation of the pure solvent using the same energy method (PME/GPU) and a similar box size, and then compute the average solvent-solvent energy per solvent molecule.
+The most accurate method to compute a reference energy is to run a GIST calculation of the pure solvent using the same energy method (PME/GPU), a similar box size, at the same temperature and pressure and then compute the average solvent-solvent energy per solvent molecule.
 

### 2024-06-11 manuscript/manuscript.tex

--- 2024-06-11 manuscript/manuscript.tex
@@
 
-In this tutorial, we aim to guide the reader through the most common steps involved in a GIST analysis at the example of the streptavidin-biotin complex.
+In this tutorial, we aim to guide the reader through the most common steps involved in a GIST analysis using the streptavidin-biotin complex as an demonstrative system.
 Furthermore, we discuss the theory of GIST with a focus on practical aspects.
 We highlight potential pitfalls and show how to avoid technical difficulties.
-We assume familiarity with molecular dynamics simulations as well as the AmberTools  package \cite{Case2023-ambertools}.
+We assume familiarity with molecular dynamics simulations as well as the AmberTools package.
 
@@
 Tutorials requiring specific software packages must provide instructions and files for the referenced version of the software.
-The main implementation of GIST is part of the MD analysis software \software{cpptraj} released with AmberTools \cite{amber24, Case2023-ambertools}.
+The main implementation of GIST is part of the MD analysis software \software{cpptraj} released with AmberTools \cite{amber22}.
 
@@
 The Amber simulation engine is used in this tutorial to run the MD simulations.
-While Amber is free for non-commercial use, it might not always be available in industry settings.
+While Amber is free for non-commercial use, it might not always be available in industrial settings.
 In that case, other MD engines can be used as long as topologies and trajectories can be produced in a format supported by \software{cpptraj}.
-If using GROMACS \cite{Abraham2015-gromacs,Pronk2013-gromacs}, we propose to prepare the topology with AmberTools \cite{Case2023-ambertools} and convert it using \software{acpype.py} \cite{Sousa_da_Silva2012-acpype}.
+If using GROMACS \cite{Abraham2015-gromacs,Pronk2013-gromacs}, we propose to prepare the topology with AmberTools and convert it using \software{acpype.py} \cite{Sousa_da_Silva2012-acpype}.
 %We should actually test that!
@@
 Now, you can finally run the actual GIST analysis.
-Center the molecule as above, to make sure that its position matches the grid parameters you decided on.
+
+Center the molecule as above, to verify that its position matches the grid parameters you decided on.
 We recommend using either the GPU or MPI-accelerated PME implementation of GIST.

### 2024-06-12 manuscript/manuscript.tex

--- 2024-06-11 manuscript/manuscript.tex
@@
 Solvation thermodynamics govern many biological and chemical processes involving, e.g., aqueous solutions, hydrophobic effects, or partitioning between different environments.
+
 Interaction with water are of special importance as many biological processes occur in aqueous environment \cite{Privalov2017-water-review}.
@@
 
-Grid inhomogeneous solvation theory (GIST) is a method to compute the free energy of solvation, with applications in the context of drug design and biophysical property description.
+Grid inhomogeneous solvation theory (GIST) \cite{Nguyen2012} is a method to compute the free energy of solvation, with applications in the context of drug design and biophysical property description.
 It estimates the solvent density from an MD trajectory and computes localized entropic and enthalpic contributions on a grid around the solute molecule.
-The entropy is computed from the density distribution of the solvent around the solute molecule, in a first-order approximation.
-A schematic of this is shown in Figure~\ref{fig:carbazole}.
+Different from most other methods, these contributions are therefore spatially resolved, improving interpretability and allowing better insights into the nature of solvation around the investigated solute.
+The entropy is computed from the density distribution of the solvent around the solute molecule, in a first-order approximation. A schematic of this is shown in Figure~\ref{fig:carbazole}. 
+The calculation of higher-order entropy terms is computationally demanding but was realized in recent works \cite{Nguyen2016-gist-second-order, Waibl2021-gist-salt}.
 Nevertheless, solvation free energies of small molecule solutes can be computed at high accuracy using linear scaling factors for the higher-order terms, such as $\Delta S = 0.6\, \Delta S^\text{1st-order}$ for water \cite{Chen2021,Waibl2022-gist-solvents}.
@@
 GIST is an implementation of IST on a three-dimensional grid.
-It can be seen as a more general version of WaterMap \cite{Abel2008-watermap,Young2007-watermap} that provides solvation thermodynamics on a space filling grid, rather than high-density clusters in a binding pocket.
+Unlike the related method WaterMap \cite{Abel2008-watermap,Young2007-watermap}, it provides solvation thermodynamics on an entire grid, rather than only high-density clusters in a binding pocket.
 This implies a higher versatility, but also puts more responsibility on the user in terms of correct post-processing.
 
-Since the density is computed from an MD simulation, GIST can accurately treat confined water regions, which are challenging for implicit-solvent methods such as GBSA and self-consistent theories such as 3D-RISM \cite{Kovalenko1998-3drism}.
-While the need for MD simulations comes at a large computational cost, GIST is still more efficient than end-states approaches such as TI \cite{Kirkwood1935-ti} or FEP \cite{Zwanzig1954-reweighting}, with applications to large systems such as serine proteases \cite{Kraml2019-gigist} or antibody fragments \cite{Waibl2021-gist-antibodies}.
+Since the density is computed from an explicit solvent MD simulation, GIST provides an accurate picture of solvation at an atomistic level, even for demanding systems such as drug binding sites in proteins.
+While the need for MD simulations comes at a large computational cost, GIST is still more efficient than end-state approaches such as TI \cite{Kirkwood1935-ti} or FEP \cite{Zwanzig1954-reweighting}, with applications to large systems such as serine proteases \cite{Kraml2019-gigist} or antibody fragments \cite{Waibl2021-gist-antibodies}.
 On the other hand, the first-order approximation means that the solvation entropy is less accurately described than in TI or FEP \cite{Chen2021,Waibl2021-gist-salt}. 
-%The 3D-RISM approach allows for a decomposition into entropic and enthalpic terms as well as a full localization of the underlying distributions to the solute molecule structure.
-%Here you would explain what problem you are tackling and briefly motivate your work.
-%
-%In this particular template, we have removed most of the usage examples which occur in \texttt{sample-document.tex} to provide a minimal template you can modify; however, we retain a couple of examples illustrating more unusual features of our templates/article class, such as the checklists, and information on algorithms and pseudocode.
-%
-%Keep in mind, as you prepare your manuscript, that you should plan for a representative image  which will be used to highlight your article on the journal website and publications.
-Usually, this would be one of your figures, but it must also be uploaded separately upon article submission.
-We give specific guidelines for this image on the journal website in the section on article submission (see \url{https://livecomsjournal.github.io/authors/policies/index.html#article-submission}).
-%
-%Additionally, for well-formatted manuscripts, we recommend that you let LaTeX handle figure/table placement for you as much as possible, so please avoid specifying strenuous float instructions like `[h!]` and `[H]` as much as possible.
+Similar to the self-consistent integral theories, however, GIST only considers a single solute conformation and  multiple computations of representative structures are necessary to describe ensemble properties.
 
@@
 The purpose of this tutorial is to help users run GIST calculations and interpret the results.
-The tutorial is structured around two examples: a GIST calculation for the small molecule solute biotin, and a study of biotin-streptavidin binding.
-We provide a GIST workflow that can be easily adapted to different systems, an introduction to the method and the theory behind it, and an overview of the different implementations of GIST.
-
-We start with a basic GIST analysis at the example of biotin.
+The tutorial is structured around two examples: a GIST calculation for the small molecule biotin, and a GIST study of biotin-streptavidin binding.
+We provide a workflow that can be easily adapted to different systems, an introduction to the method and the theory behind it, and an overview of the different implementations of GIST.
+
+We start with a basic GIST analysis of biotin.
 Building on the basics, we present a more advanced GIST study of the streptavidin-biotin complex, reflecting on many common aspects of GIST studies.
@@
 
-Often, the region of interest will be defined as the surrounding of a solute molecule, which might for example be a ligand in the binding pocket or the whole protein.
-Using \inlinecode{cpptraj's } \inlinecode{bound} command, the region of interest can be output to a file.
-For example, here is the command for a region of 10\AA{} around the ligand LIG:
+Additionally, the GIST calculation will usually be preceded by preprocessing steps, such as centering the trajectory or reimaging to account for periodic boundary conditions.
+Often, the region of interest will be defined as the surrounding of a molecule, which might for example be a ligand in the binding pocket or the whole protein. 
+Using \inlinecode{cpptraj's } \inlinecode{bounds} command, the region of interest can be output to a file. For example, here is the command for a region of 10\AA{} around the ligand LIG:
+
 \begin{lstlisting}[style=cpptraj]
@@
 
-Note that the solute molecule should not move significantly for a GIST calculation, since IST does not take the solute molecule's degrees of freedom into account.
-This is typically achieved using restraints on the solute heavy atoms.
 This command tells \software{cpptraj} to run a GIST calculation using the specified region and print the output in a file called \inlinecode{gist.dat}.
-You will need to load and image your trajectory before running this command.
-An overview of how to work with \software{cpptraj} is provided in the program's documentation \cite{cpptraj_doc}, AMBER tutorials \cite{amber_tut_cpptraj} or on the AMBER-hub website \cite{amber_hub}.
+An overview of how to work with \software{cpptraj} is provided in the program's documentation \cite{cpptraj_doc}, AMBER tutorials \cite{amber_tut_cpptraj} and on the AMBER-hub website \cite{amber_hub}.
 
@@
 \begin{table}[h]
-	\caption{Reference density $\rho_0$ and mean water-water energy $E_{ww, norm}$ for various water models from NpT simulations of N$_{WAT}$ water solvent molecules at \SI{300}{\kelvin} and 1bar.}\label{tab:ref_densities}
+
+	\label{tab:ref_values}
+	\caption{Reference number density $\rho_0$ and mean water-water energy $E_{ww, norm}$ for various water models from NPT simulations of N$_{WAT}$ water solvent molecules at \SI{300}{\kelvin} and 1bar.}\label{tab:ref_densities}
 	\small
@@
 First, extract biotin from the crystal structure and generate a topology and coordinate file compatible with Amber or a simulation package of your choice.
+
 The solute molecule should be solvated with a sufficiently large solvent box. 
+For periodic boundary conditions, the size of the solvent box should be large enough for the solvent to behave bulk-like so as to reduce artifact from solvent-mediated interactions between the periodic images.
+This corresponds to a distance between solute and box boundary of at least 12\AA{} for water, though we recommend 15\AA{} or higher.
 Carefully check the employed force field, protonation state and partial charges.
@@
 Solvate the parametrized biotin, making sure to choose a water box of large enough size so that the water on the box edges behaves bulk-like (a wall distance of at least 15\AA is recommended).
-A good starting point (though a bit dated) for the parametrization and solvation of small molecule solutes is the Amber tutorial 4b (\url{http://ambermd.org/tutorials/basic/tutorial4b/}).
+A good starting point for the parametrization and solvation of small molecules is the Amber tutorial 4b (\url{http://ambermd.org/tutorials/basic/tutorial4b/}).
 
 \subsubsection{Equilibration}
-Equilibrate your solvated biotin in an NVT or NpT ensemble.
-Even when running the simulation in NVT, it is advisable to include an NpT equilibration step to adjust the box size beforehand.
+Equilibrate your solvated biotin in an NVT or NPT ensemble.
+Even when running the simulation in NVT, it is advisable to include an NPT equilibration step to adjust the box size beforehand.
 Use enough equilibration time to allow the solvent to relax. 
-For water, \SI{0.5}{\nano\second} are usually sufficient to equilibrate the density, but narrow binding pockets can take longer.
+For water, \SI{0.5}{\nano\second} is usually sufficient to equilibrate the density, but narrow binding pockets can take longer.
 
@@
 \subsubsection{GIST analysis}
-To start our GIST analysis of biotin, we'll first have to decide on the parameters defining the GIST grid during the analyis.
-We recommend that you center the biotin solute molecule to the coordinate origin:
+
+To start our GIST analysis of biotin, we'll first have to decide on the parameters defining the GIST grid during the analysis.
+We will need a solute structure without water for the analysis of our future GIST results. 
+As we ran our MD simulation with periodic boundaries, we always need to reimage all molecules to the original box using cpptraj's \inlinecode{autoimage} function. 
+We also recommend that you center the biotin solute molecule to the coordinate origin, as this simplifies the definition of our grid origin later on:
+
 \begin{lstlisting}[style=cpptraj]
@@
 trajout biotin-centered.pdb 
-\end{lstlisting}
-To decide on a grid size, find the extent (minimum and maximum) of the x-, \mbox{y-,} and z-coordinates of the solute molecule.
+run
+\end{lstlisting}
+
+In this code snippet, the first two lines load the topology and coordinates of the first frame of our MD simulation. 
+In the third line, we reimage the coordinates to the first box and center the biotin at the origin. 
+Then, we strip all water solvent molecules in the fourth line and write out a pdb file in the fith line. 
+Finally, we start the analysis with the "run" command.
+
+To decide on a grid size, find the extent (minimum and maximum) of the x-, \mbox{y-,} and z-coordinates of the molecule.
+
 To get bulk-like water properties at the grid border, a buffer distance of \SI{15.0}{\angstrom} is reasonable for biotin.
@@
 bounds !(:WAT) dx 0.5 offset 30 name MyGrid out bounds.dat
-\end{lstlisting}
-Here, we calculate the bounds around the whole protein for a grid with \SI{0.5}{\angstrom} side-lenght voxels and a buffer zone of \SI{15.0}{\angstrom}.
-Note that the offset is given in terms of number of additional voxels, not voxel dimension.
-By varying the atom mask in the \inlinecode{bound} command, it is possible to center the grid on a region of interest, similiar to the \software{FindCentroid.py} script.
-
-Now, you can finally run the actual GIST analysis.
-
-Center the solute molecule as above, to verify that its position matches the grid parameters you decided on.
+run
+\end{lstlisting}
+In the above snippet, we calculate the bounds around the whole protein for a grid with \SI{0.5}{\angstrom} side-length voxels and a buffer zone of \SI{15.0}{\angstrom}.
+Note that the offset is given in terms of number of additional voxels, not length dimension.
+By varying the atom mask in the \inlinecode{bound} command, it is possible to center the grid on a region of interest, similar to the \software{FindCentroid.py} script.
+
+Now, you can run the actual GIST analysis.
+Reimage and center the solute molecule as above, to verify that its position matches the grid parameters you decided on.
+
 We recommend using either the GPU or MPI-accelerated PME implementation of GIST.
@@
 Note that, due to the restraints, some solvent sites might be inaccessible to the bulk solvent.
-To investigate such positions, make sure to explicitly place water solvent molecules there.
+To investigate such positions, make sure to explicitly place water solvent molecules there. %TODO add MC/MD or mupt and sources
+
 It is good practice to check your restrained MD simulation or solvent populations obtained by GIST to confirm whether all relevant sites were properly solvated.
 
-When calculating solvation free energy differences between a complex and the corresponding monomers (the dissociated state), there are two options for dealing with configurational diversity in the structures.
+When calculating solvation free energy differences between a complex and the corresponding monomers (the dissociated state), there are generally two approaches for dealing with configurational diversity.
+In the first approach, the ensemble properties of each individual part and the complex are accounted for by running individual MD simulations for each.
+The second approach assumes sufficient overlap in the conformational space of the dissociated states and the complex.
+Then, only configurations of the complex are sampled in a single MD simulation of the complex.
+Calculations for the disassociated states are then run on the extracted structures from the complex configurations.
+%TODO source MMPBSA
 One can choose to sample the complex and dissociated states individually, or to sample only one of them and assume that there is sufficient overlap in the conformations.
@@
 \end{lstlisting}
-(Adapt for biotin as necessary.)
-
-Use the \software{equilibration.py} script (or your own procedure) to perform short NVT and NpT equilibration runs on the complex structure.
-Note that the equilibration script contains unrestrained NpT steps, because restraining two solute molecules at the same time might interfere with the pressure equilibration.
-
-Then split the equilibrated structure into biotin and streptavidin parts using \software{cpptraj\_remove\_mol.sh}.
+Adapt the above input file for biotin by striping molecule 1 instead of 2 and replacing the path names in the \inlinecode{strip} and  \inlinecode{trajout} commands.
+Alternatively, we provide a small script named \software{cpptraj\_remove\_mol.sh} for this procedure.
+
 After that, use \software{equilibration.py -R} to run short restrained 
@@
 
-\subsubsection{MD Simulation}
-The MD can be run as described above for each monomer and the complex.
-
 \subsubsection{GIST analysis of Streptavidin-Biotin}
+
+After running the MD simulations as described in section \ref{sec:md_simulation}, proceed with the GIST analysis.
 For easier post-processing, here we use grids that contain the whole solute molecule.
+
 Smaller grids encompassing only the binding site could be used for streptavidin and the complex, but this would complicate the analysis.
@@
 Calculate the GIST grid parameters as outlined in the biotin section for the complex and streptavidin and run the GIST analysis.
-For larger systems like whole proteins you can usually get away with a smaller buffer region compared to small molecule solutes.
-To test whether your grid is large enough look at solvent properties close to the grid edges, e.g. the solvent-solvent energies.
+
+For larger systems like whole proteins you can usually get away with a smaller buffer region for the GIST grid compared to small molecule solutes.
+To test whether your grid is large enough, look at solvent properties close to the grid edges, e.g. the solvent-solvent energies.
+If these correspond to the respective bulk properties, it can be assumed that your grid is large enough to capture all perturbations of the solute on the solvent.
+
+Note that a different box-size or number of solvent molecules can skew the bulk properties slightly compared to the reference values provided above.
+In that case, checking for convergence of the solvent properties near the grid borders is sufficient.
+You might want to modify your reference values according to your results, as described in the next section.
 
@@
 Therefore, the reference energy can be obtained by binning the voxels by their distance to the solute molecule, and evaluate Equation~\ref{eq_ewwref} within each bin.
-If this value converges, this can be used as $E_{ww}^\textit{ref}$.
+If this value converges with increasing distance from the solute, this can be used as $E_{ww}^\textit{ref}$.
 
@@
 In a plot like Figure~\ref{fig_ewwref} based on the \inlinecode{A\_dens} column, the values should tend to zero.
-However, it is more informative to plot the cumulative free energy contribution against the distance to the solute molecule, to check the convergence.
+
+However, it is more informative to plot the cumulative free energy contribution against the distance to the solute molecule, to check the convergence (see Figure~\ref{fig_radial_convergence}).
+
 It the curve flattens out, the converged value is your final $\dgsolv$\@.
-Otherwise, you might need to tweak the $E_{ww}$ reference, or introduce a reference value for the entropy.
+If the curve diverges even for large integration distances you might need to tweak the $E_{ww}$ reference, or introduce a reference value for the entropy.
 
@@
 Even though streptavidin-biotin is known to be a very stable complex, the solvation energy favors the dissociation.
-This is expected: if the solute molecules were not able to form any strong interactions, but interacted favorably with the water instead, they would not form a complex.
+
+This is expected: the binding of biotin to streptavidin is faciliated over a large number of H-bonds and over polar interactions. 
+As such, water favourably solvates both biotin and streptavidin and an energetic penalty is paid when displacing the water during the binding process.
 
@@
 
-For the entropy, this leads to an infinite expansion of correlations of increasing order.
-In a slightly simplified view, the first order is the solvent density $g_\textit{sw}\left(\textbf{r},\omega \right)$ at each position $\textbf{r}$ and orientation $\omega$.
-The first-order entropy omits all solvent-solvent (and higher) correlations, while the solute-solvent correlation is taken into account because the coordinates are relative to the solute molecule.
+The entropy can be approximated as an infinite expansion of correlations of increasing order.
+In a slightly simplified view, the first order is the log solvent density $ln\left(g_\textit{sw}\left(\textbf{r},\omega \right)\right)$ at each position $\textbf{r}$ and orientation $\omega$.
+The first-order entropy omits all solvent-solvent (and higher) correlations, while the solute-solvent correlation is taken into account because the coordinates are relative to the solute.
 Thus, it is called the solute-water entropy $S_{sw}$.

### 2024-06-12 manuscript/manuscript.tex
Fix cheatsheet positioning, remove install section 8.1

@@
 In that case, checking for convergence of the solvent properties near the grid borders is sufficient but you might want to modify your reference values according to your results then as described in the next section.
-
 \subsubsection{Reference Values and Radial Convergence}
@@
 
-\subsection{Installation}
-GIST calculations in cpptraj can be sped up through various levels of 
-parallelization.
-If cpptraj was compiled with the '-cuda' configure flag set, the GPU accelerated version of GIST is used automatically. 
-In addition, OpenMP and MPI parallelization are available if cpptraj was compiled the necessary options.
-
-For detailed instructions on how to install \software{cpptraj}, we refer the reader to the AmberMD GitHub page \url{https://github.com/Amber-MD/cpptraj}.
+%\subsection{Installation}
+%GIST calculations in cpptraj can be sped up through various levels of 
+%parallelization.
+%If cpptraj was compiled with the '-cuda' configure flag set, the GPU accelerated version of GIST is used automatically. 
+%In addition, OpenMP and MPI parallelization are available if cpptraj was compiled the necessary options. 
+%
+%For detailed instructions on how to install \software{cpptraj}, we refer the reader to the AmberMD GitHub page \url{https://github.com/Amber-MD/cpptraj}.
 
@@
 
+\bigskip
 \textbf{Output}\\
@@
 % Here is a single-column checklist that consists of multiple sub-checklists
-\begin{Checklists}[h]
+\begin{Checklists}[H]
 

### 2024-06-12 manuscript/manuscript.tex
Add MKGs conflic of interest
Add python package/notebook notice to abstract

@@
 In this tutorial, we aim to guide the reader through the most common steps involved in a GIST analysis using the streptavidin-biotin complex as an demonstrative system.
+Towards this, Jupyter notebooks and python packages are provided to simplify the analysis.
 Furthermore, we discuss the theory of GIST with a focus on practical aspects.
 We highlight potential pitfalls and show how to avoid technical difficulties.
-We assume familiarity with molecular dynamics simulations as well as the AmberTools  package \cite{Case2023-ambertools}.
+For this tutorial, familiarity with molecular dynamics simulations as well as the AmberTools package \cite{Case2023-ambertools} is assumed.
 
@@
 
-Declare any potentially conflicting interests here, whether or not they pose an actual conflict in your view.
+MKG has an equity interest in and is a cofounder and scientiﬁc advisor of VeraChem LLC.
 

### 2024-06-13 manuscript/manuscript.tex
Add GCMC/NCMC reference to bib and text
Add MMPBSA/Endstate citations

@@
 Note that, due to the restraints, some solvent sites might be inaccessible to the bulk solvent.
-To investigate such positions, make sure to explicitly place water solvent molecules there. %TODO add MC/MD or mupt and sources
+To investigate such positions, make sure to explicitly place water solvent molecules there. 
+For cases where the exact occupation of such sites is unknown, it may be helpful to include grand-canonical Monte Carlo (GCMC) or nonequilibrium candidate Monte Carlo (NCMC) sampling steps in the equilibration or production stage \cite{Ge-GCMC2022,Melling-NCMC2023}.
 
@@
 Calculations for the disassociated states are then run on the extracted structures from the complex configurations.
-%TODO source MMPBSA
-One can choose to sample the complex and dissociated states individually, or to sample only one of them and assume that there is sufficient overlap in the conformations.
+One can choose to sample the complex and dissociated states individually, or to sample only one of them and assume that there is sufficient overlap in the conformations \cite{Genheden-EndState2012,Wang-MMPBSA2019}.
 

### 2024-06-13 manuscript/manuscript.tex
Remove nonsensical sentence
Adjustments to the checklists

@@
 \usepackage{siunitx}
+\usepackage{float}
 \DeclareSIUnit\Molar{M}
@@
 The expected output of this code is shown in Figure~\ref{fig_ewwref}.
-%TODO redo figure caption
+%TODO redo figure caption, yaxis label
 \begin{figure}
@@
 If you also visualize the individual histograms, you will notice that the difference converges much better with increasing radius than the single contributions.
-This is because the cutoff contains regions that are close to the atom, since the integration region does not comprise the whole protein.
-%TODO this sentence doesn't make sense
 
@@
 You can also plot the difference between the complex and monomer contributions against the distance to biotin, using the complex coordinates for the holo structure:
+
 \begin{lstlisting}[style=python]
@@
 
-\section{Cheat sheet}
-%This will be a cheatsheet containing all options for cpptrajs GIST implementation
-The checklist ``The CPPTRAJ GIST action'' provides a cheat sheet summarizing the important options to the GIST action and its output, as well as frequently used equations.
-
-%\subsection{Installation}
-%GIST calculations in cpptraj can be sped up through various levels of 
-%parallelization.
-%If cpptraj was compiled with the '-cuda' configure flag set, the GPU accelerated version of GIST is used automatically. 
-%In addition, OpenMP and MPI parallelization are available if cpptraj was compiled the necessary options. 
-%
-%For detailed instructions on how to install \software{cpptraj}, we refer the reader to the AmberMD GitHub page \url{https://github.com/Amber-MD/cpptraj}.
-
-\begin{Checklists*}
-
-\begin{checklist}{The CPPTRAJ GIST action}\label{cheatsheet}
-
-\textbf{Options}\\
-Various flags and options can be provided when running a GIST calculation in 
-cpptraj. A list of possible and required options is provided here:
-
-\begin{tabular}{@{}l p{.7\textwidth}@{}}
-\toprule
-doorder & Calculate the water order parameter [reference] for each voxel\\
-doeij & Calculate the triangular matrix representing the water-water 
-interactions between pairs of voxels.\\
-skipE & Skip all energy calculations (cannot be specified with ’doeij’)\\
-skipS & Skip all the entropy calculations.\\
-refdens <rdval> & Reference density of bulk water, used in computing 
-$g_O, g_H$, and the translational entropy. Default is 0.0334 molecules/Å3. \\
-temp <tval> & Temperature of the input trajectory. \\
-noimage &  Disable distance imaging in energy calculation. \\
-gridcntr <xval> <yval> <zval> & Coordinates (Å) of the center of the grid 
-(default 0.0, 0.0, 0.0). \\
-griddim <xval> <yval> <zval> & Grid dimensions along each coordinate axis 
-(default 40, 40, 40).\\
-gridspacn <spaceval> & Grid spacing (linear dimension of each voxel) in 
-Angstroms. Values greater than 0.75 Å are not recommended (default 0.5 Å).\\
-prefix <filename prefix> & Output file name prefix (default “gist”).\\
-ext <grid extension> & Output grid file name extension (default “.dx”). \\
-out <output> & Name of the main GIST output file. If not specified set to 
-’<prefix>-output.dat’.\\
-info <info> & Name of main GIST info file. If not specified info is 
-written to standard output. \\
-\bottomrule
-\end{tabular}
-
-\bigskip
-\textbf{Output}\\
-GIST calculations put out a variety of data sorted by voxel into a file 
-specified by the "out" keyword. Some of the output data is also automatically 
-written to "open data explorer" (.dx) files for convenient visualisation in 
-software such as VMD or PyMOL. The following columns can be found in the output 
-file:
-\begin{tabular}{@{}l l p{.8\textwidth}@{}}
-\toprule
-Name & Keyword & Description\\
-\midrule
-gX & X & For every element in the main solvent, the number density of atoms found in the voxel, in units of the bulk density.
-If the same element occurs multiple times, the bulk density is scaled accordingly.
-\\
-Esw & X &  Mean solute-water interaction energy density. \\
-Eww & X & Mean water-water interaction energy density. \\
-PME & [pme] & Water PME energy density. \\
-U\_PME & [pme] & Water PME energy density. \\
-dTStrans & X & First order translational entropy density. \\
-dTSorient & X & First order orientational entropy density . \\
-neighbor & X & Mean number of waters neighboring the water solvent molecules found in this voxel multiplied by the voxel number density. \\
-dipole & X & Magnitude of mean dipole moment (polarization). \\
-order & X & (if \inlinecode{doorder} was specified) Average Tetrahedral Order Parameter. \\
-dipolex & X & x-component of the mean water dipole moment density \\
-dipoley & X & y-component of the mean water dipole moment density \\
-dipolez & X & z-component of the mean water dipole moment density \\
-Eij & X & (if \inlinecode{doeij} was specified) Water-water interaction matrix. \\
-\bottomrule
-\end{tabular}
-\end{checklist}
-\end{Checklists*}
-
 \section{Checklists}
@@
 
+%\section{Cheat sheet}
+%This will be a cheatsheet containing all options for cpptrajs GIST implementation
+%The checklist ``The CPPTRAJ GIST action'' provides a cheat sheet summarizing the important options to the GIST action and its output, as well as frequently used equations.
+
+%\subsection{Installation}
+%GIST calculations in cpptraj can be sped up through various levels of 
+%parallelization.
+%If cpptraj was compiled with the '-cuda' configure flag set, the GPU accelerated version of GIST is used automatically. 
+%In addition, OpenMP and MPI parallelization are available if cpptraj was compiled the necessary options. 
+%
+%For detailed instructions on how to install \software{cpptraj}, we refer the reader to the AmberMD GitHub page \url{https://github.com/Amber-MD/cpptraj}.
+
+\begin{Checklists*}[h!]
+\begin{checklist}{The CPPTRAJ GIST action}\label{cheatsheet}
+
+\textbf{Options}\\
+Various flags and options can be provided when running a GIST calculation in cpptraj. 
+A list of possible and required options is provided here:
+\medskip
+
+\begin{tabular}{@{}l p{.7\textwidth}@{}}
+\toprule
+name <dataset name> & Name for output datasets in cpptraj. \\
+doorder & Calculate the water order parameter [reference] for each voxel\\
+\qquad nopl & Do not use pair list for order calculation\\
+\qquad plcut <val> & Pair list cutoff for order calculation. Default is 10 Å\\
+doeij & Calculate the triangular matrix representing the water-water 
+interactions between pairs of voxels.\\
+skipE & Skip all the energy calculations (cannot be specified with ’doeij’)\\
+skipS & Skip all the entropy calculations.\\
+refdens <val> & Reference density of bulk solvent, used in computing 
+$g_O, g_H$, and the translational entropy. Default is 0.0334 molecules/Å3. \\
+temp <val> & Temperature of the input trajectory. \\
+noimage &  Disable distance imaging in energy calculation. \\
+gridcntr <xval> <yval> <zval> & Coordinates in Å of the center of the grid. Default is 0.0, 0.0, 0.0. \\
+griddim <xval> <yval> <zval> & Grid dimensions in voxels along each coordinate axis. Default is 40, 40, 40.\\
+gridspacn <val> & Grid spacing (linear dimension of each voxel) in 
+Angstroms. Values greater than 0.75 Å are not recommended. Default is 0.5 Å.\\
+neighborcut <val> & Cutoff in Å for determining solvent O-O neighbors. Default is 3.5 Å. \\
+prefix <prefix> & Output file name prefix. Default is “gist”.\\
+ext <extension> & Output grid file name extension. Default is  “.dx”. \\
+out <file> & Name of the main GIST output file. If not specified set to “<prefix>-output.dat“.\\
+info <file> & Name of main GIST info file. If not specified info is written to standard output. \\
+floatfmt <'double'|'scientific'|'general'> & Format for floating
+point values in GIST output file. Default chooses 'fixed' or 'scientific' automatically. \\
+floatwidth <val>& Width of floating point values in GIST output file. Default is no width restriction. \\
+floatprec <val>& Precision of floating point values in GIST output file. Default is the system default. \\
+intwidth <val>& Width of integer point values in GIST output file. Default is no width restriction. \\
+oldnnvolume & Use the old reference volume for the nearest neighbor entropy, instead of the more precise new implementation. \\
+nocom & Do not use the center of mass to define the molecular position. Instead, use the first atom in rigidatoms. Use this flag to restore the behavior of old GIST runs. \\
+nnsearchlayers <val> & Number of layers of neighboring voxels that should be used when searching for nearest neighbors.
+This has to be at least 1 to obtain the correct entropy.
+Higher values can help to obtain better convergence of the translational and 6D entropy with little sampling or fine grid spacings, but increase the calculation time (default 1).
+\\
+solute <mask> & Selection mask for the solute. All other molecules will be solvent. If this is omitted, the standard solute/solvent assignment will be used. \\
+solventmols <mols> &Comma-separated list of names of solvent molecules.
+Energies will be computed per solvent molecule.
+For the entropy, only the main solvent (the first one) will be used.
+Use, e.g., solventmols WAT,NA,CL for a GIST calculation including ions.
+This needs to be specified if there is more than one solvent species.
+\\
+rigidatoms <central> <atom1> <atom2> & Specifies how to define the molecular orientation for the entropy. 
+By default, a simple heuristic will be used. 
+This works for water, but not for all solvents. 
+The atoms should be representative of the molecular orientation and should not be collinear. 
+Note that the central atom goes first. 
+For water, the default is equivalent to rigidatoms O H1 H2, corresponding to H1-O-H2 as the rigid substructure. \\
+\bottomrule
+\end{tabular}  
+\end{checklist}
+\end{Checklists*}
+
+\begin{Checklists*}[h!]
+\begin{checklist}{The CPPTRAJ GIST action}
+
+	\textbf{Options}\\
+	Various flags and options can be provided when running a GIST calculation in cpptraj. 
+	A list of possible and required options is provided here:
+	\medskip
+
+\begin{tabular}{@{}l p{.7\textwidth}@{}}
+\toprule
+nopme & Do not use particle mesh Ewald for the non-bonded calculation. This is set on default. \\
+pme & Use particle mesh Ewald for the non-bonded electrostatics calculation. The van der Waals energy will be calculated using a long-range correction for periodicity. Does not support doeij.\\
+\qquad cut <val> & Direct space cutoff for pme. Default is 3.5 Å.\\
+\qquad dsumtol <val> & Direct sum tolerance used to determine Ewald coefficient. Default is 0.00001.\\
+\qquad ewcoeff <val> & Ewald coefficient in 1/Å. \\
+\qquad erfcdx <val> & Spacing to use for the ERFC splines. Default is 0.0002 Å. \\
+\qquad skinnb & Used to determine pairlist atoms (added to cut, so pairlist cutoff is cut + skinnb); included in order to maintain consistency with results from sander. \\
+\qquad ljswidth <val> & If specified, use a force-switching form for the Lennard-Jones calculation from <cutoff>-<val> to <cutoff> \\
+\qquad order <val> & Spline order for charges. \\
+\qquad nfft <nfft1>,<nfft2>,<nfft3> & Explicitly set the
+number of FFT grid points in each dimension.
+Will be determined automatically if not
+specified. \\
+\bottomrule
+\end{tabular}   
+\bigskip
+
+\textbf{Output}\\
+GIST calculations put out a variety of data sorted by voxel into a file specified by the "out" keyword. 
+Some of the output data is also automatically written to "open data explorer" (.dx) files for convenient visualisation in software such as VMD or PyMOL.
+The following columns can be found in the output file:
+\medskip
+
+\begin{tabular}{@{}l l p{.8\textwidth}@{}}
+\toprule
+Name & Keyword & Description\\
+\midrule
+gX & X & For every element in the main solvent, the number density of atoms found in the voxel, in units of the bulk density.
+If the same element occurs multiple times, the bulk density is scaled accordingly.
+\\
+Esw & X &  Mean solute-solvent interaction energy density. \\
+Eww & X & Mean solvent-solvent interaction energy density. \\
+PME & [pme] & Solvent PME energy density. \\
+U\_PME & [pme] & Solvent PME energy density. \\
+dTStrans & X & First order translational entropy density. \\
+dTSorient & X & First order orientational entropy density . \\
+neighbor & X & Mean number of solvent molecules neighboring the solvent molecules found in this voxel multiplied by the voxel number density. \\
+dipole & X & Magnitude of mean dipole moment (polarization). \\
+order & X & (if \inlinecode{doorder} was specified) Average Tetrahedral Order Parameter. \\
+dipolex & X & x-component of the mean solvent dipole moment density \\
+dipoley & X & y-component of the mean solvent dipole moment density \\
+dipolez & X & z-component of the mean solvent dipole moment density \\
+Eij & X & (if \inlinecode{doeij} was specified) solvent-solvent interaction matrix. \\
+\bottomrule
+\end{tabular}
+\end{checklist}
+\end{Checklists*}
+
 \section*{Author Contributions}

### 2024-06-13 manuscript/manuscript.tex
Finish cheat sheet, fix placement of checklists/sheets

@@
 \usepackage{float}
+\usepackage{placeins}
 \DeclareSIUnit\Molar{M}
@@
 
-\section{Checklists}
+\section{Checklists and Cheat Sheats}
 %Tutorials do not necessarily require the use of a checklist as in Best Practices documents; however, they can include these if desired.
@@
 \end{Checklists}
-
 %\section{Cheat sheet}
@@
 %For detailed instructions on how to install \software{cpptraj}, we refer the reader to the AmberMD GitHub page \url{https://github.com/Amber-MD/cpptraj}.
-
+\FloatBarrier
 \begin{Checklists*}[h!]
@@
 
-\begin{tabular}{@{}l p{.7\textwidth}@{}}
+\begin{tabularx}{\textwidth}{@{}l X @{} }
 \toprule
+\textbf{I/O Options} &\\
 name <dataset name> & Name for output datasets in cpptraj. \\
-doorder & Calculate the water order parameter [reference] for each voxel\\
-\qquad nopl & Do not use pair list for order calculation\\
-\qquad plcut <val> & Pair list cutoff for order calculation. Default is 10 Å\\
-doeij & Calculate the triangular matrix representing the water-water 
-interactions between pairs of voxels.\\
+prefix <prefix> & Output file name prefix. Default is “gist”.\\
+ext <extension> & Output grid file name extension. Default is  “.dx”. \\
+out <file> & Name of the main GIST output file. If not specified set to “<prefix>-output.dat“.\\
+info <file> & Name of main GIST info file. If not specified info is written to standard output. \\
+floatfmt <fmt> & Format for floating point values in GIST output file. Options are 'double','scientific' or 'general'. Default chooses 'fixed' or 'scientific' automatically. \\
+floatwidth <val>& Width of floating point values in GIST output file. Default is no width restriction. \\
+floatprec <val>& Precision of floating point values in GIST output file. Default is the system default. \\
+intwidth <val>& Width of integer point values in GIST output file. Default is no width restriction. \\
+doeij & Output the triangular matrix representing the water-water 
+interactions between pairs of voxels. Not supported with PME or GPU GIST.\\
+
+\textbf{Grid Options} &\\
+gridcntr <xval> <yval> <zval> & Coordinates in Å of the center of the grid. Default is 0.0, 0.0, 0.0. \\
+griddim <xval> <yval> <zval> & Grid dimensions in voxels along each coordinate axis. Default is 40, 40, 40.\\
+gridspacn <val> & Grid spacing (linear dimension of each voxel) in 
+Angstroms. Values greater than 0.75 Å are not recommended. Default is 0.5 Å.\\
+
+\textbf{GIST Calculation Options} &\\
+
 skipE & Skip all the energy calculations (cannot be specified with ’doeij’)\\
@@
 noimage &  Disable distance imaging in energy calculation. \\
-gridcntr <xval> <yval> <zval> & Coordinates in Å of the center of the grid. Default is 0.0, 0.0, 0.0. \\
-griddim <xval> <yval> <zval> & Grid dimensions in voxels along each coordinate axis. Default is 40, 40, 40.\\
-gridspacn <val> & Grid spacing (linear dimension of each voxel) in 
-Angstroms. Values greater than 0.75 Å are not recommended. Default is 0.5 Å.\\
+
 neighborcut <val> & Cutoff in Å for determining solvent O-O neighbors. Default is 3.5 Å. \\
-prefix <prefix> & Output file name prefix. Default is “gist”.\\
-ext <extension> & Output grid file name extension. Default is  “.dx”. \\
-out <file> & Name of the main GIST output file. If not specified set to “<prefix>-output.dat“.\\
-info <file> & Name of main GIST info file. If not specified info is written to standard output. \\
-floatfmt <'double'|'scientific'|'general'> & Format for floating
-point values in GIST output file. Default chooses 'fixed' or 'scientific' automatically. \\
-floatwidth <val>& Width of floating point values in GIST output file. Default is no width restriction. \\
-floatprec <val>& Precision of floating point values in GIST output file. Default is the system default. \\
-intwidth <val>& Width of integer point values in GIST output file. Default is no width restriction. \\
+
 oldnnvolume & Use the old reference volume for the nearest neighbor entropy, instead of the more precise new implementation. \\
-nocom & Do not use the center of mass to define the molecular position. Instead, use the first atom in rigidatoms. Use this flag to restore the behavior of old GIST runs. \\
-nnsearchlayers <val> & Number of layers of neighboring voxels that should be used when searching for nearest neighbors.
-This has to be at least 1 to obtain the correct entropy.
-Higher values can help to obtain better convergence of the translational and 6D entropy with little sampling or fine grid spacings, but increase the calculation time (default 1).
+
+nnsearchlayers <val> & Layers of neighboring voxels considered in nearest neighbor search.
+Higher values may improve entropy convergence for little sampling or fine grid spacings, but increase the calculation time (default 1).
 \\
-solute <mask> & Selection mask for the solute. All other molecules will be solvent. If this is omitted, the standard solute/solvent assignment will be used. \\
-solventmols <mols> &Comma-separated list of names of solvent molecules.
-Energies will be computed per solvent molecule.
-For the entropy, only the main solvent (the first one) will be used.
-Use, e.g., solventmols WAT,NA,CL for a GIST calculation including ions.
-This needs to be specified if there is more than one solvent species.
-\\
-rigidatoms <central> <atom1> <atom2> & Specifies how to define the molecular orientation for the entropy. 
-By default, a simple heuristic will be used. 
-This works for water, but not for all solvents. 
-The atoms should be representative of the molecular orientation and should not be collinear. 
-Note that the central atom goes first. 
-For water, the default is equivalent to rigidatoms O H1 H2, corresponding to H1-O-H2 as the rigid substructure. \\
-\bottomrule
-\end{tabular}  
-\end{checklist}
-\end{Checklists*}
-
-\begin{Checklists*}[h!]
-\begin{checklist}{The CPPTRAJ GIST action}
-
-	\textbf{Options}\\
-	Various flags and options can be provided when running a GIST calculation in cpptraj. 
-	A list of possible and required options is provided here:
-	\medskip
-
-\begin{tabular}{@{}l p{.7\textwidth}@{}}
-\toprule
+
+\textbf{PME Options} & \\
 nopme & Do not use particle mesh Ewald for the non-bonded calculation. This is set  on default. \\
-pme & Use particle mesh Ewald for the non-bonded electrostatics calculation. The van der Waals energy will be calculated using a long-range correction for periodicity. Does not support doeij.\\
+pme & Use particle mesh Ewald for the non-bonded electrostatics calculation. The van der Waals energy will be calculated using a long-range correction for periodicity.\\
 \qquad cut <val> & Direct space cutoff for pme. Default is 3.5 Å.\\
@@
 specified. \\
+
+\end{tabularx}  
+\end{checklist}
+\end{Checklists*}
+\begin{Checklists*}[h!]
+\begin{checklist}{The CPPTRAJ GIST action}
+
+\textbf{Further Options}\\
+\begin{tabularx}{\textwidth}{@{} l X @{}}
+\toprule
+
+\textbf{Solvent Options} &\\
+solute <mask> & Selection mask for the solute. All other molecules will be solvent. If this is omitted, the standard solute/solvent assignment will be used. \\
+solventmols <mols> &Comma-separated list of solvent molecules residue names.
+Energies will be computed per solvent molecule.
+For the entropy, only the first solvent will be used.
+Use this for calulations with more than one solvent of interest, e.g.
+for  ions.
+\\
+rigidatoms <c> <a1> <a2> & Specifies the molecular orientation for the entropy calculation from a central atom and two additional atoms, e.g.
+O H1 H2 for water.
+By default, a simple heuristic will be used.
+Use this option if the automatically chosen atoms are collinear or do not represent the orientation well. \\
+nocom & Do not use the center of mass to define the molecular position. Instead, use the first atom in rigidatoms. Use this flag to restore the behavior of old GIST runs. \\
+\textbf{Order Calculation Options} &\\
+doorder & Calculate the water order parameter [reference] for each voxel\\
+\qquad nopl & Do not use pair list for order calculation\\
+\qquad plcut <val> & Pair list cutoff for order calculation. Default is 10 Å\\
 \bottomrule
-\end{tabular}   
-\bigskip
+\end{tabularx}   
+\medskip
 
 \textbf{Output}\\
-GIST calculations put out a variety of data sorted by voxel into a file specified by the "out" keyword. 
+GIST calculations write a variety of data sorted by voxel into an outputfile specified by the "out" keyword. 
 Some of the output data is also automatically written to "open data explorer" (.dx) files for convenient visualisation in software such as VMD or PyMOL.
+Note that some columns will be written out both with '\_norm' and '\_dens' suffixes, refereing to normalization by solvent molecule or voxel volume respectively.
 The following columns can be found in the output file:
@@
 
-\begin{tabular}{@{}l l p{.8\textwidth}@{}}
+\begin{tabularx}{\textwidth}{@{}l l X@{}}
 \toprule
@@
 \midrule
-gX & X & For every element in the main solvent, the number density of atoms found in the voxel, in units of the bulk density.
+index & & Voxel indices \\
+x, y, z & & Coordinates of the voxel centers \\
+pop & & Population of solvent in voxel over entire simulation \\
+gX &  & For every element in the main solvent, the number density of atoms found in the voxel, in units of the bulk density.
 If the same element occurs multiple times, the bulk density is scaled accordingly.
 \\
-Esw & X &  Mean solute-solvent interaction energy density. \\
-Eww & X & Mean solvent-solvent interaction energy density. \\
-PME & [pme] & Solvent PME energy density. \\
-U\_PME & [pme] & Solvent PME energy density. \\
-dTStrans & X & First order translational entropy density. \\
-dTSorient & X & First order orientational entropy density . \\
-neighbor & X & Mean number of solvent molecules neighboring the solvent molecules found in this voxel multiplied by the voxel number density. \\
-dipole & X & Magnitude of mean dipole moment (polarization). \\
-order & X & (if \inlinecode{doorder} was specified) Average Tetrahedral Order Parameter. \\
-dipolex & X & x-component of the mean solvent dipole moment density \\
-dipoley & X & y-component of the mean solvent dipole moment density \\
-dipolez & X & z-component of the mean solvent dipole moment density \\
-Eij & X & (if \inlinecode{doeij} was specified) solvent-solvent interaction matrix. \\
-\bottomrule
-\end{tabular}
+g\_mol\_Y & [solventmols]  & Density of every solvent species Y. Scaled by rho0. \\
+Esw &  &  Mean solute-solvent interaction energy. \\
+Eww &  & Mean solvent-solvent interaction energy. \\
+Esw\_mol\_Y & [solventmols]  &  Mean solute-solvent interaction energy for solvent species Y. \\
+Eww\_mol\_Y & [solventmols]  & Mean solvent-solvent interaction energy for solvent species Y. \\
+PME & [pme] & Solvent PME energy. \\
+U\_PME & [pme] & Solvent PME energy. \\
+dTStrans &  & First order translational entropy. \\
+dTSorient &  & First order orientational entropy. \\
+order &  & (if \inlinecode{doorder} was specified) Average Tetrahedral Order Parameter. \\
+dipolex &  & x-component of the mean solvent dipole moment density \\
+dipoley &  & y-component of the mean solvent dipole moment density \\
+dipolez &  & z-component of the mean solvent dipole moment density \\
+dipole &  & Magnitude of mean dipole moment density (polarization). \\
+neighbor &  & Mean number of solvent molecules neighboring the solvent molecules found in this voxel. \\
+order & & Average tetrahedral order parameter \\
+
+\end{tabularx}
 \end{checklist}
 \end{Checklists*}
-
+\FloatBarrier
 \section*{Author Contributions}

### 2024-06-13 manuscript/manuscript.tex

@@
 The MD simulations presented in this tutorial are run with the Amber simulation engine and analyzed using AmberTools \software{cpptraj}. 
-For a conceptual overview of running MD simulations with Amber, we recommend reading the Amber manual (\url{https://ambermd.org/Manuals.php}) \cite{amber22} and the Amber tutorials (\url{https://ambermd.org/Manuals.php}).
+For a conceptual overview of running MD simulations with Amber, we recommend reading the Amber manual (\url{https://ambermd.org/Manuals.php}) \cite{amber24} and the Amber tutorials (\url{https://ambermd.org/Manuals.php}).
 Previous experience of GIST is not necessary, although we recommend reading the works referenced in section~\ref{sec:theory}.
@@
 In order to run a GIST analysis, one must define the region where solvent properties will be calculated.
-
-This region is occupied by a rectangular grid with given position, size, and voxel spacing.
-The default spacing of \SI{0.5}{\angstrom} provides reasonable discretization for many use-cases.
+This region is defined by a rectangular grid with given position, size, and voxel spacing.
+The default grid spacing of \SI{0.5}{\angstrom} provides reasonable discretization for many use-cases.
 While integrals of GIST quantities over the grid are independent of the voxel size, provided sufficient sampling, individual values converge slower at high resolutions.
@@
 
-Additionally, the GIST calculation will usually be preceded by preprocessing steps, such as centering the trajectory or reimaging to account for periodic boundary conditions.
+Additionally, the GIST calculation should usually be preceded by preprocessing steps, such as centering the trajectory or reimaging to account for periodic boundary conditions.
 Often, the region of interest will be defined as the surrounding of a molecule, which might for example be a ligand in the binding pocket or the whole protein. 
@@
 
-Additionally, a helper script called \inlinecode{FindCentroid.py} for the same purpose is supplied with this tutorial, and can be used as follows:
+In addition, a helper script called \inlinecode{FindCentroid.py} for the same purpose is supplied with this tutorial, and can be used as follows:
 
@@
 The command needed to run GIST in cpptraj can be written in an input file (e.g gist.in) for convenience.
-The basic command, assuming a grid centered at (10.5 20.1 30.1) and a grid size of $30\times30\times30$ voxels with \SI{0.5}{\angstrom} spacing, is:
+The basic command, assuming a grid centered at coordinates (x=\SI{10.5}{\angstrom}, y=\SI{20.1}{\angstrom}, z=\SI{30.1}{\angstrom}) and a grid size of $30\times30\times30$ voxels with \SI{0.5}{\angstrom} spacing, is:
 
@@
 The solvent-solvent enthalpy $E_{ww}$ also needs to be referenced, but this can be done in the post-processing stage.
-The calculation of these properties and the need for referencing is discussed in-depth in section /ref{sec:theory}.
+The calculation of these properties and the need for referencing is discussed in-depth in section \ref{sec:theory}.
 
@@
 
-Due to the varying sizes of the boxes, the reference energies used in this context may not align with those specified in the Amber manual. 
-Generally, it is recommended that the size of the reference water box closely matches that of the production system. 
-Therefore, it is advisable to conduct your own reference calculations, especially for systems that are either very large or very small.
-While the impact on density is not as pronounced, deviations from the specified settings can lead to differences of up to 0.005 kcal/mol in the energy per water molecule ($E_{ww, norm}$) for water boxes.
+Due to the different sizes of the boxes, the reference energies used in this tutorial may not align with those specified in the Amber manual. 
+Generally, it is advisable to conduct your own reference calculations, especially for systems that are either very large or very small. 
+The reference values can be calculated from MD trajectories of bulk solvent.
+From this, one then calculates the average energy per solvent molecule ($E_{ww, norm}$) and the average solvent density $\rho_0$.
+Generally, it is recommended that the size of the reference solvent box closely matches that of the production system.
+While the impact on the density is not as pronounced, deviations from the production settings can lead to differences of up to 0.005 kcal/mol in the energy per solvent molecule ($E_{ww, norm}$) for water boxes.
 These discrepancies, when summing over large grid regions, have the potential to introduce significant errors due to incorrect referencing.
@@
 	\label{tab:ref_values}
-	\caption{Reference number density $\rho_0$ and mean water-water energy $E_{ww, norm}$ for various water models from NPT simulations of N$_{WAT}$ water solvent molecules at \SI{300}{\kelvin} and 1bar.}\label{tab:ref_densities}
+	\caption{Reference number density $\rho_0$ and mean solvent-solvent energy $E_{ww, norm}$ for various water models from NPT simulations of N$_{WAT}$ water solvent molecules at \SI{300}{\kelvin} and 1bar.}\label{tab:ref_densities}
 	\small
@@
 For water, gO and gH are produced.}
-
-	\item{[Esw] Mean solute-water interaction energy density.}
-	\item{[Eww] Mean water-water interaction energy density.}
-	\item{[PME] (only if PME was used) water PME energy density.}
-	\item{[dTStrans] First order translational entropy density.}
-	\item{[dTSorient] First order orientational entropy density .}
-	\item{[dTSsix] First order entropy density for the six-dimensional combined orientation and translation space.}
+	\item{[Esw] Mean solute-water interaction energy.}
+	\item{[Eww] Mean water-water interaction energy.}
+	\item{[PME] (only if PME was used) water PME energy.}
+	\item{[dTStrans] First order translational entropy.}
+	\item{[dTSorient] First order orientational entropy.}
+	\item{[dTSsix] First order entropy for the six-dimensional combined orientation and translation space.}
 	\item{[dipole] Magnitude of mean dipole moment (polarization).}
-	\item{[dipolex] x-component of the mean water dipole moment density}
-	\item{[dipoley] y-component of the mean water dipole moment density}
-	\item{[dipolez] z-component of the mean water dipole moment density}
 \end{itemize}
@@
 All calculated quantities are written to the output file (default \inlinecode{gist-output.dat}) and can be used to generate DX files for other quantities.
-A full list of all properties in the output is provided at the end of the tutorial. 
+A full list of all properties in the output is provided at the end of the tutorial in section \ref{sec:checklists}. 
 The output file is organized by voxel and sorted according to their x, y, and z coordinates, describing the center point of the voxel.
@@
 
-In the first part, we show how to run GIST calculations for small molecule solutes and interpret the output.
+In the first part, we show how to run GIST calculations for small molecule solutes and interpret the output. This provides an example of a simple GIST workflow, without too much details.
 Afterwards, we apply this knowledge to a protein binding pocket, visualize $\dgsolv$ hotspots and estimate a free energy of ligand binding using an end-states approach based on GIST.
-In addition, we discuss all steps in further detail at the end.
+In addition, we discuss all steps shown in the first example in further detail.
 
@@
 The solute molecule should be solvated with a sufficiently large solvent box. 
-For periodic boundary conditions, the size of the solvent box should be large enough for the solvent to behave bulk-like so as to reduce artifact from solvent-mediated interactions between the periodic images.
+For periodic boundary conditions, the size of the solvent box should be large enough for the solvent to behave bulk-like so as to reduce artifacts from solvent-mediated interactions between the periodic images.
 This corresponds to a distance between solute and box boundary of at least 12\AA{} for water, though we recommend 15\AA{} or higher.
@@
 Such dominant structures could either be identified from experimental ensembles or by clustering an unrestrained MD simulation.
-
 When starting from an experimental structure, care has to be taken during the preparation step.
 Make sure to choose a proper protonation state for all titratable moeties.
-
-You will need to solvate your structure in a solvent box, which should be large enough to obtain unbiased long-range electrostatics.
-A good criterion for validating the box size is to check whether the water on the outside of the box behaves bulk-like according to the GIST output.
-Note that, due to the restraints, some solvent sites might be inaccessible to the bulk solvent.
+You will need to solvate your structure in a solvent box, which should be large enough to obtain unbiased long-range electrostatics and avoid solvent-mediated interactions of periodic images.
+A good criterion for validating the box size is to check whether the solvent on the outside of the box behaves bulk-like according to the GIST output.
+Note that, due to the restraints and finite sampling time, some solvent sites might be inaccessible to the bulk solvent.
 To investigate such positions, make sure to explicitly place water solvent molecules there. 
@@
 After running the MD simulations as described in section \ref{sec:md_simulation}, proceed with the GIST analysis.
-For easier post-processing, here we use grids that contain the whole solute molecule.
-
-Smaller grids encompassing only the binding site could be used for streptavidin and the complex, but this would complicate the analysis.
-Furthermore, we recommend that you center the solute molecule to the coordinate origin.
+For easier post-processing, here we use grids that contain the whole solute molecule and the surrounding solvent.
+
+If one is only interested in solvent in the binding site, smaller grids encompassing only the binding site could be used for streptavidin and the complex.
+While this saves calculation time for the GIST analysis, it complicates the post-processing.
+Since the GIST analysis is usually less time-demanding than the MD simulation, we recommend using larger grids.
+Furthermore, we recommend that you center the solute molecule to the coordinate origin to simplify the analysis.
 Calculate the GIST grid parameters as outlined in the biotin section for the complex and streptavidin and run the GIST analysis.
@@
 If these correspond to the respective bulk properties, it can be assumed that your grid is large enough to capture all perturbations of the solute on the solvent.
-
 Note that a different box-size or number of solvent molecules can skew the bulk properties slightly compared to the reference values provided above.
@@
 
-If these correspond to the respective bulk properties, it can be assumed that your grid is large enough to fully capture the perturbation of the solvent by the solute molecule.
-
-Note that a different box-size or number of solvent molecules can skew the bulk properties slightly compared to the reference values provided above.
-In that case, checking for convergence of the solvent properties near the grid borders is sufficient but you might want to modify your reference values according to your results then as described in the next section.
 \subsubsection{Reference Values and Radial Convergence}
@@
 For quantitative analyses, it is recommended to compute an exact reference energy.
-
 The most accurate method to compute a reference energy is to run a GIST calculation of the pure solvent using the same energy method (PME/GPU), a similar box size, at the same temperature and pressure and then compute the average solvent-solvent energy per solvent molecule.
@@
 Therefore, the reference energy can be obtained by binning the voxels by their distance to the solute molecule, and evaluate Equation~\ref{eq_ewwref} within each bin.
-If this value converges with increasing distance from the solute, this can be used as $E_{ww}^\textit{ref}$.
+If this value converges with increasing distance from the solute, it can be used as $E_{ww}^\textit{ref}$.
 
@@
 
-\section{Checklists and Cheat Sheats}
+\section{Checklists and Cheat Sheats} \label{sec:checklists}
 %Tutorials do not necessarily require the use of a checklist as in Best Practices documents; however, they can include these if desired.
@@
 gridspacn <val> & Grid spacing (linear dimension of each voxel) in 
-Angstroms. Values greater than 0.75 Å are not recommended. Default is 0.5 Å.\\
-
+Angstroms. Default is 0.5 Å.\\
+rmsfit <mask> & If specified, grid will be rotated and translated to follow the atoms selected by mask. \\
 \textbf{GIST Calculation Options} &\\
@@
 
-oldnnvolume & Use the old reference volume for the nearest neighbor entropy, instead of the more precise new implementation. \\
+oldnnvolume & Use the old reference volume for the nearest neighbor entropy. \\
 

### 2024-06-19 manuscript/manuscript.tex
Fix tabel labeling
Change author order
Address Franz comments in the introduction

@@
 
+\author[1\authfn{1}]{Valentin J Hoerschinger}
 \author[1\authfn{1}]{Franz Waibl}
-\author[1\authfn{1}]{Valentin J Hoerschinger}
 \author[2]{Vjay Molino}
+\author[2]{Helmut Carter}
 \author[1]{Monica L Fern{\'a}ndez-Quintero}
@@
 In this tutorial, we aim to guide the reader through the most common steps involved in a GIST analysis using the streptavidin-biotin complex as an demonstrative system.
-Towards this, Jupyter notebooks and python packages are provided to simplify the analysis.
+Towards this, Jupyter notebooks and Python packages are provided to simplify the analysis.
 Furthermore, we discuss the theory of GIST with a focus on practical aspects.
@@
 
-Interaction with water are of special importance as many biological processes occur in aqueous environment \cite{Privalov2017-water-review}.
-The free energy of solvation is defined as the reversible work needed to insert a solute molecule into the solvent \cite{ben-naim-book}.
+Interactions with water are of special importance as many biological processes occur in aqueous environment \cite{Privalov2017-water-review}.
+The free energy of solvation is defined as the reversible work needed to insert a solute molecule into the solvent from gas phase \cite{ben-naim-book}.
 It is directly linked to the solubility of gases and drives many real-world effects such as hydrophobicity, ligand binding, and protein folding.
@@
 Grid inhomogeneous solvation theory (GIST) \cite{Nguyen2012} is a method to compute the free energy of solvation, with applications in the context of drug design and biophysical property description.
-It estimates the solvent density from an MD trajectory and computes localized entropic and enthalpic contributions on a grid around the solute molecule.
-Different from most other methods, these contributions are therefore spatially resolved, improving interpretability and allowing better insights into the nature of solvation around the investigated solute.
+It computes the solvent density from an MD trajectory and estimates localized entropic and enthalpic contributions on a grid around the solute molecule.
+Since these contributions are spatially resolved, GIST improves interpretability and offers more detailed insights into the solvation of the investigated solute. 
 The entropy is computed from the density distribution of the solvent around the solute molecule, in a first-order approximation. A schematic of this is shown in Figure~\ref{fig:carbazole}. 
@@
 Since its initial development, GIST has been used to describe the thermodynamics of water displacement in protein binding pockets \cite{Nguyen2016-gist-second-order,Ramsey2016,Balius2017-gist-ligand-discovery,Hufner-Wulsdorf2020-gist-drug-design}.
-As such, it has been deployed in systematically mapping the hydration properties of proteins found in SARS-CoV-2 \cite{Olson2020-covid-gist}.
+It has been used for a large-scale investigation of the hydration properties of proteins found in SARS-CoV-2 \cite{Olson2020-covid-gist}.
 Furthermore, recent works used GIST to describe biophysical properties of serine proteases \cite{Kraml2019-gigist}, macrocycles \cite{Kamenik2020-gist-macrocycles}, or antibody fragements \cite{Waibl2021-gist-antibodies}.
@@
 
-In this tutorial, we aim to give the reader an introduction into GIST and its application to calculate solution thermodynamics properties of interest. 
-While previous publications \cite{Ramsey2016} and tutorials\footnote{See \url{http://ambermd.org/tutorials/advanced/tutorial25/}} gave insightful pointers how to setup and run GIST calculations, the post-processing was often treated less rigorously.
+In this tutorial, we aim to give the reader an introduction to GIST and its application to calculate solution thermodynamics properties of interest. 
+While previous publications \cite{Ramsey2016} and tutorials \cite{amber_tut_gist} treated how to setup and run GIST calculations in detail, the post-processing was often treated less rigorously.
 While building on these tutorials, we provide an updated guide to calculating solvation properties using GIST to address recent improvements.
-Furthermore, we aim to establish a template for the analysis and interpretation of GIST calculations.
+Furthermore, we aim to establish best practices for the analysis and interpretation of GIST calculations.
 
@@
 Methods derived from statistical mechanics bridge this gap by providing three-dimensional resolution.
-Two major flavors of methods in this group are those based on the Ornstein--Zernike (OZ) equation \cite{Hansen2013-simple-liquids} and those based on inhomogeneous fluid solvation theory (IST) \cite{Lazaridis1998}.
+These methods most commonly are based on either the Ornstein–Zernike (OZ) equation, or \cite{Hansen2013-simple-liquids} inhomogeneous fluid solvation theory (IST) \cite{Lazaridis1998}.
 The most important example of OZ-based methods is the reference interaction site model (RISM) \cite{Chandler1972-rism} as well as its extension to three dimensions (3D-RISM) \cite{Kovalenko1998-3drism}.
@@
 Examples include grid inhomogeneous solvation theory (GIST) \cite{Nguyen2012,Ramsey2016}, WaterMap \cite{Young2007-watermap,Abel2008-watermap}, SSTMap \cite{Haider2018-sstmap}, STOW \cite{Li2012-stow} and Solvaware \cite{Huggins2016-solvaware}.
-The main limitation of those methods is their reliance on approximating the entropy through an infinite expansion in terms of the density distribution. 
+The main limitation of those methods is that the entropy is approximated through an infinite expansion in terms of the density distribution. 
 Due to the high computational effort required to converge the higher order terms, the entropy expansion is usually truncated after the first or, sometimes, the second \cite{Nguyen2016-gist-second-order,Waibl2022-gist-solvents} term, thereby introducing an error.
+%TODO add MIE citation and context
 Furthermore, IST does not consider the internal degrees of freedom of the solvent.
@@
 
-	\label{tab:ref_values}
 	\caption{Reference number density $\rho_0$ and mean solvent-solvent energy $E_{ww, norm}$ for various water models from NPT simulations of N$_{WAT}$ water solvent molecules at \SI{300}{\kelvin} and 1bar.}\label{tab:ref_densities}
@@
 The Amber manual provides reference densities and energies for several solvents.
-Furthermore, a large variety of reference values calculated with GPU-GIST are listed in table \ref{tab:ref_values}.
+Furthermore, a large variety of reference values calculated with GPU-GIST are listed in table \ref{tab:ref_densities}. 
 However, the exact energy values are different when using PME-GIST, and also depend on the box size.

### 2024-06-20 manuscript/manuscript.tex
Change findcentroid.py name, fix angstrom naming

@@
 
-Additionally, the GIST calculation should usually be preceded by preprocessing steps, such as centering the trajectory or reimaging to account for periodic boundary conditions.
+Additionally, the GIST calculation should usually be preceded by preprocessing steps, such as centering the solute molecule in the unit cell or reimaging to account for periodic boundary conditions.
 Often, the region of interest will be defined as the surrounding of a molecule, which might for example be a ligand in the binding pocket or the whole protein. 
-Using \inlinecode{cpptraj's } \inlinecode{bounds} command, the region of interest can be output to a file. For example, here is the command for a region of 10\AA{} around the ligand LIG:
+Using \inlinecode{cpptraj's } \inlinecode{bounds} command, the region of interest can be output to a file. For example, here is the command for a region of \SI{10}{\angstrom} around the ligand LIG:
 
@@
 
-In addition, a helper script called \inlinecode{FindCentroid.py} for the same purpose is supplied with this tutorial, and can be used as follows:
+In addition, a helper script called \inlinecode{findcentroid.py} for the same purpose is supplied with this tutorial, and can be used as follows:
 
 \begin{lstlisting}[style=bash]
-python FindCentroid.py -i structure.pdb
+python findcentroid.py -i structure.pdb
 \end{lstlisting}
@@
 		\midrule
-		gridcntr    & <y> <y> <z> & Coordinates of the grid center in \AA & 0 0 0  \\
+		gridcntr    & <y> <y> <z> & Coordinates of the grid center in \AA{} & 0 0 0  \\
 		griddim     & <nx> <ny> <nz> & Number of voxels per axis & 40 40 40\\
-		gridspacn & <dval> &Grid spacing / voxel dimensions in  \AA & 0.5    \\
-		refdens         & <rdval> & Reference density of the bulk water in molecules/\AA$^3$ . & 0.0334    \\
+		gridspacn & <dval> &Grid spacing / voxel dimensions in  \AA{} & 0.5    \\
+		refdens         & <rdval> & Reference density of the bulk water in molecules/\AA{}$^3$ . & 0.0334    \\
 		temp & <tval> & System temperature in Kelvin & 300 \\
@@
 		Water Model       & $\rho_0$ & $E_{ww, norm}$ & $N_\text{WAT}$ \\
-        & $/\text{\AA}^{-3}$ & $/\text{kcal}\cdot \text{mol}^{-1}$ & \\
+        & $/\text{\AA{}}^{-3}$ & $/\text{kcal}\cdot \text{mol}^{-1}$ & \\
 		\midrule
@@
 For periodic boundary conditions, the size of the solvent box should be large enough for the solvent to behave bulk-like so as to reduce artifacts from solvent-mediated interactions between the periodic images.
-This corresponds to a distance between solute and box boundary of at least 12\AA{} for water, though we recommend 15\AA{} or higher.
+This corresponds to a distance between solute and box boundary of at least \SI{12}{\angstrom} for water, though we recommend \SI{15}{\angstrom} or higher.
+%TODO add reference for 12AA, maybe Hannes paper?
 Carefully check the employed force field, protonation state and partial charges.
@@
 Note that the offset is given in terms of number of additional voxels, not length dimension.
-By varying the atom mask in the \inlinecode{bound} command, it is possible to center the grid on a region of interest, similar to the \software{FindCentroid.py} script.
+By varying the atom mask in the \inlinecode{bound} command, it is possible to center the grid on a region of interest, similar to the \software{findcentroid.py} script.
 
@@
 plt.legend()
-plt.xlabel("distance cutoff [Å]")
+plt.xlabel("distance cutoff []")
 plt.ylabel("free energy contribution [kcal/mol]")
@@
 Right: The hydration contribution to binding, evaluated as the difference between the lines in the left panel.
-The vertical dashed line is at 12 \AA\ and represents the chosen integration cutoff.}\label{fig-dg-sums}
+The vertical dashed line is at \SI{12}{\angstrom} and represents the chosen integration cutoff.}\label{fig-dg-sums}
 \end{figure}
@@
 \textbf{Grid Options} &\\
-gridcntr <xval> <yval> <zval> & Coordinates in Å of the center of the grid. Default is 0.0, 0.0, 0.0. \\
+gridcntr <xval> <yval> <zval> & Coordinates in \AA{} of the center of the grid. Default is 0.0, 0.0, 0.0. \\
 griddim <xval> <yval> <zval> & Grid dimensions in voxels along each coordinate axis. Default is 40, 40, 40.\\
 gridspacn <val> & Grid spacing (linear dimension of each voxel) in 
-Angstroms. Default is 0.5 Å.\\
+Angstroms. Default is \SI{0.5}{\angstrom}.\\
 rmsfit <mask> & If specified, grid will be rotated and translated to follow the atoms selected by mask. \\
@@
 refdens <val> & Reference density of bulk solvent, used in computing 
-$g_O, g_H$, and the translational entropy. Default is 0.0334 molecules/Å3. \\
+$g_O, g_H$, and the translational entropy. Default is 0.0334 molecules/\si{\cubic\angstrom}. \\
 temp <val> & Temperature of the input trajectory. \\
@@
 
-neighborcut <val> & Cutoff in Å for determining solvent O-O neighbors. Default is 3.5 Å. \\
+neighborcut <val> & Cutoff in \si{\angstrom} for determining solvent O-O neighbors. Default is \SI{3.5}{\angstrom}. \\
 
@@
 pme & Use particle mesh Ewald for the non-bonded electrostatics calculation. The van der Waals energy will be calculated using a long-range correction for periodicity.\\
-\qquad cut <val> & Direct space cutoff for pme. Default is 3.5 Å.\\
+\qquad cut <val> & Direct space cutoff for pme. Default is \SI{3.5}{\angstrom}.\\
 \qquad dsumtol <val> & Direct sum tolerance used to determine Ewald coefficient. Default is 0.00001.\\
 \qquad ewcoeff <val> & Ewald coefficient in 1/Å. \\
-\qquad erfcdx <val> & Spacing to use for the ERFC splines. Default is 0.0002 Å. \\
+\qquad erfcdx <val> & Spacing to use for the ERFC splines. Default is \SI{0.0002}{\angstrom}. \\
 \qquad skinnb & Used to determine pairlist atoms (added to cut, so pairlist cutoff is cut + skinnb); included in order to maintain consistency with results from sander. \\
@@
 \qquad nopl & Do not use pair list for order calculation\\
-\qquad plcut <val> & Pair list cutoff for order calculation. Default is 10 Å\\
+\qquad plcut <val> & Pair list cutoff for order calculation. Default is \SI{10}{\angstrom}\\
 \bottomrule

### 2024-06-20 manuscript/manuscript.tex
Adress most of Franz's comments

@@
 \newcommand{\software}{\texttt}
+\newcommand{\code}{\texttt}
 \newcommand{\todo}{\textcolor{red}}
@@
 It is therefore important to set the reference number density $\rho_0$ according to the solvent model used in your simulation.
-The solvent-solvent enthalpy $E_{ww}$ also needs to be referenced, but this can be done in the post-processing stage.
+The solvent-solvent enthalpy $E_{ww}$ also needs to be referenced, but this is done in the post-processing stage.
 The calculation of these properties and the need for referencing is discussed in-depth in section \ref{sec:theory}.
@@
 Densities and reference energies of typical water models can be found in the \software{cpptraj} manual.
-We also provide the densities for water models commonly used with AMBER force fields in table \ref{tab:ref_densities}.
+We also provide the densities for water models commonly used with AMBER force fields in Table~\ref{tab:ref_densities}.
 These densities were calculated from \SI{100}{\nano\second} MD simulations at \SI{1}{\bar} and \SI{300}{\kelvin} using the Berendsen barostat and the Langevin thermostat.
@@
 The reference values can be calculated from MD trajectories of bulk solvent.
-From this, one then calculates the average energy per solvent molecule ($E_{ww, norm}$) and the average solvent density $\rho_0$.
+From this, one then calculates the density-weighted average energy per solvent molecule ($E_{ww, norm}$) and the average solvent density $\rho_0$.
 Generally, it is recommended that the size of the reference solvent box closely matches that of the production system.
@@
 If the same element occurs multiple times, the bulk density is scaled such that that the expectation value for the bulk is unity.
-For water, gO and gH are produced.}
+For water, \code{gO} and \code{gH} are produced.}
 	\item{[Esw] Mean solute-water interaction energy.}
@@
 
-We recommend using either the GPU or MPI-accelerated PME implementation of GIST.
+We recommend using either the GPU implementation or MPI-accelerated PME implementation of GIST.
 While the GPU version is faster when limited CPU resources are available, the PME implementation matches the energy of the Amber MD engine more closely.
-More recently, MPI support was added for both the energy and entropy calculation\cite{Roe2023-mpi-gist}. 
-The MPI implementation scales well over multiple CPU cores and can be combined with either the GPU or PME implementations for maximum efficiency. 
+The MPI implementation \cite{Roe2023-mpi-gist} scales well over multiple CPU cores and can be combined with either the GPU or PME implementations for maximum efficiency. 
 A speed comparison of the various parallelization implementations is shown in \cite{Roe2023-mpi-gist}.
@@
 The GIST data can be accessed and modified from the gist object in various ways.
-Gisttools provides access to single or multiple voxels through a .loc property similiar to the pandas library. 
+Gisttools provides access to single or multiple voxels through a \code{.loc} property similiar to the pandas library. 
 GIST quantity data for all voxels can be accessed by directly indexing the gist object, either by providing a single name or a list of names.
-Indexing with .loc works by providing a voxel index (either a index number or list of numbers) and optionally a column index (column name or list of names).
+Indexing with \code{.loc} works by providing a voxel index (either a index number or list of numbers) and optionally a column index (column name or list of names).
 Indices start from 0 and indexing will return a single entry, pandas series or pandas dataframe depending on the query:
@@
 The Amber manual provides reference densities and energies for several solvents.
-Furthermore, a large variety of reference values calculated with GPU-GIST are listed in table \ref{tab:ref_densities}. 
+Furthermore, a large variety of reference values calculated with GPU-GIST are listed in Table~\ref{tab:ref_densities}. 
 However, the exact energy values are different when using PME-GIST, and also depend on the box size.
@@
 	\caption{Convergence of $\dgsolv$ and its contributions with increasing distance to the solute molecule.
-All quantities cumulative, i.e., summed up to the respective radius.
+All quantities are cumulative, i.e., summed up to the respective radius.
 Computed from the biotin calculation.}\label{fig_radial_convergence}
@@
 print(pd.DataFrame(results))
-
-\end{lstlisting}
+\end{lstlisting}
+%TODO: This has a higher indent than the other python snippets in the pdf, somehow. Unsure why...
 You will find that the energy disfavors binding, because we do not yet include the interaction energy between biotin and streptavidin.
@@
 However, this should be done \emph{before} the MD, since rotating the trajectory damages the periodic box information.
+%TODO: Ask Dan if this is still the case
 
@@
 Note that, e.g., \inlinecode{Esw\_mol\_NA} contains the interactions of every ``\inlinecode{NA}'' solvent molecule with the solute molecule, and \inlinecode{Esw\_mol\_NA} contains the interactions among ``\inlinecode{NA}'' solvent molecules as well as their interactions with all other solvents, divided by 2 to account for double counting.
-Furthermore, the first solvent molecule in this list will be treated as the main solvent, for which first-order entropy values will be computed.
+The entropy will be computed using only the density of the first solvent molecule.
 
@@
 Additionally, to compute the first-order entropy of the ions, as well as an approximate second order entropy, Python code is available on \url{https://github.com/liedllab/second-disorder}.
-
+%TODO: Add recommendations how to handle charges/counterions to text (not necessarily here)
 \section{Visualization}
@@
 
-\subsection{Solvent Accessible Surface Area (SASA)}
-
-The solvent accessible surface (SASA) represents the interface between regions that are occupied by the solute molecule, and those which can be accessed by the solvent.
-GIST provides a very natural way of creating a SASA using an isosurface of the oxygen density gO.
+\subsection{Solvent Accessible Surface (SAS)}
+
+The solvent accessible surface (SAS) represents the interface between regions that are occupied by the solute molecule, and those which can be accessed by the solvent.
+GIST provides a very natural way of creating a SAS using an isosurface of the oxygen density \code{gO}.
 It can be visualized in PyMOL.
@@
 	\includegraphics[width=1.0\linewidth]{figures/sasa-simple.png}
-	\caption{SASA for streptavidin focused on the binding pocket}\label{fig-streptavidin_sasa}
+	\caption{SAS for streptavidin focused on the binding pocket}\label{fig-streptavidin_sasa}
 \end{figure}
 
-Here, the SASA was generated from a simulation of water around streptavidin without biotin, leaving the binding site solvent-accessible.
+Here, the SAS was generated from a simulation of water around streptavidin without biotin, leaving the binding site solvent-accessible.
 
@@
 def select(traj, sel):
-	# Slice a Trajectory by selection mask.
+	""" 
+	Slice a Trajectory by selection mask. 
+	"""
 	return traj.atom_slice(traj.top.select(sel))
@@
 
-### Set voxels within 3.5A to 1, write out dx file
+### Set voxels within 3.5A of biotin to 1, write out dx file
 compl.loc[ind, 'BP'] = 1
@@
 \end{figure}
-
+%TODO: Hide capping group
 \section{Theory}
@@
 Here, we only provide a short overview of the theory behind GIST.
-For more detailed information, we recommend one of the more recent publications on developments in GIST \cite{Kraml2020,Chen2021}.
+For more technical information (such as implementation details), we recommend one of the more recent publications on developments in GIST \cite{Kraml2020,Chen2021,Roe2023-mpi-gist}.
 
@@
 		\begin{itemize}
-			\item GPU, direct space: generally the fastest
+			\item CPU: slow, most general (can write out $E_{ij}$ matrices)
 			\item CPU, PME: highly consistent with Amber MD, fast
-			\item CPU: slow, most general (can write out $E_{ij}$ matrices)
+			\item GPU, direct space: fastest for a single CPU core
 		\end{itemize}
+
+	All methods profit from MPI-parallelization.   
+	PME with multiple CPU cores (>4) is usually the fastest method.
 	\end{checklist}
@@
 			\item Insufficient sampling leads to high noise in the results. Use higher numbers of frames.
-			\item Autocorrelation between frames leads to negative entropy in bulk. Increase simulation time and/or reduce the number of frames.
+			\item Correlation between frames leads to negative entropy in bulk. Increase simulation time and/or reduce the number of frames.
 		\end{itemize}
@@
 name <dataset name> & Name for output datasets in cpptraj. \\
-prefix <prefix> & Output file name prefix. Default is “gist”.\\
-ext <extension> & Output grid file name extension. Default is  “.dx”. \\
-out <file> & Name of the main GIST output file. If not specified set to “<prefix>-output.dat“.\\
+prefix <prefix> & Output file name prefix. Default is 'gist'.\\
+ext <extension> & Output grid file name extension. Default is  '.dx'. \\
+out <file> & Name of the main GIST output file. If not specified set to '<prefix>-output.dat'.\\
 info <file> & Name of main GIST info file. If not specified info is written to standard output. \\
-floatfmt <fmt> & Format for floating point values in GIST output file. Options are 'double','scientific' or 'general'. Default chooses 'fixed' or 'scientific' automatically. \\
+floatfmt <fmt> & Format for floating point values in GIST output file. Options are 'double', 'scientific' or 'general'. Default chooses 'fixed' or 'scientific' automatically. \\
 floatwidth <val>& Width of floating point values in GIST output file. Default is no width restriction. \\
@@
 nnsearchlayers <val> & Layers of neighboring voxels considered in nearest neighbor search.
-Higher values may improve entropy convergence for little sampling or fine grid spacings, but increase the calculation time (default 1).
+Higher values may improve entropy convergence for little sampling or fine grid spacings, but increase the calculation time.
+Default is 1.
 \\
@@
 \textbf{Output}\\
-GIST calculations write a variety of data sorted by voxel into an outputfile specified by the "out" keyword. 
-Some of the output data is also automatically written to "open data explorer" (.dx) files for convenient visualisation in software such as VMD or PyMOL.
+GIST calculations write a variety of data sorted by voxel into an outputfile specified by the 'out' keyword. 
+Some of the output data is also automatically written to 'open data explorer' (.dx) files for convenient visualisation in software such as VMD or PyMOL.
 Note that some columns will be written out both with '\_norm' and '\_dens' suffixes, refereing to normalization by solvent molecule or voxel volume respectively.
@@
 \section*{Author Contributions}
+
 %%%%%%%%%%%%%%%%
@@
 
-\section*{Other Contributions}
-%%%%%%%%%%%%%%%
-% You should include all people who have filed issues that were
-% accepted into the paper, or that upon discussion altered what was in the paper.
-% Multiple significant contributions might mean that the contributor
-% should be moved to authorship at the discretion of the a
-%
-% See the policies ``Policies on Authorship'' section of https://livecoms.github.io for
-% more information on deciding on authorship and author order.
-%%%%%%%%%%%%%%%
-
-(Explain the contributions of any non-author contributors here)
-% We suggest you preserve this comment:
-For a more detailed description of contributions from the community and others, see the GitHub issue tracking and changelog at \githubrepository.
-
 \section*{Potentially Conflicting Interests}

### 2024-06-20 manuscript/manuscript.tex
Add author contrributions

@@
 \author[1]{Monica L Fern{\'a}ndez-Quintero}
-%\author[1,2]{Firstname Middlename Familyname}
-%\author[2]{Firstname Initials Surname}
 \author[2]{Steven Ramsey}
@@
 
-\corr{klaus.liedl@uibk.ac.at}{KRL}  % Correspondence emails.  FMS and FS are the appropriate authors initials.
+\corr{klaus.liedl@uibk.ac.at}{KRL}
 \corr{mgilson@health.ucsd.edu}{MKG}
@@
 
+%TODO Add ORCIDS
 \orcid{Author 1 name}{AAAA-BBBB-CCCC-DDDD}
@@
 \contrib[\authfn{1}]{These authors contributed equally to this work}
-%\contrib[\authfn{2}]{These authors also contributed equally to this work}
-
-%\presentadd[\authfn{3}]{Department of General, Inorganic and Theoretical Chemistry, University of Innsbruck, Austria}
-%\presentadd[\authfn{4}]{Department, Institute, Country}
 
@@
 %%%%%%%%%%%%%%%%
-
-(Explain the contributions of the different authors here)
+All authors were involved in reviewing and editing the original manuscript. \\  
+VJH and FW conceptualized the research and co-wrote the initial draft.\\
+VM contributed the original drafts of sections \ref{sec:theory} and \ref{sec:visualization}.\\
+HC helped with code curation and validation.\\
+MLFQ helped with conceptualization and gave writing input for the initial draft. \\
+SR and DR programmed cpptraj's GIST implementation and provided technical input.\\  
+KRL, MKG and TK were involved with conceptualization, funding acquisition, project management and supervision.
 

### 2024-06-24 manuscript/manuscript.tex
Update author contributions

@@
 SR and DR programmed cpptraj's GIST implementation and provided technical input.\\  
-KRL, MKG and TK were involved with conceptualization, funding acquisition, project management and supervision.
+KRL, MKG and TK were involved with conceptualization, funding acquisition, project management and supervision.\\
 

### 2024-06-27 manuscript/manuscript.tex
Add MIE, MIST and water relabeling to intro

@@
 The main limitation of those methods is that the entropy is approximated through an infinite expansion in terms of the density distribution. 
+Within GIST and other more recent methods, these entropy contributions are calulated via a k-nearest-neighbor (kNN) approach.  
 Due to the high computational effort required to converge the higher order terms, the entropy expansion is usually truncated after the first or, sometimes, the second \cite{Nguyen2016-gist-second-order,Waibl2022-gist-solvents} term, thereby introducing an error.
-%TODO add MIE citation and context
+While there are approximations available to deal with those problems, most noteworthy the mutual information expansion (MIE) and Maximum Information Spanning Tree (MIST) approaches, they are not currently implemented in GIST.
+Recent kNN implementations to calculate solvation entropy further improve phase space sampling by relabeling the solvent molecules in the simulation, treating the solvent molecules as distinguishable \cite{reinhard2007estimation,heinz2019computing,heinz2021permut,fogolari2021optimal}.For a recent review of kNN methods to calculate entropy in a molecular context, see \cite{fogolari2024k}.
+
 Furthermore, IST does not consider the internal degrees of freedom of the solvent.
@@
 For water, \SI{0.5}{\nano\second} is usually sufficient to equilibrate the density, but narrow binding pockets can take longer.
-
 A well-tested equilibration protocol can be found in the literature \cite{Roe2020-equilibration}.

### 2024-06-27 manuscript/manuscript.tex
Removed unnecessary paragraph formating

--- 2024-06-24 manuscript/manuscript.tex
@@
 The main limitation of those methods is that the entropy is approximated through an infinite expansion in terms of the density distribution. 
+Within GIST and other more recent methods, these entropy contributions are calulated via a k-nearest-neighbor (kNN) approach.  
 Due to the high computational effort required to converge the higher order terms, the entropy expansion is usually truncated after the first or, sometimes, the second \cite{Nguyen2016-gist-second-order,Waibl2022-gist-solvents} term, thereby introducing an error.
-%TODO add MIE citation and context
+While there are approximations available to deal with those problems, most noteworthy the mutual information expansion (MIE) and Maximum Information Spanning Tree (MIST) approaches, they are not currently implemented in GIST.
+Recent kNN implementations to calculate solvation entropy further improve phase space sampling by relabeling the solvent molecules in the simulation, treating the solvent molecules as distinguishable \cite{reinhard2007estimation,heinz2019computing,heinz2021permut,fogolari2021optimal}.For a recent review of kNN methods to calculate entropy in a molecular context, see \cite{fogolari2024k}.\\
 Furthermore, IST does not consider the internal degrees of freedom of the solvent.
 Therefore, IST-based approaches can only be used with sufficiently rigid solvents.
-Nevertheless, they have been applied on a wide variety of problems involving both small molecules and biological systems as solute molecules.
-
+Nevertheless, they have been applied on a wide variety of problems involving both small molecules and biological systems as solute molecules.\\
 GIST is an implementation of IST on a three-dimensional grid.
@@
 This implies a higher versatility, but also puts more responsibility on the user in terms of correct post-processing.
-
 Since the density is computed from an explicit solvent MD simulation, GIST provides an accurate picture of solvation at an atomistic level, even for demanding systems such as drug binding sites in proteins.
@@
 Additionally, we compute the binding contribution of the free energy of solvation, which requires accurate post-processing of the GIST outputs to avoid unfavorable summation of bias.
-We believe that this example study covers a wide range of applications.
-
+We believe that this example study covers a wide range of applications.\\
 Furthermore, we discuss several technical aspects of GIST studies, such as the normalization of voxel values and how to avoid double counting of energy contributions. 
 Additionally, we provide a Python library (\software{gisttools}) to unify the analysis of GIST data produced by various versions of GIST and make the post-processing of GIST results more accessible.
-
 After completing this tutorial, we expect the reader to be able to run their own GIST study.
@@
 \subsection{Background knowledge}
-This tutorial is aimed at users with a solid knowledge of molecular dynamics (MD) simulations.
-
+This tutorial is aimed at users with a solid knowledge of molecular dynamics (MD) simulations.\\
 The user should be able to run MD simulations and analyze the resulting trajectories. 
@@
 The main implementation of GIST is part of the MD analysis software \software{cpptraj} released with AmberTools \cite{amber24, Case2023-ambertools}.
-
 A recent version of \software{cpptraj} should be used. We recommend using at least version 6.15.2, which can be obtained from GitHub (\url{https://github.com/Amber-MD/cpptraj}).
@@
 \end{itemize}
-
 If the reader prefers to skip the MD and GIST calculation, we also provide the GIST output files with this tutorial, such that the post-processing can be done without any expensive calculations.
@@
 For the visualization of three-dimensional contributions we require molecular visualization software that can handle \software{OpenDX} files, such as \software{PyMol} \cite{pymol} or \software{VMD} \cite{vmd}.
-
 Most of the visualisations and results in this tutorial can also be obtained using the \software{gistpp} program, which is available from \url{https://github.com/KurtzmanLab/Gist-Post-Processing} (code and documentation).
@@
 \end{lstlisting}
-
 Additionally, the GIST calculation should usually be preceded by preprocessing steps, such as centering the solute molecule in the unit cell or reimaging to account for periodic boundary conditions.
@@
 \end{lstlisting}
-
 This command tells \software{cpptraj} to run a GIST calculation using the specified region and print the output in a file called \inlinecode{gist.dat}.
@@
 A simulation length of at least \SI{20}{\nano\second} is recommended to obtain statistically independent snapshots of the water solvent molecules. 
-
 An example Amber input file for GIST might look like this:
@@
 \end{lstlisting}
-
 This file will run a \SI{100}{\nano\second} NPT MD simulation while keeping all atoms except for hydrogens and waters restrained.
 Note that these are very standard MD settings except for the restraints.
-
 Run the MD using \software{pmemd.cuda}:
@@
 \end{lstlisting}
-
 In this code snippet, the first two lines load the topology and coordinates of the first frame of our MD simulation. 
@@
 Reimage and center the solute molecule as above, to verify that its position matches the grid parameters you decided on.
-
 We recommend using either the GPU implementation or MPI-accelerated PME implementation of GIST.
@@
 Additionally as outlined in section \ref{sec:running_GIST}, various quantities are written to OpenDX files.
-
 To interact with your GIST data, we recommend the gisttools library for python3. Open the GIST output with gisttools:
@@
 \end{lstlisting}
-
 Gisttools provides access to all quantities listed in the GIST output file.
@@
 Information about the grid is stored in the grid property of the gist object, structural data is stored as an \software{mdtraj} trajectory in the struct property:
-
+%TODO we are missing some code here
 Gisttools also provides OpenDX writing capabilities to visualize the calculated or modified GIST data.
@@
 \end{lstlisting}
-
 Section~\ref{sec:visualization} discusses the visualization of OpenDX files.
@@
 For cases where the exact occupation of such sites is unknown, it may be helpful to include grand-canonical Monte Carlo (GCMC) or nonequilibrium candidate Monte Carlo (NCMC) sampling steps in the equilibration or production stage \cite{Ge-GCMC2022,Melling-NCMC2023}.
-
 It is good practice to check your restrained MD simulation or solvent populations obtained by GIST to confirm whether all relevant sites were properly solvated.
@@
 Alternatively, we provide a small script named \software{cpptraj\_remove\_mol.sh} for this procedure.
-
 After that, use \software{equilibration.py -R} to run short restrained 
@@
 \end{equation}
-
 Alternatively, the solvent-solvent energy in a large GIST grid converges to the bulk value at sufficient distance to the solute molecule. 
@@
 \end{lstlisting}
-
 The expected output of this code is shown in Figure~\ref{fig_ewwref}.
@@
 In \software{gisttools}, this is done simply by setting \inlinecode{gist.eww\_ref}.
-
 Now, sum all free energy contributions to obtain the spatially resolved $\dgsolv$.
@@
 \end{equation}
-
 %In \software{gistpp}, use the \inlinecode{add} command, with the same syntax as the \inlinecode{div} command used above.
@@
 In a plot like Figure~\ref{fig_ewwref} based on the \inlinecode{A\_dens} column, the values should tend to zero.
-
 However, it is more informative to plot the cumulative free energy contribution against the distance to the solute molecule, to check the convergence (see Figure~\ref{fig_radial_convergence}).
-
-It the curve flattens out, the converged value is your final $\dgsolv$\@.
+If the curve flattens out, the converged value is your final $\dgsolv$\@.
 If the curve diverges even for large integration distances you might need to tweak the $E_{ww}$ reference, or introduce a reference value for the entropy.
@@
 \end{lstlisting}
-
 The expected output of this code is shown in Figure~\ref{fig_radial_convergence}.
@@
 With \software{gisttools}, you can create one using e.g., \inlinecode{gist.save\_dx('A\_dens', 'A\_dens.dx')}.
-
 It might also be interesting to visualize the average energy of a water solvent molecule at each grid voxel.
@@
 The code is not explained in detail here, but should be clear after reading the previous sections.
-
 Load the files and double check the frame numbers and reference density.
@@
 \end{lstlisting}
-
 Assign reference energies and check them for plausibility.
@@
 \end{lstlisting}
-
 Subtract a reference entropy from the \inlinecode{dTSsix} columns.
@@
 \end{lstlisting}
-
 Compute the atom positions that define the integration region.
@@
 \end{lstlisting}
-
 Now, subtract the monomer histograms from the complex, and compute the sum of your property within some cutoff distance to the solute.
@@
 \end{lstlisting}
-
 You can also plot the difference between the complex and monomer contributions against the distance to biotin, using the complex coordinates for the holo structure:
@@
 \end{lstlisting}
-
 The expected result is shown in Figure~\ref{fig-dg-sums}.
-\begin{figure}
+\begin{figure}[h]
 	\centering
@@
 \end{figure}
-
 Compute the energy (\inlinecode{Eall}) and entropy (\inlinecode{dTSsix}) contributions separately.
@@
 \end{lstlisting}
-% test edit
-
 It has been shown \cite{Chen2021,Waibl2022-gist-solvents} that the solvation entropy in water is best approximated by 0.6 times the first order entropy.
@@
 \end{table}
-
 Even though streptavidin-biotin is known to be a very stable complex, the solvation energy favors the dissociation.
-
 This is expected: the binding of biotin to streptavidin is faciliated over a large number of H-bonds and over polar interactions. 
 As such, water favourably solvates both biotin and streptavidin and an energetic penalty is paid when displacing the water during the binding process. 
-
 In this case, we find that the energy of solvation and the internal energy of the complex more or less cancel out, indicating that binding is largely entropy-driven in this case.
@@
 To include the effect of lid closing in the binding process, free energy calculation methods such as umbrella sampling can be used.
-
 To speed up the calculation, smaller GIST grids could be used by focusing on the binding pocket, or by rotating the solute molecule by its principal axes to fit the cuboid grid more exactly.
@@
 In Reference~\cite{Kraml2020}, a GPU implementation of the energy calculation was presented, which typically increases the speed by 1--2 orders of magnitude.
-
 If \software{cpptraj} is compiled with GPU support, the energy calculation uses the GPU automatically, without any additional input.
@@
 Furthermore, PME-GIST offers the best agreement between GIST energies and those observed in an classical MD simulation.
-
 To run PME-GIST, the \inlinecode{pme} flag to \inlinecode{gist} can be used in \software{cpptraj}.
@@
 In references~\cite{Kraml2020,Kamenik2020-gist-macrocycles,Waibl2022-gist-solvents}, an extension of GIST towards solvents other than water is described.
-
 To run a GIST calculation in a solvent other than water, use \inlinecode{solute} with a \software{cpptraj} selection mask to select the solute molecule.
 All other molecules will be treated as solvent.
-If there are multiple solvent species, also use the \inlinecode{solventmols} flag as described in Section~\ref{sec-salt-water}.
-
+If there are multiple solvent species, also use the \inlinecode{solventmols} flag as described in Section~\ref{sec-salt-water}. \\
 The code will automatically choose three atoms per solvent species to define the solvent orientation, and print this selection to the output.
@@
 \end{lstlisting}
-
 Note that the central atom of the \inlinecode{rigidatoms} selection goes first.
@@
 This allows treating salting-out effects, although it was shown that the salting-out coefficient is over-predicted by GIST because of the first-order entropy approximation.
-Generally, GIST should be able to treat arbitrary solvent mixtures as long as each solvent molecule is sufficiently rigid.
-
+Generally, GIST should be able to treat arbitrary solvent mixtures as long as each solvent molecule is sufficiently rigid. \\
 In the \software{cpptraj} implementation of GIST, the \inlinecode{solute} keyword specifies which components of the system are treated as a solute molecule.
@@
 Note that, e.g., \inlinecode{Esw\_mol\_NA} contains the interactions of every ``\inlinecode{NA}'' solvent molecule with the solute molecule, and \inlinecode{Esw\_mol\_NA} contains the interactions among ``\inlinecode{NA}'' solvent molecules as well as their interactions with all other solvents, divided by 2 to account for double counting.
-The entropy will be computed using only the density of the first solvent molecule.
-
+The entropy will be computed using only the density of the first solvent molecule. \\
 For instance, a GIST calculation in salt water, including the first-order water entropy, can be run as follows:
@@
 \end{lstlisting}
-
 The current implementation of the entropy calculation in \software{cpptraj} requires at least 3 atoms in the main solvent.
@@
 \end{lstlisting}
-
 A more sophisticated visualization, including the energy, entropy, and free energy of removing a single water solvent molecule, is found in Figure~\ref{fig-binding-pocket.png}
@@
 Here, the SAS was generated from a simulation of water around streptavidin without biotin, leaving the binding site solvent-accessible.
-
 \subsection{Defining the binding pocket}
 Specific regions in a protein can be defined to calculate the thermodynamic quantities for that region.
-This is significant specially when calculating the properties of water that will be expelled when a ligand binds to the protein.
-
+This is significant specially when calculating the properties of water that will be expelled when a ligand binds to the protein.\\
 For example, voxels that are part of the binding pocket can be defined using a ligand solute molecule and a distance criteria. 
-
 In section~\ref{sec:binding_contributions} we show how to integrate over regions such as the binding pocket.
@@
 %The second order is the pair density, i.e., the probability of finding two solvent molecules in positions $\textbf{r}$ and $\textbf{r}^\prime$.
-
 The solvent density is expressed relative to the bulk density $\rho^0$. %$g_\textit{sw}\left(\textbf{r},\omega \right)$ is the solute-water 
@@
 \subsubsection{Entropy Calculations in Cpptraj}
-In \software{cpptraj}, the calculation of solvation entropy is handled by two methods.
-
+In \software{cpptraj}, the calculation of solvation entropy is handled by two methods.\\
 In the first method, the solvation entropy is broken down into translational and orientational contributions.
@@
 \end{equation}
-This is exact assuming that the distribution of the orientation $\omega$ is constant within a voxel $k$, but might converge slower than computing the entropy from the combined space (see below).
-
+This is exact assuming that the distribution of the orientation $\omega$ is constant within a voxel $k$, but might converge slower than computing the entropy from the combined space (see below).\\
 The second method directly calculates the solvation entropy by evaluating the six-dimensional integral (3 for position and 3 for orientation).
@@
 \end{equation}
-
 A first nearest neighbor (NN) approach is used to evaluate the density.
 The densities are computed relative to a homogeneous distribution of solvent molecules, given a bulk density $\rho^0$.
-
 The average entropy contributions per solvent molecule in voxel $k$ are calculated as
@@
 \end{equation}
-
 The solute-solvent energy $E_{sw}$ can be expressed in terms of the solvent density and the potential $U_{sw}(\textbf{r},\omega)$ induced by the solute molecule.
@@
 \end{equation}
-
 To localize $E_{sw}$ on the three-dimensional grid, every energy term is assigned to the voxel $k$ holding the solvent, and the expectation value in Equation~\ref{eq-esw} is computed per voxel.
@@
 Hence, it can be approximated by local spatial integrals. 
-
 The solvent-solvent energy $E_{ww}$ is computed similarly.
@@
 \end{equation}
-
 However, when water is replaced from a small region $R$ of interest, such as a single water solvent molecule, almost all interactions are with water solvent molecules outside of this region.
@@
 \end{equation}
-
 When the region $R$ comprises more than one solvent molecule, interactions within this region are double-counted while interactions to the outside are not.
@@
 While this is supported by the standard GIST implementation (using the \inlinecode{doeij} flag), it is rarely done due to the large size of the $E_{ww,k\;l}$ matrix.
-
 Integrating GIST values over the whole grid corresponds to a process where all the solvent is removed into bulk.
@@
 \end{equation}
-
 During the electrostatics calculation, the system is treated as periodic and the energy is split into a short-range term $E_\text{dir}$, which is calculated in direct space using a distance cutoff, and a long-range term $E_\text{rec}$, which is calculated in reciprocal space.

### 2024-06-28 manuscript/manuscript.tex

@@
 For periodic boundary conditions, the size of the solvent box should be large enough for the solvent to behave bulk-like so as to reduce artifacts from solvent-mediated interactions between the periodic images.
-This corresponds to a distance between solute and box boundary of at least \SI{12}{\angstrom} for water, though we recommend \SI{15}{\angstrom} or higher.
-%TODO add reference for 12AA, maybe Hannes paper?
+This corresponds to a distance between solute and box boundary of at least \SI{12}{\angstrom} for water\cite{Chen2021,Waibl2022-gist-solvents} though we recommend \SI{15}{\angstrom} or higher.
 Carefully check the employed force field, protonation state and partial charges.

### 2024-06-28 manuscript/manuscript.tex
Change abstract

@@
 \begin{abstract}
-Grid inhomogeneous solvation theory (GIST) is a method to compute the free energy of solvation of a solute molecule on a 3-dimensional grid based on molecular dynamics (MD) simulations.
+Grid inhomogeneous solvation theory (GIST) is a method to compute the free energy of solvation of a solute molecule on a 3-dimensional grid based on sampling from molecular dynamics (MD) simulations.
 The high spatial resolution of the GIST output, as well as the decomposition into energy and entropy contributions, allow for highly detailed analyses on both proteins and small molecules.

### 2024-06-28 manuscript/manuscript.tex
Change Figure 1 label

@@
 Grid inhomogeneous solvation theory (GIST) is a method to compute the free energy of solvation of a solute molecule on a 3-dimensional grid based on sampling from molecular dynamics (MD) simulations.
-The high spatial resolution of the GIST output, as well as the decomposition into energy and entropy contributions, allow for highly detailed analyses on both proteins and small molecules.
+The high spatial resolution of the GIST output, as well as the decomposition into energy and entropy contributions, allow for highly detailed analyses of solvation around both proteins and small molecules contexts.
 However, this versatility also comes with a significant entry barrier for new users.
@@
     \caption{Schematic showing regions of high water density around a carbazole solute molecule.
-The energy is computed from the intermolecular interactions of a force field, while the entropy is approximated from the 3-dimensional solvent distribution around the solute molecule.}
+The energy is computed from the intermolecular interactions of a force field, while the entropy is approximated from the 3-dimensional solvent distribution around the solute molecule.
+Note that while this depiction focuses of regions of high solvent density, regions with lower density can also contain meaningful contributions to the energy and entropy.}
     \label{fig:carbazole}

### 2024-06-28 manuscript/manuscript.tex
Updates from Overleaf

--- 2024-06-28 manuscript/manuscript.tex
@@
 \usepackage{siunitx}
-\usepackage{float}
-\usepackage{placeins}
 \DeclareSIUnit\Molar{M}
@@
 
+\author[1\authfn{1}]{Franz Waibl}
 \author[1\authfn{1}]{Valentin J Hoerschinger}
-\author[1\authfn{1}]{Franz Waibl}
 \author[2]{Vjay Molino}
-\author[2]{Helmut Carter}
 \author[1]{Monica L Fern{\'a}ndez-Quintero}
-\author[2]{Steven Ramsey}
+%\author[1,2]{Firstname Middlename Familyname}
+%\author[2]{Firstname Initials Surname}
 \author[4]{Daniel R Roe}
@@
 
-\corr{klaus.liedl@uibk.ac.at}{KRL}
+\corr{klaus.liedl@uibk.ac.at}{KRL}  % Correspondence emails.  FMS and FS are the appropriate authors initials.
 \corr{mgilson@health.ucsd.edu}{MKG}
@@
 
-%TODO Add ORCIDS
 \orcid{Author 1 name}{AAAA-BBBB-CCCC-DDDD}
@@
 \contrib[\authfn{1}]{These authors contributed equally to this work}
+%\contrib[\authfn{2}]{These authors also contributed equally to this work}
+
+%\presentadd[\authfn{3}]{Department of General, Inorganic and Theoretical Chemistry, University of Innsbruck, Austria}
+%\presentadd[\authfn{4}]{Department, Institute, Country}
 
@@
 \newcommand{\software}{\texttt}
-\newcommand{\code}{\texttt}
 \newcommand{\todo}{\textcolor{red}}
@@
 \begin{abstract}
-Grid inhomogeneous solvation theory (GIST) is a method to compute the free energy of solvation of a solute molecule on a 3-dimensional grid based on molecular dynamics (MD) simulations.
+Grid inhomogeneous solvation theory (GIST) is a method to compute the free energy of solvation of a compound on a 3-dimensional grid based on molecular dynamics (MD) simulations.
 The high spatial resolution of the GIST output, as well as the decomposition into energy and entropy contributions, allow for highly detailed analyses on both proteins and small molecules.
@@
 
-In this tutorial, we aim to guide the reader through the most common steps involved in a GIST analysis using the streptavidin-biotin complex as an demonstrative system.
-Towards this, Jupyter notebooks and Python packages are provided to simplify the analysis.
+In this tutorial, we aim to guide the reader through the most common steps involved in a GIST analysis at the example of the streptavidin-biotin complex.
 Furthermore, we discuss the theory of GIST with a focus on practical aspects.
 We highlight potential pitfalls and show how to avoid technical difficulties.
-For this tutorial, familiarity with molecular dynamics simulations as well as the AmberTools package \cite{Case2023-ambertools} is assumed.
+We assume familiarity with molecular dynamics simulations as well as the AmberTools package.
 
@@
 Solvation thermodynamics govern many biological and chemical processes involving, e.g., aqueous solutions, hydrophobic effects, or partitioning between different environments.
-
-Interactions with water are of special importance as many biological processes occur in aqueous environment \cite{Privalov2017-water-review}.
-The free energy of solvation is defined as the reversible work needed to insert a solute molecule into the solvent from gas phase \cite{ben-naim-book}.
+Interaction with water are of special importance as many biological processes occur in aqueous environment \cite{Privalov2017-water-review}.
+The free energy of solvation is defined as the reversible work needed to insert a solute into the solvent \cite{ben-naim-book}.
 It is directly linked to the solubility of gases and drives many real-world effects such as hydrophobicity, ligand binding, and protein folding.
 
-Grid inhomogeneous solvation theory (GIST) \cite{Nguyen2012} is a method to compute the free energy of solvation, with applications in the context of drug design and biophysical property description.
-It computes the solvent density from an MD trajectory and estimates localized entropic and enthalpic contributions on a grid around the solute molecule.
-Since these contributions are spatially resolved, GIST improves interpretability and offers more detailed insights into the solvation of the investigated solute. 
-The entropy is computed from the density distribution of the solvent around the solute molecule, in a first-order approximation. A schematic of this is shown in Figure~\ref{fig:carbazole}. 
-The calculation of higher-order entropy terms is computationally demanding but was realized in recent works \cite{Nguyen2016-gist-second-order, Waibl2021-gist-salt}.
-Nevertheless, solvation free energies of small molecule solutes can be computed at high accuracy using linear scaling factors for the higher-order terms, such as $\Delta S = 0.6\, \Delta S^\text{1st-order}$ for water \cite{Chen2021,Waibl2022-gist-solvents}.
+Grid inhomogeneous solvation theory (GIST) is a method to compute the free energy of solvation, with applications in the context of drug design and biophysical property description.
+It estimates the solvent density from an MD trajectory and computes localized entropic and enthalpic contributions on a grid around the solute.
+The entropy is computed from the density distribution of the solvent around the solute, in a first-order approximation.
+A schematic of this is shown in Figure~\ref{fig:carbazole}.
+Nevertheless, solvation free energies of small molecules can be computed at high accuracy using linear scaling factors for the higher-order terms, such as $\Delta S = 0.6\, \Delta S^\text{1st-order}$ for water \cite{Chen2021,Waibl2022-gist-solvents}.
 
@@
     \includegraphics[width=\linewidth]{figures/carbazole-figure.png}
-    \caption{Schematic showing regions of high water density around a carbazole solute molecule.
-The energy is computed from the intermolecular interactions of a force field, while the entropy is approximated from the 3-dimensional solvent distribution around the solute molecule.}
+    \caption{Schematic showing regions of high water density around a carbazole molecule.
+The energy is computed from the intermolecular interactions of a force field, while the entropy is approximated from the 3-dimensional solvent distribution around the solute.}
     \label{fig:carbazole}
@@
 Since its initial development, GIST has been used to describe the thermodynamics of water displacement in protein binding pockets \cite{Nguyen2016-gist-second-order,Ramsey2016,Balius2017-gist-ligand-discovery,Hufner-Wulsdorf2020-gist-drug-design}.
-It has been used for a large-scale investigation of the hydration properties of proteins found in SARS-CoV-2 \cite{Olson2020-covid-gist}.
+It has also been deployed in systematically mapping the hydration properties of proteins found in SARS-CoV-2 \cite{Olson2020-covid-gist}.
 Furthermore, recent works used GIST to describe biophysical properties of serine proteases \cite{Kraml2019-gigist}, macrocycles \cite{Kamenik2020-gist-macrocycles}, or antibody fragements \cite{Waibl2021-gist-antibodies}.
@@
 
-In this tutorial, we aim to give the reader an introduction to GIST and its application to calculate solution thermodynamics properties of interest. 
-While previous publications \cite{Ramsey2016} and tutorials \cite{amber_tut_gist} treated how to setup and run GIST calculations in detail, the post-processing was often treated less rigorously.
+In this tutorial, we aim to give the reader an introduction into GIST and its application to calculate solution thermodynamics properties of interest. 
+While previous publications \cite{Ramsey2016} and tutorials\footnote{See \url{http://ambermd.org/tutorials/advanced/tutorial25/}} gave insightful pointers how to setup and run GIST calculations, the post-processing was often treated less rigorously.
 While building on these tutorials, we provide an updated guide to calculating solvation properties using GIST to address recent improvements.
-Furthermore, we aim to establish best practices for the analysis and interpretation of GIST calculations.
+Furthermore, we aim to establish a template for the analysis and interpretation of GIST calculations.
 
@@
 Data-driven methods include QSPR and UNIFAC \cite{Borhani2019-qspr,Fredenslund1975-unifac}.
-Polarizable continuum models (PCMs) \cite{Miertus1981-pcm,Klamt1993-cosmo} treat the solvent as a homogeneous phase that reacts to the electrostatic potential of the solute molecule \cite{Mennucci2010-pcm}.
+Polarizable continuum models (PCMs) \cite{Miertus1981-pcm,Klamt1993-cosmo} treat the solvent as a homogeneous phase that reacts to the electrostatic potential of the solute \cite{Mennucci2010-pcm}.
 %Molecular mechanics based approaches describe the underlying PES through empirically fit force-fields, which reduce the computational demand of these calculations tremendously.
 %This allows for the treatment of large biological systems such as proteins in aqueous solution.
-Methods such as MM/PBSA or MM/GBSA \cite{Sitkoff1994-pbsa,Kollman2000-mmpbsa} combine an implicit treatment of solvent electrostatics with structural sampling of the solute molecule.
+Methods such as MM/PBSA or MM/GBSA \cite{Sitkoff1994-pbsa,Kollman2000-mmpbsa} combine an implicit treatment of solvent electrostatics with structural sampling of the solute.
 They can quickly provide results for a large number of structures \cite{Genheden2015-mmpbsa-review}.
@@
 Methods derived from statistical mechanics bridge this gap by providing three-dimensional resolution.
-These methods most commonly are based on either the Ornstein–Zernike (OZ) equation, or \cite{Hansen2013-simple-liquids} inhomogeneous fluid solvation theory (IST) \cite{Lazaridis1998}.
+Two major flavors of methods in this group are those based on the Ornstein--Zernike (OZ) equation \cite{Hansen2013-simple-liquids} and those based on inhomogeneous fluid solvation theory (IST) \cite{Lazaridis1998}.
 The most important example of OZ-based methods is the reference interaction site model (RISM) \cite{Chandler1972-rism} as well as its extension to three dimensions (3D-RISM) \cite{Kovalenko1998-3drism}.
@@
 
-On the other hand, methods based on IST generally use a molecular solvent distribution obtained from an MD simulation to compute the free energy of hydration from its enthalpic and entropic contributions, providing easier interpretability of the resulting distributions.
+On the other hand, methods based on IST generally use a molecular solvent distribution obtained from an MD simulation to compute the free energy of hydration from its enthalpic and entropic contributions.
 Examples include grid inhomogeneous solvation theory (GIST) \cite{Nguyen2012,Ramsey2016}, WaterMap \cite{Young2007-watermap,Abel2008-watermap}, SSTMap \cite{Haider2018-sstmap}, STOW \cite{Li2012-stow} and Solvaware \cite{Huggins2016-solvaware}.
-The main limitation of those methods is that the entropy is approximated through an infinite expansion in terms of the density distribution. 
-Within GIST and other more recent methods, these entropy contributions are calulated via a k-nearest-neighbor (kNN) approach.  
-Due to the high computational effort required to converge the higher order terms, the entropy expansion is usually truncated after the first or, sometimes, the second \cite{Nguyen2016-gist-second-order,Waibl2022-gist-solvents} term, thereby introducing an error.
-While there are approximations available to deal with those problems, most noteworthy the mutual information expansion (MIE) and Maximum Information Spanning Tree (MIST) approaches, they are not currently implemented in GIST.
-Recent kNN implementations to calculate solvation entropy further improve phase space sampling by relabeling the solvent molecules in the simulation, treating the solvent molecules as distinguishable \cite{reinhard2007estimation,heinz2019computing,heinz2021permut,fogolari2021optimal}.For a recent review of kNN methods to calculate entropy in a molecular context, see \cite{fogolari2024k}.\\
+The main limitation of those methods is that they rely on an infinite expansion of the entropy in terms of the density distribution, which is truncated after the first or, sometimes, the second \cite{Nguyen2016-gist-second-order,Waibl2022-gist-solvents} term.
 Furthermore, IST does not consider the internal degrees of freedom of the solvent.
 Therefore, IST-based approaches can only be used with sufficiently rigid solvents.
-Nevertheless, they have been applied on a wide variety of problems involving both small molecules and biological systems as solute molecules.\\
+Nevertheless, they have been applied on a wide variety of problems involving both small molecules and biological systems.
+
 GIST is an implementation of IST on a three-dimensional grid.
-Unlike the related method WaterMap \cite{Abel2008-watermap,Young2007-watermap}, it provides solvation thermodynamics on an entire grid, rather than only high-density clusters in a binding pocket.
+It can be seen as a more general version of WaterMap \cite{Abel2008-watermap,Young2007-watermap} that provides solvation thermodynamics on the entire grid, rather than high-density clusters in a binding pocket.
 This implies a higher versatility, but also puts more responsibility on the user in terms of correct post-processing.
-Since the density is computed from an explicit solvent MD simulation, GIST provides an accurate picture of solvation at an atomistic level, even for demanding systems such as drug binding sites in proteins.
+
+Since the density is computed from an MD simulation, GIST can accurately treat confined water regions, which are challenging for implicit-solvent methods such as GBSA and self-consistent theories such as 3D-RISM \cite{Kovalenko1998-3drism}.
 While the need for MD simulations comes at a large computational cost, GIST is still more efficient than end-state approaches such as TI \cite{Kirkwood1935-ti} or FEP \cite{Zwanzig1954-reweighting}, with applications to large systems such as serine proteases \cite{Kraml2019-gigist} or antibody fragments \cite{Waibl2021-gist-antibodies}.
 On the other hand, the first-order approximation means that the solvation entropy is less accurately described than in TI or FEP \cite{Chen2021,Waibl2021-gist-salt}.
-Similar to the self-consistent integral theories, however, GIST only considers a single solute conformation and  multiple computations of representative structures are necessary to describe ensemble properties.
+%The 3D-RISM approach allows for a decomposition into entropic and enthalpic terms as well as a full localization of the underlying distributions to the solute structure.
+%Here you would explain what problem you are tackling and briefly motivate your work.
+%
+%In this particular template, we have removed most of the usage examples which occur in \texttt{sample-document.tex} to provide a minimal template you can modify; however, we retain a couple of examples illustrating more unusual features of our templates/article class, such as the checklists, and information on algorithms and pseudocode.
+%
+%Keep in mind, as you prepare your manuscript, that you should plan for a representative image  which will be used to highlight your article on the journal website and publications.
+Usually, this would be one of your figures, but it must also be uploaded separately upon article submission.
+We give specific guidelines for this image on the journal website in the section on article submission (see \url{https://livecomsjournal.github.io/authors/policies/index.html#article-submission}).
+%
+%Additionally, for well-formatted manuscripts, we recommend that you let LaTeX handle figure/table placement for you as much as possible, so please avoid specifying strenuous float instructions like `[h!]` and `[H]` as much as possible.
 
@@
 Additionally, we compute the binding contribution of the free energy of solvation, which requires accurate post-processing of the GIST outputs to avoid unfavorable summation of bias.
-We believe that this example study covers a wide range of applications.\\
+We believe that this example study covers a wide range of applications.
+
 Furthermore, we discuss several technical aspects of GIST studies, such as the normalization of voxel values and how to avoid double counting of energy contributions. 
 Additionally, we provide a Python library (\software{gisttools}) to unify the analysis of GIST data produced by various versions of GIST and make the post-processing of GIST results more accessible.
+
 After completing this tutorial, we expect the reader to be able to run their own GIST study.
@@
 \subsection{Background knowledge}
-This tutorial is aimed at users with a solid knowledge of molecular dynamics (MD) simulations.\\
+This tutorial is aimed at users with a solid knowledge of molecular dynamics (MD) simulations.
 The user should be able to run MD simulations and analyze the resulting trajectories.
-The MD simulations presented in this tutorial are run with the Amber simulation engine and analyzed using AmberTools \software{cpptraj}. 
-For a conceptual overview of running MD simulations with Amber, we recommend reading the Amber manual (\url{https://ambermd.org/Manuals.php}) \cite{amber24} and the Amber tutorials (\url{https://ambermd.org/Manuals.php}).
 Previous experience of GIST is not necessary, although we recommend reading the works referenced in section~\ref{sec:theory}.
@@
 Tutorials requiring specific software packages must provide instructions and files for the referenced version of the software.
-The main implementation of GIST is part of the MD analysis software \software{cpptraj} released with AmberTools \cite{amber24, Case2023-ambertools}.
+The main implementation of GIST is part of the MD analysis software \software{cpptraj} released with AmberTools \cite{amber22}.
+
 A recent version of \software{cpptraj} should be used. We recommend using at least version 6.15.2, which can be obtained from GitHub (\url{https://github.com/Amber-MD/cpptraj}).
@@
 The Amber simulation engine is used in this tutorial to run the MD simulations.
-While Amber is free for non-commercial use, it might not always be available in industrial settings.
+While Amber is free for non-commercial use, it might not always be available in industry settings.
 In that case, other MD engines can be used as long as topologies and trajectories can be produced in a format supported by \software{cpptraj}.
-If using GROMACS \cite{Abraham2015-gromacs,Pronk2013-gromacs}, we propose to prepare the topology with AmberTools \cite{Case2023-ambertools} and convert it using \software{acpype.py} \cite{Sousa_da_Silva2012-acpype}.
+If using GROMACS \cite{Abraham2015-gromacs,Pronk2013-gromacs}, we propose to prepare the topology with AmberTools and convert it using \software{acpype.py} \cite{Sousa_da_Silva2012-acpype}.
 %We should actually test that!
@@
 	\item \software{mdtraj} (\url{https://www.mdtraj.org/}, $\geq$ 1.9.7) \cite{McGibbon2015-mdtraj}
-	\item \software{matplotlib} (\url{https://matplotlib.org/}, $\geq$ 3.7.0) \cite{Hunter2007-matplotlib}
-	\item \software{numpy} (\url{https://numpy.org/}, $\geq$ 1.23.5) \cite{harris2020-numpy}
-	\item \software{pandas} (\url{https://pandas.pydata.org/}, $\geq$ 1.5.3) \cite{pandas-2023_8239932}
 \end{itemize}
+
 If the reader prefers to skip the MD and GIST calculation, we also provide the GIST output files with this tutorial, such that the post-processing can be done without any expensive calculations.
@@
 For the visualization of three-dimensional contributions we require molecular visualization software that can handle \software{OpenDX} files, such as \software{PyMol} \cite{pymol} or \software{VMD} \cite{vmd}.
+
 Most of the visualisations and results in this tutorial can also be obtained using the \software{gistpp} program, which is available from \url{https://github.com/KurtzmanLab/Gist-Post-Processing} (code and documentation).
 \software{Gistpp} can directly manipulate \software{OpenDX} files from the GIST output.
-Its usage and some examples were shown in a previous tutorial on GIST \cite{Ramsey2016}.
+Its usage and some examples were shown in a previous tutorial on GIST\cite{Ramsey2016}.
 
@@
 	\item \software{cpptraj}.
-	\item A topology of the solute molecule of interest in a solvent box.
+	\item A topology of the molecule of interest in a solvent box.
 	\item An MD trajectory of that system.
 \end{itemize}
-Note that the solute should not move significantly for a GIST calculation, since IST does not take the solute degrees of freedom into account.
-This is typically achieved by applying restraints on the solute heavy atoms during the MD simulation.
 GIST analysis can be divided in three steps: defining a region of interest, running GIST using \software{cpptraj}, and post-processing and analysis.
@@
 In order to run a GIST analysis, one must define the region where solvent properties will be calculated.
-This region is defined by a rectangular grid with given position, size, and voxel spacing.
-The default grid spacing of \SI{0.5}{\angstrom} provides reasonable discretization for many use-cases.
-While integrals of GIST quantities over the grid are independent of the voxel size, provided sufficient sampling, individual values converge slower at high resolutions.
+This region is represented by a rectangular grid with given position, size, and voxel spacing.
+The default value of \SI{0.5}{\angstrom} provides reasonable discretization for many use-cases.
+While integrals of GIST quantities over the grid are independent of the voxel size, individual values converge slower at high resolutions.
+
 The grid is defined by parameters of the \inlinecode{gist} command in \software{cpptraj}.
-The grid center is defined by \inlinecode{gridcntr}, the voxel number in x, y, and z direction by \inlinecode{griddim}, and the spacing (voxel side-length) in \AA{} by \inlinecode{gridspacn}:
+The grid center is defined by \inlinecode{gridcntr}, the voxel number in x, y, and z direction by \inlinecode{griddim}, and the spacing (voxel side-length) in \AA{} by \inlinecode{gridspacn}.
 
@@
 \end{lstlisting}
-Additionally, the GIST calculation should usually be preceded by preprocessing steps, such as centering the solute molecule in the unit cell or reimaging to account for periodic boundary conditions.
+
 Often, the region of interest will be defined as the surrounding of a molecule, which might for example be a ligand in the binding pocket or the whole protein.
-Using \inlinecode{cpptraj's } \inlinecode{bounds} command, the region of interest can be output to a file. For example, here is the command for a region of \SI{10}{\angstrom} around the ligand LIG:
-
+Using \inlinecode{cpptraj's } \inlinecode{bound} command, the region of interest can be output to a file.
+For example, here is the command for a region of 10\AA{} around the ligand LIG:
 \begin{lstlisting}[style=cpptraj]
@@
 \end{lstlisting}
-Note that the \inlinecode{bounds} command needs to be supplied with the exact same trajectory as used in the GIST calculation to generate correct boundaries.
-Therefore, any preprocessing steps need to match between the \inlinecode{bounds} command and the following \inlinecode{gist} calculation.
-
-In addition, a helper script called \inlinecode{findcentroid.py} for the same purpose is supplied with this tutorial, and can be used as follows:
+Note that any loading and preprocessing, such as imaging, needs to be run before this command and needs to be the exact same as is used for the \inlinecode{gist} command.
+Additionally, a helper script called \inlinecode{FindCentroid.py} for the same purpose is supplied with this tutorial, and can be used as follows:
 
 \begin{lstlisting}[style=bash]
-python findcentroid.py -i structure.pdb
+python FindCentroid.py -i structure.pdb
 \end{lstlisting}
@@
 The command needed to run GIST in cpptraj can be written in an input file (e.g gist.in) for convenience.
-The basic command, assuming a grid centered at coordinates (x=\SI{10.5}{\angstrom}, y=\SI{20.1}{\angstrom}, z=\SI{30.1}{\angstrom}) and a grid size of $30\times30\times30$ voxels with \SI{0.5}{\angstrom} spacing, is:
+The basic command, assuming a given grid center and a grid size of $30\times30\times30$ voxels with \SI{0.5}{\angstrom} spacing, is:
 
@@
 \end{lstlisting}
+
+Note that the solute should not move significantly for a GIST calculation, since IST does not take the solute degrees of freedom into account.
+This is typically achieved using restraints on the solute heavy atoms.
 This command tells \software{cpptraj} to run a GIST calculation using the specified region and print the output in a file called \inlinecode{gist.dat}.
-An overview of how to work with \software{cpptraj} is provided in the program's documentation \cite{cpptraj_doc}, AMBER tutorials \cite{amber_tut_cpptraj} and on the AMBER-hub website \cite{amber_hub}.
-
-To use OpenMP, MPI or GPU-accelerated (using CUDA) versions of cpptraj, a different executable name such as \inlinecode{cpptraj.OMP}, \inlinecode{cpptraj.MPI} or \inlinecode{cpptraj.cuda} is often used.
+You will need to load and image your trajectory before running this command.
+An overview of how to work with \software{cpptraj} is provided in the program's documentation \cite{cpptraj_doc}, AMBER tutorials \cite{amber_tut_cpptraj} or on the AMBER-hub website \cite{amber_hub}.
+
+To use OpenMP or GPU-accelerated (using CUDA) versions of cpptraj, a different executable name such as \inlinecode{cpptraj.OMP} or \inlinecode{cpptraj.cuda} is often used.
 Therefore, make sure to use the right compilation and executable, to get full benefit of the various improvements to  GIST.
@@
 		\midrule
-		gridcntr    & <y> <y> <z> & Coordinates of the grid center in \AA{} & 0 0 0  \\
+		gridcntr    & <y> <y> <z> & Coordinates of the grid center in \AA & 0 0 0  \\
 		griddim     & <nx> <ny> <nz> & Number of voxels per axis & 40 40 40\\
-		gridspacn & <dval> &Grid spacing / voxel dimensions in  \AA{} & 0.5    \\
-		refdens         & <rdval> & Reference density of the bulk water in molecules/\AA{}$^3$ . & 0.0334    \\
+		gridspacn & <dval> &Grid spacing / voxel dimensions in  \AA & 0.5    \\
+		refdens         & <rdval> & Reference density of the bulk water in molecules/\AA$^3$ . & 0.0334    \\
 		temp & <tval> & System temperature in Kelvin & 300 \\
@@
 \end{table}
-In GIST, the calculated properties are derived by referencing against the bulk solvent.
-It is therefore important to set the reference number density $\rho_0$ according to the solvent model used in your simulation.
-The solvent-solvent enthalpy $E_{ww}$ also needs to be referenced, but this is done in the post-processing stage.
-The calculation of these properties and the need for referencing is discussed in-depth in section \ref{sec:theory}.
+
+It is important to set the reference density $\rho_0$ according to the solvent model used in your simulation.
+The solvent-solvent enthalpy $E_{ww}$ also needs to be referenced, but this can be done in the post-processing stage.
 
 Densities and reference energies of typical water models can be found in the \software{cpptraj} manual.
-We also provide the densities for water models commonly used with AMBER force fields in Table~\ref{tab:ref_densities}.
+We also provide the densities for water models commonly used with AMBER force fields in table \ref{tab:ref_densities}.
 These densities were calculated from \SI{100}{\nano\second} MD simulations at \SI{1}{\bar} and \SI{300}{\kelvin} using the Berendsen barostat and the Langevin thermostat.
@@
 
-Due to the different sizes of the boxes, the reference energies used in this tutorial may not align with those specified in the Amber manual. 
-Generally, it is advisable to conduct your own reference calculations, especially for systems that are either very large or very small. 
-The reference values can be calculated from MD trajectories of bulk solvent.
-From this, one then calculates the density-weighted average energy per solvent molecule ($E_{ww, norm}$) and the average solvent density $\rho_0$.
-Generally, it is recommended that the size of the reference solvent box closely matches that of the production system.
-While the impact on the density is not as pronounced, deviations from the production settings can lead to differences of up to 0.005 kcal/mol in the energy per solvent molecule ($E_{ww, norm}$) for water boxes.
+Due to the varying sizes of the boxes, the reference energies used in this context may not align with those specified in the Amber manual.
+Generally, it is recommended that the size of the reference water box closely matches that of the production system.
+Therefore, it is advisable to conduct your own reference calculations, especially for systems that are either very large or very small.
+While the impact on density is not as pronounced, deviations from the specified settings can lead to differences of up to 0.005 kcal/mol in the $E_{ww, norm}$ value for water boxes.
 These discrepancies, when summing over large grid regions, have the potential to introduce significant errors due to incorrect referencing.
@@
 \begin{table}[h]
-
-	\caption{Reference number density $\rho_0$ and mean solvent-solvent energy $E_{ww, norm}$ for various water models from NPT simulations of N$_{WAT}$ water solvent molecules at \SI{300}{\kelvin} and 1bar.}\label{tab:ref_densities}
+	\caption{Reference density $\rho_0$ and mean water-water energy $E_{ww, norm}$ for various water models from NpT simulations of N$_{WAT}$ water molecules at \SI{300}{\kelvin} and 1bar.}\label{tab:ref_densities}
 	\small
@@
 		Water Model       & $\rho_0$ & $E_{ww, norm}$ & $N_\text{WAT}$ \\
-        & $/\text{\AA{}}^{-3}$ & $/\text{kcal}\cdot \text{mol}^{-1}$ & \\
+        & $/\text{\AA}^{-3}$ & $/\text{kcal}\cdot \text{mol}^{-1}$ & \\
 		\midrule
@@
 	\item{[gX] For every element in the main solvent, the number density of atoms found in the voxel, in units of the bulk density.
-If the same element occurs multiple times, the bulk density is scaled such that that the expectation value for the bulk is unity.
-For water, \code{gO} and \code{gH} are produced.}
-	\item{[Esw] Mean solute-water interaction energy.}
-	\item{[Eww] Mean water-water interaction energy.}
-	\item{[PME] (only if PME was used) water PME energy.}
-	\item{[dTStrans] First order translational entropy.}
-	\item{[dTSorient] First order orientational entropy.}
-	\item{[dTSsix] First order entropy for the six-dimensional combined orientation and translation space.}
+If the same element occurs multiple times, the bulk density is scaled accordingly.
+For water, gO and gH are produced.}
+	\item{[Esw] Mean solute-water interaction energy density.}
+	\item{[Eww] Mean water-water interaction energy density.}
+	\item{[PME] (only if PME was used) water PME energy density.}
+	\item{[dTStrans] First order translational entropy density.}
+	\item{[dTSorient] First order orientational entropy density .}
+	\item{[dTSsix] First order entropy density for the six-dimensional combined orientation and translation space.}
 	\item{[dipole] Magnitude of mean dipole moment (polarization).}
+	\item{[dipolex] x-component of the mean water dipole moment density}
+	\item{[dipoley] y-component of the mean water dipole moment density}
+	\item{[dipolez] z-component of the mean water dipole moment density}
 \end{itemize}
@@
 All calculated quantities are written to the output file (default \inlinecode{gist-output.dat}) and can be used to generate DX files for other quantities.
-A full list of all properties in the output is provided at the end of the tutorial in section \ref{sec:checklists}. 
-The output file is organized by voxel and sorted according to their x, y, and z coordinates, describing the center point of the voxel.
+A full list of all properties in the output is provided at the end of the tutorial.
+
+The output is organized by voxel and sorted according to their x, y, and z coordinates, describing the center point of the voxel.
 Energy and entropy quantities are reported in two different ways.
 The \inlinecode{\_dens} quantities are normalized by the voxel volume and can be used to compute integrals.
-The \inlinecode{\_norm} quantities are instead normalized by the solvent number density in the respective voxel averaged over all considered frames, and are useful to visualize local water properties.
+The \inlinecode{\_norm} quantities are instead normalized by the number of solvents in the respective voxel averaged over all considered frames, and are useful to visualize local water properties.
 To convert from a \inlinecode{\_norm} quantity $X^k_{norm}$ in voxel $k$ to a \inlinecode{\_dens} quantity $X^k_{dens}$, the following equation can be applied:
@@
 \begin{enumerate}
-	\item Set up and run GIST calculations for a small molecule solute (biotin) and a protein (streptavidin) binding pocket.
+	\item Set up and run GIST calculations for a small molecule (biotin) and a protein (streptavidin) binding pocket.
 	\item Compute the free energy of solvation for biotin using GIST.
@@
 
-In the first part, we show how to run GIST calculations for small molecule solutes and interpret the output. This provides an example of a simple GIST workflow, without too much details.
-Afterwards, we apply this knowledge to a protein binding pocket, visualize $\dgsolv$ hotspots and estimate a free energy of ligand binding using an end-states approach based on GIST.
-In addition, we discuss all steps shown in the first example in further detail.
+In the first part, we show how to run GIST calculations for small molecules and interpret the output.
+Afterwards, we apply this knowledge to a protein binding pocket, visualize $\dgsolv$ hotspots and estimate a free energy of ligand binding using an end-state approach based on GIST.
+In addition, we discuss all steps in further detail at the end.
 
@@
 \subsection{Running GIST for Biotin}
-In this example, we will be running a simple GIST calculation of the biotin solute molecule to get familiar with the usage of GIST in \software{cpptraj}.
+In this example, we will be running a simple GIST calculation of the biotin molecule to get familiar with the usage of GIST in \software{cpptraj}.
 \subsubsection{System preparation}
 First, extract biotin from the crystal structure and generate a topology and coordinate file compatible with Amber or a simulation package of your choice.
-
-The solute molecule should be solvated with a sufficiently large solvent box. 
-For periodic boundary conditions, the size of the solvent box should be large enough for the solvent to behave bulk-like so as to reduce artifacts from solvent-mediated interactions between the periodic images.
-This corresponds to a distance between solute and box boundary of at least \SI{12}{\angstrom} for water\cite{Chen2021,Waibl2022-gist-solvents} though we recommend \SI{15}{\angstrom} or higher.
+The molecule should be solvated with a sufficiently large solvent box.
 Carefully check the employed force field, protonation state and partial charges.
@@
 Solvate the parametrized biotin, making sure to choose a water box of large enough size so that the water on the box edges behaves bulk-like (a wall distance of at least 15\AA is recommended).
-A good starting point for the parametrization and solvation of small molecules is the Amber tutorial 4b (\url{http://ambermd.org/tutorials/basic/tutorial4b/}).
+A good starting point (though a bit dated) for the parametrization and solvation of small molecules is the Amber tutorial 4b (\url{http://ambermd.org/tutorials/basic/tutorial4b/}).
 
 \subsubsection{Equilibration}
-Equilibrate your solvated biotin in an NVT or NPT ensemble.
-Even when running the simulation in NVT, it is advisable to include an NPT equilibration step to adjust the box size beforehand.
+Equilibrate your solvated biotin in an NVT or NpT ensemble.
+Even when running the simulation in NVT, it is advisable to include an NpT equilibration step to adjust the box size beforehand.
 Use enough equilibration time to allow the solvent to relax. 
-For water, \SI{0.5}{\nano\second} is usually sufficient to equilibrate the density, but narrow binding pockets can take longer.
+For water, \SI{0.5}{\nano\second} are usually sufficient to equilibrate the density, but narrow binding pockets can take longer.
+
 A well-tested equilibration protocol can be found in the literature \cite{Roe2020-equilibration}.
@@
 \subsubsection{MD simulation}
-\label{sec:md_simulation}
+
 Now, you can run a restrained MD simulation based on the equilibrated structure of biotin for the subsequent GIST analysis.
-Restraints should be used during the MD because IST theory is not defined for moving solute molecules.
+Restraints should be used during the MD because IST theory is not defined for moving solutes.
 We apply a harmonic restraint of \SI{100}{\kilo\calorie\per\mole\per\angstrom\squared} to all heavy atoms to keep the system in place.
 We run \SI{100}{\nano\second} of simulation and store the conformations every \SI{100}{\pico\second}.
-A simulation length of at least \SI{20}{\nano\second} is recommended to obtain statistically independent snapshots of the water solvent molecules. 
+A simulation length of at least \SI{20}{\nano\second} is recommended to obtain statistically independent snapshots of the water molecules.
+
 An example Amber input file for GIST might look like this:
 \begin{lstlisting}[style=amber-in]
-restrained 100 ns NPT
+restrained 100 ns NpT
 &cntrl
@@
 \end{lstlisting}
-This file will run a \SI{100}{\nano\second} NPT MD simulation while keeping all atoms except for hydrogens and waters restrained.
+
+This file will run a \SI{100}{\nano\second} NpT MD simulation while keeping all atoms except for hydrogens and waters restrained.
 Note that these are very standard MD settings except for the restraints.
+
 Run the MD using \software{pmemd.cuda}:
@@
 \subsubsection{GIST analysis}
-
-To start our GIST analysis of biotin, we'll first have to decide on the parameters defining the GIST grid during the analysis.
-We will need a solute structure without water for the analysis of our future GIST results. 
-As we ran our MD simulation with periodic boundaries, we always need to reimage all molecules to the original box using cpptraj's \inlinecode{autoimage} function. 
-We also recommend that you center the biotin solute molecule to the coordinate origin, as this simplifies the definition of our grid origin later on:
-
+To start our GIST analysis of biotin, we'll first have to decide on the parameters defining the GIST grid during the analyis.
+We recommend that you center the biotin molecule to the coordinate origin:
 \begin{lstlisting}[style=cpptraj]
@@
 trajout biotin-centered.pdb
-run
-\end{lstlisting}
-In this code snippet, the first two lines load the topology and coordinates of the first frame of our MD simulation. 
-In the third line, we reimage the coordinates to the first box and center the biotin at the origin. 
-Then, we strip all water solvent molecules in the fourth line and write out a pdb file in the fith line. 
-Finally, we start the analysis with the "run" command.
-
+\end{lstlisting}
 To decide on a grid size, find the extent (minimum and maximum) of the x-, \mbox{y-,} and z-coordinates of the molecule.
-
 To get bulk-like water properties at the grid border, a buffer distance of \SI{15.0}{\angstrom} is reasonable for biotin.
@@
 bounds !(:WAT) dx 0.5 offset 30 name MyGrid out bounds.dat
-run
-\end{lstlisting}
-In the above snippet, we calculate the bounds around the whole protein for a grid with \SI{0.5}{\angstrom} side-length voxels and a buffer zone of \SI{15.0}{\angstrom}. 
-Note that the offset is given in terms of number of additional voxels, not length dimension.
-By varying the atom mask in the \inlinecode{bound} command, it is possible to center the grid on a region of interest, similar to the \software{findcentroid.py} script.
-
-Now, you can run the actual GIST analysis.
-Reimage and center the solute molecule as above, to verify that its position matches the grid parameters you decided on.
-We recommend using either the GPU implementation or MPI-accelerated PME implementation of GIST.
-While the GPU version is faster when limited CPU resources are available, the PME implementation matches the energy of the Amber MD engine more closely.
-The MPI implementation \cite{Roe2023-mpi-gist} scales well over multiple CPU cores and can be combined with either the GPU or PME implementations for maximum efficiency. 
-A speed comparison of the various parallelization implementations is shown in \cite{Roe2023-mpi-gist}.
+\end{lstlisting}
+Here, we calculate the bounds around the whole protein for a grid with \SI{0.5}{\angstrom} side-lenght voxels and a buffer zone of \SI{15.0}{\angstrom}.
+Note that the offset is given in terms of number of additional voxels, not voxel dimension.
+By varying the atom mask in the \inlinecode{bound} command, it is possible to center the grid on a region of interest, similiar to the \software{FindCentroid.py} script.
+
+Now, you can finally run the actual GIST analysis.
+Center the molecule as above, to make sure that its position matches the grid parameters you decided on.
+We recommend using either the PME or GPU implementation of GIST.
+For the GPU version, cpptraj needs to be compiled with the \inlinecode{-cuda} flag.
+While the GPU version is generally faster, the PME implementation matches the energy of the Amber MD engine more closely.
 
@@
 \subsubsection{Working with the GIST results}
-Running the GIST analysis above generates an output data file (gist.dat) containing all calculated quantities for each voxel in the grid. 
+Running the GIST analysis above, an output data file (gist.dat) containing all calculated quantities for each voxel in the system is produced. 
 Additionally as outlined in section \ref{sec:running_GIST}, various quantities are written to OpenDX files.
+
 To interact with your GIST data, we recommend the gisttools library for python3. Open the GIST output with gisttools:
@@
 The GIST data can be accessed and modified from the gist object in various ways.
-Gisttools provides access to single or multiple voxels through a \code{.loc} property similiar to the pandas library. 
-GIST quantity data for all voxels can be accessed by directly indexing the gist object, either by providing a single name or a list of names.
-Indexing with \code{.loc} works by providing a voxel index (either a index number or list of numbers) and optionally a column index (column name or list of names).
+Gisttools provides access to single or multiple voxels through a .loc property similiar to the pandas library. 
+GIST quantity data for all voxels can be accessed through directly indexing the gist object, either by providing a single name or a list of names.
+Indexing with .loc works by providing a voxel index (either a index number or list of numbers) and optionally a column index (column name or list of names).
 Indices start from 0 and indexing will return a single entry, pandas series or pandas dataframe depending on the query:
@@
 \end{lstlisting}
+
 Gisttools provides access to all quantities listed in the GIST output file.
-Additionally, it calculates the total enthalpy per voxel (Eall) as a sum of Eww and Esw, and the free energy of solvation (A) by subtracting $\Delta$TSsix from Eall.
+Additionally, it calculates the total enthalpy per voxel (Eall) as a sum of Eww and Esw, and the free energy of solvation (A) by subtracting dTSsix from Eall.
 
 Information about the grid is stored in the grid property of the gist object, structural data is stored as an \software{mdtraj} trajectory in the struct property:
-%TODO we are missing some code here
-Gisttools also provides OpenDX writing capabilities to visualize the calculated or modified GIST data.
+
+Gisttools also provides minimal OpenDX writing capabilities to visualize the calculated or modified GIST data.
 For visualization, you'll most likely want to write out the density-normalized GIST quantities (X\_dens):
@@
 \end{lstlisting}
-Section~\ref{sec:visualization} discusses the visualization of OpenDX files.
-
+
+This tutorial discusses the visualization of OpenDX files in Section~\ref{sec:visualization}.
 \subsection{Applying GIST for the Streptavidin-Biotin complex}
-In this section we will apply GIST to the full complex and discuss each step of the process in more detail to outline potential pitfalls.
+In this section we will apply GIST to the full complex and discuss each step of the process in more detail to outline pitfalls.
 We will afterwards introduce advanced analysis workflows which allow you to modify and better visualize the resulting data.
@@
 In our example, we only use a single equilibrated structure.
-A more sophisticated approach would be to identify several dominant structures in the ensemble and run a separate MD simulation and GIST calculation for each structure. 
-Such dominant structures could either be identified from experimental ensembles or by clustering an unrestrained MD simulation. 
+A more sophisticated approach would be to identify several dominant structures in the ensemble -- either from experimental ensembles or by clustering an MD simulation.
+
 When starting from an experimental structure, care has to be taken during the preparation step.
 Make sure to choose a proper protonation state for all titratable moeties.
-You will need to solvate your structure in a solvent box, which should be large enough to obtain unbiased long-range electrostatics and avoid solvent-mediated interactions of periodic images.
-A good criterion for validating the box size is to check whether the solvent on the outside of the box behaves bulk-like according to the GIST output.
-Note that, due to the restraints and finite sampling time, some solvent sites might be inaccessible to the bulk solvent.
-To investigate such positions, make sure to explicitly place water solvent molecules there. 
-For cases where the exact occupation of such sites is unknown, it may be helpful to include grand-canonical Monte Carlo (GCMC) or nonequilibrium candidate Monte Carlo (NCMC) sampling steps in the equilibration or production stage \cite{Ge-GCMC2022,Melling-NCMC2023}.
+
+You will need to solvate your structure in a solvent box, which should be large enough to obtain unbiased long-range electrostatics.
+A good criterion for validating the box size is to check whether the water on the outside of the box behaves bulk-like according to the GIST output.
+Note that, due to the restraints, some solvent sites might be inaccessible to the bulk solvent.
+To investigate such positions, make sure to explicitly place water molecules there.
 It is good practice to check your restrained MD simulation or solvent populations obtained by GIST to confirm whether all relevant sites were properly solvated.
 
-When calculating solvation free energy differences between a complex and the corresponding monomers (the dissociated state), there are generally two approaches for dealing with configurational diversity.
-In the first approach, the ensemble properties of each individual part and the complex are accounted for by running individual MD simulations for each.
-The second approach assumes sufficient overlap in the conformational space of the dissociated states and the complex. 
-Then, only configurations of the complex are sampled in a single MD simulation of the complex. 
-Calculations for the disassociated states are then run on the extracted structures from the complex configurations.
-One can choose to sample the complex and dissociated states individually, or to sample only one of them and assume that there is sufficient overlap in the conformations \cite{Genheden-EndState2012,Wang-MMPBSA2019}.
+When calculating solvation free energy differences between a complex and the corresponding monomers (the dissociated state), there are two options for dealing with configurational diversity in the structures.
+One can choose to sample the complex and dissociated states individually, or to sample only one of them and assume that there is sufficient overlap in the conformations.
 
@@
 \end{lstlisting}
-Adapt the above input file for biotin by striping molecule 1 instead of 2 and replacing the path names in the \inlinecode{strip} and  \inlinecode{trajout} commands.
-Alternatively, we provide a small script named \software{cpptraj\_remove\_mol.sh} for this procedure.
+(Adapt for biotin as necessary.)
+
+Use the \software{equilibration.py} script (or your own procedure) to perform short NVT and NpT equilibration runs on the complex structure.
+Note that the equilibration script contains unrestrained NpT steps, because restraining two molecules at the same time might interfere with the pressure equilibration.
+
+Then split the equilibrated structure into biotin and streptavidin parts using \software{cpptraj\_remove\_mol.sh}.
 After that, use \software{equilibration.py -R} to run short restrained 
@@
 systems.
-By using restraints, we can ensure that the monomer conformations match the complex.
+By using restraints, we can make sure that the monomer conformations match the complex.
+
+\subsubsection{MD Simulation}
+The MD can be run as described above for each monomer and the complex.
 
 \subsubsection{GIST analysis of Streptavidin-Biotin}
-
-After running the MD simulations as described in section \ref{sec:md_simulation}, proceed with the GIST analysis.
-For easier post-processing, here we use grids that contain the whole solute molecule and the surrounding solvent.
-
-If one is only interested in solvent in the binding site, smaller grids encompassing only the binding site could be used for streptavidin and the complex.
-While this saves calculation time for the GIST analysis, it complicates the post-processing.
-Since the GIST analysis is usually less time-demanding than the MD simulation, we recommend using larger grids.
-Furthermore, we recommend that you center the solute molecule to the coordinate origin to simplify the analysis.
+For easier post-processing, here we use grids that contain the whole molecule.
+Smaller grids encompassing only the binding site could be used for streptavidin and the complex, but this would complicate the analysis.
+Furthermore, we recommend that you center the molecule to the coordinate origin.
 Calculate the GIST grid parameters as outlined in the biotin section for the complex and streptavidin and run the GIST analysis.
-
-For larger systems like whole proteins you can usually get away with a smaller buffer region for the GIST grid compared to small molecule solutes.
-To test whether your grid is large enough, look at solvent properties close to the grid edges, e.g. the solvent-solvent energies.
+For larger systems like whole proteins you can usually get away with a smaller buffer region compared to small molecules.
+To test whether your grid is large enough look at solvent properties close to the grid edges, e.g. the solvent-solvent energies.
 If these correspond to the respective bulk properties, it can be assumed that your grid is large enough to capture all perturbations of the solute on the solvent.
 Note that a different box-size or number of solvent molecules can skew the bulk properties slightly compared to the reference values provided above.
-In that case, checking for convergence of the solvent properties near the grid borders is sufficient.
-You might want to modify your reference values according to your results, as described in the next section.
+In that case, checking for convergence of the solvent properties near the grid borders is sufficient but you might want to modify your reference values according to your results then as described in the next section.
+
+For the GIST analysis, we recommend using either the PME or GPU implementation of GIST.
+For the GPU version, cpptraj needs to be compiled with the \inlinecode{-cuda} flag.
+While the GPU version is generally faster, the PME implementation matches the energy of the Amber MD engine more closely.
 
 \subsubsection{Reference Values and Radial Convergence}
-All GIST quantities should be expressed relative to bulk as outlined in section \ref{sec:theory}.
+All GIST quantities should be expressed relative to bulk.
 This is automatically done by \software{cpptraj} for the entropy, using the bulk density specified by \inlinecode{refdens}.
@@
 The Amber manual provides reference densities and energies for several solvents.
-Furthermore, a large variety of reference values calculated with GPU-GIST are listed in Table~\ref{tab:ref_densities}. 
 However, the exact energy values are different when using PME-GIST, and also depend on the box size.
 For quantitative analyses, it is recommended to compute an exact reference energy.
-The most accurate method to compute a reference energy is to run a GIST calculation of the pure solvent using the same energy method (PME/GPU), a similar box size, at the same temperature and pressure and then compute the average solvent-solvent energy per solvent molecule.
+
+The most accurate method to compute a reference energy is to run a GIST calculation of the pure solvent using the same energy method (PME/GPU) and a similar box size, and then compute the average solvent-solvent energy per molecule.
 
@@
 \end{equation}
-Alternatively, the solvent-solvent energy in a large GIST grid converges to the bulk value at sufficient distance to the solute molecule. 
-Therefore, the reference energy can be obtained by binning the voxels by their distance to the solute molecule, and evaluate Equation~\ref{eq_ewwref} within each bin.
-If this value converges with increasing distance from the solute, it can be used as $E_{ww}^\textit{ref}$.
+
+Alternatively, the solvent-solvent energy in a large GIST grid converges to the bulk value at sufficient distance to the solute.
+Therefore, the reference energy can be obtained by binning the voxels by their distance to the solute, and evaluate Equation~\ref{eq_ewwref} within each bin.
+If this value converges, this can be used as $E_{ww}^\textit{ref}$.
 
@@
 \end{lstlisting}
+
 The expected output of this code is shown in Figure~\ref{fig_ewwref}.
-%TODO redo figure caption, yaxis label
+
 \begin{figure}
@@
 	\includegraphics[width=0.8\linewidth]{figures/Eww_convergence.eps}
-	\caption{Convergence of $E_{ww}$ in the complex calculation with increasing distance to the solute molecule (streptavidin-biotin).
+	\caption{Convergence of $E_{ww}$ in the complex calculation with increasing distance to the solute (streptavidin-biotin).
 The horizontal line shows the automatically computed reference energy.}\label{fig_ewwref}
@@
 In \software{gisttools}, this is done simply by setting \inlinecode{gist.eww\_ref}.
+
 Now, sum all free energy contributions to obtain the spatially resolved $\dgsolv$.
@@
 \end{equation}
+
 %In \software{gistpp}, use the \inlinecode{add} command, with the same syntax as the \inlinecode{div} command used above.
-After that, check whether your free energy contributions become negligible far away from the solute molecule.
+After that, check whether your free energy contributions become negligible far away from the solute.
 In a plot like Figure~\ref{fig_ewwref} based on the \inlinecode{A\_dens} column, the values should tend to zero.
-However, it is more informative to plot the cumulative free energy contribution against the distance to the solute molecule, to check the convergence (see Figure~\ref{fig_radial_convergence}).
-If the curve flattens out, the converged value is your final $\dgsolv$\@.
-If the curve diverges even for large integration distances you might need to tweak the $E_{ww}$ reference, or introduce a reference value for the entropy.
+However, it is more informative to plot the cumulative free energy contribution against the distance to the solute, to check the convergence.
+It the curve flattens out, the converged value is your final $\dgsolv$\@.
+Otherwise, you might need to tweak the $E_{ww}$ reference, or introduce a reference value for the entropy.
 
@@
 import matplotlib.pyplot as plt
-import numpy as np
-# adapt eww_ref according to the used solvent model! 
-gist = load_gist_file('gist.dat',
-    struct='solute-centered.pdb', eww_ref=-9.533)
+# adapt eww_ref!
+gist = load_gist_file("gist.dat",
+    struct="solute-centered.pdb", eww_ref=-9.533)
 bins, (dg, esw, eww, s) = gist.rdf(
-    ['A_dens', 'Esw_dens', 'Eww_dens', 'dTSsix_dens'],
-    bins=100, rmax=20., normalize='none')
-plt.plot(bins, np.cumsum(eww), label='Eww')
-plt.plot(bins, np.cumsum(esw), label='Esw')
-plt.plot(bins, np.cumsum(dg), label='dG')
-plt.plot(bins, np.cumsum(s), label='dS')
-plt.legend()
-plt.xlabel("distance cutoff []")
-plt.ylabel("free energy contribution [kcal/mol]")
-\end{lstlisting}
+    ["A_dens", "Esw_dens", "Eww_dens", "dTSsix_dens"],
+    bins=100, rmax=20., normalize="none")
+plt.plot(bins, np.cumsum(dg), label="dG")
+plt.plot(bins, np.cumsum(eww), label="Eww")
+plt.plot(bins, np.cumsum(esw), label="Esw")
+plt.plot(bins, np.cumsum(s), label="dS")
+\end{lstlisting}
+
 The expected output of this code is shown in Figure~\ref{fig_radial_convergence}.
@@
 	\includegraphics[width=0.8\linewidth]{figures/A_E_S_convergence.eps}
-	\caption{Convergence of $\dgsolv$ and its contributions with increasing distance to the solute molecule.
-All quantities are cumulative, i.e., summed up to the respective radius.
+	\caption{Convergence of $\dgsolv$ and its contributions with increasing distance to the solute.
+All quantities cumulative, i.e., summed up to the respective radius.
 Computed from the biotin calculation.}\label{fig_radial_convergence}
@@
 %With \software{gistpp}, you can use the file from the convergence check. Additional OpenDX files can be produced using \inlinecode{gistpp -op makedx}.
-With \software{gisttools}, you can create one using e.g., \inlinecode{gist.save\_dx('A\_dens', 'A\_dens.dx')}.
-It might also be interesting to visualize the average energy of a water solvent molecule at each grid voxel.
+With \software{gisttools}, you can create one using e.g., \inlinecode{gist.save\_dx("A\_dens", "A\_dens.dx")}.
+
+It might also be interesting to visualize the average energy of a water molecule at each grid voxel.
 This quantity is given by $2E_{ww} + E_{sw}$.
@@
 The code is not explained in detail here, but should be clear after reading the previous sections.
+
 Load the files and double check the frame numbers and reference density.
@@
 Assign reference energies and check them for plausibility.
-
 \begin{lstlisting}[style=python]
 biotin.eww_ref = biotin.detect_reference_value()
-print('Biotin:', biotin.eww_ref)
+print("Biotin:", biotin.eww_ref)
 strept.eww_ref = strept.detect_reference_value()
-print('Streptavidin:', strept.eww_ref)
+print("Streptavidin:", strept.eww_ref)
 compl.eww_ref = compl.detect_reference_value()
-print('Complex:', compl.eww_ref)
-\end{lstlisting}
+print("Complex:", compl.eww_ref)
+\end{lstlisting}
+
 Subtract a reference entropy from the \inlinecode{dTSsix} columns.
-
 \begin{lstlisting}[style=python]
@@
 \end{lstlisting}
+
 Compute the atom positions that define the integration region.
@@
 \end{lstlisting}
-Now, subtract the monomer histograms from the complex, and compute the sum of your property within some cutoff distance to the solute.
-If you also visualize the individual histograms, you will notice that the difference converges much better with increasing radius than the single contributions.
+
+Now, subtract the monomer rdfs from the complex, and compute the integrals.
+If you also visualize the individual rdfs, you will notice that the difference converges much better with increasing radius than the contributions.
+This is because the cutoff contains regions that are close to the atom, since the integration region does not comprise the whole protein.
 
@@
 \end{lstlisting}
+
+Compute the energy (\inlinecode{Eall}) and entropy (\inlinecode{dTSsix}) contributions separately.
+You will find that the energy disfavors binding, because we do not yet include the interaction energy between biotin and streptavidin.
+To make this easier, gisttool includes a function to integrate in a radius around atom centers:
+\begin{lstlisting}[style=python]
+centers = {"biotin": biotin_x, 
+	   "strept": strept_x, 
+	   "compl": compl_x}
+gist_objs = {"biotin": biotin, 
+             "strept": strept, 
+             "compl": compl}
+cols = ['Eall_dens','dTSsix_dens']
+rmax = 12
+results = {
+    name: gist.integrate_around(
+        cols, rmax, centers[name])
+    for name, gist in gist_objs.items()}
+display(pd.DataFrame(results))
+
+\end{lstlisting}
 You can also plot the difference between the complex and monomer contributions against the distance to biotin, using the complex coordinates for the holo structure:
-
 \begin{lstlisting}[style=python]
@@
 \end{lstlisting}
+You can also plot the difference between the complex and monomer contributions against the distance to biotin, using the complex coordinates for the holo structure.
 The expected result is shown in Figure~\ref{fig-dg-sums}.
-\begin{figure}[h]
+\begin{figure}
 	\centering
@@
 Right: The hydration contribution to binding, evaluated as the difference between the lines in the left panel.
-The vertical dashed line is at \SI{12}{\angstrom} and represents the chosen integration cutoff.}\label{fig-dg-sums}
+The vertical dashed line is at 12 \AA\ and represents the chosen integration cutoff.}\label{fig-dg-sums}
 \end{figure}
-Compute the energy (\inlinecode{Eall}) and entropy (\inlinecode{dTSsix}) contributions separately.
- To make this easier, gisttool includes a function to integrate in a radius around atom centers:
-\begin{lstlisting}[style=python]
-import pandas as pd
-centers = {'biotin': biotin_x, 
-	'strept': strept_x, 
-	'compl': compl_x}
-gist_objs = {'biotin': biotin, 
-	'strept': strept, 
-	'compl': compl}
-cols = ['Eall_dens','dTSsix_dens']
-rmax = 12
-results = {
-	name: gist.integrate_around(
-	cols, rmax, centers[name])
-	for name, gist in gist_objs.items()}
-print(pd.DataFrame(results))
-\end{lstlisting}
-%TODO: This has a higher indent than the other python snippets in the pdf, somehow. Unsure why...
-You will find that the energy disfavors binding, because we do not yet include the interaction energy between biotin and streptavidin.
+
 Using the \inlinecode{energy} command in \software{cpptraj}, you can compute this interaction energy.
-We recommend using PME in combination with PME-GIST, but not with GPU-GIST to be in line with the different ways the energy is calculated in the two GIST methods.
+We recommend using PME in combination with PME-GIST, but not with GPU-GIST\@.
 An example \software{cpptraj} input might look like:
@@
 \end{lstlisting}
+% test edit
+
 It has been shown \cite{Chen2021,Waibl2022-gist-solvents} that the solvation entropy in water is best approximated by 0.6 times the first order entropy.
@@
 \end{table}
+
 Even though streptavidin-biotin is known to be a very stable complex, the solvation energy favors the dissociation.
-This is expected: the binding of biotin to streptavidin is faciliated over a large number of H-bonds and over polar interactions. 
-As such, water favourably solvates both biotin and streptavidin and an energetic penalty is paid when displacing the water during the binding process. 
+This is expected: if the molecules were not able to form any interactions, but interacted fully with the water, they would not form a complex.
+
 In this case, we find that the energy of solvation and the internal energy of the complex more or less cancel out, indicating that binding is largely entropy-driven in this case.
@@
 To include the effect of lid closing in the binding process, free energy calculation methods such as umbrella sampling can be used.
-To speed up the calculation, smaller GIST grids could be used by focusing on the binding pocket, or by rotating the solute molecule by its principal axes to fit the cuboid grid more exactly.
+
+To speed up the calculation, smaller GIST grids could be used by focusing on the binding pocket, or by rotating the molecule by its principal axes to fit the cuboid grid more exactly.
 However, this should be done \emph{before} the MD, since rotating the trajectory damages the periodic box information.
-%TODO: Ask Dan if this is still the case
 
@@
 In Reference~\cite{Kraml2020}, a GPU implementation of the energy calculation was presented, which typically increases the speed by 1--2 orders of magnitude.
+
 If \software{cpptraj} is compiled with GPU support, the energy calculation uses the GPU automatically, without any additional input.
@@
 For typical systems, this implementation is slightly slower than GPU-GIST, but much faster than the original CPU code.
-Furthermore, PME-GIST offers the best agreement between GIST energies and those observed in an classical MD simulation.
+Furthermore, PME-GIST offers the best agreement between GIST energies and those observed in an MD simulation.
+
 To run PME-GIST, the \inlinecode{pme} flag to \inlinecode{gist} can be used in \software{cpptraj}.
 The output will contain two additional columns: \inlinecode{PME\_dens} and \inlinecode{PME\_norm}.
-They contain the potential of solute molecule and solvent evaluated at the solvent positions and divided by two.
+They contain the potential of solute and solvent evaluated at the solvent positions and divided by two.
 Furthermore, the \inlinecode{Eww} and \inlinecode{Esw} columns are also computed using PME, and are reported as usual.
 For typical use cases, we recommend using \inlinecode{Eww} and \inlinecode{Esw} over the \inlinecode{PME} column.
-
-\subsection{MPI parallelization}
-The most recent addition to GIST is the MPI parallelization of both energy and entropy calculations\cite{Roe2023-mpi-gist}.
-Since the MPI parallelization is orthogonal to the other improvements, it can be used with both PME or GPU accelerated GIST.
 \subsection{GIST with non-water solvents}
-In references~\cite{Kraml2020,Kamenik2020-gist-macrocycles,Waibl2022-gist-solvents}, an extension of GIST towards solvents other than water is described.
-To run a GIST calculation in a solvent other than water, use \inlinecode{solute} with a \software{cpptraj} selection mask to select the solute molecule.
+In References~\cite{Kraml2020,Kamenik2020-gist-macrocycles,Waibl2022-gist-solvents}, an extension of GIST towards solvents other than water is described.
+
+To run a GIST calculation in a solvent other than water, use \inlinecode{solute} with a \software{cpptraj} selection mask.
 All other molecules will be treated as solvent.
-If there are multiple solvent species, also use the \inlinecode{solventmols} flag as described in Section~\ref{sec-salt-water}. \\
-The code will automatically choose three atoms per solvent species to define the solvent orientation, and print this selection to the output.
+If there are multiple solvent species, also use the \inlinecode{solventmols} flag as described in Section~\ref{sec-salt-water}.
+
+The code will automatically choose three atoms per molecule to define the solvent orientation, and print this selection to the output.
 Sometimes, however, the automatic selection is not optimal.
@@
 \end{lstlisting}
+
 Note that the central atom of the \inlinecode{rigidatoms} selection goes first.
@@
 This allows treating salting-out effects, although it was shown that the salting-out coefficient is over-predicted by GIST because of the first-order entropy approximation.
-Generally, GIST should be able to treat arbitrary solvent mixtures as long as each solvent molecule is sufficiently rigid. \\
-In the \software{cpptraj} implementation of GIST, the \inlinecode{solute} keyword specifies which components of the system are treated as a solute molecule.
+Generally, GIST should be able to treat arbitrary solvent mixtures as long as each molecule is sufficiently rigid.
+
+In the \software{cpptraj} implementation of GIST, the \inlinecode{solute} keyword specifies which components of the system are treated as a solute.
 Everything else will be treated as solvent.
 If the solvent contains more than one molecular species, a list of solvent molecule names (i.e., ``\inlinecode{solventmols WAT,NA,CL}'') must be specified.
-For each solvent molecule in this list, densities (e.g., \inlinecode{g\_mol\_NA}) and energies (e.g., \inlinecode{Eww\_mol\_NA}) will be computed and written to .dx files.
-Note that, e.g., \inlinecode{Esw\_mol\_NA} contains the interactions of every ``\inlinecode{NA}'' solvent molecule with the solute molecule, and \inlinecode{Esw\_mol\_NA} contains the interactions among ``\inlinecode{NA}'' solvent molecules as well as their interactions with all other solvents, divided by 2 to account for double counting.
-The entropy will be computed using only the density of the first solvent molecule. \\
+For each molecule in this list, densities (e.g., \inlinecode{g\_mol\_NA}) and energies (e.g., \inlinecode{Eww\_mol\_NA}) will be computed and written to .dx files.
+Note that, e.g., \inlinecode{Esw\_mol\_NA} contains the interactions of every ``\inlinecode{NA}'' molecule with the solute, and \inlinecode{Esw\_mol\_NA} contains the interactions among ``\inlinecode{NA}'' molecules as well as their interactions with all other solvents, divided by 2 to account for double counting.
+Furthermore, the first molecule in this list will be treated as the main solvent, for which first-order entropy values will be computed.
+
 For instance, a GIST calculation in salt water, including the first-order water entropy, can be run as follows:
@@
 \end{lstlisting}
+
 The current implementation of the entropy calculation in \software{cpptraj} requires at least 3 atoms in the main solvent.
-Note that the calculation of the reference value is complicated by the introduction of additional solvents.
-For an overview of how to correctly handle such calculations, refer to reference \cite{Waibl2021-gist-salt}.
-Additionally, to compute the first-order entropy of the ions, as well as an approximate second order entropy, Python code is available on \url{https://github.com/liedllab/second-disorder}.
-%TODO: Add recommendations how to handle charges/counterions to text (not necessarily here)
+To compute the first-order entropy of the ions, as well as an approximate second order entropy, Python code is available on \url{https://github.com/liedllab/second-disorder}.
+
 \section{Visualization}
@@
 \end{lstlisting}
-A more sophisticated visualization, including the energy, entropy, and free energy of removing a single water solvent molecule, is found in Figure~\ref{fig-binding-pocket.png}
+
+A more sophisticated visualization, including the energy, entropy, and free energy of removing a single water molecule, is found in Figure~\ref{fig-binding-pocket.png}
 
@@
 
-\subsection{Solvent Accessible Surface (SAS)}
-
-The solvent accessible surface (SAS) represents the interface between regions that are occupied by the solute molecule, and those which can be accessed by the solvent.
-GIST provides a very natural way of creating a SAS using an isosurface of the oxygen density \code{gO}.
+\subsection{Solvent Accessible Surface Area (SASA)}
+
+The solvent accessible surface (SASA) represents the interface between regions that are occupied by the solute, and those which can be accessed by the solvent.
+GIST provides a very natural way of creating a SASA using an isosurface of the oxygen density gO.
 It can be visualized in PyMOL.
@@
 	\includegraphics[width=1.0\linewidth]{figures/sasa-simple.png}
-	\caption{SAS for streptavidin focused on the binding pocket}\label{fig-streptavidin_sasa}
+	\caption{SASA for streptavidin focused on the binding pocket}\label{fig-streptavidin_sasa}
 \end{figure}
 
-Here, the SAS was generated from a simulation of water around streptavidin without biotin, leaving the binding site solvent-accessible.
+Here, the SASA was generated from a simulation of water around streptavidin without biotin, leaving the binding site solvent-accessible.
+
 \subsection{Defining the binding pocket}
 Specific regions in a protein can be defined to calculate the thermodynamic quantities for that region.
-This is significant specially when calculating the properties of water that will be expelled when a ligand binds to the protein.\\
-For example, voxels that are part of the binding pocket can be defined using a ligand solute molecule and a distance criteria. 
+This is significant specially when calculating the properties of water that will be expelled when a ligand binds to the protein.
+
+For example, voxels that are part of the binding pocket can be defined using a ligand molecule and a distance criteria.
+
 In section~\ref{sec:binding_contributions} we show how to integrate over regions such as the binding pocket.
@@
 \begin{lstlisting}[style=python]
-import gisttools as gt
 def select(traj, sel):
-	""" 
-	Slice a Trajectory by selection mask. 
-	"""
+	"Slice a Trajectory by selection mask."
 	return traj.atom_slice(traj.top.select(sel))
@@
 
-### Set voxels within 3.5A of biotin to 1, write out dx file
+### Set voxels within 3.5A to 1, write out dx file
 compl.loc[ind, 'BP'] = 1
@@
 \end{figure}
-%TODO: Hide capping group
+
 \section{Theory}
@@
 Here, we only provide a short overview of the theory behind GIST.
-For more technical information (such as implementation details), we recommend one of the more recent publications on developments in GIST \cite{Kraml2020,Chen2021,Roe2023-mpi-gist}.
+For more detailed information, we recommend one of the more recent publications on developments in GIST \cite{Kraml2020,Chen2021}.
 
@@
 %$\rho^\textit{o}$ is the bulk number density, $g_\textit{sw}\left(\textbf{r},\omega \right)$ is the solute-water 
-%pair correlation function in the solute's frame of reference, \textbf{r} is the coordinates of the oxygen atom of the water solvent molecule,  
+%pair correlation function in the solute's frame of reference, \textbf{r} is the coordinates of the oxygen atom of the water molecule,  
 %$\omega$ represents the Euler angles, and $\frac{1}{8\pi^2}$ is the volume of the orientational space, which acts as a normalization constant.
@@
 
-IST expresses the free energy of solvation in terms of the solvent density in a coordinate system defined by the solute molecule.
-The solute molecule is kept fixed in space, following the standard solvation process established by Ben-Naim \cite{ben-naim-book}.
-
-The entropy can be approximated as an infinite expansion of correlations of increasing order.
-In a slightly simplified view, the first order is the log solvent density $ln\left(g_\textit{sw}\left(\textbf{r},\omega \right)\right)$ at each position $\textbf{r}$ and orientation $\omega$.
+IST expresses the free energy of solvation in terms of the solvent density in a coordinate system defined by the solute.
+The solute is kept fixed in space, following the standard solvation process established by Ben-Naim \cite{ben-naim-book}.
+
+For the entropy, this leads to an infinite expansion of correlations of increasing order.
+In a slightly simplified view, the first order is the solvent density $g_\textit{sw}\left(\textbf{r},\omega \right)$ at each position $\textbf{r}$ and orientation $\omega$.
 The first-order entropy omits all solvent-solvent (and higher) correlations, while the solute-solvent correlation is taken into account because the coordinates are relative to the solute.
@@
 %The second order is the pair density, i.e., the probability of finding two solvent molecules in positions $\textbf{r}$ and $\textbf{r}^\prime$.
+
 The solvent density is expressed relative to the bulk density $\rho^0$. %$g_\textit{sw}\left(\textbf{r},\omega \right)$ is the solute-water 
-%pair correlation function in the solute's frame of reference, \textbf{r} is the coordinates of the oxygen atom of the water solvent molecule,  
+%pair correlation function in the solute's frame of reference, \textbf{r} is the coordinates of the oxygen atom of the water molecule,  
 %$\omega$ represents the Euler angles, and 
 The entropy integral is also normalized by $8\pi^2$, which is the volume of the orientational space.
-In bulk or at a high distance to the solute molecule, the quantity $g_\textit{sw}\left(\textbf{r},\omega \right)$ approaches one, which leads to zero first-order entropy. 
+In bulk or at a high distance to the solute, the quantity $g_\textit{sw}\left(\textbf{r},\omega \right)$ approaches one, which leads to zero first-order entropy. 
 Therefore, no reference entropy is needed unless there is a numeric bias, e.g., when the MD frames are statistically dependent because the time between them is insufficient.
@@
 \subsubsection{Entropy Calculations in Cpptraj}
-In \software{cpptraj}, the calculation of solvation entropy is handled by two methods.\\
+In \software{cpptraj}, the calculation of solvation entropy is handled by two methods.
+
 In the first method, the solvation entropy is broken down into translational and orientational contributions.
@@
 \end{equation}
-This is exact assuming that the distribution of the orientation $\omega$ is constant within a voxel $k$, but might converge slower than computing the entropy from the combined space (see below).\\
+This is exact assuming that the distribution of the orientation $\omega$ is constant within a voxel $k$, but might converge slower than computing the entropy from the combined space (see below).
+
 The second method directly calculates the solvation entropy by evaluating the six-dimensional integral (3 for position and 3 for orientation).
@@
 \end{equation}
+
 A first nearest neighbor (NN) approach is used to evaluate the density.
 The densities are computed relative to a homogeneous distribution of solvent molecules, given a bulk density $\rho^0$.
+
 The average entropy contributions per solvent molecule in voxel $k$ are calculated as
@@
 \end{equation}
-where $N_\textit{k}$ is the number of water solvent molecules found in voxel k, $\gamma$ is Euler's constant accounting for the bias of the naive entropy estimator, and $g_\textit{NN}$ is the nearest neighbor estimate of the density.
+where $N_\textit{k}$ is the number of water molecules found in voxel k, $\gamma$ is Euler's constant accounting for the bias of the naive entropy estimator, and $g_\textit{NN}$ is the nearest neighbor estimate of the density.
 
@@
 \end{equation}
-The solute-solvent energy $E_{sw}$ can be expressed in terms of the solvent density and the potential $U_{sw}(\textbf{r},\omega)$ induced by the solute molecule.
+
+The solute-solvent energy $E_{sw}$ can be expressed in terms of the solvent density and the potential $U_{sw}(\textbf{r},\omega)$ induced by the solute.
 In practice, it is evaluated as the expectation value $\langle\cdot\rangle$ of the pairwise force field energy $U_{ij}$ between all solvent molecules in voxel $k$ and all solute atoms.
@@
 \end{equation}
+
 To localize $E_{sw}$ on the three-dimensional grid, every energy term is assigned to the voxel $k$ holding the solvent, and the expectation value in Equation~\ref{eq-esw} is computed per voxel.
-Similar to the entropy integrals, the solute-water solvation energy decays with increasing distance to the solute molecule.
+Similar to the entropy integrals, the solute-water solvation energy decays with increasing distance to the solute.
 Hence, it can be approximated by local spatial integrals.
+
 The solvent-solvent energy $E_{ww}$ is computed similarly.
@@
 In contrast to $E_{sw}$, $E_{ww}$ does not tend to zero in bulk.
-Therefore, a reference corresponding to the average energy of a bulk water solvent molecule needs to be subtracted.
+Therefore, a reference corresponding to the average energy of a bulk water molecule needs to be subtracted.
 The referenced solvent-solvent energy will be denoted as $E_{ww}^\text{corr}$.
@@
 \end{equation}
-However, when water is replaced from a small region $R$ of interest, such as a single water solvent molecule, almost all interactions are with water solvent molecules outside of this region.
+
+However, when water is replaced from a small region $R$ of interest, such as a single water molecule, almost all interactions are with water molecules outside of this region.
 Therefore, there is no double counting, and the full solvent-solvent energy should be used.
@@
 \end{equation}
-When the region $R$ comprises more than one solvent molecule, interactions within this region are double-counted while interactions to the outside are not.
+
+When the region $R$ comprises more than one molecule, interactions within this region are double-counted while interactions to the outside are not.
 This could be solved using the solvent-solvent energy between each pair of voxels $E_{ww,k\;l}$:
@@
 While this is supported by the standard GIST implementation (using the \inlinecode{doeij} flag), it is rarely done due to the large size of the $E_{ww,k\;l}$ matrix.
+
 Integrating GIST values over the whole grid corresponds to a process where all the solvent is removed into bulk.
@@
 \end{equation}
+
 During the electrostatics calculation, the system is treated as periodic and the energy is split into a short-range term $E_\text{dir}$, which is calculated in direct space using a distance cutoff, and a long-range term $E_\text{rec}$, which is calculated in reciprocal space.
@@
 
-\section{Checklists and Cheat Sheats} \label{sec:checklists}
+\section{Cheat sheet}
+%This will be a cheatsheet containing all options for cpptrajs GIST implementation
+The checklist ``The CPPTRAJ GIST action'' provides a cheat sheet summarizing the important options to the GIST action and its output, as well as frequently used equations.
+
+\subsection{Installation}
+GIST calculations in cpptraj can be sped up through various levels of 
+parallelization.
+If cpptraj was compiled with the '-cuda' configure flag set, the GPU accelerated version of GIST is used automatically. 
+Furthermore, if cpptraj was compiled with OpenMP support, the entropy calculation will be parallelized.
+
+For detailed instructions on how to install \software{cpptraj}, we refer the reader to the AmberMD GitHub page \url{https://github.com/Amber-MD/cpptraj}.
+
+\begin{Checklists*}
+
+\begin{checklist}{The CPPTRAJ GIST action}\label{cheatsheet}
+
+\textbf{Options}\\
+Various flags and options can be provided when running a GIST calculation in 
+cpptraj. A list of possible and required options is provided here:
+
+\begin{tabular}{@{}l p{.7\textwidth}@{}}
+\toprule
+doorder & Calculate the water order parameter [reference] for each voxel\\
+doeij & Calculate the triangular matrix representing the water-water 
+interactions between pairs of voxels.\\
+skipE & Skip all energy calculations (cannot be specified with ’doeij’)\\
+skipS & Skip all the entropy calculations.\\
+refdens <rdval> & Reference density of bulk water, used in computing 
+$g_O, g_H$, and the translational entropy. Default is 0.0334 molecules/Å3. \\
+temp <tval> & Temperature of the input trajectory. \\
+noimage &  Disable distance imaging in energy calculation. \\
+gridcntr <xval> <yval> <zval> & Coordinates (Å) of the center of the grid 
+(default 0.0, 0.0, 0.0). \\
+griddim <xval> <yval> <zval> & Grid dimensions along each coordinate axis 
+(default 40, 40, 40).\\
+gridspacn <spaceval> & Grid spacing (linear dimension of each voxel) in 
+Angstroms. Values greater than 0.75 Å are not recommended (default 0.5 Å).\\
+prefix <filename prefix> & Output file name prefix (default “gist”).\\
+ext <grid extension> & Output grid file name extension (default “.dx”). \\
+out <output> & Name of the main GIST output file. If not specified set to 
+’<prefix>-output.dat’.\\
+info <info> & Name of main GIST info file. If not specified info is 
+written to standard output. \\
+\bottomrule
+\end{tabular}
+
+\textbf{Output}\\
+GIST calculations put out a variety of data sorted by voxel into a file 
+specified by the "out" keyword. Some of the output data is also automatically 
+written to "open data explorer" (.dx) files for convenient visualisation in 
+software such as VMD or PyMOL. The following columns can be found in the output 
+file:
+\begin{tabular}{@{}l l p{.8\textwidth}@{}}
+\toprule
+Name & Keyword & Description\\
+\midrule
+gX & X & For every element in the main solvent, the number density of atoms found in the voxel, in units of the bulk density.
+If the same element occurs multiple times, the bulk density is scaled accordingly.
+\\
+Esw & X &  Mean solute-water interaction energy density. \\
+Eww & X & Mean water-water interaction energy density. \\
+PME & [pme] & Water PME energy density. \\
+U\_PME & [pme] & Water PME energy density. \\
+dTStrans & X & First order translational entropy density. \\
+dTSorient & X & First order orientational entropy density . \\
+neighbor & X & Mean number of waters neighboring the water molecules found in this voxel multiplied by the voxel number density. \\
+dipole & X & Magnitude of mean dipole moment (polarization). \\
+order & X & (if \inlinecode{doorder} was specified) Average Tetrahedral Order Parameter. \\
+dipolex & X & x-component of the mean water dipole moment density \\
+dipoley & X & y-component of the mean water dipole moment density \\
+dipolez & X & z-component of the mean water dipole moment density \\
+Eij & X & (if \inlinecode{doeij} was specified) Water-water interaction matrix. \\
+\bottomrule
+\end{tabular}
+\end{checklist}
+\end{Checklists*}
+
+\section{Checklists}
 %Tutorials do not necessarily require the use of a checklist as in Best Practices documents; however, they can include these if desired.
@@
 % Here is a single-column checklist that consists of multiple sub-checklists
-\begin{Checklists}[H]
+\begin{Checklists}[h]
 
@@
 \begin{itemize}
+	\item GPU, direct space: generally the fastest
+	\item CPU, PME: highly consistent with Amber MD, fast
 	\item CPU: slow, most general (can write out $E_{ij}$ matrices)
-			\item CPU, PME: highly consistent with Amber MD, fast
-			\item GPU, direct space: fastest for a single CPU core
 \end{itemize}
-
-	All methods profit from MPI-parallelization.   
-	PME with multiple CPU cores (>4) is usually the fastest method.
 \end{checklist}
@@
 	\item Insufficient sampling leads to high noise in the results. Use higher numbers of frames.
-			\item Correlation between frames leads to negative entropy in bulk. Increase simulation time and/or reduce the number of frames.
+	\item Autocorrelation between frames leads to negative entropy in bulk. Increase simulation time and/or reduce the number of frames.
 \end{itemize}
@@
 \end{Checklists}
-%\section{Cheat sheet}
-%This will be a cheatsheet containing all options for cpptrajs GIST implementation
-%The checklist ``The CPPTRAJ GIST action'' provides a cheat sheet summarizing the important options to the GIST action and its output, as well as frequently used equations.
-
-%\subsection{Installation}
-%GIST calculations in cpptraj can be sped up through various levels of 
-%parallelization.
-%If cpptraj was compiled with the '-cuda' configure flag set, the GPU accelerated version of GIST is used automatically. 
-%In addition, OpenMP and MPI parallelization are available if cpptraj was compiled the necessary options. 
-%
-%For detailed instructions on how to install \software{cpptraj}, we refer the reader to the AmberMD GitHub page \url{https://github.com/Amber-MD/cpptraj}.
-\FloatBarrier
-\begin{Checklists*}[h!]
-\begin{checklist}{The CPPTRAJ GIST action}\label{cheatsheet}
-
-\textbf{Options}\\
-Various flags and options can be provided when running a GIST calculation in cpptraj. 
-A list of possible and required options is provided here:
-\medskip
-
-\begin{tabularx}{\textwidth}{@{}l X @{} }
-\toprule
-\textbf{I/O Options} &\\
-name <dataset name> & Name for output datasets in cpptraj. \\
-prefix <prefix> & Output file name prefix. Default is 'gist'.\\
-ext <extension> & Output grid file name extension. Default is  '.dx'. \\
-out <file> & Name of the main GIST output file. If not specified set to '<prefix>-output.dat'.\\
-info <file> & Name of main GIST info file. If not specified info is written to standard output. \\
-floatfmt <fmt> & Format for floating point values in GIST output file. Options are 'double', 'scientific' or 'general'. Default chooses 'fixed' or 'scientific' automatically. \\
-floatwidth <val>& Width of floating point values in GIST output file. Default is no width restriction. \\
-floatprec <val>& Precision of floating point values in GIST output file. Default is the system default. \\
-intwidth <val>& Width of integer point values in GIST output file. Default is no width restriction. \\
-doeij & Output the triangular matrix representing the water-water 
-interactions between pairs of voxels. Not supported with PME or GPU GIST.\\
-
-\textbf{Grid Options} &\\
-gridcntr <xval> <yval> <zval> & Coordinates in \AA{} of the center of the grid. Default is 0.0, 0.0, 0.0. \\
-griddim <xval> <yval> <zval> & Grid dimensions in voxels along each coordinate axis. Default is 40, 40, 40.\\
-gridspacn <val> & Grid spacing (linear dimension of each voxel) in 
-Angstroms. Default is \SI{0.5}{\angstrom}.\\
-rmsfit <mask> & If specified, grid will be rotated and translated to follow the atoms selected by mask. \\
-\textbf{GIST Calculation Options} &\\
-
-skipE & Skip all the energy calculations (cannot be specified with ’doeij’)\\
-skipS & Skip all the entropy calculations.\\
-refdens <val> & Reference density of bulk solvent, used in computing 
-$g_O, g_H$, and the translational entropy. Default is 0.0334 molecules/\si{\cubic\angstrom}. \\
-temp <val> & Temperature of the input trajectory. \\
-noimage &  Disable distance imaging in energy calculation. \\
-
-neighborcut <val> & Cutoff in \si{\angstrom} for determining solvent O-O neighbors. Default is \SI{3.5}{\angstrom}. \\
-
-oldnnvolume & Use the old reference volume for the nearest neighbor entropy. \\
-
-nnsearchlayers <val> & Layers of neighboring voxels considered in nearest neighbor search.
-Higher values may improve entropy convergence for little sampling or fine grid spacings, but increase the calculation time.
-Default is 1.
-\\
-
-\textbf{PME Options} & \\
-nopme & Do not use particle mesh Ewald for the non-bonded calculation. This is set  on default. \\
-pme & Use particle mesh Ewald for the non-bonded electrostatics calculation. The van der Waals energy will be calculated using a long-range correction for periodicity.\\
-\qquad cut <val> & Direct space cutoff for pme. Default is \SI{3.5}{\angstrom}.\\
-\qquad dsumtol <val> & Direct sum tolerance used to determine Ewald coefficient. Default is 0.00001.\\
-\qquad ewcoeff <val> & Ewald coefficient in 1/Å. \\
-\qquad erfcdx <val> & Spacing to use for the ERFC splines. Default is \SI{0.0002}{\angstrom}. \\
-\qquad skinnb & Used to determine pairlist atoms (added to cut, so pairlist cutoff is cut + skinnb); included in order to maintain consistency with results from sander. \\
-\qquad ljswidth <val> & If specified, use a force-switching form for the Lennard-Jones calculation from <cutoff>-<val> to <cutoff> \\
-\qquad order <val> & Spline order for charges. \\
-\qquad nfft <nfft1>,<nfft2>,<nfft3> & Explicitly set the
-number of FFT grid points in each dimension.
-Will be determined automatically if not
-specified. \\
-
-\end{tabularx}  
-\end{checklist}
-\end{Checklists*}
-\begin{Checklists*}[h!]
-\begin{checklist}{The CPPTRAJ GIST action}
-
-\textbf{Further Options}\\
-\begin{tabularx}{\textwidth}{@{} l X @{}}
-\toprule
-
-\textbf{Solvent Options} &\\
-solute <mask> & Selection mask for the solute. All other molecules will be solvent. If this is omitted, the standard solute/solvent assignment will be used. \\
-solventmols <mols> &Comma-separated list of solvent molecules residue names.
-Energies will be computed per solvent molecule.
-For the entropy, only the first solvent will be used.
-Use this for calulations with more than one solvent of interest, e.g.
-for  ions.
-\\
-rigidatoms <c> <a1> <a2> & Specifies the molecular orientation for the entropy calculation from a central atom and two additional atoms, e.g.
-O H1 H2 for water.
-By default, a simple heuristic will be used.
-Use this option if the automatically chosen atoms are collinear or do not represent the orientation well. \\
-nocom & Do not use the center of mass to define the molecular position. Instead, use the first atom in rigidatoms. Use this flag to restore the behavior of old GIST runs. \\
-\textbf{Order Calculation Options} &\\
-doorder & Calculate the water order parameter [reference] for each voxel\\
-\qquad nopl & Do not use pair list for order calculation\\
-\qquad plcut <val> & Pair list cutoff for order calculation. Default is \SI{10}{\angstrom}\\
-\bottomrule
-\end{tabularx}   
-\medskip
-
-\textbf{Output}\\
-GIST calculations write a variety of data sorted by voxel into an outputfile specified by the 'out' keyword. 
-Some of the output data is also automatically written to 'open data explorer' (.dx) files for convenient visualisation in software such as VMD or PyMOL.
-Note that some columns will be written out both with '\_norm' and '\_dens' suffixes, refereing to normalization by solvent molecule or voxel volume respectively.
-The following columns can be found in the output file:
-\medskip
-
-\begin{tabularx}{\textwidth}{@{}l l X@{}}
-\toprule
-Name & Keyword & Description\\
-\midrule
-index & & Voxel indices \\
-x, y, z & & Coordinates of the voxel centers \\
-pop & & Population of solvent in voxel over entire simulation \\
-gX &  & For every element in the main solvent, the number density of atoms found in the voxel, in units of the bulk density.
-If the same element occurs multiple times, the bulk density is scaled accordingly.
-\\
-g\_mol\_Y & [solventmols]  & Density of every solvent species Y. Scaled by rho0. \\
-Esw &  &  Mean solute-solvent interaction energy. \\
-Eww &  & Mean solvent-solvent interaction energy. \\
-Esw\_mol\_Y & [solventmols]  &  Mean solute-solvent interaction energy for solvent species Y. \\
-Eww\_mol\_Y & [solventmols]  & Mean solvent-solvent interaction energy for solvent species Y. \\
-PME & [pme] & Solvent PME energy. \\
-U\_PME & [pme] & Solvent PME energy. \\
-dTStrans &  & First order translational entropy. \\
-dTSorient &  & First order orientational entropy. \\
-order &  & (if \inlinecode{doorder} was specified) Average Tetrahedral Order Parameter. \\
-dipolex &  & x-component of the mean solvent dipole moment density \\
-dipoley &  & y-component of the mean solvent dipole moment density \\
-dipolez &  & z-component of the mean solvent dipole moment density \\
-dipole &  & Magnitude of mean dipole moment density (polarization). \\
-neighbor &  & Mean number of solvent molecules neighboring the solvent molecules found in this voxel. \\
-order & & Average tetrahedral order parameter \\
-
-\end{tabularx}
-\end{checklist}
-\end{Checklists*}
-\FloatBarrier
+
 \section*{Author Contributions}
-
 %%%%%%%%%%%%%%%%
@@
 %%%%%%%%%%%%%%%%
-All authors were involved in reviewing and editing the original manuscript. \\  
-VJH and FW conceptualized the research and co-wrote the initial draft.\\
-VM contributed the original drafts of sections \ref{sec:theory} and \ref{sec:visualization}.\\
-HC helped with code curation and validation.\\
-MLFQ helped with conceptualization and gave writing input for the initial draft. \\
-SR and DR programmed cpptraj's GIST implementation and provided technical input.\\  
-KRL, MKG and TK were involved with conceptualization, funding acquisition, project management and supervision.\\
+
+(Explain the contributions of the different authors here)
 
@@
 see the GitHub issue tracking and changelog at \githubrepository.
+
+\section*{Other Contributions}
+%%%%%%%%%%%%%%%
+% You should include all people who have filed issues that were
+% accepted into the paper, or that upon discussion altered what was in the paper.
+% Multiple significant contributions might mean that the contributor
+% should be moved to authorship at the discretion of the a
+%
+% See the policies ``Policies on Authorship'' section of https://livecoms.github.io for
+% more information on deciding on authorship and author order.
+%%%%%%%%%%%%%%%
+
+(Explain the contributions of any non-author contributors here)
+% We suggest you preserve this comment:
+For a more detailed description of contributions from the community and others, see the GitHub issue tracking and changelog at \githubrepository.
 
@@
 
-MKG has an equity interest in and is a cofounder and scientiﬁc advisor of VeraChem LLC.
+Declare any potentially conflicting interests here, whether or not they pose an actual conflict in your view.
 

### 2024-06-30 manuscript/manuscript.tex
Adress Steves comments in intro

--- 2024-06-28 manuscript/manuscript.tex
@@
 The calculation of higher-order entropy terms is computationally demanding but was realized in recent works \cite{Nguyen2016-gist-second-order, Waibl2021-gist-salt}.
-Nevertheless, solvation free energies of small molecule solutes can be computed at high accuracy using linear scaling factors for the higher-order terms, such as $\Delta S = 0.6\, \Delta S^\text{1st-order}$ for water \cite{Chen2021,Waibl2022-gist-solvents}.
+It has been found that solvation free energies of small molecule solutes can be computed at high accuracy using linear scaling factors for the higher-order terms, such as $\Delta S = 0.6\, \Delta S^\text{1st-order}$ for water \cite{Chen2021,Waibl2022-gist-solvents}.
 
@@
 
-On the other hand, methods based on IST generally use a molecular solvent distribution obtained from an MD simulation to compute the free energy of hydration from its enthalpic and entropic contributions, providing easier interpretability of the resulting distributions.
+Methods based on IST use a molecular solvent distribution obtained from an MD simulation to compute the free energy of hydration from its enthalpic and entropic contributions, providing easier interpretability of the resulting distributions.
 Examples include grid inhomogeneous solvation theory (GIST) \cite{Nguyen2012,Ramsey2016}, WaterMap \cite{Young2007-watermap,Abel2008-watermap}, SSTMap \cite{Haider2018-sstmap}, STOW \cite{Li2012-stow} and Solvaware \cite{Huggins2016-solvaware}.
@@
 The main implementation of GIST is part of the MD analysis software \software{cpptraj} released with AmberTools \cite{amber24, Case2023-ambertools}.
-A recent version of \software{cpptraj} should be used. We recommend using at least version 6.15.2, which can be obtained from GitHub (\url{https://github.com/Amber-MD/cpptraj}).
+A recent version of \software{cpptraj} should be used. We recommend using at least version 6.24, which was released with AmberTools24 \cite{Case2023-ambertools}.
 

### 2024-07-01 manuscript/manuscript.tex
Visualisation changes
Changed visualizations for figures 4 and 6 ('norm' binding pocket and 'dens' binding pocket) and added pymol script for figure 4

@@
 	\centering
-	\includegraphics[width=0.8\linewidth]{figures/binding_pocket_S_-2_E_-2_E_3.png}
-	\caption{Free energy contributions in the streptavidin binding pocket.
-Red: high solvent energy (at +3 kcal/mol).
-Blue: low solvent energy (at -2 kcal/mol).
-Green: low solvent entropy $T\Delta S^{six}$ (at -2 kcal/mol).}\label{fig_binding_pocket_pymol}
+	\includegraphics[width=0.8\linewidth]{figures/binding_pocket_norm_quants.png}
+	\caption{Free energy contributions in the streptavidin binding pocket. Red: high solvent energy (at +2.5 
+	kcal/mol/molecule). Green: low solvent energy (at -2.5 kcal/mol/molecule). Blue: low solvent entropy $T\Delta S^{six}$ 
+	(at -2.5 kcal/mol/molecule).}\label{fig_binding_pocket_pymol}
 \end{figure}
@@
 \begin{table}[h]
-	\caption{Free energy contributions for the monomers and the dimer in kcal$\cdot$mol$^{-1}$.
-``Diff'' is calculated as $E^\text{internal} + \Delta E^\text{GIST} - \Delta S^\text{scaled}$.}\label{tab_dg_monomers_dimer}
+	\caption{Free energy contributions for the monomers and the dimer in kcal$\cdot$mol$^{-1}$. ``Diff'' is calculated as 
+	$E^\text{internal} + \Delta E^\text{GIST} - \Delta S^\text{scaled}$. ``total'' is calculated as 
+	``complex''-``streptavidin'' - ``biotin''.}\label{tab_dg_monomers_dimer}
 	\small
@@
 		biotin       & -24.0   &  -98.1 &  -35.1 &  -21.1 &  -101.0 \\
-		Diff         & -98.7   &   97.7 &   44.4 &   26.6 &   -27.6 \\
+		\midrule
+		total         & -98.7   &   97.7 &   44.4 &   26.6 &   -27.6 \\
 		\bottomrule
@@
 	\caption{Biotin in the streptavidin binding pocket, with isosurfaces based on the GIST calculation on the holo structure. 
-Green: negative entropy ($T\Delta S$).
-Blue: negative energy (based on $E_{sw} + 2E_{ww}$).
-Red: positive energy.
-Isolevels are ($\pm$) 0.25, 0.5, 1.0, 1.5, 2.0, 2.5, and 3.0, except for the positive energy, where they are divided by two.}\label{fig-binding-pocket.png}
+	Blue: negative entropy ($T\Delta S$). Green: negative energy (based on $E_{sw} + 2E_{ww}$). Red: positive energy. 
+	Isolevels are ($\pm$) 0.25, 0.5, 1.0, 1.5, 2.0, 2.5, and 3.0 \si{\kcal\per\mole\per\angstrom\squared}, except for the 
+	positive energy, where they are divided by 
+	two.}\label{fig-binding-pocket.png}
 \end{figure}
@@
 
-MKG has an equity interest in and is a cofounder and scientiﬁc advisor of VeraChem LLC.
+MKG has an equity interest in and is a cofounder and scientific advisor of 
+VeraChem LLC.
 

### 2024-07-01 manuscript/manuscript.tex
Update to figure 6
Split figure into two visualisations, one for energy and one for entropy

@@
 
+\usepackage{caption}
+\usepackage{subcaption}
 \usepackage{listings}
@@
 \end{lstlisting}
-A more sophisticated visualization, including the energy, entropy, and free energy of removing a single water solvent molecule, is found in Figure~\ref{fig-binding-pocket.png}
+A more sophisticated visualization, including the energy, entropy, and free energy of removing a single water solvent 
+molecule, is found in Figure~\ref{fig-binding-pocket}
 
@@
 	\centering
-	\includegraphics[width=\linewidth]{figures/binding-pocket-energy-levels.png}
+	\begin{subfigure}[b]{0.45\textwidth}
+	\includegraphics[width=\linewidth]{figures/binding-pocket-E.png}
+	\caption{Green: negative energy (based on $E_{sw} + 2E_{ww}$). Red: positive energy.}
+	\label{fig-binding-pocket-energy}
+    \end{subfigure}
+	\hfill
+	\begin{subfigure}[b]{0.3\textwidth}
+		\includegraphics[width=\linewidth]{figures/binding-pocket-S.png}
+		\caption{Blue: negative entropy}
+		\label{fig-binding-pocket-entropy}
+	\end{subfigure}
 	\caption{Biotin in the streptavidin binding pocket, with isosurfaces based on the GIST calculation on the holo structure. 
-	Blue: negative entropy ($T\Delta S$). Green: negative energy (based on $E_{sw} + 2E_{ww}$). Red: positive energy. 
-	Isolevels are ($\pm$) 0.25, 0.5, 1.0, 1.5, 2.0, 2.5, and 3.0 \si{\kcal\per\mole\per\angstrom\squared}, except for the 
-	positive energy, where they are divided by 
-	two.}\label{fig-binding-pocket.png}
+	Isolevels shown are ($\pm$) 0.25, 0.5, 1.0, 1.5, 2.0, 2.5, and 3.0 \si{\kcal\per\mole\per\angstrom\square}, except for 
+	the positive energy, where they are divided by two.}
+	\label{fig-binding-pocket}
 \end{figure}
@@
 	\includegraphics[width=1.0\linewidth]{figures/streptavidin_gO_high_surf.png}
-	\caption{Oxygen density around streptavidin at an isolevel of twice the reference density}\label{fig-streptavidin_gO}
+	\caption{Oxygen density around streptavidin at an isolevel of twice the reference density}
+	\label{fig-streptavidin_gO}
 \end{figure}

### 2024-07-01 manuscript/manuscript.tex
Fix subfigures, some formatting

@@
     \caption{Schematic showing regions of high water density around a carbazole solute molecule.
-The energy is computed from the intermolecular interactions of a force field, while the entropy is approximated from the 3-dimensional solvent distribution around the solute molecule.
-Note that while this depiction focuses of regions of high solvent density, regions with lower density can also contain meaningful contributions to the energy and entropy.}
+Based on an MD simulation of solvent around the solute molecule, the energy is computed from the intermolecular interactions of a force field, while the entropy is approximated from the 3-dimensional solvent distribution around the solute molecule.
+Note that while this depiction focuses on regions of high solvent density, regions with lower density can also contain meaningful contributions to the overall energy and entropy.}
     \label{fig:carbazole}
@@
 
+\pagebreak %TODO: maybe remove this by better formatting
+
 \section{Running GIST}
@@
 \end{lstlisting}
-Additionally, the GIST calculation should usually be preceded by preprocessing steps, such as centering the solute molecule in the unit cell or reimaging to account for periodic boundary conditions.
-Often, the region of interest will be defined as the surrounding of a molecule, which might for example be a ligand in the binding pocket or the whole protein. 
-Using \inlinecode{cpptraj's } \inlinecode{bounds} command, the region of interest can be output to a file. For example, here is the command for a region of \SI{10}{\angstrom} around the ligand LIG:
+Additionally, the GIST calculation should usually be preceded by centering the solute molecule in the unit cell and reimaging to account for periodic boundary conditions.
+Often, the region of interest will be defined as the surrounding of a molecule, for example a ligand in the binding pocket or a whole protein. 
+Using \inlinecode{cpptraj's } \inlinecode{bounds} command, boundaries can be output to a file. For example, here is the command for a boundary of \SI{10}{\angstrom} around the ligand LIG:
 
@@
 \end{lstlisting}
-Note that the \inlinecode{bounds} command needs to be supplied with the exact same trajectory as used in the GIST calculation to generate correct boundaries.
-Therefore, any preprocessing steps need to match between the \inlinecode{bounds} command and the following \inlinecode{gist} calculation.
-
-In addition, a helper script called \inlinecode{findcentroid.py} for the same purpose is supplied with this tutorial, and can be used as follows:
-
+\begin{minipage}{\linewidth} %TODO: maybe find a better way to keep listing on same page?
+Make sure that the \inlinecode{bounds} command is supplied with the exact same trajectory as in the GIST calculation.
+Any preprocessing steps need to match between the \inlinecode{bounds} command and the \inlinecode{gist} calculation.
+A helper script called \inlinecode{findcentroid.py} for the same purpose is supplied with this tutorial, and can be used as follows:
+~
 \begin{lstlisting}[style=bash]
@@
 \end{lstlisting}
-
+\end{minipage}
 \paragraph{Run GIST using cpptraj}
@@
 	\begin{subfigure}[b]{0.45\textwidth}
-	\includegraphics[width=\linewidth]{figures/binding-pocket-E.png}
+	\includegraphics[width=\linewidth]{figures/binding_pocket_E.png}
 	\caption{Green: negative energy (based on $E_{sw} + 2E_{ww}$). Red: positive energy.}
@@
 	\hfill
-	\begin{subfigure}[b]{0.3\textwidth}
-		\includegraphics[width=\linewidth]{figures/binding-pocket-S.png}
+	\begin{subfigure}[b]{0.45\textwidth}
+		\includegraphics[width=\linewidth]{figures/binding_pocket_S.png}
 		\caption{Blue: negative entropy}
@@
 	\caption{Biotin in the streptavidin binding pocket, with isosurfaces based on the GIST calculation on the holo structure. 
-	Isolevels shown are ($\pm$) 0.25, 0.5, 1.0, 1.5, 2.0, 2.5, and 3.0 \si{\kcal\per\mole\per\angstrom\square}, except for 
-	the positive energy, where they are divided by two.}
+	Isolevels shown are ($\pm$) 0.25, 0.5, 1.0, 1.5, 2.0, 2.5, and 3.0 \si{\kilo\calorie\per\mole\per\square\angstrom}, except for 
+	the positive energy, where they are divided by two to visually account for its overall lower density.}
 	\label{fig-binding-pocket}
@@
 	\centering
-	\includegraphics[width=1.0\linewidth]{figures/streptavidin_gO_high_surf.png}
+	\includegraphics[width=1\linewidth]{figures/streptavidin_gO_high_surf.png}
 	\caption{Oxygen density around streptavidin at an isolevel of twice the reference density}
@@
 	\centering
-	\includegraphics[width=1.0\linewidth]{figures/sasa-simple.png}
+	\includegraphics[width=1\linewidth]{figures/sasa-simple.png}
 	\caption{SAS for streptavidin focused on the binding pocket}\label{fig-streptavidin_sasa}

### 2024-07-01 manuscript/manuscript.tex
More formatting, adress a comment of Steve

@@
 The calculation of these properties and the need for referencing is discussed in-depth in section \ref{sec:theory}.
-
 Densities and reference energies of typical water models can be found in the \software{cpptraj} manual.
@@
 Generally, it is advisable to conduct your own reference calculations, especially for systems that are either very large or very small. 
-The reference values can be calculated from MD trajectories of bulk solvent.
-From this, one then calculates the density-weighted average energy per solvent molecule ($E_{ww, norm}$) and the average solvent density $\rho_0$.
+The reference values can be extracted from MD trajectories of bulk solvent by calculating the density-weighted average energy per solvent molecule ($E_{ww, norm}$) and the average solvent density $\rho_0$.
 Generally, it is recommended that the size of the reference solvent box closely matches that of the production system.
@@
 \end{table}
-
+\pagebreak %TODO: pagebreak
 \paragraph{Postprocessing and Analyzing GIST results}
@@
 
+\pagebreak
 \section{Tutorial}
@@
 	\item Set up and run GIST calculations for a small molecule solute (biotin) and a protein (streptavidin) binding pocket.
-	\item Compute the free energy of solvation for biotin using GIST.
+	\item Compute biotin's free energy of solvation using GIST.
 	\item Visualize local contributions to $\dgsolv$ in the streptavidin binding pocket and around biotin.
@@
 In the first part, we show how to run GIST calculations for small molecule solutes and interpret the output. This provides an example of a simple GIST workflow, without too much details.
-Afterwards, we apply this knowledge to a protein binding pocket, visualize $\dgsolv$ hotspots and estimate a free energy of ligand binding using an end-states approach based on GIST.
+Afterwards, we apply this to a protein binding pocket and estimate a free energy of ligand binding using an end-states approach based on GIST.
 In addition, we discuss all steps shown in the first example in further detail.
@@
 The streptavidin-biotin complex binding is one of the strongest known non-covalent interactions involving a small molecule.
-Therefore, the combination of streptavidin and biotin is used routinely in biotechnology \cite{Dundas2013-streptavidin-review}, and the interaction between them has been investigated in detail \cite{McConnell2021-biotin}.
+Therefore, the combination of streptavidin and biotin is used routinely as a validation case \cite{Dundas2013-streptavidin-review}, and the interaction between them has been investigated in detail \cite{McConnell2021-biotin}.
 

### 2024-07-01 manuscript/manuscript.tex
Finish formatting, change BP vis (fig 9)

@@
 First, extract biotin from the crystal structure and generate a topology and coordinate file compatible with Amber or a simulation package of your choice.
-
 The solute molecule should be solvated with a sufficiently large solvent box. 
 For periodic boundary conditions, the size of the solvent box should be large enough for the solvent to behave bulk-like so as to reduce artifacts from solvent-mediated interactions between the periodic images.
-This corresponds to a distance between solute and box boundary of at least \SI{12}{\angstrom} for water\cite{Chen2021,Waibl2022-gist-solvents} though we recommend \SI{15}{\angstrom} or higher.
+This corresponds to a distance between solute and box boundary of at least \SI{12}{\angstrom} for water \cite{Chen2021,Waibl2022-gist-solvents} though we recommend \SI{15}{\angstrom} or higher.
 Carefully check the employed force field, protonation state and partial charges.
@@
 Use enough equilibration time to allow the solvent to relax. 
-For water, \SI{0.5}{\nano\second} is usually sufficient to equilibrate the density, but narrow binding pockets can take longer.
 A well-tested equilibration protocol can be found in the literature \cite{Roe2020-equilibration}.
@@
 \end{figure}
-
+\pagebreak %TODO pagebreak
 \subsubsection{Visualizing $\dgsolv$}
@@
 We recommend using a \software{Jupyter Notebook} for the following analyses.
-The code is not explained in detail here, but should be clear after reading the previous sections.
 Load the files and double check the frame numbers and reference density.
@@
 \end{lstlisting}
+\pagebreak %TODO pagebreak
 Subtract a reference entropy from the \inlinecode{dTSsix} columns.
@@
 	Isolevels shown are ($\pm$) 0.25, 0.5, 1.0, 1.5, 2.0, 2.5, and 3.0 \si{\kilo\calorie\per\mole\per\square\angstrom}, except for 
-	the positive energy, where they are divided by two to visually account for its overall lower density.}
+	the positive energy, where they are divided by two to visually account for its overall lower density. Only voxels within \SI{4}{\angstrom} of biotin are shown.}
 	\label{fig-binding-pocket}
@@
 %TODO: Hide capping group
+\pagebreak %TODO pagebreak
 \section{Theory}
@@
 \end{equation}
-This is exact assuming that the distribution of the orientation $\omega$ is constant within a voxel $k$, but might converge slower than computing the entropy from the combined space (see below).\\
+This is exact assuming that the distribution of the orientation $\omega$ is constant within a voxel $k$, but might converge slower than computing the entropy from the combined space (see below).
 The second method directly calculates the solvation entropy by evaluating the six-dimensional integral (3 for position and 3 for orientation).
@@
 MLFQ helped with conceptualization and gave writing input for the initial draft. \\
-SR and DR programmed cpptraj's GIST implementation and provided technical input.\\  
+SR and DR contributed technical input.\\  
 KRL, MKG and TK were involved with conceptualization, funding acquisition, project management and supervision.\\
@@
 %%%%%%%
-FMS acknowledges the support of NSF grant CHE-1111111.
 

### 2024-07-01 manuscript/manuscript.tex
Add missing code in 4.3.5, fix code indent

@@
 Information about the grid is stored in the grid property of the gist object, structural data is stored as an \software{mdtraj} trajectory in the struct property:
-%TODO we are missing some code here
+\begin{lstlisting}[style=python]
+gist.grid # Grid object
+gist.struct # mdtraj.Trajectory object
+\end{lstlisting}
 Gisttools also provides OpenDX writing capabilities to visualize the calculated or modified GIST data.
@@
 \end{lstlisting}
-%TODO: This has a higher indent than the other python snippets in the pdf, somehow. Unsure why...
 You will find that the energy disfavors binding, because we do not yet include the interaction energy between biotin and streptavidin.

### 2024-07-08 manuscript/manuscript.tex
Enforce figure placement

@@
 \lstdefinestyle{code}{}
-\lstdefinestyle{cpptraj}{title={\centering Cpptraj}}
+\lstdefinestyle{cpptraj}{title={\centering cpptraj}}
 \lstdefinestyle{bash}{title={\centering Command-line}}
@@
 
-Interactions with water are of special importance as many biological processes occur in aqueous environment \cite{Privalov2017-water-review}.
+Interactions with water are of special importance as many biological processes occur in an aqueous environment \cite{Privalov2017-water-review}.
 The free energy of solvation is defined as the reversible work needed to insert a solute molecule into the solvent from gas phase \cite{ben-naim-book}.
@@
 Since these contributions are spatially resolved, GIST improves interpretability and offers more detailed insights into the solvation of the investigated solute. 
-The entropy is computed from the density distribution of the solvent around the solute molecule, in a first-order approximation. A schematic of this is shown in Figure~\ref{fig:carbazole}. 
+The entropy is computed from orientational and translational distributions of the solvent around the solute molecule, in a first-order approximation.
+A schematic of this is shown in Figure~\ref{fig:carbazole}.
 The calculation of higher-order entropy terms is computationally demanding but was realized in recent works \cite{Nguyen2016-gist-second-order, Waibl2021-gist-salt}.
@@
     \caption{Schematic showing regions of high water density around a carbazole solute molecule.
-Based on an MD simulation of solvent around the solute molecule, the energy is computed from the intermolecular interactions of a force field, while the entropy is approximated from the 3-dimensional solvent distribution around the solute molecule.
+Based on an MD simulation of solvent around the solute molecule, the energy is computed from the intermolecular interactions of a force field, while the entropy is approximated from the 6-dimensional orientational and solvational solvent distribution $\rho$ around the solute molecule.
 Note that while this depiction focuses on regions of high solvent density, regions with lower density can also contain meaningful contributions to the overall energy and entropy.}
@@
 
-GIST has been implemented on the GPU, substantially improving its performance on large systems \cite{Kraml2019-gigist}.
+GIST has been implemented on the GPU, substantially improving its performance for large grids \cite{Kraml2019-gigist}.
 Furthermore, an implementation based on the particle mesh Ewald (PME) \cite{Darden1993-pme} method improves both the efficiency and the agreement with popular molecular dynamics engines \cite{Chen2021}.
@@
 In this tutorial, we aim to give the reader an introduction to GIST and its application to calculate solution thermodynamics properties of interest. 
-While previous publications \cite{Ramsey2016} and tutorials \cite{amber_tut_gist} treated how to setup and run GIST calculations in detail, the post-processing was often treated less rigorously.
+While previous publications \cite{Ramsey2016} and tutorials \cite{amber_tut_gist} treated how to setup and run GIST calculations in detail, the post-processing was often treated in less detail.
 While building on these tutorials, we provide an updated guide to calculating solvation properties using GIST to address recent improvements.
@@
 Furthermore, IST does not consider the internal degrees of freedom of the solvent.
-Therefore, IST-based approaches can only be used with sufficiently rigid solvents.
+Therefore, IST-based approaches currently can only be used with sufficiently rigid solvents, although a recent paper provides the theory needed to generalize to flexible solvent molecules \cite{gilson-freeenergy-2024}.
 Nevertheless, they have been applied on a wide variety of problems involving both small molecules and biological systems as solute molecules.\\
@@
 The tutorial is structured around two examples: a GIST calculation for the small molecule biotin, and a GIST study of biotin-streptavidin binding. 
-We provide a workflow that can be easily adapted to different systems, an introduction to the method and the theory behind it, and an overview of the different implementations of GIST.
-
-We start with a basic GIST analysis of biotin.
-Building on the basics, we present a more advanced GIST study of the streptavidin-biotin complex, reflecting on many common aspects of GIST studies.
+Whereas the biotin example gives an introduction to the basics of GIST calculations, the full biotin-streptavidin analysis showcases details and considerations commonly found in more advanced GIST studies.
 We show how to interpret the three-dimensional contributions of solvation free energy in a binding pocket and around the ligand.
 Additionally, we compute the binding contribution of the free energy of solvation, which requires accurate post-processing of the GIST outputs to avoid unfavorable summation of bias.
-We believe that this example study covers a wide range of applications.\\
-Furthermore, we discuss several technical aspects of GIST studies, such as the normalization of voxel values and how to avoid double counting of energy contributions. 
+In total, we aim to provide a workflow that can be easily adapted to different systems, an introduction to the method and the theory behind it, and an overview of the different implementations of GIST.\\
+Furthermore, we discuss several technical aspects of GIST studies, such as the normalization of voxel values. 
 Additionally, we provide a Python library (\software{gisttools}) to unify the analysis of GIST data produced by various versions of GIST and make the post-processing of GIST results more accessible.
@@
 To use OpenMP, MPI or GPU-accelerated (using CUDA) versions of cpptraj, a different executable name such as \inlinecode{cpptraj.OMP}, \inlinecode{cpptraj.MPI} or \inlinecode{cpptraj.cuda} is often used.
-Therefore, make sure to use the right compilation and executable, to get full benefit of the various improvements to GIST.
+Therefore, make sure to use the right compilation and executable, to get full benefit of the various improvements to GIST as implemented in cpptraj.
 The \inlinecode{gist} action has a large number of options to modify the GIST calculation or its output.
@@
 	\item Visualize local contributions to $\dgsolv$ in the streptavidin binding pocket and around biotin.
-	\item Compute the hydration contribution to the binding. Combined with the biotin--streptavidin interaction energy, the binding free energy can be estimated.
+	\item Compute the hydration contribution of the biotin, streptavidin and complex to the binding. Combined with the biotin--streptavidin interaction energy, the binding free energy can be estimated.
 	\item Explore advantages and disadvantages of the GIST method when used to calculate  $\dgsolv$
@@
 Now, you can run a restrained MD simulation based on the equilibrated structure of biotin for the subsequent GIST analysis.
-Restraints should be used during the MD because IST theory is not defined for moving solute molecules.
+Restraints should be used during the MD because IST theory is not well-defined for moving solute molecules.
 We apply a harmonic restraint of \SI{100}{\kilo\calorie\per\mole\per\angstrom\squared} to all heavy atoms to keep the system in place.
@@
 To get bulk-like water properties at the grid border, a buffer distance of \SI{15.0}{\angstrom} is reasonable for biotin.
-The grid parameters can be directly calculated using \software{Cpptraj's} \inlinecode{bound} command:
+The grid parameters can be directly calculated using \software{cpptraj's} \inlinecode{bound} command:
 \begin{lstlisting}[style=cpptraj]
@@
 The GIST data can be accessed and modified from the gist object in various ways.
-Gisttools provides access to single or multiple voxels through a \code{.loc} property similiar to the pandas library. 
+Gisttools provides access to single or multiple voxels through a \code{.loc} property similar to the pandas library. 
 GIST quantity data for all voxels can be accessed by directly indexing the gist object, either by providing a single name or a list of names.
@@
 \end{lstlisting}
-Gisttools provides access to all quantities listed in the GIST output file.
-Additionally, it calculates the total enthalpy per voxel (Eall) as a sum of Eww and Esw, and the free energy of solvation (A) by subtracting $\Delta$TSsix from Eall.
-
-Information about the grid is stored in the grid property of the gist object, structural data is stored as an \software{mdtraj} trajectory in the struct property:
+Gisttools also calculates the total enthalpy per voxel (Eall) as a sum of Eww and Esw, and the free energy of solvation (A) by subtracting $\Delta$TSsix from Eall.
+
+Information about the grid is stored in the grid property of the gist object and structural data (supplied as a pdb of the solute molecule) is stored as an \software{mdtraj} trajectory in the struct property:
 \begin{lstlisting}[style=python]
@@
 When starting from an experimental structure, care has to be taken during the preparation step.
-Make sure to choose a proper protonation state for all titratable moeties.
+Make sure to choose a proper protonation state for all titratable moieties.
 You will need to solvate your structure in a solvent box, which should be large enough to obtain unbiased long-range electrostatics and avoid solvent-mediated interactions of periodic images.
@@
 Note that, due to the restraints and finite sampling time, some solvent sites might be inaccessible to the bulk solvent.
-To investigate such positions, make sure to explicitly place water solvent molecules there. 
+To investigate such positions, make sure that the number of explicitly placed solvent molecules in such sites matches with experiment or your best guess. 
 For cases where the exact occupation of such sites is unknown, it may be helpful to include grand-canonical Monte Carlo (GCMC) or nonequilibrium candidate Monte Carlo (NCMC) sampling steps in the equilibration or production stage \cite{Ge-GCMC2022,Melling-NCMC2023}.
@@
 \end{lstlisting}
-You can also plot the difference between the complex and monomer contributions against the distance to biotin, using the complex coordinates for the holo structure:
+Finally, you can also plot the difference between the complex and monomer contributions against the distance to biotin, using the complex coordinates for the holo structure:
 
@@
 The expected result is shown in Figure~\ref{fig-dg-sums}.
-\begin{figure}[h]
+\begin{figure}[H]
 	\centering
@@
 \end{lstlisting}
-It has been shown \cite{Chen2021,Waibl2022-gist-solvents} that the solvation entropy in water is best approximated by 0.6 times the first order entropy.
-Also compute the scaled entropy and check the effect on the binding affinity.
+It has been shown \cite{Chen2021,Waibl2022-gist-solvents} that the solvation entropy in water is best approximated by 0.6 times the first order entropy provided by GIST.
+So we also compute the scaled entropy and check the effect on the binding affinity.
 The expected results are summarized in Table~\ref{tab_dg_monomers_dimer}.
@@
 		\toprule
-		System       & E(internal) & $\Delta E^\textit{GIST}$ & $T\Delta S^\textit{GIST}$ & $T\Delta S^\textit{scaled}$ & Diff \\
+		System       & $E^{internal}$ & $\Delta E^\textit{GIST}$ & $T\Delta S^\textit{GIST}$ & $T\Delta S^\textit{scaled}$ & Diff \\
 		\midrule
@@
 \end{table}
-Even though streptavidin-biotin is known to be a very stable complex, the solvation energy favors the dissociation.
+Even though streptavidin-biotin is known to be a very stable complex, the free solvation energy ($\Delta E^{GIST}-T\Delta S^{scaled}$) favors the dissociation.
 This is expected: the binding of biotin to streptavidin is faciliated over a large number of H-bonds and over polar interactions. 
 As such, water favourably solvates both biotin and streptavidin and an energetic penalty is paid when displacing the water during the binding process. 
-In this case, we find that the energy of solvation and the internal energy of the complex more or less cancel out, indicating that binding is largely entropy-driven in this case.
+In this case, we find that the changes in the energy of solvation and of the internal energy $\Delta E^{internal}$ more or less cancel out, indicating that binding is largely entropy-driven in this case.
 At first, this is surprising, since isothermal titration calorimetry (ITC) of biotin-streptavidin shows strong enthalpic binding contributions \cite{mpye2020-biotin-itc,hyre2006-biotin-itc}.
@@
 A minimal PyMOL input to visualize a dx file is shown here.
-For reference, an input structure is loaded from a file named \inlinecode{streptavidin.pdb}.
-The oxygen density is visualized at an isolevel of 2 (i.e., twice the reference density).
+This loads an input structure from a file named \inlinecode{streptavidin.pdb} and visualizes the oxygen density at an isolevel of 2 (i.e., twice the reference density).
 The expected output is shown in Figure~\ref{fig-streptavidin_gO}.
@@
 A more sophisticated visualization, including the energy, entropy, and free energy of removing a single water solvent 
-molecule, is found in Figure~\ref{fig-binding-pocket}
-
+molecule, is found in Figure~\ref{fig-binding-pocket}.
+\begin{figure}
+	\centering
+	\includegraphics[width=1\linewidth]{figures/streptavidin_gO_high_surf.png}
+	\caption{Oxygen density around streptavidin at an isolevel of twice the reference density (i.e., bulk)}
+	\label{fig-streptavidin_gO}
+\end{figure}
 \begin{figure}
@@
 
-\begin{figure}
-	\centering
-	\includegraphics[width=1\linewidth]{figures/streptavidin_gO_high_surf.png}
-	\caption{Oxygen density around streptavidin at an isolevel of twice the reference density}
-	\label{fig-streptavidin_gO}
-\end{figure}
-
 \subsection{Solvent Accessible Surface (SAS)}
-
 The solvent accessible surface (SAS) represents the interface between regions that are occupied by the solute molecule, and those which can be accessed by the solvent.
@@
 It can be visualized in PyMOL. 
+Here, the SAS was generated from a simulation of water around streptavidin without biotin, leaving the binding site solvent-accessible.
 The expected output is shown in Figure~\ref{fig-streptavidin_sasa}.
@@
 \end{figure}
-
-Here, the SAS was generated from a simulation of water around streptavidin without biotin, leaving the binding site solvent-accessible.
 \subsection{Defining the binding pocket}
@@
 This is significant specially when calculating the properties of water that will be expelled when a ligand binds to the protein.\\
-For example, voxels that are part of the binding pocket can be defined using a ligand solute molecule and a distance criteria. 
-In section~\ref{sec:binding_contributions} we show how to integrate over regions such as the binding pocket.
+For example, voxels that are located within the binding pocket can be defined using a ligand solute molecule and a distance criteria. 
+In section~\ref{sec:binding_contributions}, we show how to integrate over regions such as the binding pocket.
 To visualize the considered region, we can create a OpenDX file where we set the considered voxels to 1 and all others to 0.
@@
 
-\subsubsection{Entropy Calculations in Cpptraj}
+\subsubsection{Entropy Calculations in cpptraj}
 In \software{cpptraj}, the calculation of solvation entropy is handled by two methods.\\
@@
 This depends on the local environment and on \emph{what} the solvent would be replaced by.
-Therefore, the effect of reorganization can not be directly included in the GIST method, but must be judged for each case individually.
-
+Therefore, the effect of reorganization must be judged for each case individually.
+One way to treat reorganization is an end-state approach as shown in the tutorial: all states before and after solvent reorganization are considered.
+In the tutorial, this corresponds to the solvation of biotin and streptavidin seperated (initial states) and in complex (end-state).
+Note that all solvent contributions need to be considered to get thermodynamically accurate results in such a case.
+In practice this therefore requires the integration over the whole grid or at least all voxels with solvent properties different from bulk (i.e.
+setting a large cutoff).
 \subsubsection{PME energy}

### 2024-07-08 manuscript/manuscript.tex
Remove entropy eq and add more description to SAS

@@
 GIST provides a very natural way of creating a SAS using an isosurface of the oxygen density \code{gO}.
-It can be visualized in PyMOL. 
+At a very low isovalue, an isosurface of the water's oxygen or hydrogen shows areas the water can reach, essentially creating a SAS. 
 Here, the SAS was generated from a simulation of water around streptavidin without biotin, leaving the binding site solvent-accessible.
@@
 
-\begin{equation}
-g_\textit{vox} \left( \textbf{r}, \omega \right) \approx g_\textit{vox}(\textbf{r}) g_\textit{vox}(\omega)
-\end{equation}
-A first nearest neighbor (NN) approach is used to evaluate the density.
-The densities are computed relative to a homogeneous distribution of solvent molecules, given a bulk density $\rho^0$.
+%\begin{equation}
+%g_\textit{vox} \left( \textbf{r}, \omega \right) \approx g_\textit{vox}(\textbf{r}) g_\textit{vox}(\omega)
+%\end{equation}
+A first nearest neighbor (NN) approach is used to evaluate the entropy based on the orientational and translational distributions.
+The distributions are computed relative to a homogeneous distribution of solvent molecules, given a bulk density $\rho^0$.
 The average entropy contributions per solvent molecule in voxel $k$ are calculated as
@@
 The solute-solvent energy $E_{sw}$ can be expressed in terms of the solvent density and the potential $U_{sw}(\textbf{r},\omega)$ induced by the solute molecule.
-In practice, it is evaluated as the expectation value $\langle\cdot\rangle$ of the pairwise force field energy $U_{ij}$ between all solvent molecules in voxel $k$ and all solute atoms.
+In practice, $E_{sw}$ is computed as the expectation value $\langle\cdot\rangle$ of the pairwise force field energy $U_{ij}$ between all solvent molecules in voxel $k$ and all solute atoms.
 
@@
 \subsubsection{PME energy}
-
 In the original version of GIST, the energies are calculated based on the the minimum image convention.

### 2024-07-10 manuscript/manuscript.tex
Some cleanup in section 1 - Change figure placements in manuscript to be in line with pdf - small text changes - remove even more paragraphs and replace with line breaks - fix citation for Gilson 2024

@@
 \corr{thomas.kurtzman@lehman.cuny.edu}{TK}
-
-%TODO Add ORCIDS
-\orcid{Author 1 name}{AAAA-BBBB-CCCC-DDDD}
-\orcid{Author 2 name}{EEEE-FFFF-GGGG-HHHH}
+%TODO Vjays ORCID is missing
+\orcid{Valentin J Hoerschinger}{0000-0002-4469-3238}
+\orcid{Franz Waibl}{0000-0003-0527-0803}
+\orcid{Vjay Molino}{}
+\orcid{Helmut Carter}{0000-0003-0273-4107}
+\orcid{Monica L Fern{\'a}ndez-Quintero}{0000-0002-6811-6283}
+\orcid{Steven Ramsey}{0000-0001-7441-3228}
+\orcid{Daniel R Roe}{0000-0002-5834-2447}
+\orcid{Klaus R Liedl}{0000-0002-0985-2299}
+\orcid{Michael K Gilson}{0000-0002-3375-1738}
+\orcid{Tom Kurtzman}{0000-0003-0900-772X}
 
@@
 
-\section{Introduction}
-Solvation thermodynamics govern many biological and chemical processes involving, e.g., aqueous solutions, hydrophobic effects, or partitioning between different environments.
-
-Interactions with water are of special importance as many biological processes occur in an aqueous environment \cite{Privalov2017-water-review}.
-The free energy of solvation is defined as the reversible work needed to insert a solute molecule into the solvent from gas phase \cite{ben-naim-book}.
-It is directly linked to the solubility of gases and drives many real-world effects such as hydrophobicity, ligand binding, and protein folding.
-
-Grid inhomogeneous solvation theory (GIST) \cite{Nguyen2012} is a method to compute the free energy of solvation, with applications in the context of drug design and biophysical property description.
-It computes the solvent density from an MD trajectory and estimates localized entropic and enthalpic contributions on a grid around the solute molecule.
-Since these contributions are spatially resolved, GIST improves interpretability and offers more detailed insights into the solvation of the investigated solute. 
-The entropy is computed from orientational and translational distributions of the solvent around the solute molecule, in a first-order approximation.
-A schematic of this is shown in Figure~\ref{fig:carbazole}.
-The calculation of higher-order entropy terms is computationally demanding but was realized in recent works \cite{Nguyen2016-gist-second-order, Waibl2021-gist-salt}.
-It has been found that solvation free energies of small molecule solutes can be computed at high accuracy using linear scaling factors for the higher-order terms, such as $\Delta S = 0.6\, \Delta S^\text{1st-order}$ for water \cite{Chen2021,Waibl2022-gist-solvents}.
-
-\begin{figure}
+\begin{figure}[H]
 	\centering
@@
 \end{figure}
-
+\section{Introduction}
+Solvation thermodynamics govern many biological and chemical processes involving, e.g., aqueous solutions, hydrophobic effects, or partitioning between different environments.
+Interactions with water are of special importance as many biological processes occur in an aqueous environment \cite{Privalov2017-water-review}.
+The free energy of solvation is defined as the reversible work needed to insert a solute molecule into the solvent from gas phase \cite{ben-naim-book}.
+It is directly linked to the solubility of gases and drives many real-world effects such as hydrophobicity, ligand binding, and protein folding.
+
+Grid inhomogeneous solvation theory (GIST) \cite{Nguyen2012} is a method to compute the free energy of solvation, with applications in the context of drug design and biophysical property description.
+It computes the solvent density from an MD trajectory and estimates localized entropic and enthalpic contributions on a grid around the solute molecule.
+Since these contributions are spatially resolved, GIST improves interpretability and offers more detailed insights into the solvation of the investigated solute.
+A schematic of this is shown in Figure~\ref{fig:carbazole}.  
+The entropy is computed from orientational and translational distributions of the solvent around the solute molecule, in a first-order approximation. 
+The calculation of higher-order entropy terms is computationally demanding but was realized in recent works \cite{Nguyen2016-gist-second-order, Waibl2021-gist-salt}.
+It has been found that solvation free energies of small molecule solutes can be computed at high accuracy using linear scaling factors for the higher-order terms, such as $\Delta S = 0.6\, \Delta S^\text{1st-order}$ for water \cite{Chen2021,Waibl2022-gist-solvents}.
 Since its initial development, GIST has been used to describe the thermodynamics of water displacement in protein binding pockets \cite{Nguyen2016-gist-second-order,Ramsey2016,Balius2017-gist-ligand-discovery,Hufner-Wulsdorf2020-gist-drug-design}.
@@
 Methods such as MM/PBSA or MM/GBSA \cite{Sitkoff1994-pbsa,Kollman2000-mmpbsa} combine an implicit treatment of solvent electrostatics with structural sampling of the solute molecule.
-They can quickly provide results for a large number of structures \cite{Genheden2015-mmpbsa-review}.
-
+They can quickly provide results for a large number of structures \cite{Genheden2015-mmpbsa-review}.\\
 On the other hand, explicit solvation methods model the solvent using individual molecules.
@@
 However, those methods are unable to provide a spatial interpretation of the results.
-Furthermore, splitting the free energy into enthalpy and entropy contributions is challenging \cite{Peter2004-alchemical-entropy}.
-%Assuming sufficient sampling, they can accurately predict the free energy of solvation in the boundaries of the assumptions underlying their forcefield. 
-%However, the seperation of enthalpic and entropic contributions to the free energy of solvation is difficult with these methods and localizing these contributions not easily possible.
-
+Furthermore, splitting the free energy into enthalpy and entropy contributions is challenging \cite{Peter2004-alchemical-entropy}.\\
 Methods derived from statistical mechanics bridge this gap by providing three-dimensional resolution.
@@
 While there are approximations available to deal with those problems, most noteworthy the mutual information expansion (MIE) and Maximum Information Spanning Tree (MIST) approaches, they are not currently implemented in GIST.
-Recent kNN implementations to calculate solvation entropy further improve phase space sampling by relabeling the solvent molecules in the simulation, treating the solvent molecules as distinguishable \cite{reinhard2007estimation,heinz2019computing,heinz2021permut,fogolari2021optimal}.For a recent review of kNN methods to calculate entropy in a molecular context, see \cite{fogolari2024k}.\\
+Recent kNN implementations to calculate solvation entropy further improve phase space sampling by relabeling the solvent molecules in the simulation, treating the solvent molecules as distinguishable \cite{reinhard2007estimation,heinz2019computing,heinz2021permut,fogolari2021optimal}.
+For a recent review of kNN methods to calculate entropy in a molecular context, see \cite{fogolari2024k}.
 Furthermore, IST does not consider the internal degrees of freedom of the solvent.
-Therefore, IST-based approaches currently can only be used with sufficiently rigid solvents, although a recent paper provides the theory needed to generalize to flexible solvent molecules \cite{gilson-freeenergy-2024}.
+Therefore, IST-based approaches currently can only be used with sufficiently rigid solvents, although a recent paper provides the theory needed to generalize to flexible solvent molecules \cite{gilson2024freeenergy}.
 Nevertheless, they have been applied on a wide variety of problems involving both small molecules and biological systems as solute molecules.\\
@@
 On the other hand, the first-order approximation means that the solvation entropy is less accurately described than in TI or FEP \cite{Chen2021,Waibl2021-gist-salt}. 
-Similar to the self-consistent integral theories, however, GIST only considers a single solute conformation and  multiple computations of representative structures are necessary to describe ensemble properties.
+Similar to the self-consistent integral theories, GIST only considers a single solute conformation and  multiple computations of representative structures are necessary to describe ensemble properties.
 
@@
 The reference values can be extracted from MD trajectories of bulk solvent by calculating the density-weighted average energy per solvent molecule ($E_{ww, norm}$) and the average solvent density $\rho_0$.
-Generally, it is recommended that the size of the reference solvent box closely matches that of the production system.
+It is recommended that the size of the reference solvent box closely matches that of the production system.
 While the impact on the density is not as pronounced, deviations from the production settings can lead to differences of up to 0.005 kcal/mol in the energy per solvent molecule ($E_{ww, norm}$) for water boxes.

### 2024-07-10 manuscript/manuscript.tex
Cleanup of section 2

@@
 The MD simulations presented in this tutorial are run with the Amber simulation engine and analyzed using AmberTools \software{cpptraj}. 
-For a conceptual overview of running MD simulations with Amber, we recommend reading the Amber manual (\url{https://ambermd.org/Manuals.php}) \cite{amber24} and the Amber tutorials (\url{https://ambermd.org/Manuals.php}).
-Previous experience of GIST is not necessary, although we recommend reading the works referenced in section~\ref{sec:theory}.
-
+For a conceptual overview of running MD simulations with Amber, we recommend reading the Amber manual (\url{https://ambermd.org/Manuals.php}) \cite{amber24} and the Amber tutorials (\url{https://ambermd.org/tutorials}).
+Previous experience of GIST is not necessary, although we recommend reading the works referenced in section~\ref{sec:theory}.\\
 The presented analyses are run in Python.
@@
 \subsection{Software/system requirements}
-%Tutorials should clearly define what system and/or software requirements the researcher will need to complete the tutorial (e.g., VMD version 1.9 or newer, AMBER, etc.).
-Tutorials requiring specific software packages must provide instructions and files for the referenced version of the software.
 The main implementation of GIST is part of the MD analysis software \software{cpptraj} released with AmberTools \cite{amber24, Case2023-ambertools}.
@@
 If using GROMACS \cite{Abraham2015-gromacs,Pronk2013-gromacs}, we propose to prepare the topology with AmberTools \cite{Case2023-ambertools} and convert it using \software{acpype.py} \cite{Sousa_da_Silva2012-acpype}.
-%We should actually test that!
 Also note that compressed trajectory formats such as \inlinecode{xtc} might bias the entropy calculation due to the loss of precision for the atom positions.
@@
 We additionally provide a Jupyter notebook \cite{Kluyver2016-jupyter,Granger2021-jupyter} with the presented analyses. 
-%We recommend JupyterLab to work with the notebook. Jupyter is available at \url{https://jupyter.org/}.
-
 For the visualization of three-dimensional contributions we require molecular visualization software that can handle \software{OpenDX} files, such as \software{PyMol} \cite{pymol} or \software{VMD} \cite{vmd}.
@@
 \label{sec:running_GIST}
-%\subsection{Running the GIST analysis}
 To run a GIST analysis, the following is required:
@@
 The grid is defined by parameters of the \inlinecode{gist} command in \software{cpptraj}.
-The grid center is defined by \inlinecode{gridcntr}, the voxel number in x, y, and z direction by \inlinecode{griddim}, and the spacing (voxel side-length) in \AA{} by \inlinecode{gridspacn}:
+The grid center is defined by its center coordinates \inlinecode{gridcntr}, the voxel number in x, y, and z direction by \inlinecode{griddim}, and the spacing (voxel side-length) in \AA{} by \inlinecode{gridspacn}:
 
 \begin{lstlisting}[style=cpptraj]
-gist gridcntr <x> <y> <z> \
-griddim <Nx> <Ny> <Nz> gridspacn <val>
+gist gridcntr <x> <y> <z> griddim <Nx> <Ny> <Nz> gridspacn <val>
 \end{lstlisting}

### 2024-07-10 manuscript/manuscript.tex
More text changes in sec 4&5, adds make commands for md and gist

@@
 \paragraph{Postprocessing and Analyzing GIST results}
-Cpptraj outputs quantities in OpenDX (.dx) file format which can be visualized in Pymol or VMD.
+Cpptraj outputs grid quantities in OpenDX (.dx) file format which can be visualized in Pymol or VMD.
 Aside from thermodynamic quantities, GIST also calculates the number density of each atomic element in the solvent.
@@
 \end{enumerate}
-
 In the first part, we show how to run GIST calculations for small molecule solutes and interpret the output. This provides an example of a simple GIST workflow, without too much details.
@@
 To decide on a grid size, find the extent (minimum and maximum) of the x-, \mbox{y-,} and z-coordinates of the molecule.
-
-To get bulk-like water properties at the grid border, a buffer distance of \SI{15.0}{\angstrom} is reasonable for biotin.
+To get bulk-like water properties at the grid border, a buffer distance of \SI{15}{\angstrom} is reasonable for biotin.
 The grid parameters can be directly calculated using \software{cpptraj's} \inlinecode{bound} command:
@@
 \end{lstlisting}
-Gisttools also calculates the total enthalpy per voxel (Eall) as a sum of Eww and Esw, and the free energy of solvation (A) by subtracting $\Delta$TSsix from Eall.
-
+Gisttools also calculates the total enthalpy per voxel (Eall) as a sum of Eww and Esw, and the free energy of solvation (A) by subtracting $\Delta$TSsix from Eall.\\
 Information about the grid is stored in the grid property of the gist object and structural data (supplied as a pdb of the solute molecule) is stored as an \software{mdtraj} trajectory in the struct property:
@@
 \subsubsection{GIST analysis of Streptavidin-Biotin}
-
+To create input \code{bash} scripts for the MD simulations and GIST analysis, run the following commands in the \code{code} folder:
+\begin{lstlisting}[style=bash]
+	make gist-md-inputs
+	make gist-inputs
+\end{lstlisting}
 After running the MD simulations as described in section \ref{sec:md_simulation}, proceed with the GIST analysis.
 For easier post-processing, here we use grids that contain the whole solute molecule and the surrounding solvent.
-
 If one is only interested in solvent in the binding site, smaller grids encompassing only the binding site could be used for streptavidin and the complex.
@@
 The solute-solvent energy $E_{sw}$ naturally tends towards zero in bulk.
-However, the solvent-solvent energy $E_{ww}$ needs to be referenced.
-
+However, the solvent-solvent energy $E_{ww}$ needs to be referenced.\\
 The Amber manual provides reference densities and energies for several solvents.
@@
 \end{figure}
-
 After choosing an appropriate reference value, we need to subtract this value from $E_{ww}$.
@@
 \end{figure}
-\pagebreak %TODO pagebreak
+
 \subsubsection{Visualizing $\dgsolv$}
-Next, we want to visualize the 3-dimensional contributions to the free energy of hydration, as well as the entropy and enthalpy contributions.
+Next, we visualize the 3-dimensional contributions to the free energy of hydration, as well as the entropy and enthalpy contributions.
 You can use PyMol\cite{pymol} or VMD\cite{vmd} to visualize the .dx files from GIST\@.
-%With \software{gistpp}, you can use the file from the convergence check. Additional OpenDX files can be produced using \inlinecode{gistpp -op makedx}.
 With \software{gisttools}, you can create one using e.g., \inlinecode{gist.save\_dx('A\_dens', 'A\_dens.dx')}.
@@
 \end{lstlisting}
-
 The OpenDX files can then be imported into your molecular viewer of choice.
-%TODO we are missing a pymol script here
-%Adapt the above PyMOL script to visualize the average energy per solvent at a positive and a negative isolevel, as well as regions of negative entropy per solvent.
 A visualization of entropy and energy contributions to the free energy of solvation of the streptavidin binding pocket is shown in Figure~\ref{fig_binding_pocket_pymol}. 
@@
 \end{figure}
-
-%You should find that there are several regions in the binding pocket that feature a high solvent energy and/or low entropy.
-%However, such regions also exist outside of the binding pocket, and there are also low-energy regions within the binding pocket.
-%Therefore, the strong affinity to biotin cannot be explained only by hydration, and the direct interaction energy also needs to be taken into account.
 
@@
 Therefore, we align the complex structure to streptavidin and use the shifted biotin coordinates to define the integration region.
-
 We recommend using a \software{Jupyter Notebook} for the following analyses.
@@
 \end{lstlisting}
-\pagebreak %TODO pagebreak
 Subtract a reference entropy from the \inlinecode{dTSsix} columns.

### 2024-07-10 manuscript/manuscript.tex
Remove pdf from gitignore

@@
 We recommend using PME in combination with PME-GIST, but not with GPU-GIST to be in line with the different ways the energy is calculated in the two GIST methods.
-An example \software{cpptraj} input might look like:
+An example \software{cpptraj} is then:
 \begin{lstlisting}[style=cpptraj]
@@
 Even though streptavidin-biotin is known to be a very stable complex, the free solvation energy ($\Delta E^{GIST}-T\Delta S^{scaled}$) favors the dissociation.
-This is expected: the binding of biotin to streptavidin is faciliated over a large number of H-bonds and over polar interactions. 
+This is expected: the binding of biotin to streptavidin is facilitated over a large number of H-bonds and over polar interactions. 
 As such, water favourably solvates both biotin and streptavidin and an energetic penalty is paid when displacing the water during the binding process. 
@@
 To speed up the calculation, smaller GIST grids could be used by focusing on the binding pocket, or by rotating the solute molecule by its principal axes to fit the cuboid grid more exactly.
-However, this should be done \emph{before} the MD, since rotating the trajectory damages the periodic box information.
+However, this should be done \emph{before} the MD simulation, since rotating the trajectory damages the periodic box information.
 %TODO: Ask Dan if this is still the case
@@
 \subsection{MPI parallelization}
-The most recent addition to GIST is the MPI parallelization of both energy and entropy calculations\cite{Roe2023-mpi-gist}.
+The most recent addition to GIST is the MPI parallelization of both energy and entropy calculations \cite{Roe2023-mpi-gist}.
 Since the MPI parallelization is orthogonal to the other improvements, it can be used with both PME or GPU accelerated GIST.
@@
 The latter is probably more relevant since it can incorporate hydrogen bonding effects.
-Assuming that the alcohol hydrogen is called H1, this could be specified as follows:
+Assuming that the alcohol hydrogen atom is called H1, this could be specified as follows:
 

### 2024-07-10 manuscript/manuscript.tex

@@
 We apply a harmonic restraint of \SI{100}{\kilo\calorie\per\mole\per\angstrom\squared} to all heavy atoms to keep the system in place.
+Note that this is a rather strong restraint.
+While lower values are possible, make sure that the solute molecule is kept rigid during the simulation.
+As a minimum value, we recommend \SI{100}{\kilo\calorie\per\mole\per\angstrom\squared} restraints.
 We run \SI{100}{\nano\second} of simulation and store the conformations every \SI{100}{\pico\second}.
@@
 	\includegraphics[width=\linewidth]{figures/binding_pocket_E.png}
-	\caption{Green: negative energy (based on $E_{sw} + 2E_{ww}$). Red: positive energy.}
+	\caption{GIST solvent energy (based on $E_{sw} + 2E_{ww}$).}
 	\label{fig-binding-pocket-energy}
@@
 		\includegraphics[width=\linewidth]{figures/binding_pocket_S.png}
-		\caption{Blue: negative entropy}
+		\caption{GIST solvent entropy}
 		\label{fig-binding-pocket-entropy}
@@
 	\caption{Biotin in the streptavidin binding pocket, with isosurfaces based on the GIST calculation on the holo structure. 
-	Isolevels shown are ($\pm$) 0.25, 0.5, 1.0, 1.5, 2.0, 2.5, and 3.0 \si{\kilo\calorie\per\mole\per\square\angstrom}, except for 
-	the positive energy, where they are divided by two to visually account for its overall lower density. Only voxels within \SI{4}{\angstrom} of biotin are shown.}
+	Isolevels shown are ($\pm$) 0.25, 0.5, 1.0, 1.5, 2.0, 2.5, and 3.0 \si{\kilo\calorie\per\mole\per\cubic\angstrom}, except for 
+	the positive energy, where they are divided by two to visually account for its overall lower density.
+The negative solvent energgy is shown in green, positive solvent energy in red and negative entropy in blue.
+Only voxels within \SI{4}{\angstrom} of biotin are shown.}
 	\label{fig-binding-pocket}
@@
 			\item Number of analyzed frames: \num{10000} (\num{100000})
-			\item Restraints: typically \SI{100}{\kcalPerMolASqr} on solute heavy atoms.
+			\item Restraints: typically \SI{100}{\kcalPerMolASqr} on solute heavy atoms. Minimum: \SI{10}{\kcalPerMolASqr}
 		\end{itemize}

### 2024-07-10 manuscript/manuscript.tex
Fix minimum restraint number, checklist formatting

@@
 While lower values are possible, make sure that the solute molecule is kept rigid during the simulation.
-As a minimum value, we recommend \SI{100}{\kilo\calorie\per\mole\per\angstrom\squared} restraints.
+As a minimum value, we recommend \SI{10}{\kilo\calorie\per\mole\per\angstrom\squared} restraints.
 We run \SI{100}{\nano\second} of simulation and store the conformations every \SI{100}{\pico\second}.
@@
 			\item Number of analyzed frames: \num{10000} (\num{100000})
-			\item Restraints: typically \SI{100}{\kcalPerMolASqr} on solute heavy atoms. Minimum: \SI{10}{\kcalPerMolASqr}
+			\item Restraints: minimum \SI{10}{\kcalPerMolASqr} on solute heavy atoms. (typically \SI{100}{\kcalPerMolASqr})
 		\end{itemize}

### 2024-07-11 manuscript/manuscript.tex

@@
 \corr{thomas.kurtzman@lehman.cuny.edu}{TK}
-%TODO Vjays ORCID is missing
 \orcid{Valentin J Hoerschinger}{0000-0002-4469-3238}
 \orcid{Franz Waibl}{0000-0003-0527-0803}
-\orcid{Vjay Molino}{}
+\orcid{Vjay Molino}{0009-0005-2055-9295}
 \orcid{Helmut Carter}{0000-0003-0273-4107}
@@
 
-In this tutorial, we aim to guide the reader through the most common steps involved in a GIST analysis using the streptavidin-biotin complex as an demonstrative system.
+In this tutorial, we aim to guide the reader through the most common steps involved in a GIST analysis using the streptavidin-biotin complex as a demonstrative system.
 Towards this, Jupyter notebooks and Python packages are provided to simplify the analysis.
 Furthermore, we discuss the theory of GIST with a focus on practical aspects.
-We highlight potential pitfalls and show how to avoid technical difficulties.
-For this tutorial, familiarity with molecular dynamics simulations as well as the AmberTools package \cite{Case2023-ambertools} is assumed.
-
-%This particular document provides a skeleton illustrating key sections for a Tutorial document.
-%Please see the sample \texttt{sample-document.tex} in \url{github.com/livecomsjournal/article_templates/templates} for additional information on and examples of using the LiveCoMS LaTeX class.
-%Here we also assume familiarity with LaTeX and knowledge of how to include figures, tables, etc.; if you want examples, see the sample just referenced.
-%
-%In your work, in this particular slot, please provide an abstract of no more than 250 words.
-%Your abstract should explain the main contributions of your article, and should not contain any material that is not included in the main text.
-%Please note that your abstract, plus the authorship material following it, must not extend beyond the title page or modifications to the LaTeX class will likely be needed.
+We highlight potential pitfalls and provide strategies to avoid technical difficulties.
+This tutorial assumes familiarity with molecular dynamics simulations and the AmberTools package.
+
 \end{abstract}
@@
 
-%Tutorials should endeavor to cover the specific task at hand, and also highlight how the steps might need to be modified (or additional care might need to be taken at particular points) to handle more general cases.
-%
-%The scope of the tutorial, as well as the expected proficiencies / outcomes for researchers who complete the tutorial, should be clearly defined.
-%This will often happen in a specific section or subsection in the article itself.
-
 \section{Prerequisites}
@@
 
-\pagebreak %TODO: maybe remove this by better formatting
+\pagebreak %TODO: pagebreak, maybe remove these by better formatting
 
@@
 However, this should be done \emph{before} the MD simulation, since rotating the trajectory damages the periodic box information.
-%TODO: Ask Dan if this is still the case
 
@@
 \end{figure}
-%TODO: Hide capping group
+
 \pagebreak %TODO pagebreak
@@
 \subsubsection{PME energy}
-In the original version of GIST, the energies are calculated based on the the minimum image convention.
+In the original version of GIST, the energies are calculated based on the minimum image convention.
 In PME-GIST, the electrostatic energy $E_\text{elec}$ is calculated using the particle mesh Ewald (PME) method, which yields energies that are highly consistent with the Amber MD engine \cite{Chen2021}.

### 2024-07-23 manuscript/manuscript.tex
Add FWF Funding Information
small changes, mostly to grammar, but some effect the meaning of sentences

@@
 Furthermore, an implementation based on the particle mesh Ewald (PME) \cite{Darden1993-pme} method improves both the efficiency and the agreement with popular molecular dynamics engines \cite{Chen2021}.
-Yet further changes allow the inclusion of other rigid solvents than water \cite{Kraml2020,Waibl2022-gist-solvents} or the use of salt-water mixtures as solvents \cite{Waibl2021-gist-salt}.
+Other extensions of GIST allow the inclusion of rigid solvents other than water \cite{Kraml2020,Waibl2022-gist-solvents} or the use of salt-water mixtures as solvents \cite{Waibl2021-gist-salt}.
 
@@
 Since the density is computed from an explicit solvent MD simulation, GIST provides an accurate picture of solvation at an atomistic level, even for demanding systems such as drug binding sites in proteins.
-While the need for MD simulations comes at a large computational cost, GIST is still more efficient than end-state approaches such as TI \cite{Kirkwood1935-ti} or FEP \cite{Zwanzig1954-reweighting}, with applications to large systems such as serine proteases \cite{Kraml2019-gigist} or antibody fragments \cite{Waibl2021-gist-antibodies}.
+While all MD simulations require a large computational cost, GIST is still more efficient than end-state approaches such as TI \cite{Kirkwood1935-ti} or FEP \cite{Zwanzig1954-reweighting}, with applications to large systems such as serine proteases \cite{Kraml2019-gigist} or antibody fragments \cite{Waibl2021-gist-antibodies}.
 On the other hand, the first-order approximation means that the solvation entropy is less accurately described than in TI or FEP \cite{Chen2021,Waibl2021-gist-salt}. 
@@
 The tutorial is structured around two examples: a GIST calculation for the small molecule biotin, and a GIST study of biotin-streptavidin binding. 
-Whereas the biotin example gives an introduction to the basics of GIST calculations, the full biotin-streptavidin analysis showcases details and considerations commonly found in more advanced GIST studies.
+Biotin is used as a simple example to introduce the basics of GIST calculations, whereas the full biotin-streptavidin analysis showcases details and considerations commonly found in more advanced GIST studies.
 We show how to interpret the three-dimensional contributions of solvation free energy in a binding pocket and around the ligand.
@@
 
-Furthermore, a recent Python version (3.6 or newer) should be available with the following packages installed:
+Furthermore, a recent Python version (3.6 or newer) should be installed, along with the following packages:
 \begin{itemize}
@@
 We additionally provide a Jupyter notebook \cite{Kluyver2016-jupyter,Granger2021-jupyter} with the presented analyses. 
-For the visualization of three-dimensional contributions we require molecular visualization software that can handle \software{OpenDX} files, such as \software{PyMol} \cite{pymol} or \software{VMD} \cite{vmd}.
+The visualization of three-dimensional contributions to solvation free energy requires molecular visualization software that can handle \software{OpenDX} files, such as \software{PyMol} \cite{pymol} or \software{VMD} \cite{vmd}.
 Most of the visualisations and results in this tutorial can also be obtained using the \software{gistpp} program, which is available from \url{https://github.com/KurtzmanLab/Gist-Post-Processing} (code and documentation).
 \software{Gistpp} can directly manipulate \software{OpenDX} files from the GIST output.
-Its usage and some examples were shown in a previous tutorial on GIST \cite{Ramsey2016}.
+Its usage and some examples were presented in a previous tutorial on GIST \cite{Ramsey2016}.
 
@@
 Additionally, the GIST calculation should usually be preceded by centering the solute molecule in the unit cell and reimaging to account for periodic boundary conditions.
-Often, the region of interest will be defined as the surrounding of a molecule, for example a ligand in the binding pocket or a whole protein. 
+Often, the region of interest will be defined as the surrounding of a molecule, for example a ligand in a binding pocket, or an entire protein. 
 Using \inlinecode{cpptraj's } \inlinecode{bounds} command, boundaries can be output to a file. For example, here is the command for a boundary of \SI{10}{\angstrom} around the ligand LIG:
@@
 GIST is implemented in \software{cpptraj} which requires the topology and the trajectory of the MD simulation to be analyzed.
-The command needed to run GIST in cpptraj can be written in an input file (e.g gist.in) for convenience.
+The command needed to run GIST in cpptraj can be written in an input file (e.g gist.in) for convenience and reproducibility.
 The basic command, assuming a grid centered at coordinates (x=\SI{10.5}{\angstrom}, y=\SI{20.1}{\angstrom}, z=\SI{30.1}{\angstrom}) and a grid size of $30\times30\times30$ voxels with \SI{0.5}{\angstrom} spacing, is:

### 2024-07-23 manuscript/manuscript.tex
grammar

@@
 Since the density is computed from an explicit solvent MD simulation, GIST provides an accurate picture of solvation at an atomistic level, even for demanding systems such as drug binding sites in proteins.
-While all MD simulations require a large computational cost, GIST is still more efficient than end-state approaches such as TI \cite{Kirkwood1935-ti} or FEP \cite{Zwanzig1954-reweighting}, with applications to large systems such as serine proteases \cite{Kraml2019-gigist} or antibody fragments \cite{Waibl2021-gist-antibodies}.
+While all MD simulations require a large computational cost, GIST is more efficient than end-state approaches such as TI \cite{Kirkwood1935-ti} or FEP \cite{Zwanzig1954-reweighting}, with applications to large systems such as serine proteases \cite{Kraml2019-gigist} or antibody fragments \cite{Waibl2021-gist-antibodies}.
 On the other hand, the first-order approximation means that the solvation entropy is less accurately described than in TI or FEP \cite{Chen2021,Waibl2021-gist-salt}. 

### 2024-07-23 manuscript/manuscript.tex
grammar

@@
 The tutorial is structured around two examples: a GIST calculation for the small molecule biotin, and a GIST study of biotin-streptavidin binding. 
-Biotin is used as a simple example to introduce the basics of GIST calculations, whereas the full biotin-streptavidin analysis showcases details and considerations commonly found in more advanced GIST studies.
+Biotin is used as a simple example to introduce the basics of GIST calculations, whereas the full biotin-streptavidin analysis highlights details and considerations relevant to more advanced GIST studies.
 We show how to interpret the three-dimensional contributions of solvation free energy in a binding pocket and around the ligand.

### 2024-08-28 manuscript/manuscript.tex
Adress typos noted by reviewer 1

@@
 Furthermore, an implementation based on the particle mesh Ewald (PME) \cite{Darden1993-pme} method improves both the efficiency and the agreement with popular molecular dynamics engines \cite{Chen2021}.
-Other extensions of GIST allow the inclusion of rigid solvents other than water \cite{Kraml2020,Waibl2022-gist-solvents} or the use of salt-water mixtures as solvents \cite{Waibl2021-gist-salt}.
+Yet further changes allow the inclusion of other rigid solvents than water \cite{Kraml2020,Waibl2022-gist-solvents} or the use of salt-water mixtures as solvents \cite{Waibl2021-gist-salt}.
 
@@
 Since the density is computed from an explicit solvent MD simulation, GIST provides an accurate picture of solvation at an atomistic level, even for demanding systems such as drug binding sites in proteins.
-While all MD simulations require a large computational cost, GIST is more efficient than end-state approaches such as TI \cite{Kirkwood1935-ti} or FEP \cite{Zwanzig1954-reweighting}, with applications to large systems such as serine proteases \cite{Kraml2019-gigist} or antibody fragments \cite{Waibl2021-gist-antibodies}.
+While the need for MD simulations comes at a large computational cost, GIST is still more efficient than end-state approaches such as TI \cite{Kirkwood1935-ti} or FEP \cite{Zwanzig1954-reweighting}, with applications to large systems such as serine proteases \cite{Kraml2019-gigist} or antibody fragments \cite{Waibl2021-gist-antibodies}.
 On the other hand, the first-order approximation means that the solvation entropy is less accurately described than in TI or FEP \cite{Chen2021,Waibl2021-gist-salt}. 
@@
 The tutorial is structured around two examples: a GIST calculation for the small molecule biotin, and a GIST study of biotin-streptavidin binding. 
-Biotin is used as a simple example to introduce the basics of GIST calculations, whereas the full biotin-streptavidin analysis highlights details and considerations relevant to more advanced GIST studies.
+Whereas the biotin example gives an introduction to the basics of GIST calculations, the full biotin-streptavidin analysis showcases details and considerations commonly found in more advanced GIST studies.
 We show how to interpret the three-dimensional contributions of solvation free energy in a binding pocket and around the ligand.
@@
 
-Furthermore, a recent Python version (3.6 or newer) should be installed, along with the following packages:
+Furthermore, a recent Python version (3.6 or newer) should be available with the following packages installed:
 \begin{itemize}
@@
 We additionally provide a Jupyter notebook \cite{Kluyver2016-jupyter,Granger2021-jupyter} with the presented analyses. 
-The visualization of three-dimensional contributions to solvation free energy requires molecular visualization software that can handle \software{OpenDX} files, such as \software{PyMol} \cite{pymol} or \software{VMD} \cite{vmd}.
+For the visualization of three-dimensional contributions we require molecular visualization software that can handle \software{OpenDX} files, such as \software{PyMol} \cite{pymol} or \software{VMD} \cite{vmd}.
 Most of the visualisations and results in this tutorial can also be obtained using the \software{gistpp} program, which is available from \url{https://github.com/KurtzmanLab/Gist-Post-Processing} (code and documentation).
 \software{Gistpp} can directly manipulate \software{OpenDX} files from the GIST output.
-Its usage and some examples were presented in a previous tutorial on GIST \cite{Ramsey2016}.
+Its usage and some examples were shown in a previous tutorial on GIST \cite{Ramsey2016}.
 
@@
 Additionally, the GIST calculation should usually be preceded by centering the solute molecule in the unit cell and reimaging to account for periodic boundary conditions.
-Often, the region of interest will be defined as the surrounding of a molecule, for example a ligand in a binding pocket, or an entire protein. 
+Often, the region of interest will be defined as the surrounding of a molecule, for example a ligand in the binding pocket or a whole protein. 
 Using \inlinecode{cpptraj's } \inlinecode{bounds} command, boundaries can be output to a file. For example, here is the command for a boundary of \SI{10}{\angstrom} around the ligand LIG:
@@
 GIST is implemented in \software{cpptraj} which requires the topology and the trajectory of the MD simulation to be analyzed.
-The command needed to run GIST in cpptraj can be written in an input file (e.g gist.in) for convenience and reproducibility.
+The command needed to run GIST in cpptraj can be written in an input file (e.g gist.in) for convenience.
 The basic command, assuming a grid centered at coordinates (x=\SI{10.5}{\angstrom}, y=\SI{20.1}{\angstrom}, z=\SI{30.1}{\angstrom}) and a grid size of $30\times30\times30$ voxels with \SI{0.5}{\angstrom} spacing, is:
@@
 In the third line, we reimage the coordinates to the first box and center the biotin at the origin. 
-Then, we strip all water solvent molecules in the fourth line and write out a pdb file in the fith line. 
+Then, we strip all water solvent molecules in the fourth line and write out a pdb file in the fifth line. 
 Finally, we start the analysis with the "run" command.
@@
 \end{lstlisting}
-Adapt the above input file for biotin by striping molecule 1 instead of 2 and replacing the path names in the \inlinecode{strip} and  \inlinecode{trajout} commands.
+Adapt the above input file for biotin by stripping the molecule 1 instead of 2 and replacing the path names in the \inlinecode{strip} and  \inlinecode{trajout} commands.
 Alternatively, we provide a small script named \software{cpptraj\_remove\_mol.sh} for this procedure.
@@
 	\caption{Free energy contributions for the monomers and the dimer in kcal$\cdot$mol$^{-1}$. ``Diff'' is calculated as 
-	$E^\text{internal} + \Delta E^\text{GIST} - \Delta S^\text{scaled}$. ``total'' is calculated as 
+	$E^\text{internal} + \Delta E^\text{GIST} - T\Delta S^\text{scaled}$. ``total'' is calculated as 
 	``complex''-``streptavidin'' - ``biotin''.}\label{tab_dg_monomers_dimer}
@@
 	the positive energy, where they are divided by two to visually account for its overall lower density.
-The negative solvent energgy is shown in green, positive solvent energy in red and negative entropy in blue.
+The negative solvent energy is shown in green, positive solvent energy in red and negative entropy in blue.
 Only voxels within \SI{4}{\angstrom} of biotin are shown.}
@@
 Specific regions in a protein can be defined to calculate the thermodynamic quantities for that region.
-This is significant specially when calculating the properties of water that will be expelled when a ligand binds to the protein.\\
+This is significant especially when calculating the properties of water that will be expelled when a ligand binds to the protein.\\
 For example, voxels that are located within the binding pocket can be defined using a ligand solute molecule and a distance criteria. 
@@
 %%%%%%%
-
+This research was funded in whole or in part by the Austrian Science Fund (FWF) [10.55776/P34518].
 \section*{Author Information}

### 2024-08-28 manuscript/manuscript.tex
Moves cpptraj references/intro to start of software section (R2.3)
Adds explicit section labels to the 'Tutorial Data' section (R2.4)

@@
 A recent version of \software{cpptraj} should be used. We recommend using at least version 6.24, which was released with AmberTools24 \cite{Case2023-ambertools}. 
+An overview of how to work with \software{cpptraj} is provided in the program's documentation \cite{cpptraj_doc}, AMBER tutorials \cite{amber_tut_cpptraj} and on the AMBER-hub website \cite{amber_hub}.
 
@@
 This command tells \software{cpptraj} to run a GIST calculation using the specified region and print the output in a file called \inlinecode{gist.dat}.
-An overview of how to work with \software{cpptraj} is provided in the program's documentation \cite{cpptraj_doc}, AMBER tutorials \cite{amber_tut_cpptraj} and on the AMBER-hub website \cite{amber_hub}.
 
@@
 We will use the crystal structure of streptavidin (PDB code: 1STP \cite{Weber1989-streptavidin-structure}) to start our calculations.
-For the beginners section, we will use the biotin ligand extracted from the crystal structure.
-For the second part, we will use the whole crystal structure.
+For section \ref{subsec:gist_biotin}, we will use the biotin ligand extracted from the crystal structure.
+For the second part in section \ref{subsec:gist_complex}, we will use the whole crystal structure.
 For the sake of simplicity, you can download prepared and solvated Amber topologies and structures based on 1STP by running
@@
 \subsection{Running GIST for Biotin}
+\label{subsec:gist_biotin}
 In this example, we will be running a simple GIST calculation of the biotin solute molecule to get familiar with the usage of GIST in \software{cpptraj}.
@@
 \subsection{Applying GIST for the Streptavidin-Biotin complex}
+\label{subsec:gist_complex}
 In this section we will apply GIST to the full complex and discuss each step of the process in more detail to outline potential pitfalls.

### 2024-08-28 manuscript/manuscript.tex
Fix variable naming (Eww instead of Esw) error (R2.15)

@@
 For each solvent molecule in this list, densities (e.g., \inlinecode{g\_mol\_NA}) and energies (e.g., \inlinecode{Eww\_mol\_NA}) will be computed and written to .dx files.
-Note that, e.g., \inlinecode{Esw\_mol\_NA} contains the interactions of every ``\inlinecode{NA}'' solvent molecule with the solute molecule, and \inlinecode{Esw\_mol\_NA} contains the interactions among ``\inlinecode{NA}'' solvent molecules as well as their interactions with all other solvents, divided by 2 to account for double counting.
+Note that, e.g., \inlinecode{Esw\_mol\_NA} contains the interactions of every ``\inlinecode{NA}'' solvent molecule with the solute molecule, and \inlinecode{Eww\_mol\_NA} contains the interactions among ``\inlinecode{NA}'' solvent molecules as well as their interactions with all other solvents, divided by 2 to account for double counting.
 The entropy will be computed using only the density of the first solvent molecule. \\

### 2025-04-13 manuscript/manuscript.tex

--- 2024-08-28 manuscript/manuscript.tex
@@
 Since the density is computed from an explicit solvent MD simulation, GIST provides an accurate picture of solvation at an atomistic level, even for demanding systems such as drug binding sites in proteins.
-While the need for MD simulations comes at a large computational cost, GIST is still more efficient than end-state approaches such as TI \cite{Kirkwood1935-ti} or FEP \cite{Zwanzig1954-reweighting}, with applications to large systems such as serine proteases \cite{Kraml2019-gigist} or antibody fragments \cite{Waibl2021-gist-antibodies}.
+While all MD simulations require a large computational cost, GIST is more efficient than end-state approaches such as TI \cite{Kirkwood1935-ti} or FEP \cite{Zwanzig1954-reweighting}, with applications to large systems such as serine proteases \cite{Kraml2019-gigist} or antibody fragments \cite{Waibl2021-gist-antibodies}.
 On the other hand, the first-order approximation means that the solvation entropy is less accurately described than in TI or FEP \cite{Chen2021,Waibl2021-gist-salt}. 
@@
 The tutorial is structured around two examples: a GIST calculation for the small molecule biotin, and a GIST study of biotin-streptavidin binding. 
-Whereas the biotin example gives an introduction to the basics of GIST calculations, the full biotin-streptavidin analysis showcases details and considerations commonly found in more advanced GIST studies.
+Biotin is used as a simple example to introduce the basics of GIST calculations, whereas the full biotin-streptavidin analysis highlights details and considerations relevant to more advanced GIST studies.
 We show how to interpret the three-dimensional contributions of solvation free energy in a binding pocket and around the ligand.
@@
 In the third line, we reimage the coordinates to the first box and center the biotin at the origin. 
-Then, we strip all water solvent molecules in the fourth line and write out a pdb file in the fifth line. 
+Then, we strip all water solvent molecules in the fourth line and write out a pdb file in the fith line. 
 Finally, we start the analysis with the "run" command.
@@
 \end{lstlisting}
-Adapt the above input file for biotin by stripping the molecule 1 instead of 2 and replacing the path names in the \inlinecode{strip} and  \inlinecode{trajout} commands.
+Adapt the above input file for biotin by striping molecule 1 instead of 2 and replacing the path names in the \inlinecode{strip} and  \inlinecode{trajout} commands.
 Alternatively, we provide a small script named \software{cpptraj\_remove\_mol.sh} for this procedure.
@@
 	\caption{Free energy contributions for the monomers and the dimer in kcal$\cdot$mol$^{-1}$. ``Diff'' is calculated as 
-	$E^\text{internal} + \Delta E^\text{GIST} - T\Delta S^\text{scaled}$. ``total'' is calculated as 
+	$E^\text{internal} + \Delta E^\text{GIST} - \Delta S^\text{scaled}$. ``total'' is calculated as 
 	``complex''-``streptavidin'' - ``biotin''.}\label{tab_dg_monomers_dimer}
@@
 	the positive energy, where they are divided by two to visually account for its overall lower density.
-The negative solvent energy is shown in green, positive solvent energy in red and negative entropy in blue.
+The negative solvent energgy is shown in green, positive solvent energy in red and negative entropy in blue.
 Only voxels within \SI{4}{\angstrom} of biotin are shown.}
@@
 Specific regions in a protein can be defined to calculate the thermodynamic quantities for that region.
-This is significant especially when calculating the properties of water that will be expelled when a ligand binds to the protein.\\
+This is significant specially when calculating the properties of water that will be expelled when a ligand binds to the protein.\\
 For example, voxels that are located within the binding pocket can be defined using a ligand solute molecule and a distance criteria. 

### 2025-05-15 manuscript/manuscript.tex
Various fixes
1) Fix SI package warning by adding \bar explicitly
2) Clarify intro sentence on link between free energy of solvation and hydrophobicity
3) Clarify use of restraints

--- 2024-08-28 manuscript/manuscript.tex
@@
 \usepackage{placeins}
+\DeclareSIUnit{\bar}{bar}
 \DeclareSIUnit\Molar{M}
@@
 The free energy of solvation is defined as the reversible work needed to insert a solute molecule into the solvent from gas phase \cite{ben-naim-book}.
-It is directly linked to the solubility of gases and drives many real-world effects such as hydrophobicity, ligand binding, and protein folding.
+It is directly linked to the solubility of solutes and drives many real-world effects. 
+For example, solvation thermodynamics govern aqueous solubility and hydrophobic effects, which in turn influence phenomena like partitioning between environments, ligand binding, and protein folding.
 
@@
 Now, you can run a restrained MD simulation based on the equilibrated structure of biotin for the subsequent GIST analysis.
-Restraints should be used during the MD because IST theory is not well-defined for moving solute molecules.
-We apply a harmonic restraint of \SI{100}{\kilo\calorie\per\mole\per\angstrom\squared} to all heavy atoms to keep the system in place.
+Restraints should be used during the MD because IST theory is defined for rigid solute molecules.
+In practice, this means that the solute molecule should not move significantly during the MD simulation.
+This is tipically achieved with harmonic positional restraints applied to the heavy atoms of the solute, tethering them to their initial or reference coordinates (e.g., from a crystal structure or an equilibrated structure).
+In AMBER,  the harmonic restraint potential takes the form $U = k_r * (r - r_0)^2$ for each cartesian coordinate of each restrained atom, where $r_0$ is its reference position. 
+The force constant $k_r$ should be chosen strong enough to prevent significant deviation.
+Here, we apply a harmonic restraint of \SI{100}{\kilo\calorie\per\mole\per\angstrom\squared} to all heavy atoms to keep the system in place. 
 Note that this is a rather strong restraint. 
-While lower values are possible, make sure that the solute molecule is kept rigid during the simulation.
+While weaker restraints are possible, make sure that the solute molecule is kept rigid during the simulation. 
 As a minimum value, we recommend \SI{10}{\kilo\calorie\per\mole\per\angstrom\squared} restraints.
+Similarly, the solute could also be kept rigid using positional constraints, RMSD restraints, or other methods.
+However, depending on the method, additional post-processing may be required to keep the solute at the same position in the grid.
+To make sure that the solute is not moving significantly, we recommend to check the root mean square deviation (RMSD) of the solute heavy atoms during the simulation.
+The RMSD should be less than \SI{0.5}{\angstrom} for the solute heavy atoms, preferably less than \SI{0.1}{\angstrom}.
 We run \SI{100}{\nano\second} of simulation and store the conformations every \SI{100}{\pico\second}.

### 2025-05-15 manuscript/manuscript.tex
More revisions
1) Add recommendations for the grid size and boundaries
2) Fix some sentences on the same line
3) Fix some typos
4) Add subsection on how to deal with ensembles

@@
 Often, the region of interest will be defined as the surrounding of a molecule, for example a ligand in the binding pocket or a whole protein. 
-Using \inlinecode{cpptraj's } \inlinecode{bounds} command, boundaries can be output to a file. For example, here is the command for a boundary of \SI{10}{\angstrom} around the ligand LIG:
+We recommend to use grids with a similiar size to the simulation box, as the GIST calculation will usually not be the time limiting step and a region of interest can then be chosen freely.
+However, if computation resources and disk space are limited, it might be beneficial to reduce the grid size.
+The minimum recommended grid size depends on the solvent and its interactions with the solute.
+Generally, a grid encompassing 2 to 3 solvent layers around the solute is sufficient. 
+For TIP3P water, this corresponds to a buffer zone of around \SI{10}{\angstrom} around the solute. 
+However, larger grids might be necessary for example for strongly charged solutes.
+
+Using \inlinecode{cpptraj's } \inlinecode{bounds} command, boundaries can be output to a file. 
+For example, here is the command for a boundary of \SI{10}{\angstrom} around the ligand LIG:
 
@@
 In practice, this means that the solute molecule should not move significantly during the MD simulation.
-This is tipically achieved with harmonic positional restraints applied to the heavy atoms of the solute, tethering them to their initial or reference coordinates (e.g., from a crystal structure or an equilibrated structure).
+This is typically achieved with harmonic positional restraints applied to the heavy atoms of the solute, tethering them to their initial or reference coordinates (e.g., from a crystal structure or an equilibrated structure).
 In AMBER,  the harmonic restraint potential takes the form $U = k_r * (r - r_0)^2$ for each cartesian coordinate of each restrained atom, where $r_0$ is its reference position. 
@@
 We will afterwards introduce advanced analysis workflows which allow you to modify and better visualize the resulting data.
-\subsubsection{System Preparation and Equilibration}
+\subsubsection{System Preparation}
 If you want to skip this section, run 
@@
 The choice of initial structure is crucial to the GIST results.
-Crystal structures may not represent the dominant solution structure for a given target.
-In our example, we only use a single equilibrated structure.
-A more sophisticated approach would be to identify several dominant structures in the ensemble and run a separate MD simulation and GIST calculation for each structure. 
-Such dominant structures could either be identified from experimental ensembles or by clustering an unrestrained MD simulation. 
 When starting from an experimental structure, care has to be taken during the preparation step.
@@
 
+\subsubsection{Dealing with Conformational Heterogeneity}
+Crystal structures may not represent the dominant solution structure for a given target.
+In our example, we only use a single equilibrated structure.
+However, in practice the target may be flexible and a single structure may not be representative of the conformational ensemble.
+A more sophisticated approach would identify several dominant structures in the ensemble and run a separate MD simulation and GIST calculation for each structure. 
+Such dominant structures could either be identified from experimental ensembles or by clustering an unrestrained MD simulation. 
+When the probabilities of the different conformations are known, one can then use a weighted average of the GIST results to obtain a free energy of solvation for the ensemble \cite{Waibl2021-gist-antibodies}:
+\begin{equation}
+	\Delta G_{solv} = k_BT \cdot ln \int p(q_u) \cdot e^{ \frac{-\Delta G_solv(q_u)}{k_BT}} dq_u
+\end{equation}
+While averaging over multiple conformations has been shown to improve agreement with experimental observables \cite{Kamenik2020-gist-macrocycles,Waibl2021-gist-antibodies}, there are some caveats to this approach.
+First, running GIST analyses for each conformation is computationally expensive and time-consuming, since each conformation needs their own MD simulation.
+In cases where the conformational ensemble is large, it may be difficult to sample all relevant conformations. 
+If accurate solvation thermodynamics are desired for such cases, other methods which inherently include the solute confromational ensemble may be more appropriate.
+Second, this approach suffers from the curse of exponential averaging, where the free energy of solvation is dominated by very few conformations due to the exponential reweighting \cite{Ekberg2021-exponential}.
+For cases where the number of sampled conformations is small and the samples are badly chosen, this can lead to large errors in the free energy of solvation.
+Often a simple weighted average is then a better estimator:
+\begin{equation}
+	\Delta G_{solv} = \frac{\sum_i p(q_u) \cdot \Delta G_solv(q_u)}{\sum_i p(q_u)}
+\end{equation}
+While dealing with conformational heterogeneity is therefore definitely possible with GIST, it is not trivial and remains challenging.
+However, GIST is at its best when the solute conformational ensemble is small and well-defined, such as in the case of small molecules or rigid protein-ligand complexes.
+It remains unmatched when one is interested in the local solvation properties of a single dominant conformation, e.g. to elucidate the solvation thermodynamics of crystal waters in a x-ray structure.
+
+Concerning the end-state approach shown in this tutorial, we of course don't need to only worry about the conformational heterogeneity of a single solute molecule, but about multiple similar conformational ensembles.
 When calculating solvation free energy differences between a complex and the corresponding monomers (the dissociated state), there are generally two approaches for dealing with configurational diversity.
@@
 Calculations for the disassociated states are then run on the extracted structures from the complex configurations.
-One can choose to sample the complex and dissociated states individually, or to sample only one of them and assume that there is sufficient overlap in the conformations \cite{Genheden-EndState2012,Wang-MMPBSA2019}.
+One can choose to sample the ensembles of the complex and dissociated states individually, or to sample only one of them and assume that there is sufficient overlap in the conformations \cite{Genheden-EndState2012,Wang-MMPBSA2019}.
 
 Here, we do not include any sampling of the conformational states.
-It is therefore important that the molecular conformations in the complex and dissociated states match exactly.
+It is therefore important that the molecular conformations as well as the GIST grids in the complex and dissociated states match exactly.
+\subsubsection{Equilibration}
 Use the \software{equilibration.py} script (or your own procedure) to perform short NVT and NPT equilibration runs on the complex structure.

### 2025-05-15 manuscript/manuscript.tex
Add reference val equations

@@
 Often, the region of interest will be defined as the surrounding of a molecule, for example a ligand in the binding pocket or a whole protein. 
-We recommend to use grids with a similiar size to the simulation box, as the GIST calculation will usually not be the time limiting step and a region of interest can then be chosen freely.
+We recommend using GIST grids with a similar size to the simulation box, as the GIST calculation will usually not be the time limiting step and a region of interest can then be chosen freely in post-analysis.
 However, if computation resources and disk space are limited, it might be beneficial to reduce the grid size.
@@
 For TIP3P water, this corresponds to a buffer zone of around \SI{10}{\angstrom} around the solute. 
-However, larger grids might be necessary for example for strongly charged solutes.
+Larger grids might be necessary, for example for strongly charged solutes.
 
@@
 Generally, it is advisable to conduct your own reference calculations, especially for systems that are either very large or very small. 
-The reference values can be extracted from MD trajectories of bulk solvent by calculating the density-weighted average energy per solvent molecule ($E_{ww, norm}$) and the average solvent density $\rho_0$.
+The reference values can be extracted from unrestrained MD trajectories of pure bulk solvent by calculating the average potential energy per solvent molecule ($E_{ww, norm}$) and the average solvent density $\rho_0$:
+\begin{equation}
+	E_{ww, norm} = \frac{\bar{E}_{pot,\ total}}{N_{WAT}}
+\end{equation}	
+\begin{equation}
+	\rho_0 = \frac{N_{WAT}}{\bar{V}_{box}}
+\end{equation}
 It is recommended that the size of the reference solvent box closely matches that of the production system.

### 2025-05-16 manuscript/manuscript.tex
More revisions, again
1) Redid Figure 4, added one version with and without entropy referencing
2) Rewrote the paragraphs referencing Fig4, as the implications changed

@@
 If the curve flattens out, the converged value is your final $\dgsolv$\@.
-If the curve diverges even for large integration distances you might need to tweak the $E_{ww}$ reference, or introduce a reference value for the entropy.
-
+If the curve diverges even for large integration distances, it depends on which of the contributing parts diverges. 
 \begin{lstlisting}[style=python]
@@
 	gist = load_gist_file('gist.dat',
-    struct='solute-centered.pdb', eww_ref=-9.533)
+		struct='solute-centered.pdb', eww_ref=-9.5398)
 	bins, (dg, esw, eww, s) = gist.rdf(
@@
 		bins=100, rmax=20., normalize='none')
-plt.plot(bins, np.cumsum(eww), label='Eww')
-plt.plot(bins, np.cumsum(esw), label='Esw')
-plt.plot(bins, np.cumsum(dg), label='dG')
-plt.plot(bins, np.cumsum(s), label='dS')
+	plt.plot(bins, np.cumsum(eww), label='$\Delta$E$_{ww}$')
+	plt.plot(bins, np.cumsum(esw), label='$\Delta$E$_{sw}$')
+	plt.plot(bins, np.cumsum(dg), label='$\Delta$G')
+	plt.plot(bins, np.cumsum(s), label='T$\Delta$S')
 	plt.legend()
-plt.xlabel("distance cutoff []")
+	plt.xlabel("distance cutoff [$\AA$]")
 	plt.ylabel("free energy contribution [kcal/mol]")
@@
 The expected output of this code is shown in Figure~\ref{fig_radial_convergence}.
-Especially the solvent-solvent energy is not perfectly flat, indicating that the reference value is not optimal.
-%but is also complicated because the solute is charged.
-Both the entropy and $E_{sw}$ can be considered converged around \SI{10}{\angstrom}.
-While this is not true for $E_{ww}$, we will see below that the energy difference upon binding converges more readily than its individual contributions.
-
 \begin{figure}
 	\centering
-	\includegraphics[width=0.8\linewidth]{figures/A_E_S_convergence.eps}
+	\includegraphics[width=0.8\linewidth]{figures/biotin-contributions_noref.png}
 	\caption{Convergence of $\dgsolv$ and its contributions with increasing distance to the solute molecule.
@@
 Computed from the biotin calculation.}\label{fig_radial_convergence}
+\end{figure}
+Here we see two effects: First, that the solute-solvent energy $E_{sw}$ converges only slowly, due to the charged nature of biotin.
+Second, that the solvent entropy diverges, even at large distances.
+Most of the time, this happens due to a bias in the Nearest-Neighbor entropy calculation\cite{Chen2021,Huggins2014-entropy}.
+This can be fixed by either calculating the bias per water molecule (as was done in \cite{Chen2021}) or referencing the entropy in the same manner as the $E_{ww, norm}$:
+\begin{lstlisting}[style=python]
+	def reference_entropy(gf):
+    if 'dTSsix_unref_norm' not in gf.data.columns:
+        gf['dTSsix_unref_norm'] = gf['dTSsix_norm']
+        gf['dTSsix_unref_dens'] = gf['dTSsix_dens']
+    refval = gf.detect_reference_value('dTSsix_unref_dens')
+    gf['dTSsix_norm'] = gf.get_referenced('dTSsix_unref_norm', refval)
+    gf['dTSsix_dens'] = gf.get_referenced('dTSsix_unref_dens', refval)
+
+reference_entropy(biotin)
+\end{lstlisting}
+When we recreate the plot from Figure~\ref{fig_radial_convergence} with the referenced entropy in Figure~\ref{fig_radial_convergence_ref}, we see that the divergence is removed.
+
+While $E_{ww}$ converges nicely here when using the tabulated $E_{ww, norm}$, if it diverges you might need to tweak the $E_{ww}$ reference by either using the automatic referencing, using a different tabulated $E_{ww, norm}$ value or running a pure bulk solvent reference calculation.
+\begin{figure}
+	\centering
+	\includegraphics[width=0.8\linewidth]{figures/biotin-contributions_ref.png}
+	\caption{Convergence of $\dgsolv$ and its contributions with increasing distance to the solute molecule.
+The entropy was referenced, so it converges to zero for voxels far away from the solute.}\label{fig_radial_convergence_ref}
 \end{figure}

### 2025-05-16 manuscript/manuscript.tex
More Clarifications
1) Clarified the use of gridcntr in section 3
2) Clarified common error of Eww being wrong

@@
 \end{lstlisting}
-Additionally, the GIST calculation should usually be preceded by centering the solute molecule in the unit cell and reimaging to account for periodic boundary conditions.
+If the solute is already centered at the origin (0, 0, 0), the grid center does not need to be specified.
+As such, the GIST calculation should usually be preceded by centering the solute molecule in the unit cell and reimaging to account for periodic boundary conditions.
 Often, the region of interest will be defined as the surrounding of a molecule, for example a ligand in the binding pocket or a whole protein. 
@@
 
-While $E_{ww}$ converges nicely here when using the tabulated $E_{ww, norm}$, if it diverges you might need to tweak the $E_{ww}$ reference by either using the automatic referencing, using a different tabulated $E_{ww, norm}$ value or running a pure bulk solvent reference calculation.
+While $E_{ww}$ converges nicely here when using the tabulated $E_{ww, norm}$, if it diverges one might need to tweak the $E_{ww}$ reference by either using the automatic referencing, using a different tabulated $E_{ww, norm}$ value or running a pure bulk solvent reference calculation.
+The most common error for strong divergence of $E_{ww}$ is that the reference value is set to a value of a different solvent model.
+
 \begin{figure}

### 2025-05-16 manuscript/manuscript.tex
Added signposting
1) Direct  links to cpptraj, amber manual
2) More explanation for bound command

@@
 Its usage and some examples were shown in a previous tutorial on GIST \cite{Ramsey2016}.
-
+For help with the listed cpptraj commands, AMBER atom mask syntax and other AMBER-specific commands, we recommend the following webresources:\\
+\begin{itemize}
+\item The website https://amberhub.chpc.utah.edu/ which is managed by the Cheatham III group, 
+\item The cpptraj manual at https://raw.githubusercontent.com/Amber-MD/cpptraj/master/doc/CpptrajManual.pdf 
+\item and the AMBER manual \cite{amber24} at https://ambermd.org/Manuals.php.
+\end{itemize}
 \pagebreak %TODO: pagebreak, maybe remove these by better formatting
@@
 Using \inlinecode{cpptraj's } \inlinecode{bounds} command, boundaries can be output to a file. 
-For example, here is the command for a boundary of \SI{10}{\angstrom} around the ligand LIG:
-
+For example, here is the command for a boundary of \SI{15}{\angstrom} around the ligand LIG:
 \begin{lstlisting}[style=cpptraj]
-bounds :LIG dx 0.5 offset 20 name Grid out bounds.dat
-\end{lstlisting}
-\begin{minipage}{\linewidth} %TODO: maybe find a better way to keep listing on same page?
+bounds :LIG dx 0.5 offset 30 name Grid out bounds.dat
+\end{lstlisting}
+The atom mask \inlinecode{:LIG} specifies the residue named LIG to build the boundaries around, with dx and offset providing the actual dimensions of the boundaries, i.e. an offset of 30 times 0.5\AA.
+\begin{minipage}{\linewidth} 
 Make sure that the \inlinecode{bounds} command is supplied with the exact same trajectory as in the GIST calculation. 
@@
 This will download the tutorial files, including both the manuscript and the examples in the \inlinecode{code} folder.
-
+Additionally, a link to download the original MD trajectory files is provided in the GitHub repository.
 \subsection{Running GIST for Biotin}
@@
 autoimage !(:WAT) origin
-bounds !(:WAT) dx 0.5 offset 30 name MyGrid out bounds.dat
+bounds !(:WAT) dx 0.5 offset 30 name Grid out bounds.dat
 run
@@
 In the above snippet, we calculate the bounds around the whole protein for a grid with \SI{0.5}{\angstrom} side-length voxels and a buffer zone of \SI{15.0}{\angstrom}. 
+The atom mask !(:WAT) specifies that the grid should be calculated around all atoms that are not water residues.
 Note that the offset is given in terms of number of additional voxels, not length dimension. 

### 2025-05-18 manuscript/manuscript.tex
Add MD repository links

@@
 If the reader prefers to skip the MD and GIST calculation, we also provide the GIST output files with this tutorial, such that the post-processing can be done without any expensive calculations.
+Additionally, the MD simulation files and raw GIST output files can be found at \url{https://doi.org/10.48323/4mbrd-67m83}, to reproduce the results shown in this tutorial.
 We additionally provide a Jupyter notebook \cite{Kluyver2016-jupyter,Granger2021-jupyter} with the presented analyses. 
@@
 This will download the tutorial files, including both the manuscript and the examples in the \inlinecode{code} folder.
-Additionally, a link to download the original MD trajectory files is provided in the GitHub repository.
+Additionally, the original MD trajectory files and raw GIST output files are provided under the DOI \url{https://doi.org/10.48323/4mbrd-67m83}.
 \subsection{Running GIST for Biotin}

### 2025-05-19 manuscript/manuscript.tex
Add more solvation methods to intro
Add mentions to SZMAP, SPAM and the cell theories to intro.

@@
 A schematic of this is shown in Figure~\ref{fig:carbazole}.
+The enthalpic contributions are computed from the intermolecular interactions of the solvent with the solute and as well as other solvent. 
+As they are derived from the force field, pairwise additivity of the underlying intermolecular interactions is assumed.
 The entropy is computed from orientational and translational distributions of the solvent around the solute molecule, in a first-order approximation. 
 The calculation of higher-order entropy terms is computationally demanding but was realized in recent works \cite{Nguyen2016-gist-second-order, Waibl2021-gist-salt}.
-It has been found that solvation free energies of small molecule solutes can be computed at high accuracy using linear scaling factors for the higher-order terms, such as $\Delta S = 0.6\, \Delta S^\text{1st-order}$ for water \cite{Chen2021,Waibl2022-gist-solvents}.
+However, it has been found that solvation free energies of small molecule solutes can be computed at high accuracy using linear scaling factors for the higher-order terms, such as $\Delta S = 0.6\, \Delta S^\text{1st-order}$ for water \cite{Chen2021,Waibl2022-gist-solvents}.
 Since its initial development, GIST has been used to describe the thermodynamics of water displacement in protein binding pockets \cite{Nguyen2016-gist-second-order,Ramsey2016,Balius2017-gist-ligand-discovery,Hufner-Wulsdorf2020-gist-drug-design}.
@@
 Methods derived from statistical mechanics bridge this gap by providing three-dimensional resolution. 
+It is important to note that while the total solvation free energy is a well-defined thermodynamic quantity, its spatial decomposition is not unique and depends on the theoretical framework\cite{Persson02092022}.
 These methods most commonly are based on either the Ornstein–Zernike (OZ) equation, or \cite{Hansen2013-simple-liquids} inhomogeneous fluid solvation theory (IST) \cite{Lazaridis1998}.
@@
 Similar to the self-consistent integral theories, GIST only considers a single solute conformation and multiple computations of representative structures are necessary to describe ensemble properties.
+
+Aside from the previously mentioned methods, there is a large variety of other methods to compute enthalpy and entropy contributions to the solvation free energy, as recently reviewed \cite{Heyden2019-localsolvation,Mukherjee2022-ReviewSpatialWatersBiomolecule}.
+Approaches like SZMAP \cite{Bayden2015-SZMAP} trade the accuracy of all-atom MD simulation for speed by using a faster probe-based approach.
+SPAM ("maps" in reverse) is more akin to WaterMaps, investigating distinct hydration sites based on high local water densities \cite{Cui2013-SPAM}. 
+However, it approximates the water entropies from the water interaction enthalpy distributions.
+Another group of methods builds on the cell theory for liquids \cite{henchman2003partition}, which derives independent effective potentials for each molecule in its' neighbors mean field.
+Conceptually close to GIST is Grid Cell Theory (GCT), which also adopts a grid-based approach but differs in the assumptions made its' entropy calculations \cite{Gerogiokas2014-GCT}.
+More recently, Multiscale Cell Correlation (MCC) has been developed, generalizing and pushing the method from simple liquids up to large biomolecules \cite{henchman2007free, Hensen2014, Ali2021-hc, Kalayan2022-MCCBiomol}.
 

### 2025-05-19 manuscript/manuscript.tex
Small changes to the requirement section

@@
 A recent version of \software{cpptraj} should be used. We recommend using at least version 6.24, which was released with AmberTools24 \cite{Case2023-ambertools}. 
-An overview of how to work with \software{cpptraj} is provided in the program's documentation \cite{cpptraj_doc}, AMBER tutorials \cite{amber_tut_cpptraj} and on the AMBER-hub website \cite{amber_hub}.
+An overview of how to work with \software{cpptraj} is provided in the program's documentation \cite{cpptraj_doc}, AMBER manual \cite{amber24}, AMBER tutorials \cite{amber_tut_cpptraj} and on the AMBER-hub website \cite{amber_hub}.
 
@@
 \end{itemize}
-If the reader prefers to skip the MD and GIST calculation, we also provide the GIST output files with this tutorial, such that the post-processing can be done without any expensive calculations.
+If the reader prefers to skip the MD and GIST calculation, GIST output files are provided with this tutorial, such that the post-processing can be done without any expensive calculations.
 Additionally, the MD simulation files and raw GIST output files can be found at \url{https://doi.org/10.48323/4mbrd-67m83}, to reproduce the results shown in this tutorial.
@@
 Its usage and some examples were shown in a previous tutorial on GIST \cite{Ramsey2016}.
-For help with the listed cpptraj commands, AMBER atom mask syntax and other AMBER-specific commands, we recommend the following webresources:\\
-\begin{itemize}
-\item The website https://amberhub.chpc.utah.edu/ which is managed by the Cheatham III group, 
-\item The cpptraj manual at https://raw.githubusercontent.com/Amber-MD/cpptraj/master/doc/CpptrajManual.pdf 
-\item and the AMBER manual \cite{amber24} at https://ambermd.org/Manuals.php.
-\end{itemize}
 \pagebreak %TODO: pagebreak, maybe remove these by better formatting

### 2025-05-19 manuscript/manuscript.tex
Include Flowchart & Update Fig5

--- 2025-05-18 manuscript/manuscript.tex
@@
 If the reader prefers to skip the MD and GIST calculation, we also provide the GIST output files with this tutorial, such that the post-processing can be done without any expensive calculations.
-Additionally, the MD simulation files and raw GIST output files can be found at \url{https://doi.org/10.48323/4mbrd-67m83}, to reproduce the results shown in this tutorial.
 We additionally provide a Jupyter notebook \cite{Kluyver2016-jupyter,Granger2021-jupyter} with the presented analyses. 
@@
 This will download the tutorial files, including both the manuscript and the examples in the \inlinecode{code} folder.
-Additionally, the original MD trajectory files and raw GIST output files are provided under the DOI \url{https://doi.org/10.48323/4mbrd-67m83}.
+Additionally, a link to download the original MD trajectory files is provided in the GitHub repository.
 \subsection{Running GIST for Biotin}
@@
 \subsubsection{Visualizing $\dgsolv$}
-Next, we visualize the 3-dimensional contributions to the free energy of hydration, as well as the entropy and enthalpy contributions.
+Next, we visualize the enthalpic contributions to the free energy of solvation.
 You can use PyMol\cite{pymol} or VMD\cite{vmd} to visualize the .dx files from GIST\@.
@@
 import gisttools as gt
+# Load the GIST data, make sure to set the correct reference energy
 gist = gt.gist.load_gist_file('gist.dat',
@@
 The OpenDX files can then be imported into your molecular viewer of choice.
-A visualization of entropy and energy contributions to the free energy of solvation of the streptavidin binding pocket is shown in Figure~\ref{fig_binding_pocket_pymol}. 
+A visualization of enthalpy contributions to the free energy of solvation of the streptavidin binding pocket is shown in Figure~\ref{fig_binding_pocket_pymol}. 
+Enthalpywise, the streptavidin binding pocket is notably hydrophobic (red) in most parts except for where biotin is also interacting polarly with the protein (green).
+Nothe that this is "norm" data, so not density-weighted but rather comparing voxels on a per-molecule basis.
+The surfaces are therefore disjointed since even voxels seldomly sampled by the water are shown.
 PyMOL input scripts to generate such visualizations are available in the GitHub repository for this tutorial.
@@
 	\includegraphics[width=0.8\linewidth]{figures/binding_pocket_norm_quants.png}
-	\caption{Free energy contributions in the streptavidin binding pocket. Red: high solvent energy (at +2.5 
-	kcal/mol/molecule). Green: low solvent energy (at -2.5 kcal/mol/molecule). Blue: low solvent entropy $T\Delta S^{six}$ 
-	(at -2.5 kcal/mol/molecule).}\label{fig_binding_pocket_pymol}
+	\caption{Enthalpy contributions using "norm" data in the streptavidin binding pocket overlayed with biotin. Red: high solvent energy (at +3 
+	kcal/mol/molecule).
+Green: low solvent energy (at -3 kcal/mol/molecule).
+The enthalpy density is based on the apo structure and is shown only when within 1.5\AA of biotin.}\label{fig_binding_pocket_pymol}
 \end{figure}
@@
 
-\section{Checklists and Cheat Sheats} \label{sec:checklists}
+\section{Checklists, Cheat Sheats ad Flowcharts} \label{sec:checklists}
 %Tutorials do not necessarily require the use of a checklist as in Best Practices documents; however, they can include these if desired.
@@
 \end{Checklists*}
+
+\begin{Checklists*}
+	\includegraphics[width=0.9\linewidth,page=1]{figures/flowchart_GIST.pdf}
+\end{Checklists*}
+	\pagebreak
+	\begin{Checklists*}
+	\includegraphics[width=0.9\linewidth,page=2]{figures/flowchart_GIST.pdf}
+\end{Checklists*}
 \FloatBarrier
+
 \section*{Author Contributions}

### 2025-05-20 manuscript/manuscript.tex
Replace dG with dA
Plots not done yet

@@
 %%% Shortcuts and macros
-\newcommand{\dgsolv}{\Delta G_\textup{solv}}
+\newcommand{\dasolv}{\Delta A_\textup{solv}}
 \newcommand{\software}{\texttt}
@@
 Due to the high computational effort required to converge the higher order terms, the entropy expansion is usually truncated after the first or, sometimes, the second \cite{Nguyen2016-gist-second-order,Waibl2022-gist-solvents} term, thereby introducing an error.
-While there are approximations available to deal with those problems, most noteworthy the mutual information expansion (MIE) and Maximum Information Spanning Tree (MIST) approaches, they are not currently implemented in GIST.
+While there are approximations to deal with those problems, most noteworthy the mutual information expansion (MIE) and Maximum Information Spanning Tree (MIST) approaches, they are not currently implemented in GIST.
 Recent kNN implementations to calculate solvation entropy further improve phase space sampling by relabeling the solvent molecules in the simulation, treating the solvent molecules as distinguishable \cite{reinhard2007estimation,heinz2019computing,heinz2021permut,fogolari2021optimal}.
@@
 Similar to the self-consistent integral theories, GIST only considers a single solute conformation and  multiple computations of representative structures are necessary to describe ensemble properties.
+Additionally, like most other MD- and force-field based methods, a number of approximations are assumed, such as the pairwise additivity of the intermolecular interactions, the neglect of polarization and quantum effects, as well as the neglect of the solute's internal degrees of freedom.
+Recent perspectives on the various limitations inherent to the force-field based MD approach were provided by \cite{van2021effect} and \cite{van2024methods}.
 
@@
 	\item Compute biotin's free energy of solvation using GIST.
-	\item Visualize local contributions to $\dgsolv$ in the streptavidin binding pocket and around biotin.
+	\item Visualize local contributions to $\dasolv$ in the streptavidin binding pocket and around biotin.
 	\item Compute the hydration contribution of the biotin, streptavidin and complex to the binding. Combined with the biotin--streptavidin interaction energy, the binding free energy can be estimated.
-	\item Explore advantages and disadvantages of the GIST method when used to calculate  $\dgsolv$
+	\item Explore advantages and disadvantages of the GIST method when used to calculate $\dasolv$
 \end{enumerate}
@@
 However, depending on the method, additional post-processing may be required to keep the solute at the same position in the grid.
-To make sure that the solute is not moving significantly, we recommend to check the root mean square deviation (RMSD) of the solute heavy atoms during the simulation.
+To make sure that the solute is not moving significantly, we recommend checking the root mean square deviation (RMSD) of the solute heavy atoms during the simulation.
 The RMSD should be less than \SI{0.5}{\angstrom} for the solute heavy atoms, preferably less than \SI{0.1}{\angstrom}.
@@
 \end{lstlisting}
-Gisttools also calculates the total enthalpy per voxel (Eall) as a sum of Eww and Esw, and the free energy of solvation (A) by subtracting $\Delta$TSsix from Eall.\\
+Gisttools also calculates the total enthalpy per voxel (Eall) as a sum of Eww and Esw, and the free energy of solvation (A) by subtracting T$\Delta$Ssix from Eall.\\
 Information about the grid is stored in the grid property of the gist object and structural data (supplied as a pdb of the solute molecule) is stored as an \software{mdtraj} trajectory in the struct property:
@@
 \begin{equation}
-	\Delta G_{solv} = k_BT \cdot ln \int p(q_u) \cdot e^{ \frac{-\Delta G_solv(q_u)}{k_BT}} dq_u
+	\Delta A_{solv} = k_BT \cdot ln \int p(q_u) \cdot e^{ \frac{-\Delta A_{solv}(q_u)}{k_BT}} dq_u
 \end{equation}
@@
 \begin{equation}
-	\Delta G_{solv} = \frac{\sum_i p(q_u) \cdot \Delta G_solv(q_u)}{\sum_i p(q_u)}
+	\Delta A_{solv} = \frac{\sum_i p(q_u) \cdot \Delta A_{solv}(q_u)}{\sum_i p(q_u)}
 \end{equation}
@@
 In \software{gisttools}, this is done simply by setting \inlinecode{gist.eww\_ref}.
-Now, sum all free energy contributions to obtain the spatially resolved $\dgsolv$.
+Now, sum all free energy contributions to obtain the spatially resolved $\dasolv$.
 In \software{gisttools}, this is done automatically and can be accessed using the \inlinecode{A\_dens} and \inlinecode{A\_norm} data rows.
@@
 \begin{equation}
-	\dgsolv(\coordinate) = \Delta E_{ww}(\coordinate) + \Delta E_{sw}(\coordinate) - T\Delta S^\text{six}(\coordinate)
+	\dasolv(\coordinate) = \Delta E_{ww}(\coordinate) + \Delta E_{sw}(\coordinate) - T\Delta S^\text{six}(\coordinate)
 \end{equation}
@@
 However, it is more informative to plot the cumulative free energy contribution against the distance to the solute molecule, to check the convergence (see Figure~\ref{fig_radial_convergence}).
-If the curve flattens out, the converged value is your final $\dgsolv$\@.
+If the curve flattens out, the converged value is your final $\dasolv$\@.
 If the curve diverges even for large integration distances, it depends on which of the contributing parts diverges. 
@@
 		struct='solute-centered.pdb', eww_ref=-9.5398)
-	bins, (dg, esw, eww, s) = gist.rdf(
+	bins, (da, esw, eww, s) = gist.rdf(
 		['A_dens', 'Esw_dens', 'Eww_dens', 'dTSsix_dens'],
@@
 	plt.plot(bins, np.cumsum(esw), label='$\Delta$E$_{sw}$')
-	plt.plot(bins, np.cumsum(dg), label='$\Delta$G')
+	plt.plot(bins, np.cumsum(da), label='$\Delta$A)
 	plt.plot(bins, np.cumsum(s), label='T$\Delta$S')
@@
 	\includegraphics[width=0.8\linewidth]{figures/biotin-contributions_noref.png}
-	\caption{Convergence of $\dgsolv$ and its contributions with increasing distance to the solute molecule.
+	\caption{Convergence of $\dasolv$ and its contributions with increasing distance to the solute molecule.
 All quantities are cumulative, i.e., summed up to the respective radius.
@@
 	\includegraphics[width=0.8\linewidth]{figures/biotin-contributions_ref.png}
-	\caption{Convergence of $\dgsolv$ and its contributions with increasing distance to the solute molecule.
+	\caption{Convergence of $\dasolv$ and its contributions with increasing distance to the solute molecule.
 The entropy was referenced, so it converges to zero for voxels far away from the solute.}\label{fig_radial_convergence_ref}
@@
 
-\subsubsection{Visualizing $\dgsolv$}
+\subsubsection{Visualizing $\dasolv$}
 Next, we visualize the enthalpic contributions to the free energy of solvation.
@@
 \label{sec:binding_contributions}
-In the next step, calculate the $\dgsolv$ contributions around biotin in each system (biotin, streptavidin, and complex).
+In the next step, calculate the $\dasolv$ contributions around biotin in each system (biotin, streptavidin, and complex).
 Note that our integration region does not fully reach into bulk, and some of the boundary regions are close to streptavidin. 
@@
 plt.xlabel('distance to biotin [Å]')
-plt.ylabel('dG contribution [kcal/mol]')
+plt.ylabel('$\Delta A$ contribution [kcal/mol]')
 \end{lstlisting}
@@
 ax1.set_xlabel('distance to biotin [Å]')
-ax1.set_ylabel('$\Delta G$ contributions [Å]')
+ax1.set_ylabel('$\Delta A$ contributions [Å]')
 ax1.grid()
@@
 ax2.set_xlabel('distance to biotin [Å]')
-ax2.set_ylabel('$\Delta \Delta G$ [kcal/mol]')
+ax2.set_ylabel('$\Delta \Delta A$ [kcal/mol]')
 ax2.grid()
@@
 	\includegraphics[width=1.0\linewidth]{figures/deltaG-difference.png}
-	\caption{Left: $\dgsolv$ contributions around the location of biotin in the binding pocket, evaluated in the biotin, streptavidin, and complex systems.
+	\caption{Left: $\dasolv$ contributions around the location of biotin in the binding pocket, evaluated in the biotin, streptavidin, and complex systems.
 Each line represents a cumulative sum plotted against the distance to biotin.
@@
 \label{sec:theory}
-GIST is an implementation of inhomogeneous fluid solvation theory (IST) \cite{Lazaridis1998} that discretizes the free energy of solvation $\dgsolv$ on a three-dimensional grid. 
+GIST is an implementation of inhomogeneous fluid solvation theory (IST) \cite{Lazaridis1998} that discretizes the free energy of solvation $\dasolv$ on a three-dimensional grid. 
 It was first devised by Nguyen et al.\ \cite{Nguyen2012} and its implementation in \software{cpptraj} was thoroughly described by Ramsey et al.\ \cite{Ramsey2016}.
@@
 For more technical information (such as implementation details), we recommend one of the more recent publications on developments in GIST \cite{Kraml2020,Chen2021,Roe2023-mpi-gist}.
-
+\subsection{On thermodynamic quantities}
+Throughout this tutorial, we have refered to the free energy of solvation as $\dasolv$.
+Strictly speaking, the GIST enthalpic term is an energy not an enthalpy, and the GIST free energy of solvation is not the Gibbs free energy of solvation $\DeltaG_solv$ but the Helmholtz free energy of solvation $\DeltaA_solv$.
+However, the difference between the two is negligible for condensed phase systems at ambient conditions, since the PV term is small under these conditions.
+As such, $E \approx H$ and $\dasolv \approx \DeltaG_solv$.
+Similarly, the entropy columms are named dTS, rather than the more correct $T\DeltaS$.'
 \subsection{Solvation Entropy}
@@
 
+For a more detailed description of the PME-GIST implementation and the pairwise decomposition of the PME energy, we refer to the original publication \cite{Chen2021}.
 \section{Checklists, Cheat Sheats ad Flowcharts} \label{sec:checklists}
@@
 
-	\begin{checklist}{Obtaining absolute $\Delta G_\textup{solv}$}
+	\begin{checklist}{Obtaining absolute $\Delta A_\textup{solv}$}
 		\begin{itemize}

### 2025-05-20 manuscript/manuscript.tex
Add code comments and reference to Ben-Naims blunder

@@
 \end{lstlisting}
+For a full explenation of the input parameters, please refer to the Amber manual \cite{amber24}.
 This file will run a \SI{100}{\nano\second} NPT MD simulation while keeping all atoms except for hydrogens and waters restrained.
@@
 \begin{lstlisting}[style=cpptraj]
+# Load the topology ('parm') and coordinates of the first frame
 parm solvated.parm7 
 trajin md-01.nc 1 1 1 
+# Reimage and center the solute (not water residues) to the origin
 autoimage !(:WAT) origin 
+# Strip all water molecules and write out a pdb file
 strip :WAT 
@@
 \begin{lstlisting}[style=cpptraj]
+# Load the topology ('parm') and coordinates of the first frame
 parm solvated.parm7
 trajin md-01.nc 1 1 1
+# Reimage and center the solute (not water residues) to the origin
 autoimage !(:WAT) origin
+# Calculate the grid bounds around the solute with a buffer of 15\AA
+# and a voxel size of 0.5\AA
 bounds !(:WAT) dx 0.5 offset 30 name Grid out bounds.dat
@@
 \begin{lstlisting}[style=cpptraj]
+# Load the topology ('parm') and coordinates of the full trajectory
 parm solvated.parm7
 trajin md-01.nc
+# Reimage and center the solute (not water residues) to the origin
 autoimage !(:WAT) origin
+# Run GIST with a grid centered at the origin
+# and a grid size of 100x100x100 voxels with 0.5\AA spacing
 # optionally remove the 'pme' option
@@
 import matplotlib.pyplot as plt
+# Load the GIST data, setting the reference water-water energy directly
 gist = load_gist_file(
@@
 eww_ref=-9.539)
+# Print the number of frames and the reference density
 print(f'n_frames = {gist.n_frames}')
@@
 \begin{lstlisting}[style=python]
+# GIST objects contain pandas dataframes
 gist.loc[0, 'population'] # single value
@@
 \begin{lstlisting}[style=python]
+# Write out the column A_dens (the free solvation energy) to a dx file
 gist.save_dx('A_dens','A.dx')
@@
 \begin{lstlisting}[style=cpptraj]
+# Load the topology ('parm') and coordinates of the complex after equilibration
 parm complex/solvated.parm7
 trajin complex/equil/EQUIL-DONE.rst
+# Remove the biotin molecule (Nr.2) from the complex, save the topology and coordinates
 strip ^2 parmout streptavidin/solvated.parm7
@@
 Thus, it is called the solute-water entropy $S_{sw}$.
+Note that this definition is different and more direct than the one used originally by Ben-Naim \cite{ben-naim-book}, as recently discussed in \cite{heinz2023solvent}.
 %The second order is the pair density, i.e., the probability of finding two solvent molecules in positions $\textbf{r}$ and $\textbf{r}^\prime$.

### 2025-05-20 manuscript/manuscript.tex
Fix some Latex Bugs, add some more code comments

@@
 autoimage !(:WAT) origin
-# Calculate the grid bounds around the solute with a buffer of 15\AA
-# and a voxel size of 0.5\AA
+# Calculate the grid bounds with a buffer of 15Å
+# and a voxel size of 0.5Å
 bounds !(:WAT) dx 0.5 offset 30 name Grid out bounds.dat
@@
 A speed comparison of the various parallelization implementations is shown in \cite{Roe2023-mpi-gist}.
+The reason for the many different implementations is that GIST is a very expensive calculation, and historically, the CPU version was the only one available.
+The GPU version improved the speed of GIST calculations significantly, allowing for grids encompassing whole biomolecules for the first time.
+The PME version came after with the intention to match the energy of the AMBER MD engine more closely to compare with other free energy methods.
+Most recently, the MPI parallelization provided another significant speedup for all implementations.
 
@@
 However, in practice the target may be flexible and a single structure may not be representative of the conformational ensemble.
-A more sophisticated approach would identify several dominant structures in the ensemble and run a separate MD simulation and GIST calculation for each structure. 
+A more sophisticated approach would identify several dominanÅt structures in the ensemble and run a separate MD simulation and GIST calculation for each structure. 
 Such dominant structures could either be identified from experimental ensembles or by clustering an unrestrained MD simulation. 
@@
 import matplotlib.pyplot as plt
+# Load the GIST data, setting the reference water-water energy directly
 gist = load_gist_file(
     'gist.dat', struct='solute-centered.pdb')
+# Run an rdf calculation to show the unref. Eww 
 bins, eww = gist.rdf(
@@
 plt.plot(bins, eww)
+# Run the Eww autodetection
 eww_auto = gist.detect_reference_value()
@@
 	def reference_entropy(gf):
+	'''
+	Reference the entropy to zero at large distances.
+	'''
     if 'dTSsix_unref_norm' not in gf.data.columns:
@@
     struct='solute-centered.pdb', eww_ref=-9.533)
+# Generate a new column with the energy without accounting for double counting
 gist['E_norm'] = gist['Eww_norm'] * 2 + gist['Esw_norm']
+# Set the energy to zero for all voxels with no population
 gist.loc[gist['population'] == 0, 'E_norm'] = 0
+# Save the new energy to a dx file
 gist.save_dx('E_norm', 'gist-E-per-mol-norm.dx')
@@
 import matplotlib.pyplot as plt
-
+# Load the GIST data, setting no reference energy
+# since we will calculate it later
 compl = load_gist_file('complex/gist/gist.dat', \
@@
 \begin{lstlisting}[style=python]
+# Set the reference density for the solvent by automatic detection
 biotin.eww_ref = biotin.detect_reference_value()
@@
 \begin{lstlisting}[style=python]
-def reference_entropy(gf):
-    if 'dTSsix_unref_norm' not in gf.data.columns:
-        gf['dTSsix_unref_norm'] = gf['dTSsix_norm']
-        gf['dTSsix_unref_dens'] = gf['dTSsix_dens']
-    gf['dTSsix_norm'] = gf['dTSsix_unref_norm'] \
-        - gf.detect_reference_value('dTSsix_unref_dens')
-    gf['dTSsix_dens'] = gf.norm2dens(gf['dTSsix_norm'])
 reference_entropy(biotin)
@@
 strept_mask = 'not resname BTN and not resname WAT and not element H'
+# select the biotin atoms
 compl_x = select(compl.struct, biotin_mask).xyz[0] * 10.
-biotin_x = select(biotin.struct, biotin_mask).xyz[0]*10.
+biotin_x = select(biotin.struct, biotin_mask).xyz[0] * 10.
+# align the complex to streptavidin
+# and select the streptavidin atoms
 aligned = compl.struct[:].superpose(strept.struct, \
@@
 
+# Now we can calculate the radial distribution functions
 bins, biotin_rdf = biotin.rdf( \
@@
 \begin{lstlisting}[style=python]
+# Endstate approach: complex - monomers
 difference = compl_rdf - biotin_rdf - strept_rdf
+# Set integration cutoff
 cutoff = 12
+# Integrate the difference
 integral = difference[bins < cutoff].sum()
 print('Integral = {}'.format(integral))
+# Plot the cumulative sum of the difference
 plt.plot(bins, np.cumsum(difference))
@@
 import pandas as pd
+# Integrate the energy and entropy contributions
 centers = {'biotin': biotin_x, 
@@
 \begin{lstlisting}[style=cpptraj]
+# Load the topology and trajectory of the complex
 parm solvated.parm7
 trajin md-01.nc 1 last 100
+# Calculate the interaction energy of the complex and the monomers
 energy complex ^1,2 # etype pme
@@
 Throughout this tutorial, we have refered to the free energy of solvation as $\dasolv$.
-Strictly speaking, the GIST enthalpic term is an energy not an enthalpy, and the GIST free energy of solvation is not the Gibbs free energy of solvation $\DeltaG_solv$ but the Helmholtz free energy of solvation $\DeltaA_solv$.
+Strictly speaking, the GIST enthalpic term is an energy not an enthalpy, and the GIST free energy of solvation is not the Gibbs free energy of solvation $\Delta G_solv$ but the Helmholtz free energy of solvation $\Delta A_solv$.
 However, the difference between the two is negligible for condensed phase systems at ambient conditions, since the PV term is small under these conditions.
-As such, $E \approx H$ and $\dasolv \approx \DeltaG_solv$.
-Similarly, the entropy columms are named dTS, rather than the more correct $T\DeltaS$.'
+As such, $E \approx H$ and $\dasolv \approx \Delta G_solv$.
+Similarly, the entropy columms are named dTS, rather than the more correct $T\Delta S$.
 \subsection{Solvation Entropy}
@@
 Note that this definition is different and more direct than the one used originally by Ben-Naim \cite{ben-naim-book}, as recently discussed in \cite{heinz2023solvent}.
-%The second order is the pair density, i.e., the probability of finding two solvent molecules in positions $\textbf{r}$ and $\textbf{r}^\prime$.
+%The second order is the pairF density, i.e., the probability of finding two solvent molecules in positions $\textbf{r}$ and $\textbf{r}^\prime$.
 The solvent density is expressed relative to the bulk density $\rho^0$. %$g_\textit{sw}\left(\textbf{r},\omega \right)$ is the solute-water 

### 2025-05-20 manuscript/manuscript.tex
Small changes

@@
 \end{lstlisting}
-The atom mask \inlinecode{:LIG} specifies the residue named LIG to build the boundaries around, with dx and offset providing the actual dimensions of the boundaries, i.e. an offset of 30 times 0.5\AA.
-\begin{minipage}{\linewidth} 
+The atom mask \inlinecode{:LIG} specifies the residue named LIG to build the boundaries around, with dx and offset providing the actual dimensions of the boundaries, i.e. an offset of 30 times 0.5Å.
 Make sure that the \inlinecode{bounds} command is supplied with the exact same trajectory as in the GIST calculation. 
@@
 A helper script called \inlinecode{findcentroid.py} for the same purpose is supplied with this tutorial, and can be used as follows:
-~
 \begin{lstlisting}[style=bash]
@@
 \end{lstlisting}
-\end{minipage}
 \paragraph{Run GIST using cpptraj}
@@
 For more technical information (such as implementation details), we recommend one of the more recent publications on developments in GIST \cite{Kraml2020,Chen2021,Roe2023-mpi-gist}.
-\subsection{On thermodynamic quantities}
+\subsection{On Thermodynamic Quantities}
 Throughout this tutorial, we have refered to the free energy of solvation as $\dasolv$.

### 2025-05-23 manuscript/manuscript.tex

@@
 Due to the high computational effort required to converge the higher order terms, the entropy expansion is usually truncated after the first or, sometimes, the second \cite{Nguyen2016-gist-second-order,Waibl2022-gist-solvents} term, thereby introducing an error.
-While there are approximations to deal with those problems, most noteworthy the mutual information expansion (MIE) and Maximum Information Spanning Tree (MIST) approaches, they are not currently implemented in GIST.
+
+While there are approximations available to deal with those problems, most noteworthy the mutual information expansion (MIE) and Maximum Information Spanning Tree (MIST) approaches, they are not currently implemented in GIST.
 Recent kNN implementations to calculate solvation entropy further improve phase space sampling by relabeling the solvent molecules in the simulation, treating the solvent molecules as distinguishable \cite{reinhard2007estimation,heinz2019computing,heinz2021permut,fogolari2021optimal}.
@@
 On the other hand, the first-order approximation means that the solvation entropy is less accurately described than in TI or FEP \cite{Chen2021,Waibl2021-gist-salt}.
+
 Similar to the self-consistent integral theories, GIST only considers a single solute conformation and multiple computations of representative structures are necessary to describe ensemble properties.
-Additionally, like most other MD- and force-field based methods, a number of approximations are assumed, such as the pairwise additivity of the intermolecular interactions, the neglect of polarization and quantum effects, as well as the neglect of the solute's internal degrees of freedom.
-Recent perspectives on the various limitations inherent to the force-field based MD approach were provided by \cite{van2021effect} and \cite{van2024methods}.
+
+Aside from the previously mentioned methods, there is a large variety of other methods to compute enthalpy and entropy contributions to the solvation free energy, as recently reviewed \cite{Heyden2019-localsolvation,Mukherjee2022-ReviewSpatialWatersBiomolecule}.
+Approaches like SZMAP \cite{Bayden2015-SZMAP} trade the accuracy of all-atom MD simulation for speed by using a faster probe-based approach.
+SPAM ("maps" in reverse) is more akin to WaterMaps, investigating distinct hydration sites based on high local water densities \cite{Cui2013-SPAM}. 
+However, it approximates the water entropies from the water interaction enthalpy distributions.
+Another group of methods builds on the cell theory for liquids \cite{henchman2003partition}, which derives independent effective potentials for each molecule in its' neighbors mean field.
+Conceptually close to GIST is Grid Cell Theory (GCT), which also adopts a grid-based approach but differs in the assumptions made its' entropy calculations \cite{Gerogiokas2014-GCT}.
+More recently, Multiscale Cell Correlation (MCC) has been developed, generalizing and pushing the method from simple liquids up to large biomolecules \cite{henchman2007free, Hensen2014, Ali2021-hc, Kalayan2022-MCCBiomol}.
 
@@
 \end{itemize}
-If the reader prefers to skip the MD and GIST calculation, we also provide the GIST output files with this tutorial, such that the post-processing can be done without any expensive calculations.
+If the reader prefers to skip the MD and GIST calculation, GIST output files are provided with this tutorial, such that the post-processing can be done without any expensive calculations.
+Additionally, the MD simulation files and raw GIST output files can be found at \url{https://doi.org/10.48323/4mbrd-67m83}, to reproduce the results shown in this tutorial.
 We additionally provide a Jupyter notebook \cite{Kluyver2016-jupyter,Granger2021-jupyter} with the presented analyses. 

### 2025-05-23 manuscript/manuscript.tex
Final changes
Fixes typos, figure captions, some sentences as well

@@
 \DeclareSIUnit{\bar}{bar}
-\DeclareSIUnit\Molar{M}
-\DeclareSIUnit\calorie{cal}
-\DeclareSIUnit\kcalPerMolASqr{\kilo\calorie\per\mole\per\angstrom\squared}
-%\DeclareSIUnit{\Molar}{M}
-\DeclareSIUnit\angstrom{\text{\AA}}
+\DeclareSIUnit{\Molar}{M}
+\DeclareSIUnit{\calorie}{cal}
+\DeclareSIUnit{\kcalPerMolASqr}{\kilo\calorie\per\mole\per\angstrom\squared}
+\DeclareSIUnit{\angstrom}{\text{\AA}}
 %\DeclareSIPrefix\micro{\text{\textmu}}{0}
-\sisetup{
-	%math-rm=\mathsf,
-	text-rm=\sffamily
-}
+%\sisetup{
+%	%math-rm=\mathsf,
+%	text-rm=\sffamily
+%}
 
@@
 \usepackage{tabularx}
-
+\usepackage{microtype}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \lstdefinestyle{code}{}
-\lstdefinestyle{cpptraj}{title={\centering cpptraj}}
+\lstdefinestyle{cpptraj}{title={\centering Cpptraj}}
 \lstdefinestyle{bash}{title={\centering Command-line}}
 \lstdefinestyle{amber-in}{title={\centering Amber input}}
-\lstdefinestyle{pymol}{title={\centering Pymol}}
+\lstdefinestyle{pymol}{title={\centering PyMOL}}
 \lstdefinestyle{python}{
@@
 Grid inhomogeneous solvation theory (GIST) is a method to compute the free energy of solvation of a solute molecule on a 3-dimensional grid based on sampling from molecular dynamics (MD) simulations.
-The high spatial resolution of the GIST output, as well as the decomposition into energy and entropy contributions, allow for highly detailed analyses of solvation around both proteins and small molecules contexts.
+The high spatial resolution of the GIST output, as well as the decomposition into energy and entropy contributions, allow for highly detailed analyses of solvation around both proteins and small molecules.
 However, this versatility also comes with a significant entry barrier for new users.
@@
 In this tutorial, we aim to guide the reader through the most common steps involved in a GIST analysis using the streptavidin-biotin complex as a demonstrative system.
-Towards this, Jupyter notebooks and Python packages are provided to simplify the analysis.
+Towards this, Jupyter notebooks and a Python package (gisttools) are provided to simplify the analysis.
 Furthermore, we discuss the theory of GIST with a focus on practical aspects.
@@
 	\caption{Schematic showing regions of high water density around a carbazole solute molecule.
-Based on an MD simulation of solvent around the solute molecule, the energy is computed from the intermolecular interactions of a force field, while the entropy is approximated from the 6-dimensional orientational and solvational solvent distribution $\rho$ around the solute molecule.
+Based on an MD simulation of solvent around the solute molecule, the energy is computed from the intermolecular interactions of a force field, while the entropy is approximated from the 6-dimensional orientational and translational solvent distribution $\rho$ around the solute molecule.
 Note that while this depiction focuses on regions of high solvent density, regions with lower density can also contain meaningful contributions to the overall energy and entropy.}
@@
 Interactions with water are of special importance as many biological processes occur in an aqueous environment \cite{Privalov2017-water-review}.
-The free energy of solvation is defined as the reversible work needed to insert a solute molecule into the solvent from gas phase \cite{ben-naim-book}.
+The free energy of solvation is defined as the reversible work needed to insert a solute molecule into the solvent from the gas phase \cite{ben-naim-book}.
 It is directly linked to the solubility of solutes and drives many real-world effects. 
@@
 A schematic of this is shown in Figure~\ref{fig:carbazole}.
-The enthalpic contributions are computed from the intermolecular interactions of the solvent with the solute and as well as other solvent. 
+The enthalpic contributions are computed from the intermolecular interactions of the solvent with the solute as well as between solvent molecules. 
 As they are derived from the force field, pairwise additivity of the underlying intermolecular interactions is assumed.
@@
 The calculation of higher-order entropy terms is computationally demanding but was realized in recent works \cite{Nguyen2016-gist-second-order, Waibl2021-gist-salt}.
-However, it has been found that solvation free energies of small molecule solutes can be computed at high accuracy using linear scaling factors for the higher-order terms, such as $\Delta S = 0.6\, \Delta S^\text{1st-order}$ for water \cite{Chen2021,Waibl2022-gist-solvents}.
+However, it has been found that solvation free energies of small molecule solutes can be computed at high accuracy using linear scaling factors for the higher-order terms. 
+A common approximation is to scale the first-order entropy, for example, $\Delta S = 0.6\, \Delta S^\text{1st-order}$ for water \cite{Chen2021,Waibl2022-gist-solvents}.
 Since its initial development, GIST has been used to describe the thermodynamics of water displacement in protein binding pockets \cite{Nguyen2016-gist-second-order,Ramsey2016,Balius2017-gist-ligand-discovery,Hufner-Wulsdorf2020-gist-drug-design}.
 It has been used for a large-scale investigation of the hydration properties of proteins found in SARS-CoV-2 \cite{Olson2020-covid-gist}.
-Furthermore, recent works used GIST to describe biophysical properties of serine proteases \cite{Kraml2019-gigist}, macrocycles \cite{Kamenik2020-gist-macrocycles}, or antibody fragements \cite{Waibl2021-gist-antibodies}.
+Furthermore, recent works used GIST to describe biophysical properties of serine proteases \cite{Kraml2019-gigist}, macrocycles \cite{Kamenik2020-gist-macrocycles}, or antibody fragments \cite{Waibl2021-gist-antibodies}.
 
@@
 In this tutorial, we aim to give the reader an introduction to GIST and its application to calculate solution thermodynamics properties of interest. 
-While previous publications \cite{Ramsey2016} and tutorials \cite{amber_tut_gist} treated how to setup and run GIST calculations in detail, the post-processing was often treated in less detail. 
+While previous publications \cite{Ramsey2016} and tutorials \cite{amber_tut_gist} treated how to set up and run GIST calculations in detail, the post-processing was often treated in less detail. 
 While building on these tutorials, we provide an updated guide to calculating solvation properties using GIST to address recent improvements.
@@
 The main limitation of those methods is that the entropy is approximated through an infinite expansion in terms of the density distribution. 
-Within GIST and other more recent methods, these entropy contributions are calulated via a k-nearest-neighbor (kNN) approach.  
+Within GIST and other more recent methods, these entropy contributions are calculated via a k-nearest-neighbor (kNN) approach.  
 Due to the high computational effort required to converge the higher order terms, the entropy expansion is usually truncated after the first or, sometimes, the second \cite{Nguyen2016-gist-second-order,Waibl2022-gist-solvents} term, thereby introducing an error.
@@
 Another group of methods builds on the cell theory for liquids \cite{henchman2003partition}, which derives independent effective potentials for each molecule in its' neighbors mean field.
-Conceptually close to GIST is Grid Cell Theory (GCT), which also adopts a grid-based approach but differs in the assumptions made its' entropy calculations \cite{Gerogiokas2014-GCT}.
+Conceptually close to GIST is Grid Cell Theory (GCT), which also adopts a grid-based approach but differs in the assumptions made in its entropy calculations \cite{Gerogiokas2014-GCT}.
 More recently, Multiscale Cell Correlation (MCC) has been developed, generalizing and pushing the method from simple liquids up to large biomolecules \cite{henchman2007free, Hensen2014, Ali2021-hc, Kalayan2022-MCCBiomol}.
@@
 We show how to interpret the three-dimensional contributions of solvation free energy in a binding pocket and around the ligand.
-Additionally, we compute the binding contribution of the free energy of solvation, which requires accurate post-processing of the GIST outputs to avoid unfavorable summation of bias.
+Additionally, we compute the contribution of solvation free energy to binding, which requires accurate post-processing of the GIST outputs to avoid unfavorable summation of bias.
 In total, we aim to provide a workflow that can be easily adapted to different systems, an introduction to the method and the theory behind it, and an overview of the different implementations of GIST.\\
@@
 We additionally provide a Jupyter notebook \cite{Kluyver2016-jupyter,Granger2021-jupyter} with the presented analyses. 
-For the visualization of three-dimensional contributions we require molecular visualization software that can handle \software{OpenDX} files, such as \software{PyMol} \cite{pymol} or \software{VMD} \cite{vmd}.
+For the visualization of three-dimensional contributions we require molecular visualization software that can handle \software{OpenDX} files, such as \software{PyMOL} \cite{pymol} or \software{VMD} \cite{vmd}.
 Most of the visualisations and results in this tutorial can also be obtained using the \software{gistpp} program, which is available from \url{https://github.com/KurtzmanLab/Gist-Post-Processing} (code and documentation).
 \software{Gistpp} can directly manipulate \software{OpenDX} files from the GIST output. 
-Its usage and some examples were shown in a previous tutorial on GIST \cite{Ramsey2016}.
+%Its usage and some examples were shown in a previous tutorial on GIST \cite{Ramsey2016}.
 \pagebreak %TODO: pagebreak, maybe remove these by better formatting
-
 \section{Running GIST}
@@
 This region is defined by a rectangular grid with given position, size, and voxel spacing.
-The default grid spacing of \SI{0.5}{\angstrom} provides reasonable discretization for many use-cases.
+The default grid spacing of \qty{0.5}{\angstrom} provides reasonable discretization for many use-cases.
 While integrals of GIST quantities over the grid are independent of the voxel size, provided sufficient sampling, individual values converge slower at high resolutions.
@@
 Generally, a grid encompassing 2 to 3 solvent layers around the solute is sufficient. 
-For TIP3P water, this corresponds to a buffer zone of around \SI{10}{\angstrom} around the solute. 
+For TIP3P water, this corresponds to a buffer zone of around \qty{10}{\angstrom} around the solute. 
 Larger grids might be necessary, for example for strongly charged solutes.
@@
 Using \inlinecode{cpptraj's } \inlinecode{bounds} command, boundaries can be output to a file. 
-For example, here is the command for a boundary of \SI{15}{\angstrom} around the ligand LIG:
+For example, here is the command for a boundary of \qty{15}{\angstrom} around the ligand LIG:
 \begin{lstlisting}[style=cpptraj]
@@
 The command needed to run GIST in cpptraj can be written in an input file (e.g gist.in) for convenience.
-The basic command, assuming a grid centered at coordinates (x=\SI{10.5}{\angstrom}, y=\SI{20.1}{\angstrom}, z=\SI{30.1}{\angstrom}) and a grid size of $30\times30\times30$ voxels with \SI{0.5}{\angstrom} spacing, is:
+The basic command, assuming a grid centered at coordinates (x=\qty{10.5}{\angstrom}, y=\qty{20.1}{\angstrom}, z=\qty{30.1}{\angstrom}) and a grid size of $30\times30\times30$ voxels with \qty{0.5}{\angstrom} spacing, is:
 
@@
 \end{lstlisting}
-This command tells \software{cpptraj} to run a GIST calculation using the specified region and print the output in a file called \inlinecode{gist.dat}.
+This command tells \software{cpptraj} to run a GIST calculation using the specified region and print the grid data in a file called \inlinecode{gist.dat}.
+In addition, .dx files for some default thermodynamic densities are generated as well.
 
@@
 		\midrule
-		gridcntr    & <y> <y> <z> & Coordinates of the grid center in \AA{} & 0 0 0  \\
+		gridcntr    & <x> <y> <z> & Coordinates of the grid center in \AA{} & 0 0 0  \\
 		griddim     & <nx> <ny> <nz> & Number of voxels per axis & 40 40 40\\
@@
 We also provide the densities for water models commonly used with AMBER force fields in Table~\ref{tab:ref_densities}.
-These densities were calculated from \SI{100}{\nano\second} MD simulations at \SI{1}{\bar} and \SI{300}{\kelvin} using the Berendsen barostat and the Langevin thermostat.
-The water boxes were prepared with \software{tleap} to a size of \SI{100}{\angstrom} x \SI{100}{\angstrom} x \SI{100}{\angstrom} before pressure equilibration.
+These densities were calculated from \qty{100}{\nano\second} MD simulations at \qty{1}{\bar} and \qty{300}{\kelvin} using the Berendsen barostat and the Langevin thermostat.
+The water boxes were prepared with \software{tleap} to a size of \qty{100}{\angstrom} x \qty{100}{\angstrom} x \qty{100}{\angstrom} before pressure equilibration.
 
@@
 
-	\caption{Reference number density $\rho_0$ and mean solvent-solvent energy $E_{ww, norm}$ for various water models from NPT simulations of N$_{WAT}$ water solvent molecules at \SI{300}{\kelvin} and 1bar.}\label{tab:ref_densities}
+	\caption{Reference number density $\rho_0$ and mean solvent-solvent energy $E_{ww, norm}$ for various water models from NPT simulations of N$_{WAT}$ water molecules at \qty{300}{\kelvin} and 1bar.}\label{tab:ref_densities}
 	\small
@@
 \paragraph{Postprocessing and Analyzing GIST results}
-Cpptraj outputs grid quantities in OpenDX (.dx) file format which can be visualized in Pymol or VMD.
+Cpptraj outputs grid quantities in OpenDX (.dx) file format which can be visualized in PyMOL or VMD.
 Aside from thermodynamic quantities, GIST also calculates the number density of each atomic element in the solvent.
@@
 	\item{[gX] For every element in the main solvent, the number density of atoms found in the voxel, in units of the bulk density.
-If the same element occurs multiple times, the bulk density is scaled such that that the expectation value for the bulk is unity.
+If the same element occurs multiple times, the bulk density is scaled such that the expectation value for the bulk is unity.
 For water, \code{gO} and \code{gH} are produced.}
@@
 \newcommand{\appr}{{\mathord{\sim}}}
-In nature, streptavidin occurs as a tetramer, and binds biotin with a binding constant of \SI{\sim e-14}{\per\Molar} \cite{Dundas2013-streptavidin-review}.
+In nature, streptavidin occurs as a tetramer, and binds biotin with a binding constant of \qty{\sim e14}{\per\Molar} \cite{Dundas2013-streptavidin-review}.
 Streptavidin has a so-called flap region, which closes the binding pocket after biotin binding.
 Here, we will investigate the binding of biotin to a closed streptavidin monomer for simplicity.
-The binding to the closed conformation of the tetramer has been estimated to have a free energy of \SI{-26.6}{\kilo\calorie\per\mol} \cite{Bansal2018-biotin}.
+The binding to the closed conformation of the tetramer has been estimated to have a free energy of \qty{-26.6}{\kilo\calorie\per\mol} \cite{Bansal2018-biotin}.
 
@@
 For periodic boundary conditions, the size of the solvent box should be large enough for the solvent to behave bulk-like so as to reduce artifacts from solvent-mediated interactions between the periodic images.
-This corresponds to a distance between solute and box boundary of at least \SI{12}{\angstrom} for water \cite{Chen2021,Waibl2022-gist-solvents} though we recommend \SI{15}{\angstrom} or higher.
+This corresponds to a distance between solute and box boundary of at least \qty{12}{\angstrom} for water \cite{Chen2021,Waibl2022-gist-solvents} though we recommend \qty{15}{\angstrom} or higher.
 Carefully check the employed force field, protonation state and partial charges.
@@
 The force constant $k_r$ should be chosen strong enough to prevent significant deviation.
-Here, we apply a harmonic restraint of \SI{100}{\kilo\calorie\per\mole\per\angstrom\squared} to all heavy atoms to keep the system in place. 
+Here, we apply a harmonic restraint of \qty{100}{\kilo\calorie\per\mole\per\angstrom\squared} to all heavy atoms to keep the system in place. 
 Note that this is a rather strong restraint. 
 While weaker restraints are possible, make sure that the solute molecule is kept rigid during the simulation. 
-As a minimum value, we recommend \SI{10}{\kilo\calorie\per\mole\per\angstrom\squared} restraints.
+As a minimum value, we recommend \qty{10}{\kilo\calorie\per\mole\per\angstrom\squared} restraints.
 Similarly, the solute could also be kept rigid using positional constraints, RMSD restraints, or other methods.
@@
 To make sure that the solute is not moving significantly, we recommend checking the root mean square deviation (RMSD) of the solute heavy atoms during the simulation.
-The RMSD should be less than \SI{0.5}{\angstrom} for the solute heavy atoms, preferably less than \SI{0.1}{\angstrom}.
-We run \SI{100}{\nano\second} of simulation and store the conformations every \SI{100}{\pico\second}.
-A simulation length of at least \SI{20}{\nano\second} is recommended to obtain statistically independent snapshots of the water solvent molecules. 
+The RMSD should be less than \qty{0.5}{\angstrom} for the solute heavy atoms, preferably less than \qty{0.1}{\angstrom}.
+We run \qty{100}{\nano\second} of simulation and store the conformations every \qty{100}{\pico\second}.
+A simulation length of at least \qty{20}{\nano\second} is recommended to obtain statistically independent snapshots of the water solvent molecules. 
 An example Amber input file for GIST might look like this:
@@
 For a full explenation of the input parameters, please refer to the Amber manual \cite{amber24}.
-This file will run a \SI{100}{\nano\second} NPT MD simulation while keeping all atoms except for hydrogens and waters restrained.
+Note the restraintmask, which specifies that all atoms except for hydrogens and water should be restrained, i.e. all heavy atoms of the solute.
+This file will run a \qty{100}{\nano\second} NPT MD simulation while keeping all atoms except for hydrogens and waters restrained.
 Note that these are very standard MD settings except for the restraints.
@@
 To decide on a grid size, find the extent (minimum and maximum) of the x-, \mbox{y-,} and z-coordinates of the molecule.
-To get bulk-like water properties at the grid border, a buffer distance of \SI{15}{\angstrom} is reasonable for biotin.
+To get bulk-like water properties at the grid border, a buffer distance of \qty{15}{\angstrom} is reasonable for biotin.
 The grid parameters can be directly calculated using \software{cpptraj's} \inlinecode{bound} command:
@@
 \end{lstlisting}
-In the above snippet, we calculate the bounds around the whole protein for a grid with \SI{0.5}{\angstrom} side-length voxels and a buffer zone of \SI{15.0}{\angstrom}. 
+First, we load the topology and coordinates of the first frame of our MD simulation (1 1 1 meaning start=1, stop=1 and stride=1).
+Then, we reimage the coordinates and center the solute at the origin.
+Here, dx 0.5 sets a voxel spacing of 0.5 Å for the boundary determination, and offset 30 adds a buffer of 30 voxels (i.e., 30 * 0.5 Å = 15 Å) to each dimension.
 The atom mask !(:WAT) specifies that the grid should be calculated around all atoms that are not water residues.
@@
 # and a grid size of 100x100x100 voxels with 0.5\AA spacing
-# optionally remove the 'pme' option
+# The 'pme' option enables PME-GIST calculations.
 gist griddim 100 100 100 out gist.dat \
@@
 import matplotlib.pyplot as plt
-# Load the GIST data, setting the reference water-water energy directly
+# Load the GIST data, setting the reference water-water energy directly to -9.539 for TIP3P water
 gist = load_gist_file(
@@
 Such dominant structures could either be identified from experimental ensembles or by clustering an unrestrained MD simulation. 
-When the probabilities of the different conformations are known, one can then use a weighted average of the GIST results to obtain a free energy of solvation for the ensemble \cite{Waibl2021-gist-antibodies}:
+When the probabilities of the different conformations $p(q_u)$ are known, one can then use a weighted average of the GIST results to obtain a free energy of solvation $\Delta A_{solv}$ for the ensemble \cite{Waibl2021-gist-antibodies}:
 \begin{equation}
@@
 trajin complex/equil/EQUIL-DONE.rst
-# Remove the biotin molecule (Nr.2) from the complex, save the topology and coordinates
+# Assuming biotin is residue 2 in the complex topology (selection ^2):
+# Remove biotin from the complex, save the topology and coordinates
 strip ^2 parmout streptavidin/solvated.parm7
@@
 Therefore, the reference energy can be obtained by binning the voxels by their distance to the solute molecule, and evaluate Equation~\ref{eq_ewwref} within each bin.
-If this value converges with increasing distance from the solute, it can be used as $E_{ww}^\textit{ref}$.
+If this value converges with increasing distance from the solute, it can be used as $E_{ww, ref}^\textit{ref}$.
 
@@
 eww_auto = gist.detect_reference_value()
-plt.axhline(eww_auto, color='k')
+plt.axhline(eww_auto, color='k', linestyle='--')
 plt.gca().set(
-    xlabel='distance [A]', ylabel='E_ww per mol')
+    xlabel='Distance [$\AA$]', ylabel='Avg. E$_{ww}$ per molecule [kcal/mol]', xlim=(1, 15))
 plt.show()
@@
 	\centering
-	\includegraphics[width=0.8\linewidth]{figures/Eww_convergence.eps}
+	\includegraphics[width=0.8\linewidth]{figures/Eww_convergence.png}
 	\caption{Convergence of $E_{ww}$ in the complex calculation with increasing distance to the solute molecule (streptavidin-biotin).
@@
 Make sure to subtract the reference from the normalized (\_norm) data.
-In \software{gisttools}, this is done simply by setting \inlinecode{gist.eww\_ref}.
+In \software{gisttools}, this is done simply by setting \inlinecode{gist.eww\_ref}:
+\begin{lstlisting}[style=python]
+gist.eww_ref = gist.detect_reference_value()
+\end{lstlisting}
 Now, sum all free energy contributions to obtain the spatially resolved $\dasolv$.
@@
 	plt.plot(bins, np.cumsum(esw), label='$\Delta$E$_{sw}$')
-	plt.plot(bins, np.cumsum(da), label='$\Delta$A)
+	plt.plot(bins, np.cumsum(da), label='$\Delta$A')
 	plt.plot(bins, np.cumsum(s), label='T$\Delta$S')
@@
 Next, we visualize the enthalpic contributions to the free energy of solvation.
-You can use PyMol\cite{pymol} or VMD\cite{vmd} to visualize the .dx files from GIST\@.
+You can use PyMOL\cite{pymol} or VMD\cite{vmd} to visualize the .dx files from GIST\@.
 With \software{gisttools}, you can create one using e.g., \inlinecode{gist.save\_dx('A\_dens', 'A\_dens.dx')}.
@@
 Enthalpywise, the streptavidin binding pocket is notably hydrophobic (red) in most parts except for where biotin is also interacting polarly with the protein (green).
-Nothe that this is "norm" data, so not density-weighted but rather comparing voxels on a per-molecule basis.
+Note that this is "norm" data, so not density-weighted but rather comparing voxels on a per-molecule basis.
 The surfaces are therefore disjointed since even voxels seldomly sampled by the water are shown.
@@
 Green: low solvent energy (at -3 kcal/mol/molecule).
-The enthalpy density is based on the apo structure and is shown only when within 1.5\AA of biotin.}\label{fig_binding_pocket_pymol}
+The enthalpy density is based on the apo structure and is shown only when within \qty{1.5}{\angstrom} of biotin.}\label{fig_binding_pocket_pymol}
 \end{figure}
@@
 	\centering
-	\includegraphics[width=1.0\linewidth]{figures/deltaG-difference.png}
+	\includegraphics[width=1.0\linewidth]{figures/deltaA-difference.png}
 	\caption{Left: $\dasolv$ contributions around the location of biotin in the binding pocket, evaluated in the biotin, streptavidin, and complex systems.
@@
 Right: The hydration contribution to binding, evaluated as the difference between the lines in the left panel.
-The vertical dashed line is at \SI{12}{\angstrom} and represents the chosen integration cutoff.}\label{fig-dg-sums}
+The vertical dashed line is at \qty{12}{\angstrom} and represents the chosen integration cutoff.}\label{fig-dg-sums}
 \end{figure}
@@
 trajin md-01.nc 1 last 100
-# Calculate the interaction energy of the complex and the monomers
+# Calculate the interaction energy of the complex (^1,2) and the monomers (^1 for streptavidin, ^2 for biotin)
 energy complex ^1,2 # etype pme
@@
 
-Prior computational studies suggest a $\Delta G$ of \SI{-26.6}{\kilo\calorie\per\mole} for the binding of biotin into the closed conformation of streptavidin \cite{Bansal2018-biotin}.
+Prior computational studies suggest a $\Delta G$ of \qty{-26.6}{\kilo\calorie\per\mole} for the binding of biotin into the closed conformation of streptavidin \cite{Bansal2018-biotin}.
 This indicates that our result is in good agreement with literature.
@@
 Furthermore, PME-GIST offers the best agreement between GIST energies and those observed in an classical MD simulation.
-To run PME-GIST, the \inlinecode{pme} flag to \inlinecode{gist} can be used in \software{cpptraj}.
+To run PME-GIST, the \inlinecode{pme} flag to the \inlinecode{gist} command can be used in \software{cpptraj}.
 The output will contain two additional columns: \inlinecode{PME\_dens} and \inlinecode{PME\_norm}.
@@
 In references~\cite{Kraml2020,Kamenik2020-gist-macrocycles,Waibl2022-gist-solvents}, an extension of GIST towards solvents other than water is described.
-To run a GIST calculation in a solvent other than water, use \inlinecode{solute} with a \software{cpptraj} selection mask to select the solute molecule.
+To run a GIST calculation in a solvent other than water, use the \inlinecode{solute} flag with a \software{cpptraj} selection mask to select the solute molecule (e.g.
+\inlinecode solute :1-5 if the first five residues are the solute).
 All other molecules will be treated as solvent.
@@
 \end{lstlisting}
-Note that the central atom of the \inlinecode{rigidatoms} selection goes first.
-
+Here, \inlinecode{solute \^{}1} defines residue 1 as the solute. 
+The \inlinecode{rigidatoms O1 H1 C1} keyword specifies atoms (e.g., in methanol solvent) to define the solvent's orientation, with O1 being the central atom.
 \subsection{GIST with salt-water mixtures}
@@
 	\caption{Biotin in the streptavidin binding pocket, with isosurfaces based on the GIST calculation on the holo structure. 
-	Isolevels shown are ($\pm$) 0.25, 0.5, 1.0, 1.5, 2.0, 2.5, and 3.0 \si{\kilo\calorie\per\mole\per\cubic\angstrom}, except for 
-	the positive energy, where they are divided by two to visually account for its overall lower density.
+	Isolevels shown are ($\pm$) 0.25, 0.5, 1.0, 1.5, 2.0, 2.5, and 3.0 \unit{\kilo\calorie\per\mole\per\cubic\angstrom}, except for the positive energy, where they are divided by two to visually account for its overall lower density.
 	The negative solvent energy is shown in green, positive solvent energy in red and negative entropy in blue.
-Only voxels within \SI{4}{\angstrom} of biotin are shown.}
+	Only voxels within \qty{4}{\angstrom} of biotin are shown.}
 	\label{fig-binding-pocket}
@@
 To visualize the considered region, we can create a OpenDX file where we set the considered voxels to 1 and all others to 0.
-Here, we show what this would look like for a radius of \SI{3.5}{\angstrom} around the heavy atoms of biotin.
+Here, we show what this would look like for a radius of \qty{3.5}{\angstrom} around the heavy atoms of biotin.
 
@@
 	return traj.atom_slice(traj.top.select(sel))
+
 ### Load GIST data
 compl = gt.gist.load_gist_file(
-	'complex/gist/gistoutlong.dat', 
+	'complex/gist/gist.dat', 
 	struct='complex/gist/gist_nowat.pdb')
@@
 	\includegraphics[width=1.0\linewidth]{streptavidin_bp_close.png}
-	\caption{Volume of streptavidin defined around \SI{3.5}{\angstrom} of biotin. }\label{streptavidin_volume}
+	\caption{Volume of streptavidin defined around \qty{3.5}{\angstrom} of biotin. }\label{streptavidin_volume}
 \end{figure}
@@
 As such, $E \approx H$ and $\dasolv \approx \Delta G_solv$.
-Similarly, the entropy columms are named dTS, rather than the more correct $T\Delta S$.
+Similarly, the entropy columns are named dTS, rather than the more correct $T\Delta S$, due to historical reasons.
 \subsection{Solvation Entropy}
@@
 \begin{equation}
-	\Delta E^\text{R}_{ww} = \sum_{k}^\text{voxels} \frac{E_{ww,k}^\text{corr}}{2}
+	\Delta E_{ww} = \sum_{k}^\text{voxels} \frac{E_{ww,k}^\text{corr}}{2}
 %	\Delta E^R_{ww} = n^R E^\text{bulk}_{ww} - E^\textit{R,corr}_{ww}
@@
 		\begin{itemize}
-			\item Simulation time: \SIrange{10}{20}{\nano\second} (\SI{100}{\nano\second})
+			\item Simulation time: \qtyrange{10}{20}{\nano\second} (\qty{100}{\nano\second})
 			\item Number of analyzed frames: \num{10000} (\num{100000})
-			\item Restraints: minimum \SI{10}{\kcalPerMolASqr} on solute heavy atoms. (typically \SI{100}{\kcalPerMolASqr})
+			\item Restraints: minimum \qty{10}{\kcalPerMolASqr} on solute heavy atoms. (typically \qty{100}{\kcalPerMolASqr})
 		\end{itemize}
@@
 \textbf{Grid Options} &\\
-gridcntr <xval> <yval> <zval> & Coordinates in \AA{} of the center of the grid. Default is 0.0, 0.0, 0.0. \\
-griddim <xval> <yval> <zval> & Grid dimensions in voxels along each coordinate axis. Default is 40, 40, 40.\\
+gridcntr <xval> <yval> <zval> & Coordinates in \AA{} of the center of the grid. Default is 0.0 0.0 0.0 \\
+griddim <xval> <yval> <zval> & Grid dimensions in voxels along each coordinate axis. Default is 40 40 40.\\
 gridspacn <val> & Grid spacing (linear dimension of each voxel) in 
-Angstroms. Default is \SI{0.5}{\angstrom}.\\
+Angstroms. Default is \qty{0.5}{\angstrom}.\\
 rmsfit <mask> & If specified, grid will be rotated and translated to follow the atoms selected by mask. \\
@@
 refdens <val> & Reference density of bulk solvent, used in computing 
-$g_O, g_H$, and the translational entropy. Default is 0.0334 molecules/\si{\cubic\angstrom}. \\
+$g_O, g_H$, and the translational entropy. Default is 0.0334 molecules/\unit{\cubic\angstrom}. \\
 temp <val> & Temperature of the input trajectory. \\
@@
 
-neighborcut <val> & Cutoff in \si{\angstrom} for determining solvent O-O neighbors. Default is \SI{3.5}{\angstrom}. \\
+neighborcut <val> & Cutoff in \unit{\angstrom} for determining solvent neighbors. Default is \qty{3.5}{\angstrom}, typical for water O-O. \\
 
@@
 pme & Use particle mesh Ewald for the non-bonded electrostatics calculation. The van der Waals energy will be calculated using a long-range correction for periodicity.\\
-\qquad cut <val> & Direct space cutoff for pme. Default is \SI{3.5}{\angstrom}.\\
+\qquad cut <val> & Direct space cutoff for pme. Default is \qty{8.0}{\angstrom}.\\
 \qquad dsumtol <val> & Direct sum tolerance used to determine Ewald coefficient. Default is 0.00001.\\
 \qquad ewcoeff <val> & Ewald coefficient in 1/Å. \\
-\qquad erfcdx <val> & Spacing to use for the ERFC splines. Default is \SI{0.0002}{\angstrom}. \\
+\qquad erfcdx <val> & Spacing to use for the ERFC splines. Default is \qty{0.0002}{\angstrom}. \\
 \qquad skinnb & Used to determine pairlist atoms (added to cut, so pairlist cutoff is cut + skinnb); included in order to maintain consistency with results from sander. \\
@@
 \qquad nopl & Do not use pair list for order calculation\\
-\qquad plcut <val> & Pair list cutoff for order calculation. Default is \SI{10}{\angstrom}\\
+\qquad plcut <val> & Pair list cutoff for order calculation. Default is \qty{10}{\angstrom}\\
 \bottomrule
@@
 dTSorient &  & First order orientational entropy. \\
-order &  & (if \inlinecode{doorder} was specified) Average Tetrahedral Order Parameter. \\
+order &  & (if \inlinecode{doorder} was specified) Average Tetrahedral Order Parameter \cite{chatterjee2008computational}. \\
 dipolex &  & x-component of the mean solvent dipole moment density \\
@@
 neighbor &  & Mean number of solvent molecules neighboring the solvent molecules found in this voxel. \\
-order & & Average tetrahedral order parameter \\
 

### 2025-05-23 manuscript/manuscript.tex

@@
 Since the density is computed from an explicit solvent MD simulation, GIST provides an accurate picture of solvation at an atomistic level, even for demanding systems such as drug binding sites in proteins.
-While the need for MD simulations comes at a large computational cost, GIST is still more efficient than end-state approaches such as TI \cite{Kirkwood1935-ti} or FEP \cite{Zwanzig1954-reweighting}, with applications to large systems such as serine proteases \cite{Kraml2019-gigist} or antibody fragments \cite{Waibl2021-gist-antibodies}.
+While all MD simulations require a large computational cost, GIST is more efficient than end-state approaches such as TI \cite{Kirkwood1935-ti} or FEP \cite{Zwanzig1954-reweighting}, with applications to large systems such as serine proteases \cite{Kraml2019-gigist} or antibody fragments \cite{Waibl2021-gist-antibodies}.
 On the other hand, the first-order approximation means that the solvation entropy is less accurately described than in TI or FEP \cite{Chen2021,Waibl2021-gist-salt}.
@@
 We additionally provide a Jupyter notebook \cite{Kluyver2016-jupyter,Granger2021-jupyter} with the presented analyses. 
-For the visualization of three-dimensional contributions we require molecular visualization software that can handle \software{OpenDX} files, such as \software{PyMOL} \cite{pymol} or \software{VMD} \cite{vmd}.
 Most of the visualisations and results in this tutorial can also be obtained using the \software{gistpp} program, which is available from \url{https://github.com/KurtzmanLab/Gist-Post-Processing} (code and documentation).
@@
 \end{lstlisting}
+
 If the solute is already centered at the origin (0, 0, 0), the grid center does not need to be specified.
 As such, the GIST calculation should usually be preceded by centering the solute molecule in the unit cell and reimaging to account for periodic boundary conditions.
-Often, the region of interest will be defined as the surrounding of a molecule, for example a ligand in the binding pocket or a whole protein. 
+Often, the region of interest will be defined as the surrounding of a molecule, for example a ligand in the binding pocket, or an entire protein.
 We recommend using GIST grids with a similar size to the simulation box, as the GIST calculation will usually not be the time limiting step and a region of interest can then be chosen freely in post-analysis.
@@
 GIST is implemented in \software{cpptraj} which requires the topology and the trajectory of the MD simulation to be analyzed.
-The command needed to run GIST in cpptraj can be written in an input file (e.g gist.in) for convenience.
+
+The command needed to run GIST in cpptraj can be written in an input file (e.g gist.in) for convenience and reproducibility.
 The basic command, assuming a grid centered at coordinates (x=\qty{10.5}{\angstrom}, y=\qty{20.1}{\angstrom}, z=\qty{30.1}{\angstrom}) and a grid size of $30\times30\times30$ voxels with \qty{0.5}{\angstrom} spacing, is:

### 2025-05-23 manuscript/manuscript.tex
Fix pandas citation
Fix my name
Since I am now married, and all...
Final finishes
1) Correct -TdS to actually show - not +
2) Fix overlong sentences poking out of coding sections

--- 2025-05-23 manuscript/manuscript.tex
@@
 
-\author[1\authfn{1}]{Valentin J Hoerschinger}
+\author[1\authfn{1}]{Valentin J Egger-Hoerschinger}
 \author[1\authfn{1}]{Franz Waibl}
@@
 \corr{thomas.kurtzman@lehman.cuny.edu}{TK}
-\orcid{Valentin J Hoerschinger}{0000-0002-4469-3238}
+\orcid{Valentin J Egger-Hoerschinger}{0000-0002-4469-3238}
 \orcid{Franz Waibl}{0000-0003-0527-0803}
@@
 	\item \software{numpy} (\url{https://numpy.org/}, $\geq$ 1.23.5) \cite{harris2020-numpy}
-	\item \software{pandas} (\url{https://pandas.pydata.org/}, $\geq$ 1.5.3) \cite{pandas-2023_8239932}
+	\item \software{pandas} (\url{https://pandas.pydata.org/}, $\geq$ 1.5.3) \cite{pandas-2023_8239932,mckinney-proc-scipy-2010}
 \end{itemize}
@@
 \end{table}
-\pagebreak %TODO: pagebreak
+
 \paragraph{Postprocessing and Analyzing GIST results}
@@
 
-\pagebreak
 \section{Tutorial}
@@
 import matplotlib.pyplot as plt
-# Load the GIST data, setting the reference water-water energy directly to -9.539 for TIP3P water
+# Load the GIST data, setting the reference water-water energy
 gist = load_gist_file(
@@
 \begin{lstlisting}[style=python]
-# Write out the column A_dens (the free solvation energy) to a dx file
+# Write the column A_dens (the free solvation energy) to a dx file
 gist.save_dx('A_dens','A.dx')
@@
 \begin{lstlisting}[style=cpptraj]
-# Load the topology ('parm') and coordinates of the complex after equilibration
+# Load complex topology and coordinates after equilibration
 parm complex/solvated.parm7
 trajin complex/equil/EQUIL-DONE.rst
-# Assuming biotin is residue 2 in the complex topology (selection ^2):
-# Remove biotin from the complex, save the topology and coordinates
+# Assuming biotin is residue 2 in the complex topology:
+# Remove biotin (^2) from the complex, save topology and coordinates
 strip ^2 parmout streptavidin/solvated.parm7
@@
 Due to the conceptual similarity with classical radial distribution functions, describing the density of distances between two particle types, this function is exposed to the user as the \inlinecode{rdf} method.
-
 \begin{lstlisting}[style=python]
@@
 import matplotlib.pyplot as plt
-# Load the GIST data, setting the reference water-water energy directly
+# Load the GIST data without Eww set
 gist = load_gist_file(
@@
 plt.gca().set(
-    xlabel='Distance [$\AA$]', ylabel='Avg. E$_{ww}$ per molecule [kcal/mol]', xlim=(1, 15))
+    xlabel='Distance [$\AA$]', 
+	ylabel='Avg. E$_{ww}$ per molecule [kcal/mol]', 
+	xlim=(1, 15))
 plt.show()
@@
     struct='solute-centered.pdb', eww_ref=-9.533)
-# Generate a new column with the energy without accounting for double counting
+# Calculate the total energy without double counting
 gist['E_norm'] = gist['Eww_norm'] * 2 + gist['Esw_norm']
@@
 trajin md-01.nc 1 last 100
-# Calculate the interaction energy of the complex (^1,2) and the monomers (^1 for streptavidin, ^2 for biotin)
+# Calculate the interaction energy of the complex (^1,2) 
+# and the monomers (^1 for streptavidin, ^2 for biotin)
 energy complex ^1,2 # etype pme
@@
 All authors were involved in reviewing and editing the original manuscript. \\  
-VJH and FW conceptualized the research and co-wrote the initial draft.\\
+VJEH and FW conceptualized the research and co-wrote the initial draft.\\
 VM contributed the original drafts of sections \ref{sec:theory} and \ref{sec:visualization}.\\
